### 2022-02-17 README.Rmd
Add test jittering code

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# bicla-R

<!-- badges: start -->
<!-- badges: end -->

The goal of bicla-R is to store code and (in the releases) data for estimating cycling potential and influencing policy.

# Input data

The key datasets are as follows:

- TRIPS_original.Rds, the original OD dataset

# Jittered trip data

### 2022-02-17 code/test-code/jittering.R

@@ -1,0 +1,10 @@
+# Aim: test out Jittering in study region
+remotes::install_github("atumworld/odr")
+library(tidyverse)
+
+od_all = readRDS("TRIPSmode_freguesias.Rds")
+zones = readRDS("FREGUESIASgeo.Rds")
+od_all_sf = od::od_to_sf(od_all, zones)
+plot(od_all_sf)
+
+osm_data_region = osmextract::oe_get_network(place = "Portugal", mode = "cycling", )

### 2022-02-18 code/test-code/jittering.R
Add jittering code

@@ -1,10 +1,55 @@
 # Aim: test out Jittering in study region
 remotes::install_github("atumworld/odr")
 library(tidyverse)
+library(tmap)
+tmap_mode("view")
 
 od_all = readRDS("TRIPSmode_freguesias.Rds")
 zones = readRDS("FREGUESIASgeo.Rds")
 od_all_sf = od::od_to_sf(od_all, zones)
-plot(od_all_sf)
+# plot(od_all_sf)
+od_all_sf_top = od_all_sf %>% 
+  filter(DICOFREor11 != DICOFREde11) %>% 
+  top_n(n = 100, wt = Total)
 
-osm_data_region = osmextract::oe_get_network(place = "Portugal", mode = "cycling", )
+# osm_data_national = osmextract::oe_get_network(place = "Portugal", mode = "cycling", force_download = TRUE)
+# nrow(osm_data_national) # 1 million rows, 1106127 (2022-02)
+# osm_data_region = osm_data_national[zones, ]
+# nrow(osm_data_region) # 142201
+# saveRDS(osm_data_region, "osm_data_region.Rds")
+# piggyback::pb_upload("osm_data_region.Rds")
+osm_data_region = readRDS("osm_data_region.Rds")
+osm_data_region_top = osm_data_region %>% 
+  sample_n(500)
+
+tm_shape(zones) +
+  tm_borders() +
+  tm_shape(osm_data_region_top) +
+  tm_lines("highway")
+
+summary(od_all$Total)
+set.seed(42)
+od_all_jittered = odr::odr_jitter(
+  od = od_all,
+  zones = zones,
+  subpoints = osm_data_region,
+  disaggregation_key = "Total",
+  disaggregation_threshold = 50
+)
+nrow(od_all_jittered) # 110008
+saveRDS(od_all_jittered, "od_all_jittered_50.Rds")
+piggyback::pb_upload("od_all_jittered_50.Rds")
+
+# case study zones
+small_netowrk = sf::st_transform(slopes::lisbon_road_network, 4326)
+lisbon_central = zones[small_netowrk, ]
+qtm(lisbon_central)
+
+od_jittered_central = od_all_jittered %>%
+  filter(DICOFREde11 %in% lisbon_central$Dicofre)
+qtm(od_jittered_central)
+
+od_jittered_central = od_all_jittered[lisbon_central, ]
+od_jittered_top = od_jittered_central %>% 
+  sample_n(size = 1000)
+qtm(od_jittered_top)

### 2022-02-22 README.Rmd

@@
   collapse = TRUE,
-  comment = "#>"
+  comment = "#>",
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE
 )
+library(sf)
 ```
@@
 
-# Jittered trip data
+```{r}
+od_all = readRDS("TRIPSmode_freguesias.Rds")
+zones = readRDS("FREGUESIASgeo.Rds")
+od_all_sf = od::od_to_sf(od_all, zones)
+od_all_jittered = readRDS("od_all_jittered_50.Rds")
+```
+
+The OD datasets, before and after jittering, are shown below.
+
+```{r jitteredoverview, out.width="50%", fig.show='hold'}
+plot(od_all_sf$geometry, lwd = 0.2)
+plot(od_all_jittered$geometry, lwd = 0.1)
+```

### 2022-04-18 README.Rmd
update initial data to 2020

@@
 
-# bicla-R
+# biclar
 
@@
 
-The goal of bicla-R is to store code and (in the releases) data for estimating cycling potential and influencing policy.
+The goal of biclar is to store code and (in the releases) data for estimating cycling potential and influencing policy.
 
@@
 
-- TRIPS_original.Rds, the original OD dataset
+- [CAOP 2020](https://www.dgterritorio.gov.pt/cartografia/cartografia-tematica/caop?language=en) - Official limits of Portuguese areas.
+- TRIPS_original.Rds, the original OD dataset, from INE (2018)
 

### 2022-04-26 README.Rmd
structure of biclar readme
populate with some details

--- 2022-02-22 README.Rmd
@@
 output: github_document
+bibliography: refs.bib
 ---
@@
 
-# bicla-R
+# biclar
 
 <!-- badges: start -->
+
 <!-- badges: end -->
 
-The goal of bicla-R is to store code and (in the releases) data for estimating cycling potential and influencing policy.
+The goal of `biclar` is to store code and (in the releases) data for estimating cycling potential and influencing policy.
+
+**biclar** is a tool for the design and assessment of different scenarios of the cycling network models in the Lisbon metropolitan area (LMA).
 
@@
 
-- TRIPS_original.Rds, the original OD dataset
+-   Trips dataset with Origin and Destination, at *Freguesia* level, disaggregated by transport mode, from @IMOB  
+-   [CAOP 2020](https://www.dgterritorio.gov.pt/cartografia/cartografia-tematica/caop?language=en) - Official limits of Portuguese areas.
+-   Road network from [OpenStreetMap](https://www.openstreetmap.org/#map=11/38.7053/-9.1585)  
+-   Main public transport interfaces at Lisbon Metro Region, provided by [Transportes Metropolitanos de Lisboa](https://www.tmlmobilidade.pt/)
+
+<!-- todo: include shape of transport hubs in releases -->
+
+# Cenarios for cycling uptake
+
+## Baseline
+
+The baseline scenario makes use of the 2018 mobility survey data in LMA.  
+We considered all trips between _Freguesias_.
+
+<!-- todo: include map of baseline scenario -->
+
+See vignette [baseline scenario]() to see how this was modeled.
+
+## ENMAC targets
+
+The National targets for cycling uptake were set to:
+
+-   4% of all trips should be made by bicycle by 2025
+-   10% of all trips should be made by bicycle by 2030
+
+Cycling trips should replace car trips directly.
+
+See vignette [ENMAC scenario]() to see how this was modeled.
+
+## Intermodal trips
+
+See vignette [Intermodal scenario]() to see how this was modeled.
+
+## E-bikes investment policy
+
+See vignette [E-bike scenario]() to see how this was modeled.
+
+# Methods
+
+## PCT - Propensity to Cycle Tool
+
+`biclar` uses the methods developed in [PCT.bike](https://pct.bike) [@Lovelace2017] for cycling uptake estimation and data visualization.
+
+## Jittering
+
+For the disagregation of OD pairs at _Freguesias_ level, we use [OD Jittering](https://github.com/atumworld/odrust) [@Jittering2022] method, which better suits walking and cycling trips modelling (shorter distances), instead of relying on centroids that concentrate all the trips between areas.
+
+The OD datasets, before and after jittering, are shown below.
 
@@
 
-The OD datasets, before and after jittering, are shown below.
-
 ```{r jitteredoverview, out.width="50%", fig.show='hold'}
@@
 ```
+
+## Cycling routes
+
+Use of [CyclingStreets.net](https://cyclinstreets.net) ([R package](https://rpackage.cyclestreets.net/)) for fast and quiet bike routes for baseline scenario.  
+For e-bike scenario, we developed a proper algorithm, considering the topography (and [`slopes`](https://docs.ropensci.org/slopes) package).
+
+## Intermodal trips
+
+We made use and developed a methodology that considers replacing long trips by bike + train or ferry trips (`name of the package`)
+
+<!-- todo: include Robin's reflection about multi-modal trips replacement -->
+
+## Estimation of socioeconomic benefits
+
+Health Economic Assessment Tool ([HEAT v5.0](https://www.heatwalkingcycling.org/#how_heat_works)) for walking and cycling by WHO.
+
+# Results
+
+## Cycling uptake in LMA and by Municipality
+
+## Comparision with the cycling network plans by Municipality
+
+# References {-}

### 2022-04-26 README.Rmd
links to scenarios articles

@@
 bibliography: refs.bib
+
 ---
@@
 
-See vignette [baseline scenario]() to see how this was modeled.
+See vignette [baseline scenario](articles/0_baseline_scenario.html) to see how this was modeled.
 
@@
 
-See vignette [ENMAC scenario]() to see how this was modeled.
+See vignette [ENMAC scenario](articles/1_enmac_scenario.html) to see how this was modeled.
 
@@
 
-See vignette [Intermodal scenario]() to see how this was modeled.
+See vignette [Intermodal scenario](articles/2_intermodal_scenario.html) to see how this was modeled.
 
@@
 
-See vignette [E-bike scenario]() to see how this was modeled.
+See vignette [E-bike scenario](articles/3_ebikes_scenario.html) to see how this was modeled.
 
@@
 
+<!-- figs not showing on pkgdown -->
+
 ## Cycling routes
@@
 
-We made use and developed a methodology that considers replacing long trips by bike + train or ferry trips (`name of the package`)
-
-<!-- todo: include Robin's reflection about multi-modal trips replacement -->
+We made use and developed a [methodology](https://github.com/npct/rail) that considers replacing long trips by bike + train or ferry trips.
 

### 2022-04-26 code/test-code/intermodal.R
prepare public transit intermodality code
add train and ferry data

@@ -1,0 +1,30 @@
+# Multi-modal intefaces for scenario 2
+
+
+# Read and clean data -------------------------------------------------------------------------
+
+library(tidyverse)
+library(sf)
+# train_stations = st_read("D:/GIS/TML/dados/Interfaces/estacoes_comboio.shp")
+# train_stations = train_stations %>% st_transform(crs = 3857) %>% select(ID, NAME, geometry)
+# 
+# mapview::mapview(train_stations)
+# 
+# saveRDS(train_stations, "train_stations.Rds")
+# piggyback::pb_upload("train_stations.Rds")
+
+train_stations = readRDS("train_stations.Rds")
+
+# ferry_stations = st_read("D:/GIS/TML/dados/Interfaces/estacoes_fluviais.shp")
+# ferry_stations = ferry_stations %>% st_transform(crs = 3857) %>% select(OBJECTID, POI_NAME, geometry)
+# 
+# mapview::mapview(ferry_stations)
+# 
+# ferry_stations = ferry_stations %>% filter(!ferry_stations$POI_NAME %in% c("Fluvial Tróia", "Fluvial Setúbal")) #remove the ones that are extra AML
+# 
+# names(ferry_stations) = names(train_stations)
+# 
+# saveRDS(ferry_stations, "ferry_stations.Rds")
+# piggyback::pb_upload("ferry_stations.Rds")
+
+ferry_stations = readRDS("ferry_stations.Rds")

### 2022-04-26 README.Rmd
Update readme with figures in right place

--- 2022-04-18 README.Rmd
@@
   comment = "#>",
+  fig.path = "man/figures/README-",
   echo = FALSE,
   message = FALSE,
-  warning = FALSE
+  warning = FALSE,
+  out.width = "100%"
 )
+```
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r, include = FALSE}
 library(sf)

### 2022-04-26 README.Rmd
add cycling plans in LMA - readme
maybe too heavy to show. pointo to map with url.

@@
 -   Main public transport interfaces at Lisbon Metro Region, provided by [Transportes Metropolitanos de Lisboa](https://www.tmlmobilidade.pt/)
-
-<!-- todo: include shape of transport hubs in releases -->
 
@@
 
-<!-- figs not showing on pkgdown -->
-
 ## Cycling routes
@@
 
+Compare the modeled cycling networks (segments overlapping) with expansion plans, by municipality.
+
+<!-- todo: this is too heavy -->
+
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+planned_network = sf::st_read("https://github.com/U-Shift/biclar/releases/download/0.0.1/CyclingPlans2021.gpkg", quiet = TRUE)
+
+existing_network = sf::st_read("https://github.com/U-Shift/biclar/releases/download/0.0.1/CyclingNetwork2021.gpkg", quiet = TRUE)
+
+greens2 = c("#1A7832", "#AFD4A0") #color pal
+
+library(tmap)
+tmap_mode("view")
+tm_layout(title = "Existing and planned cycling networkn in LMA", panel.show = T) +
+  tm_add_legend("fill",
+                labels = "Planned",
+                col = "#E3879E",
+                z = 3) +
+tm_shape(planned_network) +
+  tm_sf(lwd = 1.5,
+        #   color = "pink",
+        col = "#E3879E",
+        lty = "dashed",
+        alpha = 0.8 ,
+        legend.show = T,
+        title.col = "Planned") +
+tm_shape(existing_network) +
+  tm_sf(col = "Tipologia",
+           lwd = 2.5,
+           palette = greens2,
+        title.col = "Existing by typology",
+        popup.vars = c(
+      "Município" = "Municipio",
+      "Tipologia" = "Tipologia",
+      "Função" = "Funcao",      
+     "Sentido" = "Sentido", 
+     "Comprimento" = "Lenght", 
+     "ID" = "ID"
+    ),
+  )
+
+```
+
 # References {-}

### 2022-04-26 code/test-code/intermodal.R
some to show enmac4
baseline and enmac4 without jittering

@@ -28,3 +28,13 @@ train_stations = readRDS("train_stations.Rds")
 # piggyback::pb_upload("ferry_stations.Rds")
 
 ferry_stations = readRDS("ferry_stations.Rds")
+
+library(tmap)
+tmap_mode("view")
+
+tm_shape(train_stations) +
+  tm_sf(col = "blue", size = 0.3) +
+  tm_shape(ferry_stations) +
+  tm_sf(col = "red", size = 0.3)
+
+          

### 2022-04-27 README.Rmd
some to show enmac4
baseline and enmac4 without jittering
add libraries in 0

@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
+```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
 planned_network = sf::st_read("https://github.com/U-Shift/biclar/releases/download/0.0.1/CyclingPlans2021.gpkg", quiet = TRUE)
@@
 
+```{r echo=FALSE}
+knitr::include_graphics("man/figures/existingplanned.png")
+```
+
+We can view it in an [interactive map here](https://ushift.tecnico.ulisboa.pt/content/tml/RedeExistentePrevista.html).
+
 # References {-}

### 2022-06-03 code/test-code/batch_CycleSteets.R
test batch

@@ -1,0 +1,31 @@
+# Compute the routing for jittered od pairs, for fast and quiet routes
+
+# devtools::install_github("cyclestreets/cyclestreets-r") #update for batch routing function
+# see: https://rpackage.cyclestreets.net/reference/batch.html
+
+library(dplyr)
+library(sf)
+library(cyclestreets)
+
+## with a threshold of 100 jittered trips, resulting 57356 od pairs
+
+od_jittered_100 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/od_all_jittered_100.Rds"))
+
+od_jittered_test = od_jittered_100 %>% slice_sample(n = 100)
+plot(od_jittered_test$geometry, lwd = 0.1)
+
+routes_jittered = batch(od_jittered_test,
+                        name = "biclar test",
+                        strategies = "fastest", #or quietest
+                        minDistance = 200,
+                        maxDistance = 10000, #change here the assumed max euclidean distance
+                        filename = "biclar_test",
+                        includeJsonOutput = 1, #0 - only summary info like time and dist
+                        emailOnCompletion = "temospena@gmail.com",
+                        username = "temospena",
+                        password = Sys.getenv("CYCLESTREETS"),
+                        base_url = "https://api.cyclestreets.net/v2/batchroutes.createjob",
+                        id = "start"
+)
+
+plot(routes_jittered$geometry, lwd = 0.1)                       

### 2022-06-03 code/test-code/jittering.R

@@ -35,11 +35,16 @@ od_all_jittered = odjitter::jitter(
   zones = zones,
   subpoints = osm_data_region,
   disaggregation_key = "Total",
-  disaggregation_threshold = 50
+  disaggregation_threshold = 50 #maybe increase to 100 for the prototype?
 )
 nrow(od_all_jittered) # 110008
+# nrow(od_all_jittered) # 57356 - with threshold 100
 saveRDS(od_all_jittered, "od_all_jittered_50.Rds")
+# saveRDS(od_all_jittered, "od_all_jittered_100.Rds")
 piggyback::pb_upload("od_all_jittered_50.Rds")
+# piggyback::pb_upload("od_all_jittered_100.Rds")
+
+
 
 # case study zones
 small_netowrk = sf::st_transform(slopes::lisbon_road_network, 4326)

### 2022-06-03 code/test-code/batch_CycleSteets.R
batch script update
password and pat are not the same thing
#19

@@ -23,9 +23,10 @@ routes_jittered = batch(od_jittered_test,
                         includeJsonOutput = 1, #0 - only summary info like time and dist
                         emailOnCompletion = "temospena@gmail.com",
                         username = "temospena",
-                        password = Sys.getenv("CYCLESTREETS"),
+                        password = Sys.getenv("CYCLESTREETS_PW"), #pw for CS account
                         base_url = "https://api.cyclestreets.net/v2/batchroutes.createjob",
-                        id = "start"
+                        id = "start",
+                        pat = Sys.getenv("CYCLESTREETS") #API key from CS for this project
 )
 
 plot(routes_jittered$geometry, lwd = 0.1)                       

### 2022-06-03 code/test-code/batch_CycleSteets.R
hide someparameters
#19
still not working after 5min waiting for 10 desire lines

@@ -11,7 +11,7 @@ library(cyclestreets)
 
 od_jittered_100 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/od_all_jittered_100.Rds"))
 
-od_jittered_test = od_jittered_100 %>% slice_sample(n = 100)
+od_jittered_test = od_jittered_100 %>% slice_sample(n = 10)
 plot(od_jittered_test$geometry, lwd = 0.1)
 
 routes_jittered = batch(od_jittered_test,
@@ -21,12 +21,13 @@ routes_jittered = batch(od_jittered_test,
                         maxDistance = 10000, #change here the assumed max euclidean distance
                         filename = "biclar_test",
                         includeJsonOutput = 1, #0 - only summary info like time and dist
-                        emailOnCompletion = "temospena@gmail.com",
+                        # emailOnCompletion = "temospena@gmail.com",
                         username = "temospena",
                         password = Sys.getenv("CYCLESTREETS_PW"), #pw for CS account
-                        base_url = "https://api.cyclestreets.net/v2/batchroutes.createjob",
-                        id = "start",
+                        # base_url = "https://api.cyclestreets.net/v2/batchroutes.createjob",
+                        # id = 326, #number of job id 326
                         pat = Sys.getenv("CYCLESTREETS") #API key from CS for this project
 )
 
 plot(routes_jittered$geometry, lwd = 0.1)                       
+

### 2022-06-06 code/test-code/batch_CycleSteets.R
Update batch_CycleSteets.R

@@ -12,6 +12,7 @@ library(cyclestreets)
 od_jittered_100 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/od_all_jittered_100.Rds"))
 
 od_jittered_test = od_jittered_100 %>% slice_sample(n = 10)
+od_jittered_test$id = 1:nrow(od_jittered_test)
 plot(od_jittered_test$geometry, lwd = 0.1)
 
 routes_jittered = batch(od_jittered_test,
@@ -26,8 +27,17 @@ routes_jittered = batch(od_jittered_test,
                         password = Sys.getenv("CYCLESTREETS_PW"), #pw for CS account
                         # base_url = "https://api.cyclestreets.net/v2/batchroutes.createjob",
                         # id = 326, #number of job id 326
-                        pat = Sys.getenv("CYCLESTREETS") #API key from CS for this project
+                        # id = "id",
+                        pat = Sys.getenv("CYCLESTREETS"), #API key from CS for this project
+                        serverId = 1
 )
 
 plot(routes_jittered$geometry, lwd = 0.1)                       
 
+
+
+# try in cyclestreets online tool -------------------------------------------------------------
+
+st_write(od_jittered_test, "od_biclar_10.geojson", delete_dsn = TRUE)
+routes_jittered_online = read.csv("lisbon_test-data.csv")
+routes_jittered_online = st_as_sf(routes_jittered_online, st_geometry("route"))

### 2022-06-07 README.Rmd
typo enmac link
move things to ebike
imob validation and batch test

@@
 
-See vignette [ENMAC scenario](articles/1_enmac_scenario.html) to see how this was modeled.
+See vignette [ENMAC scenario](articles/1_emnac_scenario.html) to see how this was modeled.
 
@@
 Use of [CyclingStreets.net](https://cyclinstreets.net) ([R package](https://rpackage.cyclestreets.net/)) for fast and quiet bike routes for baseline scenario.
-For e-bike scenario, we developed a proper algorithm, considering the topography (and [`slopes`](https://docs.ropensci.org/slopes) package).
+
+For e-bike scenario, we developed a proper algorithm, considering the topography, and using  [`slopes`](https://docs.ropensci.org/slopes) package.
 

### 2022-06-07 code/test-code/batch_CycleSteets.R

@@ -40,4 +40,6 @@ plot(routes_jittered$geometry, lwd = 0.1)
 
 st_write(od_jittered_test, "od_biclar_10.geojson", delete_dsn = TRUE)
 routes_jittered_online = read.csv("lisbon_test-data.csv")
-routes_jittered_online = st_as_sf(routes_jittered_online, st_geometry("route"))
+cyclestreets::
+# routes_jittered_online = json2sf_cs(routes_jittered_online)
+# routes_jittered_online = st_as_sf(routes_jittered_online, st_geometry("route"))

### 2022-06-10 code/test-code/add_scenarioENMAC_rosa.R
enmac scenario tidy up script

@@ -1,36 +1,20 @@
 # scenarios applied to jittered routes
 
 routes_jittered_quietest500 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds"))
-routes_jittered_quietest500$id2 = 1:nrow(routes_jittered_quietest500)
+routes_jittered_quietest500$id2 = 1:nrow(routes_jittered_quietest500) #inique ID
 
 #cycling potential function
-od_top_enmac = routes_jittered_quietest500
+routes_enmac = routes_jittered_quietest500
 ENMAC4 = 0.04 # 4%
 ENMAC10 = 0.10 # 10#
-od_top_enmac$Bikeper = od_top_enmac$Bike / od_top_enmac$Total
+routes_enmac$Bikeper = routes_enmac$Bike / routes_enmac$Total
 
-od_top_enmac_aside = od_top_enmac %>% filter(Bikeper >= ENMAC10) #don't mess with the ones that already reach target
-od_top_enmac = od_top_enmac %>% filter(Bikeper < ENMAC10)
+routes_enmac$new_cyc4 = ifelse(routes_enmac$Bikeper >= ENMAC4, routes_enmac$Bike, ENMAC4 * routes_enmac$Total - routes_enmac$Bike)
+routes_enmac$new_cyc10 = ifelse(routes_enmac$Bikeper >= ENMAC10, routes_enmac$Bike, ENMAC10 * routes_enmac$Total - routes_enmac$Bike)
+routes_enmac$new_car4 = ifelse(routes_enmac$Bikeper >= ENMAC4, routes_enmac$Car + routes_enmac$CarP, (routes_enmac$Car + routes_enmac$CarP) - routes_enmac$new_cyc4)
+routes_enmac$new_car10 = ifelse(routes_enmac$Bikeper >= ENMAC10, routes_enmac$Car + routes_enmac$CarP, (routes_enmac$Car + routes_enmac$CarP) - routes_enmac$new_cyc10)
 
-od_top_enmac$new_cyc = ENMAC10 * od_top_enmac$Total - od_top_enmac$Bike
-od_top_enmac$new_carT = (od_top_enmac$Car + od_top_enmac$CarP) - od_top_enmac$new_cyc
-
-od_top_enmac_aside$new_cyc = od_top_enmac_aside$Bike
[…]

### 2022-06-10 code/test-code/batch_CycleSteets.R
batch both ways

@@ -12,9 +12,9 @@ library(tmap)
 ## with a threshold of 100 jittered trips, resulting 57356 od pairs
 
 # od_jittered_100 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/od_all_jittered_100.Rds"))
-# od_jittered_test = od_jittered_100
+od_jittered_test = od_jittered_100
   # slice_sample(n = 10)
-od_jittered_test = readRDS("od_all_jittered_500.Rds")
+# od_jittered_test = readRDS("od_all_jittered_500.Rds")
 
 od_jittered_test$distance = as.numeric(st_length(od_jittered_test))
 od_jittered_filter = od_jittered_test %>%
@@ -22,7 +22,7 @@ od_jittered_filter = od_jittered_test %>%
   filter(Total > 10)
 
 # Check what % of existing bike trips are captured
-sum(od_jittered_filter$Bike) / sum(od_jittered_test$Bike)
+sum(od_jittered_filter$Bike) / sum(od_jittered_test$Bike) #91.68% with jittered_100
 
 od_jittered_filter$id = 1:nrow(od_jittered_filter)
 
@@ -46,20 +46,23 @@ routes_jittered_quietest = route(
 saveRDS(routes_jittered_quietest, "routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds")
 
 ## end robin
+
+st_write(od_jittered_filter, "od_jittered_filter.geojson")
   
-samplejitter1 = od_jittered_filter %>% slice_head(n= 10000) #maximum allowed - job id 425 quiestest / 430 fastest
-samplejitter2 = od_jittered_filter %>% filter(id == c(10001:20000)) # job id 426 quiestest / 431 fastest
-samplejitter3 = od_jittered_filter %>% filter(id == c(20001:30000)) # job id 427 quiestest / 432 fastest
-samplejitter4 = od_jittered_filter %>% filter(id == c(30001:40000)) # job id 428 quiestest / 433 fastest
-samplejitter5 = od_jittered_filter %>% filter(id == c(40001:50000)) # job id 429 quiestest / 434 fastest
+samplejitter1 = od_jittered_filter %>% slice_head(n= 10000) #maximum allowed - job id 425 quiestest / 430 fastest / 438 fastest AND quietest, both diections
[…]

### 2022-06-10 R/data.R
Add pkg

@@ -1,0 +1,11 @@
+#' Example route network
+#'
+#' @aliases cents_sf
+#' @examples
+#' rnet
+#'
+#' @docType data
+#' @keywords datasets
+#' @name rnet_lisbon
+#' @format An sf object
+NULL

### 2022-06-10 R/rnet.R

@@ -1,19 +1,38 @@
-# 
-# rnet = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/rnet_quietest_threshold_500_max_9km_total_max_total_10.Rds"))
-# zones_lisbon = zones %>% 
-#   filter(Concelho == "LISBOA")
-# rnet = rnet[zones_lisbon, ]
-# # rnet = rnet %>% 
-# #   filter(Baseline > 50)
-# fs::file_size("data-raw/rnet_lisbon.geojson") # 931 kb
-# central_lisbon = zonebuilder::zb_zone(x = "Lisbon", n_circles = 1)
-# rnet = rnet[central_lisbon, , op = sf::st_within]
-# sf::write_sf(rnet, "data-raw/rnet_lisbon.geojson")
-
-
-tm_rnet = function(rnet, palette, min_width = 1, max_width = 9, lwd = names(rnet)[1], col = names(rnet)[1], max_features = 10000) {
-  tmap::tm_shape(rnet) +
-    tmap::tm_lines(id = NULL)
+#' Visualise route networks with tmap
+#'
+#' @param rnet 
+#' @param palette 
+#' @param lwd_multiplier The width of the largest line relative to the thinnest line
+#' @param scale The width of the widest line in the dataset 
+#' @param lwd 
+#' @param col 
+#' @param max_features 
+#'
+#' @return A tmap object
+#' @export
+#'
+#' @examples
+#' tm_rnet(rnet_lisbon, lwd = "ENMAC4")
+tm_rnet = function(
+    rnet,
+    palette = "mako",
+    lwd_multiplier = 4,
+    scale = 5,
+    lwd = names(rnet)[1],
+    col = names(rnet)[2],
+    max_features = 10000
+    ) {
+  rnet_names = names(rnet)
+  rnet_names = rnet_names[-grep(pattern = "geom|id", x = rnet_names)]
+  lwd_scaled = rnet[[lwd]] - min(rnet[[lwd]])
+  lwd_scaled = lwd_scaled / max(lwd_scaled)
+  lwd_scaled = lwd_scaled * (lwd_multiplier - 1)
+  lwd_scaled = lwd_scaled + 1
+  rnet$lwd = lwd_scaled
+  m = tmap::tm_shape(rnet) +
+    tmap::tm_lines(id = NULL, lwd = "lwd", scale = 5, popup.vars = rnet_names,
[…]

### 2022-06-10 code/test-code/add_scenarioENMAC_rosa.R

@@ -1,7 +1,6 @@
 # scenarios applied to jittered routes
 
 routes_jittered_quietest500 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds"))
-routes_jittered_quietest500$id2 = 1:nrow(routes_jittered_quietest500) #inique ID
 
 #cycling potential function
 routes_enmac = routes_jittered_quietest500
@@ -17,4 +16,5 @@ routes_enmac$new_car10 = ifelse(routes_enmac$Bikeper >= ENMAC10, routes_enmac$Ca
 #export
 routes_jittered_quietest500_ENMAC410 = routes_enmac
 saveRDS(routes_jittered_quietest500_ENMAC410, "routes_jittered_quietest500_ENMAC410.Rds")
+
 

### 2022-06-10 code/test-code/batch_CycleSteets.R

@@ -78,6 +78,7 @@ routes_jittered = batch(samplejitter5,
 
 # Baseline route network --------------------------------------------------
 
+# For the baseline
 rnet_baseline_raw = overline(routes_jittered_quietest, attrib = c("Bike", "quietness"), fun = c(mean, sum))
 rnet_baseline = rnet_baseline_raw %>% 
   transmute(Quietness = quietness_fn1, Baseline = round(Bike_fn2)) %>% 
@@ -85,13 +86,28 @@ rnet_baseline = rnet_baseline_raw %>%
 nrow(rnet_baseline) # 14k
 write_rds(rnet_baseline, "rnet_quietest_threshold_500_max_9km_total_max_total_10.Rds")
 
+# Generate ENMAC scenario rnet - see ad_scenarioENMAC_rosa.R file --------
+# file.edit("code/test-code/add_scenarioENMAC_rosa.R")
+routes_jittered_quietest500_ENMAC410 = readRDS("routes_jittered_quietest500_ENMAC410.Rds")
+rnet_enmac_raw = overline(
+  routes_jittered_quietest500_ENMAC410,
+  attrib = c("Bike", "quietness", "new_cyc4", "new_cyc10"),
+  fun = c("mean", "sum")
+  )
+names(rnet_enmac_raw)
+rnet_enmac_full = rnet_enmac_raw %>% 
+  transmute(Quietness = quietness_mean, Baseline = Bike_sum, ENMAC4 = new_cyc4_sum, ENMAC10 = new_cyc10_sum) %>%
+  mutate_if(is.numeric, round) 
+nrow(rnet_enmac_full) # 59 k
+sum(routes_jittered_quietest500_ENMAC410$new_cyc4) / sum(routes_jittered_quietest500_ENMAC410$Total)
+write_rds(rnet_enmac_full, "rnet_enmac_full.Rds")
+rnet_enmac_region = rnet_enmac_full %>% 
+  slice_max(order_by = ENMAC10, n = 10000)
+write_rds(rnet_enmac_region, "rnet_enmac_region_top_10000.Rds")
+
+
 # Create results for one municipality
 head(zones)
-
-
-tm_shape(rnet_baseline) +
-  tm_lines(lwd = "Baseline", col = "Quietness", palette = "-Reds", breaks = c(0, 50, 75, 100), scale = 5)
-
 
 # try in cyclestreets online tool -------------------------------------------------------------
 

### 2022-06-10 data-raw/rnet.R

@@ -1,0 +1,11 @@
+## code to prepare `rnet` dataset goes here
+library(tidyverse)
+# rnet = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/rnet_quietest_threshold_500_max_9km_total_max_total_10.Rds"))
+rnet = readRDS("rnet_enmac_region_top_10000.Rds")
+rnet = rnet %>% 
+ filter(Baseline > 10)
+central_lisbon = zonebuilder::zb_zone(x = "Lisbon", n_circles = 1)
+rnet_lisbon = rnet[central_lisbon, , op = sf::st_within]
+summary(rnet_lisbon)
+plot(rnet_lisbon)
+usethis::use_data(rnet_lisbon, overwrite = TRUE)

### 2022-06-10 R/rnet.R
Update with working tm_rnet

@@ -1,4 +1,8 @@
 #' Visualise route networks with tmap
+#' 
+#' This function is for visualising route networks.
+#' It was originally developed to show quietness, as defined by
+#' [CycleStreets](https://www.cyclestreets.net/help/journey/howitworks/)
 #'
 #' @param rnet 
 #' @param palette 
@@ -12,15 +16,17 @@
 #' @export
 #'
 #' @examples
-#' tm_rnet(rnet_lisbon, lwd = "ENMAC4")
+#' tm_rnet(rnet_lisbon, lwd = "ENMAC4", col = "Quietness")
 tm_rnet = function(
     rnet,
-    palette = "mako",
+    palette = "-temperature_diverging",
     lwd_multiplier = 4,
     scale = 5,
     lwd = names(rnet)[1],
     col = names(rnet)[2],
-    max_features = 10000
+    max_features = 10000,
+    breaks = c(0, 25, 50, 75, 100),
+    labels = c("0-24", "25-49", "50-75", "75+")
     ) {
   rnet_names = names(rnet)
   rnet_names = rnet_names[-grep(pattern = "geom|id", x = rnet_names)]
@@ -29,9 +35,20 @@ tm_rnet = function(
   lwd_scaled = lwd_scaled * (lwd_multiplier - 1)
   lwd_scaled = lwd_scaled + 1
   rnet$lwd = lwd_scaled
+  pal = cols4all::c4a(palette = palette, n = length(breaks) - 1)
+  names(pal) = labels
+  rnet$cols = cut(rnet[[col]], breaks = breaks, labels = labels)
+  
+  # if(packageVersion("tmap") < "4.0") {
   m = tmap::tm_shape(rnet) +
-    tmap::tm_lines(id = NULL, lwd = "lwd", scale = 5, popup.vars = rnet_names,
-                   palette = palette)
+    tmap::tm_lines(id = NULL, lwd = "lwd", scale = 5, popup.vars = rnet_names, col = "cols", palette = pal)
+  # } else {
+  #   # Note: not working
+  # m = tmap::tm_shape(rnet) +
+  #   tmap::tm_lines(lwd = "lwd", lwd.scale = 5,
+  #                  col.scale = palette)
+  # tmap::tmap_mode("view")
+  # }
   tmap::tmap_leaflet(m)
 }
 

### 2022-06-11 code/test-code/add_scenarioENMAC_rosa.R
Add fast network

@@ -1,9 +1,10 @@
 # scenarios applied to jittered routes
 
-routes_jittered_quietest500 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds"))
+# routes_jittered = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds"))
+routes_jittered = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds"))
 
 #cycling potential function
-routes_enmac = routes_jittered_quietest500
+routes_enmac = routes_jittered
 ENMAC4 = 0.04 # 4%
 ENMAC10 = 0.10 # 10#
 routes_enmac$Bikeper = routes_enmac$Bike / routes_enmac$Total
@@ -14,7 +15,7 @@ routes_enmac$new_car4 = ifelse(routes_enmac$Bikeper >= ENMAC4, routes_enmac$Car 
 routes_enmac$new_car10 = ifelse(routes_enmac$Bikeper >= ENMAC10, routes_enmac$Car + routes_enmac$CarP, (routes_enmac$Car + routes_enmac$CarP) - routes_enmac$new_cyc10)
 
 #export
-routes_jittered_quietest500_ENMAC410 = routes_enmac
-saveRDS(routes_jittered_quietest500_ENMAC410, "routes_jittered_quietest500_ENMAC410.Rds")
+routes_jittered_ENMAC410 = routes_enmac
+saveRDS(routes_jittered_ENMAC410, "routes_jittered_ENMAC410_fastest_500.Rds")
 
 

### 2022-06-11 code/test-code/batch_CycleSteets.R

@@ -12,9 +12,9 @@ library(tmap)
 ## with a threshold of 100 jittered trips, resulting 57356 od pairs
 
 # od_jittered_100 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/od_all_jittered_100.Rds"))
-od_jittered_test = od_jittered_100
+# od_jittered_test = od_jittered_100
   # slice_sample(n = 10)
-# od_jittered_test = readRDS("od_all_jittered_500.Rds")
+od_jittered_test = readRDS("od_all_jittered_500.Rds")
 
 od_jittered_test$distance = as.numeric(st_length(od_jittered_test))
 od_jittered_filter = od_jittered_test %>%
@@ -35,15 +35,23 @@ routes_jittered_test = route(
   plan = "quietest"
 )
 
-plot(routes_jittered_test$geometry)                       
-
 routes_jittered_quietest = route(
   l = od_jittered_filter,
   route_fun = journey,
   plan = "quietest"
 )
 
+routes_jittered_fastest = route(
+  l = od_jittered_filter,
+  route_fun = journey,
+  plan = "fastest"
+)
+
+saveRDS(routes_jittered_fastest, "routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds")
 saveRDS(routes_jittered_quietest, "routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds")
+system("gh release list")
+system("gh release upload 0.0.1 routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds")
+system("gh release upload 0.0.1 routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds")
 
 ## end robin
 
@@ -78,19 +86,20 @@ routes_jittered = batch(samplejitter5,
 
 # Baseline route network --------------------------------------------------
 
-# For the baseline
-rnet_baseline_raw = overline(routes_jittered_quietest, attrib = c("Bike", "quietness"), fun = c(mean, sum))
-rnet_baseline = rnet_baseline_raw %>% 
-  transmute(Quietness = quietness_fn1, Baseline = round(Bike_fn2)) %>% 
-  filter(Baseline > 10)
-nrow(rnet_baseline) # 14k
[…]

### 2022-06-11 code/test-code/batch_CycleSteets.R
Update map

@@ -52,6 +52,12 @@ saveRDS(routes_jittered_quietest, "routes_jittered_quietest_threshold_500_max_9k
 system("gh release list")
 system("gh release upload 0.0.1 routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds")
 system("gh release upload 0.0.1 routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds")
+gh_release_upload = function(file, tag = "0.0.1") {
+  msg = glue::glue("gh release upload {tag} {file}")
+  message("Running this command:\n", msg)
+  system(msg)
+}
+
 
 ## end robin
 
@@ -117,10 +123,15 @@ rnet_enmac_region = rnet_enmac_full %>%
 # write_rds(rnet_enmac_region, "rnet_enmac_region_fastest_top_20000.Rds")
 
 # Visualise the results ------
-rnet_enmac_region = readRDS("rnet_enmac_region_fastest_top_20000.Rds")
-m = tm_rnet(rnet_enmac_region, lwd = "ENMAC10", col = "Quietness", palette = "johnson")
-tmap_save(m, "m_rnet_enmac_region_fastest_top_20000.html")
-htmlwidgets::saveWidget(m, "m_rnet_enmac_region_fastest_top_20000.html")
+rnet_enmac_region = readRDS("rnet_enmac_region_top_20000.Rds")
+# rnet_enmac_region = readRDS("rnet_enmac_region_fastest_top_20000.Rds")
+m = tm_rnet(rnet_enmac_region %>% slice_max(ENMAC4, n = 1000), lwd = "ENMAC10", col = "Quietness", palette = "johnson")
+m = tm_rnet(rnet_enmac_region, lwd = "ENMAC10", col = "Quietness", palette = "-mako", scale = 30)
+# htmlwidgets::saveWidget(m, "m_rnet_enmac_region_fastest_top_20000.html")
+htmlwidgets::saveWidget(m, "pkgdown/assets/m_rnet_enmac_region_fastest.html")
+gh_release_upload("m_rnet_enmac_region_fastest_top_20000.html")
+gh_release_upload("m_rnet_enmac_region_quietest_top_20000.html")
+
 
 # Create results for one municipality
 head(zones)

### 2022-06-11 R/rnet.R
Add working scale argument

@@ -17,6 +17,7 @@
 #'
 #' @examples
 #' tm_rnet(rnet_lisbon, lwd = "ENMAC4", col = "Quietness")
+#' tm_rnet(rnet_lisbon, lwd = "ENMAC4", col = "Quietness", scale = 30)
 tm_rnet = function(
     rnet,
     palette = "-temperature_diverging",
@@ -41,7 +42,7 @@ tm_rnet = function(
   
   # if(packageVersion("tmap") < "4.0") {
   m = tmap::tm_shape(rnet) +
-    tmap::tm_lines(id = NULL, lwd = "lwd", scale = 5, popup.vars = rnet_names, col = "cols", palette = pal)
+    tmap::tm_lines(id = NULL, lwd = "lwd", scale = scale, popup.vars = rnet_names, col = "cols", palette = pal)
   # } else {
   #   # Note: not working
   # m = tmap::tm_shape(rnet) +

### 2022-06-11 R/rnet.R
interactive results Readme

@@ -11,6 +11,7 @@
 #' @param lwd 
 #' @param col 
 #' @param max_features 
+#' @param title.col
 #'
 #' @return A tmap object
 #' @export
@@ -27,7 +28,8 @@ tm_rnet = function(
     col = names(rnet)[2],
     max_features = 10000,
     breaks = c(0, 25, 50, 75, 100),
-    labels = c("0-24", "25-49", "50-75", "75+")
+    labels = c("0-24", "25-49", "50-75", "75+"),
+    title.col = "Quietness"
     ) {
   rnet_names = names(rnet)
   rnet_names = rnet_names[-grep(pattern = "geom|id", x = rnet_names)]
@@ -42,7 +44,14 @@ tm_rnet = function(
   
   # if(packageVersion("tmap") < "4.0") {
   m = tmap::tm_shape(rnet) +
-    tmap::tm_lines(id = NULL, lwd = "lwd", scale = scale, popup.vars = rnet_names, col = "cols", palette = pal)
+    tmap::tm_lines(id = NULL,
+                   lwd = "lwd",
+                   scale = scale,
+                   popup.vars = rnet_names,
+                   col = "cols",
+                   palette = pal,
+                   title.col = title.col
+                   )
   # } else {
   #   # Note: not working
   # m = tmap::tm_shape(rnet) +

### 2022-06-11 R/rnet.R
Multi

@@ -16,8 +16,9 @@
 #' @export
 #'
 #' @examples
-#' tm_rnet(rnet_lisbon, lwd = "ENMAC4", col = "Quietness")
-#' tm_rnet(rnet_lisbon, lwd = "ENMAC4", col = "Quietness", scale = 30)
+#' rnet = rnet_lisbon
+#' tm_rnet(rnet, lwd = "ENMAC4", col = "Quietness")
+#' tm_rnet(rnet, lwd = "ENMAC4", col = "Quietness", scale = 30)
 tm_rnet = function(
     rnet,
     palette = "-temperature_diverging",
@@ -52,5 +53,42 @@ tm_rnet = function(
   # }
   tmap::tmap_leaflet(m)
 }
+#' @export
+#' @aliases tm_rnet
+tm_rnet_multi = function(
+    rnet,
+    palette = "-temperature_diverging",
+    lwd_multiplier = 4,
+    scale = 5,
+    lwd = names(rnet)[1],
+    col = names(rnet)[2],
+    max_features = 10000,
+    breaks = c(0, 25, 50, 75, 100),
+    labels = c("0-24", "25-49", "50-75", "75+")
+) {
+  rnet_names = names(rnet)
+  rnet_names = rnet_names[-grep(pattern = "geom|id", x = rnet_names)]
+  lwd_scaled = rnet[[lwd]] - min(rnet[[lwd]])
+  lwd_scaled = lwd_scaled / max(lwd_scaled)
+  lwd_scaled = lwd_scaled * (lwd_multiplier - 1)
+  lwd_scaled = lwd_scaled + 1
+  rnet$lwd = lwd_scaled
+  pal = cols4all::c4a(palette = palette, n = length(breaks) - 1)
+  names(pal) = labels
+  rnet$cols = cut(rnet[[col]], breaks = breaks, labels = labels)
+  
+  # if(packageVersion("tmap") < "4.0") {
+  m = tmap::tm_shape(rnet) +
+    tmap::tm_lines(id = NULL, lwd = "lwd", scale = scale, popup.vars = rnet_names, col = "cols", palette = pal)
+  # } else {
+  #   # Note: not working
+  # m = tmap::tm_shape(rnet) +
+  #   tmap::tm_lines(lwd = "lwd", lwd.scale = 5,
+  #                  col.scale = palette)
+  # tmap::tmap_mode("view")
+  # }
+  tmap::tmap_leaflet(m)
+}
+
 
 # tm_rnet(rnet)

### 2022-06-12 code/test-code/export18maps.R
protoptype fill

@@ -1,40 +1,43 @@
 ## export separeted html map for each 18 municipalities
 
+library(dplyr)
+library(sf)
+library(biclar)
+
 MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))
 MUNICIPIOS = MUNICIPIOSgeo$Concelho
 
+REDEregion = readRDS("rnet_enmac_quietest_full.Rds")
+# REDE = readRDS("rnet_enmac_fastest_full.Rds")
 
-REDE = #SET ROUTES FILE HERE
+scenario = "baseline" #baseline, enmac4, enmac10
+route = "quiet" #quiet, fast
 
 for(i in 1:length(MUNICIPIOS)){
   mun = MUNICIPIOS[i]
-  BUFFER = MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(500)
-  REDE = REDE %>% sf::st_crop(BUFFER) #%>% filter(Concelho == mun) 
+  BUFFER = MUNICIPIOSgeo %>% filter(Concelho == mun) %>% st_buffer(500)
+  REDE = REDEregion %>% st_filter(BUFFER)
+  REDE = REDE %>% filter(Baseline >= quantile(REDE$Baseline, 0.60)) #try other slices
   
   #make and export 18 maps
-  m = tm_layout(title = mun, panel.show = T) +
-    biclar::tm_rnet(REDE)
-    # REPLACE HERE THE TMAP FUNCION FOR ZOOM AT MUNICIPALITIES
-    #
-    # tm_shape(REDE) +
-    # tm_sf(col = "Tipologia",
-    #       lwd = 2.5,
-    #       palette = greens2,
-    #       title.col = "Rede Existente por tipologia",
-    #       popup.vars = c(
-    #         "Município" = "Municipio",
-    #         "Tipologia" = "Tipologia",
-    #         "Função" = "Funcao",      
-    #         "Sentido" = "Sentido", 
-    #         "Comprimento" = "Lenght", 
-    #         "ID" = "ID"
-    #       ),
-    # )  
+  m = tm_rnet(
+    REDE,
+    lwd = "ENMAC10", #Baseline, ENMAC4, ENMAC10
+    col = "Quietness",
[…]

### 2022-06-12 code/test-code/vizualize_maps.R

@@ -1,4 +1,7 @@
 # Visualise the results - experiments ------
+
+# remotes::install_github("u-shift/biclar")
+
 library(tmap)
 library(biclar)
 rnet_enmac_region = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/rnet_enmac_region_fastest_top_20000.Rds"))

### 2022-06-12 code/test-code/export18maps.R
quiet scenarios and prototype
and knit README

@@ -8,16 +8,23 @@ MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download
 MUNICIPIOS = MUNICIPIOSgeo$Concelho
 
 REDEregion = readRDS("rnet_enmac_quietest_full.Rds")
-# REDE = readRDS("rnet_enmac_fastest_full.Rds")
+# REDEregion = readRDS("rnet_enmac_fastest_full.Rds")
 
-scenario = "baseline" #baseline, enmac4, enmac10
+scenario = "enmac10" #baseline, enmac4, enmac10
 route = "quiet" #quiet, fast
+
+#Quiet, Baseline - DONE
+#Quiet, ENMAC4 - DONE
+#Quiet, ENMAC10 - DONE
+#Fast, Baseline
+#Fast, ENMAC4
+#Fast, ENMAC10
 
 for(i in 1:length(MUNICIPIOS)){
   mun = MUNICIPIOS[i]
   BUFFER = MUNICIPIOSgeo %>% filter(Concelho == mun) %>% st_buffer(500)
   REDE = REDEregion %>% st_filter(BUFFER)
-  REDE = REDE %>% filter(Baseline >= quantile(REDE$Baseline, 0.60)) #try other slices
+  REDE = REDE %>% filter(ENMAC10 >= quantile(REDE$ENMAC10, 0.60)) #try other slices, change here scenario
   
   #make and export 18 maps
   m = tm_rnet(
@@ -34,10 +41,5 @@ for(i in 1:length(MUNICIPIOS)){
 
 }
 
-#Quiet, Baseline - DONE
-#Quiet, ENMAC4
-#Quiet, ENMAC10
-#Fast, Baseline
-#Fast, ENMAC4
-#Fast, ENMAC10
+
 

### 2022-06-12 code/test-code/export18maps.R
fast routes 18export

@@ -7,33 +7,33 @@ library(biclar)
 MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))
 MUNICIPIOS = MUNICIPIOSgeo$Concelho
 
-REDEregion = readRDS("rnet_enmac_quietest_full.Rds")
-# REDEregion = readRDS("rnet_enmac_fastest_full.Rds")
+# REDEregion = readRDS("rnet_enmac_quietest_full.Rds") #quiet
+REDEregion = readRDS("rnet_enmac_fastest_full.Rds") #fast
 
 scenario = "enmac10" #baseline, enmac4, enmac10
-route = "quiet" #quiet, fast
+route = "fast" #quiet, fast
 
 #Quiet, Baseline - DONE
 #Quiet, ENMAC4 - DONE
 #Quiet, ENMAC10 - DONE
-#Fast, Baseline
-#Fast, ENMAC4
+#Fast, Baseline - DONE
+#Fast, ENMAC4 - DONE
 #Fast, ENMAC10
 
 for(i in 1:length(MUNICIPIOS)){
-  mun = MUNICIPIOS[i]
-  BUFFER = MUNICIPIOSgeo %>% filter(Concelho == mun) %>% st_buffer(500)
+  mun = MUNICIPIOS[i] #test with Lisbon 6
+  BUFFER = MUNICIPIOSgeo %>% filter(Concelho == mun) %>% st_buffer(500) #maybe reduce to 200? 
   REDE = REDEregion %>% st_filter(BUFFER)
-  REDE = REDE %>% filter(ENMAC10 >= quantile(REDE$ENMAC10, 0.60)) #try other slices, change here scenario
+  REDE = REDE %>% filter(ENMAC10 >= quantile(REDE$ENMAC10, 0.70)) #0.6 for quiet, 0.7 for fast, change here scenario
   
   #make and export 18 maps
   m = tm_rnet(
     REDE,
     lwd = "ENMAC10", #Baseline, ENMAC4, ENMAC10
     col = "Quietness",
-    palette = "-mako", # "mako", "reds" - reds for fastest, mako for quietest
-    scale = 12,
-    lwd_multiplier = 12 #15 used in region
+    palette = "-linear_yl_rd_bk", # "mako", "linear_yl_rd_bk" - linear_yl_rd_bk for fastest, mako for quietest (tried reds, rocket, burg)
[…]

### 2022-06-12 code/test-code/export18maps.R
vfxira

@@ -42,4 +42,4 @@ for(i in 1:length(MUNICIPIOS)){
 }
 
 
-
+# Manually open «vila franca xira html, save as, vfxira, and delete the previous ones

### 2022-06-12 code/test-code/export18maps.R
vfxira
it was stactic before

@@ -20,11 +20,11 @@ route = "fast" #quiet, fast
 #Fast, ENMAC4 - DONE
 #Fast, ENMAC10
 
-for(i in 1:length(MUNICIPIOS)){
+for(i in 1:(length(MUNICIPIOS)-1)){
   mun = MUNICIPIOS[i] #test with Lisbon 6
   BUFFER = MUNICIPIOSgeo %>% filter(Concelho == mun) %>% st_buffer(500) #maybe reduce to 200? 
   REDE = REDEregion %>% st_filter(BUFFER)
-  REDE = REDE %>% filter(ENMAC10 >= quantile(REDE$ENMAC10, 0.70)) #0.6 for quiet, 0.7 for fast, change here scenario
+  REDE = REDE %>% filter(ENMAC10 >= quantile(REDE$ENMAC10, 0.70)) #0.6 for quiet, 0.7 for fast, change here scenario. (Barreiro should be 0.65)
   
   #make and export 18 maps
   m = tm_rnet(
@@ -42,4 +42,28 @@ for(i in 1:length(MUNICIPIOS)){
 }
 
 
-# Manually open «vila franca xira html, save as, vfxira, and delete the previous ones
+## Manually save  vila franca xira , as, vfxira ######
+
+# REDEregion = readRDS("rnet_enmac_quietest_full.Rds") #quiet
+REDEregion = readRDS("rnet_enmac_fastest_full.Rds") #fast
+
+scenario = "baseline" #baseline, enmac4, enmac10
+route = "fast" #quiet, fast
+
+mun = MUNICIPIOS[18] #Vila Franca de Xira
+BUFFER = MUNICIPIOSgeo %>% filter(Concelho == mun) %>% st_buffer(500) #maybe reduce to 200? 
+REDE = REDEregion %>% st_filter(BUFFER)
+REDE = REDE %>% filter(Baseline >= quantile(REDE$Baseline, 0.70)) #0.6 for quiet, 0.7 for fast, change here scenario. (Barreiro should be 0.65)
+
+#make and export 18 maps
+m = tm_rnet(
+  REDE,
+  lwd = "Baseline", #Baseline, ENMAC4, ENMAC10
+  col = "Quietness",
+  palette = "-linear_yl_rd_bk", # "mako", "linear_yl_rd_bk" - linear_yl_rd_bk for fastest, mako for quietest (tried reds, rocket, burg)
+  scale = 15,
[…]

### 2022-06-12 code/test-code/export18maps.R
update vfxira baseline quiet
the previous one was not vfx

@@ -44,23 +44,23 @@ for(i in 1:(length(MUNICIPIOS)-1)){
 
 ## Manually save  vila franca xira , as, vfxira ######
 
-# REDEregion = readRDS("rnet_enmac_quietest_full.Rds") #quiet
-REDEregion = readRDS("rnet_enmac_fastest_full.Rds") #fast
+REDEregion = readRDS("rnet_enmac_quietest_full.Rds") #quiet
+# REDEregion = readRDS("rnet_enmac_fastest_full.Rds") #fast
 
 scenario = "baseline" #baseline, enmac4, enmac10
-route = "fast" #quiet, fast
+route = "quiet" #quiet, fast
 
 mun = MUNICIPIOS[18] #Vila Franca de Xira
 BUFFER = MUNICIPIOSgeo %>% filter(Concelho == mun) %>% st_buffer(500) #maybe reduce to 200? 
 REDE = REDEregion %>% st_filter(BUFFER)
-REDE = REDE %>% filter(Baseline >= quantile(REDE$Baseline, 0.70)) #0.6 for quiet, 0.7 for fast, change here scenario. (Barreiro should be 0.65)
+REDE = REDE %>% filter(Baseline >= quantile(REDE$Baseline, 0.60)) #0.6 for quiet, 0.7 for fast, change here scenario. (Barreiro should be 0.65)
 
 #make and export 18 maps
 m = tm_rnet(
   REDE,
   lwd = "Baseline", #Baseline, ENMAC4, ENMAC10
   col = "Quietness",
-  palette = "-linear_yl_rd_bk", # "mako", "linear_yl_rd_bk" - linear_yl_rd_bk for fastest, mako for quietest (tried reds, rocket, burg)
+  palette = "-mako", # "mako", "linear_yl_rd_bk" - linear_yl_rd_bk for fastest, mako for quietest (tried reds, rocket, burg)
   scale = 15,
   lwd_multiplier = 15 #15 used in region and fast, 12 for quiet
 )

### 2022-06-14 code/test-code/batch_CycleSteets.R
Code for #23

@@ -41,14 +41,22 @@ routes_jittered_quietest = route(
   plan = "quietest"
 )
 
+routes_jittered_balanced = route(
+  l = od_jittered_filter,
+  route_fun = journey,
+  plan = "balanced"
+)
+
+
 routes_jittered_fastest = route(
   l = od_jittered_filter,
   route_fun = journey,
   plan = "fastest"
 )
 
-saveRDS(routes_jittered_fastest, "routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds")
 saveRDS(routes_jittered_quietest, "routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds")
+saveRDS(routes_jittered_balanced, "routes_jittered_balanced_threshold_500_max_9km_total_max_total_10.Rds")
+saveRDS(routes_jittered_fastest, "routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds")
 system("gh release list")
 system("gh release upload 0.0.1 routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds")
 system("gh release upload 0.0.1 routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds")

### 2022-06-15 R/rnet.R
Balanced routes, close #23

@@ -88,7 +88,9 @@ tm_rnet_multi = function(
   
   # if(packageVersion("tmap") < "4.0") {
   m = tmap::tm_shape(rnet) +
-    tmap::tm_lines(id = NULL, lwd = "lwd", scale = scale, popup.vars = rnet_names, col = "cols", palette = pal)
+    tmap::tm_lines(id = NULL, lwd = lwd, scale = scale, popup.vars = rnet_names, col = "cols", palette = pal)
+  tmap_mode("view")
+  m
   # } else {
   #   # Note: not working
   # m = tmap::tm_shape(rnet) +

### 2022-06-15 code/test-code/add_scenarioENMAC_rosa.R

@@ -1,7 +1,8 @@
 # scenarios applied to jittered routes
 
 # routes_jittered = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds"))
-routes_jittered = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds"))
+# routes_jittered = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds"))
+routes_jittered = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds"))
 
 
 #cycling potential function
@@ -17,6 +18,7 @@ routes_enmac$new_car10 = ifelse(routes_enmac$Bikeper >= ENMAC10, routes_enmac$Ca
 
 #export
 routes_jittered_ENMAC410 = routes_enmac
-saveRDS(routes_jittered_ENMAC410, "routes_jittered_ENMAC410_fastest_500.Rds")
+# saveRDS(routes_jittered_ENMAC410, "routes_jittered_ENMAC410_fastest_500.Rds")
+saveRDS(routes_jittered_ENMAC410, "routes_jittered_ENMAC410_balanced_500.Rds")
 
 

### 2022-06-15 code/test-code/batch_CycleSteets.R

@@ -9,6 +9,12 @@ library(cyclestreets)
 library(stplanr)
 library(tmap)
 
+gh_release_upload = function(file, tag = "0.0.1") {
+  msg = glue::glue("gh release upload {tag} {file}")
+  message("Running this command:\n", msg)
+  system(msg)
+}
+
 ## with a threshold of 100 jittered trips, resulting 57356 od pairs
 
 # od_jittered_100 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/od_all_jittered_100.Rds"))
@@ -60,12 +66,7 @@ saveRDS(routes_jittered_fastest, "routes_jittered_fastest_threshold_500_max_9km_
 system("gh release list")
 system("gh release upload 0.0.1 routes_jittered_quietest_threshold_500_max_9km_total_max_total_10.Rds")
 system("gh release upload 0.0.1 routes_jittered_fastest_threshold_500_max_9km_total_max_total_10.Rds")
-gh_release_upload = function(file, tag = "0.0.1") {
-  msg = glue::glue("gh release upload {tag} {file}")
-  message("Running this command:\n", msg)
-  system(msg)
-}
-
+gh_release_upload(file = "routes_jittered_balanced_threshold_500_max_9km_total_max_total_10.Rds")
 
 ## end robin
 
@@ -112,7 +113,8 @@ routes_jittered = batch(samplejitter5,
 # file.edit("code/test-code/add_scenarioENMAC_rosa.R")
 # routes_jittered_500_ENMAC410 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/routes_jittered_quietest500_ENMAC410.Rds"))
 # routes_jittered_500_ENMAC410 = readRDS("routes_jittered_quietest500_ENMAC410.Rds")
-routes_jittered_500_ENMAC410 = readRDS("routes_jittered_fastest500_ENMAC410.Rds")
+# routes_jittered_500_ENMAC410 = readRDS("routes_jittered_fastest500_ENMAC410.Rds")
+routes_jittered_500_ENMAC410 = readRDS("routes_jittered_ENMAC410_balanced_500.Rds")
 rnet_enmac_raw = overline(
   routes_jittered_500_ENMAC410,
[…]

### 2022-06-23 code/test-code/vizualize_maps.R
colorscale fast region

@@ -10,7 +10,7 @@ rnet_enmac_region = readRDS(url("https://github.com/U-Shift/biclar/releases/down
 m = tm_rnet(rnet_enmac_region,
         lwd = "ENMAC10", #Baseline, ENMAC4, ENMAC10
         col = "Quietness",
-        palette = "-mako", # "johnson", "mako", "burg", "reds" - reds for fastest, mako for quietest
+        palette = "-linear_yl_rd_bk", # "linear_yl_rd_bk" "johnson", "mako", "burg", "reds" - reds for fastest, mako for quietest
         scale = 15,
         lwd_multiplier = 15
         )

### 2022-07-17 README.Rmd
Update map
add funding TML

@@
 
+#### ENMAC Scenario
+
+See [here](region_enmac4_quiet.html) for full map.
+
+```{r}
+# knitr::include_url("m_rnet_enmac_region_fastest.html")
+```
+
+<iframe src="region_enmac4_quiet.html" width="800" height="400px" data-external="1">
+</iframe>
+
+See [here](articles/propotype.html) for results for each Municipality.
+
+#### E-bike Scenario
+
+<!-- todo -->
+
+#### Intermodality Scenario
+
+<!-- todo -->
+
 ## Comparision with the cycling network plans by Municipality
@@
 
+# Funding
+
+This project is funded by [TML - Transportes Metropolitanos de Lisboa](https://www.tmlmobilidade.pt/projetos-e-atividades/planos-e-estudos/rede-ciclavel-metropolitana-estudo-modelacao-e-ferramenta-de-apoio-ao-planeamento-e-decisao/).
+
 # References {-}

### 2022-09-02 README.Rmd
LMA typo

@@
 -   Road network from [OpenStreetMap](https://www.openstreetmap.org/#map=11/38.7053/-9.1585)  
--   Main public transport interfaces at Lisbon Metro Region, provided by [Transportes Metropolitanos de Lisboa](https://www.tmlmobilidade.pt/)
+-   Main public transport interfaces at Lisbon Metropolitan Area, provided by [Transportes Metropolitanos de Lisboa](https://www.tmlmobilidade.pt/)
 

### 2022-11-19 code/test-code/r5r_filter_overline_routes_CORRECT.R
re-run with correct distance filter for 1&2

@@ -6,6 +6,7 @@ library(stplanr)
 # for cenarios 1 ------------------------------------------------------------------------------
 
 ## LTS4
+routes_r5r_100jit_lts4__ferry_elev = readRDS("routes_r5r_100jit_lts4__ferry_elev_raw.Rds")
 # filter for conventional bike
 table(routes_r5r_100jit_lts4__ferry_elev$mode)
 # BICYCLE   FERRY 
@@ -25,11 +26,11 @@ routes_ferry4_filtered = routes_ferry4_filtered %>% filter(distance <= 5000) #fa
 routes_ferry4_filtered = routes_r5r_100jit_lts4__ferry_elev %>% filter(id %in% routes_ferry4_filtered$id)
 table(routes_ferry4_filtered$mode)
 # BICYCLE   FERRY 
-#   25493      89
-# 2.522k    7695
+#   25491      88
+# 2.522k    7626
 # 2.506k (2  pernas da mesma viagem, first & last mile) 
 
-# sum(routes_ferry4_filtered$Total[routes_ferry4_filtered$mode == "BICYCLE"])
+# sum(routes_ferry4_filtered$Total[routes_ferry4_filtered$mode == "FERRY"])
 
 
 # get potential
@@ -53,8 +54,6 @@ sum(routes_ferry4_filtered$new_cyc4) #89k
 sum(routes_ferry4_filtered$cyc10) #253k
 sum(routes_ferry4_filtered$new_cyc10) #240k
 
-#### CONTINUAR AQUI 15nov ####
-#erro nos cenarios 2 e 3
 
 rnet_ferry4_overline = routes_ferry4_filtered %>%
   stplanr::overline(c("Bike", "Total", "new_cyc4", "cyc4", "new_cyc10", "cyc10"),
@@ -190,8 +189,6 @@ sum(routes_ferry3_filtered$new_cyc4) #82k
 sum(routes_ferry3_filtered$cyc10) #235k
 sum(routes_ferry3_filtered$new_cyc10) #221k
 
-## CONTINUAR AQUI
-
 
 rnet_ferry3_overline = routes_ferry3_filtered %>%
   stplanr::overline(c("Bike", "Total", "new_cyc4", "cyc4", "new_cyc10", "cyc10"),
@@ -289,10 +286,10 @@ routes_ferry4_ebike_filtered = routes_ferry4_ebike_filtered %>% filter(distance 
[…]

### 2022-11-19 HEAT/code/haas_api_test0.R
HEAT api integration - tests

@@ -1,0 +1,63 @@
+# HEAT API tests
+
+# ucc_
+# The UI collects "scenario" details (aka use case criteria) (incl. geographic location, selected modes of transport,
+# type of assessment, etc.). These UI inputs are prefixed with ucc_ (for use case criteria).
+# In the UI all validity checks and warnings are carried out - here and only here at the present moment
+# (incl. maximum levels of active travel, minimum population size, etc.)
+
+
+# webapp_inputs - this is a RDS file which should be downloaded from the "Export page" of the UI.
+# It stores UI inputs with values. The values of this can be modified before they are sent to the API.
+# Please modify these with great care. Since the business logic is not a part of the model itself,
+# improper combination or values of variables (inputs) won't be detected, resulting in incorrect results.
+# The general rule is that any ucc_ UI inputs should not be modified manually
+# (since these shape the UI pages and the layout of input fields, and the corresponding collection of UI
+# inputs with values and the data structures used by the model). Any other values of inputs can be changed 
+# - such as numeric values. Still, constraints which are available in the UI (as warnings or errors)
+# should be respected to get reliable results from the model. Note that webapp_inputs as RDS has to
+# be downloaded from the UI corresponding to the model version used (at the present moment only heat_5_0 is available).
+# Please use "Inspect" tool in the web browser to get the names of the inputs.
+# raw_
+
+
+
+
+
+# 1304
+
+value10yrs = 395000000
+value10yrs / sum(HEAT13_bike$new_cyc4) 
+#4819.92 € / novo ciclista
+
+heat1304_webappinput = readRDS("HEAT/1304_webapp_input.rds")
+heat1304_genericdata = readRDS("HEAT/1304_generic_data_relevant_with_calc_params.rds")
+
+
+## Input and returned data as json
+
+# read RDS exported from UI
+webapp_input <- heat1304_webappinput
+
+# to get data prepared by "results"
+# input data as "json"
+# returned data as "json"
+resp <- 
+  httr::POST("https://api.heatwalkingcycling.org/apiv1/heat_5_0/results?output_format=json",
+             httr::set_cookies("user" = user_key),
+             body = jsonlite::toJSON(list("webapp_input" = webapp_input)),
+             encode = "multipart")
+
+# convert from "json" to R format
+from_json <- jsonlite::fromJSON(httr::content(resp, "text")) #parse?
+
+
+# convert from "jsonl-serialized" to R format
[…]

### 2022-11-19 HEAT/code/haas_prerun.R

@@ -1,0 +1,49 @@
+library(magrittr)
+library(devtools)
+
+source("D:/GIS/TML/HEAT_heatr_api/R/models_paths.R")
+source("D:/GIS/TML/HEAT_heatr_api/R/allowed_methods.R")
+source("D:/GIS/TML/HEAT_heatr_api/R/parsers.R")
+source("D:/GIS/TML/HEAT_heatr_api/R/serializer.R")
+
+user_key <- "1525a17bfa86007eb7788adcec47a4ea62414b75"
+log_file <- "./logs/plum.log"
+
+# register custom parser
+
+if(!("jsonlite_serialized" %in% plumber::registered_parsers())){
+  
+  plumber::register_parser("jsonlite_serialized",
+                           parser_jsonlite_serialized,
+                           fixed = "application/jsonl-serialized"
+  )
+}
+
+function(req) {
+  log_data <-
+    paste0(
+      as.character(Sys.time()), "--",
+      req[["REQUEST_METHOD"]], "--",
+      req[["PATH_INFO"]], "--",
+      req[["HTTP_USER_AGENT"]], "--",
+      req[["REMOTE_ADDR"]], "##",
+      names(req[["cookies"]])
+    )
+  
+  write(log_data, log_file, append = TRUE)
+  
+  if (req$PATH_INFO == "/apiv1/readme.html" ||
+      req$PATH_INFO == "/apiv1/favicon.ico") {
+    return(plumber::forward())
+  }
+  
+  if ("cookies" %in% names(req) &&
+      length(req[["cookies"]]) > 0 &&
+      "user" %in% names(req[["cookies"]]) &&
+      req[["cookies"]][["user"]] == user_key
+  ) {
+    return(plumber::forward())
+  }
+  
+  return(list(error = "No cookie provided!"))
+}

### 2022-11-19 HEAT/code/haas_reprex.R

@@ -1,0 +1,95 @@
+
+
+# read RDS exported from UI
+webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_webapp_input.rds"))
+genericdata = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_generic_data_relevant_with_calc_params.rds"))
+
+
+## 1. Input and returned data as json
+
+# to get data prepared by "results"
+# input data as "json"
+# returned data as "json"
+resp <- 
+  httr::POST("https://api.heatwalkingcycling.org/apiv1/heat_5_0/results?output_format=json",
+             httr::set_cookies("user" = Sys.getenv("HEAT_COOKIE")),
+             body = jsonlite::toJSON(list("webapp_input" = webapp_input)),
+             encode = "multipart")
+
+# convert from "json" to R format
+from_json <- jsonlite::fromJSON(httr::content(resp, "text")) #parse?
+str(from_json$results)
+
+# Check final result value
+res = as.numeric(gsub(" ","",from_json$results$moneytotaldisc)) #remove spaces and convert to to numeric
+res[26]-res[27]-res[28]+res[29]
+
+
+
+
+
+
+
+## 2. Input and returned data as jsonl-serialized
+
+# read RDS exported from UI
+webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_webapp_input.rds"))
+
+
+# to get data prepared by "results"
+# input data as "jsonl-serialized"
+# returned data as "jsonl-serialized"
+resp <- 
+  httr::POST("https://api.heatwalkingcycling.org/apiv1/heat_5_0/results?output_format=jsonl_serialized",
+             httr::set_cookies("user" = Sys.getenv("HEAT_COOKIE")),
+             body = jsonlite::serializeJSON(list("webapp_input" = webapp_input)),
+             httr::content_type("application/jsonl-serialized"),
+             encode = "multipart")
+
+# convert from "jsonl-serialized" to R format
+from_serialized_json <- jsonlite::unserializeJSON(httr::content(resp, "text"))
+str(from_serialized_json$results)
+
+# Check final result value
+res = as.numeric(gsub(" ","",from_serialized_json$results$moneytotaldisc)) #remove spaces and convert to to numeric
+res[26]-res[27]-res[28]+res[29]
+
+
+
+
[…]

### 2022-11-19 HEAT/code/haas_values_change.R

@@ -1,0 +1,52 @@
+# Assessment intro
+heat1304_webappinput[["ucc_country_location_id"]] = "2264397" #Portugal
+heat1304_webappinput[["ucc_location_id"]] = "2267057" # Lisboa
+heat1304_webappinput[["ucc_activemode"]] = "bike" # how to select more than 1? c("walk", "bike")
+heat1304_webappinput[["ucc_comparisoncases"]] = "twocases" #alternative "onecase"
+heat1304_webappinput[["ucc_yearref"]] = 2018 # reference year
+heat1304_webappinput[["ucc_yearcf"]] = 2024 # comparison year
+heat1304_webappinput[["ucc_assessmenttime"]] = 10 # over how many years should the impacts be calculated
+heat1304_webappinput[["ucc_pathway"]] = c("pa", "ap", "crash", "carbon") # which impact to be considered?
+heat1304_webappinput[["ucc_motormodebasic"]] = "car" # modes to be considered. opt in basic categories: c("car", "pt")
+heat1304_webappinput[["ucc_trafficcondition"]] = "euromeancongestionurban" # alternatives: "somecongestion", "heavycongestion", "euromeancongestionrural"
+
+# Active modes data (ref = Reference, cf = Comparison)
+heat1304_webappinput[["raw_activemode_bike_ref_source"]] = "model" # alternatives: "scenario" (Hypotetical scenario), "popsurvey" (Population survey), "intercept" (Intercept survey), "counts" (Count data), "model" (Modeled Data), "app" (App-based data)
+heat1304_webappinput[["raw_activemode_bike_ref_unit-label"]] = "modeshare" # alternatives: "min" (Minutes), "hrs" (Hours), "km" (Km), "trips" (Trips), "modeshare" (Mode Share) # other options to select after may show up, depending on the data source
+heat1304_webappinput[["raw_activemode_bike_ref_modesharepercent"]] = 0.52 # percentage of CYCLING trips in the reference scenario
+heat1304_webappinput[["raw_activemode_bike_ref_alltrips"]] = 2317371 # total travel volume of ALL MODES in the reference scenario
+heat1304_webappinput[["raw_activemode_bike_ref_alltripsunit"]] = "trips" # unit of the travel volume. alternatives: "duration" (in minutes), "distance" (in km)
+heat1304_webappinput[["raw_activemode_bike_ref_tripkm"]] = 2.7 # average length of CYCLING trips for reference scenario
[…]

### 2022-11-19 code/test-code/heat-prep2.R

@@ -1,0 +1,288 @@
+#### CAR ONLY TO COMPARE ESTIMATES ####
+
+library(dplyr)
+options(java.parameters = '-Xmx40G') 
+library(r5r)
+r5r_lts_intermodalityALL = setup_r5(data_path = "r5r/allmodes2/", elevation = "MINETTI") #to create new, delete network.dat in the folder. otherwise just load it
+departure_datetime = lubridate::force_tz(departure_datetime, "WET") #make it western european time
+
+
+
+routes_r5r_100jit_car = detailed_itineraries(
+  r5r_lts_intermodalityALL,
+  origins = od_jittered_100filter_OR,
+  destinations = od_jittered_100filter_DE,
+  mode = "CAR", #only car
+  mode_egress = "WALK",
+  departure_datetime = departure_datetime, 
+  max_trip_duration = 120L, #in minutes
+  shortest_path = TRUE, #FALSE? fastest or multiple alternatives?
+  all_to_all = FALSE,
+  n_threads = Inf,
+  verbose = FALSE,
+  progress = TRUE,
+  drop_geometry = TRUE,
+  output_dir = NULL
+)
+
+
+routes_r5r_100jit_car = routes_r5r_100jit_car %>% mutate(id = as.integer(from_id)) %>%
+  select(id, total_duration, total_distance, segment, mode, segment_duration, distance, route) %>%
+  left_join(od_jittered_100filter %>% st_drop_geometry(), by="id")
+
+names(routes_r5r_100jit_car)[7] = "distance"
+names(routes_r5r_100jit_car)[18] = "eucl_distance"
+
+saveRDS(routes_r5r_100jit_car, "routes_r5r_100jit_car_raw.Rds")
+
+summary(routes_r5r_100jit_car$total_distance)
+# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+# 15    3316    6748   10734   13805  104572 
+#doesnt make sence for the reality
+
+
+
+# População movel AML = 80.4%
+# Deslocacoes / pessoa / dia = 2.6 AML
+
+
+
+# 1.3
+
[…]

### 2022-11-19 code/test-code/r5r_intermodality.R
Update r5r_intermodality.R

@@ -409,3 +409,8 @@ saveRDS(routes_r5r_100jit_lts3__ferry_elev, "routes_r5r_100jit_lts3__ferry_elev_
 
 ###### STOP HERE #######
 # see r5r_filter_overline_routes.R for nex steps
+
+
+
+
+

### 2022-11-20 code/test-code/quietness-networks.R
run again match rnet join osm for 1&2

@@ -19,22 +19,22 @@ local_crs = "EPSG:3857"
 # rnet_allmodesNSub4_overline_morethan100_clean = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/rnet_allmodesNSub4_overline_morethan100_clean.Rds"))
 
 #enmac
-# rnet = rnet_ferry3_overline_morethan20_clean 
-# rnet = rnet_ferry3_overline_morethan100_clean 
-# rnet = rnet_ferry3_overline_morethan200_clean 
-# rnet = rnet_ferry4_overline_morethan20_clean 
-# rnet = rnet_ferry4_overline_morethan100_clean 
-# rnet = rnet_ferry4_overline_morethan200_clean 
+rnet = rnet_ferry3_overline_morethan20_clean
+# rnet = rnet_ferry3_overline_morethan100_clean #done
+# rnet = rnet_ferry3_overline_morethan200_clean #done
+# rnet = rnet_ferry4_overline_morethan20_clean #done
+# rnet = rnet_ferry4_overline_morethan100_clean #done
+# rnet = rnet_ferry4_overline_morethan200_clean #done
 # 
 # #ebikes
-# rnet = rnet_ferry3_ebike_overline_morethan50_clean 
-# rnet = rnet_ferry3_ebike_overline_morethan100_clean 
-# rnet = rnet_ferry3_ebike_overline_morethan200_clean 
-# rnet = rnet_ferry3_ebike_overline_morethan400_clean 
-# rnet = rnet_ferry4_ebike_overline_morethan50_clean 
-# rnet = rnet_ferry4_ebike_overline_morethan100_clean 
-# rnet = rnet_ferry4_ebike_overline_morethan200_clean 
-# rnet = rnet_ferry4_ebike_overline_morethan400_clean
+# rnet = rnet_ferry3_ebike_overline_morethan50_clean #done
+# rnet = rnet_ferry3_ebike_overline_morethan100_clean #done
+# rnet = rnet_ferry3_ebike_overline_morethan200_clean #done
+# rnet = rnet_ferry3_ebike_overline_morethan400_clean #done
+# rnet = rnet_ferry4_ebike_overline_morethan50_clean #done
+# rnet = rnet_ferry4_ebike_overline_morethan100_clean #done
[…]

### 2022-11-20 code/test-code/quietness-networks.R
finishing join osm

@@ -19,7 +19,7 @@ local_crs = "EPSG:3857"
 # rnet_allmodesNSub4_overline_morethan100_clean = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/rnet_allmodesNSub4_overline_morethan100_clean.Rds"))
 
 #enmac
-rnet = rnet_ferry3_overline_morethan20_clean
+# rnet = rnet_ferry3_overline_morethan20_clean #done
 # rnet = rnet_ferry3_overline_morethan100_clean #done
 # rnet = rnet_ferry3_overline_morethan200_clean #done
 # rnet = rnet_ferry4_overline_morethan20_clean #done
@@ -101,7 +101,7 @@ rnet_joined$slope = scales::label_percent(accuracy = 0.1, sufix = " %")(rnet_joi
 
 
 
-saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan20_clean_tags.Rds")
+# saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan20_clean_tags.Rds") #done
 # saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan100_clean_tags.Rds") #done
 # saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan200_clean_tags.Rds") #done
 # saveRDS(rnet_joined, "export2/rnet_ferry4_overline_morethan20_clean_tags.Rds") #done

### 2022-11-20 code/biclarwww/index.Rmd
biclar website

---
title: "biclaR"
output:
  html_document:
    theme:
      bg: "#FFFFFF"
      fg: "#000000"
      highlight: "#000000"
      primary: "#000000"
      base_font:
        google: "Prompt"
      code_font:
        google: "JetBrains Mono"
    # css: style.css
    includes:
      after_body: img/footer.html
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo=FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

# has_pub_key = nchar(Sys.getenv("MAPBOX_PUBLIC_TOKEN")) > 5
# knitr::opts_chunk$set(eval = has_pub_key)

library(sf)
library(stringr)
library(dplyr)
library(leaflet)
# library(mapboxapi)
library(DT)
library(htmlwidgets)
```

<!-- Content for landing page-->

##  Ferramenta de apoio ao pleanemanto da rede ciclável na Área Metropolitana de Lisboa

##### Nota: este projeto está em fase _beta_ e a informação deste site deve ser considerada como _work in progress_.

```{r pkg, eval=FALSE, include=FALSE}
# TODO this will contain more variables we need, primarily % cycling
# Load counties geometry
counties_raw = readRDS("counties.Rds")
counties_raw = counties_raw %>%
  mutate(county_name = stringr::str_to_title(COUNTY))
# Load pcycle_data at the county level
county_trips = read.csv("county-trips.csv")
# counties_index = county_trips %>%
#   select(county_name, pcycle_baseline)
counties_index = counties_raw %>%
  dplyr::left_join(county_trips)
# counties_index = counties_index %>%
#   dplyr::rename(Name = COUNTY) %>%
#   dplyr::mutate(Name =  stringr::str_to_sentence(Name)) %>%
#   dplyr::mutate(url = paste0(
#     "<a href='",
#     tolower(Name) ,
#     "/' target='_blank'>",
#     Name,
#     "</a>"
#   ))
# counties_index = sf::st_as_sf(counties_index)

municipios_index = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))%>% #it was counties_index -> change to a table with mode statistics!
  dplyr::rename(Name = Concelho) %>%
  dplyr::mutate(Name = stringr::str_to_sentence(Name)) %>%
  dplyr::mutate(url = paste0(
  "<a href='",
  tolower(Name) ,
  "/' target='_blank'>",
  Name,
  "</a>"
)) 
municipios_index = sf::st_as_sf(municipios_index)
```

# {.tabset}

## Mapa

test content

```{r eval=FALSE, include=FALSE}
tippy::tippy_this(elementId = "word", tooltip = "At low zoom, the map shows the % cycle uptake per county; at high zoom, it displays a cycle friendliness measure, with a score between 0 (very unfriendly) and 100 (very friendly). Both measures are shown for each given scenario. See the FAQs for more details")
palette = c(
  '#d73027',
  '#f46d43',
  '#fdae61',
  '#fee090',
  '#ffffbf',
  '#e0f3f8',
  '#abd9e9',
  '#74add1',
  '#4575b4'
)
breaks = c(0, 1, 2, 3, 5, 10, 20, 30, 40, Inf)
pal_baseline = leaflet::colorBin(palette, bins = breaks,
                                 domain = counties_index$pcycle_baseline)
pal_near = leaflet::colorBin(palette, bins = breaks,
                             domain = counties_index$pcycle_near)
pal_climate = leaflet::colorBin(palette, bins = breaks,
                                domain = counties_index$pcycle_climate)
pal_dutch = leaflet::colorBin(palette, bins = breaks,
                              domain = counties_index$pcycle_godutch)
pal_ebike = leaflet::colorBin(palette, bins = breaks,
                              domain = counties_index$pcycle_ebike)
schools_raw = readRDS("schools_all.Rds")
# main map
leaflet::leaflet(counties_index) %>%
  leaflet::addProviderTiles(
    providers$CartoDB.Positron,
    group = "Grey basemap",
    options = leaflet::providerTileOptions(minZoom = 5, maxZoom = 22)
  ) %>%
  leaflet::addPolygons(
    data = counties_index %>% sf::st_transform(crs = 4326),
    color = "black",
    fillColor = "transparent",
    group = "Borders",
    popup = paste0("View county results: ", counties_index$url),
    opacity = 0.7,
    weight = 1.2,
    label = ~ Name,
    labelOptions = leaflet::labelOptions(noHide = F, permanent = TRUE, direction = "auto",sticky = FALSE, textsize = "1.5em")) %>%
  leaflet::addCircles(
    data = schools_raw,
    group = "Schools",
    color = "black",
    opacity = 0.8,
    fillOpacity = 0.3
  ) %>%
  #polygons
  mapboxapi::addMapboxTiles(
    style_id = "cl77hy22i000m14pa2r440vgi",
    username = "nathanaelisamapper",
    group = "Baseline scenario",
    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>%
  mapboxapi::addMapboxTiles(
    style_id = "cl77h0i4v001u15o1cp39sk37",
    username = "nathanaelisamapper",
    group = "Near Market",
    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>%
  mapboxapi::addMapboxTiles(
    style_id = "cl77hqtom000214qrk6ceqeur",
    username = "nathanaelisamapper",
    group = "Climate Action Plan",
    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>%
  mapboxapi::addMapboxTiles(
    style_id = "cl77hvhpx000d14nz0uhmsp40",
    username = "nathanaelisamapper",
    group = "Ebike",
    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>%
  mapboxapi::addMapboxTiles(
    style_id = "cl77i0bqh003i16n2cijgn6r1",
    username = "nathanaelisamapper",
    group = "Go Dutch",
    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>%
  #rnet
  mapboxapi::addMapboxTiles(
    style_id = "cl66fjj97000a15ru2hhss06g",
    username = "nathanaelisamapper",
    group = "Baseline scenario",
    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>%
  mapboxapi::addMapboxTiles(
    style_id = "cl66f8ax6001i14ti630sk73z",
    username = "nathanaelisamapper",
    group = "Climate Action Plan",
    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>%   mapboxapi::addMapboxTiles(
    style_id = "cl66foarr001d14lmc88wyu42",
    username = "nathanaelisamapper",
    group = "Near Market",
    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  )  %>% mapboxapi::addMapboxTiles(
    style_id = "cl66flaic000114t385c8spxn",
    username = "nathanaelisamapper",
    group = "Ebike",
    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>% mapboxapi::addMapboxTiles(
    style_id = "cl66fmn9j001g14lkbone3hye",
    username = "nathanaelisamapper",
    group = "Go Dutch",
    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
  ) %>%
  leaflet::setView(lng = -8.2,
                   lat = 53.5,
                   zoom = 7) %>%
  leaflet::addLegend(
    "bottomleft",
    colors = c("#882255",  "#CC6677", "#44AA99", "#117733"),
    labels = c("0 to 25", "25 to 50", "50 to 75", "75 to 100"),
    title = "Cycle friendliness",
    group = "Friend",
    opacity = 0.7
  ) %>%
  leaflet::addLegend(
    "bottomleft",
    colors = c(
      '#d73027',
      '#f46d43',
      '#fdae61',
      '#fee090',
      '#ffffbf',
      '#e0f3f8',
      '#abd9e9',
      '#74add1',
      '#4575b4'
    ),
    labels = c(
      "0 to 1",
      "1 to 2",
      "2 to 3",
      "3 to 5",
      "5 to 10",
      "10 to 20",
      "20 to 30",
      "30 to 40",
      "40 to more"
    ),
    title = "Percent cycling",
    group = "Base",
    opacity = 0.7
  ) %>%
  leaflet::addLegend(
    title = "Get Started:",
    group = "Help",
    position = "bottomright",
    colors = c("white",
               "white"),
    labels = c(
      "Zoom in to explore the network",
      "Click on a county for detailed results"
    )
  ) %>%
  addLayersControl(
    baseGroups  = c(
      "Baseline scenario",
      "Near Market",
      "Climate Action Plan",
      "Go Dutch",
      "Ebike"
    ),
    overlayGroups = c("Schools"),
    options = leaflet::layersControlOptions(collapsed = FALSE)
  ) %>%
  leaflet::groupOptions(group = "Help", zoomLevels = 7:8)  %>%
    leaflet::groupOptions(group = "Base", zoomLevels = 1:9) %>%
  leaflet::groupOptions(group = "Friend", zoomLevels = 10:20) %>%
  leaflet::hideGroup("Schools")  %>%
  htmlwidgets::onRender("
        function() {
            $('.leaflet-control-layers-base').prepend('<label style=\"text-align:center;font-size:larger;\"><b>Scenarios</b></label>');
        }
    ")
```

**O que mostra este mapa? **  
Por enquanto nada.

------------------------------------------------------------------------

## Dados por Município

add table with link to each municipality

Construir tabela com os dados de populacao, Bike, cyc4, ebike, etc

```{r}
# Aim: Build table for homepage
# dn = "popupCycleways/v1"
# built_counties = list.files(dn)
# set.seed(42)
municipios_index %>%
  sf::st_drop_geometry() %>%
  # dplyr::select(url, Population, "% cycling (Baseline)" = pcycle_baseline, "% cycling (Near Market)" = pcycle_near,  "% cycling (Climate Action Plan)" = pcycle_climate,  "% cycling (Go Dutch)" = pcycle_godutch,  "% cycling (Ebike)" = pcycle_ebike) %>%
  # dplyr::rename(Name = url) %>%
  dplyr::arrange(Name) %>%
  # dplyr::mutate_if(is.numeric, round, 1) %>%
  DT::datatable(escape = -2, filter = 'none', options = list(pageLength = 18,
                                                             responsive = TRUE,
                                                             extensions = c('Responsive'))) 
```

------------------------------------------------------------------------

## Sobre

### Objetivos da ferramenta biclaR

This information can be used to inform decisions about where best to invest in new or improved cycle routes.

### A equipa biclaR

*   Transportes Metropolitanos de Lisboa (coordenação): XX XXX XXXX.

*   Instituto Superior Técnico - Universidade de Lisboa (desenvolvimento): Filipe Moura, Rosa Félix.

*   Institute for Transport Studies - University of Leeds (apoio ao desenvolvimento): Robin Lovelace.

### Agradecimentos

*   Grupo de Trabalho: ...
*   Conselho Consultuvo: ....
*   CycleStreets
*   r5r

### Feedback

You can give feedback on the tool in the [online survey](nada).

### Contactos

Enviar um email para...

------------------------------------------------------------------------

## FAQ

Short answers to key frequently answered questions (FAQ) are provided below.
See the full FAQ page for more details and answers to questions about how the tool works.

### O que é a ferramenta biclar?

tal tal.

### Qual o objetivo da ferramenta biclar?

tal tal tal.

------------------------------------------------------------------------

```{r eval=FALSE, include=FALSE}
# ?
setTimeout(function(){
  $('.loader-bg').fadeToggle();
}, 1000);
```

### 2022-11-20 code/biclarwww/municipio_index.Rmd
biclar website

<!-- landing page for each municipio -->

# Municipio X

See map below

```{r}
#read data
transit_all = sf::st_read("https://github.com/U-Shift/biclar/releases/download/0.0.1/transit.gpkg", quiet = TRUE)

library(tmap)
tmap_mode("view")

tm_shape(transit_all) +
  tm_sf(col = "mode", size = 0.3)
```

### 2022-11-20 code/biclarwww/build.R

@@ -1,0 +1,33 @@
+# Build the website 
+
+
+# Read in data --------------------------------------------------------------------------------
+
+
+
+
+
+
+
+# Create outputs for each region --------------------------------------------------------------
+
+
+rmarkdown::render(input = "code/biclarwww/index.Rmd", output_dir = "biclarwww", output_file = "index.html") #run this
+
+
+# code to build the site
+
+# i = "lisboa"
+
+for (i in municipios){
+  
+  pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio
+  dir.create(pub_folder) #comment this one??
+  rmarkdown::render(input = "code/biclarwww/municipio_index.Rmd", output_dir = pub_folder, output_file = "index.html")
+  # rmarkdown::render(input = "county-stats.Rmd", output_dir = pub_folder, output_file = "county-stats.html")
+  # rmarkdown::render(input = "baseline.Rmd", output_dir = pub_folder, output_file = "baseline.html")
+  # rmarkdown::render(input = "route-types.Rmd", output_dir = pub_folder, output_file = "route-types.html")
+  # rmarkdown::render(input = "downloads.Rmd", output_dir = pub_folder, output_file = "downloads.html")
+  
+}
+

### 2022-11-20 code/biclarwww/index.Rmd
font www
de acordo com o site da TML

@@
       base_font:
-        google: "Prompt"
+        google: "Lexend" #Prompt
       code_font:

### 2022-11-20 HEAT/code/haas_reprex.R
HEAT function multiple runs

@@ -83,7 +83,7 @@ res[26]-res[27]-res[28]+res[29]
 #> [1] 235260000 
 #> 
 #> # ESTE É o unico que dá resultados semelhantes com o do UI, pois usa o genericdata (onde tem o occupancy car rate, velocidades assumidas, e % de shifting de que modos
-# mas só fuinciona pelo reprex :(
+# 
 
 
 

### 2022-11-20 HEAT/code/haas_values_change.R

@@ -1,52 +1,77 @@
 # Assessment intro
-heat1304_webappinput[["ucc_country_location_id"]] = "2264397" #Portugal
-heat1304_webappinput[["ucc_location_id"]] = "2267057" # Lisboa
-heat1304_webappinput[["ucc_activemode"]] = "bike" # how to select more than 1? c("walk", "bike")
-heat1304_webappinput[["ucc_comparisoncases"]] = "twocases" #alternative "onecase"
-heat1304_webappinput[["ucc_yearref"]] = 2018 # reference year
-heat1304_webappinput[["ucc_yearcf"]] = 2024 # comparison year
-heat1304_webappinput[["ucc_assessmenttime"]] = 10 # over how many years should the impacts be calculated
-heat1304_webappinput[["ucc_pathway"]] = c("pa", "ap", "crash", "carbon") # which impact to be considered?
-heat1304_webappinput[["ucc_motormodebasic"]] = "car" # modes to be considered. opt in basic categories: c("car", "pt")
-heat1304_webappinput[["ucc_trafficcondition"]] = "euromeancongestionurban" # alternatives: "somecongestion", "heavycongestion", "euromeancongestionrural"
+webapp_input[["ucc_country_location_id"]] = "2264397" #Portugal
+webapp_input[["ucc_location_id"]] = "2267057" # Lisboa
+webapp_input[["ucc_activemode"]] = "bike" # how to select more than 1? c("walk", "bike")
+webapp_input[["ucc_comparisoncases"]] = "twocases" #alternative "onecase"
+webapp_input[["ucc_yearref"]] = 2018 # reference year
+webapp_input[["ucc_yearcf"]] = 2024 # comparison year
+webapp_input[["ucc_assessmenttime"]] = 10 # over how many years should the impacts be calculated
+webapp_input[["ucc_pathway"]] = c("pa", "ap", "crash", "carbon") # which impact to be considered?
+webapp_input[["ucc_motormodebasic"]] = "car" # modes to be considered. opt in basic categories: c("car", "pt")
+webapp_input[["ucc_trafficcondition"]] = "euromeancongestionurban" # alternatives: "somecongestion", "heavycongestion", "euromeancongestionrural"
 
 # Active modes data (ref = Reference, cf = Comparison)
[…]

### 2022-11-20 code/test-code/heat-prep.R

@@ -1,0 +1,212 @@
+# prepare tables for HEAT SE estimations
+
+
+# PM25 por municipio, incluindo AML
+PM25 = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "PM25")
+PM25$PM25 = round(PM25$PM25, 2)
+PM25$location_id = as.character(PM25$location_id)
+
+# VSL Portugal 2017
+VSL_intusdPPP = 3355000
+VSL_eurotousd = 2323000 #? 
+VSL_usdMER = 2194000
+#https://www.oecd-ilibrary.org/environment/mortality-risk-valuation-in-environment-health-and-transport-policies_9789264130807-en
+usd_to_eur = 0.96858759 #20nov2022
+
+# CENARIOS combinations
+CENARIOS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "CENARIOS")
+CENARIOS$dist_max[CENARIOS$Estrategia == "Convencional"] = 5000 
+CENARIOS$dist_max[CENARIOS$Estrategia == "Ebike"] = 10000 
+CENARIOS$dist_max[CENARIOS$Estrategia == "Intermodal"] = 5000 
+
+# File paths
+PATHS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "paths")
+
+
+# proportion in traffic vs. away from major roads. carspeed > 30 ?? 50 is too high, as a road in 40 is still in traffic.
+
+
+
+
+# script to get values -------------------------------------------------------------------------
+
+# Code == X
+
+Code_Hass = CENARIOS$Code[1:8] #fazer por enquanto só AML e senários 1 e 2
+#para os cenários dos municípios, acrescentar codigo que filtre as viagens com inicio lá, e a rnet within + buffer
+# HEATbind = HEAT #last result without function
+
+
+for (i in Code_Hass){
+
+HEAT = CENARIOS %>% 
+  mutate(Year_ref = `0_Ano`,
+         Year_cf = `1_Ano`) %>% 
+  select(NMunicipio, Municipio, Code, ENMAC, dist_max, Year_ref, Year_cf) %>%
+  filter(Code == i) %>% 
+  left_join(PATHS) %>% 
+  left_join(PM25) %>% 
+  mutate(ENMAC = ENMAC/100)
+
+HEAT_bike = readRDS(HEAT$routes_filepath)
+max_euc = max(HEAT_bike$eucl_distance)
+
[…]

### 2022-11-20 code/test-code/heat-prep2.R

@@ -224,7 +224,7 @@ weighted.mean(routes_car_filtered$distance, w= routes_car_filtered$car10) # 7.47
 
 rnet_joined = readRDS("export2/rnet_ferry3_ebike_overline_morethan50_clean_tags.Rds")
 summary(rnet_joined$carspeed)
-a = rnet_joined %>% filter(carspeed > 50)
+a = rnet_joined %>% filter(carspeed > 30)
 round(sum(a$Bike) / sum(rnet_joined$Bike), 2) #2% stress roads
 
 

### 2022-11-21 code/test-code/quietness-networks.R
run again for scenario 1

@@ -19,12 +19,12 @@ local_crs = "EPSG:3857"
 # rnet_allmodesNSub4_overline_morethan100_clean = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/rnet_allmodesNSub4_overline_morethan100_clean.Rds"))
 
 #enmac
-# rnet = rnet_ferry3_overline_morethan20_clean #done
+# rnet = rnet_ferry3_overline_morethan15_clean #done
+# rnet = rnet_ferry3_overline_morethan50_clean #done
 # rnet = rnet_ferry3_overline_morethan100_clean #done
-# rnet = rnet_ferry3_overline_morethan200_clean #done
-# rnet = rnet_ferry4_overline_morethan20_clean #done
+# rnet = rnet_ferry4_overline_morethan15_clean #done
+# rnet = rnet_ferry4_overline_morethan50_clean #done
 # rnet = rnet_ferry4_overline_morethan100_clean #done
-# rnet = rnet_ferry4_overline_morethan200_clean #done
 # 
 # #ebikes
 # rnet = rnet_ferry3_ebike_overline_morethan50_clean #done
@@ -101,12 +101,12 @@ rnet_joined$slope = scales::label_percent(accuracy = 0.1, sufix = " %")(rnet_joi
 
 
 
-# saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan20_clean_tags.Rds") #done
+# saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan15_clean_tags.Rds") #done
+# saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan50_clean_tags.Rds") #done
 # saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan100_clean_tags.Rds") #done
-# saveRDS(rnet_joined, "export2/rnet_ferry3_overline_morethan200_clean_tags.Rds") #done
-# saveRDS(rnet_joined, "export2/rnet_ferry4_overline_morethan20_clean_tags.Rds") #done
+# saveRDS(rnet_joined, "export2/rnet_ferry4_overline_morethan15_clean_tags.Rds") #done
+# saveRDS(rnet_joined, "export2/rnet_ferry4_overline_morethan50_clean_tags.Rds") #done
 # saveRDS(rnet_joined, "export2/rnet_ferry4_overline_morethan100_clean_tags.Rds") #done
[…]

### 2022-11-21 code/test-code/r5r_filter_overline_routes_CORRECT.R

@@ -63,11 +63,20 @@ saveRDS(rnet_ferry4_overline, "routes_ferry4_overline.Rds")
 saveRDS(routes_ferry4_filtered, "routes_ferry4_preoverline.Rds")
 
 ## filter segments with less than X trips?
-summary(rnet_ferry4_overline$new_cyc4)
+summary(rnet_ferry4_overline$cyc4)
+# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+# 0.774   7.979  19.923  42.091  51.866 692.216 
 
-rnet_ferry4_overline_morethan20 = rnet_ferry4_overline %>% filter(cyc4 >=20) #só sesimbra e mafra
-rnet_ferry4_overline_morethan100 = rnet_ferry4_overline %>% filter(cyc4 >=100)
-rnet_ferry4_overline_morethan200 = rnet_ferry4_overline %>% filter(cyc4 >=200) #estruturante
+rnet_ferry4_overline_morethan15 = rnet_ferry4_overline %>% filter(cyc4 >=15) #só sesimbra e mafra
+rnet_ferry4_overline_morethan50 = rnet_ferry4_overline %>% filter(cyc4 >=50) 
+rnet_ferry4_overline_morethan100 = rnet_ferry4_overline %>% filter(cyc4 >=100) # estruturante
+
+# rnet_ferry4_overline_morethan20 = rnet_ferry4_overline %>% filter(cyc4 >=20) #só sesimbra e mafra
+# rnet_ferry4_overline_morethan30 = rnet_ferry4_overline %>% filter(cyc4 >=30)
+# rnet_ferry4_overline_morethan60 = rnet_ferry4_overline %>% filter(cyc4 >=60) 
+# rnet_ferry4_overline_morethan80 = rnet_ferry4_overline %>% filter(cyc4 >=80) 
+# rnet_ferry4_overline_morethan200 = rnet_ferry4_overline %>% filter(cyc4 >=200) #estruturante
+
 
 
 #rnet clean to get rid of small pieces - 100
@@ -86,54 +95,69 @@ rnet_ferry4_overline_morethan100_clean = rnet_ferry4_overline_morethan100_clean 
 rm(rnet_ferry4_overline_morethan100_group)
 
 
-#rnet clean to get rid of small pieces - 20
-rnet_ferry4_overline_morethan20_clean = rnet_ferry4_overline_morethan20 %>%
[…]

### 2022-11-21 code/test-code/r5r_intermodality.R

@@ -73,87 +73,87 @@ departure_datetime = lubridate::force_tz(departure_datetime, "WET") #make it wes
 
 
 
-#routing with LTS3 (enthused and confident) - all transit modes (except Bus and subway)
-routes_r5r_100jit_lts3__intermod_NoSub_elev = detailed_itineraries(
-  r5r_lts_intermodalityALL,
-  origins = od_jittered_100filter_OR,
-  destinations = od_jittered_100filter_DE,
-  mode = c("BICYCLE", "RAIL", "TRAM", "FERRY", "BUS"), #remove subway
-  mode_egress = "BICYCLE",
-  departure_datetime = departure_datetime,
-  # time_window = 1L,
-  # suboptimal_minutes = 0L,
-  fare_structure = NULL,
-  max_fare = Inf,
-  max_walk_time = Inf,
-  max_bike_time = 30L, #25min no total - será pouco ou muito?
-  max_trip_duration = 120L, #in minutes
-  # walk_speed = 3.6,
-  bike_speed = 14, #higher to be competitive with PT
-  max_rides = 1, #max public transit rides for the same trip
-  max_lts = 3, #1 - quietest, 4 - hardcore
-  shortest_path = TRUE, #FALSE? fastest or multiple alternatives?
-  all_to_all = FALSE,
-  n_threads = Inf,
-  verbose = FALSE,
-  progress = TRUE,
-  drop_geometry = FALSE,
-  output_dir = NULL
-)
-
-saveRDS(routes_r5r_100jit_lts3__intermod_NoSub_elev, "routes_r5r_100jit_lts3__intermodALL_NoSubBus_elev_resultsraw.Rds")
-
-routes_r5r_100jit_lts3__intermod_NoSub_elev = routes_r5r_100jit_lts3__intermod_NoSub_elev %>% mutate(id = as.integer(from_id)) %>%
-  select(id, total_duration, total_distance, segment, mode, segment_duration, distance, route, geometry) %>%
-  left_join(od_jittered_100filter %>% st_drop_geometry(), by="id")
-
-names(routes_r5r_100jit_lts3__intermod_NoSub_elev)[7] = "distance"
-names(routes_r5r_100jit_lts3__intermod_NoSub_elev)[18] = "eucl_distance"
-
[…]

### 2022-11-21 code/test-code/quietness-networks.R
Update quietness-networks.R

@@ -128,7 +128,7 @@ rnet_joined$slope = scales::label_percent(accuracy = 0.1, sufix = " %")(rnet_joi
 ####
 
 # library(biclar)
-rnet_joined = readRDS("export2/rnet_ferry4_ebike_overline_morethan100_clean_tags.Rds")
+rnet_joined = readRDS("export2/rnet_ferry4_ebike_overline_morethan50_clean_tags.Rds")
 tm_rnet(rnet_joined, #filtrar barreiro [MUNICIPIOSgeo %>% filter(Concelho == "Barreiro"), , op = sf::st_within]
         lwd = "cyc4", #Baseline, ENMAC4, ENMAC10
         col = "quietness",

### 2022-11-21 code/test-code/heat-prep_municipios.R
loop for HaaS

@@ -22,35 +22,41 @@ CENARIOS$dist_max[CENARIOS$Estrategia == "Intermodal"] = 5000
 # File paths
 PATHS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "paths")
 
-
-# proportion in traffic vs. away from major roads. carspeed > 30 ?? 50 is too high, as a road in 40 is still in traffic.
-
-
+# DICOFRE
+Relacao1116 = readxl::read_excel("D:/GIS/IMOB/GIS/InfExtra_Concelhos_Freguesias_CAOP2019/RelacaoFreguesiasINE20112016.xlsx", 
+                            col_types = c("text", "skip", "text", 
+                                          "skip"))
+Relacao1116 = Relacao1116 %>% fill(DICOFRE16) %>%
+  mutate(DTCC = substr(DICOFRE16, 1, 4))%>% #ficar com primeiros 4 digitos
+  filter(DTCC %in% as.character(PM25$DTCC))
 
 
 # script to get values -------------------------------------------------------------------------
 
 # Code == X
 
-Code_Hass = CENARIOS$Code[1:8] #fazer por enquanto só AML e senários 1 e 2
+Code_Hass = CENARIOS %>% filter(Estrategia != "Intermodal") %>% filter(Municipio !="AML") %>% select(Code) #fazer por enquanto só AML e senários 1 e 2
 #para os cenários dos municípios, acrescentar codigo que filtre as viagens com inicio lá, e a rnet within + buffer
 # HEATbind = HEAT #last result without function
 
 
-for (i in Code_Hass){
-  
-# i = 1304
+for (i in Code_Hass$Code){
 
+i = 31304 #test: há muitos NA quando se junta o DTCC
+  
 HEAT = CENARIOS %>% 
   mutate(Year_ref = `0_Ano`,
          Year_cf = `1_Ano`) %>% 
   select(NMunicipio, Municipio, Code, ENMAC, dist_max, Year_ref, Year_cf) %>%
   filter(Code == i) %>% 
-  left_join(PATHS) %>% 
+  mutate(CodePath = as.integer(stringr::str_sub(Code, -4, -1))) %>% 
+  left_join(PATHS, by=c("CodePath" = "Code")) %>% 
[…]

### 2022-11-22 code/test-code/heat-prep.R
HEAT with VSL in PPP usd

@@ -7,11 +7,11 @@ PM25$PM25 = round(PM25$PM25, 2)
 PM25$location_id = as.character(PM25$location_id)
 
 # VSL Portugal 2017
-VSL_intusdPPP = 3355000
+usd_to_eur = 0.976430 #21nov2022
+VSL_intusdPPP = 3055358/usd_to_eur #fonte: ANSR 2021. original = 3355000
 VSL_eurotousd = 2323000 #? 
 VSL_usdMER = 2194000
 #https://www.oecd-ilibrary.org/environment/mortality-risk-valuation-in-environment-health-and-transport-policies_9789264130807-en
-usd_to_eur = 0.96858759 #20nov2022
 
 # CENARIOS combinations
 CENARIOS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "CENARIOS")
@@ -101,8 +101,14 @@ HEAT = HEAT %>% mutate(Bike = round(sum(HEAT_bike$Bike)),
 
 # HaaS 
 
-webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_webapp_input.rds"))
-genericdata = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_generic_data_relevant_with_calc_params.rds"))
+# # for MER VSL value
+# webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_webapp_input.rds"))
+# genericdata = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_generic_data_relevant_with_calc_params.rds"))
+
+# for PPP VSL value
+webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/162410_webapp_input.rds"))
+genericdata = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/162410_generic_data_relevant_with_calc_params.rds"))
+
 
 
 # Assessment intro
@@ -168,9 +174,9 @@ webapp_input[["pollution_calc"]] = HEAT$PM25 # PM2.5 concentration fot the locat
 # webapp_input[["crashdata_used_fatality_rate_bike_cf"]] = 1 # Fatality rate for cycling in the comparison case, as integer of fatalities / 100 million km
 
 # Value of Statistical Life VSL
[…]

### 2022-11-25 code/test-code/add_heat_to_rnet.R
add SE values to rnets

@@ -1,0 +1,63 @@
+# heat in lines
+
+# names(HEAT_aml_redux)
+# names(rnet_tags)
+
+# read data
+
+HEAT_aml_redux = readRDS("HEAT/HEAT_AML_PPP_123_redux.Rds")
+RNETS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "rnet")
+
+
+for (i in 1:nrow(RNETS)){
+
+rnet_tags = readRDS(RNETS$path[i])
+code = RNETS$Code[i]
+HEAT_aml = HEAT_aml_redux %>% filter(Code == code)
+
+# Mortality10_aml = as.numeric(gsub(" ","", HEAT_aml$Mortality10))/HEAT_aml$Bike_new #este não faz sentido por segmento
+tCO2eq10_aml = as.numeric(gsub(" ","", HEAT_aml$CO2eq10))/HEAT_aml$Bike_new
+Econommic10_aml = HEAT_aml$value_newcyc_eur
+
+if (HEAT_aml$ENMAC == 0.04) {
+  rnet_tags = rnet_tags %>%
+    select(-c(cyc10, new_cyc10)) %>%
+    rename(cyc = cyc4,
+           new_cyc = new_cyc4)
+  
+} else {
+  rnet_tags = rnet_tags %>%
+    select(-c(cyc4, new_cyc4)) %>%
+    rename(cyc = cyc10,
+           new_cyc = new_cyc10)
+} 
+    
+
+rnet_tags = rnet_tags %>%
+  select(-carspeedmax) %>% 
+  mutate(Beneficios10_keuro = round(Econommic10_aml*new_cyc/1000), #em milhares de euro
+         tCO2eq10 = round(tCO2eq10_aml*new_cyc))  
+
+summary(rnet_tags$Beneficios10)
+summary(rnet_tags$tCO2eq10)
+
+path = paste0("export_rnet/rnet_aml_", code, "_", RNETS$Rede[i], ".Rds")
+path_sf = paste0("export_rnet/rnet_aml_", code, "_", RNETS$Rede[i], ".gpkg")
+
+saveRDS(rnet_tags, path)
+sf::st_write(rnet_tags, path_sf, delete_dsn = TRUE)
+
+print(path)
+
+}
+
+
+# test map
+
+rnet_test = sf::st_read("export_rnet/rnet_aml_3410_10.gpkg")
[…]

### 2022-11-25 code/test-code/emmission_valued_TP.R

@@ -1,0 +1,6 @@
+# HEAT cenario 3 - only the TP leg
+
+
+CAR_TP_TRANSF_results <- readxl::read_excel("HEAT/CAR_TP_TRANSF_results.xlsm", sheet = "Values", range = "A1:N5")
+
+# join with HEAT AML redux

### 2022-11-25 code/test-code/color_pick.R
maps colors and viz tests

@@ -1,0 +1,14 @@
+# colors scenarios and quietness levels
+
+
+# test map
+
+rnet_test = sf::st_read("export_rnet/rnet_aml_3410_10.gpkg")
+biclar::tm_rnet(rnet_test[MUNICIPIOSgeo %>% filter(Concelho == "Barreiro"), , op = sf::st_within],
+                lwd = "cyc", #Baseline, ENMAC4, ENMAC10
+                col = "quietness",
+                palette = "-yl_gn_bu", # "linear_yl_rd_bk", "mako", "burg", "reds" - reds for fastest, mako for quietest
+                scale = 15, #30 para lisboa?
+                lwd_multiplier = 15)
+
+

### 2022-11-25 code/test-code/donut_map.R

@@ -1,0 +1,41 @@
+# try to replicate donut_maps
+# https://github.com/mtennekes/donutmaps
+
+TRIPSmode_municipal = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/TRIPSmode_municipal.Rds"))
+CENTROIDS_municipios = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/CENTROIDS_municipios.Rds"))
+ZONAMENTO = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/ZONAMENTO_imob.Rds"))
+ZONAMENTO$DTMN_DSG[ZONAMENTO$DTMN_DSG == "Setúbal"] = "Setubal"
+
+
+DonutTrips = TRIPSmode_municipal %>% 
+  group_by(Origem, Destino) %>% 
+  summarise(viagens = round(sum(viagens))) %>% 
+  left_join(ZONAMENTO %>%
+              select(DTMN, DTMN_DSG) %>%
+              distinct(), 
+            by = c("Origem" = "DTMN_DSG")) %>% 
+  rename(DTMN_or = DTMN) %>% 
+  left_join(ZONAMENTO %>% select(DTMN, DTMN_DSG) %>% distinct(), by = c("Destino" = "DTMN_DSG")) %>% 
+  rename(DTMN_de = DTMN) %>% 
+  select(DTMN_or, DTMN_de, viagens, Origem, Destino)
+
+CENTROIDS_municipios = CENTROIDS_municipios %>% select(DTMN11, geometry)
+
+
+DonutTrips = stplanr::od2line(flow = DonutTrips,
+                              zones = CENTROIDS_municipios)
+
+
+library(tmap)
+library(donutmaps)
+
+bm = bake_donuts(x = DonutTrips,
+            var = "viagens",
+            groupname = "Origem",
+            pal = CBS_pal,
+            donut_size_min = 30000, donut_size_max = 400000,
+            flow_th = 500, flow_max = 20000, flow_buffer = 500, flow_scale = 10,
+            donut_scale = 1.75)
+
+#não funciona, só com a versão tmap 4
+# https://mtennekes.github.io/tmap4/index.html

### 2022-11-25 code/test-code/emmission_valued_TP.R
add replaced TP emissions

@@ -4,3 +4,25 @@
 CAR_TP_TRANSF_results <- readxl::read_excel("HEAT/CAR_TP_TRANSF_results.xlsm", sheet = "Values", range = "A1:N5")
 
 # join with HEAT AML redux
+
+HEAT_aml_results = HEAT_aml_redux %>%
+  left_join(CAR_TP_TRANSF_results %>% select(Code, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, money_emissions_eur))
+
+
+saveRDS(HEAT_aml_results, "HEAT/HEAT_aml_results.Rds")
+
+
+
+
+
+# verificar -----------------------------------------------------------------------------------
+
+# sum(HEAT_bike$distance*HEAT_bike$new_cyc)
+# dist_bike_0km= sum(HEAT_bike$distance*HEAT_bike$new_cyc)/1000 #148407.1 km/dia
+# 
+# HEAT$CO2eq #5919 ton / ano
+# # 5919*1000000/365 = 16216438 g/dia
+# 
+# 16216438 / dist_bike_0km = 109.27 g/km 
+# 
+# #está certo!

### 2022-11-25 code/biclarwww/index.Rmd
web tab municipios

--- 2022-11-20 code/biclarwww/index.Rmd
@@
       base_font:
-        google: "Prompt"
+        google: "Lexend" #Prompt
       code_font:
         google: "JetBrains Mono"
-    # css: style.css
+    css: css/style.css
     includes:
@@
 
-##  Ferramenta de apoio ao pleanemanto da rede ciclável na Área Metropolitana de Lisboa
+##  Ferramenta de apoio ao planeamento da rede ciclável na Área Metropolitana de Lisboa
 
@@
 
-municipios_index = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))%>% #it was counties_index -> change to a table with mode statistics!
-  dplyr::rename(Name = Concelho) %>%
-  dplyr::mutate(Name = stringr::str_to_sentence(Name)) %>%
+municipios_index = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds")) %>% #it was counties_index -> change to a table with mode statistics!
+  # dplyr::mutate(county_name = stringr::str_to_title(Concelho)) %>% #maiusculas
+  dplyr::rename(Municipio = Concelho) %>%
+  # dplyr::mutate(Municipio = stringr::str_to_sentence(Municipio)) %>% #minusculas
   dplyr::mutate(url = paste0(
-  "<a href='",
-  tolower(Name) ,
-  "/' target='_blank'>",
-  Name,
-  "</a>"
+  "<a href='", tolower(Municipio) , "/' target='_blank'>",  Municipio,  "</a>"
 )) 
@@
 
-## Dados por Município
-
-add table with link to each municipality
+## AML
 
@@
 
-```{r}
+```{r eval=FALSE, include=FALSE}
 # Aim: Build table for homepage
@@
                                                              extensions = c('Responsive'))) 
+```
+
+------------------------------------------------------------------------
+
+## Municípios
+
+Construir tabela com os dados de populacao, Bike, cyc4, ebike, etc, por municipio
+
+```{r}
+# Aim: Build table for homepage
+# dn = "popupCycleways/v1"
+# built_counties = list.files(dn)
+# set.seed(42)
+municipios_index %>%
+  sf::st_drop_geometry() %>%
+  dplyr::left_join(HEAT_municipios_PPP_redux %>% filter(CodePath == 1304)) %>% 
+  dplyr::select(url, Pop21, Total, Bike, Bike_new, CO2eq10, Economic10) %>%
+  dplyr::mutate(Economic10 = Economic10/1000,
+                CO2eq10 = as.numeric(gsub(" ","", CO2eq10))) %>% 
+  # Population, "% cycling (Baseline)" = pcycle_baseline, "% cycling (Near Market)" = pcycle_near,  "% cycling (Climate Action Plan)" = pcycle_climate,  "% cycling (Go Dutch)" = pcycle_godutch,  "% cycling (Ebike)" = pcycle_ebike) %>%
+  dplyr::rename(Municipio = url) %>%
+  dplyr::arrange(Municipio) %>%
+  # dplyr::mutate_if(is.numeric, round, 1) %>%
+  DT::datatable(escape = -1, #para colocar os links, a contar com os rownames
+                filter = 'none',
+                rownames = FALSE,
+                colnames = c("Município", "População 2021", "Viagens /dia", "Viagens Bicicleta", "Novas Viagens Bicicleta 4%",
+                             "CO2eq evitado em 10 anos (ton)", "Benefícios estimados em 10 anos (milhares €)"),
+                options = list(dom = "t",
+                               pageLength = 18,
+                               responsive = TRUE,
+                              extensions = c('Responsive'))) %>% 
+  DT::formatRound(c(2:7),0, mark = " ")
 ```

### 2022-12-03 code/biclarwww/municipio_index.Rmd
font www
de acordo com o site da TML
progress with maps

@@
-<!-- landing page for each municipio -->
+---
+title: "Resultados por município"
+output:
+  html_document:
+    theme:
+      bg: "#FFFFFF"
+      fg: "#000000"
+      highlight: "#000000"
+      primary: "#000000"
+      base_font:
+        google: "Lexend"
+      code_font:
+        google: "JetBrains Mono"
+    css: css/style.css
+    includes:
+      after_body: img/footer.html
+---
 
-# Municipio X
-
-See map below
-
-```{r}
-#read data
-transit_all = sf::st_read("https://github.com/U-Shift/biclar/releases/download/0.0.1/transit.gpkg", quiet = TRUE)
-
+```{r setup, include=FALSE}
+library(dplyr)
 library(tmap)
@@
 
-tm_shape(transit_all) +
-  tm_sf(col = "mode", size = 0.3)
+knitr::opts_chunk$set(echo = FALSE,
+                      message = FALSE,
+                      warning = FALSE)
+
+fmt_num = function(x) format(round(sum(x)), big.mark=" ")
 ```
+
+```{r chamadas}
+i = "barreiro"
+mun = stringr::str_to_title(i)
+proposicao = ifelse(mun %in% c("Amadora", "Moita"), "Na",
+                    ifelse(mun %in% c("Barreiro", "Montijo", "Seixal"), "No",
+                           "Em"))
+proposicaoD = ifelse(mun %in% c("Amadora", "Moita"), "da",
+                    ifelse(mun %in% c("Barreiro", "Montijo", "Seixal"), "do",
+                           "de"))
+```
+
+```{r}
+MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))
+
+HEAT_municipios_redux = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/HEAT_municipios_123_redux.Rds"))
+HEAT_mun = HEAT_municipios_redux %>% filter(Municipio == mun)
+
+# mapa11 = sf::st_read("export_rnet/rnet_aml_1410_50.gpkg")
+mapa11 = sf::st_read("D:/GIS/TML/biclar/export_rnet/rnet_aml_1310_50.gpkg", quiet = TRUE)
+mapa11 = mapa11[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10)
+mapa12 = sf::st_read("D:/GIS/TML/biclar/export_rnet/rnet_aml_1410_50.gpkg", quiet = TRUE)
+mapa12 = mapa12[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10)
+
+# mapa2 = sf::st_read("export_rnet/rnet_aml_2410_100.gpkg")
+mapa2 = sf::st_read("D:/GIS/TML/biclar/export_rnet/rnet_aml_2410_100.gpkg", quiet = TRUE)
+mapa2 = mapa2[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within]
+
+# mapa3 = sf::st_read("export_rnet/rnet_aml_3410_30.gpkg")
+mapa3 = sf::st_read("D:/GIS/TML/biclar/export_rnet/rnet_aml_3410_30.gpkg", quiet = TRUE)
+mapa3 = mapa3[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within]
+
+```
+
+<!-- landing page for each municipio -->
+
+# `r mun`
+
+`r proposicao` `r mun`, as viagens são assim.
+
+Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMOB 2018), para **4%** de viagens em bicicleta (meta 2024), e para **10%** de viagens em bicicleta (meta 2030), transferidas diretamente de viagens em automóvel.
+
+Cada mapa refere-se a cada uma das 3 estimativas:
+
+* viagens até 5 km com ligação em bicicleta (_`r fmt_num(HEAT_mun[4,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+* viagens até 10 km com ligação em bicicleta (_`r fmt_num(HEAT_mun[8,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+* viagens até 5 km em bicicleta de ou até uma interface de transportes (_`r fmt_num(HEAT_mun[12,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+
+### Até 5 km
+
+```{r}
+biclar::tm_rnet(mapa1,
+                lwd = "cyc", #Baseline, ENMAC4, ENMAC10
+                col = "quietness",
+                palette = "-mako", 
+                scale = 15, #30 para lisboa?
+                lwd_multiplier = 15) 
+```
+
+```{r}
+m = tmap::tm_shape(mapa11, name = "Potencial 4%") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 12,
+                   lwd_multiplier = 15,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  # "Em bicicleta 2018" = "Bike",
+                                  "% bicicleta 2018" = "Bikeper",
+                                  "Em bicicleta 4%" = "cyc",
+                                  "Novos ciclistas 4%" = "new_cyc",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2 evitado [ton/ano]" = "tCO2eq"
+                                  ),
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-mako", n = 5),
+                   breaks = c(0, 20, 40, 60, 80, 100),
+                   labels = c("0 - 19", "20 - 39", "40 - 59", "60 - 79", "80 - 100"),
+                   textNA = "sem info",
+                   title.col = "Tranquilidade"
+                   ) +
+  tmap::tm_shape(mapa12, name = "Potencial 10%") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 12,
+                   lwd_multiplier = 15,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  # "Em bicicleta 2018" = "Bike",
+                                  "% bicicleta 2018" = "Bikeper",
+                                  "Em bicicleta 10%" = "cyc",
+                                  "Novos ciclistas 10%" = "new_cyc",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2 evitado [ton/ano]" = "tCO2eq"
+                                  ),
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-reds", n = 5),
+                   breaks = c(0, 20, 40, 60, 80, 100),
+                   labels = c("0 - 19", "20 - 39", "40 - 59", "60 - 79", "80 - 100"),
+                   textNA = "sem info",
+                   title.col = "Tranquilidade",
+                   ) +
+  tm_view(view.legend.position = c("right", "bottom"),
+          control.position = c("right", "top"),
+          leaflet.options = highlightOptions(color = "white", bringToFront = TRUE))
+tmap::tmap_leaflet(m) %>% 
+  # leaflet::addPolylines(highlightOptions = highlightOptions(color = "white", bringToFront = TRUE)) %>%
+  leaflet::addLayersControl(overlayGroups = c("Potencial 4%", "Potencial 10%"),
+                                                    options = layersControlOptions(collapsed = FALSE))
+```
+
+### Até 10 km (e-bike)
+
+```{r}
+biclar::tm_rnet(mapa2,
+                lwd = "cyc", #Baseline, ENMAC4, ENMAC10
+                col = "quietness",
+                palette = "-linear_yl_rd_bk", 
+                scale = 15, #30 para lisboa?
+                lwd_multiplier = 15)
+```
+
+### Até 5 km intermodal
+
+```{r}
+biclar::tm_rnet(mapa3,
+                lwd = "cyc", #Baseline, ENMAC4, ENMAC10
+                col = "quietness",
+                palette = "-burg", 
+                scale = 15, #30 para lisboa?
+                lwd_multiplier = 15)
+```
+
+> Nota: Estes mapas referem-se às ligações mais _rápidas_. Para ligações mais _tranquilas_, ver a página de [donwloads](downloads.html).
+
+## Resumo
+
+```{r}
+# tabela com potencial para cada cenario +  estimativas SE
+```

### 2022-12-03 code/biclarwww/build.R

@@ -24,7 +24,7 @@ for (i in municipios){
   pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio
   dir.create(pub_folder) #comment this one??
   rmarkdown::render(input = "code/biclarwww/municipio_index.Rmd", output_dir = pub_folder, output_file = "index.html")
-  # rmarkdown::render(input = "county-stats.Rmd", output_dir = pub_folder, output_file = "county-stats.html")
+  # rmarkdown::render(input = "stats.Rmd", output_dir = pub_folder, output_file = "stats.html")
   # rmarkdown::render(input = "baseline.Rmd", output_dir = pub_folder, output_file = "baseline.html")
   # rmarkdown::render(input = "route-types.Rmd", output_dir = pub_folder, output_file = "route-types.html")
   # rmarkdown::render(input = "downloads.Rmd", output_dir = pub_folder, output_file = "downloads.html")

### 2022-12-03 code/biclarwww/municipio_index.Rmd
getting individual files better

--- 2022-11-20 code/biclarwww/municipio_index.Rmd
@@
+---
+title: "Resultados por município"
+output:
+  html_document:
+    theme:
+      bg: "#FFFFFF"
+      fg: "#000000"
+      highlight: "#000000"
+      primary: "#000000"
+      base_font:
+        google: "Lexend"
+      code_font:
+        google: "JetBrains Mono"
+    css: css/style.css
+    includes:
+      after_body: img/footer.html
+---
+
+```{r setup, include=FALSE}
+library(dplyr)
+library(tmap)
+tmap_mode("view")
+library(leaflet)
+
+knitr::opts_chunk$set(echo = FALSE,
+                      message = FALSE,
+                      warning = FALSE)
+
+fmt_num = function(x) format(round(sum(x)), big.mark=" ")
+```
+
+```{r chamadas}
+i = "barreiro" #delete this when running for all!!!
+mun = stringr::str_to_title(i)
+
+if (i == "vila franca de xira") {
+    i = "vfxira"
+  }
+
+proposicao = ifelse(mun %in% c("Amadora", "Moita"), "Na",
+                    ifelse(mun %in% c("Barreiro", "Montijo", "Seixal"), "No",
+                           "Em"))
+proposicaoD = ifelse(mun %in% c("Amadora", "Moita"), "da",
+                    ifelse(mun %in% c("Barreiro", "Montijo", "Seixal"), "do",
+                           "de"))
+```
+
+```{r}
+MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))
+
+HEAT_municipios_redux = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/HEAT_municipios_123_redux.Rds"))
+HEAT_mun = HEAT_municipios_redux %>% filter(Municipio == mun)
+
+# cenario 1
+
+if (i %in% c("mafra", "sesimbra", "alcochete")){
+
+  mapa11r = sf::st_read("export_rnet/rnet_aml_1404_15.gpkg", quiet = TRUE)
+  mapa11r = mapa11r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+  mapa11l = sf::st_read("export_rnet/rnet_aml_1304_15.gpkg", quiet = TRUE)
+  mapa11l = mapa11l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+  mapa12r = sf::st_read("export_rnet/rnet_aml_1410_15.gpkg", quiet = TRUE)
+  mapa12r = mapa12r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+  mapa12l = sf::st_read("export_rnet/rnet_aml_1310_15.gpkg", quiet = TRUE)
+  mapa12l = mapa12l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+
+} else {
+
+mapa11r = sf::st_read("export_rnet/rnet_aml_1404_50.gpkg", quiet = TRUE)
+mapa11r = mapa11r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa11l = sf::st_read("export_rnet/rnet_aml_1304_50.gpkg", quiet = TRUE)
+mapa11l = mapa11l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa12r = sf::st_read("export_rnet/rnet_aml_1410_50.gpkg", quiet = TRUE)
+mapa12r = mapa12r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa12l = sf::st_read("export_rnet/rnet_aml_1310_50.gpkg", quiet = TRUE)
+mapa12l = mapa12l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+
+}
+
+# cenario 2
+
+if (i %in% c("mafra", "sesimbra", "alcochete")){
+
+mapa21r = sf::st_read("export_rnet/rnet_aml_1404_50.gpkg", quiet = TRUE)
+mapa21r = mapa21r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa21l = sf::st_read("export_rnet/rnet_aml_1304_50.gpkg", quiet = TRUE)
+mapa21l = mapa21l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa22r = sf::st_read("export_rnet/rnet_aml_1410_50.gpkg", quiet = TRUE)
+mapa22r = mapa22r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa22l = sf::st_read("export_rnet/rnet_aml_1310_50.gpkg", quiet = TRUE)
+mapa22l = mapa22l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+
+} else if (i %in% c("almada", "oeiras", "lisboa", "odivelas", "amadora")){
+
+mapa21r = sf::st_read("export_rnet/rnet_aml_1404_200.gpkg", quiet = TRUE)
+mapa21r = mapa21r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa21l = sf::st_read("export_rnet/rnet_aml_1304_200.gpkg", quiet = TRUE)
+mapa21l = mapa21l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa22r = sf::st_read("export_rnet/rnet_aml_1410_200.gpkg", quiet = TRUE)
+mapa22r = mapa22r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa22l = sf::st_read("export_rnet/rnet_aml_1310_200.gpkg", quiet = TRUE)
+mapa22l = mapa22l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+
+} else {
+
+mapa21r = sf::st_read("export_rnet/rnet_aml_1404_100.gpkg", quiet = TRUE)
+mapa21r = mapa21r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa21l = sf::st_read("export_rnet/rnet_aml_1304_100.gpkg", quiet = TRUE)
+mapa21l = mapa21l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa22r = sf::st_read("export_rnet/rnet_aml_1410_100.gpkg", quiet = TRUE)
+mapa22r = mapa22r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa22l = sf::st_read("export_rnet/rnet_aml_1310_100.gpkg", quiet = TRUE)
+mapa22l = mapa22l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+
+}
+
+# cenario 3
+
+if (i %in% c("lisboa", "almada", "cascais", "amadora", "sintra", "oeiras")){
+
+  mapa31r = sf::st_read("export_rnet/rnet_aml_3404_30.gpkg", quiet = TRUE)
+mapa31r = mapa31r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa31l = sf::st_read("export_rnet/rnet_aml_3304_30.gpkg", quiet = TRUE)
+mapa31l = mapa31l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa32r = sf::st_read("export_rnet/rnet_aml_3410_30.gpkg", quiet = TRUE)
+mapa32r = mapa32r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa32l = sf::st_read("export_rnet/rnet_aml_3310_30.gpkg", quiet = TRUE)
+mapa32l = mapa32l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+
+} else {
+
+mapa31r = sf::st_read("export_rnet/rnet_aml_3404_10.gpkg", quiet = TRUE)
+mapa31r = mapa31r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa31l = sf::st_read("export_rnet/rnet_aml_3304_10.gpkg", quiet = TRUE)
+mapa31l = mapa31l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa32r = sf::st_read("export_rnet/rnet_aml_3410_10.gpkg", quiet = TRUE)
+mapa32r = mapa32r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+mapa32l = sf::st_read("export_rnet/rnet_aml_3310_10.gpkg", quiet = TRUE)
+mapa32l = mapa32l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+  mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+
+}
+
+```
+
+```{r minimos}
+if (i %in% c("mafra", "sesimbra", "alcochete")){ 
+  minimo1 = 15} else {
+    minimo1 = 50 
+  }
+
+if (i %in% c("mafra", "sesimbra", "alcochete")){ 
+  minimo2 = 50} else if (i %in% c("almada", "oeiras", "lisboa", "odivelas", "amadora")){
+    minimo2 = 200 } else {
+    minimo2 = 100
+  }
+
+if (i %in% c("lisboa", "almada", "cascais", "amadora", "sintra", "oeiras")){
+  minimo3 = 30} else {
+    minimo3 = 10
+  }
+
+```
+
 <!-- landing page for each municipio -->
 
-# Municipio X
+# `r mun`
 
-See map below
+`r proposicao` `r mun`, as viagens são assim.
+
+Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMOB 2018), para **4%** de viagens em bicicleta (meta 2024), e para **10%** de viagens em bicicleta (meta 2030), transferidas diretamente de viagens em automóvel.
+
+Cada mapa refere-se a cada uma das 3 estimativas:
+
+* viagens até 5 km com ligação em bicicleta (_`r fmt_num(HEAT_mun[4,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+* viagens até 10 km com ligação em bicicleta elétrica (_`r fmt_num(HEAT_mun[8,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+* viagens até 5 km em bicicleta de ou até uma interface de transportes (_`r fmt_num(HEAT_mun[12,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+
+<!-- Passar isto para FAQ? -->
+Os mapas apresentam a informação gráfica para duas redes - a rede ciclável para viagens mais diretas, e a rede para viagens mais seguras e tranquilas (conforme a rede viária existente).
+A largura das linhas varia consoante o número de ciclistas potencial para cada um dos cenários (4% e 10% de ciclistas), e essa informação é visível ao passar com o rato em cima da linha.
+A cor das linhas varia com o nível de tranquilidade ([_quietness_](https://www.cyclestreets.net/help/journey/howitworks/#quietness)), sendo as mais escuras aquelas menos seguras para circular em bicicleta - indicando com maior clareza quais os segmentos que necessitam de intervenção para se tornarem mais seguros e potencialmente trazerem mais ciclistas: os **mais escuros e mais largos**.
+
+Ao clicar em cada linha, uma janela abre-se com mais informação sobre as viagens estimadas que podem passar por aquele segmento, incluindo os benefícios socio-ambientais estimados, e as características físicas daquele segmento.
+
+É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacionail de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf)), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel.
+
+### Até 5 km
+
+```{r mapa1}
+m1 = tmap::tm_shape(mapa11r, name = "Potencial 4% - direto") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 12,
+                   lwd_multiplier = 15,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  # "Em bicicleta 2018" = "Bike",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 4%" = "cyc",
+                                  # "Novos ciclistas 4%" = "new_cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0), #
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-linear_yl_rd_bk", n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   textNA = "sem info",
+                   title.col = "Tranquilidade"
+                   ) +
+  tmap::tm_shape(mapa11l, name = "Potencial 4% - seguro") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 12,
+                   lwd_multiplier = 15,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 4%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0), #
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-mako", n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   textNA = "sem info",
+                   title.col = "Tranquilidade"
+                   ) +
+  tmap::tm_shape(mapa12r, name = "Potencial 10% - direto") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 15,
+                   lwd_multiplier = 18,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 10%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0),
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-linear_yl_rd_bk",  n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   legend.col.show = FALSE,
+                   textNA = "sem info",
+                   title.col = "Tranquilidade",
+                   ) +
+  tmap::tm_shape(mapa12l, name = "Potencial 10% - seguro") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 15,
+                   lwd_multiplier = 18,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 10%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0),
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-mako",  n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   legend.col.show = FALSE,
+                   textNA = "sem info",
+                   title.col = "Tranquilidade",
+                   ) +
+  tm_view(view.legend.position = c("right", "bottom"),
+          control.position = c("right", "top")
+          # title = "Viagens até 5 km em bicicleta"
+          # leaflet.options = highlightOptions(color = "white", bringToFront = TRUE)
+          )
+m1leaf = tmap::tmap_leaflet(m1) %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
+  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
+                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
+                                              "Potencial 4% - seguro", "Potencial 10% - seguro"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro"))
+
+m1leaf
+
+htmlwidgets::saveWidget(m1leaf, file = paste0("biclarwww/", i,"/mapa_cenario1.html"))
+
+```
+
+Ver mapa em [ecrã inteiro](mapa_cenario1.html).
+
+> Este mapa apresenta os segmentos com mais de `r minimo1` viagens potenciais em bicicleta. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
+
+### Até 10 km (e-bike)
+
+```{r mapa2}
+m2 = tmap::tm_shape(mapa21r, name = "Potencial 4% - direto") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 12,
+                   lwd_multiplier = 15,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  # "Em bicicleta 2018" = "Bike",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 4%" = "cyc",
+                                  # "Novos ciclistas 4%" = "new_cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0), #
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-linear_yl_rd_bk", n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   textNA = "sem info",
+                   title.col = "Tranquilidade"
+                   ) +
+  tmap::tm_shape(mapa21l, name = "Potencial 4% - seguro") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 12,
+                   lwd_multiplier = 15,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 4%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0), #
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-mako", n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   textNA = "sem info",
+                   title.col = "Tranquilidade"
+                   ) +
+  tmap::tm_shape(mapa22r, name = "Potencial 10% - direto") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 15,
+                   lwd_multiplier = 18,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 10%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0),
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-linear_yl_rd_bk",  n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   legend.col.show = FALSE,
+                   textNA = "sem info",
+                   title.col = "Tranquilidade",
+                   ) +
+  tmap::tm_shape(mapa22l, name = "Potencial 10% - seguro") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 15,
+                   lwd_multiplier = 18,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 10%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0),
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-mako",  n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   legend.col.show = FALSE,
+                   textNA = "sem info",
+                   title.col = "Tranquilidade",
+                   ) +
+  tm_view(view.legend.position = c("right", "bottom"),
+          control.position = c("right", "top")
+          # title = "Viagens até 5 km em bicicleta"
+          # leaflet.options = highlightOptions(color = "white", bringToFront = TRUE)
+          )
+m2leaf = tmap::tmap_leaflet(m2) %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
+  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
+                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
+                                              "Potencial 4% - seguro", "Potencial 10% - seguro"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro"))
+
+m2leaf
+
+htmlwidgets::saveWidget(m2leaf, file = paste0("biclarwww/", i,"/mapa_cenario2.html"))
+
+```
+
+Ver mapa em [ecrã inteiro](mapa_cenario2.html).
+
+> Este mapa mostra os segmentos com mais de `r minimo2` viagens potenciais em bicicleta. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
+
+### Até 5 km intermodal
+
+```{r mapa3}
+m3 = tmap::tm_shape(mapa31r, name = "Potencial 4% - direto") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 12,
+                   lwd_multiplier = 15,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  # "Em bicicleta 2018" = "Bike",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 4%" = "cyc",
+                                  # "Novos ciclistas 4%" = "new_cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0), #
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-burg", n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   textNA = "sem info",
+                   title.col = "Tranquilidade"
+                   ) +
+  tmap::tm_shape(mapa31l, name = "Potencial 4% - seguro") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 12,
+                   lwd_multiplier = 15,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 4%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0), #
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-dark_mint", n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   textNA = "sem info",
+                   title.col = "Tranquilidade"
+                   ) +
+  tmap::tm_shape(mapa32r, name = "Potencial 10% - direto") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 15,
+                   lwd_multiplier = 18,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 10%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0),
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-burg",  n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   legend.col.show = FALSE,
+                   textNA = "sem info",
+                   title.col = "Tranquilidade",
+                   ) +
+  tmap::tm_shape(mapa32l, name = "Potencial 10% - seguro") +
+    tmap::tm_lines(id = "cyc",
+                   lwd = "cyc",
+                   alpha = 0.9,
+                   scale = 15,
+                   lwd_multiplier = 18,
+                   popup.vars = c("Viagens 2018" = "Total",
+                                  "Ciclistas cenário base" = "bike18",
+                                  "Ciclistas cenário 10%" = "cyc",
+                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
+                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
+                                  "Tranquilidade" = "quietness",
+                                  "Velocidade auto [km/h]" = "carspeed",
+                                  "Declive médio" = "slope"
+                                  ),
+                   popup.format = list(big.mark = " ", digits = 0),
+                   col = "quietness",
+                   palette = cols4all::c4a(palette = "-dark_mint",  n = 4),
+                   breaks = c(0, 25, 50, 75, 100),
+                   legend.format = list(text.separator = "a"),
+                   legend.col.show = FALSE,
+                   textNA = "sem info",
+                   title.col = "Tranquilidade",
+                   ) +
+  tm_view(view.legend.position = c("right", "bottom"),
+          control.position = c("right", "top")
+          # title = "Viagens até 5 km em bicicleta"
+          # leaflet.options = highlightOptions(color = "white", bringToFront = TRUE)
+          )
+m3leaf = tmap::tmap_leaflet(m3) %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
+  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
+                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
+                                              "Potencial 4% - seguro", "Potencial 10% - seguro"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro"))
+
+m3leaf
+
+htmlwidgets::saveWidget(m3leaf, file = paste0("biclarwww/", i,"/mapa_cenario3.html"))
+
+```
+
+Ver mapa em [ecrã inteiro](mapa_cenario3.html).
+
+> Este mapa mostra os segmentos com mais de `r minimo3` viagens potenciais em bicicleta. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
+
+## Resumo
 
 ```{r}
-#read data
-transit_all = sf::st_read("https://github.com/U-Shift/biclar/releases/download/0.0.1/transit.gpkg", quiet = TRUE)
+# tabela com potencial para cada cenario +  estimativas SE
+```
 
-library(tmap)
-tmap_mode("view")
+## Downloads
 
-tm_shape(transit_all) +
-  tm_sf(col = "mode", size = 0.3)
+```{r}
+download_folder = paste0("biclarwww/", i)
+
+path_sf_1r = paste0(download_folder, "/", i, "_rededireta_cenario1.gpkg")
+path_sf_1l = paste0(download_folder, "/", i, "_redesegura_cenario1.gpkg")
+
+path_sf_2r = paste0(download_folder, "/", i, "_rededireta_cenario2.gpkg")
+path_sf_2l = paste0(download_folder, "/", i, "_redesegura_cenario2.gpkg")
+
+path_sf_3r = paste0(download_folder, "/", i, "_rededireta_cenario3.gpkg")
+path_sf_3l = paste0(download_folder, "/", i, "_redesegura_cenario3.gpkg")
+
 ```
+
+Os mapas acima estão disponíveis para download, em formato _GeoPackage_ (.gpkg), que pode ser ligo em qualquer SIG moderno.
+É recomendável o software de acesso livre [QGIS](https://qgis.org/pt_PT/site/) para leitura, visualização e análise de dados em detalhe.
+
+1.    **Viagens até 5 km em bicicleta**
+    *   [Rede direta](path_sf_1r)
+    *   [Rede segura](path_sf_1l)
+2.    **Viagens até 10 km em bicicleta elétrica**
+    *   [Rede direta](path_sf_2r)
+    *   [Rede segura](path_sf_2l)
+3.    **Viagens intermodais até 5 km em bicicleta**
+    *   [Rede direta](path_sf_3r)
+    *   [Rede segura](path_sf_3l)
+
+Para cada município, os dados disponíveis para download referem-se à rede ciclável potencial com um _buffer_ de 300 metros para além dos limites do município, para que sejam consideradas as ligações ao município vizinho.
+( _Excepto no cenário 3_)
+
+A rede disponibilizada poderá não corresponder à que é apresentada nos mapas acima, que mostram apenas os segmentos com mais de X viagens potenciais em bicicleta.
+Para visualizar os segmentos dependendo do número de viagens potenciais, utilizar os filtros do SIG, por exemplo, para visualizar só os segmentos com mais de 500 viagens, ou todos os segmentos com mais de 10 viagens em bicileta.

### 2022-12-03 code/biclarwww/build.R

@@ -17,12 +17,12 @@ rmarkdown::render(input = "code/biclarwww/index.Rmd", output_dir = "biclarwww", 
 
 # code to build the site
 
-# i = "lisboa"
+i = "barreiro"
 
 for (i in municipios){
   
   pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio
-  dir.create(pub_folder) #comment this one??
+  # dir.create(pub_folder) #comment this one??
   rmarkdown::render(input = "code/biclarwww/municipio_index.Rmd", output_dir = pub_folder, output_file = "index.html")
   # rmarkdown::render(input = "stats.Rmd", output_dir = pub_folder, output_file = "stats.html")
   # rmarkdown::render(input = "baseline.Rmd", output_dir = pub_folder, output_file = "baseline.html")

### 2022-12-03 code/test-code/color_pick.R

@@ -12,3 +12,11 @@ biclar::tm_rnet(rnet_test[MUNICIPIOSgeo %>% filter(Concelho == "Barreiro"), , op
                 lwd_multiplier = 15)
 
 
+
+rnet_test = sf::st_read("export_rnet/rnet_aml_1410_50.gpkg")
+biclar::tm_rnet(rnet_test,
+                lwd = "cyc", #Baseline, ENMAC4, ENMAC10
+                col = "quietness",
+                palette = "-yl_gn_bu", # "linear_yl_rd_bk", "mako", "burg", "reds" - reds for fastest, mako for quietest
+                scale = 15, #30 para lisboa?
+                lwd_multiplier = 15)

### 2022-12-03 code/test-code/r5r_filter_overline_routes_CORRECT.R

@@ -903,7 +903,7 @@ rnet_allmodesNSub3_overline_morethan50_clean = rnet_allmodesNSub3_overline_moret
   filter(group %in% groupp) #more than 10 segments connected
 rm(rnet_allmodesNSub3_overline_morethan50_group, group)
 
-tm_rnet(rnet_allmodesNSub3_overline_morethan50_clean,
+tm_rnet(rnet_allmodesNSub3_overline_morethan30_clean,
         lwd = "Bike", #Baseline, ENMAC4, ENMAC10
         # col = "Quietness",
         palette = "-burg", # "linear_yl_rd_bk" "johnson", "mako", "burg", "reds" - reds for fastest, mako for quietest

### 2022-12-04 code/biclarwww/index.Rmd
minor adjustments website
solve Vila Franca Xira

@@
 ))
+
+municipios_index$url[Municipio == "Vila Franca de Xira"] = paste0("<a href='vfxira/' target='_blank'>",  Municipio,  "</a>")
+
 municipios_index = sf::st_as_sf(municipios_index)
+
 ```
@@
 test content
+
+```{r}
+#read data
+transit_all = sf::st_read("https://github.com/U-Shift/biclar/releases/download/0.0.1/transit.gpkg", quiet = TRUE)
+
+library(tmap)
+tmap_mode("view")
+
+tm_shape(transit_all) +
+  tm_sf(col = "mode", size = 0.3)
+```
 
@@
 
-------------------------------------------------------------------------
+<!-- ------------------------------------------------------------------------ -->
 
@@
 *   Conselho Consultuvo: ....
+*   Matheus Correia, OpenStreetMap Portugal
 *   CycleStreets

### 2022-12-04 code/biclarwww/build.R

@@ -21,7 +21,11 @@ i = "barreiro"
 
 for (i in municipios){
   
-  pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio
+  if (i = "vila franca de xira"){
+    pub_folder = paste0("biclarwww/", "vfxira")
+      } else {
+    pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio
+  }
   # dir.create(pub_folder) #comment this one??
   rmarkdown::render(input = "code/biclarwww/municipio_index.Rmd", output_dir = pub_folder, output_file = "index.html")
   # rmarkdown::render(input = "stats.Rmd", output_dir = pub_folder, output_file = "stats.html")

### 2022-12-04 code/biclarwww/municipio_index.Rmd
lower threshold for Palmela

@@
 
-if (i %in% c("mafra", "sesimbra", "alcochete")){
+if (i %in% c("mafra", "sesimbra", "alcochete", "palmela")){
 
@@
 ```{r minimos}
-if (i %in% c("mafra", "sesimbra", "alcochete")){ 
+if (i %in% c("mafra", "sesimbra", "alcochete", "palmela")){ 
   minimo1 = 15} else {
@@
 
-if (i %in% c("mafra", "sesimbra", "alcochete")){ 
+if (i %in% c("mafra", "sesimbra", "alcochete", "palmela")){ 
   minimo2 = 50} else if (i %in% c("almada", "oeiras", "lisboa", "odivelas", "amadora")){

### 2022-12-05 code/biclarwww/municipio_index.Rmd
website plots

@@
 
-```{r setup, include=FALSE}
+```{r setup, include=FALSE, cache=FALSE}
 library(dplyr)
@@
                       message = FALSE,
-                      warning = FALSE)
+                      cache = TRUE, #hide if needed
+                      warning = FALSE
+                      )
+
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) #to use the project root as the working directory
+
+options(encoding = "UTF-8")
 
@@
 ```{r chamadas}
-i = "barreiro" #delete this when running for all!!!
+# i = "barreiro" #delete this when running for all!!!
 mun = stringr::str_to_title(i)
@@
     i = "vfxira"
+    mun = "Vila Franca de Xira"
   }
@@
 
-```{r}
+```{r dados}
+TRIPSmode_municipio = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/TRIPSmode_municipal.Rds"))
+
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Car"] = "Automovel"
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Bike"] = "Bicicleta"
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Motorcycle"] = "Motociclo"
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Walk"] = "A Pe"
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Transit"] = "Transportes Publicos"
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Other"] = "Outro"
+
+# TRIPSmode_municipio$Origem[TRIPSmode_municipio$Origem == "Setubal"] = "Setúbal" #resolver o encoding!
+# TRIPSmode_municipio$Destino[TRIPSmode_municipio$Destino == "Setubal"] = "Setúbal"
+
+TRIPSmode_mun = TRIPSmode_municipio %>%
+  filter(Origem == mun | Destino == mun) %>% 
+  group_by(modo) %>%
+  summarise(viagens = round(sum(viagens))) %>% 
+  mutate(percentagem = round(100*viagens/sum(viagens),1))
+
+TRIPSmode_mun_intr = TRIPSmode_municipio %>%
+  filter(Origem == mun | Destino == mun)
+TRIPSmode_mun_intr$Inter[TRIPSmode_mun_intr$Origem == TRIPSmode_mun_intr$Destino] = "Intramunicipal"
+TRIPSmode_mun_intr$Inter[is.na(TRIPSmode_mun_intr$Inter)] = "Intermunicipal"
+TRIPSmode_mun_intr = TRIPSmode_mun_intr %>% 
+  group_by(modo, Inter) %>%
+  summarise(viagens = round(sum(viagens)))
+
+viagensinter = sum(TRIPSmode_mun_intr$viagens[TRIPSmode_mun_intr$Inter == "Intermunicipal"])
+viagensintra = sum(TRIPSmode_mun_intr$viagens[TRIPSmode_mun_intr$Inter == "Intramunicipal"])
+
+TRIPSmode_mun_intr$percentagem[TRIPSmode_mun_intr$Inter == "Intermunicipal"] =  round(100*TRIPSmode_mun_intr$viagens[TRIPSmode_mun_intr$Inter == "Intermunicipal"]/viagensinter, 1)
+
+TRIPSmode_mun_intr$percentagem[TRIPSmode_mun_intr$Inter == "Intramunicipal"] =  round(100*TRIPSmode_mun_intr$viagens[TRIPSmode_mun_intr$Inter == "Intramunicipal"]/viagensintra, 1)
+
+TRIPSmode_mun_intr = bind_rows(TRIPSmode_mun %>% mutate(Inter = "Todas"),  TRIPSmode_mun_intr)
+
+```
+
+```{r makemaps}
 MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))
@@
 
-mapa21r = sf::st_read("export_rnet/rnet_aml_1404_50.gpkg", quiet = TRUE)
+mapa21r = sf::st_read("export_rnet/rnet_aml_2404_50.gpkg", quiet = TRUE)
 mapa21r = mapa21r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa21l = sf::st_read("export_rnet/rnet_aml_1304_50.gpkg", quiet = TRUE)
+mapa21l = sf::st_read("export_rnet/rnet_aml_2304_50.gpkg", quiet = TRUE)
 mapa21l = mapa21l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa22r = sf::st_read("export_rnet/rnet_aml_1410_50.gpkg", quiet = TRUE)
+mapa22r = sf::st_read("export_rnet/rnet_aml_2410_50.gpkg", quiet = TRUE)
 mapa22r = mapa22r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa22l = sf::st_read("export_rnet/rnet_aml_1310_50.gpkg", quiet = TRUE)
+mapa22l = sf::st_read("export_rnet/rnet_aml_2310_50.gpkg", quiet = TRUE)
 mapa22l = mapa22l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
 
-mapa21r = sf::st_read("export_rnet/rnet_aml_1404_200.gpkg", quiet = TRUE)
+mapa21r = sf::st_read("export_rnet/rnet_aml_2404_200.gpkg", quiet = TRUE)
 mapa21r = mapa21r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa21l = sf::st_read("export_rnet/rnet_aml_1304_200.gpkg", quiet = TRUE)
+mapa21l = sf::st_read("export_rnet/rnet_aml_2304_200.gpkg", quiet = TRUE)
 mapa21l = mapa21l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa22r = sf::st_read("export_rnet/rnet_aml_1410_200.gpkg", quiet = TRUE)
+mapa22r = sf::st_read("export_rnet/rnet_aml_2410_200.gpkg", quiet = TRUE)
 mapa22r = mapa22r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa22l = sf::st_read("export_rnet/rnet_aml_1310_200.gpkg", quiet = TRUE)
+mapa22l = sf::st_read("export_rnet/rnet_aml_2310_200.gpkg", quiet = TRUE)
 mapa22l = mapa22l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
 
-mapa21r = sf::st_read("export_rnet/rnet_aml_1404_100.gpkg", quiet = TRUE)
+mapa21r = sf::st_read("export_rnet/rnet_aml_2404_100.gpkg", quiet = TRUE)
 mapa21r = mapa21r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa21l = sf::st_read("export_rnet/rnet_aml_1304_100.gpkg", quiet = TRUE)
+mapa21l = sf::st_read("export_rnet/rnet_aml_2304_100.gpkg", quiet = TRUE)
 mapa21l = mapa21l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa22r = sf::st_read("export_rnet/rnet_aml_1410_100.gpkg", quiet = TRUE)
+mapa22r = sf::st_read("export_rnet/rnet_aml_2410_100.gpkg", quiet = TRUE)
 mapa22r = mapa22r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
          bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa22l = sf::st_read("export_rnet/rnet_aml_1310_100.gpkg", quiet = TRUE)
+mapa22l = sf::st_read("export_rnet/rnet_aml_2310_100.gpkg", quiet = TRUE)
 mapa22l = mapa22l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
 
-`r proposicao` `r mun`, as viagens são assim.
+Segundo o Inquérito à Mobilidade  ([IMOB](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt)) realizado em 2018 pelo INE, as viagens com origem e/ou destino `r stringr::str_to_lower(proposicao)` `r mun` têm as seguintes características.
+
+```{r mododist, fig.ncol=2, out.width="50%"}
+# gráfico com caracterização viagens IMOB intra municipais e intermonicipais
+
+TRIPSmode_mun_intr$modo = factor(TRIPSmode_mun_intr$modo, 
+                                 levels = c("Automovel", "Transportes Publicos", "Motociclo", 
+                                            "Outro", "A Pe", "Bicicleta"))
+
+col_modes = c("#fe5f55", "#ffd166", "#8E794A", "grey", "#457b9d", "#90be6d")
+
+ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = percentagem/100, x = Inter)) +
+  geom_bar(position = position_fill(reverse = FALSE), stat = "identity") +
+  scale_fill_manual(values = col_modes, name = "Modo") +
+  scale_y_continuous(labels = scales::percent) +
+  geom_text(aes(label = ifelse(percentagem > 6, paste0(percentagem,"%"), "")),
+    position = position_stack(vjust = 0.5, reverse = FALSE))+
+  theme_minimal() +
+  theme(axis.title.y = element_blank(), axis.title.x = element_blank(), legend.title = element_blank(), legend.position = "none") +
+  coord_flip()
+
+ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = viagens, x = Inter)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(values = col_modes, name = "Modo") +
+  scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
+  geom_text(aes(label = ifelse(viagens/1000 > 10, paste0(round(viagens/1000)," k"), "")),
+    position = position_stack(vjust = 0.5, reverse = FALSE))+
+  theme_minimal() +
+  theme(axis.title.y = element_blank(), legend.title = element_text("Modo"), legend.position = "right") +
+  coord_flip()
+
+```
+
+Das cerca de `r fmt_num(round(sum(TRIPSmode_mun$viagens)/1000))`mil viagens diárias `r stringr::str_to_lower(proposicao)` `r mun`, apenas **`r TRIPSmode_mun[3,3]`% das viagens são realizadas em bicicleta** (`r round(TRIPSmode_mun[3,2])`).
+`r round(TRIPSmode_mun[2,2]/1000)`mil são feitas em automóvel (`r TRIPSmode_mun[2,3]`%), `r round(TRIPSmode_mun[1,2]/1000)`mil a pé (`r TRIPSmode_mun[1,3]`%), `r round(TRIPSmode_mun[6,2]/1000)`mil em Transportes Públicos (`r TRIPSmode_mun[6,3]`%) e `r round(TRIPSmode_mun[4,2]+TRIPSmode_mun[5,2])` em outros modos (`r TRIPSmode_mun[4,3]+TRIPSmode_mun[5,3]`%).
+
+Diariamente, cerca de `r fmt_num(round(viagensintra/1000))`mil viagens têm origem e destino `r stringr::str_to_lower(proposicao)` `r mun`, e cerca de `r fmt_num(round(viagensinter/1000))`mil viagens têm origem ou destino em municípios diferentes.
+
+## Rede ciclável potencial
 
@@
 <!-- Passar isto para FAQ? -->
+
+<details>
+  <summary>**Como ler estes mapas?**</summary>
 Os mapas apresentam a informação gráfica para duas redes - a rede ciclável para viagens mais diretas, e a rede para viagens mais seguras e tranquilas (conforme a rede viária existente).  
@@
 É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacionail de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf)), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel.
+</details>
 
-### Até 5 km
+#### Até 5 km
 
-```{r mapa1}
+Este mapa apresenta os segmentos com mais de `r minimo1` viagens potenciais em bicicleta. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
+
+```{r mapa1, cache=FALSE}
 m1 = tmap::tm_shape(mapa11r, name = "Potencial 4% - direto") +
@@
 
-> Este mapa apresenta os segmentos com mais de `r minimo1` viagens potenciais em bicicleta. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
+#### Até 10 km (e-bike)
 
-### Até 10 km (e-bike)
+Este mapa mostra os segmentos com mais de `r minimo2` viagens potenciais em bicicleta elétrica. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
 
-```{r mapa2}
+```{r mapa2, cache=FALSE}
 m2 = tmap::tm_shape(mapa21r, name = "Potencial 4% - direto") +
@@
                    col = "quietness",
-                   palette = cols4all::c4a(palette = "-linear_yl_rd_bk", n = 4),
+                   palette = cols4all::c4a(palette = "-reds3", n = 4),
                    breaks = c(0, 25, 50, 75, 100),
@@
                    col = "quietness",
-                   palette = cols4all::c4a(palette = "-mako", n = 4),
+                   palette = cols4all::c4a(palette = "-hokusai2", n = 4),
                    breaks = c(0, 25, 50, 75, 100),
@@
                    col = "quietness",
-                   palette = cols4all::c4a(palette = "-linear_yl_rd_bk",  n = 4),
+                   palette = cols4all::c4a(palette = "-reds3",  n = 4),
                    breaks = c(0, 25, 50, 75, 100),
@@
                    col = "quietness",
-                   palette = cols4all::c4a(palette = "-mako",  n = 4),
+                   palette = cols4all::c4a(palette = "-hokusai2",  n = 4),
                    breaks = c(0, 25, 50, 75, 100),
@@
 
-> Este mapa mostra os segmentos com mais de `r minimo2` viagens potenciais em bicicleta. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
+#### Até 5 km intermodal
 
-### Até 5 km intermodal
+Este mapa mostra os segmentos com mais de `r minimo3` viagens potenciais em bicicleta, em que parte da viagem é realizada em transportes públicos (barco, comboio, metro sul do Tejo, ou carris metropolitana).
+Os segmentos deste mapa dizem respeito apenas à parte da viagem que é realizada em bicicleta, num total de até 5 km no conjunto da primeira e última etapa da viagem intermodal.
+Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
 
-```{r mapa3}
+```{r mapa3, cache=FALSE}
 m3 = tmap::tm_shape(mapa31r, name = "Potencial 4% - direto") +
@@
 
-> Este mapa mostra os segmentos com mais de `r minimo3` viagens potenciais em bicicleta. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
-
 ## Resumo
 
-```{r}
+Tabela com potencial para cada cenário + estimativas impactes socio-ambientais.
+
+```{r tabela}
 # tabela com potencial para cada cenario +  estimativas SE
+
+DT::datatable(head(iris),
+              rownames = FALSE,
+              filter = 'none',
+              options = list(dom = "t",
+                               pageLength = 5,
+                               responsive = TRUE,
+                              extensions = c('Responsive'))
+)
 ```
@@
 
-```{r}
-download_folder = paste0("biclarwww/", i)
+```{r downloads}
 
-path_sf_1r = paste0(download_folder, "/", i, "_rededireta_cenario1.gpkg")
-path_sf_1l = paste0(download_folder, "/", i, "_redesegura_cenario1.gpkg")
+path_sf_1r = paste0(i, "_rededireta_cenario1.gpkg") #how works best?
+path_sf_1l = paste0(i, "_redesegura_cenario1.gpkg")
 
-path_sf_2r = paste0(download_folder, "/", i, "_rededireta_cenario2.gpkg")
-path_sf_2l = paste0(download_folder, "/", i, "_redesegura_cenario2.gpkg")
+path_sf_2r = paste0(i, "_rededireta_cenario2.gpkg")
+path_sf_2l = paste0(i, "_redesegura_cenario2.gpkg")
 
-path_sf_3r = paste0(download_folder, "/", i, "_rededireta_cenario3.gpkg")
-path_sf_3l = paste0(download_folder, "/", i, "_redesegura_cenario3.gpkg")
+path_sf_3r = paste0(i, "_rededireta_cenario3.gpkg")
+path_sf_3l = paste0(i, "_redesegura_cenario3.gpkg")
 
@@
 1.    **Viagens até 5 km em bicicleta**
-    *   [Rede direta](path_sf_1r)
-    *   [Rede segura](path_sf_1l)
+
+  *   [Rede direta](`r path_sf_1r`)
+  *   [Rede segura](`r path_sf_1l`)
+
 2.    **Viagens até 10 km em bicicleta elétrica**
-    *   [Rede direta](path_sf_2r)
-    *   [Rede segura](path_sf_2l)
+
+  *   [Rede direta](`r path_sf_2r`)
+  *   [Rede segura](`r path_sf_2l`)
+
 3.    **Viagens intermodais até 5 km em bicicleta**
-    *   [Rede direta](path_sf_3r)
-    *   [Rede segura](path_sf_3l)
 
-Para cada município, os dados disponíveis para download referem-se à rede ciclável potencial com um _buffer_ de 300 metros para além dos limites do município, para que sejam consideradas as ligações ao município vizinho.
-( _Excepto no cenário 3_)
+  *   [Rede direta](`r path_sf_3r`)
+  *   [Rede segura](`r path_sf_3l`)
+
+Para cada município, os dados disponíveis para download referem-se à rede ciclável potencial com uma margem de 300 metros para além dos limites do município, para que sejam consideradas as ligações ao município vizinho.
+(_Exceto no cenário 3_)
 

### 2022-12-05 code/biclarwww/build.R

@@ -4,18 +4,18 @@
 # Read in data --------------------------------------------------------------------------------
 
 
+# página principal
+rmarkdown::render(input = "code/biclarwww/index.Rmd", output_dir = "biclarwww", output_file = "index.html") #run this
 
 
+# página da AML - demora cerca de 5min a render. Agora sem 2 mapas é rápido (1min)
+rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", output_dir = "biclarwww/aml", output_file = "index.html") #run this
 
 
 
 # Create outputs for each region --------------------------------------------------------------
 
 
-rmarkdown::render(input = "code/biclarwww/index.Rmd", output_dir = "biclarwww", output_file = "index.html") #run this
-
-
-# code to build the site
 
 i = "barreiro"
 
@@ -34,4 +34,5 @@ for (i in municipios){
   # rmarkdown::render(input = "downloads.Rmd", output_dir = pub_folder, output_file = "downloads.html")
   
 }
+
 

### 2022-12-05 code/test-code/simple_statistics.R

@@ -49,3 +49,174 @@ p1 = ggplot(routes_quietest_grouped) +
 ggplotly(p1) 
 
 
+
+# prepare for website -------------------------------------------------------------------------
+
+# routes1r = readRDS("routes_ferry4_preoverline.Rds") #não. o que se quer mostrar é viagens também com mais de 5km
+routes1r = readRDS("routes_r5r_100jit_lts4__ferry_elev_raw.Rds")
+# routes2r = readRDS("routes_ferry4_ebike_preoverline.Rds") #não faz sentido pois é igual ao primeiro
+routes3r = readRDS("routes_r5r_100jit_lts4__intermodALL_NoSub_elev_raw.Rds")
+
+# ENMAC4 = 0.04 # 4%
+# ENMAC10 = 0.10 # 10#
+
+
+routes1r = routes1r %>% 
+  sf::st_drop_geometry() %>% 
+  ungroup() %>% 
+  filter(mode == "BICYCLE") %>% 
+  group_by(id) %>% 
+  summarise(Bike = Bike,
+            Car = Car + CarP,
+            Total = Total,
+            distance = sum(distance),
+            Bikeper = Bike / Total) %>% 
+  ungroup() %>%
+  mutate(
+    cyc4 = ifelse(Bikeper >= ENMAC4, Bike, ENMAC4 * Total),
+    new_cyc4 = ifelse(Bikeper >= ENMAC4, 0, cyc4 - Bike),
+    cyc10 = ifelse(Bikeper >= ENMAC10, Bike, ENMAC10 * Total),
+    new_cyc10 = ifelse(Bikeper >= ENMAC10, 0, cyc10 - Bike))
+
+
+#filter by municipio
+
+routes1r_plot = routes1r %>% 
+  mutate(length_km = distance/1000) %>% 
+  mutate(distance_band = cut(length_km, 0:20)) %>% 
+  group_by(distance_band) %>% 
+  summarise(cyc_dist_km = mean(length_km),
+            trips_cyc_baseline_1 = sum(Bike),
+            trips_cyc_enmac4_1 = sum(cyc4),
+            trips_cyc_enmac10_1 = sum(cyc10),
+            trips_all_baseline_1 = sum(Total)) %>% 
+  mutate_if(is.numeric, round, digits = 2) %>% 
+  mutate_at(c(3:6), round, digits = 0)
+
+
+
+
+routes3rid = routes3r %>% 
[…]

### 2022-12-06 code/biclarwww/municipio_index.Rmd
tables in www

@@
 
+if (i == "setubal"){
+  mun = "Setúbal"
+}
+
+dontrun = i %in% c("alcochete", "mafra", "montijo") # não mostrar mapa 3
+
 proposicao = ifelse(mun %in% c("Amadora", "Moita"), "Na",
@@
 
-TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Car"] = "Automovel"
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Car"] = "Automóvel"
 TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Bike"] = "Bicicleta"
 TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Motorcycle"] = "Motociclo"
-TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Walk"] = "A Pe"
-TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Transit"] = "Transportes Publicos"
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Walk"] = "A Pé"
+TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Transit"] = "Transportes Públicos"
 TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Other"] = "Outro"
 
-# TRIPSmode_municipio$Origem[TRIPSmode_municipio$Origem == "Setubal"] = "Setúbal" #resolver o encoding!
-# TRIPSmode_municipio$Destino[TRIPSmode_municipio$Destino == "Setubal"] = "Setúbal"
+TRIPSmode_municipio$Origem[TRIPSmode_municipio$Origem == "Setubal"] = "Setúbal" # por causa do acento
+TRIPSmode_municipio$Destino[TRIPSmode_municipio$Destino == "Setubal"] = "Setúbal"
 
@@
 TRIPSmode_mun_intr$modo = factor(TRIPSmode_mun_intr$modo, 
-                                 levels = c("Automovel", "Transportes Publicos", "Motociclo", 
-                                            "Outro", "A Pe", "Bicicleta"))
+                                 levels = c("Automóvel", "Transportes Públicos", "Motociclo", 
+                                            "Outro", "A Pé", "Bicicleta"))
 
@@
 
-htmlwidgets::saveWidget(m1leaf, file = paste0("biclarwww/", i,"/mapa_cenario1.html"))
-
 ```
@@
 
-htmlwidgets::saveWidget(m2leaf, file = paste0("biclarwww/", i,"/mapa_cenario2.html"))
-
 ```
@@
 
-```{r mapa3, cache=FALSE}
+```{r mapa3, cache=FALSE, eval=dontrun}
 m3 = tmap::tm_shape(mapa31r, name = "Potencial 4% - direto") +
@@
 
-htmlwidgets::saveWidget(m3leaf, file = paste0("biclarwww/", i,"/mapa_cenario3.html"))
-
 ```
@@
 
+```{r exportmaps}
+htmlwidgets::saveWidget(m1leaf, file = paste0("biclarwww/", i,"/mapa_cenario1.html"))
+
+htmlwidgets::saveWidget(m2leaf, file = paste0("biclarwww/", i,"/mapa_cenario2.html"))
+
+htmlwidgets::saveWidget(m3leaf, file = paste0("biclarwww/", i,"/mapa_cenario3.html"))
+```
+
 ## Resumo
 
-Tabela com potencial para cada cenário + estimativas impactes socio-ambientais.
+A tabela mostra o potencial ciclável `r stringr::str_to_lower(proposicao)` `r mun` em número de novas viagens em bicicleta potenciais, para cada meta da Estratégia Nacional ENMAC, para cada cenário, e para cada tipo de rotas otimizadas.
+
+Para cada um destes cenários, é possível ver também de forma agregada qual o potencial socio-ambiental estimado, em toneladas de CO~2~eq evitado anualmente, e em benefícios sociais.
+
+<details>
+  <summary>**Como são estimados os benefícios socio-ambientais?**</summary>
+Os benefícios socio-ambientais são estimados em duas vertentes, utilizando os métodos e a ferramenta [_HEAT for Cycling_](https://www.heatwalkingcycling.org/), da Organização Mundial de Saúde.
+
+A componente ambiental é medida em toneladas de CO~2~eq evitado, por transferência de viagens equivalentes em automóvel, com a mesma origem e destino, mas não necessariamente com o mesmo percurso.
+
+Os impactes sociais agregados, na componente da saúde, são estimados em termos de redução/aumento da mortalidade por aumento de atividade física, por exposição à poluição atomosférica, e por exposição ao risco de sinistralidade rodoviária.
+Este valor é por fim monetizado utilizando o _Valor Estatístico da Vida_ (€3.055.358, segundo [ANSR 2021](http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O%20Impacto%20Economico%20e%20Social%20da%20Sinistralidade%20-%20PT.pdf)), estimado para 10 anos, com uma taxa de desconto de 5% e inflação de 3%.
+</details>
 
@@
 
-DT::datatable(head(iris),
+tabelas_mun = readRDS("HEAT/tabelas_mun.Rds")
+tabelas_mun = tabelas_mun %>% filter(Municipio == mun) %>%
+  mutate(Percursos = recode(LTS, `3` = "seguro", `4` = "direto")) %>% 
+  select(c(4,2,11,6:10)) %>% 
+  arrange(Meta_ENMAC)
+
+DT::datatable(tabelas_mun,
+                filter = 'none',
                 rownames = FALSE,
-              filter = 'none',
+                colnames = c("Meta ENMAC", "Cenário", "Tipo de Percurso", "Total Viagens /dia",
+                             "Viagens Bicicleta 2018", "Novas Viagens Bicicleta",
+                             "CO2eq evitado (ton/ano)", "Benefícios estimados em 10 anos (milhares €)"),
+                extensions = "RowGroup",
                 options = list(dom = "t",
-                               pageLength = 5,
+                               pageLength = 12,
                                responsive = TRUE,
-                              extensions = c('Responsive'))
-)
+                               rowGroup = list(dataSrc = 0), #agrupar a primeira coluna
+                               extensions = c('Responsive'))) %>% 
+    DT::formatRound(c(4:8),0, mark = " ")
+
+cicmacx4 = max(tabelas_mun$Bike_new[tabelas_mun$Meta_ENMAC=="4%"])
+cicmacx10 = max(tabelas_mun$Bike_new[tabelas_mun$Meta_ENMAC=="10%"])
+co2macx4 = max(tabelas_mun$CO2eq[tabelas_mun$Meta_ENMAC=="4%"])
+co2macx10 = max(tabelas_mun$CO2eq[tabelas_mun$Meta_ENMAC=="10%"])
+euromacx4 = max(tabelas_mun$Economic10[tabelas_mun$Meta_ENMAC=="4%"])
+euromacx10 = max(tabelas_mun$Economic10[tabelas_mun$Meta_ENMAC=="10%"])
+
 ```
+
+`r proposicao` `r mun`, a **meta de 4%** de ciclistas diários equivale a ter **`r fmt_num(cicmacx4)` ciclistas** em viagens até 10 km, com uma redução potencial de **`r fmt_num(co2macx4)` toneladas** de CO~2~eq por ter menos viagens em automóvel na mesma razão, com potenciais benefícios para a sociedade de **`r fmt_num(euromacx4/1000)` milhões €** ao fim de 10 anos.
+
+Já a **meta de 10%** de ciclistas diários equivale a ter **`r fmt_num(cicmacx10)` ciclistas** em viagens até 10 km, com uma redução potencial de **`r fmt_num(co2macx10)` toneladas** de CO~2~eq por ter menos viagens em automóvel na mesma razão, com potenciais benefícios para a sociedade de **`r fmt_num(euromacx10/1000)` milhões €** ao fim de 10 anos.
 
@@
 
-A rede disponibilizada poderá não corresponder à que é apresentada nos mapas acima, que mostram apenas os segmentos com mais de X viagens potenciais em bicicleta.
+`r emo::ji("magnifying")` A rede disponibilizada poderá não corresponder à que é apresentada nos mapas acima, que mostram apenas os segmentos com mais de X viagens potenciais em bicicleta.
 Para visualizar os segmentos dependendo do número de viagens potenciais, utilizar os filtros do SIG, por exemplo, para visualizar só os segmentos com mais de 500 viagens, ou todos os segmentos com mais de 10 viagens em bicileta.

### 2022-12-06 code/biclarwww/build.R

@@ -5,11 +5,11 @@
 
 
 # página principal
-rmarkdown::render(input = "code/biclarwww/index.Rmd", output_dir = "biclarwww", output_file = "index.html") #run this
+rmarkdown::render(input = "code/biclarwww/index.Rmd", encoding = "UTF-8", output_dir = "biclarwww", output_file = "index.html") #run this
 
 
 # página da AML - demora cerca de 5min a render. Agora sem 2 mapas é rápido (1min)
-rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", output_dir = "biclarwww/aml", output_file = "index.html") #run this
+rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", encoding = "UTF-8", output_dir = "biclarwww/aml", output_file = "index.html") #run this
 
 
 
@@ -27,7 +27,7 @@ for (i in municipios){
     pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio
   }
   # dir.create(pub_folder) #comment this one??
-  rmarkdown::render(input = "code/biclarwww/municipio_index.Rmd", output_dir = pub_folder, output_file = "index.html")
+  rmarkdown::render(input = "code/biclarwww/municipio_index.Rmd", encoding = "UTF-8", output_dir = pub_folder, output_file = "index.html")
   # rmarkdown::render(input = "stats.Rmd", output_dir = pub_folder, output_file = "stats.html")
   # rmarkdown::render(input = "baseline.Rmd", output_dir = pub_folder, output_file = "baseline.html")
   # rmarkdown::render(input = "route-types.Rmd", output_dir = pub_folder, output_file = "route-types.html")

### 2022-12-06 code/test-code/haas_stats.R

@@ -88,8 +88,6 @@ sum(HEAT_aml_PPP_redux_143$Economic) #37882000
 format(sum(HEAT_aml_PPP_redux_143$Economic10), scientific=F) #336.770.000
 
 
-
-
 HEAT_municipios_PPP_redux_2104 = HEAT_municipios_PPP_redux %>%
   filter(Estrategia == "ate 10km e-bike",
          Meta_ENMAC == "10%",
@@ -109,3 +107,8 @@ sum(as.numeric(gsub(" ","",HEAT_aml_PPP_redux_2104$Mortality))) #90
 sum(as.numeric(gsub(" ","",HEAT_aml_PPP_redux_2104$CO2eq))) #65420
 sum(HEAT_aml_PPP_redux_2104$Economic) #284690000
 format(sum(HEAT_aml_PPP_redux_2104$Economic10), scientific=F) #2.537.000.000
+
+
+
+
+saveRDS(HEAT_municipios_PPP_redux, "HEAT/HEAT_municipios_123_redux.Rds")

### 2022-12-06 code/biclarwww/municipio_index.Rmd
conditional eval
#35

@@
 
-dontrun = i %in% c("alcochete", "mafra", "montijo") # não mostrar mapa 3
+dontrun = !i %in% c("alcochete", "mafra", "montijo") # não mostrar mapa 3
+run = i %in% c("alcochete", "mafra", "montijo") # mostrar mensagem mapa 3
 
@@
 
+```{r conditional_block, echo=FALSE, results='asis', eval=run}
+cat("
+
+> Lamentamos, mas nesta área não existe rede ciclável potencial para o cenário 3, com as caraceterísticas definidas.")
+```
+
 Ver mapa em [ecrã inteiro](mapa_cenario3.html).
@@
 
-cicmacx4 = max(tabelas_mun$Bike_new[tabelas_mun$Meta_ENMAC=="4%"])
-cicmacx10 = max(tabelas_mun$Bike_new[tabelas_mun$Meta_ENMAC=="10%"])
-co2macx4 = max(tabelas_mun$CO2eq[tabelas_mun$Meta_ENMAC=="4%"])
-co2macx10 = max(tabelas_mun$CO2eq[tabelas_mun$Meta_ENMAC=="10%"])
-euromacx4 = max(tabelas_mun$Economic10[tabelas_mun$Meta_ENMAC=="4%"])
-euromacx10 = max(tabelas_mun$Economic10[tabelas_mun$Meta_ENMAC=="10%"])
+cicmacx4 = max(tabelas_mun$Bike_new[tabelas_mun$Meta_ENMAC=="4%"], na.rm = TRUE)
+cicmacx10 = max(tabelas_mun$Bike_new[tabelas_mun$Meta_ENMAC=="10%"], na.rm = TRUE)
+co2macx4 = max(tabelas_mun$CO2eq[tabelas_mun$Meta_ENMAC=="4%"], na.rm = TRUE)
+co2macx10 = max(tabelas_mun$CO2eq[tabelas_mun$Meta_ENMAC=="10%"], na.rm = TRUE)
+euromacx4 = max(tabelas_mun$Economic10[tabelas_mun$Meta_ENMAC=="4%"], na.rm = TRUE)
+euromacx10 = max(tabelas_mun$Economic10[tabelas_mun$Meta_ENMAC=="10%"], na.rm = TRUE)
 

### 2022-12-06 code/biclarwww/build.R

@@ -21,7 +21,7 @@ i = "barreiro"
 
 for (i in municipios){
   
-  if (i = "vila franca de xira"){
+  if (i == "vila franca de xira"){
     pub_folder = paste0("biclarwww/", "vfxira")
       } else {
     pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio

### 2022-12-06 code/biclarwww/index.Rmd
website plots
try new tab
table and disable responsive

@@
 
+knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) #to use the project root as the working directory
+
 library(sf)
@@
 library(htmlwidgets)
+
+fmt_num = function(x) format(round(sum(x)), big.mark=" ")
 ```
 
-<!-- Content for landing page-->
-
-##  Ferramenta de apoio ao planeamento da rede ciclável na Área Metropolitana de Lisboa
+<h2 style="text-align: center;">Ferramenta de apoio ao planeamento da rede ciclável<br> na Área Metropolitana de Lisboa</h2>
 
 ##### Nota: este projeto está em fase _beta_ e a informação deste site deve ser considerada como _work in progress_.
+
+<!-- Content for landing page-->
 
@@
 
+library(tidyverse)
 library(tmap)
@@
 
-Construir tabela com os dados de populacao, Bike, cyc4, ebike, etc
+Ver página com detalhes, mapas e mais resultados para a Área Metropolitana de Lisboa: [AML](aml/index.html){target="_blank"}
 
-```{r eval=FALSE, include=FALSE}
-# Aim: Build table for homepage
-# dn = "popupCycleways/v1"
-# built_counties = list.files(dn)
-# set.seed(42)
-municipios_index %>%
-  sf::st_drop_geometry() %>%
-  # dplyr::select(url, Population, "% cycling (Baseline)" = pcycle_baseline, "% cycling (Near Market)" = pcycle_near,  "% cycling (Climate Action Plan)" = pcycle_climate,  "% cycling (Go Dutch)" = pcycle_godutch,  "% cycling (Ebike)" = pcycle_ebike) %>%
-  # dplyr::rename(Name = url) %>%
-  dplyr::arrange(Name) %>%
-  # dplyr::mutate_if(is.numeric, round, 1) %>%
-  DT::datatable(escape = -2, filter = 'none', options = list(pageLength = 18,
+```{r amlprep}
+tabelas_aml = readRDS("HEAT/tabelas_aml.Rds")
+tabelas_aml$Estrategia = gsub("ate", "até", tabelas_aml$Estrategia)
+POPaml = tabelas_aml$Pop21[1]
+tabelas_aml = tabelas_aml %>%
+  mutate(Percursos = recode(LTS, `3` = "seguro", `4` = "direto")) %>% 
+  filter(Percursos == "direto") %>% 
+  select(c(1,5,6,7,3)) %>% 
+  pivot_wider(names_from = Meta_ENMAC,
+              values_from = Bike_new) %>% 
+  arrange(Estrategia)
+```
+
+A AML tem uma população de `r fmt_num(POPaml)`. Diariamente são realizadas 5.33 milhões de viagens, em vários modos de transporte, das quais apenas **25 479 são realizadas em bicicleta** (0.5%).
+
+A tabela informa sobre o potencial ciclável em cada cenário considerado neste estudo, para ligações entre freguesias, realizáveis em bicicleta, e para percursos preferencialmente diretos.
+
+```{r tabelaaml}
+
+DT::datatable(tabelas_aml,
+                filter = 'none',
+                rownames = FALSE,
+                colnames = c("Cenário", "Total Viagens /dia",
+                             "Viagens Bicicleta 2018", "Viagens Bicicleta 4%",
+                              "Viagens Bicicleta 10%"),
+                # extensions = "RowGroup",
+                extensions = "Responsive",
+                options = list(dom = "t",
+                               pageLength = 3,
                                responsive = TRUE,
-                                                             extensions = c('Responsive'))) 
+                               # rowGroup = list(dataSrc = 0), #agrupar a primeira coluna
+                               extensions = c('Responsive'))) %>% 
+    DT::formatRound(c(2:5),0, mark = " ")
+
 ```
+
+Para todos os detalhes sobre estes resultados, ver a página da [Área Metropolitana de Lisboa](aml/index.html){target="_blank"}
 
@@
 
-Construir tabela com os dados de populacao, Bike, cyc4, ebike, etc, por municipio
+A tabela informa sobre o potencial ciclável em cada cenário considerado neste estudo, para ligações entre freguesias, realizáveis em bicicleta, e para percursos preferencialmente diretos.
 
-```{r}
-# Aim: Build table for homepage
-# dn = "popupCycleways/v1"
-# built_counties = list.files(dn)
-# set.seed(42)
-municipios_index %>%
+Para um maior detalhe sobre os resultados de cada município, consultar a página individual, carregando no link do município, na tabela seguinte.
+
+```{r dados}
+TRIPSmode_municipio = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/TRIPSmode_municipal.Rds"))
+
+TRIPSall_mun = TRIPSmode_municipio %>%
+  group_by(Origem) %>%
+  summarise(viagens = round(sum(viagens)))
+```
+
+```{r tabelamun}
+
+municipios_taburl = municipios_index %>%
   sf::st_drop_geometry() %>%
-  dplyr::left_join(HEAT_municipios_PPP_redux %>% filter(CodePath == 1304)) %>% 
-  dplyr::select(url, Pop21, Total, Bike, Bike_new, CO2eq10, Economic10) %>%
-  dplyr::mutate(Economic10 = Economic10/1000,
-                CO2eq10 = as.numeric(gsub(" ","", CO2eq10))) %>% 
-  # Population, "% cycling (Baseline)" = pcycle_baseline, "% cycling (Near Market)" = pcycle_near,  "% cycling (Climate Action Plan)" = pcycle_climate,  "% cycling (Go Dutch)" = pcycle_godutch,  "% cycling (Ebike)" = pcycle_ebike) %>%
-  dplyr::rename(Municipio = url) %>%
-  dplyr::arrange(Municipio) %>%
-  # dplyr::mutate_if(is.numeric, round, 1) %>%
-  DT::datatable(escape = -1, #para colocar os links, a contar com os rownames
+  dplyr::left_join(HEAT_municipios_PPP_redux) %>% 
+  mutate(Percursos = recode(LTS, `3` = "seguro", `4` = "direto")) %>% 
+  filter(Percursos == "direto") %>% 
+  left_join(TRIPSall_mun, by =c("Municipio" = "Origem")) %>% 
+  dplyr::select(url, Pop21, viagens, Total, Bike, Bike_new, Estrategia, Meta_ENMAC) %>%
+  pivot_wider(names_from = c(Estrategia,Meta_ENMAC),
+              values_from = c(Total, Bike, Bike_new)) %>% 
+  select(-c(5,7,9,11,13,15)) %>% 
+  select(c(1:3,4,7,10,11,5,8,12,13,6,9,14,15)) %>% 
+  select(-c(4,8,12)) %>% 
+  dplyr::rename(Municipio = url,
+                # Total_5km = `Total_ate 5 km_4%`,
+                Bike_5km = `Bike_ate 5 km_4%`,
+                Bike_new4_5km = `Bike_new_ate 5 km_4%`,
+                Bike_new10_5km = `Bike_new_ate 5 km_10%`,
+                # Total_10km = `Total_ate 10km e-bike_4%`,
+                Bike_10km = `Bike_ate 10km e-bike_4%`,
+                Bike_new4_10km = `Bike_new_ate 10km e-bike_4%`,
+                Bike_new10_10km = `Bike_new_ate 10km e-bike_10%`,
+                # Total_5km_intermod = `Total_intermodal ate 5km _4%`,
+                Bike_5km_intermod = `Bike_intermodal ate 5km _4%`,
+                Bike_new4_intermod = `Bike_new_intermodal ate 5km _4%`,
+                Bike_new10_intermod = `Bike_new_intermodal ate 5km _10%`
+                ) %>%
+  dplyr::arrange(Municipio)
+
+DT::datatable(municipios_taburl,
+                escape = -1, #para colocar os links, a contar com os rownames
                 filter = 'none',
                 rownames = FALSE,
-                colnames = c("Município", "População 2021", "Viagens /dia", "Viagens Bicicleta", "Novas Viagens Bicicleta 4%",
-                             "CO2eq evitado em 10 anos (ton)", "Benefícios estimados em 10 anos (milhares €)"),
+                colnames = c("Município", "População 2021", "Total Viagens /dia 2018",
+                             "Viagens Bicicleta 2018 até 5km", "Viagens Bicicleta 4% até 5km", "Viagens Bicicleta 10% até 5km", "Viagens Bicicleta 2018 até 10km e-bike", "Viagens Bicicleta 4% até 10km e-bike", "Viagens Bicicleta 10% até 10km e-bike","Viagens Bicicleta 2018 até 5km intermodal", "Viagens Bicicleta 4% até 5km intermodal", "Viagens Bicicleta 10% até 5km intermodal"),
                 options = list(dom = "t",
                                pageLength = 18,
-                               responsive = TRUE,
+                               # width = "100%",
+                               scrollX = TRUE,
+                               # responsive = TRUE,
                               extensions = c('Responsive'))) %>% 
-  DT::formatRound(c(2:7),0, mark = " ")
+  DT::formatRound(c(2:12),0, mark = " ")
 ```

### 2022-12-06 code/biclarwww/municipio_index.Rmd
target blank

@@
 
-Segundo o Inquérito à Mobilidade  ([IMOB](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt)) realizado em 2018 pelo INE, as viagens com origem e/ou destino `r stringr::str_to_lower(proposicao)` `r mun` têm as seguintes características.
+Segundo o Inquérito à Mobilidade  ([IMOB](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2018 pelo INE, as viagens com origem e/ou destino `r stringr::str_to_lower(proposicao)` `r mun` têm as seguintes características.
 
@@
 A largura das linhas varia consoante o número de ciclistas potencial para cada um dos cenários (4% e 10% de ciclistas), e essa informação é visível ao passar com o rato em cima da linha.  
-A cor das linhas varia com o nível de tranquilidade ([_quietness_](https://www.cyclestreets.net/help/journey/howitworks/#quietness)), sendo as mais escuras aquelas menos seguras para circular em bicicleta - indicando com maior clareza quais os segmentos que necessitam de intervenção para se tornarem mais seguros e potencialmente trazerem mais ciclistas: os **mais escuros e mais largos**.
+A cor das linhas varia com o nível de tranquilidade ([_quietness_](https://www.cyclestreets.net/help/journey/howitworks/#quietness){target="_blank"}), sendo as mais escuras aquelas menos seguras para circular em bicicleta - indicando com maior clareza quais os segmentos que necessitam de intervenção para se tornarem mais seguros e potencialmente trazerem mais ciclistas: os **mais escuros e mais largos**.
 
@@
 
-É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacionail de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf)), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel.
+É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacionail de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel.
 </details>
@@
 
-Ver mapa em [ecrã inteiro](mapa_cenario1.html).
+Ver mapa em [ecrã inteiro](mapa_cenario1.html){target="_blank"}.
 
@@
 
-Ver mapa em [ecrã inteiro](mapa_cenario2.html).
+Ver mapa em [ecrã inteiro](mapa_cenario2.html){target="_blank"}.
 
@@
 
-Ver mapa em [ecrã inteiro](mapa_cenario3.html).
+Ver mapa em [ecrã inteiro](mapa_cenario3.html){target="_blank"}.
 
@@
   <summary>**Como são estimados os benefícios socio-ambientais?**</summary>
-Os benefícios socio-ambientais são estimados em duas vertentes, utilizando os métodos e a ferramenta [_HEAT for Cycling_](https://www.heatwalkingcycling.org/), da Organização Mundial de Saúde.
+Os benefícios socio-ambientais são estimados em duas vertentes, utilizando os métodos e a ferramenta [_HEAT for Cycling_](https://www.heatwalkingcycling.org/){target="_blank"}, da Organização Mundial de Saúde.
 
@@
 Os impactes sociais agregados, na componente da saúde, são estimados em termos de redução/aumento da mortalidade por aumento de atividade física, por exposição à poluição atomosférica, e por exposição ao risco de sinistralidade rodoviária.
-Este valor é por fim monetizado utilizando o _Valor Estatístico da Vida_ (€3.055.358, segundo [ANSR 2021](http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O%20Impacto%20Economico%20e%20Social%20da%20Sinistralidade%20-%20PT.pdf)), estimado para 10 anos, com uma taxa de desconto de 5% e inflação de 3%.
+Este valor é por fim monetizado utilizando o _Valor Estatístico da Vida_ (€3.055.358, segundo [ANSR 2021](http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O%20Impacto%20Economico%20e%20Social%20da%20Sinistralidade%20-%20PT.pdf){target="_blank"}), estimado para 10 anos, com uma taxa de desconto de 5% e inflação de 3%.
 </details>
@@
 tabelas_mun = readRDS("HEAT/tabelas_mun.Rds")
+tabelas_mun$Estrategia = gsub("ate", "até", tabelas_mun$Estrategia)
 tabelas_mun = tabelas_mun %>% filter(Municipio == mun) %>%
@@
 Os mapas acima estão disponíveis para download, em formato _GeoPackage_ (.gpkg), que pode ser ligo em qualquer SIG moderno.
-É recomendável o software de acesso livre [QGIS](https://qgis.org/pt_PT/site/) para leitura, visualização e análise de dados em detalhe.
+É recomendável o software de acesso livre [QGIS](https://qgis.org/pt_PT/site/){target="_blank"} para leitura, visualização e análise de dados em detalhe.
 

### 2022-12-06 code/biclarwww/index.Rmd
headers in table municipios

@@
 
+sketch = htmltools::withTags(table( #repetir header
+  class = 'display',
+  thead(
+    tr(
+      th(rowspan = 2, 'Município'),
+      th(rowspan = 2, 'População 2021'),
+      th(rowspan = 2, 'Total Viagens /dia 2018'),
+      th(colspan = 3, 'Viagens Bicicleta até 5km'),
+      th(colspan = 3, 'Viagens Bicicleta até 10km e-bike'),
+      th(colspan = 3, 'Viagens Bicicleta até 5km intermodal')
+    ),
+    tr(
+      lapply(rep(c("2018", "Meta 4%", "Meta 10%"), 3), th)
+    )
+  )
+))
+
 DT::datatable(municipios_taburl,
@@
                 rownames = FALSE,
-                colnames = c("Município", "População 2021", "Total Viagens /dia 2018",
-                             "Viagens Bicicleta 2018 até 5km", "Viagens Bicicleta 4% até 5km", "Viagens Bicicleta 10% até 5km", "Viagens Bicicleta 2018 até 10km e-bike", "Viagens Bicicleta 4% até 10km e-bike", "Viagens Bicicleta 10% até 10km e-bike","Viagens Bicicleta 2018 até 5km intermodal", "Viagens Bicicleta 4% até 5km intermodal", "Viagens Bicicleta 10% até 5km intermodal"),
+                container = sketch,
                 options = list(dom = "t",
@@
   DT::formatRound(c(2:12),0, mark = " ")
+
 ```

### 2022-12-06 code/biclarwww/index.Rmd
minor typos

@@
   pivot_wider(names_from = Meta_ENMAC,
-              values_from = Bike_new) %>% 
-  arrange(Estrategia)
+              values_from = Bike_new) 
 ```

### 2022-12-06 code/biclarwww/municipio_index.Rmd
minor typos

@@
 
-É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacionail de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel.
+É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel.
 </details>

### 2022-12-09 code/biclarwww/index.Rmd
table PT emissions

@@
 municipios_index$url[Municipio == "Vila Franca de Xira"] = paste0("<a href='vfxira/' target='_blank'>",  Municipio,  "</a>")
+municipios_index$url[Municipio == "Setubal"] = paste0("<a href='setubal/' target='_blank'>",  "Setúbal",  "</a>")
 
@@
 ## Mapa
-
-test content
-
-```{r}
-#read data
-transit_all = sf::st_read("https://github.com/U-Shift/biclar/releases/download/0.0.1/transit.gpkg", quiet = TRUE)
-
-library(tidyverse)
-library(tmap)
-tmap_mode("view")
-
-tm_shape(transit_all) +
-  tm_sf(col = "mode", size = 0.3)
-```
 
@@
 
+```{r mapahome}
+MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))
+TRIPSmode_municipio = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/TRIPSmode_municipal.Rds"))
+municipios_index$url[municipios_index$Municipio == "Vila Franca de Xira"] = paste0("<a href='vfxira/' target='_blank'>", "Vila Franca de Xira",  "</a>")
+municipios_index$url[municipios_index$Municipio == "Setubal"] = paste0("<a href='setubal/' target='_blank'>",  "Setúbal",  "</a>")
+
+TRIPSall_mun = TRIPSmode_municipio %>%
+  group_by(Origem) %>%
+  summarise(viagens = round(sum(viagens))) %>% 
+  ungroup()
+
+TRIPSbike_mun = TRIPSmode_municipio %>%
+  group_by(Origem, modo) %>%
+  summarise(viagens = round(sum(viagens))) %>% 
+  filter(modo == "Bike") %>% 
+  rename(Bike = viagens) %>% 
+  left_join(TRIPSall_mun) %>% 
+  mutate(percent = 100*Bike/viagens) %>% 
+  rename(Municipio = Origem)
+
+MUNICIPIOSgeo_mapa = municipios_index %>% left_join(TRIPSbike_mun) %>% select(-modo) %>% 
+  mutate(percent_label = paste0(round(percent,1), " %"))
+
+pal0 = colorNumeric("viridis", domain=MUNICIPIOSgeo_mapa$percent)
+
+m0 = leaflet::leaflet(MUNICIPIOSgeo_mapa) %>%
+  leaflet::addPolygons(
+    data = MUNICIPIOSgeo_mapa,
+    color = "white",
+    fillColor = ~pal0(percent),
+    group = "Municípios",
+    popup = paste0("Ver resultados: ", MUNICIPIOSgeo_mapa$url),
+    fillOpacity = 0.5,
+    opacity = 0.9,
+    weight = 1.5,
+    highlightOptions = highlightOptions(color = "black", bringToFront = TRUE, weight = 3),
+    label = ~ Municipio,
+    labelOptions = leaflet::labelOptions(noHide = F, permanent = TRUE, direction = "auto",sticky = FALSE, textsize = "1.5em")) %>%
+  leaflet::addLegend(
+    "bottomright",
+    pal = pal0,
+    values = ~percent,
+    opacity = 1.0,
+    title = "% viagens bicicleta 2018",
+    group = "Municípios"
+  ) %>%
+  leaflet::addLegend(
+    title = "Como começar:",
+    group = "Help",
+    position = "bottomleft",
+    colors = c("white",
+               "white"),
+    labels = c(
+      "Zoom para explorar a rede ciclável existente",
+      "Clicar num município para ver os resultados detalhados da rede potencial"
+    )
+  ) %>%
+    leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>%
+    leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
+    leaflet::addProviderTiles("OpenStreetMap.Mapnik", group = "OpenStreetMap") %>% 
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satelite"),
+                            overlayGroups = c("Municípios", "Rede Ciclável existente"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Rede Ciclável existente"))
+  # htmlwidgets::onRender("
+  #       function() {
+  #           $('.leaflet-control-layers-base').prepend('<label style=\"text-align:center;font-size:larger;\"><b>Base</b></label>');
+  #       }
+  #   ")
+
+m0
+```
+
 **O que mostra este mapa? **  
-Por enquanto nada.
+Este mapa mostra a percentagem de viagens com origem em cada município que são realizadas em biciceta.
+
+<!-- ------------------------------------------------------------------------ -->
+
+## Cenários
+
+Em que consistem os cenários utilizados nesta modelação?
+
+### Cenário Base
+
+Na AML, são realizadas cerca de 5 milhões de viagens todos os dias, em vários modos.
+O número de viagens em bicicleta, segundo o IMOB [2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
+
+Este é o cenário base, ou de referência, utilizado apenas para comparação com os cenários seguintes - de onde advém o **potencial ciclável**.
+
+### Cenário 1 - Viagens até 5 km
+
+Este cenário inclui todas as viagens que podem ser realizadas entre freguesias até 5 km em bicicleta.
+Inclui também viagens que utilizam o transporte fluvial (barcos) para unir ambas as margens, embora esse troço não seja considerado no total dos 5km.
+
+Este é um cenário conservador / realista.
+
+### Cenário 2 - Viagens até 10 km bicicleta elétrica
+
+Este cenário inclui todas as viagens que podem ser realizadas entre freguesias até 10 km em bicicleta, considerando para tal que os ciclistas poderão utilizar uma bicicleta elétrica.
+Inclui também viagens que utilizam o transporte fluvial (barcos) para unir ambas as margens, embora esse troço não seja considerado no total dos 10km.
+
+Este abrange uma maior gama de utilizadores, e tem em consideração possíveis políticas de incentivo à aquisição de bicicletas elétricas.
+
+### Cenário 3 - Viagens até 5 km de ou até uma interface de transportes públicos
+
+Este cenário inclui todas as viagens que podem ser realizadas entre freguesias em transportes públicos, cuja primeira e última _milha_ da viagem podem ser feitas em bicicleta, até uma extensão total de 5 km - por exemplo: 1 km de bicicleta até ao interface A, e 4 km de bicicleta entre o interface B e o destino.
+Como transportes públicos, são considerados apenas o comboio, barco, metro sul do Tejo, e algumas carreiras da Carris Metropolitana. 
+De uma forma conservadora, assumimos também que não são possíveis transferências entre transportes ou entre linhas/carreiras.
+
+Este valoriza a utilização da bicicleta como complemento ao transporte público.
 
@@
 ```{r amlprep}
-tabelas_aml = readRDS("HEAT/tabelas_aml.Rds")
+# tabelas_aml = readRDS("HEAT/tabelas_aml.Rds")
+tabelas_aml = readRDS("D:/GIS/TML/biclar/HEAT/tabelas_aml.Rds")
+
 tabelas_aml$Estrategia = gsub("ate", "até", tabelas_aml$Estrategia)
@@
               values_from = Bike_new)
+
 ```
@@
 
-This information can be used to inform decisions about where best to invest in new or improved cycle routes.
+O biclar é uma ferramenta online que informa os decisores sobre o planeamento e investimentos em infraestrutura ciclável, de modo a atingir as metas da Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}).
+
+Na ótica do decisor, pretende-se que esta informe os decisores, em comparação com o cenário atual de utilização de bicicleta, sobre:
+
+*   Quais as ligações prioritárias e percursos entre Freguesias e entre Municípios, a investir com infraestrutura ciclável, consoante o cenário adotado e a rede (primária ou secundária);
+*   Qual o incremento potencial de utilizadores de bicicleta de cada infraestrutura ciclável, bem como qual a substituição modal (redução de outros modos);
+*   Quais os impactes socioambientais previstos a médio e longo prazo (1 e 10 anos) em termos ambientais (CO~2~eq evitado), e sociais .
+
+É também possível saber a informação agregada por [município](Municípios), para cada cenário escolhido, sobre todos os valores acima descritos.
 
@@
 
-You can give feedback on the tool in the [online survey](nada).
+Para reportar algum problema com a ferramenta, ou colocar questões, pode abrir um _issue_ no nosso [repositório de dados abertos](https://github.com/U-Shift/biclarwww/issues).
 
@@
 
-Short answers to key frequently answered questions (FAQ) are provided below.
-See the full FAQ page for more details and answers to questions about how the tool works.
+Pergutas e respostas simplificadas.
+Ver o [relatório metodológico](nada) para mais detalhes sobre o funcionamento desta ferramenta..
 

### 2022-12-09 code/biclarwww/build.R
run all

@@ -18,6 +18,11 @@ rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", encoding = "UTF-8", ou
 
 
 i = "barreiro"
+i = "lisboa"
+i = "odivelas"
+i = "setubal"
+
+# for (i in municipios){
 
 for (i in municipios){
   
@@ -26,13 +31,10 @@ for (i in municipios){
       } else {
     pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio
   }
-  # dir.create(pub_folder) #comment this one??
+  # dir.create(pub_folder) #comment this one after run first time
   rmarkdown::render(input = "code/biclarwww/municipio_index.Rmd", encoding = "UTF-8", output_dir = pub_folder, output_file = "index.html")
-  # rmarkdown::render(input = "stats.Rmd", output_dir = pub_folder, output_file = "stats.html")
-  # rmarkdown::render(input = "baseline.Rmd", output_dir = pub_folder, output_file = "baseline.html")
-  # rmarkdown::render(input = "route-types.Rmd", output_dir = pub_folder, output_file = "route-types.html")
-  # rmarkdown::render(input = "downloads.Rmd", output_dir = pub_folder, output_file = "downloads.html")
-  
+
+  print(i)
 }
 
 

### 2022-12-09 code/test-code/heat-prep_car.R
data for table emissions

@@ -362,7 +362,7 @@ CAR_TP_TRANSF_lts4 = routes_r5r_100jit_car_TP_lts4_ONLYTPsegment %>% mutate(LTS 
 
 #Bind
 CAR_TP_TRANSF = rbind(CAR_TP_TRANSF_lts3, CAR_TP_TRANSF_lts4)
-
+saveRDS(CAR_TP_TRANSF, "HEAT/CAR_TP_TRANSF.Rds")
 
 
 

### 2022-12-09 code/biclarwww/municipio_index.Rmd
muncipios scenario3 specifics

@@
 
-dontrun = !i %in% c("alcochete", "mafra", "montijo") # não mostrar mapa 3
-run = i %in% c("alcochete", "mafra", "montijo") # mostrar mensagem mapa 3
+dontrun = !i %in% c("alcochete", "mafra", "montijo", "odivelas", "sesimbra") # não mostrar mapa 3
+run = i %in% c("alcochete", "mafra", "montijo", "odivelas", "sesimbra") # mostrar mensagem mapa 3
 
@@
 ```{r makemaps}
+
+if (i == "setubal"){
+  mun = "Setubal"
+}
+
 MUNICIPIOSgeo = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/MUNICIPIOSgeo.Rds"))
@@
   mapa12l = sf::st_read("export_rnet/rnet_aml_1310_15.gpkg", quiet = TRUE)
+  mapa12l = mapa12l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+
+} else if (i %in% c("lisboa")){
+
+  mapa11r = sf::st_read("export_rnet/rnet_aml_1404_100.gpkg", quiet = TRUE)
+  mapa11r = mapa11r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+  mapa11l = sf::st_read("export_rnet/rnet_aml_1304_100.gpkg", quiet = TRUE)
+  mapa11l = mapa11l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+  mapa12r = sf::st_read("export_rnet/rnet_aml_1410_100.gpkg", quiet = TRUE)
+  mapa12r = mapa12r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+         bike18 = paste0("(", Bikeper, ") ", Bike))
+  mapa12l = sf::st_read("export_rnet/rnet_aml_1310_100.gpkg", quiet = TRUE)
   mapa12l = mapa12l[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
@@
 if (i %in% c("mafra", "sesimbra", "alcochete", "palmela")){ 
-  minimo1 = 15} else {
+  minimo1 = 15} else if (i %in% c("lisboa")){ 
+    minimo1 = 100 
+  } else {
     minimo1 = 50 
@@
     minimo3 = 10
+  }
+
+if (i == "setubal"){
+  mun = "Setúbal"
 }
@@
 
+if (i == "lisboa"){
+  ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = viagens, x = Inter)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(values = col_modes, name = "Modo") +
+  scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
+  geom_text(aes(label = ifelse(viagens/1000 > 200, paste0(round(viagens/1000)," k"), "")),
+    position = position_stack(vjust = 0.5, reverse = FALSE))+
+  theme_minimal() +
+  theme(axis.title.y = element_blank(), legend.title = element_text("Modo"), legend.position = "right") +
+  coord_flip()
+} else {
   ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = viagens, x = Inter)) +
@@
   coord_flip()
+}
 
@@
 Das cerca de `r fmt_num(round(sum(TRIPSmode_mun$viagens)/1000))`mil viagens diárias `r stringr::str_to_lower(proposicao)` `r mun`, apenas **`r TRIPSmode_mun[3,3]`% das viagens são realizadas em bicicleta** (`r round(TRIPSmode_mun[3,2])`).
-`r round(TRIPSmode_mun[2,2]/1000)`mil são feitas em automóvel (`r TRIPSmode_mun[2,3]`%), `r round(TRIPSmode_mun[1,2]/1000)`mil a pé (`r TRIPSmode_mun[1,3]`%), `r round(TRIPSmode_mun[6,2]/1000)`mil em Transportes Públicos (`r TRIPSmode_mun[6,3]`%) e `r round(TRIPSmode_mun[4,2]+TRIPSmode_mun[5,2])` em outros modos (`r TRIPSmode_mun[4,3]+TRIPSmode_mun[5,3]`%).
+`r round(TRIPSmode_mun[2,2]/1000)`mil são feitas em automóvel (`r TRIPSmode_mun[2,3]`%), `r round(TRIPSmode_mun[1,2]/1000)`mil a pé (`r TRIPSmode_mun[1,3]`%), `r round(TRIPSmode_mun[6,2]/1000)`mil em Transportes Públicos (`r TRIPSmode_mun[6,3]`%) e `r fmt_num(round(TRIPSmode_mun[4,2]+TRIPSmode_mun[5,2]))` em outros modos (`r TRIPSmode_mun[4,3]+TRIPSmode_mun[5,3]`%).
 
@@
           control.position = c("right", "top")
-          # title = "Viagens até 5 km em bicicleta"
-          # leaflet.options = highlightOptions(color = "white", bringToFront = TRUE)
           )
+
 m1leaf = tmap::tmap_leaflet(m1) %>% 
@@
 ```{r mapa3, cache=FALSE, eval=dontrun}
+if (i == "lisboa"){
+  mapa31r = mapa31r %>% filter(cyc < 2400) #otherwise we cant see scale lwd
+  mapa31l = mapa31l %>% filter(cyc < 2400) #otherwise we cant see scale lwd
+  mapa32r = mapa32r %>% filter(cyc < 5400) #otherwise we cant see scale lwd
+  mapa32l = mapa32l %>% filter(cyc < 5400) #otherwise we cant see scale lwd
+}
+
 m3 = tmap::tm_shape(mapa31r, name = "Potencial 4% - direto") +
@@
                    alpha = 0.9,
-                   scale = 12,
-                   lwd_multiplier = 15,
+                   scale = 30,
+                   lwd_multiplier = 18,
                    popup.vars = c("Viagens 2018" = "Total",
@@
                    alpha = 0.9,
-                   scale = 12,
-                   lwd_multiplier = 15,
+                   scale = 40,
+                   lwd_multiplier = 18,
                    popup.vars = c("Viagens 2018" = "Total",
@@
                    alpha = 0.9,
-                   scale = 15,
+                   scale = 30,
                    lwd_multiplier = 18,
@@
                    alpha = 0.9,
-                   scale = 15,
+                   scale = 40,
                    lwd_multiplier = 18,
@@
 
+if (i == "setubal"){
+  mun = "Setubal"
+}
+
 tabelas_mun = readRDS("HEAT/tabelas_mun.Rds")
@@
 
+if (i == "setubal"){
+  mun = "Setúbal"
+}
 ```

### 2022-12-09 code/biclarwww/index.Rmd
tweaks municipios

@@
 ```{r amlprep}
-# tabelas_aml = readRDS("HEAT/tabelas_aml.Rds")
-tabelas_aml = readRDS("D:/GIS/TML/biclar/HEAT/tabelas_aml.Rds")
+tabelas_aml = readRDS("HEAT/tabelas_aml.Rds")
 
@@
 
-A tabela informa sobre o potencial ciclável em cada cenário considerado neste estudo, para ligações entre freguesias, realizáveis em bicicleta, e para percursos preferencialmente diretos.
+A tabela informa sobre o potencial ciclável em cada [cenário](Cenários) considerado neste estudo, para ligações entre freguesias, realizáveis em bicicleta, e para percursos preferencialmente diretos.
 
@@
                 # extensions = "RowGroup",
-                extensions = "Responsive",
-                options = list(dom = "t",
+                extensions = c("Responsive", "Buttons"),
+                options = list(dom = "tB",
+                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
                                pageLength = 3,
@@
                 container = sketch,
-                options = list(dom = "t",
+                extensions = "Buttons",
+                options = list(dom = "tB",
+                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
                                pageLength = 18,

### 2022-12-09 code/biclarwww/municipio_index.Rmd
tweaks municipios

@@
 
-dontrun = !i %in% c("alcochete", "mafra", "montijo", "odivelas", "sesimbra") # não mostrar mapa 3
-run = i %in% c("alcochete", "mafra", "montijo", "odivelas", "sesimbra") # mostrar mensagem mapa 3
+dontrun = !i %in% c("alcochete", "mafra", "montijo", "odivelas") # não mostrar mapa 3
+run = i %in% c("alcochete", "mafra", "montijo", "odivelas") # mostrar mensagem mapa 3
 
@@
 
-} else if (i %in% c("lisboa")){
+} else if (i %in% c("lisboa", "amadora", "oeiras")){
 
@@
 if (i %in% c("mafra", "sesimbra", "alcochete", "palmela")){ 
-  minimo1 = 15} else if (i %in% c("lisboa")){ 
+  minimo1 = 15} else if (i %in% c("lisboa", "amadora", "oeiras")){ 
     minimo1 = 100 
@@
   geom_text(aes(label = ifelse(viagens/1000 > 200, paste0(round(viagens/1000)," k"), "")),
+    position = position_stack(vjust = 0.5, reverse = FALSE))+
+  theme_minimal() +
+  theme(axis.title.y = element_blank(), legend.title = element_text("Modo"), legend.position = "right") +
+  coord_flip()
+} else if (i %in% c("amadora", "cascais", "loures", "oeiras")){
+  ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = viagens, x = Inter)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(values = col_modes, name = "Modo") +
+  scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
+  geom_text(aes(label = ifelse(viagens/1000 > 19, paste0(round(viagens/1000)," k"), "")),
+    position = position_stack(vjust = 0.5, reverse = FALSE))+
+  theme_minimal() +
+  theme(axis.title.y = element_blank(), legend.title = element_text("Modo"), legend.position = "right") +
+  coord_flip()
+} else if (i %in% c("sintra")){
+  ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = viagens, x = Inter)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(values = col_modes, name = "Modo") +
+  scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
+  geom_text(aes(label = ifelse(viagens/1000 > 25, paste0(round(viagens/1000)," k"), "")),
     position = position_stack(vjust = 0.5, reverse = FALSE))+
@@
 
+if (i == "sesimbra"){
+  scale1 = 20
+  scale2 = 30
+} else {
+  scale1 = 30
+  scale2 = 40
+}
+
 m3 = tmap::tm_shape(mapa31r, name = "Potencial 4% - direto") +
@@
                    alpha = 0.9,
-                   scale = 30,
+                   scale = scale1,
                    lwd_multiplier = 18,
@@
                    alpha = 0.9,
-                   scale = 40,
+                   scale = scale1,
                    lwd_multiplier = 18,
@@
                    alpha = 0.9,
-                   scale = 30,
+                   scale = scale2,
                    lwd_multiplier = 18,
@@
                    alpha = 0.9,
-                   scale = 40,
+                   scale = scale2,
                    lwd_multiplier = 18,
@@
                              "CO2eq evitado (ton/ano)", "Benefícios estimados em 10 anos (milhares €)"),
-                extensions = "RowGroup",
-                options = list(dom = "t",
+                extensions = c("RowGroup", "Buttons"),
+                options = list(dom = "tB",
+                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
                                pageLength = 12,

### 2022-12-09 code/biclarwww/build.R

@@ -16,7 +16,7 @@ rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", encoding = "UTF-8", ou
 # Create outputs for each region --------------------------------------------------------------
 
 
-
+i = "oeiras"
 i = "barreiro"
 i = "lisboa"
 i = "odivelas"
@@ -26,14 +26,21 @@ i = "setubal"
 
 for (i in municipios){
   
+  # dir.create(pub_folder) #comment this one after run first time
+  
   if (i == "vila franca de xira"){
     pub_folder = paste0("biclarwww/", "vfxira")
       } else {
     pub_folder = paste0("biclarwww/", i) #this needs to be updated for each municipio
-  }
-  # dir.create(pub_folder) #comment this one after run first time
+      }
+  
+  if (i %in% c("alcochete", "mafra", "montijo", "odivelas")){
+    rmarkdown::render(input = "code/biclarwww/municipio_minor_index.Rmd", encoding = "UTF-8", output_dir = pub_folder, output_file = "index.html")
+    
+  } else {
   rmarkdown::render(input = "code/biclarwww/municipio_index.Rmd", encoding = "UTF-8", output_dir = pub_folder, output_file = "index.html")
-
+  }
+  
   print(i)
 }
 

### 2022-12-14 code/biclarwww/municipio_index.Rmd
rename column "Novas" to "potencial"

@@
                 colnames = c("Meta ENMAC", "Cenário", "Tipo de Percurso", "Total Viagens /dia",
-                             "Viagens Bicicleta 2018", "Novas Viagens Bicicleta",
+                             "Viagens Bicicleta 2018", "Viagens Bicicleta Potencial",
                              "CO2eq evitado (ton/ano)", "Benefícios estimados em 10 anos (milhares €)"),

### 2022-12-14 code/biclarwww/index.Rmd
replace socioambiental by social and ambiental

@@
 *   Qual o incremento potencial de utilizadores de bicicleta de cada infraestrutura ciclável, bem como qual a substituição modal (redução de outros modos);
-*   Quais os impactes socioambientais previstos a médio e longo prazo (1 e 10 anos) em termos ambientais (CO~2~eq evitado), e sociais .
+*   Quais os impactes previstos a médio e longo prazo (1 e 10 anos) em termos ambientais (CO~2~eq evitado), e sociais.
 

### 2022-12-14 code/biclarwww/municipio_index.Rmd
replace socioambiental by social and ambiental

@@
 
-Ao clicar em cada linha, uma janela abre-se com mais informação sobre as viagens estimadas que podem passar por aquele segmento, incluindo os benefícios socio-ambientais estimados, e as características físicas daquele segmento.
+Ao clicar em cada linha, uma janela abre-se com mais informação sobre as viagens estimadas que podem passar por aquele segmento, incluindo os benefícios sociais e ambientais estimados, e as características físicas daquele segmento.
 
@@
 
-Para cada um destes cenários, é possível ver também de forma agregada qual o potencial socio-ambiental estimado, em toneladas de CO~2~eq evitado anualmente, e em benefícios sociais.
+Para cada um destes cenários, é possível ver também de forma agregada qual o benefício potencial estimado, em toneladas de CO~2~eq evitado anualmente, e em benefícios sociais.
 
 <details>
-  <summary>**Como são estimados os benefícios socio-ambientais?**</summary>
-Os benefícios socio-ambientais são estimados em duas vertentes, utilizando os métodos e a ferramenta [_HEAT for Cycling_](https://www.heatwalkingcycling.org/){target="_blank"}, da Organização Mundial de Saúde.
+  <summary>**Como são estimados os benefícios sociais e ambientais?**</summary>
+Os benefícios potenciais são estimados em duas vertentes, utilizando os métodos e a ferramenta [_HEAT for Cycling_](https://www.heatwalkingcycling.org/){target="_blank"}, da Organização Mundial de Saúde.
 

### 2022-12-15 code/biclarwww/aml_index.Rmd
add missing files
necessary to build the biclarwww website

---
title: "Resultados para a AML"
output:
  html_document:
    theme:
      bg: "#FFFFFF"
      fg: "#000000"
      highlight: "#000000"
      primary: "#000000"
      base_font:
        google: "Lexend"
      code_font:
        google: "JetBrains Mono"
    # toc: true
    # toc_float:
    #   toc_collapsed: true
    # toc_depth: 3
    css: css/style.css
    includes:
      after_body: img/footer.html
---

<!-- landing page for aml -->

```{r setup, include=FALSE, cache=FALSE}
library(dplyr)
library(tmap)
tmap_mode("view")
library(leaflet)
library(ggplot2)
library(plotly)
library(scales)

knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file()) #to use the project root as the working directory

knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      cache = TRUE, #hide if needed
                      warning = FALSE
                      )

fmt_num = function(x) format(round(sum(x)), big.mark=" ")
```

```{r chamadas}
i = "aml"
mun = "Área Metropolitana de Lisboa"

proposicao = "Na"
proposicaoD = "Da"
```

```{r dados}
TRIPSmode_municipio = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/TRIPSmode_municipal.Rds"))

TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Car"] = "Automóvel"
TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Bike"] = "Bicicleta"
TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Motorcycle"] = "Motociclo"
TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Walk"] = "A Pé"
TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Transit"] = "Transportes Públicos"
TRIPSmode_municipio$modo[TRIPSmode_municipio$modo == "Other"] = "Outro"

TRIPSmode_aml = TRIPSmode_municipio %>%
  group_by(modo) %>%
  summarise(viagens = round(sum(viagens))) %>% 
  mutate(percentagem = round(100*viagens/sum(viagens),1))

TRIPSmode_aml_intr = TRIPSmode_municipio
TRIPSmode_aml_intr$Inter[TRIPSmode_aml_intr$Origem == TRIPSmode_aml_intr$Destino] = "Intramunicipal"
TRIPSmode_aml_intr$Inter[is.na(TRIPSmode_aml_intr$Inter)] = "Intermunicipal"
TRIPSmode_aml_intr = TRIPSmode_aml_intr %>% 
  group_by(modo, Inter) %>%
  summarise(viagens = round(sum(viagens)))

viagensinter = sum(TRIPSmode_aml_intr$viagens[TRIPSmode_aml_intr$Inter == "Intermunicipal"])
viagensintra = sum(TRIPSmode_aml_intr$viagens[TRIPSmode_aml_intr$Inter == "Intramunicipal"])

TRIPSmode_aml_intr$percentagem[TRIPSmode_aml_intr$Inter == "Intermunicipal"] =  round(100*TRIPSmode_aml_intr$viagens[TRIPSmode_aml_intr$Inter == "Intermunicipal"]/viagensinter, 1)

TRIPSmode_aml_intr$percentagem[TRIPSmode_aml_intr$Inter == "Intramunicipal"] =  round(100*TRIPSmode_aml_intr$viagens[TRIPSmode_aml_intr$Inter == "Intramunicipal"]/viagensintra, 1)

TRIPSmode_aml_intr = bind_rows(TRIPSmode_aml %>% mutate(Inter = "Todas"),  TRIPSmode_aml_intr)

```

```{r makemaps}

HEAT_aml_redux = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/HEAT_aml_123_redux.Rds"))

# cenario 1

  mapa11r = sf::st_read("export_rnet/rnet_aml_1404_100.gpkg", quiet = TRUE)
  mapa11r = mapa11r %>% 
    mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
  mapa11l = sf::st_read("export_rnet/rnet_aml_1304_100.gpkg", quiet = TRUE)
  mapa11l = mapa11l %>% 
    mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
  mapa12r = sf::st_read("export_rnet/rnet_aml_1410_100.gpkg", quiet = TRUE)
  mapa12r = mapa12r %>% 
    mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
  mapa12l = sf::st_read("export_rnet/rnet_aml_1310_100.gpkg", quiet = TRUE)
  mapa12l = mapa12l %>% 
    mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))

# cenario 2

mapa21r = sf::st_read("export_rnet/rnet_aml_2404_200.gpkg", quiet = TRUE)
mapa21r = mapa21r %>% 
  mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
mapa21l = sf::st_read("export_rnet/rnet_aml_2304_200.gpkg", quiet = TRUE)
mapa21l = mapa21l%>% 
  mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
mapa22r = sf::st_read("export_rnet/rnet_aml_2410_200.gpkg", quiet = TRUE)
mapa22r = mapa22r %>% 
  mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
mapa22l = sf::st_read("export_rnet/rnet_aml_2310_200.gpkg", quiet = TRUE)
mapa22l = mapa22l %>% 
  mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))

# cenario 3

mapa31r = sf::st_read("export_rnet/rnet_aml_3404_50.gpkg", quiet = TRUE)
mapa31r = mapa31r %>% 
  mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
mapa31l = sf::st_read("export_rnet/rnet_aml_3304_50.gpkg", quiet = TRUE)
mapa31l = mapa31l %>% 
  mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
mapa32r = sf::st_read("export_rnet/rnet_aml_3410_30.gpkg", quiet = TRUE)
mapa32r = mapa32r %>% 
  mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))
mapa32l = sf::st_read("export_rnet/rnet_aml_3310_30.gpkg", quiet = TRUE)
mapa32l = mapa32l %>% 
  mutate(tCO2eq = tCO2eq10/10,
         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
         bike18 = paste0("(", Bikeper, ") ", Bike))

```

```{r minimos}

minimo1 = 100

minimo2 = 200

minimo3 = 50

```

# `r mun`

Segundo o Inquérito à Mobilidade  ([IMOB](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2018 pelo INE, as viagens dentro da `r mun` têm as seguintes características.

```{r mododist, fig.ncol=2, out.width="50%"}
#gráfico com bistribuição de viagens por modo na AML segundo IMOB 2018

TRIPSmode_aml_intr$modo = factor(TRIPSmode_aml_intr$modo, 
                                 levels = c("Automóvel", "Transportes Públicos", "Motociclo", 
                                            "Outro", "A Pé", "Bicicleta"))

col_modes = c("#fe5f55", "#ffd166", "#8E794A", "grey", "#457b9d", "#90be6d")

ggplot(TRIPSmode_aml_intr, aes(fill = modo, y = percentagem/100, x = Inter)) +
  geom_bar(position = position_fill(reverse = FALSE), stat = "identity") +
  scale_fill_manual(values = col_modes, name = "Modo") +
  scale_y_continuous(labels = scales::percent) +
  geom_text(aes(label = ifelse(percentagem > 6, paste0(percentagem,"%"), "")),
    position = position_stack(vjust = 0.5, reverse = FALSE))+
  theme_minimal() +
  theme(axis.title.y = element_blank(), axis.title.x = element_blank(), legend.title = element_blank(), legend.position = "none") +
  coord_flip()

ggplot(TRIPSmode_aml_intr, aes(fill = modo, y = viagens, x = Inter)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = col_modes, name = "Modo") +
  scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
  geom_text(aes(label = ifelse(viagens > 500000, paste0(round(viagens/1000000,1)," M"), "")),
    position = position_stack(vjust = 0.5, reverse = FALSE))+
  theme_minimal() +
  theme(axis.title.y = element_blank(), legend.title = element_text("Modo"), legend.position = "right") +
  coord_flip()

```

Das cerca de `r round(sum(TRIPSmode_aml$viagens)/1000000,1)` milhões de viagens diárias `r stringr::str_to_lower(proposicao)` `r mun`, apenas `r TRIPSmode_aml[3,3]`% das viagens são realizadas em bicicleta (`r fmt_num(round(TRIPSmode_aml[3,2]))`). `r round(TRIPSmode_aml[2,2]/1000000,1)` milhões são feitas em automóvel (`r TRIPSmode_aml[2,3]`%), `r round(TRIPSmode_aml[1,2]/1000000,1)` milhões a pé (`r TRIPSmode_aml[1,3]`%), `r fmt_num(round(TRIPSmode_aml[6,2]/1000))`mil em Transportes Públicos (`r TRIPSmode_aml[6,3]`%) e `r round((TRIPSmode_aml[4,2]+TRIPSmode_aml[5,2])/1000)`mil em outros modos (`r TRIPSmode_aml[4,3]+TRIPSmode_aml[5,3]`%).

A figura seguinte mostra a proporção de viagens que é realizada com origem e destino no mesmo município (`r round(viagensintra/1000000,1)` milhões viagens), e as viagens com origem e destino em municípios diferentes (`r round(viagensinter/1000000,1)` milhões viagens). Como se pode observar, para todos os municípios, a maioria das viagens realizadas é intra municipal, emboras as viagens intermunicipais são bastante expressivas, principalmente em municípios como Odivelas, Oeiras, Amadora e Loures.

```{r circus, out.width="50%", fig.align='center'}
circus = "https://github.com/U-Shift/biclar/raw/master/man/figures/aml_circos.png"
knitr::include_graphics(circus, error = FALSE)
```

As metas da Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), estabelecem que dentro dos centros urbanos **4%** das viagens devem ser realizadas em bicicleta até 2024, e **10%** até 2030, e que as mesmas devem ser substituídas de viagens atualmente realizadas em automóvel.
Partindo do número de viagens realizadas em bicicleta em 2018 (cenário de referência, ou _cenário base_), é possível estimar o número de viagens com potencial para serem substituídas do automóvel para a bicicleta para cada cenário aqui avaliado.

O gráfico seguinte mostra o potencial de viagens a serem realizadas em bicicleta diariamente, por gama de distâncias de viagem, e para os cenários avaliados.

```{r plottly}
# gráfico com caracterização viagens IMOB intra municipais e intermonicipais

routes_13r_plot = readRDS("routes_13r_plot.Rds")

p2 = ggplot(routes_13r_plot) + 
  geom_smooth(span = 0.2, aes(cyc_dist_km,
                              trips_cyc_enmac10_1,
                              color="ENMAC 10%", 
                              group =1,
                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac10_1))) + 
  # geom_smooth(span = 0.2, aes(cyc_dist_km,
  #                             cumsum(trips_cyc_enmac10_1),
  #                             color="ENMAC 10%", 
  #                             group =1,
  #                             text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac10_1)), 
  #             linetype = "dashed") +
  geom_smooth(span = 0.2, aes(cyc_dist_km,
                              trips_cyc_enmac4_1,
                              color="ENMAC 4%",
                              group =2,
                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac4_1))) +
  # geom_smooth(span = 0.2, aes(cyc_dist_km,
  #                             cumsum(trips_cyc_enmac4_1),
  #                             color="ENMAC 4%",
  #                             group =2,
  #                             text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac4_1)), 
  #             linetype = "dashed") +
  geom_smooth(span = 0.3, aes(cyc_dist_km,
                              trips_cyc_enmac10_3,
                              color="Intermodal ENMAC 10%", 
                              group =1,
                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac10_3))) +
  # geom_smooth(span = 0.3, aes(cyc_dist_km,
  #                             cumsum(trips_cyc_enmac10_3),
  #                             color="Intermodal ENMAC 10%", 
  #                             group =1,
  #                             text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac10_3)), 
  #             linetype = "dashed") +
  geom_smooth(span = 0.3, aes(cyc_dist_km,
                              trips_cyc_enmac4_3,
                              color="Intermodal ENMAC 4%",
                              group =2,
                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac4_3))) +
  # geom_smooth(span = 0.3, aes(cyc_dist_km,
  #                             cumsum(trips_cyc_enmac4_3),
  #                             color="Intermodal ENMAC 4%",
  #                             group =2,
  #                             text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac4_3)), 
  #             linetype = "dashed")  +
  geom_smooth(span = 0.2, aes(cyc_dist_km,
                              trips_cyc_baseline_1,
                              color="IMOB 2018", 
                              group =3,
                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_baseline_1))) +
  geom_smooth(span = 0.2, aes(cyc_dist_km,
                              cumsum(trips_cyc_baseline_1),
                              color="IMOB 2018", 
                              group =3,
                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_baseline_1)), 
              linetype = "dashed") +
  # geom_smooth(span = 0.2, aes(cyc_dist_km,
  #                             trips_all_baseline_1,
  #                             color="Todos os modos", 
  #                             group =3,
  #                             text = paste("Distancia:", cyc_dist_km, "<br>Viagens total:", trips_all_baseline_1))) +
  scale_y_continuous("Viagens em bicictela /dia", labels = label_number(suffix = " k", scale = 1e-3)) +
  scale_x_continuous("Distancia de viagem (km)", breaks = seq(0,30, by=5)) +
  labs(title="", color='Cenario') +
  theme_minimal() +  
  guides(color=guide_legend(override.aes=list(fill=NA)), lty=guide_legend(override.aes=list(fill=NA))) +# removes the grey background in legend categories 
  scale_color_manual(values = c("blue", "darkgreen", "black", "red", "orange"))
  # scale_linetype_manual(name = "valor", values = c("solid", "dashed"), labels = c("as it is", "comulativo")) #tipo de linha
ggplotly(p2) %>% config(displaylogo = FALSE) %>% 
  layout(hovermode = "x") %>% style(hoverinfo = "y")

```

A linha a tracejado representa os valor comulativo de número de viagens em bicicleta por dia do IMOB, para cada distância. Como se pode observar, o maior potencial de viagens em bicicleta situa-se em viagens até 10 km, sendo que os primeiros 5 km são muito expressivos.

## Rede ciclável potencial

Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMOB 2018), para **4%** de viagens em bicicleta (meta 2024), e para **10%** de viagens em bicicleta (meta 2030), transferidas diretamente de viagens em automóvel.

Cada mapa refere-se a cada uma das 3 estimativas:

* viagens até 5 km com ligação em bicicleta (_`r fmt_num(HEAT_aml_redux[4,13])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
* viagens até 10 km com ligação em bicicleta elétrica (_`r fmt_num(HEAT_aml_redux[8,13])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
* viagens até 5 km em bicicleta de ou até uma interface de transportes (_`r fmt_num(HEAT_aml_redux[12,13])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)

<!-- Passar isto para FAQ? -->

<details>
  <summary>**Como ler estes mapas?**</summary>
Os mapas apresentam a informação gráfica para duas redes - a rede ciclável para viagens mais diretas, e a rede para viagens mais seguras e tranquilas (conforme a rede viária existente).  
A largura das linhas varia consoante o número de ciclistas potencial para cada um dos cenários (4% e 10% de ciclistas), e essa informação é visível ao passar com o rato em cima da linha.  
A cor das linhas varia com o nível de tranquilidade ([_quietness_](https://www.cyclestreets.net/help/journey/howitworks/#quietness){target="_blank"}), sendo as mais escuras aquelas menos seguras para circular em bicicleta - indicando com maior clareza quais os segmentos que necessitam de intervenção para se tornarem mais seguros e potencialmente trazerem mais ciclistas: os **mais escuros e mais largos**.

Ao clicar em cada linha, uma janela abre-se com mais informação sobre as viagens estimadas que podem passar por aquele segmento, incluindo os benefícios sociais e ambientais estimados, e as características físicas daquele segmento.

É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel. 
</details>

#### Até 5 km

Este mapa apresenta os segmentos com mais de `r minimo1` viagens potenciais em bicicleta. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].

```{r mapa1, cache=FALSE, eval=FALSE}
m1 = tmap::tm_shape(mapa11r, name = "Potencial 4% - direto") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 12,
                   lwd_multiplier = 15,
                   popup.vars = c("Viagens 2018" = "Total",
                                  # "Em bicicleta 2018" = "Bike",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 4%" = "cyc",
                                  # "Novos ciclistas 4%" = "new_cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0), #
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-linear_yl_rd_bk", n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   textNA = "sem info",
                   title.col = "Tranquilidade"
                   ) +
  tmap::tm_shape(mapa11l, name = "Potencial 4% - seguro") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 12,
                   lwd_multiplier = 15,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 4%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0), #
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-mako", n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   textNA = "sem info",
                   title.col = "Tranquilidade"
                   ) +
  tmap::tm_shape(mapa12r, name = "Potencial 10% - direto") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 15,
                   lwd_multiplier = 18,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 10%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0),
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-linear_yl_rd_bk",  n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   legend.col.show = FALSE,
                   textNA = "sem info",
                   title.col = "Tranquilidade",
                   ) +
  tmap::tm_shape(mapa12l, name = "Potencial 10% - seguro") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 15,
                   lwd_multiplier = 18,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 10%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0),
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-mako",  n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   legend.col.show = FALSE,
                   textNA = "sem info",
                   title.col = "Tranquilidade",
                   ) +
  tm_view(view.legend.position = c("right", "bottom"),
          control.position = c("right", "top")
          # title = "Viagens até 5 km em bicicleta"
          # leaflet.options = highlightOptions(color = "white", bringToFront = TRUE)
          )
m1leaf = tmap::tmap_leaflet(m1) %>% 
  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
                                              "Potencial 4% - seguro", "Potencial 10% - seguro"),
                            options = layersControlOptions(collapsed = FALSE)) %>% 
  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro"))

m1leaf

```

> Ver o mapa em [ecrã inteiro](mapa_cenario1.html){target="_blank"}.

#### Até 10 km (e-bike)

Este mapa mostra os segmentos com mais de `r minimo2` viagens potenciais em bicicleta elétrica. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].

```{r mapa2, cache=FALSE, eval=FALSE}
m2 = tmap::tm_shape(mapa21r, name = "Potencial 4% - direto") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 12,
                   lwd_multiplier = 15,
                   popup.vars = c("Viagens 2018" = "Total",
                                  # "Em bicicleta 2018" = "Bike",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 4%" = "cyc",
                                  # "Novos ciclistas 4%" = "new_cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0), #
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-reds3", n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   textNA = "sem info",
                   title.col = "Tranquilidade"
                   ) +
  tmap::tm_shape(mapa21l, name = "Potencial 4% - seguro") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 12,
                   lwd_multiplier = 15,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 4%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0), #
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-hokusai2", n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   textNA = "sem info",
                   title.col = "Tranquilidade"
                   ) +
  tmap::tm_shape(mapa22r, name = "Potencial 10% - direto") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 15,
                   lwd_multiplier = 18,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 10%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0),
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-reds3",  n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   legend.col.show = FALSE,
                   textNA = "sem info",
                   title.col = "Tranquilidade",
                   ) +
  tmap::tm_shape(mapa22l, name = "Potencial 10% - seguro") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 15,
                   lwd_multiplier = 18,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 10%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0),
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-hokusai2",  n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   legend.col.show = FALSE,
                   textNA = "sem info",
                   title.col = "Tranquilidade",
                   ) +
  tm_view(view.legend.position = c("right", "bottom"),
          control.position = c("right", "top")
          # title = "Viagens até 5 km em bicicleta"
          # leaflet.options = highlightOptions(color = "white", bringToFront = TRUE)
          )
m2leaf = tmap::tmap_leaflet(m2) %>% 
  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
                                              "Potencial 4% - seguro", "Potencial 10% - seguro"),
                            options = layersControlOptions(collapsed = FALSE)) %>% 
  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro"))

m2leaf

```

> Ver o mapa em [ecrã inteiro](mapa_cenario2.html){target="_blank"}.

#### Até 5 km intermodal

Este mapa mostra os segmentos com mais de `r minimo3` viagens potenciais em bicicleta, em que parte da viagem é realizada em transportes públicos (barco, comboio, metro sul do Tejo, ou carris metropolitana). Os segmentos deste mapa dizem respeito apenas à parte da viagem que é realizada em bicicleta, num total de até 5 km no conjunto da primeira e última etapa da viagem intermodal. Para um maior ou menor detalhe da rede, ver a secção de [Downloads].

```{r mapa3, cache=FALSE}
mapa31r = mapa31r %>% filter(cyc < 2400) #otherwise we cant see scale lwd
mapa31l = mapa31l %>% filter(cyc < 2400) #otherwise we cant see scale lwd
mapa32r = mapa32r %>% filter(cyc < 5400) #otherwise we cant see scale lwd
mapa32l = mapa32l %>% filter(cyc < 5400) #otherwise we cant see scale lwd

m3 = tmap::tm_shape(mapa31r, name = "Potencial 4% - direto") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 30,
                   lwd_multiplier = 18,
                   popup.vars = c("Viagens 2018" = "Total",
                                  # "Em bicicleta 2018" = "Bike",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 4%" = "cyc",
                                  # "Novos ciclistas 4%" = "new_cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0), #
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-burg", n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   textNA = "sem info",
                   title.col = "Tranquilidade"
                   ) +
  tmap::tm_shape(mapa31l, name = "Potencial 4% - seguro") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 40,
                   lwd_multiplier = 18,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 4%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0), #
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-dark_mint", n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   textNA = "sem info",
                   title.col = "Tranquilidade"
                   ) +
  tmap::tm_shape(mapa32r, name = "Potencial 10% - direto") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 30,
                   lwd_multiplier = 18,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 10%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0),
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-burg",  n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   legend.col.show = FALSE,
                   textNA = "sem info",
                   title.col = "Tranquilidade",
                   ) +
  tmap::tm_shape(mapa32l, name = "Potencial 10% - seguro") +
    tmap::tm_lines(id = "cyc",
                   lwd = "cyc",
                   alpha = 0.9,
                   scale = 40,
                   lwd_multiplier = 18,
                   popup.vars = c("Viagens 2018" = "Total",
                                  "Ciclistas cenário base" = "bike18",
                                  "Ciclistas cenário 10%" = "cyc",
                                  "Benefícios a 10 anos [milhares €]" = "Beneficios10_keuro",
                                  "CO2eq evitado [ton/ano]" = "tCO2eq",
                                  "Tranquilidade" = "quietness",
                                  "Velocidade auto [km/h]" = "carspeed",
                                  "Declive médio" = "slope"
                                  ),
                   popup.format = list(big.mark = " ", digits = 0),
                   col = "quietness",
                   palette = cols4all::c4a(palette = "-dark_mint",  n = 4),
                   breaks = c(0, 25, 50, 75, 100),
                   legend.format = list(text.separator = "a"),
                   legend.col.show = FALSE,
                   textNA = "sem info",
                   title.col = "Tranquilidade",
                   ) +
  tm_view(view.legend.position = c("right", "bottom"),
          control.position = c("right", "top")
          # title = "Viagens até 5 km em bicicleta"
          # leaflet.options = highlightOptions(color = "white", bringToFront = TRUE)
          )
m3leaf = tmap::tmap_leaflet(m3) %>% 
  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
                                              "Potencial 4% - seguro", "Potencial 10% - seguro"),
                            options = layersControlOptions(collapsed = FALSE)) %>% 
  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro"))

m3leaf

```

Ver mapa em [ecrã inteiro](mapa_cenario3.html){target="_blank"}.

```{r exportmaps}
htmlwidgets::saveWidget(m1leaf, file = paste0("biclarwww/", i,"/mapa_cenario1.html"))

htmlwidgets::saveWidget(m2leaf, file = paste0("biclarwww/", i,"/mapa_cenario2.html"))

htmlwidgets::saveWidget(m3leaf, file = paste0("biclarwww/", i,"/mapa_cenario3.html"))
```

## Resumo

A tabela mostra o potencial ciclável `r stringr::str_to_lower(proposicao)` `r mun` em número de novas viagens em bicicleta potenciais, para cada meta da Estratégia Nacional ENMAC, para cada cenário, e para cada tipo de rotas otimizadas.

Para cada um destes cenários, é possível ver também de forma agregada qual o benefício potencial estimado, em toneladas de CO~2~eq evitado anualmente, e em benefícios sociais.

<details>
  <summary>**Como são estimados os benefícios sociais e ambientais?**</summary>
Os benefícios potenciais são estimados em duas vertentes, utilizando os métodos e a ferramenta [_HEAT for Cycling_](https://www.heatwalkingcycling.org/){target="_blank"}, da Organização Mundial de Saúde.

A componente ambiental é medida em toneladas de CO~2~eq evitado, por transferência de viagens equivalentes em automóvel, com a mesma origem e destino, mas não necessariamente com o mesmo percurso.

Os impactes sociais agregados, na componente da saúde, são estimados em termos de redução/aumento da mortalidade por aumento de atividade física, por exposição à poluição atomosférica, e por exposição ao risco de sinistralidade rodoviária.  
Este valor é por fim monetizado utilizando o _Valor Estatístico da Vida_ (€3.055.358, segundo [ANSR 2021](http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O%20Impacto%20Economico%20e%20Social%20da%20Sinistralidade%20-%20PT.pdf){target="_blank"}), estimado para 10 anos, com uma taxa de desconto de 5% e inflação de 3%. 
</details>

```{r tabela}
# tabela com potencial para cada cenario +  estimativas SE

tabelas_aml = readRDS("HEAT/tabelas_aml.Rds")
tabelas_aml$Estrategia = gsub("ate", "até", tabelas_aml$Estrategia)
tabelas_aml = tabelas_aml %>%
  mutate(Percursos = recode(LTS, `3` = "seguro", `4` = "direto")) %>% 
  select(c(3,1,10,5:9)) %>% 
  arrange(Meta_ENMAC)

DT::datatable(tabelas_aml,
                filter = 'none',
                rownames = FALSE,
                colnames = c("Meta ENMAC", "Cenário", "Tipo de Percurso", "Total Viagens /dia",
                             "Viagens Bicicleta 2018", "Viagens Bicicleta Potencial",
                             "CO2eq evitado (ton/ano)", "Benefícios estimados em 10 anos (milhares €)"),
                extensions = c("RowGroup", "Buttons"),
                options = list(dom = "tB",
                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
                               pageLength = 12,
                               responsive = TRUE,
                               rowGroup = list(dataSrc = 0), #agrupar a primeira coluna
                               extensions = c('Responsive'))) %>% 
    DT::formatRound(c(4:8),0, mark = " ")

cicmacx4 = max(tabelas_aml$Bike_new[tabelas_aml$Meta_ENMAC=="4%"])
cicmacx10 = max(tabelas_aml$Bike_new[tabelas_aml$Meta_ENMAC=="10%"])
co2macx4 = max(tabelas_aml$CO2eq[tabelas_aml$Meta_ENMAC=="4%"])
co2macx10 = max(tabelas_aml$CO2eq[tabelas_aml$Meta_ENMAC=="10%"])
euromacx4 = max(tabelas_aml$Economic10[tabelas_aml$Meta_ENMAC=="4%"])
euromacx10 = max(tabelas_aml$Economic10[tabelas_aml$Meta_ENMAC=="10%"])

```

`r proposicao` `r mun`, a **meta de 4%** de ciclistas diários equivale a ter **`r fmt_num(cicmacx4/1000)` mil ciclistas** em viagens até 10 km, com uma redução potencial de **`r fmt_num(co2macx4/1000)` mil toneladas** de CO~2~eq por ter menos viagens em automóvel na mesma razão, com potenciais benefícios para a sociedade de **`r fmt_num(euromacx4/1000)` milhões €** ao fim de 10 anos.

Já a **meta de 10%** de ciclistas diários equivale a ter **`r fmt_num(cicmacx10/1000)` mil ciclistas** em viagens até 10 km, com uma redução potencial de **`r fmt_num(co2macx10/1000)` mil toneladas** de CO~2~eq por ter menos viagens em automóvel na mesma razão, com potenciais benefícios para a sociedade de **`r round(euromacx10/1000000, 1)` mil milhões €** ao fim de 10 anos.

Relativamente ao cenário 3, de transferência modal de viagens em automóvel que são passíveis de serem realizadas em **intermodalidade entre bicicleta + transportes públicos** (viáveis), e cuja distância em bicicleta não é superior a 5km no total da primeira e última parte da viagem, a tabela seguinte mostra o potencial de transferência estimado para cada modo de transporte público, bem como as **estimativas de consumo de CO~2~eq evitado anualmente** por trasferência do automóvel, considerando também o consumo dos transportes públicos, no cômputo dessas viagens.

```{r tabela2}
CAR_TP_TRANSF = readRDS("HEAT/CAR_TP_TRANSF.Rds")
HEAT_aml_results2 = readRDS("HEAT/HEAT_aml_results.Rds")

TPtable_car = CAR_TP_TRANSF %>%
  group_by(LTS, mode_TP) %>% 
  summarise(TP_4 = sum(TP_4), TP_10 = sum(TP_10)
            # Total = sum(Total), Bike = sum(Bike), Car = sum(Car), TP = sum(TP), 
            # Bike_4 = sum(Bike_4), Bike_10 = sum(Bike_10)
            ) %>% 
  ungroup() %>% 
  pivot_longer(c(3,4), names_to = "ENMAC", values_to = "viagens") %>% 
  pivot_wider(names_from = mode_TP, values_from = viagens) %>% 
  mutate (Total = rowSums(.[3:6])) %>% 
  as.data.frame()

TPtable_aml = HEAT_aml_results %>% 
  filter(Estrategia == "intermodal ate 5km ") %>% 
  select(33:40) %>%
  bind_cols(TPtable_car) %>% 
  select(`LTS...1`, Meta_ENMAC, Total, FERRY, RAIL, BUS, TRAM,
         CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, money_emissions_eur) %>%
  mutate(Percursos = recode(`LTS...1`, `3` = "seguro", `4` = "direto")) %>% 
  select(14, 2:13) %>% 
  data.frame()

sketch2 = htmltools::withTags(table( #repetir header
  class = 'display',
  thead(
    tr(
      th(rowspan = 2, 'Tipo de Percurso'),
      th(rowspan = 2, 'Meta ENMAC'),
      th(colspan = 5, 'Viagens TP em complemento com bicicleta'),
      th(colspan = 5, 'Emissões evitadas por substituição do automóvel (ton / ano)'),
      th(rowspan = 2, 'Valorização monetária (€)')
    ),
    tr(
      lapply(rep(c("Total", "Barco", "Comboio", "Autocarro", "Metro superfície",
                   "CO2eq", "CO", "PM10", "NOx", "VOC"), 1), th)
    )
  )
))

DT::datatable(TPtable_aml,
                filter = 'none',
                rownames = FALSE,
                container = sketch2,
                # colnames = c("Tipo de Percurso", "Meta ENMAC",  "Total", "Barco", "Comboio", "Autocarro",
                #              "Metro superfície", "CO2eq", "CO", "PM10", "NOx", "VOC",
                #              "Valorização monetária (€)"),
                extensions = c("RowGroup", "Buttons"),
                options = list(dom = "tB",
                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
                               pageLength = 4,
                               responsive = TRUE,
                               extensions = c('Responsive'))) %>% 
    DT::formatRound(c(3:9,11,13),0, mark = " ") %>% 
    DT::formatRound(c(10,12),1)

```

Estes resultados indicam que a transferência da viagem em automóvel para bicicleta + transportes públicos, dentro dos critétrios considerados (max 5km em bicicleta na soma da primeira e última _milha_, e sem transferências de modo ou linha), poderá ter **benefícios potenciais estimados entre `r round(TPtable_aml[1,13]/1000000,1)` e `r round(TPtable_aml[4,13]/1000000,1)` milhões de €**.

Neste cenário, `r fmt_num(round(min(TPtable_aml$Total/1000)))`mil a `r fmt_num(round(max(TPtable_aml$Total/1000)))`mil viagens podem ser substituídas do automóvel para Bicicleta + Transportes Públicos.

O modo com maior potencial de transferência modal, em complemento com a bicicleta, é o **comboio**, com `r round(TPtable_aml[2,5]/1000)`mil viagens

## Downloads

```{r downloads}

path_sf_1r = paste0(i, "_rededireta_cenario1.gpkg") 
path_sf_1l = paste0(i, "_redesegura_cenario1.gpkg")

path_sf_2r = paste0(i, "_rededireta_cenario2.gpkg")
path_sf_2l = paste0(i, "_redesegura_cenario2.gpkg")

path_sf_3r = paste0(i, "_rededireta_cenario3.gpkg")
path_sf_3l = paste0(i, "_redesegura_cenario3.gpkg")

```

Os mapas acima estão disponíveis para download, em formato _GeoPackage_ (.gpkg), que pode ser ligo em qualquer SIG moderno. É recomendável o software de acesso livre [QGIS](https://qgis.org/pt_PT/site/){target="_blank"} para leitura, visualização e análise de dados em detalhe.

1.    **Viagens até 5 km em bicicleta**

  *   [Rede direta](`r path_sf_1r`)
  *   [Rede segura](`r path_sf_1l`)

2.    **Viagens até 10 km em bicicleta elétrica**

  *   [Rede direta](`r path_sf_2r`)
  *   [Rede segura](`r path_sf_2l`)

3.    **Viagens intermodais até 5 km em bicicleta**

  *   [Rede direta](`r path_sf_3r`)
  *   [Rede segura](`r path_sf_3l`)

`r emo::ji("magnifying")` A rede disponibilizada poderá não corresponder à que é apresentada nos mapas acima, que mostram apenas os segmentos com mais de X viagens potenciais em bicicleta. Para visualizar os segmentos dependendo do número de viagens potenciais, utilizar os filtros do SIG, por exemplo, para visualizar só os segmentos com mais de 500 viagens, ou todos os segmentos com mais de 10 viagens em bicileta.

### 2022-12-15 code/biclarwww/municipio_minor_index.Rmd
add missing files
necessary to build the biclarwww website

--- 2022-12-14 code/biclarwww/municipio_index.Rmd
@@
 
-if (i == "vila franca de xira") {
-    i = "vfxira"
-    mun = "Vila Franca de Xira"
-}
-
-if (i == "setubal"){
-  mun = "Setúbal"
-}
-
-dontrun = !i %in% c("alcochete", "mafra", "montijo", "odivelas") # não mostrar mapa 3
-run = i %in% c("alcochete", "mafra", "montijo", "odivelas") # mostrar mensagem mapa 3
+dontrun = !i %in% c("alcochete", "mafra", "montijo", "odivelas", "sesimbra") # não mostrar mapa 3
+run = i %in% c("alcochete", "mafra", "montijo", "odivelas", "sesimbra") # mostrar mensagem mapa 3
 
@@
 
-} else if (i %in% c("lisboa", "amadora", "oeiras")){
+} else if (i %in% c("lisboa")){
 
@@
 
-# cenario 3
-
-if (i %in% c("lisboa", "almada", "cascais", "amadora", "sintra", "oeiras")){
-
-  mapa31r = sf::st_read("export_rnet/rnet_aml_3404_30.gpkg", quiet = TRUE)
-mapa31r = mapa31r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
-  mutate(tCO2eq = tCO2eq10/10,
-         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
-         bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa31l = sf::st_read("export_rnet/rnet_aml_3304_30.gpkg", quiet = TRUE)
-mapa31l = mapa31l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
-  mutate(tCO2eq = tCO2eq10/10,
-         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
-         bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa32r = sf::st_read("export_rnet/rnet_aml_3410_30.gpkg", quiet = TRUE)
-mapa32r = mapa32r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
-  mutate(tCO2eq = tCO2eq10/10,
-         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
-         bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa32l = sf::st_read("export_rnet/rnet_aml_3310_30.gpkg", quiet = TRUE)
-mapa32l = mapa32l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
-  mutate(tCO2eq = tCO2eq10/10,
-         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
-         bike18 = paste0("(", Bikeper, ") ", Bike))
-
-} else {
-
-mapa31r = sf::st_read("export_rnet/rnet_aml_3404_10.gpkg", quiet = TRUE)
-mapa31r = mapa31r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
-  mutate(tCO2eq = tCO2eq10/10,
-         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
-         bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa31l = sf::st_read("export_rnet/rnet_aml_3304_10.gpkg", quiet = TRUE)
-mapa31l = mapa31l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
-  mutate(tCO2eq = tCO2eq10/10,
-         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
-         bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa32r = sf::st_read("export_rnet/rnet_aml_3410_10.gpkg", quiet = TRUE)
-mapa32r = mapa32r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
-  mutate(tCO2eq = tCO2eq10/10,
-         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
-         bike18 = paste0("(", Bikeper, ") ", Bike))
-mapa32l = sf::st_read("export_rnet/rnet_aml_3310_10.gpkg", quiet = TRUE)
-mapa32l = mapa32l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
-  mutate(tCO2eq = tCO2eq10/10,
-         quietness = as.integer(quietness), #não funciona pra efeitos da legenda
-         bike18 = paste0("(", Bikeper, ") ", Bike))
-
-}
+# 
+# # cenario 3
+# 
+# if (i %in% c("lisboa", "almada", "cascais", "amadora", "sintra", "oeiras")){
+#   
+#   mapa31r = sf::st_read("export_rnet/rnet_aml_3404_30.gpkg", quiet = TRUE)
+# mapa31r = mapa31r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+#   mutate(tCO2eq = tCO2eq10/10,
+#          quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+#          bike18 = paste0("(", Bikeper, ") ", Bike))
+# mapa31l = sf::st_read("export_rnet/rnet_aml_3304_30.gpkg", quiet = TRUE)
+# mapa31l = mapa31l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+#   mutate(tCO2eq = tCO2eq10/10,
+#          quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+#          bike18 = paste0("(", Bikeper, ") ", Bike))
+# mapa32r = sf::st_read("export_rnet/rnet_aml_3410_30.gpkg", quiet = TRUE)
+# mapa32r = mapa32r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+#   mutate(tCO2eq = tCO2eq10/10,
+#          quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+#          bike18 = paste0("(", Bikeper, ") ", Bike))
+# mapa32l = sf::st_read("export_rnet/rnet_aml_3310_30.gpkg", quiet = TRUE)
+# mapa32l = mapa32l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+#   mutate(tCO2eq = tCO2eq10/10,
+#          quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+#          bike18 = paste0("(", Bikeper, ") ", Bike))
+#   
+#   
+# } else {
+# 
+# mapa31r = sf::st_read("export_rnet/rnet_aml_3404_10.gpkg", quiet = TRUE)
+# mapa31r = mapa31r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+#   mutate(tCO2eq = tCO2eq10/10,
+#          quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+#          bike18 = paste0("(", Bikeper, ") ", Bike))
+# mapa31l = sf::st_read("export_rnet/rnet_aml_3304_10.gpkg", quiet = TRUE)
+# mapa31l = mapa31l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+#   mutate(tCO2eq = tCO2eq10/10,
+#          quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+#          bike18 = paste0("(", Bikeper, ") ", Bike))
+# mapa32r = sf::st_read("export_rnet/rnet_aml_3410_10.gpkg", quiet = TRUE)
+# mapa32r = mapa32r[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+#   mutate(tCO2eq = tCO2eq10/10,
+#          quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+#          bike18 = paste0("(", Bikeper, ") ", Bike))
+# mapa32l = sf::st_read("export_rnet/rnet_aml_3310_10.gpkg", quiet = TRUE)
+# mapa32l = mapa32l[MUNICIPIOSgeo %>% filter(Concelho == mun), , op = sf::st_within] %>% 
+#   mutate(tCO2eq = tCO2eq10/10,
+#          quietness = as.integer(quietness), #não funciona pra efeitos da legenda
+#          bike18 = paste0("(", Bikeper, ") ", Bike))
+# 
+# }
 
@@
 if (i %in% c("mafra", "sesimbra", "alcochete", "palmela")){ 
-  minimo1 = 15} else if (i %in% c("lisboa", "amadora", "oeiras")){ 
+  minimo1 = 15} else if (i %in% c("lisboa")){ 
     minimo1 = 100 
@@
 
-if (i %in% c("lisboa", "almada", "cascais", "amadora", "sintra", "oeiras")){
-  minimo3 = 30} else {
-    minimo3 = 10
-  }
+# if (i %in% c("lisboa", "almada", "cascais", "amadora", "sintra", "oeiras")){
+#   minimo3 = 30} else {
+#     minimo3 = 10
+#   }
 
@@
 
-if (i == "lisboa"){
+if (i %in% c("alcochete", "montijo")){
   ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = viagens, x = Inter)) +
@@
   scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
-  geom_text(aes(label = ifelse(viagens/1000 > 200, paste0(round(viagens/1000)," k"), "")),
-    position = position_stack(vjust = 0.5, reverse = FALSE))+
-  theme_minimal() +
-  theme(axis.title.y = element_blank(), legend.title = element_text("Modo"), legend.position = "right") +
-  coord_flip()
-} else if (i %in% c("amadora", "cascais", "loures", "oeiras")){
-  ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = viagens, x = Inter)) +
-  geom_bar(stat = "identity") +
-  scale_fill_manual(values = col_modes, name = "Modo") +
-  scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
-  geom_text(aes(label = ifelse(viagens/1000 > 19, paste0(round(viagens/1000)," k"), "")),
-    position = position_stack(vjust = 0.5, reverse = FALSE))+
-  theme_minimal() +
-  theme(axis.title.y = element_blank(), legend.title = element_text("Modo"), legend.position = "right") +
-  coord_flip()
-} else if (i %in% c("sintra")){
-  ggplot(TRIPSmode_mun_intr, aes(fill = modo, y = viagens, x = Inter)) +
-  geom_bar(stat = "identity") +
-  scale_fill_manual(values = col_modes, name = "Modo") +
-  scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
-  geom_text(aes(label = ifelse(viagens/1000 > 25, paste0(round(viagens/1000)," k"), "")),
+  geom_text(aes(label = ifelse(viagens/1000 > 5, paste0(round(viagens/1000)," k"), "")),
     position = position_stack(vjust = 0.5, reverse = FALSE))+
@@
   scale_y_continuous("Viagens / dia", labels = scales::comma_format(big.mark = ' ')) +
-  geom_text(aes(label = ifelse(viagens/1000 > 10, paste0(round(viagens/1000)," k"), "")),
+  geom_text(aes(label = ifelse(viagens/1000 > 12, paste0(round(viagens/1000)," k"), "")),
     position = position_stack(vjust = 0.5, reverse = FALSE))+
@@
 * viagens até 10 km com ligação em bicicleta elétrica (_`r fmt_num(HEAT_mun[8,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
-* viagens até 5 km em bicicleta de ou até uma interface de transportes (_`r fmt_num(HEAT_mun[12,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+<!-- * viagens até 5 km em bicicleta de ou até uma interface de transportes (_`r fmt_num(HEAT_mun[12,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_) -->
 
@@
 
-Este mapa mostra os segmentos com mais de `r minimo3` viagens potenciais em bicicleta, em que parte da viagem é realizada em transportes públicos (barco, comboio, metro sul do Tejo, ou carris metropolitana).
+<!-- Este mapa mostra os segmentos com mais de `r minimo3` viagens potenciais em bicicleta, em que parte da viagem é realizada em transportes públicos (barco, comboio, metro sul do Tejo, ou carris metropolitana).
 Os segmentos deste mapa dizem respeito apenas à parte da viagem que é realizada em bicicleta, num total de até 5 km no conjunto da primeira e última etapa da viagem intermodal.
 Para um maior ou menor detalhe da rede, ver a secção de [Downloads].
+-->
 
@@
 
-if (i == "sesimbra"){
-  scale1 = 20
-  scale2 = 30
-} else {
-  scale1 = 30
-  scale2 = 40
-}
-
 m3 = tmap::tm_shape(mapa31r, name = "Potencial 4% - direto") +
@@
                    alpha = 0.9,
-                   scale = scale1,
+                   scale = 30,
                    lwd_multiplier = 18,
@@
                    alpha = 0.9,
-                   scale = scale1,
+                   scale = 40,
                    lwd_multiplier = 18,
@@
                    alpha = 0.9,
-                   scale = scale2,
+                   scale = 30,
                    lwd_multiplier = 18,
@@
                    alpha = 0.9,
-                   scale = scale2,
+                   scale = 40,
                    lwd_multiplier = 18,
@@
 
-```{r conditional_block, echo=FALSE, results='asis', eval=run}
-cat("
+> Lamentamos, mas nesta área não existe rede ciclável potencial para o cenário 3, com as caraceterísticas definidas.
 
-> Lamentamos, mas nesta área não existe rede ciclável potencial para o cenário 3, com as caraceterísticas definidas.")
-```
-
-Ver mapa em [ecrã inteiro](mapa_cenario3.html){target="_blank"}.
+<!-- Ver mapa em [ecrã inteiro](mapa_cenario3.html){target="_blank"}. -->
 
@@
 
-htmlwidgets::saveWidget(m3leaf, file = paste0("biclarwww/", i,"/mapa_cenario3.html"))
+# htmlwidgets::saveWidget(m3leaf, file = paste0("biclarwww/", i,"/mapa_cenario3.html"))
 ```
@@
 tabelas_mun = tabelas_mun %>% filter(Municipio == mun) %>%
+  filter(Estrategia != "intermodal até 5km ") %>% #só nestes casos pequenos
   mutate(Percursos = recode(LTS, `3` = "seguro", `4` = "direto")) %>% 
@@
 
-  *   [Rede direta](`r path_sf_3r`)
-  *   [Rede segura](`r path_sf_3l`)
+  *   _`r proposicao` `r mun` não existe um mínimo de segmentos suficientes com as características deste cenário._
+
+  <!-- *   [Rede direta](`r path_sf_3r`) -->
+  <!-- *   [Rede segura](`r path_sf_3l`) -->
 
 Para cada município, os dados disponíveis para download referem-se à rede ciclável potencial com uma margem de 300 metros para além dos limites do município, para que sejam consideradas as ligações ao município vizinho.
-(_Exceto no cenário 3_)
 

### 2022-12-15 code/biclarwww/aml_index.Rmd
minor updates to www

@@
                               text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_baseline_1))) +
-  geom_smooth(span = 0.2, aes(cyc_dist_km,
-                              cumsum(trips_cyc_baseline_1),
-                              color="IMOB 2018", 
-                              group =3,
-                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_baseline_1)), 
-              linetype = "dashed") +
   # geom_smooth(span = 0.2, aes(cyc_dist_km,
-  #                             trips_all_baseline_1,
-  #                             color="Todos os modos", 
+  #                             cumsum(trips_cyc_baseline_1),
+  #                             color="IMOB 2018", 
   #                             group =3,
-  #                             text = paste("Distancia:", cyc_dist_km, "<br>Viagens total:", trips_all_baseline_1))) +
+  #                             text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_baseline_1)), 
+  #             linetype = "dashed") +
   scale_y_continuous("Viagens em bicictela /dia", labels = label_number(suffix = " k", scale = 1e-3)) +
-  scale_x_continuous("Distancia de viagem (km)", breaks = seq(0,30, by=5)) +
-  labs(title="", color='Cenario') +
+  scale_x_continuous("Distância de viagem (km)", breaks = seq(0,30, by=5)) +
+  labs(title="", color='Cenário') +
   theme_minimal() +  
@@
 
-A linha a tracejado representa os valor comulativo de número de viagens em bicicleta por dia do IMOB, para cada distância.
-Como se pode observar, o maior potencial de viagens em bicicleta situa-se em viagens até 10 km, sendo que os primeiros 5 km são muito expressivos.
+As linhas a tracejado, no gráfico seguinte, representam os valores comulativos de número de viagens em bicicleta por dia do IMOB, para cada distância.
+Como se pode observar, o maior potencial de viagens em bicicleta situa-se em viagens até 10 km, sendo que os primeiros 5 km são muito expressivos, e os ganhos potenciais após os 10 km são residuais.
+
+```{r plottly2}
+# gráfico com caracterização viagens IMOB intra municipais e intermonicipais
+
+p3 = ggplot(routes_13r_plot) + 
+  geom_smooth(span = 0.2, aes(cyc_dist_km,
+                              cumsum(trips_cyc_enmac10_1),
+                              color="ENMAC 10%",
+                              group =1,
+                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac10_1)),
+              linetype = "dashed") +
+  geom_smooth(span = 0.2, aes(cyc_dist_km,
+                              cumsum(trips_cyc_enmac4_1),
+                              color="ENMAC 4%",
+                              group =2,
+                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac4_1)),
+              linetype = "dashed") +
+  geom_smooth(span = 0.3, aes(cyc_dist_km,
+                              cumsum(trips_cyc_enmac10_3),
+                              color="Intermodal ENMAC 10%",
+                              group =1,
+                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac10_3)),
+              linetype = "dashed") +
+  geom_smooth(span = 0.3, aes(cyc_dist_km,
+                              cumsum(trips_cyc_enmac4_3),
+                              color="Intermodal ENMAC 4%",
+                              group =2,
+                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_enmac4_3)),
+              linetype = "dashed")  +
+  geom_smooth(span = 0.2, aes(cyc_dist_km,
+                              cumsum(trips_cyc_baseline_1),
+                              color="IMOB 2018", 
+                              group =3,
+                              text = paste("Distancia:", cyc_dist_km, "<br>Viagens em bicicleta:", trips_cyc_baseline_1)), 
+              linetype = "dashed") +
+  scale_y_continuous("Viagens em bicictela /dia (cumulativo)", labels = label_number(suffix = " k", scale = 1e-3)) +
+  scale_x_continuous("Distância de viagem (km)", breaks = seq(0,30, by=5)) +
+  labs(title="", color='Cenário') +
+  theme_minimal() +  
+  guides(color=guide_legend(override.aes=list(fill=NA)), lty=guide_legend(override.aes=list(fill=NA))) +
+  scale_color_manual(values = c("blue", "darkgreen", "black", "red", "orange"))
+
+ggplotly(p3) %>% config(displaylogo = FALSE) %>% 
+  layout(hovermode = "x") %>% style(hoverinfo = "y")
+
+```
+
+<details>
+  <summary>**Porque é que não ha viagens com mais de 12 km no cenário intermodal?**</summary>
+Esta limitação tem a ver com a forma como foram estimadas as viagens passíveis de serem realizadas em bicicleta nos vários cenários, em que no cenário intermodal considerámos uma duração máxima razoável de 2 horas.
+A soma das distâncias da primeira e última _milha_ ("em cima de uma bicicleta") nessas viagens não ultrapassa os 12 km.
+</details>
 

### 2022-12-15 code/test-code/heat-prep.R

@@ -203,8 +203,8 @@ webapp_input[["ap_inflationrate"]] = 3 # inflation rate
 # Adjusting default and background values
 genericdata$value[genericdata$parameter_name == "occupancy_rate_car"] = 1.6 #car occupancy rate for all purposes. default: 2
 genericdata$value[genericdata$parameter_name == "default_activemode_bike_fromcar"] = 100 # percentage of trips shifting from CAR. default: 30
-genericdata$value[genericdata$parameter_name == "default_activemode_bike_frompt"] = 0 # percentage of trips shifting from CAR. default: 50
-genericdata$value[genericdata$parameter_name == "default_activemode_bike_fromwalk"] = 0 # percentage of trips shifting from CAR. default: 20
+genericdata$value[genericdata$parameter_name == "default_activemode_bike_frompt"] = 0 # percentage of trips shifting from PT default: 50
+genericdata$value[genericdata$parameter_name == "default_activemode_bike_fromwalk"] = 0 # percentage of trips shifting from Walk. default: 20
 
 
 

### 2022-12-15 code/test-code/quietness-networks.R

@@ -128,11 +128,11 @@ rnet_joined$slope = scales::label_percent(accuracy = 0.1, sufix = " %")(rnet_joi
 ####
 
 # library(biclar)
-rnet_joined = readRDS("export2/rnet_ferry4_ebike_overline_morethan50_clean_tags.Rds")
+rnet_joined = readRDS("export2/rnet_allmodesNSub4_overline_morethan10_clean_tags.Rds")
 tm_rnet(rnet_joined, #filtrar barreiro [MUNICIPIOSgeo %>% filter(Concelho == "Barreiro"), , op = sf::st_within]
         lwd = "cyc4", #Baseline, ENMAC4, ENMAC10
         col = "quietness",
-        palette = "-reds", # "linear_yl_rd_bk" "johnson", "mako", "burg", "reds" - reds for fastest, mako for quietest
+        palette = "-burg", # "linear_yl_rd_bk" "johnson", "mako", "burg", "reds" - reds for fastest, mako for quietest
         scale = 15,
         lwd_multiplier = 15
 )

### 2022-12-15 code/test-code/r5r_filter_overline_routes_CORRECT.R

@@ -853,7 +853,7 @@ rnet_allmodesNSub3_overline_morethan30_clean = rnet_allmodesNSub3_overline_moret
   filter(group %in% groupp) #more than 10 segments connected
 rm(rnet_allmodesNSub3_overline_morethan30_group, group)
 
-tm_rnet(rnet_allmodesNSub3_overline_morethan30_clean,
+tm_rnet(rnet_allmodesNSub3_overline_morethan10_clean,
         lwd = "Bike", #Baseline, ENMAC4, ENMAC10
         # col = "Quietness",
         palette = "-burg", # "linear_yl_rd_bk" "johnson", "mako", "burg", "reds" - reds for fastest, mako for quietest

### 2022-12-19 code/biclarwww/aml_index.Rmd
maps 1 and 2 as image to lighten
improvments to website version

@@
 # `r mun`
+
+## Cenário Base
 
@@
 Esta limitação tem a ver com a forma como foram estimadas as viagens passíveis de serem realizadas em bicicleta nos vários cenários, em que no cenário intermodal considerámos uma duração máxima razoável de 2 horas.
-A soma das distâncias da primeira e última _milha_ ("em cima de uma bicicleta") nessas viagens não ultrapassa os 12 km.
+A soma das distâncias e durações da primeira e última _milha_ ("em cima de uma bicicleta") nessas viagens não ultrapassa os 12 km.
 </details>
+
+```{r sideplottly}
+plotly::subplot(p2, p3) %>% 
+  config(displaylogo = FALSE) %>% 
+  layout(hovermode = "x") %>%
+  style(hoverinfo = "y")
+```
 
@@
 
+[![Mapa AML cenário 1](img/aml_cenario1_cover.png)](mapa_cenario1.html){target="_blank"}
+
 > Ver o mapa em [ecrã inteiro](mapa_cenario1.html){target="_blank"}.
@@
 ```
+
+[![Mapa AML cenário 2](img/aml_cenario2_cover.png)](mapa_cenario2.html){target="_blank"}
 
@@
   group_by(LTS, mode_TP) %>% 
-  summarise(TP_4 = sum(TP_4), TP_10 = sum(TP_10)
-            # Total = sum(Total), Bike = sum(Bike), Car = sum(Car), TP = sum(TP), 
-            # Bike_4 = sum(Bike_4), Bike_10 = sum(Bike_10)
-            ) %>% 
+  summarise(TP_4 = sum(TP_4), TP_10 = sum(TP_10)) %>% 
   ungroup() %>% 
@@
 `r emo::ji("magnifying")` A rede disponibilizada poderá não corresponder à que é apresentada nos mapas acima, que mostram apenas os segmentos com mais de X viagens potenciais em bicicleta.
-Para visualizar os segmentos dependendo do número de viagens potenciais, utilizar os filtros do SIG, por exemplo, para visualizar só os segmentos com mais de 500 viagens, ou todos os segmentos com mais de 10 viagens em bicileta.
+Para visualizar os segmentos dependendo do número de viagens potenciais, utilizar os filtros do SIG, por exemplo, para visualizar só os segmentos com mais de 500 viagens, ou todos os segmentos com mais de 10 viagens em bicicleta.

### 2022-12-19 code/biclarwww/index.Rmd
minor updates to www
improvments to website version

@@
 Na AML, são realizadas cerca de 5 milhões de viagens todos os dias, em vários modos.
-O número de viagens em bicicleta, segundo o IMOB [2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
+O número de viagens em bicicleta, segundo o IMOB ([2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
 
@@
 
+## FAQ
+
+Pergutas e respostas simplificadas.
+
+### O que é a ferramenta biclar?
+
+O biclar é uma ferramenta online que informa os decisores sobre o planeamento e investimentos em infraestrutura ciclável, de modo a atingir as metas da Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}).
+
+### Qual o objetivo da ferramenta biclar?
+
+Na ótica do decisor, pretende-se que esta informe os decisores, em comparação com o cenário atual de utilização de bicicleta, sobre:
+
+*   Quais as ligações prioritárias e percursos entre Freguesias e entre Municípios, a investir com infraestrutura ciclável, consoante o cenário adotado e a rede (primária ou secundária);
+*   Qual o incremento potencial de utilizadores de bicicleta de cada infraestrutura ciclável, bem como qual a substituição modal (redução de outros modos);
+*   Quais os impactes previstos a médio e longo prazo (1 e 10 anos) em termos ambientais (CO~2~eq evitado), e sociais.
+
+### Como ler os mapas disponibilizados?
+
+Os mapas apresentam a informação gráfica para duas redes - a rede ciclável para viagens mais diretas, e a rede para viagens mais seguras e tranquilas (conforme a rede viária existente).  
+A largura das linhas varia consoante o número de ciclistas potencial para cada um dos cenários (4% e 10% de ciclistas), e essa informação é visível ao passar com o rato em cima da linha.  
+A cor das linhas varia com o nível de tranquilidade ([_quietness_](https://www.cyclestreets.net/help/journey/howitworks/#quietness){target="_blank"}), sendo as mais escuras aquelas menos seguras para circular em bicicleta - indicando com maior clareza quais os segmentos que necessitam de intervenção para se tornarem mais seguros e potencialmente trazerem mais ciclistas: os **mais escuros e mais largos**.
+
+Ao clicar em cada linha, uma janela abre-se com mais informação sobre as viagens estimadas que podem passar por aquele segmento, incluindo os benefícios sociais e ambientais estimados, e as características físicas daquele segmento.
+
+É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel.
+
+### Posso descarregar os mapas em formato tipo _shapefile_?
+
+Sim! Na secção de downloas de cada página individual dos municípios, ou de toda a AML.
+
+Os mapas acima estão disponíveis em formato _GeoPackage_ (.gpkg), que pode ser ligo em qualquer SIG moderno.
+É recomendável o software de acesso livre QGIS para leitura, visualização e análise de dados em detalhe.
+
+### Quero ver uma rede com mais detalhe, é possível?
+
+A rede disponibilizada poderá não corresponder à que é apresentada nos mapas, que mostram apenas os segmentos com mais de X viagens potenciais em bicicleta no cenrário de 4%.
+Disponibilizamos a rede com maior detalhe possível para todos os cenários, com um mínimo de **15** viagens/segmento no cenário 1, **50** viagens/segmento no cenário 2, e **10** viagens/segmento no cenário 3.
+Poderíamos disponibilizar a rede sem qualquer mínimo, mas em termos de planeamento de infraestruturas cicláveis, consideramos que é mais útil informar sobre aquelas com maior potencial de transferência modal.
+
+Para visualizar os segmentos dependendo do número de viagens potenciais: utilizar os filtros do SIG, por exemplo, para visualizar só os segmentos com mais de 500 viagens, ou todos os segmentos com mais de 10 viagens em bicicleta.
+
+### Quero ver a rede ciclável com maior potencial de transferência modal, é possível?
+
+Sim, tal como descrito na questão anterior, para visualizar os segmentos dependendo do número de viagens potenciais: utilizar os filtros do SIG, por exemplo, para visualizar só os segmentos com mais de 500 viagens.
+Quanto maior for o número mínimo de viagens a visualizar, mais importante ou prioritária é a rede (em termos de potencial de viagens transferidas do automóvel).
+
+Mas atenção que convém olhar também para os níveis de tranquilidade dos segmentos, cruzanto ambas as informações para encontrar a rede prioritária: aquela com maior potencial de tranferência & aquela que é atualmente menos segura para circular em bicicleta.
+
+### O que é o nível de tranquilidade?
+
+O nível de tranquilidade é estimado numa escala de 0 a 100 (em que o 100 é o nível mais tranquilo), e é definido pelo _CycleStreets_ (ver [_quietness_](https://www.cyclestreets.net/help/journey/howitworks/#quietness){target="_blank"}).
+
+Os caminhos fora de estrada, por exemplo em parques e jardins, têm o nível máximo de tranquilidade.
+Ruas calmas e com infrastrutura ciclável terão um nível ligeiramente inferior, e ruas principais com muito tráfego terão níveis muito baixos de tranquilidade.
+
+### As linhas dos mapas sugerem que devo panlear uma infraestrutura ciclável exatamente ali?
+
+Não.
+Os mapas mostram **sugestões de percursos**, mais diretos ou mais seguros, entre freguesias.
+Poderá haver outras soluções a ter em conta, como por exemplo a construção de uma ponte ou de uma passagem inferior, a abertura de um caminho inexistente por um jardim, etc.
+
+Para cada contexto poderão ser tomadas decisões e escolhidas soluções de entre uma variada panóplia.
+Recomendamos a consulta dos mais recentes manuais internacionais para ajudar a decidir sobre o tipo de solução escolhida.
+
+### Como são estimados os benefícios sociais e ambientais?
+
+Os benefícios potenciais são estimados em duas vertentes, utilizando os métodos e a ferramenta [_HEAT for Cycling_](https://www.heatwalkingcycling.org/){target="_blank"}, da Organização Mundial de Saúde.
+
+A componente ambiental é medida em toneladas de CO~2~eq evitado, por transferência de viagens equivalentes em automóvel, com a mesma origem e destino, mas não necessariamente com o mesmo percurso.
+
+Os impactes sociais agregados, na componente da saúde, são estimados em termos de redução/aumento da mortalidade por aumento de atividade física, por exposição à poluição atomosférica, e por exposição ao risco de sinistralidade rodoviária.
+Este valor é por fim monetizado utilizando o _Valor Estatístico da Vida_ (€3.055.358, segundo [ANSR 2021](http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O%20Impacto%20Economico%20e%20Social%20da%20Sinistralidade%20-%20PT.pdf){target="_blank"}), estimado para 10 anos, com uma taxa de desconto de 5% e inflação de 3%.
+
+### Posso fazer o mesmo tipo de análise para outra cidade ou área metropolitana?
+
+Claro!
+
+Este é um dos pontos-chave deste projeto: a sua reproducibilidade, transparência, e adaptação para outros contextos ou novos cenários.
+
+Todo o código utilizado está disponivel no repositório de dados abertos criado com este projeto, em [github.com/U-Shift/biclar](https://github.com/U-Shift/biclar){target="_blank"}.
+
+Tendo dados de viagens para caracterizar o cenário base (ou dados simulados), e mais alguns necessários dependendo dos cenários a considerar, é possível sim.
+
+### Gostava de avaliar o potencial ciclável para um outro cenário, é possível?
+
+Sim, ver a resposta à questão anterior.
+
+Sugerimos começar por fazer download do [projeto e scripts utilizados](https://github.com/U-Shift/biclar){target="_blank"}, carregando em _Code_ e _Download ZIP_.
+Alguns dados estão disponíveis na parte de [_Releases_](https://github.com/U-Shift/biclar/releases){target="_blank"}.
+Sendo um utilizador GitHub, é também possível simplesmente fazer um [_Fork_](https://github.com/U-Shift/biclar/fork){target="_blank"} ao projeto para a sua conta.
+A partir daí, e tendo algumas noções de _R_, é possível adaptar o código para criar outros cenários.
+
+Se considerar que o novo cenário desenvolvido deveria também constar nesta ferramenta, é possível fazer um [_pull request_](https://github.com/U-Shift/biclar/pulls){target="_blank"}, e a equipa biclaR irá avaliar os contributos e decidir sobre a sua integração.
+Esta é a beleza do código-aberto e da colaboração `r emo::ji("slightly_smiling_face")`.
+
+> Ver o [relatório metodológico](nada) para mais detalhes sobre o funcionamento desta ferramenta.
+
+<!-- ------------------------------------------------------------------------ -->
+
 ## Sobre
@@
 
-É também possível saber a informação agregada por [município](Municípios), para cada cenário escolhido, sobre todos os valores acima descritos.
+É também possível saber a informação agregada por [município](Municípios) ou para toda a [AML](AML), para cada cenário escolhido, sobre todos os valores acima descritos.
 
@@
 *   Grupo de Trabalho: ...
-*   Conselho Consultuvo: ....
+*   Conselho Consultivo: ....
 *   Matheus Correia, OpenStreetMap Portugal
@@
 *   r5r
+*   ATUMIE team
 
@@
 
-<!-- ------------------------------------------------------------------------ -->
+### Última atualização
 
-## FAQ
-
-Pergutas e respostas simplificadas.
-Ver o [relatório metodológico](nada) para mais detalhes sobre o funcionamento desta ferramenta..
-
-### O que é a ferramenta biclar?
-
-tal tal.
-
-### Qual o objetivo da ferramenta biclar?
-
-tal tal tal.
+Este website foi atualizado pela última vez a `r Sys.Date()`, pelas `r format(as.POSIXct(Sys.time()), format = "%H:%M")`.
 

### 2022-12-19 code/test-code/simple_statistics.R

@@ -220,3 +220,33 @@ p2 = ggplot(routes_13r_plot) +
   scale_color_manual(values = c("blue", "darkgreen", "black", "red", "orange")) #add  "black" if using all_modes
 ggplotly(p2) %>% config(displaylogo = FALSE) %>% 
   layout(hovermode = "x") %>% style(hoverinfo = "y")
+
+
+
+
+# Car -----------------------------------------------------------------------------------------
+
+
+routescar_plot = routes1r %>% 
+  mutate(length_km = distance/1000) %>% 
+  mutate(distance_band = cut(length_km, 0:20)) %>% 
+  group_by(distance_band) %>% 
+  summarise(cyc_dist_km = mean(length_km),
+            trips_cyc_baseline_1 = sum(Bike),
+            trips_car_baseline_1 = sum(Car),
+            trips_cyc_enmac4_1 = sum(cyc4),
+            trips_cyc_enmac10_1 = sum(cyc10),
+            trips_all_baseline_1 = sum(Total)) %>% 
+  mutate_if(is.numeric, round, digits = 2) %>% 
+  mutate_at(c(3:7), round, digits = 0)
+
+
+
+# juntar
+routes_car_plot =
+  routes1r_plot %>% 
+  left_join(routescar_plot %>% select(-cyc_dist_km))
+
+routes_car_plot2 =
+  routescar_plot %>% 
+  left_join(routes1r_plot %>% select(-cyc_dist_km))

### 2022-12-19 code/biclarwww/index.Rmd
FAQ correct
Corrigido com sugestões do @filmoura

@@
 
+### O biclaR faz um estudo de procura para o modo ciclável na AML?
+
+Não.
+Como explicado na questão anterior, o biclaR identifica quais as ligações prioritárias a estabelecer entre pares origem-destino numa determinada geografia por forma a acomodar a transferência de viagens em modos motorizados para a bicicleta.
+Para tal são definidos cenários e pressupostos para essa transferência potencial.
+Um estudo de procura implicaria uma estimativa da probabilidade de transferência dos modos motorizados para a bicicleta, não sendo esse o objectivo deste projeto.
+
 ### Como ler os mapas disponibilizados?
@@
 
-É possível ver mais do que uma opção de rota em simultâneo (direta / segura), por exemplo para comparar alternativas, e níveis de potencial ciclável, estabelecidos pela Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), em 4 e 10% nos centros urbanos, transferidos de viagens em automóvel.
+É possível ver mais do que uma opção de rota em simultâneo (direta / segura).
+Por exemplo, podem-se comparar alternativas e cenários de transferência de 4% e 10% das viagens em automóvel privado para a bicicleta, nos centros urbanos, tal como estabelecido pela Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}).
 
@@
 
-Sim! Na secção de downloas de cada página individual dos municípios, ou de toda a AML.
+Sim! Na secção de downloads de cada página individual dos municípios, ou de toda a AML.
 
-Os mapas acima estão disponíveis em formato _GeoPackage_ (.gpkg), que pode ser ligo em qualquer SIG moderno.
+Estes mapas estão disponíveis em formato _GeoPackage_ (.gpkg), que pode ser ligo em qualquer SIG moderno.
 É recomendável o software de acesso livre QGIS para leitura, visualização e análise de dados em detalhe.
@@
 
-A rede disponibilizada poderá não corresponder à que é apresentada nos mapas, que mostram apenas os segmentos com mais de X viagens potenciais em bicicleta no cenrário de 4%.
+A rede disponibilizada poderá não corresponder à que é apresentada nos mapas, que mostram apenas os segmentos com uma quantidade mínima de viagens potenciais em bicicleta no cenário de 4%.
 Disponibilizamos a rede com maior detalhe possível para todos os cenários, com um mínimo de **15** viagens/segmento no cenário 1, **50** viagens/segmento no cenário 2, e **10** viagens/segmento no cenário 3.
@@
 
-### As linhas dos mapas sugerem que devo panlear uma infraestrutura ciclável exatamente ali?
+### As linhas dos mapas sugerem que devo planear uma infraestrutura ciclável exatamente ali?
 
@@
 Recomendamos a consulta dos mais recentes manuais internacionais para ajudar a decidir sobre o tipo de solução escolhida.
+
+### Como posso avaliar se uma rua ou avenida é adequada para incluir na rede ciclável futura?
+
+Deverá localizar esse segmento da rede viária nos mapas que identificam as ligações prioritárias e percursos para os vários cenários que pretende analisar, e verificar se foi selecionado nessa rede, ou se é uma boa alternativa (próxima) a uma das ligações identificadas.
+
+Poderá avaliar igualmente qual o potencial de transferência que oferece, se se trata de uma ligação mais segura ou mais direta, e que impactes potenciais se pode esperar, designadamente em termos de impactes sociais e ambientais.
 
@@
 
-Este é um dos pontos-chave deste projeto: a sua reproducibilidade, transparência, e adaptação para outros contextos ou novos cenários.
+Este é um dos pontos-chave deste projeto: a sua reprodutibilidade, transparência, e adaptação para outros contextos ou novos cenários.
 

### 2022-12-19 code/biclarwww/aml_index.Rmd
aml plot side by site

@@
   # scale_linetype_manual(name = "valor", values = c("solid", "dashed"), labels = c("as it is", "comulativo")) #tipo de linha
-ggplotly(p2) %>% config(displaylogo = FALSE) %>% 
-  layout(hovermode = "x") %>% style(hoverinfo = "y")
+# ggplotly(p2) %>% config(displaylogo = FALSE) %>% 
+#   layout(hovermode = "x") %>% style(hoverinfo = "y")
 
 ```
-
-As linhas a tracejado, no gráfico seguinte, representam os valores comulativos de número de viagens em bicicleta por dia do IMOB, para cada distância.
-Como se pode observar, o maior potencial de viagens em bicicleta situa-se em viagens até 10 km, sendo que os primeiros 5 km são muito expressivos, e os ganhos potenciais após os 10 km são residuais.
-
 ```{r plottly2}
@@
 
-ggplotly(p3) %>% config(displaylogo = FALSE) %>% 
-  layout(hovermode = "x") %>% style(hoverinfo = "y")
+```
+```{r sideplottly}
+plotly::subplot(p2, p3, shareX = TRUE, titleY = TRUE) %>% 
+  config(displaylogo = FALSE) %>% 
+  layout(hovermode = "x") %>%
+  style(hoverinfo = "y")
+```
 
-```
+As linhas a tracejado representam os valores comulativos de número de viagens em bicicleta por dia do IMOB, para cada distância.
+Como se pode observar, o maior potencial de viagens em bicicleta situa-se em viagens até 10 km, sendo que os primeiros 5 km são muito expressivos, e os ganhos potenciais após os 10 km são residuais.
 
@@
 </details>
-
-```{r sideplottly}
-plotly::subplot(p2, p3) %>% 
-  config(displaylogo = FALSE) %>% 
-  layout(hovermode = "x") %>%
-  style(hoverinfo = "y")
-```
 
@@
 Para visualizar os segmentos dependendo do número de viagens potenciais, utilizar os filtros do SIG, por exemplo, para visualizar só os segmentos com mais de 500 viagens, ou todos os segmentos com mais de 10 viagens em bicicleta.
+
+#### Rede ciclável existente e prevista
+
+É também disponibilizado um [mapa da rede ciclável existente e prevista](https://ushift.tecnico.ulisboa.pt/content/tml/RedeExistentePrevista.html){target="_blank"} na AML em **Julho 2022**, e os seus ficheiros para transferência:
+
+*   [Rede ciclável existente, por tipologia](https://ushift.tecnico.ulisboa.pt/content/tml/RedeExistente.gpkg)
+*   [Rede ciclável prevista](https://ushift.tecnico.ulisboa.pt/content/tml/RedePrevista.gpkg)

### 2022-12-19 code/biclarwww/build.R
tidy build

@@ -1,7 +1,7 @@
 # Build the website 
 
 
-# Read in data --------------------------------------------------------------------------------
+# Main pages ----------------------------------------------------------------------------------
 
 
 # página principal
@@ -15,12 +15,11 @@ rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", encoding = "UTF-8", ou
 
 # Create outputs for each region --------------------------------------------------------------
 
-
-i = "oeiras"
+# i = "oeiras"
 i = "barreiro"
-i = "lisboa"
-i = "odivelas"
-i = "setubal"
+# i = "lisboa"
+# i = "odivelas"
+# i = "setubal"
 
 # for (i in municipios){
 

### 2023-02-27 code/biclarwww/index.Rmd
minor index update
AML to amL

@@
 
+library(tidyverse)
 library(sf)
@@
 
-<h2 style="text-align: center;">Ferramenta de apoio ao planeamento da rede ciclável<br> na Área Metropolitana de Lisboa</h2>
+<h2 style="text-align: center;">Ferramenta de apoio ao planeamento da rede ciclável<br> na área metropolitana de Lisboa</h2>
 
@@
 **O que mostra este mapa? **  
-Este mapa mostra a percentagem de viagens com origem em cada município que são realizadas em biciceta.
+Este mapa mostra a percentagem de viagens com origem em cada município que são realizadas em bicicleta.
 
@@
 
-Ver página com detalhes, mapas e mais resultados para a Área Metropolitana de Lisboa: [AML](aml/index.html){target="_blank"}
+Ver página com detalhes, mapas e mais resultados para a área metropolitana de Lisboa: [AML](aml/index.html){target="_blank"}
 
@@
 
-Para todos os detalhes sobre estes resultados, ver a página da [Área Metropolitana de Lisboa](aml/index.html){target="_blank"}
+Para todos os detalhes sobre estes resultados, ver a página da [área metropolitana de Lisboa](aml/index.html){target="_blank"}
 
@@
 Para tal são definidos cenários e pressupostos para essa transferência potencial.
+
 Um estudo de procura implicaria uma estimativa da probabilidade de transferência dos modos motorizados para a bicicleta, não sendo esse o objectivo deste projeto.

### 2023-02-27 code/biclarwww/aml_index.Rmd
AML to amL
IMOB to IMob

@@
 i = "aml"
-mun = "Área Metropolitana de Lisboa"
+mun = "área metropolitana de Lisboa"
 
@@
 
-Segundo o Inquérito à Mobilidade  ([IMOB](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2018 pelo INE, as viagens dentro da `r mun` têm as seguintes características.
+Segundo o Inquérito à Mobilidade  ([IMob](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2018 pelo INE, as viagens dentro da `r mun` têm as seguintes características.
 
 ```{r mododist, fig.ncol=2, out.width="50%"}
-#gráfico com bistribuição de viagens por modo na AML segundo IMOB 2018
+#gráfico com bistribuição de viagens por modo na AML segundo IMob 2018
 
@@
 
-As linhas a tracejado representam os valores comulativos de número de viagens em bicicleta por dia do IMOB, para cada distância.
+As linhas a tracejado representam os valores comulativos de número de viagens em bicicleta por dia do IMob, para cada distância.
 Como se pode observar, o maior potencial de viagens em bicicleta situa-se em viagens até 10 km, sendo que os primeiros 5 km são muito expressivos, e os ganhos potenciais após os 10 km são residuais.
@@
 
-Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMOB 2018), para **4%** de viagens em bicicleta (meta 2024), e para **10%** de viagens em bicicleta (meta 2030), transferidas diretamente de viagens em automóvel.
+Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMob 2018), para **4%** de viagens em bicicleta (meta 2024), e para **10%** de viagens em bicicleta (meta 2030), transferidas diretamente de viagens em automóvel.
 

### 2023-02-27 code/biclarwww/index.Rmd
IMOB to IMob

@@
 Na AML, são realizadas cerca de 5 milhões de viagens todos os dias, em vários modos.
-O número de viagens em bicicleta, segundo o IMOB ([2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
+O número de viagens em bicicleta, segundo o IMob ([2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
 

### 2023-02-27 code/biclarwww/municipio_index.Rmd
IMOB to IMob

@@
 
-Segundo o Inquérito à Mobilidade  ([IMOB](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2018 pelo INE, as viagens com origem e/ou destino `r stringr::str_to_lower(proposicao)` `r mun` têm as seguintes características.
+Segundo o Inquérito à Mobilidade  ([IMob](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2018 pelo INE, as viagens com origem e/ou destino `r stringr::str_to_lower(proposicao)` `r mun` têm as seguintes características.
 

### 2023-02-27 code/biclarwww/aml_index.Rmd
typos

@@
 
-Relativamente ao cenário 3, de transferência modal de viagens em automóvel que são passíveis de serem realizadas em **intermodalidade entre bicicleta + transportes públicos** (viáveis), e cuja distância em bicicleta não é superior a 5km no total da primeira e última parte da viagem, a tabela seguinte mostra o potencial de transferência estimado para cada modo de transporte público, bem como as **estimativas de consumo de CO~2~eq evitado anualmente** por trasferência do automóvel, considerando também o consumo dos transportes públicos, no cômputo dessas viagens.
+Relativamente ao cenário 3, de transferência modal de viagens em automóvel que são passíveis de serem realizadas em **intermodalidade entre bicicleta + transportes públicos**, e cuja distância em bicicleta não é superior a 5km no total da primeira e última parte da viagem, a tabela seguinte mostra o potencial de transferência estimado para cada modo de transporte público, bem como as **estimativas de consumo de CO~2~eq evitado anualmente** por transferência do automóvel, considerando também o consumo dos transportes públicos, no cômputo dessas viagens.
 
@@
 
-Estes resultados indicam que a transferência da viagem em automóvel para bicicleta + transportes públicos, dentro dos critétrios considerados (max 5km em bicicleta na soma da primeira e última _milha_, e sem transferências de modo ou linha), poderá ter **benefícios potenciais estimados entre `r round(TPtable_aml[1,13]/1000000,1)` e `r round(TPtable_aml[4,13]/1000000,1)` milhões de €**.
+Estes resultados indicam que a transferência da viagem em automóvel para bicicleta + transportes públicos, dentro dos critérios considerados (max 5 km em bicicleta na soma da primeira e última _milha_, e sem transferências de modo ou linha), poderá ter **benefícios potenciais estimados entre `r round(TPtable_aml[1,13]/1000000,1)` e `r round(TPtable_aml[4,13]/1000000,1)` milhões de €**.
 
@@
 
-O modo com maior potencial de transferência modal, em complemento com a bicicleta, é o **comboio**, com `r round(TPtable_aml[2,5]/1000)`mil viagens
+O modo com maior potencial de transferência modal, em complemento com a bicicleta, é o **comboio**, com `r round(TPtable_aml[2,5]/1000)` mil viagens
 
@@
 
-Os mapas acima estão disponíveis para download, em formato _GeoPackage_ (.gpkg), que pode ser ligo em qualquer SIG moderno.
+Os mapas acima estão disponíveis para download, em formato _GeoPackage_ (.gpkg), que pode ser lido em qualquer SIG moderno.
 É recomendável o software de acesso livre [QGIS](https://qgis.org/pt_PT/site/){target="_blank"} para leitura, visualização e análise de dados em detalhe.

### 2023-02-27 code/biclarwww/index.Rmd
typos

@@
 
-Pergutas e respostas simplificadas.
+Perguntas e respostas simplificadas.
 
@@
 
-O biclar é uma ferramenta online que informa os decisores sobre o planeamento e investimentos em infraestrutura ciclável, de modo a atingir as metas da Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}).
+O biclaR é uma ferramenta online que informa os decisores sobre o planeamento e investimentos em infraestrutura ciclável, de modo a atingir as metas da Estratégia Nacional para a Mobilidade Ativa Ciclável 2020-2030 ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}).
 
-Na ótica do decisor, pretende-se que esta informe os decisores, em comparação com o cenário atual de utilização de bicicleta, sobre:
+Pretende-se que esta ferramenta informe os decisores, em comparação com o cenário atual de utilização de bicicleta, sobre:
 
@@
 *   Qual o incremento potencial de utilizadores de bicicleta de cada infraestrutura ciclável, bem como qual a substituição modal (redução de outros modos);
-*   Quais os impactes previstos a médio e longo prazo (1 e 10 anos) em termos ambientais (CO~2~eq evitado), e sociais.
+*   Quais os impactes previstos a curto e longo prazo (1 e 10 anos) em termos ambientais (CO~2~eq evitado) e sociais.
 
@@
 
-*   Transportes Metropolitanos de Lisboa (coordenação): XX XXX XXXX.
+*   TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.
+(coordenação): António Sérgio Manso Pinheiro (Diretor do Departamento de Estudos e Projetos), Catarina Tavares Marcelino (Coordenadora), Camila Garcia, Conceição Bandarrinha, Luísa Nogueira, Pedro Machado.
 
@@
 *   r5r
+*   Thomas Götshi, HEAT for Cycling
 *   ATUMIE team

### 2023-02-27 code/biclarwww/index.Rmd
internal links and typos

@@
 
-Na AML, são realizadas cerca de 5 milhões de viagens todos os dias, em vários modos.
-O número de viagens em bicicleta, segundo o IMob ([2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
+Na área metropolitana de Lisboa são realizadas diariamente cerca de 5.3 milhões de viagens, em vários modos.
+O número de viagens em bicicleta, segundo o Inquérito à Mobilidade ([IMob 2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
 
@@
 
-Ver página com detalhes, mapas e mais resultados para a área metropolitana de Lisboa: [AML](aml/index.html){target="_blank"}
+Ver página com detalhes, mapas e mais resultados para a área metropolitana de Lisboa em [AML](aml/index.html){target="_blank"}
 
@@
 
-A AML tem uma população de `r fmt_num(POPaml)`. Diariamente são realizadas 5.33 milhões de viagens, em vários modos de transporte, das quais apenas **25 479 são realizadas em bicicleta** (0.5%).
+A área metropolitana de Lisboa tem uma população de `r fmt_num(POPaml)` habitantes, segundo os Censos 2021.
+Diariamente são realizadas 5.33 milhões de viagens, em vários modos de transporte, das quais apenas **25 479 são realizadas em bicicleta** (0.5%).
 
-A tabela informa sobre o potencial ciclável em cada [cenário](Cenários) considerado neste estudo, para ligações entre freguesias, realizáveis em bicicleta, e para percursos preferencialmente diretos.
+A tabela seguinte informa sobre o potencial ciclável em cada [cenário](#Cenários) considerado neste estudo, para ligações entre freguesias, realizáveis em bicicleta, e para percursos preferencialmente diretos.
 
@@
 
-É também possível saber a informação agregada por [município](Municípios) ou para toda a [AML](AML), para cada cenário escolhido, sobre todos os valores acima descritos.
+É também possível saber a informação agregada por [município](#Municípios) ou para toda a [AML](#AML), para cada cenário escolhido, sobre todos os valores acima descritos.
 

### 2023-02-27 code/biclarwww/index.Rmd
internal links

@@
 
-A tabela seguinte informa sobre o potencial ciclável em cada [cenário](#Cenários) considerado neste estudo, para ligações entre freguesias, realizáveis em bicicleta, e para percursos preferencialmente diretos.
+A tabela seguinte informa sobre o potencial ciclável em cada [cenário](#cenários) considerado neste estudo, para ligações entre freguesias, realizáveis em bicicleta, e para percursos preferencialmente diretos.
 
@@
 
-É também possível saber a informação agregada por [município](#Municípios) ou para toda a [AML](#AML), para cada cenário escolhido, sobre todos os valores acima descritos.
+É também possível saber a informação agregada por [município](#municípios) ou para toda a [AML](#aml), para cada cenário escolhido, sobre todos os valores acima descritos.
 

### 2023-02-27 code/biclarwww/aml_index.Rmd
export tables without thousand separator
buttons = list(list(extend="excel", exportOptions = list(orthogonal = "export"))),

@@
                 options = list(dom = "tB",
-                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
+                               buttons = list(list(extend='copy',text="Copiar"),
+                                              list(extend="excel", exportOptions = list(orthogonal = "export"))),
                                pageLength = 12,
@@
                 options = list(dom = "tB",
-                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
+                               buttons = list(list(extend='copy',text="Copiar"),
+                                              list(extend="excel", exportOptions = list(orthogonal = "export"))),
                                pageLength = 4,

### 2023-02-27 code/biclarwww/index.Rmd
export tables without thousand separator
buttons = list(list(extend="excel", exportOptions = list(orthogonal = "export"))),

@@
                 options = list(dom = "tB",
-                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
+                               buttons = list(list(extend='copy',text="Copiar"),
+                                              list(extend="excel", exportOptions = list(orthogonal = "export"))), #para salvar como tabela original sem espaços de separador milhares
                                pageLength = 3,
@@
                 options = list(dom = "tB",
-                               buttons = list(list(extend='copy',text="Copiar"), "excel"),
+                               buttons = list(list(extend='copy',text="Copiar"),
+                                              list(extend="excel", exportOptions = list(orthogonal = "export"))),
                                pageLength = 18,

### 2023-02-27 code/biclarwww/index.Rmd
acknowledgments
https://github.com/U-Shift/biclarwww/issues/1

@@
 
-Na ótica do decisor, pretende-se que esta informe os decisores, em comparação com o cenário atual de utilização de bicicleta, sobre:
+Pretende-se que esta ferramenta informe os decisores, em comparação com o cenário atual de utilização de bicicleta, sobre:
 
@@
 *   Qual o incremento potencial de utilizadores de bicicleta de cada infraestrutura ciclável, bem como qual a substituição modal (redução de outros modos);
-*   Quais os impactes previstos a médio e longo prazo (1 e 10 anos) em termos ambientais (CO~2~eq evitado), e sociais.
+*   Quais os impactes previstos a curto e longo prazo (1 e 10 anos) em termos ambientais (CO~2~eq evitado) e sociais.
 
-### O biclaR faz um estudo de procura para o modo ciclável na AML?
+### O biclaR faz um estudo de procura para o modo ciclável na área metropolitana de Lisboa?
 
@@
 Os impactes sociais agregados, na componente da saúde, são estimados em termos de redução/aumento da mortalidade por aumento de atividade física, por exposição à poluição atomosférica, e por exposição ao risco de sinistralidade rodoviária.
-Este valor é por fim monetizado utilizando o _Valor Estatístico da Vida_ (€3.055.358, segundo [ANSR 2021](http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O%20Impacto%20Economico%20e%20Social%20da%20Sinistralidade%20-%20PT.pdf){target="_blank"}), estimado para 10 anos, com uma taxa de desconto de 5% e inflação de 3%.
+Este valor é, por fim, monetizado utilizando o _Valor Estatístico da Vida_ (€3.055.358, segundo [ANSR 2021](http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O%20Impacto%20Economico%20e%20Social%20da%20Sinistralidade%20-%20PT.pdf){target="_blank"}), estimado para 10 anos, com uma taxa de desconto de 5% e inflação de 3%.
 
@@
 
-Todo o código utilizado está disponivel no repositório de dados abertos criado com este projeto, em [github.com/U-Shift/biclar](https://github.com/U-Shift/biclar){target="_blank"}.
+Todo o código utilizado está disponível no repositório de dados abertos criado com este projeto, em [github.com/U-Shift/biclar](https://github.com/U-Shift/biclar){target="_blank"}.
 
@@
 
-*   TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.
-(coordenação): António Sérgio Manso Pinheiro (Diretor do Departamento de Estudos e Projetos), Catarina Tavares Marcelino (Coordenadora), Camila Garcia, Conceição Bandarrinha, Luísa Nogueira, Pedro Machado.
+*   **Coordenação**: TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.
+- António Sérgio Manso Pinheiro (Diretor do Departamento de Estudos e Projetos), Catarina Tavares Marcelino (Coordenadora), Camila Garcia, Conceição Bandarrinha, Luísa Nogueira, Pedro Machado.
 
-*   Instituto Superior Técnico - Universidade de Lisboa (desenvolvimento): Filipe Moura, Rosa Félix.
+*   **Desenvolvimento**: Instituto Superior Técnico - Universidade de Lisboa - Filipe Moura, Rosa Félix.
 
-*   Institute for Transport Studies - University of Leeds (apoio ao desenvolvimento): Robin Lovelace.
+*   **Apoio ao desenvolvimento**: Institute for Transport Studies - University of Leeds - Robin Lovelace.
 
@@
 
-*   Grupo de Trabalho: ...
-*   Conselho Consultivo: ....
-*   Matheus Correia, OpenStreetMap Portugal
+*   Grupo de Trabalho dos Municípios, composto por representantes e técnicos dos 18 municípios da área metropolitana de Lisboa
+*   Comissão de Acompanhamento do projeto biclaR, composto por várias entidades e instituições de caráter público e associativo, organizado pela TML
+*   Matheus Correia - OpenStreetMap Portugal
+*   Rafael Pereira - r5r
+*   Thomas Götshi - HEAT for Cycling
 *   CycleStreets
-*   r5r
-*   Thomas Götshi, HEAT for Cycling
 *   ATUMIE team
@@
 
-Enviar um email para...
+Para qualquer esclarecimento adicional, por favor enviar um email para info @tmlmobilidade.pt com o assunto _biclaR_.
 

### 2023-02-28 code/biclarwww/aml_index.Rmd
typos

@@
 
-Este mapa mostra os segmentos com mais de `r minimo3` viagens potenciais em bicicleta, em que parte da viagem é realizada em transportes públicos (barco, comboio, metro sul do Tejo, ou carris metropolitana).
+Este mapa mostra os segmentos com mais de `r minimo3` viagens potenciais em bicicleta, em que parte da viagem é realizada em transportes públicos (barco, comboio, metro sul do Tejo, ou Carris Metropolitana).
 Os segmentos deste mapa dizem respeito apenas à parte da viagem que é realizada em bicicleta, num total de até 5 km no conjunto da primeira e última etapa da viagem intermodal.
@@
 
-A tabela mostra o potencial ciclável `r stringr::str_to_lower(proposicao)` `r mun` em número de novas viagens em bicicleta potenciais, para cada meta da Estratégia Nacional ENMAC, para cada cenário, e para cada tipo de rotas otimizadas.
+A tabela mostra o potencial ciclável `r stringr::str_to_lower(proposicao)` `r mun` em número de novas viagens em bicicleta potenciais diárias, para cada meta da Estratégia Nacional ENMAC, para cada cenário, e para cada tipo de rotas otimizadas.
 

### 2023-02-28 code/biclarwww/index.Rmd
typos

@@
 
-Para reportar algum problema com a ferramenta, ou colocar questões, pode abrir um _issue_ no nosso [repositório de dados abertos](https://github.com/U-Shift/biclarwww/issues).
+Para reportar algum problema com a ferramenta, ou colocar questões, pode abrir um _issue_ no nosso [repositório de dados abertos](https://github.com/U-Shift/biclarwww/issues){target="_blank"}.
 

### 2023-03-02 code/biclarwww/aml_index.Rmd
2025 and typos

@@
 
-Segundo o Inquérito à Mobilidade  ([IMob](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2018 pelo INE, as viagens dentro da `r mun` têm as seguintes características.
+Segundo o Inquérito à Mobilidade ([IMob 2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2017 pelo INE, as viagens dentro da `r mun` têm as seguintes características:
 
@@
 A figura seguinte mostra a proporção de viagens que é realizada com origem e destino no mesmo município (`r round(viagensintra/1000000,1)` milhões viagens), e as viagens com origem e destino em municípios diferentes (`r round(viagensinter/1000000,1)` milhões viagens).
-Como se pode observar, para todos os municípios, a maioria das viagens realizadas é intra municipal, emboras as viagens intermunicipais são bastante expressivas, principalmente em municípios como Odivelas, Oeiras, Amadora e Loures.
+Como se pode observar, para todos os municípios, a maioria das viagens realizadas é intra municipal, embora as viagens intermunicipais sejam bastante expressivas, principalmente em municípios como Odivelas, Oeiras, Amadora e Loures.
 
@@
 
-As metas da Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), estabelecem que dentro dos centros urbanos **4%** das viagens devem ser realizadas em bicicleta até 2024, e **10%** até 2030, e que as mesmas devem ser substituídas de viagens atualmente realizadas em automóvel.
-Partindo do número de viagens realizadas em bicicleta em 2018 (cenário de referência, ou _cenário base_), é possível estimar o número de viagens com potencial para serem substituídas do automóvel para a bicicleta para cada cenário aqui avaliado.
+As metas da Estratégia Nacional para a Mobilidade Ativa Ciclável 2020-2030 ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}) estabelecem que dentro dos centros urbanos **4%** das viagens devem ser realizadas em bicicleta até 2025, e **10%** até 2030, e que as mesmas devem ser transferidas de viagens atualmente realizadas em automóvel.
+Partindo do número de viagens realizadas em bicicleta em 2018 (cenário de referência, ou _cenário base_), é possível estimar o número de viagens com potencial para serem transferidas do automóvel para a bicicleta para cada cenário aqui avaliado.
 
@@
 
-Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMob 2018), para **4%** de viagens em bicicleta (meta 2024), e para **10%** de viagens em bicicleta (meta 2030), transferidas diretamente de viagens em automóvel.
+Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMob 2018), para **4%** de viagens em bicicleta (meta ENMAC 2025), e para **10%** de viagens em bicicleta (meta ENMAC 2030), transferidas diretamente de viagens em automóvel.
 
@@
 * viagens até 10 km com ligação em bicicleta elétrica (_`r fmt_num(HEAT_aml_redux[8,13])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
-* viagens até 5 km em bicicleta de ou até uma interface de transportes (_`r fmt_num(HEAT_aml_redux[12,13])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+* viagens até 5 km em bicicleta até ou desde uma interface de transportes (_`r fmt_num(HEAT_aml_redux[12,13])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
 
@@
 
-A tabela mostra o potencial ciclável `r stringr::str_to_lower(proposicao)` `r mun` em número de novas viagens em bicicleta potenciais diárias, para cada meta da Estratégia Nacional ENMAC, para cada cenário, e para cada tipo de rotas otimizadas.
+A tabela mostra o potencial ciclável `r stringr::str_to_lower(proposicao)` `r mun` em número de novas viagens em bicicleta potenciais diárias, para cada meta da Estratégia Nacional ENMAC (4% e 10%), para cada um dos três cenários, e para cada tipo de rotas otimizadas.
 
@@
 
-Os impactes sociais agregados, na componente da saúde, são estimados em termos de redução/aumento da mortalidade por aumento de atividade física, por exposição à poluição atomosférica, e por exposição ao risco de sinistralidade rodoviária.
+Os impactes sociais agregados, na componente da saúde, são estimados em termos de redução/aumento da mortalidade por aumento de atividade física, por exposição à poluição atmosférica, e por exposição ao risco de sinistralidade rodoviária.
 Este valor é por fim monetizado utilizando o _Valor Estatístico da Vida_ (€3.055.358, segundo [ANSR 2021](http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O%20Impacto%20Economico%20e%20Social%20da%20Sinistralidade%20-%20PT.pdf){target="_blank"}), estimado para 10 anos, com uma taxa de desconto de 5% e inflação de 3%.

### 2023-03-02 code/biclarwww/index.Rmd
2025 and typos

@@
 **O que mostra este mapa? **  
-Este mapa mostra a percentagem de viagens com origem em cada município que são realizadas em bicicleta.
+Este mapa mostra a percentagem de viagens que são realizadas em bicicleta com origem em cada município .
 
@@
 Na área metropolitana de Lisboa são realizadas diariamente cerca de 5.3 milhões de viagens, em vários modos.
-O número de viagens em bicicleta, segundo o Inquérito à Mobilidade ([IMob 2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
+O número de viagens em bicicleta, segundo o Inquérito à Mobilidade realizado em 2017 ([IMob 2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
 
@@
 
-Este cenário inclui todas as viagens que podem ser realizadas entre freguesias até 5 km em bicicleta.
+Este cenário inclui todas as viagens que podem ser realizadas por bicicleta entre freguesias, até 5 km.
 Inclui também viagens que utilizam o transporte fluvial (barcos) para unir ambas as margens, embora esse troço não seja considerado no total dos 5km.
 
-Este é um cenário conservador / realista.
+Este é um **cenário conservador / realista**.
 
@@
 
-Este cenário inclui todas as viagens que podem ser realizadas entre freguesias até 10 km em bicicleta, considerando para tal que os ciclistas poderão utilizar uma bicicleta elétrica.
+Este cenário inclui todas as viagens que podem ser realizadas por bicicleta entre freguesias, até 10 km - considerando para tal que os ciclistas poderão utilizar uma **bicicleta elétrica**.
 Inclui também viagens que utilizam o transporte fluvial (barcos) para unir ambas as margens, embora esse troço não seja considerado no total dos 10km.
 
-Este abrange uma maior gama de utilizadores, e tem em consideração possíveis políticas de incentivo à aquisição de bicicletas elétricas.
+Este abrange uma maior gama de utilizadores, e tem em consideração eventuais políticas de incentivo à aquisição de bicicletas elétricas.
 
@@
 
-Este cenário inclui todas as viagens que podem ser realizadas entre freguesias em transportes públicos, cuja primeira e última _milha_ da viagem podem ser feitas em bicicleta, até uma extensão total de 5 km - por exemplo: 1 km de bicicleta até ao interface A, e 4 km de bicicleta entre o interface B e o destino.
-Como transportes públicos, são considerados apenas o comboio, barco, metro sul do Tejo, e algumas carreiras da Carris Metropolitana. 
+Este cenário inclui todas as viagens que podem ser realizadas em transportes públicos, cuja primeira e última _milha_ da viagem podem ser feitas em bicicleta, entre freguesias até uma extensão total de 5 km - por exemplo: 1 km de bicicleta até ao interface A, e 4 km de bicicleta entre o interface B e o destino.
+Como transportes públicos, são considerados o comboio, barco, metro sul do Tejo, e algumas carreiras da Carris Metropolitana - em que é possível transportar uma bicicleta de forma prática.. 
 De uma forma conservadora, assumimos também que não são possíveis transferências entre transportes ou entre linhas/carreiras.
 
-Este valoriza a utilização da bicicleta como complemento ao transporte público.
+Este valoriza a utilização da bicicleta como **complemento ao transporte público**.
 
@@
 
-O biclar é uma ferramenta online que informa os decisores sobre o planeamento e investimentos em infraestrutura ciclável, de modo a atingir as metas da Estratégia Nacional de Mobilidade Ativa Ciclável ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}).
+O biclaR é uma ferramenta online que informa os decisores sobre o planeamento e investimentos em infraestrutura ciclável, de modo a atingir as metas da Estratégia Nacional de Mobilidade Ativa Ciclável 2020-2030 ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}), que definem que até 2025 4% das viagens realizadas em meio urbano deverão ser feitas em bicicleta, e 10% até 2030 - sendo as mesmas transferidas de viagens atualmente realizadas em automóvel.
 

### 2023-03-02 code/biclarwww/municipio_index.Rmd
2025 and typos

@@
 
-Segundo o Inquérito à Mobilidade  ([IMob](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2018 pelo INE, as viagens com origem e/ou destino `r stringr::str_to_lower(proposicao)` `r mun` têm as seguintes características.
+Segundo o Inquérito à Mobilidade ([IMob 2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) realizado em 2017 pelo INE, as viagens com origem e/ou destino `r stringr::str_to_lower(proposicao)` `r mun` têm as seguintes características:
 
@@
 
-Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMob 2018), para **4%** de viagens em bicicleta (meta 2024), e para **10%** de viagens em bicicleta (meta 2030), transferidas diretamente de viagens em automóvel.
+Os mapas seguintes mostram as ligações cicláveis com maior potencial, para o **cenário base** (viagens do IMob 2018), para **4%** de viagens em bicicleta (meta 2025), e para **10%** de viagens em bicicleta (meta 2030), transferidas diretamente de viagens em automóvel.
 
@@
 * viagens até 10 km com ligação em bicicleta elétrica (_`r fmt_num(HEAT_mun[8,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
-* viagens até 5 km em bicicleta de ou até uma interface de transportes (_`r fmt_num(HEAT_mun[12,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
+* viagens até 5 km em bicicleta até ou desde uma interface de transportes (_`r fmt_num(HEAT_mun[12,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
 

### 2023-03-02 code/biclarwww/municipio_minor_index.Rmd
2025 and typos

--- 2023-02-27 code/biclarwww/municipio_minor_index.Rmd
@@
 * viagens até 10 km com ligação em bicicleta elétrica (_`r fmt_num(HEAT_mun[8,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_)
-<!-- * viagens até 5 km em bicicleta de ou até uma interface de transportes (_`r fmt_num(HEAT_mun[12,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_) -->
+<!-- * viagens até 5 km em bicicleta até ou desde uma interface de transportes (_`r fmt_num(HEAT_mun[12,14])` viagens `r stringr::str_to_lower(proposicao)` `r mun`_) -->
 
@@
 
-> Lamentamos, mas nesta área não existe rede ciclável potencial para o cenário 3, com as caraceterísticas definidas.
+> Lamentamos, mas nesta área não existe rede ciclável potencial para o cenário 3, com as caraceterísticas mínimas definidas.
 
@@
 
-A tabela mostra o potencial ciclável `r stringr::str_to_lower(proposicao)` `r mun` em número de novas viagens em bicicleta potenciais, para cada meta da Estratégia Nacional ENMAC, para cada cenário, e para cada tipo de rotas otimizadas.
+A tabela mostra o potencial ciclável `r stringr::str_to_lower(proposicao)` `r mun` em número de novas viagens em bicicleta potenciais diárias, para cada meta da Estratégia Nacional ENMAC (4% e 10%), para cada um dos cenários, e para cada tipo de rotas otimizadas, para cada cenário, e para cada tipo de rotas otimizadas.
 

### 2023-03-02 code/biclarwww/aml_index.Rmd
home button in footer

@@
     includes:
-      after_body: img/footer.html
+      after_body: img/footer_button.html
 ---
@@
 
-# `r mun`
+# Área metropolitana de Lisboa
 

### 2023-03-02 code/biclarwww/index.Rmd
home button in footer

@@
 **O que mostra este mapa? **  
-Este mapa mostra a percentagem de viagens que são realizadas em bicicleta com origem em cada município .
+Este mapa mostra a percentagem de viagens que são realizadas em bicicleta com origem em cada município.
 
@@
 
+> Nota: Os valores de população são os constantes nos resultados _provisórios_ dos Censos 2021. A diferença para os resultados definitivos é de menos de 0.02% - cerca de 500 indivíduos.
+
 <!-- ------------------------------------------------------------------------ -->
@@
 
-Pretende-se que esta ferramenta informe os decisores, em comparação com o cenário atual de utilização de bicicleta, sobre:
+Pretende-se que esta ferramenta informe os decisores, em comparação com o cenário atual de utilização de bicicleta ([IMob 2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), sobre:
 

### 2023-03-03 code/biclarwww/index.Rmd
home map tweaks

@@
 
+REDEexistente = sf::st_read("data-raw/RedeExistente.gpkg", quiet = TRUE) %>% sf::st_transform(4326)
+REDEproposta = sf::st_read("data-raw/RedePrevista.gpkg", quiet = TRUE) %>% sf::st_transform(4326)
+
+greys2 =c("grey30", "grey70")
+greys2_factor = colorFactor(greys2, REDEexistente$Tipologia)
+
 m0 = leaflet::leaflet(MUNICIPIOSgeo_mapa) %>%
+ leaflet::addProviderTiles("OpenStreetMap.Mapnik", group = "OpenStreetMap") %>% 
+      leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
+      leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>%
+
+  leaflet::addPolylines(
+    data = REDEexistente,
+    group = "Rede Ciclável existente",
+    color = ~greys2_factor(Tipologia),
+    label = ~Municipio,
+    opacity = 1,
+    weight = 2
+    ) %>%
+
+  # leaflet::addPolylines(
+  #   data = REDEproposta,
+  #   color = "#E3879E",
+  #   dashArray = "3",
+  #   group = "Rede Ciclável planeada",
+  #   weight = 1.3,
+  #   opacity = 0.8
+  #   ) %>%
+
   leaflet::addPolygons(
@@
     labelOptions = leaflet::labelOptions(noHide = F, permanent = TRUE, direction = "auto",sticky = FALSE, textsize = "1.5em")) %>%
+
   leaflet::addLegend(
@@
     group = "Municípios"
+  ) %>%  
+  leaflet::addLegend(
+    "bottomright",
+    colors = greys2, 
+    labels = c("Dedicada", "Não-dedicada"),
+    group = "Rede Ciclável existente"
+  ) %>%
+  leaflet::addLegend(
+    "bottomright",
+    colors = "#E3879E", 
+    labels = "Planeada",
+    group = "Rede Ciclável planeada"
   ) %>%
@@
   ) %>%
-    leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>%
-    leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
-    leaflet::addProviderTiles("OpenStreetMap.Mapnik", group = "OpenStreetMap") %>% 
+
   leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satelite"),
-                            overlayGroups = c("Municípios", "Rede Ciclável existente"),
+                            overlayGroups = c("Municípios", "Rede Ciclável existente", "Rede Ciclável planeada"),
                             options = layersControlOptions(collapsed = FALSE)) %>% 
-  leaflet::hideGroup(c("Rede Ciclável existente"))
+  leaflet::hideGroup(c("Rede Ciclável existente", "Rede Ciclável planeada")) 
   # htmlwidgets::onRender("
@@
 **O que mostra este mapa? **  
-Este mapa mostra a percentagem de viagens que são realizadas em bicicleta com origem em cada município.
+Este mapa mostra a percentagem de viagens que são realizadas em bicicleta com origem em cada município, em escala de cor.
+Seleccionado outras camadas, é possível visualizar a rede ciclável existente e planeada, à data de 2022.
 
@@
 
-### Cenário Base
+### **Cenário Base**
 
@@
 
-### Cenário 1 - Viagens até 5 km
+### **Cenário 1** - Viagens até 5 km
 
@@
 
-### Cenário 2 - Viagens até 10 km bicicleta elétrica
+### **Cenário 2** - Viagens até 10 km bicicleta elétrica
 
@@
 
-### Cenário 3 - Viagens até 5 km de ou até uma interface de transportes públicos
+### **Cenário 3** - Viagens até 5 km de ou até uma interface de transportes públicos
 
@@
 
+> Forma sugerida de citação:
+<!-- Félix, R., Lovelace, R., Moura, F., Pinheiro, S., Marcelino, C., Garcia, C., Bandarrinha, C., Nogueira, L., Machado, P.
+(2022).
+biclaR: Ferramenta de apoio ao planeamento da rede ciclável na área metropolitana de Lisboa.
+CERIS - Instituto Superior Técnico e Transportes Metropolitanos de Lisboa.
+Lisboa, Dezembro.
+-->
+
 ### Agradecimentos
@@
 
-Para qualquer esclarecimento adicional, por favor enviar um email para info @tmlmobilidade.pt com o assunto _biclaR_.
+Para qualquer esclarecimento adicional, por favor enviar um email para *info* *@tmlmobilidade.pt* com o assunto _biclaR_.
 

### 2023-03-03 code/biclarwww/index.Rmd
layers map + scenarios description

@@
 
-```{r eval=FALSE, include=FALSE}
-tippy::tippy_this(elementId = "word", tooltip = "At low zoom, the map shows the % cycle uptake per county; at high zoom, it displays a cycle friendliness measure, with a score between 0 (very unfriendly) and 100 (very friendly).
-Both measures are shown for each given scenario.
-See the FAQs for more details")
-palette = c(
-  '#d73027',
-  '#f46d43',
-  '#fdae61',
-  '#fee090',
-  '#ffffbf',
-  '#e0f3f8',
-  '#abd9e9',
-  '#74add1',
-  '#4575b4'
-)
-breaks = c(0, 1, 2, 3, 5, 10, 20, 30, 40, Inf)
-pal_baseline = leaflet::colorBin(palette, bins = breaks,
-                                 domain = counties_index$pcycle_baseline)
-pal_near = leaflet::colorBin(palette, bins = breaks,
-                             domain = counties_index$pcycle_near)
-pal_climate = leaflet::colorBin(palette, bins = breaks,
-                                domain = counties_index$pcycle_climate)
-pal_dutch = leaflet::colorBin(palette, bins = breaks,
-                              domain = counties_index$pcycle_godutch)
-pal_ebike = leaflet::colorBin(palette, bins = breaks,
-                              domain = counties_index$pcycle_ebike)
-schools_raw = readRDS("schools_all.Rds")
-# main map
-leaflet::leaflet(counties_index) %>%
-  leaflet::addProviderTiles(
-    providers$CartoDB.Positron,
-    group = "Grey basemap",
-    options = leaflet::providerTileOptions(minZoom = 5, maxZoom = 22)
-  ) %>%
-  leaflet::addPolygons(
-    data = counties_index %>% sf::st_transform(crs = 4326),
-    color = "black",
-    fillColor = "transparent",
-    group = "Borders",
-    popup = paste0("View county results: ", counties_index$url),
-    opacity = 0.7,
-    weight = 1.2,
-    label = ~ Name,
-    labelOptions = leaflet::labelOptions(noHide = F, permanent = TRUE, direction = "auto",sticky = FALSE, textsize = "1.5em")) %>%
-  leaflet::addCircles(
-    data = schools_raw,
-    group = "Schools",
-    color = "black",
-    opacity = 0.8,
-    fillOpacity = 0.3
-  ) %>%
-  #polygons
-  mapboxapi::addMapboxTiles(
-    style_id = "cl77hy22i000m14pa2r440vgi",
-    username = "nathanaelisamapper",
-    group = "Baseline scenario",
-    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>%
-  mapboxapi::addMapboxTiles(
-    style_id = "cl77h0i4v001u15o1cp39sk37",
-    username = "nathanaelisamapper",
-    group = "Near Market",
-    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>%
-  mapboxapi::addMapboxTiles(
-    style_id = "cl77hqtom000214qrk6ceqeur",
-    username = "nathanaelisamapper",
-    group = "Climate Action Plan",
-    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>%
-  mapboxapi::addMapboxTiles(
-    style_id = "cl77hvhpx000d14nz0uhmsp40",
-    username = "nathanaelisamapper",
-    group = "Ebike",
-    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>%
-  mapboxapi::addMapboxTiles(
-    style_id = "cl77i0bqh003i16n2cijgn6r1",
-    username = "nathanaelisamapper",
-    group = "Go Dutch",
-    options = leaflet::providerTileOptions(minZoom = 1, maxZoom = 9),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>%
-  #rnet
-  mapboxapi::addMapboxTiles(
-    style_id = "cl66fjj97000a15ru2hhss06g",
-    username = "nathanaelisamapper",
-    group = "Baseline scenario",
-    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>%
-  mapboxapi::addMapboxTiles(
-    style_id = "cl66f8ax6001i14ti630sk73z",
-    username = "nathanaelisamapper",
-    group = "Climate Action Plan",
-    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>%   mapboxapi::addMapboxTiles(
-    style_id = "cl66foarr001d14lmc88wyu42",
-    username = "nathanaelisamapper",
-    group = "Near Market",
-    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  )  %>% mapboxapi::addMapboxTiles(
-    style_id = "cl66flaic000114t385c8spxn",
-    username = "nathanaelisamapper",
-    group = "Ebike",
-    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>% mapboxapi::addMapboxTiles(
-    style_id = "cl66fmn9j001g14lkbone3hye",
-    username = "nathanaelisamapper",
-    group = "Go Dutch",
-    options = leaflet::providerTileOptions(minZoom = 9, maxZoom = 22),
-    access_token = Sys.getenv("MAPBOX_PUBLIC_TOKEN")
-  ) %>%
-  leaflet::setView(lng = -8.2,
-                   lat = 53.5,
-                   zoom = 7) %>%
-  leaflet::addLegend(
-    "bottomleft",
-    colors = c("#882255",  "#CC6677", "#44AA99", "#117733"),
-    labels = c("0 to 25", "25 to 50", "50 to 75", "75 to 100"),
-    title = "Cycle friendliness",
-    group = "Friend",
-    opacity = 0.7
-  ) %>%
-  leaflet::addLegend(
-    "bottomleft",
-    colors = c(
-      '#d73027',
-      '#f46d43',
-      '#fdae61',
-      '#fee090',
-      '#ffffbf',
-      '#e0f3f8',
-      '#abd9e9',
-      '#74add1',
-      '#4575b4'
-    ),
-    labels = c(
-      "0 to 1",
-      "1 to 2",
-      "2 to 3",
-      "3 to 5",
-      "5 to 10",
-      "10 to 20",
-      "20 to 30",
-      "30 to 40",
-      "40 to more"
-    ),
-    title = "Percent cycling",
-    group = "Base",
-    opacity = 0.7
-  ) %>%
-  leaflet::addLegend(
-    title = "Get Started:",
-    group = "Help",
-    position = "bottomright",
-    colors = c("white",
-               "white"),
-    labels = c(
-      "Zoom in to explore the network",
-      "Click on a county for detailed results"
-    )
-  ) %>%
-  addLayersControl(
-    baseGroups  = c(
-      "Baseline scenario",
-      "Near Market",
-      "Climate Action Plan",
-      "Go Dutch",
-      "Ebike"
-    ),
-    overlayGroups = c("Schools"),
-    options = leaflet::layersControlOptions(collapsed = FALSE)
-  ) %>%
-  leaflet::groupOptions(group = "Help", zoomLevels = 7:8)  %>%
-    leaflet::groupOptions(group = "Base", zoomLevels = 1:9) %>%
-  leaflet::groupOptions(group = "Friend", zoomLevels = 10:20) %>%
-  leaflet::hideGroup("Schools")  %>%
-  htmlwidgets::onRender("
-        function() {
-            $('.leaflet-control-layers-base').prepend('<label style=\"text-align:center;font-size:larger;\"><b>Scenarios</b></label>');
-        }
-    ")
-```
-
 ```{r mapahome}
@@
 
-REDEexistente = sf::st_read("data-raw/RedeExistente.gpkg", quiet = TRUE) %>% sf::st_transform(4326)
-REDEproposta = sf::st_read("data-raw/RedePrevista.gpkg", quiet = TRUE) %>% sf::st_transform(4326)
+REDEexistente = sf::st_read("https://ushift.tecnico.ulisboa.pt/content/tml/RedeExistente.gpkg", quiet = TRUE) %>%
+  sf::st_transform(4326) %>%
+  arrange(., desc(Tipologia)) #para as nao dedicadas ficarem em baixo nas intersecções
+REDEproposta = sf::st_read("https://ushift.tecnico.ulisboa.pt/content/tml/RedePrevista.gpkg", quiet = TRUE) %>%
+  sf::st_transform(4326) %>%
+  filter(!ID %in% c("1105", "2104")) %>% #troços previstos em Mafra com overlap de existentes (?)
+  group_by(Municipio, Estado) %>%
+  summarise()
 
@@
    leaflet::addPolylines(
+    data = REDEproposta,
+    color = "#E3879E",
+    dashArray = "3",
+    group = "Rede Ciclável planeada",
+    weight = 1.3,
+    opacity = 0.8,
+    label = ~Municipio,
+    labelOptions = leaflet::labelOptions(textsize = "1.2em")
+    ) %>% # Está a impedir que tudo o resto seja mostrado. Talvez tenha de juntar ambas as layers e definir simbologia só numa
+
+  leaflet::addPolylines(
     data = REDEexistente,
@@
     label = ~Municipio,
+    labelOptions = leaflet::labelOptions(textsize = "1.2em"),
     opacity = 1,
@@
     ) %>%
-
-  # leaflet::addPolylines(
-  #   data = REDEproposta,
-  #   color = "#E3879E",
-  #   dashArray = "3",
-  #   group = "Rede Ciclável planeada",
-  #   weight = 1.3,
-  #   opacity = 0.8
-  #   ) %>%
 
@@
     "bottomright",
-    colors = greys2, 
+    pal = greys2_factor, 
+    values = ~REDEexistente$Tipologia,
     labels = c("Dedicada", "Não-dedicada"),
+    # title = "Rede Ciclável existente",
+    title = "",
     group = "Rede Ciclável existente"
@@
 
-Em que consistem os cenários utilizados nesta modelação?
+Em que consistem os [cenários](#cenarios2), [metas](#metas), e [tipos de percurso](#percurso) utilizados nesta modelação?
 
-### **Cenário Base**
+### Cenários {#cenarios2}
 
-Na área metropolitana de Lisboa são realizadas diariamente cerca de 5.3 milhões de viagens, em vários modos.
-O número de viagens em bicicleta, segundo o Inquérito à Mobilidade realizado em 2017 ([IMob 2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}), é de cerca de 25 mil viagens/dia.
+#### **Cenário Base**
+
+As características das viagens Inquérito à Mobilidade realizado em 2017/2018 ([IMob 2018](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt){target="_blank"}) constituem o cenário base deste projeto.
+Apesar de ter sido realizado em período pré-pandemia, este conjunto de dados é a melhor e mais recente informação que temos em termos de mobilidade urbana nas áreas metropolitanas.
+
+Na área metropolitana de Lisboa são realizadas diariamente cerca de 5.3 milhões de viagens, em vários modos. O número de viagens em bicicleta é de cerca de 25 mil viagens/dia.
 
@@
 
-### **Cenário 1** - Viagens até 5 km
+#### **Cenário 1** - Viagens até 5 km
 
@@
 
-Este é um **cenário conservador / realista**.
+Considera-se este um **cenário conservador**, por ter em conta apenas viagens inferiores a 5 km (“em cima de uma bicicleta”).  
+Para uma primeira abordagem de captação de transferência modal, este cenário é o mais realista.
 
-### **Cenário 2** - Viagens até 10 km bicicleta elétrica
+#### **Cenário 2** - Viagens até 10 km bicicleta elétrica
 
@@
 
-### **Cenário 3** - Viagens até 5 km de ou até uma interface de transportes públicos
+#### **Cenário 3** - Viagens até 5 km de ou até uma interface de transportes públicos
 
 Este cenário inclui todas as viagens que podem ser realizadas em transportes públicos, cuja primeira e última _milha_ da viagem podem ser feitas em bicicleta, entre freguesias até uma extensão total de 5 km - por exemplo: 1 km de bicicleta até ao interface A, e 4 km de bicicleta entre o interface B e o destino.
-Como transportes públicos, são considerados o comboio, barco, metro sul do Tejo, e algumas carreiras da Carris Metropolitana - em que é possível transportar uma bicicleta de forma prática.. 
+Como transportes públicos, são considerados o comboio, barco, metro sul do Tejo, e algumas carreiras da Carris Metropolitana - em que é possível transportar uma bicicleta de forma prática. 
 De uma forma conservadora, assumimos também que não são possíveis transferências entre transportes ou entre linhas/carreiras.
@@
 Este valoriza a utilização da bicicleta como **complemento ao transporte público**.
+
+>Estes cenários partem do pressuposto que o número de viagens realizadas se mantém constante ao longo dos anos, fazendo-se apenas variar o seu modo de transporte.
+
+### Metas {#metas}
+
+As metas são as estabelecidas pela Estratégia Nacional de Mobilidade Ativa Ciclável 2020-2030 ([ENMAC](https://files.dre.pt/1s/2019/08/14700/0004600081.pdf){target="_blank"}).
+
+#### 4%
+
+Em **2025**, 4% das viagens em meio urbano deverão ser realizadas em bicicleta, resultando exclusivamente da redução da utilização do transporte automóvel privado.
+
+#### 10 %
+
+Em **2030**, 10% das viagens em meio urbano deverão ser realizadas em bicicleta, resultando exclusivamente da redução da utilização do transporte automóvel privado.
+
+### Tipos de percurso {#percurso}
+
+A identificação dos percursos entre as origens e destinos (para as linhas de desejo, ligando Freguesias da AML) são um aspeto fundamental da modelação da rede ciclável nesta metodologia, uma vez que o IMob não identifica os percursos realizados por cada viagem, sendo necessário estimá-los.
+
+A modelação dos percursos cicláveis depende das variáveis consideradas, assim como das restrições pré-definidas.
+Estas restrições podem privilegiar aspetos como velocidade e volumes baixos, rotas mais diretas, percursos menos declivosos, entre outros, adequados à utilização da bicicleta, sendo que o algoritmo segue uma avaliação que resulta da ponderação das variáveis consideradas.
+Mesmo o percurso menos declivoso, mas com maior volume de tráfego, poderá ser o indicado se a função assim o definir.
+
+O algoritmo de escolha de percursos utilizado foi o [r5r](https://ipeagit.github.io/r5r/reference/detailed_itineraries.html){target="_blank"}, que é muito flexível nas configurações de tipo de percursos estimados, nomeadamente por permitir estimar com probabilidade e incerteza os percursos que usam transportes públicos (considerando os seus horários).
+
+O r5r facilita a identificação de percursos mais diretos ou mais seguros, recorrendo para isso ao nível de stress de tráfego (_Level of Traffic Stress_, ou [LTS](https://docs.conveyal.com/learn-more/traffic-stress){target="_blank"}).
+Este varia de 1 a 4, sendo o 1 mais tranquilo - correspondendo, por exemplo, a pistas cicláveis fora de estrada, e o 4 menos tranquilo - correspondendo, por exemplo, a percursos partilhados com o tráfego motorizado.
+
+#### Direto
+
+Os percursos diretos foram estimados usando como nível de stress máximo permitido o LTS 4.
+
+#### Seguro
+
+Os percursos diretos foram estimados usando como nível de stress máximo permitido o LTS 3.  
+Por vezes não é possível encontrar um caminho de bicicleta com estas características (ente a origem e o destino), e daí o número de viagens estimadas poder ser menor para este tipo de percursos.
+
+O nível de **tranquilidade** (ou _quietness_) é outro indicador, e é estimado para cada troço com base no número de vias, velocidade máxima permitida, arborização, hierarquia e tráfego (quando a informação correspondente existe nas etiquetas do OpenStreetMap).
+Este indicador, desenvolvido pelo [Cyclestreets](https://www.cyclestreets.net/help/journey/howitworks/#quietness){target="_blank"}, é apresentado numa escala de 0 a 100, em que o zero corresponde ao nível menos seguro para se circular de bicicleta, e o 100 ao nível mais seguro e tranquilo - normalmente correspondendo a troços que incluem já infraestrutura ciclável segregada do tráfego motorizado.
 
@@
                 colnames = c("Cenário", "Total Viagens /dia",
-                             "Viagens Bicicleta 2018", "Viagens Bicicleta 4%",
+                             "Viagens Bicicleta Base", "Viagens Bicicleta 4%",
                               "Viagens Bicicleta 10%"),
@@
     tr(
-      lapply(rep(c("2018", "Meta 4%", "Meta 10%"), 3), th)
+      lapply(rep(c("Base", "Meta 4%", "Meta 10%"), 3), th)
     )
@@
 
-> Nota: Os valores de população são os constantes nos resultados _provisórios_ dos Censos 2021. A diferença para os resultados definitivos é de menos de 0.02% - cerca de 500 indivíduos.
+> Nota 1: Os valores de população são os constantes nos resultados _provisórios_ dos Censos 2021. A diferença para os resultados definitivos é de menos de 0.02% - cerca de 500 indivíduos.
+
+> Nota 2: Para alguns municípios não é possível estimar viagens intermodais com recurso à bicicleta, para os critérios mínumos definidos (valores _em branco_).
 
@@
 
-> Forma sugerida de citação:
-<!-- Félix, R., Lovelace, R., Moura, F., Pinheiro, S., Marcelino, C., Garcia, C., Bandarrinha, C., Nogueira, L., Machado, P.
+Forma sugerida de citação:
+
+>Félix, R., Lovelace, R., Moura, F.
 (2022).
-biclaR: Ferramenta de apoio ao planeamento da rede ciclável na área metropolitana de Lisboa.
+**biclaR: Ferramenta de apoio ao planeamento da rede ciclável na área metropolitana de Lisboa**.
 CERIS - Instituto Superior Técnico e Transportes Metropolitanos de Lisboa.
+[biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt).
 Lisboa, Dezembro.
--->
 

### 2023-03-03 code/biclarwww/aml_index.Rmd
replace 2018 with Base

@@
                 colnames = c("Meta ENMAC", "Cenário", "Tipo de Percurso", "Total Viagens /dia",
-                             "Viagens Bicicleta 2018", "Viagens Bicicleta Potencial",
+                             "Viagens Bicicleta Base", "Viagens Bicicleta Potencial",
                              "CO2eq evitado (ton/ano)", "Benefícios estimados em 10 anos (milhares €)"),

### 2023-03-04 code/biclarwww/index.Rmd
add why scenario 3 not show

@@
 Perguntas e respostas simplificadas.  
+Ver o [relatório metodológico](nada) para mais detalhes sobre os métodos e funcionamento desta ferramenta.
 
@@
 Poderá avaliar igualmente qual o potencial de transferência que oferece, se se trata de uma ligação mais segura ou mais direta, e que impactes potenciais se pode esperar, designadamente em termos de impactes sociais e ambientais.
+
+### Porque é que nalguns municípios não surge uma rede potencial para o cenário 3?
+
+Apesar de ter sido estimada rede ciclável potencial no cenário 3 - intermodalidade até 5km, em alguns municípios, o processo de limpeza da rede resulta em que não surja uma rede (ou segmentos) com um mínimo potencial de transferência modal nos seus territórios.
+São eles os municípios de Alcochete, Mafra, Montijo, Odivelas e Sesimbra.
+
+Isto deve-se a:
+
+*   Possibilidade de não existir interface de transportes públicos considerados nesse território;
+*   Possibilidade de as interfaces consideradas não servirem as origens-destino do IMob;
+*   Possibilidade de as etapas em bicicleta das viagens intermodais estimadas terem mais de 5km de extensão na sua totalidade; e/ou
+*   Possibilidade de o potencial de rede ciclável estimado não ter um mínimo de 10 viagens em cada segmento.
+
+Em todo o caso, é possível estimar cenários menos conservadores de intermodalidade (por exemplo com viagens até 7.5 km), em que essas redes cicláveis potenciais tenham um mínimo razoável de X viagens por segmento.
 

### 2023-03-04 code/biclarwww/municipio_minor_index.Rmd
add why scenario 3 not show

@@
 <!-- Ver mapa em [ecrã inteiro](mapa_cenario3.html){target="_blank"}. -->
+<details>
+  <summary>**Porque é que nalguns municípios não surge uma rede potencial para o cenário 3?**</summary>
+
+Apesar de ter sido estimada rede ciclável potencial no cenário 3 - intermodalidade até 5km, em alguns municípios, o processo de limpeza da rede resulta em que não surja uma rede (ou segmentos) com um mínimo potencial de transferência modal nos seus territórios.
+São eles os municípios de Alcochete, Mafra, Montijo, Odivelas e Sesimbra.
+
+Isto deve-se a:
+
+*   Possibilidade de não existir interface de transportes públicos considerados nesse território;
+*   Possibilidade de as interfaces consideradas não servirem as origens-destino do IMob;
+*   Possibilidade de as etapas em bicicleta das viagens intermodais estimadas terem mais de 5km de extensão na sua totalidade; e/ou
+*   Possibilidade de o potencial de rede ciclável estimado não ter um mínimo de 10 viagens em cada segmento.
+</details>
 

### 2023-03-04 code/biclarwww/build.R
Update build.R

@@ -16,7 +16,7 @@ rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", encoding = "UTF-8", ou
 # Create outputs for each region --------------------------------------------------------------
 
 # i = "oeiras"
-i = "barreiro"
+# i = "barreiro"
 # i = "lisboa"
 # i = "odivelas"
 # i = "setubal"

### 2023-03-04 code/biclarwww/index.Rmd
sesimbra not outlier

@@
 Apesar de ter sido estimada rede ciclável potencial no cenário 3 - intermodalidade até 5km, em alguns municípios, o processo de limpeza da rede resulta em que não surja uma rede (ou segmentos) com um mínimo potencial de transferência modal nos seus territórios.
-São eles os municípios de Alcochete, Mafra, Montijo, Odivelas e Sesimbra.
+São eles os municípios de Alcochete, Mafra, Montijo e Odivelas.
 
@@
 *   Possibilidade de as interfaces consideradas não servirem as origens-destino do IMob;
+*   Possibilidade de não ser possível estimar um percurso em bicicleta até à interface;
 *   Possibilidade de as etapas em bicicleta das viagens intermodais estimadas terem mais de 5km de extensão na sua totalidade; e/ou
 *   Possibilidade de o potencial de rede ciclável estimado não ter um mínimo de 10 viagens em cada segmento.
+
+Nalguns casos, é no entanto possível ver a rede com critérios abaixo do mínimo de viagens por segmento definidos para efeito de visualização, fazendo [download do SIG para a AML](aml/index.html#downloads) e visualizando toda a rede estimada para o cenário 3.
 

### 2023-03-04 code/biclarwww/municipio_minor_index.Rmd
sesimbra not outlier

--- 2023-03-03 code/biclarwww/municipio_minor_index.Rmd
@@
 <!-- Ver mapa em [ecrã inteiro](mapa_cenario3.html){target="_blank"}. -->
+<details>
+  <summary>**Porque é que nalguns municípios não surge uma rede potencial para o cenário 3?**</summary>
+
+Apesar de ter sido estimada rede ciclável potencial no cenário 3 - intermodalidade até 5km, em alguns municípios, o processo de limpeza da rede resulta em que não surja uma rede (ou segmentos) com um mínimo potencial de transferência modal nos seus territórios.
+Isto deve-se a:
+
+*   Possibilidade de não existir interface de transportes públicos considerados nesse território;
+*   Possibilidade de as interfaces consideradas não servirem as origens-destino do IMob;
+*   Possibilidade de não ser possível estimar um percurso em bicicleta até à interface;
+*   Possibilidade de as etapas em bicicleta das viagens intermodais estimadas terem mais de 5km de extensão na sua totalidade; e/ou
+*   Possibilidade de o potencial de rede ciclável estimado não ter um mínimo de 10 viagens em cada segmento.
+</details>
 

### 2023-03-04 code/biclarwww/build.R

@@ -20,6 +20,7 @@ rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", encoding = "UTF-8", ou
 # i = "lisboa"
 # i = "odivelas"
 # i = "setubal"
+#sesimbra é pequeno ou não?
 
 # for (i in municipios){
 

### 2023-03-04 code/biclarwww/index.Rmd
show plans in larger map

@@
 
-REDEexistente = sf::st_read("https://ushift.tecnico.ulisboa.pt/content/tml/RedeExistente.gpkg", quiet = TRUE) %>%
+# REDEexistente = sf::st_read("https://ushift.tecnico.ulisboa.pt/content/tml/RedeExistente.gpkg") %>% sf::st_transform(4326)
+# REDEproposta = sf::st_read("https://ushift.tecnico.ulisboa.pt/content/tml/RedePrevista.gpkg") %>% sf::st_transform(4326)
+
+REDEexistente = sf::st_read("data-raw/RedeExistente.gpkg", quiet = TRUE) %>%
   sf::st_transform(4326) %>%
   arrange(., desc(Tipologia)) #para as nao dedicadas ficarem em baixo nas intersecções
-REDEproposta = sf::st_read("https://ushift.tecnico.ulisboa.pt/content/tml/RedePrevista.gpkg", quiet = TRUE) %>%
+REDEproposta = sf::st_read("data-raw/RedePrevista.gpkg", quiet = TRUE) %>%
   sf::st_transform(4326) %>%

### 2023-03-18 code/biclarwww/index.Rmd
typo

@@
 
-> Nota 2: Para alguns municípios não é possível estimar viagens intermodais com recurso à bicicleta, para os critérios mínumos definidos (valores _em branco_).
+> Nota 2: Para alguns municípios não é possível estimar viagens intermodais com recurso à bicicleta, para os critérios mínimos definidos (valores _em branco_).
 

### 2023-03-18 code/biclarwww/index.Rmd
basemap Cinza + FAQ

@@
  leaflet::addProviderTiles("OpenStreetMap.Mapnik", group = "OpenStreetMap") %>% 
-      leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
+      leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
       leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>%
@@
 
-  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satelite"),
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
                             overlayGroups = c("Municípios", "Rede Ciclável existente", "Rede Ciclável planeada"),
@@
 
+### Como comparar as redes resultantes com a rede planeada?
+
+É possível fazê-lo de várias formas.
+Uma delas é abrindo o mapa em ecrã inteiro (existe um link em baixo de cada mapa que o permite fazer), e lá irá encontrar mais duas camadas de informação: a rede planeada, e a rede existente dedicada - i.e., ciclovias e pistas cicláveis existentes.
+
+Outra forma é utilizando um SIG, e descarregando os mapas dos resultados e da rede planeada e existente, que se encontra na secção de [Downloads na página da AML](aml/index.html#downloads){target="_blank"}.
+
+Atenção que, dependendo do cenário que estamos a analisar, a rede potencial pode cobrir maior ou menor parte da rede planeada por cada Município, sendo que valerá a pena confrontar os resultados dos vários cenários.
+
 ### Posso descarregar os mapas em formato tipo _shapefile_?

### 2023-03-18 code/biclarwww/municipio_index.Rmd
existing and planned infrastructure in full creen maps
as layers

@@
 m1leaf = tmap::tmap_leaflet(m1) %>% 
-  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
-  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
+  leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
                             overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
@@
 
-Ver mapa em [ecrã inteiro](mapa_cenario1.html){target="_blank"}.
+```{r mapa1more}
+# load redes e corte nos limites do municipio
+REDEproposta_mun = sf::st_read("data-raw/RedePrevista.gpkg", quiet = TRUE) %>%
+  sf::st_transform(4326) %>%
+  filter(!ID %in% c("1105", "2104")) %>% #troços previstos em Mafra com overlap de existentes (?)
+  filter(Municipio == mun) %>% 
+  group_by(Estado) %>%
+  summarise()
+
+REDEexistente_mun = sf::st_read("data-raw/RedeExistente.gpkg", quiet = TRUE) %>%
+  sf::st_transform(4326) %>%
+  filter(Municipio == mun) %>% 
+  filter(Tipologia == "Dedicada") %>% #não interessa mostrar a não dedicada
+  group_by(Funcao) %>% 
+  summarise()
+
+m1leaf_more = tmap::tmap_leaflet(m1) %>% 
+  leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
+  leaflet::addPolylines(
+    data = REDEproposta_mun,
+    color = "#E3879E", #this is not the best color fot this purpose!
+    dashArray = "3",
+    group = "Rede Ciclável planeada",
+    weight = 1.3,
+    opacity = 0.8
+    ) %>% # Está a impedir que tudo o resto seja mostrado. Talvez tenha de juntar ambas as layers e definir simbologia só numa
+
+  leaflet::addPolylines(
+    data = REDEexistente_mun,
+    # color = "grey30",
+    # color = "#4d4d4d",
+    color = "#1A7832",
+    group = "Rede Ciclável existente",
+    opacity = 1,
+    weight = 2
+  ) %>%
+
+  leaflet::addLegend(
+    "bottomright",
+    colors = "#E3879E", 
+    labels = "Planeada",
+    group = "Rede Ciclável planeada"
+  ) %>%
+    leaflet::addLegend(
+    "bottomright",
+    colors = "#4d4d4d", 
+    labels = "Existente dedicada",
+    group = "Rede Ciclável existente"
+  ) %>%
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
+                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
+                                              "Potencial 4% - seguro", "Potencial 10% - seguro",
+                                              "Rede Ciclável planeada", "Rede Ciclável existente"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro",
+                       "Rede Ciclável planeada", "Rede Ciclável existente"))
+
+```
+
+Vendo o mapa em [ecrã inteiro](mapa_cenario1.html){target="_blank"} é possível comparar os resultados com a rede existente e planeada.
 
@@
 m2leaf = tmap::tmap_leaflet(m2) %>% 
-  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
-  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
+  leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
                             overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
@@
 m2leaf
+
+```
+```{r mapa2more}
+# com rede planeada e existente em camadas extra no modo ecran inteiro
+m2leaf_more = tmap::tmap_leaflet(m2) %>% 
+  leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
+  leaflet::addPolylines(
+    data = REDEproposta_mun,
+    color = "#E3879E", #this is not the best color fot this purpose!
+    dashArray = "3",
+    group = "Rede Ciclável planeada",
+    weight = 1.3,
+    opacity = 0.8
+    ) %>% # Está a impedir que tudo o resto seja mostrado. Talvez tenha de juntar ambas as layers e definir simbologia só numa
+
+  leaflet::addPolylines(
+    data = REDEexistente_mun,
+    # color = "grey30",
+    # color = "#4d4d4d",
+    color = "#1A7832",
+    group = "Rede Ciclável existente",
+    opacity = 1,
+    weight = 2
+  ) %>%
+
+  leaflet::addLegend(
+    "bottomright",
+    colors = "#E3879E", 
+    labels = "Planeada",
+    group = "Rede Ciclável planeada"
+  ) %>%
+    leaflet::addLegend(
+    "bottomright",
+    colors = "#4d4d4d", 
+    labels = "Existente dedicada",
+    group = "Rede Ciclável existente"
+  ) %>%
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
+                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
+                                              "Potencial 4% - seguro", "Potencial 10% - seguro",
+                                              "Rede Ciclável planeada", "Rede Ciclável existente"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro",
+                       "Rede Ciclável planeada", "Rede Ciclável existente"))
 
@@
 m3leaf = tmap::tmap_leaflet(m3) %>% 
-  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satelite") %>% 
-  leaflet::addLayersControl(baseGroups = c("Esri.WorldGrayCanvas", "OpenStreetMap", "Satelite"),
+  leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
                             overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
@@
 m3leaf
+
+```
+```{r mapa3more}
+# com rede planeada e existente em camadas extra no modo ecran inteiro
+m3leaf_more = tmap::tmap_leaflet(m3) %>% 
+  leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
+  leaflet::addPolylines(
+    data = REDEproposta_mun,
+    color = "#E3879E", #this is not the best color fot this purpose!
+    dashArray = "3",
+    group = "Rede Ciclável planeada",
+    weight = 1.3,
+    opacity = 0.8
+    ) %>% # Está a impedir que tudo o resto seja mostrado. Talvez tenha de juntar ambas as layers e definir simbologia só numa
+
+  leaflet::addPolylines(
+    data = REDEexistente_mun,
+    # color = "grey30",
+    # color = "#4d4d4d",
+    color = "#1A7832",
+    group = "Rede Ciclável existente",
+    opacity = 1,
+    weight = 2
+  ) %>%
+
+  leaflet::addLegend(
+    "bottomright",
+    colors = "#E3879E", 
+    labels = "Planeada",
+    group = "Rede Ciclável planeada"
+  ) %>%
+    leaflet::addLegend(
+    "bottomright",
+    colors = "#4d4d4d", 
+    labels = "Existente dedicada",
+    group = "Rede Ciclável existente"
+  ) %>%
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
+                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
+                                              "Potencial 4% - seguro", "Potencial 10% - seguro",
+                                              "Rede Ciclável planeada", "Rede Ciclável existente"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro",
+                       "Rede Ciclável planeada", "Rede Ciclável existente"))
 
@@
 ```{r exportmaps}
-htmlwidgets::saveWidget(m1leaf, file = paste0("biclarwww/", i,"/mapa_cenario1.html"))
+htmlwidgets::saveWidget(m1leaf_more, file = paste0("biclarwww/", i,"/mapa_cenario1.html"))
 
-htmlwidgets::saveWidget(m2leaf, file = paste0("biclarwww/", i,"/mapa_cenario2.html"))
+htmlwidgets::saveWidget(m2leaf_more, file = paste0("biclarwww/", i,"/mapa_cenario2.html"))
 
-htmlwidgets::saveWidget(m3leaf, file = paste0("biclarwww/", i,"/mapa_cenario3.html"))
+htmlwidgets::saveWidget(m3leaf_more, file = paste0("biclarwww/", i,"/mapa_cenario3.html"))
 ```

### 2023-03-18 code/biclarwww/municipio_minor_index.Rmd
existing and planned infrastructure in full creen maps
as layers

@@
 m1leaf
+
+```
+
+```{r mapa1more}
+# load redes e corte nos limites do municipio
+REDEproposta_mun = sf::st_read("data-raw/RedePrevista.gpkg", quiet = TRUE) %>%
+  sf::st_transform(4326) %>%
+  filter(!ID %in% c("1105", "2104")) %>% #troços previstos em Mafra com overlap de existentes (?)
+  filter(Municipio == mun) %>% 
+  group_by(Estado) %>%
+  summarise()
+
+REDEexistente_mun = sf::st_read("data-raw/RedeExistente.gpkg", quiet = TRUE) %>%
+  sf::st_transform(4326) %>%
+  filter(Municipio == mun) %>% 
+  filter(Tipologia == "Dedicada") %>% #não interessa mostrar a não dedicada
+  group_by(Funcao) %>% 
+  summarise()
+
+m1leaf_more = tmap::tmap_leaflet(m1) %>% 
+  leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
+  leaflet::addPolylines(
+    data = REDEproposta_mun,
+    color = "#E3879E", #this is not the best color fot this purpose!
+    dashArray = "3",
+    group = "Rede Ciclável planeada",
+    weight = 1.3,
+    opacity = 0.8
+    ) %>% # Está a impedir que tudo o resto seja mostrado. Talvez tenha de juntar ambas as layers e definir simbologia só numa
+
+  leaflet::addPolylines(
+    data = REDEexistente_mun,
+    # color = "grey30",
+    # color = "#4d4d4d",
+    color = "#1A7832",
+    group = "Rede Ciclável existente",
+    opacity = 1,
+    weight = 2
+  ) %>%
+
+  leaflet::addLegend(
+    "bottomright",
+    colors = "#E3879E", 
+    labels = "Planeada",
+    group = "Rede Ciclável planeada"
+  ) %>%
+    leaflet::addLegend(
+    "bottomright",
+    colors = "#4d4d4d", 
+    labels = "Existente dedicada",
+    group = "Rede Ciclável existente"
+  ) %>%
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
+                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
+                                              "Potencial 4% - seguro", "Potencial 10% - seguro",
+                                              "Rede Ciclável planeada", "Rede Ciclável existente"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro",
+                       "Rede Ciclável planeada", "Rede Ciclável existente"))
 
@@
 m3leaf
+
+```
+```{r mapa3more, eval=dontrun}
+# com rede planeada e existente em camadas extra no modo ecran inteiro
+m3leaf_more = tmap::tmap_leaflet(m3) %>% 
+  leaflet::addProviderTiles("Esri.WorldGrayCanvas", group = "Basemap cinza") %>% 
+  leaflet::addProviderTiles("Esri.WorldImagery", group = "Satélite") %>% 
+  leaflet::addPolylines(
+    data = REDEproposta_mun,
+    color = "#E3879E", #this is not the best color fot this purpose!
+    dashArray = "3",
+    group = "Rede Ciclável planeada",
+    weight = 1.3,
+    opacity = 0.8
+    ) %>% # Está a impedir que tudo o resto seja mostrado. Talvez tenha de juntar ambas as layers e definir simbologia só numa
+
+  leaflet::addPolylines(
+    data = REDEexistente_mun,
+    # color = "grey30",
+    # color = "#4d4d4d",
+    color = "#1A7832",
+    group = "Rede Ciclável existente",
+    opacity = 1,
+    weight = 2
+  ) %>%
+
+  leaflet::addLegend(
+    "bottomright",
+    colors = "#E3879E", 
+    labels = "Planeada",
+    group = "Rede Ciclável planeada"
+  ) %>%
+    leaflet::addLegend(
+    "bottomright",
+    colors = "#4d4d4d", 
+    labels = "Existente dedicada",
+    group = "Rede Ciclável existente"
+  ) %>%
+  leaflet::addLayersControl(baseGroups = c("Basemap cinza", "OpenStreetMap", "Satélite"),
+                            overlayGroups = c("Potencial 4% - direto", "Potencial 10% - direto", 
+                                              "Potencial 4% - seguro", "Potencial 10% - seguro",
+                                              "Rede Ciclável planeada", "Rede Ciclável existente"),
+                            options = layersControlOptions(collapsed = FALSE)) %>% 
+  leaflet::hideGroup(c("Potencial 10% - direto", "Potencial 4% - seguro", "Potencial 10% - seguro",
+                       "Rede Ciclável planeada", "Rede Ciclável existente"))
 
@@
 ```{r exportmaps}
-htmlwidgets::saveWidget(m1leaf, file = paste0("biclarwww/", i,"/mapa_cenario1.html"))
+htmlwidgets::saveWidget(m1leaf_more, file = paste0("biclarwww/", i,"/mapa_cenario1.html"))
 
-htmlwidgets::saveWidget(m2leaf, file = paste0("biclarwww/", i,"/mapa_cenario2.html"))
+htmlwidgets::saveWidget(m2leaf_more, file = paste0("biclarwww/", i,"/mapa_cenario2.html"))
 

### 2023-03-18 code/biclarwww/index.Rmd
seguro is not safe but safer
warning

@@
 
-Os percursos diretos foram estimados usando como nível de stress máximo permitido o LTS 3.  
+Os percursos seguros foram estimados usando como nível de stress máximo permitido o LTS 3.  
 Por vezes não é possível encontrar um caminho de bicicleta com estas características (ente a origem e o destino), e daí o número de viagens estimadas poder ser menor para este tipo de percursos.
@@
 Este indicador, desenvolvido pelo [Cyclestreets](https://www.cyclestreets.net/help/journey/howitworks/#quietness){target="_blank"}, é apresentado numa escala de 0 a 100, em que o zero corresponde ao nível menos seguro para se circular de bicicleta, e o 100 ao nível mais seguro e tranquilo - normalmente correspondendo a troços que incluem já infraestrutura ciclável segregada do tráfego motorizado.
+
+Chama-se a atenção que o _percurso seguro_ não é necessariamente "seguro" para circular em bicicleta nas condições atuais.
+É o caminho **mais** seguro, recomendado pelo algoritmo de escolha de percursos.
 

### 2023-03-19 code/biclarwww/municipio_index.Rmd
colors of existing networks

@@
     "bottomright",
-    colors = "#4d4d4d", 
+    colors = "#1A7832", 
     labels = "Existente dedicada",

### 2023-03-19 code/biclarwww/municipio_minor_index.Rmd
colors of existing networks

@@
 
-Ver mapa em [ecrã inteiro](mapa_cenario1.html){target="_blank"}.
+Vendo o mapa em [ecrã inteiro](mapa_cenario1.html){target="_blank"} é possível comparar os resultados com a rede existente e planeada.
 

### 2023-03-19 code/biclarwww/build.R
build picky

@@ -20,9 +20,9 @@ rmarkdown::render(input = "code/biclarwww/aml_index.Rmd", encoding = "UTF-8", ou
 # i = "lisboa"
 # i = "odivelas"
 # i = "setubal"
-#sesimbra é pequeno ou não?
 
-# for (i in municipios){
+# for (i in c("alcochete", "mafra", "montijo", "odivelas")){ # run only miniors
+# for (i in municipios[6:18]){ #run only the last ones (after a crash)
 
 for (i in municipios){
   

### 2023-03-19 code/biclarwww/index.Rmd
alpha version

@@
 
-##### Nota: este projeto está em fase _beta_ e a informação deste site deve ser considerada como _work in progress_.
+<!-- ##### Nota: este projeto está em fase _beta_ e a informação deste site deve ser considerada como _work in progress_.  -->
+\br
 
@@
 Uma delas é abrindo o mapa em ecrã inteiro (existe um link em baixo de cada mapa que o permite fazer), e lá irá encontrar mais duas camadas de informação: a rede planeada, e a rede existente dedicada - i.e., ciclovias e pistas cicláveis existentes.
+Esta opção está disponível nas páginas individuais de cada Município.
 

### 2023-04-26 code/heat-prep.R
tidy up HEAT HaaS
#38
in line 86 it was `mutate(Car_new = Car + CarP - new_cyc)`
but I think it makes more sense as `mutate(Car_new = Car - new_cyc)`

@@ -4,7 +4,7 @@ library(sf)
 
 
 # locations id
-HEAT_all_locations_data = read_csv("HEAT/all_locations_data.csv") #data from https://github.com/HEAT-WHO/HEAT_data/blob/main/data/heatr/locations_data/all_locations_data.csv
+HEAT_all_locations_data = read.csv("HEAT/all_locations_data.csv") #data from https://github.com/HEAT-WHO/HEAT_data/blob/main/data/heatr/locations_data/all_locations_data.csv
 HEAT_all_locations_data = HEAT_all_locations_data %>% filter(country_location_name == "Portugal")
 HEAT_all_locations_data = HEAT_all_locations_data %>% filter(sub1_location_name %in% c("Distrito de Lisboa", "Distrito de Setúbal"))
 HEAT_all_locations_data = HEAT_all_locations_data %>% filter(is.na(city_location_id))
@@ -15,4 +15,470 @@ HEAT_all_locations_data = HEAT_all_locations_data %>% filter(is.na(city_location
 PM25 = openxlsx::read.xlsx("https://github.com/U-Shift/biclar/releases/download/0.0.1/Scenarios_Routing.xlsx", sheet = "PM25")
 PM25$PM25 = round(PM25$PM25, 2)
 PM25$location_id = as.character(PM25$location_id) # needed to use as input in HaaS
+
+# SCENARIOS combinations
+CENARIOS = openxlsx::read.xlsx("https://github.com/U-Shift/biclar/releases/download/0.0.1/Scenarios_Routing.xlsx", sheet = "CENARIOS")
+CENARIOS$dist_max[CENARIOS$Estrategia == "Convencional"] = 5000 
+CENARIOS$dist_max[CENARIOS$Estrategia == "Ebike"] = 10000 
+CENARIOS$dist_max[CENARIOS$Estrategia == "Intermodal"] = 5000 
+
+# File paths
+PATHS = openxlsx::read.xlsx("https://github.com/U-Shift/biclar/releases/download/0.0.1/Scenarios_Routing.xlsx", sheet = "paths")
+# save all files listed in "routes_filepath", "rnet_filepath", and "car_routes_filepath" in the respective folder.
+# download them from https://github.com/U-Shift/biclar/releases
+
+
+
+
+# For scenarios 1 and 2 (Conventional and Ebike) ---------------------------------------------------
+
+Code_Hass = CENARIOS %>%
+  filter(Estrategia != "Intermodal") %>% # make only for senários 1 & 2
[…]

### 2023-04-26 code/heat-prep.R
add export HEAT
#38

@@ -482,3 +482,15 @@ HEATbind_municipios_intermodal_PPP = HEATbind_municipios_intermodal
 
 
 
+
+# Join scenarios 1,2,3 ----------------------------------------------------
+
+HEAT_municipios_PPP_redux = bind_rows(HEATbind_municipios_PPP, HEATbind_municipios_intermodal_PPP) %>% 
+  select(-c(routes_filepath, rnet_filepath, car_routes_filepath)) %>% 
+  mutate(Estrategia = substr(CodePath, 1,1),
+         LTS = substr(CodePath, 2,2)) %>%
+  mutate(Meta_ENMAC = factor(ENMAC, levels = c(0.04, 0.10), labels = c("4%", "10%")),
+         Estrategia = factor(Estrategia, levels = c("1", "2", "3"), labels =c("ate 5 km", "ate 10km e-bike","intermodal ate 5km ")),
+  )
+
+saveRDS(HEAT_municipios_PPP_redux, "HEAT/HEAT_municipios_123_redux.Rds")

### 2023-04-26 code/biclarwww/index.Rmd
warning when comparing
link to methods report
#38

@@
 Perguntas e respostas simplificadas.  
-Ver o [relatório metodológico](nada) para mais detalhes sobre os métodos e funcionamento desta ferramenta.
+Ver o [relatório metodológico](https://ushift.tecnico.ulisboa.pt/content/tml/biclar/biclaR_RelatorioFinal_RedeCiclavelMetropolitana_Metodologia.pdf){target="_blank"} para mais detalhes sobre os métodos e funcionamento desta ferramenta.
 
@@
 Atenção que, dependendo do cenário que estamos a analisar, a rede potencial pode cobrir maior ou menor parte da rede planeada por cada Município, sendo que valerá a pena confrontar os resultados dos vários cenários.
+Uma vez que a rede potencial visível no site tem um mínimo de viagens por segmento, recomenda-se que uma comparação mais fina seja feita em SIG com as camadas de rede potencial completas (ver restantes FAQ).
 
@@
 
-> Ver o [relatório metodológico](nada) para mais detalhes sobre o funcionamento desta ferramenta.
+> Ver o [relatório metodológico](https://ushift.tecnico.ulisboa.pt/content/tml/biclar/biclaR_RelatorioFinal_RedeCiclavelMetropolitana_Metodologia.pdf){target="_blank"} para mais detalhes sobre o funcionamento desta ferramenta.
 

### 2023-09-16 paper/PaperTRA/PaperTRA.Rmd
setup Paper TRA 2024

---
# Supported options: 
#   sn-nature:       Style for submissions to Nature Portfolio journals
#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
#   sn-mathphys:     Math and Physical Sciences Reference Style
#   sn-aps:          American Physical Society (APS) Reference Style
#   sn-vancouver:    Vancouver Reference Style
#   sn-apa:          APA Reference Style 
#   sn-chicago:      Chicago-based Humanities Reference Style
#   default:         Default

classoptions: 
  - sn-apa      
  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  - referee       # Optional: Use double line spacing
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools
# titlerunning: Article Title runing

authors: 
  - firstname: Rosa
    lastname: Félix
    email: rosamfelix@tecnico.ulisboa.pt
    affiliation: 1
    corresponding: TRUE
    equalcont: These authors contributed equally to this work.

  - firstname: Filipe
    lastname: Moura
    email: fmoura@tecnico.ulisboa.pt
    affiliation: 1
    corresponding: FALSE
    equalcont: These authors contributed equally to this work.

  - firstname: Robin
    lastname: Lovelace
    email: r.lovelace@leeds.ac.uk
    affiliation: 2
    equalcont: These authors contributed equally to this work.

affiliations:
  - number: 1
    corresponding: TRUE
    info:
      orgdiv: CERIS - Instituto Superior Técnico
      orgname: University of Lisbon
    address:
        street: Av Rovisco Pais 1
        city: Lisbon
        postcode: 1049-001
        # state: State
        country: Portugal
  - number: 2
    corresponding: FALSE
    info:
      orgdiv: Institute for Transport Studies
      orgname: University of Leeds
    address:
        street: 34-40 University Rd
        city: Leeds
        postcode: LS2 9JT
        # state: State
        country: United Kingdom

keywords:
  - Active transport
  - Intermodality
  - First and last mile
  - Health economic assessment
  - Environmental impacts
  - Open data and methods

# Sample for structured abstract - MAX 200 words - ainda tem 274
abstract: |
  In metropolitan areas, car trips can be replaced by a combination of public transit and cycling for the first-and-last mile. This approach requires interventions to make cycling more appealing, and the resulting public investments can have significant social and environmental benefits. This paper focuses on estimating the potential for cycling + PT as a substitute for car trips in the Lisbon metropolitan area and assessing its socio-environmental impacts using open data and open source tools.

  To achieve the cycling targets set by the Portuguese cycling strategy, the Department of Transport introduced _biclaR_, a decision support tool that facilitates the design and development of a metropolitan cycling network. A scenario of intermodality introduced, and its socio-environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools. Additionally, we estimated the impacts of shifting car trips to PT and monetize them with the EU Guide to cost-benefit analysis.

  The results indicate that 20% of the current trips can be made with the bicycle + PT combination, with an additional 12% of PT trips being potentially replaced. Shifting to cycling for the first-and-last mile can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the 10-year socio-environmental benefits account for €230 to €590 million, depending on the cycling targets. For the PT leg, the transfer from car results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per year.

  The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.

bibliography: bibliography.bib

header-includes: |

output: rticles::springer_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      out.extra = "")  # This forces knitr to label all figures.
```

# Introduction {#sec1}

The Introduction section should explain the background to the study, its aims, a summary of the existing literature and why this study was necessary.

In metropolitan areas, car trips can be replaced by a combination of public transit (PT) and cycling for the first-and-last mile. This approach requires interventions and programs to make bicycling more appealing, and the resulting public investments can have significant social and environmental benefits. This paper focuses on estimating the potential for cycling + PT as a substitute for car trips in the Lisbon metropolitan area (LMA) and assessing its socio-environmental impacts using open data and open source tools.

According to the latest mobility survey conducted in 2018, the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle. Car modal share is 58.4%, while PT accounts for 15.5%. To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively), the Department of Transport introduced biclaR, a decision support tool that facilitates the design and development of a metropolitan cycling network.

The intermodality scenario considers trips that can combine PT and cycling for the first-and-last legs. Conservatively, we considered the sum of first-and-last legs up to 5 km. Furthermore, we restricted PT use to unimodal trips without transfers (although they can be included in future modeling). Finally, we only included PT modes that can practically accommodate bicycles, such as trains, ferries, trams, and intermunicipal bus lines with bike racks (Figure 1).

To obtain reliable results, we used the OpenStreetMap road network and GTFS data. The r5r R package estimated the trip duration and distance for both the original modes and the bicycle + PT combination, while the od jittering R package estimated the OD locations based on a centroid-based OD matrix.

Socio-environmental impacts were assessed using the HEAT for Cycling and the HEAT as a Service tools, from the WHO. Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA’s COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.

The results indicate that 20% of the current trips can be made with the bicycle + PT combination, with an additional 12% of PT trips being potentially replaced. Shifting to cycling for the first-and-last mile can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the 10-year socio-environmental benefits account for €230 to €590 million, depending on the cycling targets. For the PT leg, the transfer from car results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per year, or €1.4 to €3.5 million over 10 years, with trains offering the greatest potential for substitution (88%).

By making the research process publicly accessible in a code repository, this study enables the replication of similar estimates for socio-environmental impacts resulting from a modal shift from cars to bicycles + PT in other metropolitan areas. The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.

<!-- The Introduction section, of referenced text @bib1 expands on the background of the work (some overlap with the Abstract is acceptable). The introduction should not include subheadings. -->

# Methods {#sec2}

The methods section should include:

* the aim, design and setting of the study
* the characteristics of participants or description of materials
* a clear description of all processes and methodologies employed. Generic names should generally be used. When proprietary * brands are used in research, include the brand names in parentheses
* the type of statistical analysis used, including a power calculation if appropriate
* studies involving human participants, data or tissue or animals must include statement on ethics approval and consent

```{r fig1, fig.cap = "This is an example of a caption", fig.width = 5}
# plot(pressure)

```

# Results and Discussion {#sec11}

Topical subheadings are allowed. Authors must ensure that their Methods section includes adequate experimental and characterization data necessary for others in the field to reproduce their work. Authors are encouraged to include RIIDs where appropriate.

**Ethical approval declarations** (only required where applicable) Any article reporting experiment/s carried out on (i)~live vertebrate (or higher invertebrates), (ii)~humans or (iii)~human samples must include an unambiguous statement within the methods section that meets the following requirements:

1. Approval: a statement which confirms that all experimental protocols were approved by a named institutional and/or licensing committee. Please identify the approving body in the methods section

2. Accordance: a statement explicitly saying that the methods were carried out in accordance with the relevant guidelines and regulations

3. Informed consent (for experiments involving humans or human tissue samples): include a statement confirming that informed consent was obtained from all participants and/or their legal guardian/s

If your manuscript includes potentially identifying patient/participant information, or if it describes human transplantation research, or if it reports results of a clinical trial then  additional information will be required. Please visit (<https://www.nature.com/nature-research/editorial-policies>) for Nature Portfolio journals, (<https://www.springer.com/gp/authors-editors/journal-author/journal-author-helpdesk/publishing-ethics/14214>) for Springer Nature journals, or (<https://www.biomedcentral.com/getpublished/editorial-policies\#ethics+and+consent>) for BMC.

# Conclusion {#sec13}

This should state clearly the main conclusions and provide an explanation of the importance and relevance of the study to the field.

Conclusions may be used to restate your hypothesis or research question, restate your major findings, explain the relevance and the added value of your work, highlight any limitations of your study, describe future directions for research and recommendations.

In some disciplines use of Discussion or 'Conclusion' is interchangeable. It is not mandatory to use both. Please refer to Journal-level guidance for any specific requirements.

\backmatter

\bmhead{Acknowledgments}

Acknowledgments are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.

Please refer to Journal-level guidance for any specific requirements.

# Declarations {.unnumbered}

Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading `Declarations':

* Funding
* Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
* Ethics approval
* Consent to participate
* Consent for publication
* Availability of data and materials
* Code availability
* Authors' contributions

\noindent
If any of the sections are not relevant to your manuscript, please include the heading and write `Not applicable' for that section.

# References {.unnumbered}

### 2023-09-16 paper/PaperTRA/PaperTRA.tex
setup Paper TRA 2024

%Version 2.1 April 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[sn-apa,referee,pdflatex]{sn-jnl}

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

\raggedbottom

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

\title[]{Estimating the socio-environmental impacts of car substitution
by bicycle and public transit using open tools}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1]{\fnm{Rosa} \sur{Félix} }\email{\href{mailto:rosamfelix@tecnico.ulisboa.pt}{\nolinkurl{rosamfelix@tecnico.ulisboa.pt}}}

\author[1]{\fnm{Filipe} \sur{Moura} }\email{\href{mailto:fmoura@tecnico.ulisboa.pt}{\nolinkurl{fmoura@tecnico.ulisboa.pt}}}

\author[2]{\fnm{Robin} \sur{Lovelace} }\email{\href{mailto:r.lovelace@leeds.ac.uk}{\nolinkurl{r.lovelace@leeds.ac.uk}}}

  \affil*[1]{\orgdiv{CERIS - Instituto Superior
Técnico}, \orgname{University of
Lisbon}, \orgaddress{\city{Lisbon}, \country{Portugal}, \postcode{1049-001}, \street{Av
Rovisco Pais 1}}}
  \affil[2]{\orgdiv{Institute for Transport
Studies}, \orgname{University of
Leeds}, \orgaddress{\city{Leeds}, \country{United
Kingdom}, \postcode{LS2 9JT}, \street{34-40 University Rd}}}

\abstract{In metropolitan areas, car trips can be replaced by a
combination of public transit and cycling for the first-and-last mile.
This approach requires interventions to make cycling more appealing, and
the resulting public investments can have significant social and
environmental benefits. This paper focuses on estimating the potential
for cycling + PT as a substitute for car trips in the Lisbon
metropolitan area and assessing its socio-environmental impacts using
open data and open source tools.

To achieve the cycling targets set by the Portuguese cycling strategy,
the Department of Transport introduced \emph{biclaR}, a decision support
tool that facilitates the design and development of a metropolitan
cycling network. A scenario of intermodality introduced, and its
socio-environmental impacts were assessed using the \emph{HEAT for
Cycling} and the \emph{HEAT as a Service} tools. Additionally, we
estimated the impacts of shifting car trips to PT and monetize them with
the EU Guide to cost-benefit analysis.

The results indicate that 20\% of the current trips can be made with the
bicycle + PT combination, with an additional 12\% of PT trips being
potentially replaced. Shifting to cycling for the first-and-last mile
can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the
10-year socio-environmental benefits account for €230 to €590 million,
depending on the cycling targets. For the PT leg, the transfer from car
results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per
year.

The provided information on socio-economic benefits can support
policymakers in prioritizing interventions to reduce the reliance on
individual motorized transportation and effectively communicate their
decisions.}

\keywords{Active transport, Intermodality, First and last mile, Health
economic assessment, Environmental impacts, Open data and methods}

\maketitle

\hypertarget{sec1}{%
\section{Introduction}\label{sec1}}

The Introduction section should explain the background to the study, its
aims, a summary of the existing literature and why this study was
necessary.

In metropolitan areas, car trips can be replaced by a combination of
public transit (PT) and cycling for the first-and-last mile. This
approach requires interventions and programs to make bicycling more
appealing, and the resulting public investments can have significant
social and environmental benefits. This paper focuses on estimating the
potential for cycling + PT as a substitute for car trips in the Lisbon
metropolitan area (LMA) and assessing its socio-environmental impacts
using open data and open source tools.

According to the latest mobility survey conducted in 2018, the LMA
registered a total of 5.3 million daily trips, with only 0.5\% by
bicycle. Car modal share is 58.4\%, while PT accounts for 15.5\%. To
achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively), the Department
of Transport introduced biclaR, a decision support tool that facilitates
the design and development of a metropolitan cycling network.

The intermodality scenario considers trips that can combine PT and
cycling for the first-and-last legs. Conservatively, we considered the
sum of first-and-last legs up to 5 km. Furthermore, we restricted PT use
to unimodal trips without transfers (although they can be included in
future modeling). Finally, we only included PT modes that can
practically accommodate bicycles, such as trains, ferries, trams, and
intermunicipal bus lines with bike racks (Figure 1).

To obtain reliable results, we used the OpenStreetMap road network and
GTFS data. The r5r R package estimated the trip duration and distance
for both the original modes and the bicycle + PT combination, while the
od jittering R package estimated the OD locations based on a
centroid-based OD matrix.

Socio-environmental impacts were assessed using the HEAT for Cycling and
the HEAT as a Service tools, from the WHO. Additionally, we estimate the
impacts of shifting car trips to PT for the second leg of the journey
with EMEP/EEA's COPERT methodology and monetize them with the EU Guide
to cost-benefit analysis.

The results indicate that 20\% of the current trips can be made with the
bicycle + PT combination, with an additional 12\% of PT trips being
potentially replaced. Shifting to cycling for the first-and-last mile
can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the
10-year socio-environmental benefits account for €230 to €590 million,
depending on the cycling targets. For the PT leg, the transfer from car
results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per
year, or €1.4 to €3.5 million over 10 years, with trains offering the
greatest potential for substitution (88\%).

By making the research process publicly accessible in a code repository,
this study enables the replication of similar estimates for
socio-environmental impacts resulting from a modal shift from cars to
bicycles + PT in other metropolitan areas. The provided information on
socio-economic benefits can support policymakers in prioritizing
interventions to reduce the reliance on individual motorized
transportation and effectively communicate their decisions.

The Introduction section, of referenced text \citet{bib1} expands on the
background of the work (some overlap with the Abstract is acceptable).
The introduction should not include subheadings.

Springer Nature does not impose a strict layout as standard however
authors are advised to check the individual requirements for the journal
they are planning to submit to as there may be journal-level
preferences. When preparing your text please also be aware that some
stylistic choices are not supported in full text XML (publication
version), including coloured font. These will not be replicated in the
typeset article if it is accepted.

\hypertarget{sec2}{%
\section{Methods}\label{sec2}}

The methods section should include:

\begin{itemize}
\tightlist
\item
  the aim, design and setting of the study
\item
  the characteristics of participants or description of materials
\item
  a clear description of all processes and methodologies employed.
  Generic names should generally be used. When proprietary * brands are
  used in research, include the brand names in parentheses
\item
  the type of statistical analysis used, including a power calculation
  if appropriate
\item
  studies involving human participants, data or tissue or animals must
  include statement on ethics approval and consent
\end{itemize}

\begin{figure}
\includegraphics{PaperTRA_files/figure-latex/fig1-1} \caption{This is an example of a caption}\label{fig:fig1}
\end{figure}

\hypertarget{sec11}{%
\section{Results and Discussion}\label{sec11}}

Topical subheadings are allowed. Authors must ensure that their Methods
section includes adequate experimental and characterization data
necessary for others in the field to reproduce their work. Authors are
encouraged to include RIIDs where appropriate.

\textbf{Ethical approval declarations} (only required where applicable)
Any article reporting experiment/s carried out on
(i)\textasciitilde live vertebrate (or higher invertebrates),
(ii)\textasciitilde humans or (iii)\textasciitilde human samples must
include an unambiguous statement within the methods section that meets
the following requirements:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Approval: a statement which confirms that all experimental protocols
  were approved by a named institutional and/or licensing committee.
  Please identify the approving body in the methods section
\item
  Accordance: a statement explicitly saying that the methods were
  carried out in accordance with the relevant guidelines and regulations
\item
  Informed consent (for experiments involving humans or human tissue
  samples): include a statement confirming that informed consent was
  obtained from all participants and/or their legal guardian/s
\end{enumerate}

If your manuscript includes potentially identifying patient/participant
information, or if it describes human transplantation research, or if it
reports results of a clinical trial then additional information will be
required. Please visit
(\url{https://www.nature.com/nature-research/editorial-policies}) for
Nature Portfolio journals,
(\url{https://www.springer.com/gp/authors-editors/journal-author/journal-author-helpdesk/publishing-ethics/14214})
for Springer Nature journals, or
(\url{https://www.biomedcentral.com/getpublished/editorial-policies/\#ethics+and+consent})
for BMC.

\hypertarget{sec13}{%
\section{Conclusion}\label{sec13}}

This should state clearly the main conclusions and provide an
explanation of the importance and relevance of the study to the field.

Conclusions may be used to restate your hypothesis or research question,
restate your major findings, explain the relevance and the added value
of your work, highlight any limitations of your study, describe future
directions for research and recommendations.

In some disciplines use of Discussion or `Conclusion' is
interchangeable. It is not mandatory to use both. Please refer to
Journal-level guidance for any specific requirements.

\backmatter

\bmhead{Acknowledgments}

Acknowledgments are not compulsory. Where included they should be brief.
Grant or contribution numbers may be acknowledged.

Please refer to Journal-level guidance for any specific requirements.

\hypertarget{declarations}{%
\section*{Declarations}\label{declarations}}
\addcontentsline{toc}{section}{Declarations}

Some journals require declarations to be submitted in a standardised
format. Please check the Instructions for Authors of the journal to
which you are submitting to see if you need to complete this section. If
yes, your manuscript must contain the following sections under the
heading `Declarations':

\begin{itemize}
\tightlist
\item
  Funding
\item
  Conflict of interest/Competing interests (check journal-specific
  guidelines for which heading to use)
\item
  Ethics approval
\item
  Consent to participate
\item
  Consent for publication
\item
  Availability of data and materials
\item
  Code availability
\item
  Authors' contributions
\end{itemize}

\noindent If any of the sections are not relevant to your manuscript,
please include the heading and write `Not applicable' for that section.

\renewcommand\refname{References}
\bibliography{bibliography.bib}

\end{document}

### 2023-09-16 paper/PaperTRA/skeleton.tex
setup Paper TRA 2024

%Version 2.1 April 2023
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[sn-basic,pdflatex]{sn-jnl}

%%%% Standard Packages
%%<additional latex packages if required can be included here>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published
%%%%  by Springer Nature. The guidance has been prepared in partnership with
%%%%  production teams to conform to Springer Nature technical requirements.
%%%%  Editorial and presentation requirements differ among journal portfolios and
%%%%  research disciplines. You may find sections in this template are irrelevant
%%%%  to your work and are empowered to omit any such section if allowed by the
%%%%  journal you intend to submit to. The submission guidelines and policies
%%%%  of the journal take precedence. A detailed User Manual is available in the
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% Per the spinger doc, new theorem styles can be included using built in style, 
%% but it seems the don't work so commented below
%\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
\newtheorem{proposition}[theorem]{Proposition}%
%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

%% \theoremstyle{thmstyletwo}%
\theoremstyle{remark}
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

%% \theoremstyle{thmstylethree}%
\theoremstyle{definition}
\newtheorem{definition}{Definition}%

\raggedbottom

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

\title[Article Title runing]{Article Title}

%%=============================================================%%
%% Prefix	-> \pfx{Dr}
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% NatureName	-> \tanm{Poet Laureate} -> Title after name
%% Degrees	-> \dgr{MSc, PhD}
%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
%%=============================================================%%

\author*[1,2]{\pfx{Dr.} \fnm{Leading} \spfx{van} \sur{Author} \sfx{III} \tanm{Poet
Laureate} \dgr{MSc,
PhD}}\email{\href{mailto:abc@def}{\nolinkurl{abc@def}}}

\author[2]{\fnm{Second} \sur{Author} }

  \affil*[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\city{City}, \country{Country}, \postcode{100190}, \state{State}, \street{Street}}}
  \affil*[2]{\orgname{Other Organisation}}

\abstract{\textbf{Purpose}: The abstract serves both as a general
introduction to the topic and as a brief, non-technical summary of the
main results and their implications. The abstract must not include
subheadings (unless expressly permitted in the journal's Instructions to
Authors), equations or citations. As a guide the abstract should not
exceed 200 words. Most journals do not set a hard limit however authors
are advised to check the author instructions for the journal they are
submitting to.

\textbf{Methods:} The abstract serves both as a general introduction to
the topic and as a brief, non-technical summary of the main results and
their implications. The abstract must not include subheadings (unless
expressly permitted in the journal's Instructions to Authors), equations
or citations. As a guide the abstract should not exceed 200 words. Most
journals do not set a hard limit however authors are advised to check
the author instructions for the journal they are submitting to.

\textbf{Results:} The abstract serves both as a general introduction to
the topic and as a brief, non-technical summary of the main results and
their implications. The abstract must not include subheadings (unless
expressly permitted in the journal's Instructions to Authors), equations
or citations. As a guide the abstract should not exceed 200 words. Most
journals do not set a hard limit however authors are advised to check
the author instructions for the journal they are submitting to.

\textbf{Conclusion:} The abstract serves both as a general introduction
to the topic and as a brief, non-technical summary of the main results
and their implications. The abstract must not include subheadings
(unless expressly permitted in the journal's Instructions to Authors),
equations or citations. As a guide the abstract should not exceed 200
words. Most journals do not set a hard limit however authors are advised
to check the author instructions for the journal they are submitting
to.\}}

\keywords{key, dictionary, word}

\pacs[JEL Classification]{D8, H51}
\pacs[MSC Classification]{35A01, 65L10}

\maketitle

\hypertarget{sec1}{%
\section{Introduction}\label{sec1}}

The Introduction section, of referenced text \citet{bib1} expands on the
background of the work (some overlap with the Abstract is acceptable).
The introduction should not include subheadings.

Springer Nature does not impose a strict layout as standard however
authors are advised to check the individual requirements for the journal
they are planning to submit to as there may be journal-level
preferences. When preparing your text please also be aware that some
stylistic choices are not supported in full text XML (publication
version), including coloured font. These will not be replicated in the
typeset article if it is accepted.

\hypertarget{sec2}{%
\section{Results}\label{sec2}}

Sample body text. Sample body text. Sample body text. Sample body text.
Sample body text. Sample body text. Sample body text. Sample body text.

\hypertarget{sec3}{%
\section{This is an example for first level head---section
head}\label{sec3}}

\hypertarget{subsec2}{%
\subsection{This is an example for second level head---subsection
head}\label{subsec2}}

\hypertarget{subsubsec2}{%
\subsubsection{This is an example for third level head---subsubsection
head}\label{subsubsec2}}

Sample body text. Sample body text. Sample body text. Sample body text.
Sample body text. Sample body text. Sample body text.

\hypertarget{sec4}{%
\section{Equations}\label{sec4}}

Equations in \LaTeX~can either be inline or on-a-line by itself
(``display equations''). For inline equations use the \texttt{\$...\$}
commands. E.g.: The equation \(H\psi = E \psi\) is written via the
command \texttt{\$H\ \textbackslash{}psi\ =\ E\ \textbackslash{}psi\$}.

For display equations (with auto generated equation numbers) one can use
the equation or align environments:

\begin{equation}
\|\tilde{X}(k)\|^2 \leq\frac{\sum\limits_{i=1}^{p}\left\|\tilde{Y}_i(k)\right\|^2+\sum\limits_{j=1}^{q}\left\|\tilde{Z}_j(k)\right\|^2 }{p+q}.\label{eq1}
\end{equation}

where, \begin{align}
D_\mu &=  \partial_\mu - ig \frac{\lambda^a}{2} A^a_\mu \nonumber \\
F^a_{\mu\nu} &= \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^a_\nu \label{eq2}
\end{align}

Notice the use of \texttt{\textbackslash{}nonumber} in the align
environment at the end of each line, except the last, so as not to
produce equation numbers on lines where no equation numbers are
required. The \texttt{\textbackslash{}label\{\}} command should only be
used at the last line of an align environment where
\texttt{\textbackslash{}nonumber} is not used.

\begin{equation}
Y_\infty = \left( \frac{m}{\textrm{GeV}} \right)^{-3}
    \left[ 1 + \frac{3 \ln(m/\textrm{GeV})}{15}
    + \frac{\ln(c_2/5)}{15} \right]
\end{equation}

The class file also supports the use of
\texttt{\textbackslash{}mathbb\{\}},
\texttt{\textbackslash{}mathscr\{\}} and
\texttt{\textbackslash{}mathcal\{\}} commands. As such
\texttt{\textbackslash{}mathbb\{R\}},
\texttt{\textbackslash{}mathscr\{R\}} and
\texttt{\textbackslash{}mathcal\{R\}} produces \(\mathbb{R}\),
\(\mathscr{R}\) and \(\mathcal{R}\) respectively (refer
Subsubsection~\ref{subsubsec2}).

\hypertarget{sec5}{%
\section{Tables}\label{sec5}}

Tables can be inserted via the normal \texttt{knitr::kable()} function
or other table-generating packages.

\begin{table}

\caption{\label{tab:tab1}Caption text}
\centering
\begin{tabular}[t]{r|r}
\hline
temperature & pressure\\
\hline
0 & 0.0002\\
\hline
20 & 0.0012\\
\hline
40 & 0.0060\\
\hline
60 & 0.0300\\
\hline
80 & 0.0900\\
\hline
100 & 0.2700\\
\hline
\end{tabular}
\end{table}

Tables can also be inserted via the normal table and tabular
environment. To put footnotes inside tables you should use
\texttt{\textbackslash{}footnotetext{[}{]}\{...\}} tag. The footnote
appears just below the table itself (refer
Tables\textasciitilde{}\ref{tab1} and \ref{tab2}). For the corresponding
footnotemark use \texttt{\textbackslash{}footnotemark{[}...{]}}

\begin{table}[h]
\caption{Caption text}\label{tab1}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2  & Column 3 & Column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
\end{table}

\noindent The input format for the above table is as follows:

\begin{verbatim}
\begin{table}[<placement-specifier>]
\caption{<table-caption>}\label{<table-label>}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2 & Column 3 & Column 4\\
\midrule
row 1 & data 1 & data 2  & data 3 \\
row 2 & data 4 & data 5\footnotemark[1] & data 6 \\
row 3 & data 7 & data 8  & data 9\footnotemark[2]\\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote.
This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote.
This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote.
This is an example of table footnote.}
\end{table}
\end{verbatim}

\begin{table}[h]
\caption{Example of a lengthy table which is set to full textwidth}\label{tab2}
\begin{tabular*}{\textwidth}{@{\extracolsep\fill}lcccccc}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]} & \multicolumn{3}{@{}c@{}}{Element 2\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
\botrule
\end{tabular*}
\footnotetext{Note: This is an example of table footnote. This is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote.}
\footnotetext[2]{Example for a second table footnote.}
\end{table}

\vfill\eject

In case of double column layout, tables which do not fit in single
column width should be set to full text width. For this, you need to use
\texttt{\textbackslash{}begin\{table*\}} \texttt{...}
\texttt{\textbackslash{}end\{table*\}} instead of
\texttt{\textbackslash{}begin\{table\}} \texttt{...}
\texttt{\textbackslash{}end\{table\}} environment. Lengthy tables which
do not fit in textwidth should be set as rotated table. For this, you
need to use \texttt{\textbackslash{}begin\{sidewaystable\}} \texttt{...}
\texttt{\textbackslash{}end\{sidewaystable\}} instead of
\texttt{\textbackslash{}begin\{table*\}} \texttt{...}
\texttt{\textbackslash{}end\{table*\}} environment. This environment
puts tables rotated to single column width. For tables rotated to double
column width, use \texttt{\textbackslash{}begin\{sidewaystable*\}}
\texttt{...} \texttt{\textbackslash{}end\{sidewaystable*\}}.

\begin{sidewaystable}
\caption{Tables which are too long to fit, should be written using the "sidewaystable" environment as shown here}\label{tab3}
\begin{tabular*}{\textheight}{@{\extracolsep\fill}lcccccc}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]}& \multicolumn{3}{@{}c@{}}{Element\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
Projectile & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 4 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
Element 5 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 6 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
\botrule
\end{tabular*}
\footnotetext{Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
\footnotetext[1]{This is an example of table footnote.}
\end{sidewaystable}

\hypertarget{sec6}{%
\section{Figures}\label{sec6}}

As per the \LaTeX~standards you need to use eps images for
\LaTeX~compilation and \texttt{pdf/jpg/png} images for \texttt{PDFLaTeX}
compilation. Use the \texttt{dev} knitr option to use the approrpate
format. This is one of the major difference between \LaTeX~and
\texttt{PDFLaTeX}. Each image should be from a single input .eps/vector
image file. Avoid using subfigures. The command for inserting images for
\LaTeX~and \texttt{PDFLaTeX} can be generalized. The package used to
insert images in \texttt{LaTeX/PDFLaTeX} is the graphicx package.
Figures can be inserted via the normal figure environment as shown in
the below example:

\begin{figure}
\includegraphics{skeleton_files/figure-latex/fig1-1} \caption{This is an example of a caption}\label{fig:fig1}
\end{figure}

\hypertarget{sec7}{%
\section{Algorithms, Program codes and Listings}\label{sec7}}

Packages \texttt{algorithm}, \texttt{algorithmicx} and
\texttt{algpseudocode} are used for setting algorithms in \LaTeX~using
the format:

\begin{verbatim}
\begin{algorithm}
\caption{<alg-caption>}\label{<alg-label>}
\begin{algorithmic}[1]
. . .
\end{algorithmic}
\end{algorithm}
\end{verbatim}

You may refer above listed package documentations for more details
before setting \texttt{algorithm} environment. For program codes, the
``program'' package is required and the command to be used is
\texttt{\textbackslash{}begin\{program\}} \texttt{...}
\texttt{\textbackslash{}end\{program\}}. A fast exponentiation
procedure:

Similarly, for \texttt{listings}, use the \texttt{listings} package.
\texttt{\textbackslash{}begin\{lstlisting\}} \texttt{...}
\texttt{\textbackslash{}end\{lstlisting\}} is used to set environments
similar to \texttt{verbatim} environment. Refer to the
\texttt{lstlisting} package documentation for more details.

A fast exponentiation procedure:

\lstset{texcl=true,basicstyle=\small\sf,commentstyle=\small\rm,mathescape=true,escapeinside={(*}{*)}}
\begin{lstlisting}
begin
  for $i:=1$ to $10$ step $1$ do
      expt($2,i$);  
      newline() od                (*\textrm{Comments will be set flush to the right margin}*)
where
proc expt($x,n$) $\equiv$
  $z:=1$;
  do if $n=0$ then exit fi;
     do if odd($n$) then exit fi;                 
        comment: (*\textrm{This is a comment statement;}*)
        $n:=n/2$; $x:=x*x$ od;
     { $n>0$ };
     $n:=n-1$; $z:=z*x$ od;
  print($z$). 
end
\end{lstlisting}

\begin{algorithm}
\caption{Calculate $y = x^n$}\label{algo1}
\begin{algorithmic}[1]
\Require $n \geq 0 \vee x \neq 0$
\Ensure $y = x^n$ 
\State $y \Leftarrow 1$
\If{$n < 0$}\label{algln2}
        \State $X \Leftarrow 1 / x$
        \State $N \Leftarrow -n$
\Else
        \State $X \Leftarrow x$
        \State $N \Leftarrow n$
\EndIf
\While{$N \neq 0$}
        \If{$N$ is even}
            \State $X \Leftarrow X \times X$
            \State $N \Leftarrow N / 2$
        \Else[$N$ is odd]
            \State $y \Leftarrow y \times X$
            \State $N \Leftarrow N - 1$
        \EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}

\begin{minipage}{\hsize}

\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}

\begin{lstlisting}
for i:=maxint to 0 do begin \{ do nothing \} end; Write(`Case
insensitive'); Write(`Pascal keywords.');

\end{lstlisting}

\end{minipage}

\hypertarget{sec8}{%
\section{Cross referencing}\label{sec8}}

Figures and tables are labeled with a prefix (fig or tab, respectively)
plus the chunk label. Other environments such as equation and align can
be labelled via the \texttt{\textbackslash{}label\{\#label\}} command
inside or just below the \texttt{\textbackslash{}caption\{\}} command.
You can then use the label for cross-reference. As an example, consider
the chunk label declared for Figure~\ref{fig:fig1} which is fig1. To
cross-reference it, use the command
\texttt{Figure\ \textbackslash{}ref\{fig:fig1\}}, for which it comes up
as ``Figure~\ref{fig:fig1}''.

To reference line numbers in an algorithm, consider the label declared
for the line number 2 of Algorithm~\ref{algo1} is
\texttt{\textbackslash{}label\{algln2\}}. To cross-reference it, use the
command \texttt{\textbackslash{}ref\{algln2\}} for which it comes up as
line~\ref{algln2} of Algorithm~\ref{algo1}.

\hypertarget{subsec7}{%
\subsection{Details on reference citations}\label{subsec7}}

For citations of references, use~\citet{bib1} or \citep{bib2}.

\hypertarget{sec10}{%
\section{Examples for theorem like environments}\label{sec10}}

The documentclass for springer \texttt{sn-jnl.cls} contains 3 styling
that you can use to set new default for theorems and proofs type

\begin{description}
\item[\texttt{thmstyleone}]
Numbered, theorem head in bold font and theorem text in italic style
\item[\texttt{thmstyletwo}]
Numbered, theorem head in roman font and theorem text in italic style
\item[\texttt{thmstylethree}]
Numbered, theorem head in bold font and theorem text in roman style
\end{description}

For mathematics journals, theorem styles can be included as shown in the
following examples.

\begin{theorem}
Example theorem text. Example theorem text. Example theorem text.
Example theorem text. Example theorem text. Example theorem text.
Example theorem text. Example theorem text. Example theorem text.
Example theorem text. Example theorem text.

\end{theorem}

To add labels and subheadings, use LaTeX notation

\begin{theorem}[Theorem subhead]\label{thm1}
Example theorem text. Example theorem text. Example theorem text.
Example theorem text. Example theorem text. Example theorem text.
Example theorem text. Example theorem text. Example theorem text.
Example theorem text. Example theorem text.

\end{theorem}

Other environments are proposition, example, remark, definition, proof
and quote

Sample body text. Sample body text. Sample body text. Sample body text.
Sample body text. Sample body text. Sample body text. Sample body text.

\begin{proposition}
Example proposition text. Example proposition text. Example proposition
text. Example proposition text. Example proposition text. Example
proposition text. Example proposition text. Example proposition text.
Example proposition text. Example proposition text.

\end{proposition}

Sample body text. Sample body text. Sample body text. Sample body text.
Sample body text. Sample body text. Sample body text. Sample body text.

\begin{example}
Phasellus adipiscing semper elit. Proin fermentum massa ac quam. Sed
diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas
lacinia. Nam ipsum ligula, eleifend at, accumsan nec, suscipit a, ipsum.
Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem.

\end{example}

Sample body text. Sample body text. Sample body text. Sample body text.
Sample body text. Sample body text. Sample body text. Sample body text.

\begin{remark}
Phasellus adipiscing semper elit. Proin fermentum massa ac quam. Sed
diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas
lacinia. Nam ipsum ligula, eleifend at, accumsan nec, suscipit a, ipsum.
Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem.

\end{remark}

Sample body text. Sample body text. Sample body text. Sample body text.
Sample body text. Sample body text. Sample body text. Sample body text.

\begin{definition}[Definition sub head]
Example definition text. Example definition text. Example definition
text. Example definition text. Example definition text. Example
definition text. Example definition text. Example definition text.

\end{definition}

Additionally a predefined ``proof'' environment is available. This
prints a ``Proof'' head in italic font style and the ``body text'' in
roman font style with an open square at the end of each proof
environment.

\begin{proof}
Example for proof text. Example for proof text. Example for proof text.
Example for proof text. Example for proof text. Example for proof text.
Example for proof text. Example for proof text. Example for proof text.
Example for proof text.

\end{proof}

Sample body text. Sample body text. Sample body text. Sample body text.
Sample body text. Sample body text. Sample body text. Sample body text.

\hypertarget{sec11}{%
\section{Methods}\label{sec11}}

Topical subheadings are allowed. Authors must ensure that their Methods
section includes adequate experimental and characterization data
necessary for others in the field to reproduce their work. Authors are
encouraged to include RIIDs where appropriate.

\textbf{Ethical approval declarations} (only required where applicable)
Any article reporting experiment/s carried out on
(i)\textasciitilde live vertebrate (or higher invertebrates),
(ii)\textasciitilde humans or (iii)\textasciitilde human samples must
include an unambiguous statement within the methods section that meets
the following requirements:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Approval: a statement which confirms that all experimental protocols
  were approved by a named institutional and/or licensing committee.
  Please identify the approving body in the methods section
\item
  Accordance: a statement explicitly saying that the methods were
  carried out in accordance with the relevant guidelines and regulations
\item
  Informed consent (for experiments involving humans or human tissue
  samples): include a statement confirming that informed consent was
  obtained from all participants and/or their legal guardian/s
\end{enumerate}

If your manuscript includes potentially identifying patient/participant
information, or if it describes human transplantation research, or if it
reports results of a clinical trial then additional information will be
required. Please visit
(\url{https://www.nature.com/nature-research/editorial-policies}) for
Nature Portfolio journals,
(\url{https://www.springer.com/gp/authors-editors/journal-author/journal-author-helpdesk/publishing-ethics/14214})
for Springer Nature journals, or
(\url{https://www.biomedcentral.com/getpublished/editorial-policies/\#ethics+and+consent})
for BMC.

\hypertarget{sec12}{%
\section{Discussion}\label{sec12}}

Discussions should be brief and focused. In some disciplines use of
Discussion or `Conclusion' is interchangeable. It is not mandatory to
use both. Some journals prefer a section `Results and Discussion'
followed by a section `Conclusion'. Please refer to Journal-level
guidance for any specific requirements.

\hypertarget{sec13}{%
\section{Conclusion}\label{sec13}}

Conclusions may be used to restate your hypothesis or research question,
restate your major findings, explain the relevance and the added value
of your work, highlight any limitations of your study, describe future
directions for research and recommendations.

In some disciplines use of Discussion or `Conclusion' is
interchangeable. It is not mandatory to use both. Please refer to
Journal-level guidance for any specific requirements.

\backmatter

\bmhead{Supplementary information}

If your article has accompanying supplementary file/s please state so
here.

Authors reporting data from electrophoretic gels and blots should supply
the full unprocessed scans for key as part of their Supplementary
information. This may be requested by the editorial team/s if it is
missing.

Please refer to Journal-level guidance for any specific requirements.

\bmhead{Acknowledgments}

Acknowledgments are not compulsory. Where included they should be brief.
Grant or contribution numbers may be acknowledged.

Please refer to Journal-level guidance for any specific requirements.

\hypertarget{declarations}{%
\section*{Declarations}\label{declarations}}
\addcontentsline{toc}{section}{Declarations}

Some journals require declarations to be submitted in a standardised
format. Please check the Instructions for Authors of the journal to
which you are submitting to see if you need to complete this section. If
yes, your manuscript must contain the following sections under the
heading `Declarations':

\begin{itemize}
\tightlist
\item
  Funding
\item
  Conflict of interest/Competing interests (check journal-specific
  guidelines for which heading to use)
\item
  Ethics approval
\item
  Consent to participate
\item
  Consent for publication
\item
  Availability of data and materials
\item
  Code availability
\item
  Authors' contributions
\end{itemize}

\noindent If any of the sections are not relevant to your manuscript,
please include the heading and write `Not applicable' for that section.

\begin{flushleft}
Editorial Policies for:

\noindent Springer journals and proceedings:
\url{https://www.springer.com/gp/editorial-policies}

\noindent Nature Portfolio journals:
\url{https://www.nature.com/nature-research/editorial-policies}

\noindent \textit{Scientific Reports}:
\url{https://www.nature.com/srep/journal-policies/editorial-policies}

\noindent BMC journals:
\url{https://www.biomedcentral.com/getpublished/editorial-policies}

\end{flushleft}

\begin{appendices}

\hypertarget{secA1}{%
\section{Section title of first appendix}\label{secA1}}

An appendix contains supplementary information that is not an essential
part of the text itself but which may be helpful in providing a more
comprehensive understanding of the research problem or it is information
that is too cumbersome to be included in the body of the paper.

For submissions to Nature Portfolio Journals please use the heading
``Extended Data''.

\end{appendices}

\bibliography{bibliography.bib}

\end{document}

### 2023-09-21 paper/PaperTRA/PaperTRA.Rmd
references paper
move to full paper TRA

@@
 ---
-# Supported options: 
-#   sn-nature:       Style for submissions to Nature Portfolio journals
-#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
-#   sn-mathphys:     Math and Physical Sciences Reference Style
-#   sn-aps:          American Physical Society (APS) Reference Style
-#   sn-vancouver:    Vancouver Reference Style
-#   sn-apa:          APA Reference Style 
-#   sn-chicago:      Chicago-based Humanities Reference Style
-#   default:         Default
-
-classoptions: 
-  - sn-apa      
-  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
-  - referee       # Optional: Use double line spacing
-  # - lineno        # Optional: Add line numbers
-  # - iicol         # Optional: Double column layour
-
-title: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools
-# titlerunning: Article Title runing
+classoption: "runningheads"
+title: "Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
+shorttitle: "Estimating the SE impacts of car substitution by bicycle and public transit"
 
 authors: 
-  - firstname: Rosa
-    lastname: Félix
-    email: rosamfelix@tecnico.ulisboa.pt
-    affiliation: 1
-    corresponding: TRUE
-    equalcont: These authors contributed equally to this work.
+  - name: "Rosa Félix"
+    inst: "1"
+    orcid: "0000-0002-5642-6006"
+  - name: "Filipe Moura"
+    inst: "1"
+    orcid: "0000-0001-7749-8490"
+  - name: "Robin Lovelace"
+    inst: "2"
+    orcid: "0000-0001-5679-6536"
+# First names are abbreviated in the running head.
+# If there are more than two authors, 'et al.' is used.
+authorrunning: "R. Félix et al." 
+institutes: 
+  - name: "CERIS - Instituto Superior Técnico, University of Lisbon. Av Rovisco Pais 1049-001 Lisboa, Portugal"
+    email: "rosamfelix@tecnico.ulisboa.pt"
+  - name: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
 
-  - firstname: Filipe
-    lastname: Moura
-    email: fmoura@tecnico.ulisboa.pt
-    affiliation: 1
-    corresponding: FALSE
-    equalcont: These authors contributed equally to this work.
-
-  - firstname: Robin
-    lastname: Lovelace
-    email: r.lovelace@leeds.ac.uk
-    affiliation: 2
-    equalcont: These authors contributed equally to this work.
-
-affiliations:
-  - number: 1
-    corresponding: TRUE
-    info:
-      orgdiv: CERIS - Instituto Superior Técnico
-      orgname: University of Lisbon
-    address:
-        street: Av Rovisco Pais 1
-        city: Lisbon
-        postcode: 1049-001
-        # state: State
-        country: Portugal
-  - number: 2
-    corresponding: FALSE
-    info:
-      orgdiv: Institute for Transport Studies
-      orgname: University of Leeds
-    address:
-        street: 34-40 University Rd
-        city: Leeds
-        postcode: LS2 9JT
-        # state: State
-        country: United Kingdom
-
+abstract: | 
+  In metropolitan areas, car trips can be replaced by a combination of public transit and cycling for the first-and-last mile.
+This paper focuses on estimating the potential for cycling + PT as a substitute for car trips in the Lisbon metropolitan area and assessing its socio-environmental impacts using open data and open source tools.
+  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_).
+A scenario of intermodality introduced, and its socio-environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
+Additionally, the impacts of shifting car trips to PT were estimated and monetized.
+  The results indicate that 20% of the current trips can be made with the bicycle + PT combination.
+Shifting to cycling for the first-and-last mile can reduce annual CO2eq emissions from 6,000 tons/day, and the 10-year socio-environmental benefits account from €230 million.
+For the PT leg, the transfer from car results in the avoidance of at lest 8,500 tons of CO2eq emissions per year.
+  The information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.
 keywords: 
@@
   - Open data and methods
+bibliography: bibliography.bib
+# Use csl when using default citation_package for Pandoc citeproc
+# csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
+# citations with arabic numbers is prefered when using natbib
+natbiboptions: numbers
 
-# Sample for structured abstract - MAX 200 words - ainda tem 274
-abstract: |
-  In metropolitan areas, car trips can be replaced by a combination of public transit and cycling for the first-and-last mile.
-This approach requires interventions to make cycling more appealing, and the resulting public investments can have significant social and environmental benefits.
-This paper focuses on estimating the potential for cycling + PT as a substitute for car trips in the Lisbon metropolitan area and assessing its socio-environmental impacts using open data and open source tools.
+header-includes:
+  - \usepackage[hidelinks]{hyperref}
 
-  To achieve the cycling targets set by the Portuguese cycling strategy, the Department of Transport introduced _biclaR_, a decision support tool that facilitates the design and development of a metropolitan cycling network.
-A scenario of intermodality introduced, and its socio-environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
-Additionally, we estimated the impacts of shifting car trips to PT and monetize them with the EU Guide to cost-benefit analysis.
-
-  The results indicate that 20% of the current trips can be made with the bicycle + PT combination, with an additional 12% of PT trips being potentially replaced.
-Shifting to cycling for the first-and-last mile can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the 10-year socio-environmental benefits account for €230 to €590 million, depending on the cycling targets.
-For the PT leg, the transfer from car results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per year.
-
-  The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.
-
-bibliography: bibliography.bib
-
-header-includes: |
-
-output: rticles::springer_article
+output: 
+  rticles::lncs_article:
+    citation_package: default # or natbib
 ---
@@
 
-# Introduction {#sec1}
+# Introduction
 
-The Introduction section should explain the background to the study, its aims, a summary of the existing literature and why this study was necessary.
+**full paper**: 4-6 pages in length (typically up to 3,000 words).
 
@@
 
-According to the latest mobility survey conducted in 2018, the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.
-Car modal share is 58.4%, while PT accounts for 15.5%.
+According to the latest mobility survey conducted in 2018, the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle. Car modal share is 58.4%, while PT accounts for 15.5%.
 To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively), the Department of Transport introduced biclaR, a decision support tool that facilitates the design and development of a metropolitan cycling network.
+
+This research aims to present and discuss the methods used to estimate…
+
+Propensity to Cycle Tool
+
+adding up an intermodality scenario to estimate cycling potential to public transit interfaces, and thus to support planning and prioritize investments in the cycling network.
+
+# Methods
+
+## Case Study
+
+## Modeling Origin-Destination trips
+
+## Modeling intermodality
 
@@
 Furthermore, we restricted PT use to unimodal trips without transfers (although they can be included in future modeling).
-Finally, we only included PT modes that can practically accommodate bicycles, such as trains, ferries, trams, and intermunicipal bus lines with bike racks (Figure 1).
+[@bib2, @felix2023]
+
+Finally, we only included PT modes that can practically accommodate bicycles, such as trains, ferries, trams, and intermunicipal bus lines with bike racks (\ \ref{fig:map1}).
+
+```{r map1, fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area"}
+plot(pressure)
+```
 
@@
 
+## Assessing socio-environmental benefits
+
 Socio-environmental impacts were assessed using the HEAT for Cycling and the HEAT as a Service tools, from the WHO.
 Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA’s COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.
+
+Table\ \ref{tab:table} gives a summary of all heading levels.
+
+```{r table}
+knitr::kable(head(pressure), 
+             format = "latex",
+             caption = "Table captions should be placed above the tables.")
+```
+
+# Results and Discussion
 
@@
 
+Map result (\ \ref{fig:map2}).
+
+```{r map2, fig.cap="Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool)."}
+plot(pressure)
+```
+
+# Conclusion
+
 By making the research process publicly accessible in a code repository, this study enables the replication of similar estimates for socio-environmental impacts resulting from a modal shift from cars to bicycles + PT in other metropolitan areas.
@@
 
-<!-- The Introduction section, of referenced text @bib1 expands on the background of the work (some overlap with the Abstract is acceptable). The introduction should not include subheadings. -->
+### Acknowledgements {.unnumbered}
+Please place your acknowledgments at the end of the paper, preceded by an unnumbered run-in heading (i.e. 3rd-level heading).
 
-# Methods {#sec2}
-
-The methods section should include:
-
-* the aim, design and setting of the study
-* the characteristics of participants or description of materials
-* a clear description of all processes and methodologies employed. Generic names should generally be used. When proprietary * brands are used in research, include the brand names in parentheses
-* the type of statistical analysis used, including a power calculation if appropriate
-* studies involving human participants, data or tissue or animals must include statement on ethics approval and consent
-
-```{r fig1, fig.cap = "This is an example of a caption", fig.width = 5}
-# plot(pressure)
-
-```
-
-# Results and Discussion {#sec11}
-
-Topical subheadings are allowed.
-Authors must ensure that their Methods section includes adequate experimental and characterization data necessary for others in the field to reproduce their work.
-Authors are encouraged to include RIIDs where appropriate.
-
-**Ethical approval declarations** (only required where applicable) Any article reporting experiment/s carried out on (i)~live vertebrate (or higher invertebrates), (ii)~humans or (iii)~human samples must include an unambiguous statement within the methods section that meets the following requirements:
-
-1. Approval: a statement which confirms that all experimental protocols were approved by a named institutional and/or licensing committee. Please identify the approving body in the methods section
-
-2. Accordance: a statement explicitly saying that the methods were carried out in accordance with the relevant guidelines and regulations
-
-3. Informed consent (for experiments involving humans or human tissue samples): include a statement confirming that informed consent was obtained from all participants and/or their legal guardian/s
-
-If your manuscript includes potentially identifying patient/participant information, or if it describes human transplantation research, or if it reports results of a clinical trial then  additional information will be required.
-Please visit (<https://www.nature.com/nature-research/editorial-policies>) for Nature Portfolio journals, (<https://www.springer.com/gp/authors-editors/journal-author/journal-author-helpdesk/publishing-ethics/14214>) for Springer Nature journals, or (<https://www.biomedcentral.com/getpublished/editorial-policies\#ethics+and+consent>) for BMC.
-
-# Conclusion {#sec13}
-
-This should state clearly the main conclusions and provide an explanation of the importance and relevance of the study to the field.
-
-Conclusions may be used to restate your hypothesis or research question, restate your major findings, explain the relevance and the added value of your work, highlight any limitations of your study, describe future directions for research and recommendations.
-
-In some disciplines use of Discussion or 'Conclusion' is interchangeable. It is not mandatory to use both. Please refer to Journal-level guidance for any specific requirements.
-
-\backmatter
-
-\bmhead{Acknowledgments}
-
-Acknowledgments are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.
-
-Please refer to Journal-level guidance for any specific requirements.
-
-# Declarations {.unnumbered}
-
-Some journals require declarations to be submitted in a standardised format.
-Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section.
-If yes, your manuscript must contain the following sections under the heading `Declarations':
-
-* Funding
-* Conflict of interest/Competing interests (check journal-specific guidelines for which heading to use)
-* Ethics approval
-* Consent to participate
-* Consent for publication
-* Availability of data and materials
-* Code availability
-* Authors' contributions
-
-\noindent
-If any of the sections are not relevant to your manuscript, please include the heading and write `Not applicable' for that section.
+Thomas Götshi - HAAS.
 

### 2023-09-21 paper/PaperTRA/PaperTRA.tex
move to full paper TRA

@@
-%Version 2.1 April 2023
-% See section 11 of the User Manual for version history
+% This is samplepaper.tex, a sample chapter demonstrating the
+% LLNCS macro package for Springer Computer Science proceedings;
+% Version 2.21 of 2022/01/12
 %
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%                                                                 %%
-%% Please do not use \input{...} to include other tex files.       %%
-%% Submit your LaTeX manuscript as one .tex document.              %%
-%%                                                                 %%
-%% All additional figures and files should be attached             %%
-%% separately and not embedded in the \TeX\ document itself.       %%
-%%                                                                 %%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\documentclass[sn-apa,referee,pdflatex]{sn-jnl}
-
-%%%% Standard Packages
-%%<additional latex packages if required can be included here>
-
-\usepackage{graphicx}%
-\usepackage{multirow}%
-\usepackage{amsmath,amssymb,amsfonts}%
-\usepackage{amsthm}%
-\usepackage{mathrsfs}%
-\usepackage[title]{appendix}%
-\usepackage{xcolor}%
-\usepackage{textcomp}%
-\usepackage{manyfoot}%
-\usepackage{booktabs}%
-\usepackage{algorithm}%
-\usepackage{algorithmicx}%
-\usepackage{algpseudocode}%
-\usepackage{listings}%
-%%%%
-
-%%%%%=============================================================================%%%%
-%%%%  Remarks: This template is provided to aid authors with the preparation
-%%%%  of original research articles intended for submission to journals published
-%%%%  by Springer Nature. The guidance has been prepared in partnership with
-%%%%  production teams to conform to Springer Nature technical requirements.
-%%%%  Editorial and presentation requirements differ among journal portfolios and
-%%%%  research disciplines. You may find sections in this template are irrelevant
-%%%%  to your work and are empowered to omit any such section if allowed by the
-%%%%  journal you intend to submit to. The submission guidelines and policies
-%%%%  of the journal take precedence. A detailed User Manual is available in the
-%%%%  template package for technical guidance.
-%%%%%=============================================================================%%%%
-
-\raggedbottom
+\documentclass[runningheads]{llncs}
+%
+\usepackage[T1]{fontenc}
+% T1 fonts will be used to generate the final print and online PDFs,
+% so please use T1 fonts in your manuscript whenever possible.
+% Other font encondings may result in incorrect characters.
 
@@
 
+% Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\usepackage[hidelinks]{hyperref}
+
+\usepackage{graphicx}
+% Used for displaying a sample figure. If possible, figure files should
+% be included in EPS format.
+%
+% If you use the hyperref package, please uncomment the following two lines
+% to display URLs in blue roman font according to Springer's eBook style:
+\usepackage{hyperref}
+\usepackage{color}
+\renewcommand\UrlFont{\color{blue}\rmfamily}
+
 \begin{document}
 
-\title[]{Estimating the socio-environmental impacts of car substitution
-by bicycle and public transit using open tools}
+\title{Estimating the socio-environmental impacts of car substitution by
+bicycle and public transit using open tools}
+%
+\titlerunning{Estimating the SE impacts of car substitution by bicycle
+and public transit}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{Rosa Félix\inst{1}\orcidID{0000-0002-5642-6006} \and Filipe
+Moura\inst{1}\orcidID{0000-0001-7749-8490} \and Robin
+Lovelace\inst{2}\orcidID{0000-0001-5679-6536}}
 
-%%=============================================================%%
-%% Prefix	-> \pfx{Dr}
-%% GivenName	-> \fnm{Joergen W.}
-%% Particle	-> \spfx{van der} -> surname prefix
-%% FamilyName	-> \sur{Ploeg}
-%% Suffix	-> \sfx{IV}
-%% NatureName	-> \tanm{Poet Laureate} -> Title after name
-%% Degrees	-> \dgr{MSc, PhD}
-%% \author*[1,2]{\pfx{Dr} \fnm{Joergen W.} \spfx{van der} \sur{Ploeg} \sfx{IV} \tanm{Poet Laureate}
-%%                 \dgr{MSc, PhD}}\email{iauthor@gmail.com}
-%%=============================================================%%
+\authorrunning{R. Félix et al.}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
 
-\author*[1]{\fnm{Rosa} \sur{Félix} }\email{\href{mailto:rosamfelix@tecnico.ulisboa.pt}{\nolinkurl{rosamfelix@tecnico.ulisboa.pt}}}
+\institute{CERIS - Instituto Superior Técnico, University of Lisbon. Av
+Rovisco Pais 1049-001 Lisboa, Portugal\\
+\email{\href{mailto:rosamfelix@tecnico.ulisboa.pt}{\nolinkurl{rosamfelix@tecnico.ulisboa.pt}}}\\ \and Institute
+for Transport Studies, University of Leeds. 34-40 University Rd, Leeds
+LS2 9JT, UK}
 
-\author[1]{\fnm{Filipe} \sur{Moura} }\email{\href{mailto:fmoura@tecnico.ulisboa.pt}{\nolinkurl{fmoura@tecnico.ulisboa.pt}}}
+\maketitle              % typeset the header of the contribution
+%
+\begin{abstract}
+In metropolitan areas, car trips can be replaced by a combination of
+public transit and cycling for the first-and-last mile. This paper
+focuses on estimating the potential for cycling + PT as a substitute for
+car trips in the Lisbon metropolitan area and assessing its
+socio-environmental impacts using open data and open source tools. A
+decision support tool that facilitates the design and development of a
+metropolitan cycling network was developed (\emph{biclaR}). A scenario
+of intermodality introduced, and its socio-environmental impacts were
+assessed using the \emph{HEAT for Cycling} and the \emph{HEAT as a
+Service} tools. Additionally, the impacts of shifting car trips to PT
+were estimated and monetized. The results indicate that 20\% of the
+current trips can be made with the bicycle + PT combination. Shifting to
+cycling for the first-and-last mile can reduce annual CO2eq emissions
+from 6,000 tons/day, and the 10-year socio-environmental benefits
+account from €230 million. For the PT leg, the transfer from car results
+in the avoidance of at lest 8,500 tons of CO2eq emissions per year. The
+information on socio-economic benefits can support policymakers in
+prioritizing interventions to reduce the reliance on individual
+motorized transportation and effectively communicate their decisions.
 
-\author[2]{\fnm{Robin} \sur{Lovelace} }\email{\href{mailto:r.lovelace@leeds.ac.uk}{\nolinkurl{r.lovelace@leeds.ac.uk}}}
+\keywords{Active transport \and Intermodality \and First and last
+mile \and Health economic assessment \and Environmental
+impacts \and Open data and methods}
 
-  \affil*[1]{\orgdiv{CERIS - Instituto Superior
-Técnico}, \orgname{University of
-Lisbon}, \orgaddress{\city{Lisbon}, \country{Portugal}, \postcode{1049-001}, \street{Av
-Rovisco Pais 1}}}
-  \affil[2]{\orgdiv{Institute for Transport
-Studies}, \orgname{University of
-Leeds}, \orgaddress{\city{Leeds}, \country{United
-Kingdom}, \postcode{LS2 9JT}, \street{34-40 University Rd}}}
+\end{abstract}
 
-\abstract{In metropolitan areas, car trips can be replaced by a
-combination of public transit and cycling for the first-and-last mile.
-This approach requires interventions to make cycling more appealing, and
-the resulting public investments can have significant social and
-environmental benefits. This paper focuses on estimating the potential
-for cycling + PT as a substitute for car trips in the Lisbon
-metropolitan area and assessing its socio-environmental impacts using
-open data and open source tools.
-
-To achieve the cycling targets set by the Portuguese cycling strategy,
-the Department of Transport introduced \emph{biclaR}, a decision support
-tool that facilitates the design and development of a metropolitan
-cycling network. A scenario of intermodality introduced, and its
-socio-environmental impacts were assessed using the \emph{HEAT for
-Cycling} and the \emph{HEAT as a Service} tools. Additionally, we
-estimated the impacts of shifting car trips to PT and monetize them with
-the EU Guide to cost-benefit analysis.
-
-The results indicate that 20\% of the current trips can be made with the
-bicycle + PT combination, with an additional 12\% of PT trips being
-potentially replaced. Shifting to cycling for the first-and-last mile
-can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the
-10-year socio-environmental benefits account for €230 to €590 million,
-depending on the cycling targets. For the PT leg, the transfer from car
-results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per
-year.
-
-The provided information on socio-economic benefits can support
-policymakers in prioritizing interventions to reduce the reliance on
-individual motorized transportation and effectively communicate their
-decisions.}
-
-\keywords{Active transport, Intermodality, First and last mile, Health
-economic assessment, Environmental impacts, Open data and methods}
-
-\maketitle
-
-\hypertarget{sec1}{%
-\section{Introduction}\label{sec1}}
-
-The Introduction section should explain the background to the study, its
-aims, a summary of the existing literature and why this study was
-necessary.
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
+This research aims to present and discuss the methods used to
+estimate\ldots{}
+
+Propensity to Cycle Tool
+
+adding up an intermodality scenario to estimate cycling potential to
+public transit interfaces, and thus to support planning and prioritize
+investments in the cycling network.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\hypertarget{case-study}{%
+\subsection{Case Study}\label{case-study}}
+
+\hypertarget{modeling-origin-destination-trips}{%
+\subsection{Modeling Origin-Destination
+trips}\label{modeling-origin-destination-trips}}
+
+\hypertarget{modeling-intermodality}{%
+\subsection{Modeling intermodality}\label{modeling-intermodality}}
+
 The intermodality scenario considers trips that can combine PT and
@@
 to unimodal trips without transfers (although they can be included in
-future modeling). Finally, we only included PT modes that can
-practically accommodate bicycles, such as trains, ferries, trams, and
-intermunicipal bus lines with bike racks (Figure 1).
+future modeling). Félix, Lovelace, and Moura (2022)
+
+Finally, we only included PT modes that can practically accommodate
+bicycles, such as trains, ferries, trams, and intermunicipal bus lines
+with bike racks (~\ref{fig:map1}).
+
+\begin{figure}
+\includegraphics{PaperTRA_files/figure-latex/map1-1} \caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
+\end{figure}
 
@@
 
+\hypertarget{assessing-socio-environmental-benefits}{%
+\subsection{Assessing socio-environmental
+benefits}\label{assessing-socio-environmental-benefits}}
+
 Socio-environmental impacts were assessed using the HEAT for Cycling and
@@
 to cost-benefit analysis.
+
+Table~\ref{tab:table} gives a summary of all heading levels.
+
+\begin{table}
+
+\caption{\label{tab:table}Table captions should be placed above the tables.}
+\centering
+\begin{tabular}[t]{r|r}
+\hline
+temperature & pressure\\
+\hline
+0 & 0.0002\\
+\hline
+20 & 0.0012\\
+\hline
+40 & 0.0060\\
+\hline
+60 & 0.0300\\
+\hline
+80 & 0.0900\\
+\hline
+100 & 0.2700\\
+\hline
+\end{tabular}
+\end{table}
+
+\hypertarget{results-and-discussion}{%
+\section{Results and Discussion}\label{results-and-discussion}}
 
@@
 
+Map result (~\ref{fig:map2}).
+
+\begin{figure}
+\includegraphics{PaperTRA_files/figure-latex/map2-1} \caption{Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
+\end{figure}
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
 By making the research process publicly accessible in a code repository,
@@
 
-The Introduction section, of referenced text \citet{bib1} expands on the
-background of the work (some overlap with the Abstract is acceptable).
-The introduction should not include subheadings.
+\hypertarget{acknowledgements}{%
+\subsubsection*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{subsubsection}{Acknowledgements}
 
-Springer Nature does not impose a strict layout as standard however
-authors are advised to check the individual requirements for the journal
-they are planning to submit to as there may be journal-level
-preferences. When preparing your text please also be aware that some
-stylistic choices are not supported in full text XML (publication
-version), including coloured font. These will not be replicated in the
-typeset article if it is accepted.
+Please place your acknowledgments at the end of the paper, preceded by
+an unnumbered run-in heading (i.e.~3rd-level heading).
 
-\hypertarget{sec2}{%
-\section{Methods}\label{sec2}}
+Thomas Götshi - HAAS.
 
-The methods section should include:
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
-\begin{itemize}
-\tightlist
-\item
-  the aim, design and setting of the study
-\item
-  the characteristics of participants or description of materials
-\item
-  a clear description of all processes and methodologies employed.
-  Generic names should generally be used. When proprietary * brands are
-  used in research, include the brand names in parentheses
-\item
-  the type of statistical analysis used, including a power calculation
-  if appropriate
-\item
-  studies involving human participants, data or tissue or animals must
-  include statement on ethics approval and consent
-\end{itemize}
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
+Félix, Rosa, Robin Lovelace, and Filipe Moura. 2022. {``{biclaR -
+Ferramenta de apoio ao planeamento da rede ciclável na área
+metropolitana de Lisboa}.''} {CERIS - Instituto Superior Técnico and
+Transportes Metropolitanos de Lisboa}.
+\url{https://biclar.tmlmobilidade.pt}.
 
-\begin{figure}
-\includegraphics{PaperTRA_files/figure-latex/fig1-1} \caption{This is an example of a caption}\label{fig:fig1}
-\end{figure}
+\leavevmode\vadjust pre{\hypertarget{ref-bib2}{}}%
+Slifka, M. K., and J. L. Whitton. 2000. {``Clinical Implications of
+Dysregulated Cytokine Production.''} \emph{J. {M}ol. {M}ed.} 78: 74--80.
+\url{https://doi.org/10.1007/s001090000086}.
 
-\hypertarget{sec11}{%
-\section{Results and Discussion}\label{sec11}}
+\end{CSLReferences}
 
-Topical subheadings are allowed. Authors must ensure that their Methods
-section includes adequate experimental and characterization data
-necessary for others in the field to reproduce their work. Authors are
-encouraged to include RIIDs where appropriate.
-
-\textbf{Ethical approval declarations} (only required where applicable)
-Any article reporting experiment/s carried out on
-(i)\textasciitilde live vertebrate (or higher invertebrates),
-(ii)\textasciitilde humans or (iii)\textasciitilde human samples must
-include an unambiguous statement within the methods section that meets
-the following requirements:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\item
-  Approval: a statement which confirms that all experimental protocols
-  were approved by a named institutional and/or licensing committee.
-  Please identify the approving body in the methods section
-\item
-  Accordance: a statement explicitly saying that the methods were
-  carried out in accordance with the relevant guidelines and regulations
-\item
-  Informed consent (for experiments involving humans or human tissue
-  samples): include a statement confirming that informed consent was
-  obtained from all participants and/or their legal guardian/s
-\end{enumerate}
-
-If your manuscript includes potentially identifying patient/participant
-information, or if it describes human transplantation research, or if it
-reports results of a clinical trial then additional information will be
-required. Please visit
-(\url{https://www.nature.com/nature-research/editorial-policies}) for
-Nature Portfolio journals,
-(\url{https://www.springer.com/gp/authors-editors/journal-author/journal-author-helpdesk/publishing-ethics/14214})
-for Springer Nature journals, or
-(\url{https://www.biomedcentral.com/getpublished/editorial-policies/\#ethics+and+consent})
-for BMC.
-
-\hypertarget{sec13}{%
-\section{Conclusion}\label{sec13}}
-
-This should state clearly the main conclusions and provide an
-explanation of the importance and relevance of the study to the field.
-
-Conclusions may be used to restate your hypothesis or research question,
-restate your major findings, explain the relevance and the added value
-of your work, highlight any limitations of your study, describe future
-directions for research and recommendations.
-
-In some disciplines use of Discussion or `Conclusion' is
-interchangeable. It is not mandatory to use both. Please refer to
-Journal-level guidance for any specific requirements.
-
-\backmatter
-
-\bmhead{Acknowledgments}
-
-Acknowledgments are not compulsory. Where included they should be brief.
-Grant or contribution numbers may be acknowledged.
-
-Please refer to Journal-level guidance for any specific requirements.
-
-\hypertarget{declarations}{%
-\section*{Declarations}\label{declarations}}
-\addcontentsline{toc}{section}{Declarations}
-
-Some journals require declarations to be submitted in a standardised
-format. Please check the Instructions for Authors of the journal to
-which you are submitting to see if you need to complete this section. If
-yes, your manuscript must contain the following sections under the
-heading `Declarations':
-
-\begin{itemize}
-\tightlist
-\item
-  Funding
-\item
-  Conflict of interest/Competing interests (check journal-specific
-  guidelines for which heading to use)
-\item
-  Ethics approval
-\item
-  Consent to participate
-\item
-  Consent for publication
-\item
-  Availability of data and materials
-\item
-  Code availability
-\item
-  Authors' contributions
-\end{itemize}
-
-\noindent If any of the sections are not relevant to your manuscript,
-please include the heading and write `Not applicable' for that section.
-
-\renewcommand\refname{References}
-\bibliography{bibliography.bib}
+%
+% ---- Bibliography ----
 

### 2023-09-21 paper/PaperTRA_ETRR/PaperTRA.tex
move to full paper TRA

--- 2023-09-16 paper/PaperTRA/skeleton.tex
@@
 
-\documentclass[sn-basic,pdflatex]{sn-jnl}
+\documentclass[sn-apa,referee,pdflatex]{sn-jnl}
 
@@
 
-%% Per the spinger doc, new theorem styles can be included using built in style, 
-%% but it seems the don't work so commented below
-%\theoremstyle{thmstyleone}%
-\newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
-\newtheorem{proposition}[theorem]{Proposition}%
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-
-%% \theoremstyle{thmstyletwo}%
-\theoremstyle{remark}
-\newtheorem{example}{Example}%
-\newtheorem{remark}{Remark}%
-
-%% \theoremstyle{thmstylethree}%
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}%
-
 \raggedbottom
@@
 
-\title[Article Title runing]{Article Title}
+\title[]{Estimating the socio-environmental impacts of car substitution
+by bicycle and public transit using open tools}
 
@@
 
-\author*[1,2]{\pfx{Dr.} \fnm{Leading} \spfx{van} \sur{Author} \sfx{III} \tanm{Poet
-Laureate} \dgr{MSc,
-PhD}}\email{\href{mailto:abc@def}{\nolinkurl{abc@def}}}
+\author*[1]{\fnm{Rosa} \sur{Félix} }\email{\href{mailto:rosamfelix@tecnico.ulisboa.pt}{\nolinkurl{rosamfelix@tecnico.ulisboa.pt}}}
 
-\author[2]{\fnm{Second} \sur{Author} }
+\author[1]{\fnm{Filipe} \sur{Moura} }\email{\href{mailto:fmoura@tecnico.ulisboa.pt}{\nolinkurl{fmoura@tecnico.ulisboa.pt}}}
 
-  \affil*[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\city{City}, \country{Country}, \postcode{100190}, \state{State}, \street{Street}}}
-  \affil*[2]{\orgname{Other Organisation}}
+\author[2]{\fnm{Robin} \sur{Lovelace} }\email{\href{mailto:r.lovelace@leeds.ac.uk}{\nolinkurl{r.lovelace@leeds.ac.uk}}}
 
-\abstract{\textbf{Purpose}: The abstract serves both as a general
-introduction to the topic and as a brief, non-technical summary of the
-main results and their implications. The abstract must not include
-subheadings (unless expressly permitted in the journal's Instructions to
-Authors), equations or citations. As a guide the abstract should not
-exceed 200 words. Most journals do not set a hard limit however authors
-are advised to check the author instructions for the journal they are
-submitting to.
+  \affil*[1]{\orgdiv{CERIS - Instituto Superior
+Técnico}, \orgname{University of
+Lisbon}, \orgaddress{\city{Lisbon}, \country{Portugal}, \postcode{1049-001}, \street{Av
+Rovisco Pais 1}}}
+  \affil[2]{\orgdiv{Institute for Transport
+Studies}, \orgname{University of
+Leeds}, \orgaddress{\city{Leeds}, \country{United
+Kingdom}, \postcode{LS2 9JT}, \street{34-40 University Rd}}}
 
-\textbf{Methods:} The abstract serves both as a general introduction to
-the topic and as a brief, non-technical summary of the main results and
-their implications. The abstract must not include subheadings (unless
-expressly permitted in the journal's Instructions to Authors), equations
-or citations. As a guide the abstract should not exceed 200 words. Most
-journals do not set a hard limit however authors are advised to check
-the author instructions for the journal they are submitting to.
+\abstract{In metropolitan areas, car trips can be replaced by a
+combination of public transit and cycling for the first-and-last mile.
+This approach requires interventions to make cycling more appealing, and
+the resulting public investments can have significant social and
+environmental benefits. This paper focuses on estimating the potential
+for cycling + PT as a substitute for car trips in the Lisbon
+metropolitan area and assessing its socio-environmental impacts using
+open data and open source tools.
 
-\textbf{Results:} The abstract serves both as a general introduction to
-the topic and as a brief, non-technical summary of the main results and
-their implications. The abstract must not include subheadings (unless
-expressly permitted in the journal's Instructions to Authors), equations
-or citations. As a guide the abstract should not exceed 200 words. Most
-journals do not set a hard limit however authors are advised to check
-the author instructions for the journal they are submitting to.
+To achieve the cycling targets set by the Portuguese cycling strategy,
+the Department of Transport introduced \emph{biclaR}, a decision support
+tool that facilitates the design and development of a metropolitan
+cycling network. A scenario of intermodality introduced, and its
+socio-environmental impacts were assessed using the \emph{HEAT for
+Cycling} and the \emph{HEAT as a Service} tools. Additionally, we
+estimated the impacts of shifting car trips to PT and monetize them with
+the EU Guide to cost-benefit analysis.
 
-\textbf{Conclusion:} The abstract serves both as a general introduction
-to the topic and as a brief, non-technical summary of the main results
-and their implications. The abstract must not include subheadings
-(unless expressly permitted in the journal's Instructions to Authors),
-equations or citations. As a guide the abstract should not exceed 200
-words. Most journals do not set a hard limit however authors are advised
-to check the author instructions for the journal they are submitting
-to.\}}
+The results indicate that 20\% of the current trips can be made with the
+bicycle + PT combination, with an additional 12\% of PT trips being
+potentially replaced. Shifting to cycling for the first-and-last mile
+can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the
+10-year socio-environmental benefits account for €230 to €590 million,
+depending on the cycling targets. For the PT leg, the transfer from car
+results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per
+year.
 
-\keywords{key, dictionary, word}
+The provided information on socio-economic benefits can support
+policymakers in prioritizing interventions to reduce the reliance on
+individual motorized transportation and effectively communicate their
+decisions.}
 
-\pacs[JEL Classification]{D8, H51}
-\pacs[MSC Classification]{35A01, 65L10}
+\keywords{Active transport, Intermodality, First and last mile, Health
+economic assessment, Environmental impacts, Open data and methods}
 
@@
 
-The Introduction section, of referenced text \citet{bib1} expands on the
-background of the work (some overlap with the Abstract is acceptable).
-The introduction should not include subheadings.
+The Introduction section should explain the background to the study, its
+aims, a summary of the existing literature and why this study was
+necessary.
 
-Springer Nature does not impose a strict layout as standard however
-authors are advised to check the individual requirements for the journal
-they are planning to submit to as there may be journal-level
-preferences. When preparing your text please also be aware that some
-stylistic choices are not supported in full text XML (publication
-version), including coloured font. These will not be replicated in the
-typeset article if it is accepted.
+In metropolitan areas, car trips can be replaced by a combination of
+public transit (PT) and cycling for the first-and-last mile. This
+approach requires interventions and programs to make bicycling more
+appealing, and the resulting public investments can have significant
+social and environmental benefits. This paper focuses on estimating the
+potential for cycling + PT as a substitute for car trips in the Lisbon
+metropolitan area (LMA) and assessing its socio-environmental impacts
+using open data and open source tools.
+
+According to the latest mobility survey conducted in 2018, the LMA
+registered a total of 5.3 million daily trips, with only 0.5\% by
+bicycle. Car modal share is 58.4\%, while PT accounts for 15.5\%. To
+achieve the cycling targets set by the Portuguese national cycling
+strategy for 2025 and 2030 (4\% and 10\%, respectively), the Department
+of Transport introduced biclaR, a decision support tool that facilitates
+the design and development of a metropolitan cycling network.
+
+The intermodality scenario considers trips that can combine PT and
+cycling for the first-and-last legs. Conservatively, we considered the
+sum of first-and-last legs up to 5 km. Furthermore, we restricted PT use
+to unimodal trips without transfers (although they can be included in
+future modeling). Finally, we only included PT modes that can
+practically accommodate bicycles, such as trains, ferries, trams, and
+intermunicipal bus lines with bike racks (Figure 1).
+
+To obtain reliable results, we used the OpenStreetMap road network and
+GTFS data. The r5r R package estimated the trip duration and distance
+for both the original modes and the bicycle + PT combination, while the
+od jittering R package estimated the OD locations based on a
+centroid-based OD matrix.
+
+Socio-environmental impacts were assessed using the HEAT for Cycling and
+the HEAT as a Service tools, from the WHO. Additionally, we estimate the
+impacts of shifting car trips to PT for the second leg of the journey
+with EMEP/EEA's COPERT methodology and monetize them with the EU Guide
+to cost-benefit analysis.
+
+The results indicate that 20\% of the current trips can be made with the
+bicycle + PT combination, with an additional 12\% of PT trips being
+potentially replaced. Shifting to cycling for the first-and-last mile
+can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the
+10-year socio-environmental benefits account for €230 to €590 million,
+depending on the cycling targets. For the PT leg, the transfer from car
+results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per
+year, or €1.4 to €3.5 million over 10 years, with trains offering the
+greatest potential for substitution (88\%).
+
+By making the research process publicly accessible in a code repository,
+this study enables the replication of similar estimates for
+socio-environmental impacts resulting from a modal shift from cars to
+bicycles + PT in other metropolitan areas. The provided information on
+socio-economic benefits can support policymakers in prioritizing
+interventions to reduce the reliance on individual motorized
+transportation and effectively communicate their decisions.
 
 \hypertarget{sec2}{%
-\section{Results}\label{sec2}}
+\section{Methods}\label{sec2}}
 
-Sample body text. Sample body text. Sample body text. Sample body text.
-Sample body text. Sample body text. Sample body text. Sample body text.
+The methods section should include:
 
-\hypertarget{sec3}{%
-\section{This is an example for first level head---section
-head}\label{sec3}}
-
-\hypertarget{subsec2}{%
-\subsection{This is an example for second level head---subsection
-head}\label{subsec2}}
-
-\hypertarget{subsubsec2}{%
-\subsubsection{This is an example for third level head---subsubsection
-head}\label{subsubsec2}}
-
-Sample body text. Sample body text. Sample body text. Sample body text.
-Sample body text. Sample body text. Sample body text.
-
-\hypertarget{sec4}{%
-\section{Equations}\label{sec4}}
-
-Equations in \LaTeX~can either be inline or on-a-line by itself
-(``display equations''). For inline equations use the \texttt{\$...\$}
-commands. E.g.: The equation \(H\psi = E \psi\) is written via the
-command \texttt{\$H\ \textbackslash{}psi\ =\ E\ \textbackslash{}psi\$}.
-
-For display equations (with auto generated equation numbers) one can use
-the equation or align environments:
-
-\begin{equation}
-\|\tilde{X}(k)\|^2 \leq\frac{\sum\limits_{i=1}^{p}\left\|\tilde{Y}_i(k)\right\|^2+\sum\limits_{j=1}^{q}\left\|\tilde{Z}_j(k)\right\|^2 }{p+q}.\label{eq1}
-\end{equation}
-
-where, \begin{align}
-D_\mu &=  \partial_\mu - ig \frac{\lambda^a}{2} A^a_\mu \nonumber \\
-F^a_{\mu\nu} &= \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^a_\nu \label{eq2}
-\end{align}
-
-Notice the use of \texttt{\textbackslash{}nonumber} in the align
-environment at the end of each line, except the last, so as not to
-produce equation numbers on lines where no equation numbers are
-required. The \texttt{\textbackslash{}label\{\}} command should only be
-used at the last line of an align environment where
-\texttt{\textbackslash{}nonumber} is not used.
-
-\begin{equation}
-Y_\infty = \left( \frac{m}{\textrm{GeV}} \right)^{-3}
-    \left[ 1 + \frac{3 \ln(m/\textrm{GeV})}{15}
-    + \frac{\ln(c_2/5)}{15} \right]
-\end{equation}
-
-The class file also supports the use of
-\texttt{\textbackslash{}mathbb\{\}},
-\texttt{\textbackslash{}mathscr\{\}} and
-\texttt{\textbackslash{}mathcal\{\}} commands. As such
-\texttt{\textbackslash{}mathbb\{R\}},
-\texttt{\textbackslash{}mathscr\{R\}} and
-\texttt{\textbackslash{}mathcal\{R\}} produces \(\mathbb{R}\),
-\(\mathscr{R}\) and \(\mathcal{R}\) respectively (refer
-Subsubsection~\ref{subsubsec2}).
-
-\hypertarget{sec5}{%
-\section{Tables}\label{sec5}}
-
-Tables can be inserted via the normal \texttt{knitr::kable()} function
-or other table-generating packages.
-
-\begin{table}
-
-\caption{\label{tab:tab1}Caption text}
-\centering
-\begin{tabular}[t]{r|r}
-\hline
-temperature & pressure\\
-\hline
-0 & 0.0002\\
-\hline
-20 & 0.0012\\
-\hline
-40 & 0.0060\\
-\hline
-60 & 0.0300\\
-\hline
-80 & 0.0900\\
-\hline
-100 & 0.2700\\
-\hline
-\end{tabular}
-\end{table}
-
-Tables can also be inserted via the normal table and tabular
-environment. To put footnotes inside tables you should use
-\texttt{\textbackslash{}footnotetext{[}{]}\{...\}} tag. The footnote
-appears just below the table itself (refer
-Tables\textasciitilde{}\ref{tab1} and \ref{tab2}). For the corresponding
-footnotemark use \texttt{\textbackslash{}footnotemark{[}...{]}}
-
-\begin{table}[h]
-\caption{Caption text}\label{tab1}%
-\begin{tabular}{@{}llll@{}}
-\toprule
-Column 1 & Column 2  & Column 3 & Column 4\\
-\midrule
-row 1    & data 1   & data 2  & data 3  \\
-row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
-row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
-\botrule
-\end{tabular}
-\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
-\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
-\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
-\end{table}
-
-\noindent The input format for the above table is as follows:
-
-\begin{verbatim}
-\begin{table}[<placement-specifier>]
-\caption{<table-caption>}\label{<table-label>}%
-\begin{tabular}{@{}llll@{}}
-\toprule
-Column 1 & Column 2 & Column 3 & Column 4\\
-\midrule
-row 1 & data 1 & data 2  & data 3 \\
-row 2 & data 4 & data 5\footnotemark[1] & data 6 \\
-row 3 & data 7 & data 8  & data 9\footnotemark[2]\\
-\botrule
-\end{tabular}
-\footnotetext{Source: This is an example of table footnote.
-This is an example of table footnote.}
-\footnotetext[1]{Example for a first table footnote.
-This is an example of table footnote.}
-\footnotetext[2]{Example for a second table footnote.
-This is an example of table footnote.}
-\end{table}
-\end{verbatim}
-
-\begin{table}[h]
-\caption{Example of a lengthy table which is set to full textwidth}\label{tab2}
-\begin{tabular*}{\textwidth}{@{\extracolsep\fill}lcccccc}
-\toprule%
-& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]} & \multicolumn{3}{@{}c@{}}{Element 2\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
-Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
-\midrule
-Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
-Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
-\botrule
-\end{tabular*}
-\footnotetext{Note: This is an example of table footnote.
-This is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
-\footnotetext[1]{Example for a first table footnote.}
-\footnotetext[2]{Example for a second table footnote.}
-\end{table}
-
-\vfill\eject
-
-In case of double column layout, tables which do not fit in single
-column width should be set to full text width. For this, you need to use
-\texttt{\textbackslash{}begin\{table*\}} \texttt{...}
-\texttt{\textbackslash{}end\{table*\}} instead of
-\texttt{\textbackslash{}begin\{table\}} \texttt{...}
-\texttt{\textbackslash{}end\{table\}} environment. Lengthy tables which
-do not fit in textwidth should be set as rotated table. For this, you
-need to use \texttt{\textbackslash{}begin\{sidewaystable\}} \texttt{...}
-\texttt{\textbackslash{}end\{sidewaystable\}} instead of
-\texttt{\textbackslash{}begin\{table*\}} \texttt{...}
-\texttt{\textbackslash{}end\{table*\}} environment. This environment
-puts tables rotated to single column width. For tables rotated to double
-column width, use \texttt{\textbackslash{}begin\{sidewaystable*\}}
-\texttt{...} \texttt{\textbackslash{}end\{sidewaystable*\}}.
-
-\begin{sidewaystable}
-\caption{Tables which are too long to fit, should be written using the "sidewaystable" environment as shown here}\label{tab3}
-\begin{tabular*}{\textheight}{@{\extracolsep\fill}lcccccc}
-\toprule%
-& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]}& \multicolumn{3}{@{}c@{}}{Element\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
-Projectile & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
-\midrule
-Element 3 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
-Element 4 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
-Element 5 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
-Element 6 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
-\botrule
-\end{tabular*}
-\footnotetext{Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
-\footnotetext[1]{This is an example of table footnote.}
-\end{sidewaystable}
-
-\hypertarget{sec6}{%
-\section{Figures}\label{sec6}}
-
-As per the \LaTeX~standards you need to use eps images for
-\LaTeX~compilation and \texttt{pdf/jpg/png} images for \texttt{PDFLaTeX}
-compilation. Use the \texttt{dev} knitr option to use the approrpate
-format. This is one of the major difference between \LaTeX~and
-\texttt{PDFLaTeX}. Each image should be from a single input .eps/vector
-image file. Avoid using subfigures. The command for inserting images for
-\LaTeX~and \texttt{PDFLaTeX} can be generalized. The package used to
-insert images in \texttt{LaTeX/PDFLaTeX} is the graphicx package.
-Figures can be inserted via the normal figure environment as shown in
-the below example:
-
-\begin{figure}
-\includegraphics{skeleton_files/figure-latex/fig1-1} \caption{This is an example of a caption}\label{fig:fig1}
-\end{figure}
-
-\hypertarget{sec7}{%
-\section{Algorithms, Program codes and Listings}\label{sec7}}
-
-Packages \texttt{algorithm}, \texttt{algorithmicx} and
-\texttt{algpseudocode} are used for setting algorithms in \LaTeX~using
-the format:
-
-\begin{verbatim}
-\begin{algorithm}
-\caption{<alg-caption>}\label{<alg-label>}
-\begin{algorithmic}[1]
-. . .
-\end{algorithmic}
-\end{algorithm}
-\end{verbatim}
-
-You may refer above listed package documentations for more details
-before setting \texttt{algorithm} environment. For program codes, the
-``program'' package is required and the command to be used is
-\texttt{\textbackslash{}begin\{program\}} \texttt{...}
-\texttt{\textbackslash{}end\{program\}}. A fast exponentiation
-procedure:
-
-Similarly, for \texttt{listings}, use the \texttt{listings} package.
-\texttt{\textbackslash{}begin\{lstlisting\}} \texttt{...}
-\texttt{\textbackslash{}end\{lstlisting\}} is used to set environments
-similar to \texttt{verbatim} environment. Refer to the
-\texttt{lstlisting} package documentation for more details.
-
-A fast exponentiation procedure:
-
-\lstset{texcl=true,basicstyle=\small\sf,commentstyle=\small\rm,mathescape=true,escapeinside={(*}{*)}}
-\begin{lstlisting}
-begin
-  for $i:=1$ to $10$ step $1$ do
-      expt($2,i$);  
-      newline() od                (*\textrm{Comments will be set flush to the right margin}*)
-where
-proc expt($x,n$) $\equiv$
-  $z:=1$;
-  do if $n=0$ then exit fi;
-     do if odd($n$) then exit fi;                 
-        comment: (*\textrm{This is a comment statement;}*)
-        $n:=n/2$; $x:=x*x$ od;
-     { $n>0$ };
-     $n:=n-1$; $z:=z*x$ od;
-  print($z$). 
-end
-\end{lstlisting}
-
-\begin{algorithm}
-\caption{Calculate $y = x^n$}\label{algo1}
-\begin{algorithmic}[1]
-\Require $n \geq 0 \vee x \neq 0$
-\Ensure $y = x^n$ 
-\State $y \Leftarrow 1$
-\If{$n < 0$}\label{algln2}
-        \State $X \Leftarrow 1 / x$
-        \State $N \Leftarrow -n$
-\Else
-        \State $X \Leftarrow x$
-        \State $N \Leftarrow n$
-\EndIf
-\While{$N \neq 0$}
-        \If{$N$ is even}
-            \State $X \Leftarrow X \times X$
-            \State $N \Leftarrow N / 2$
-        \Else[$N$ is odd]
-            \State $y \Leftarrow y \times X$
-            \State $N \Leftarrow N - 1$
-        \EndIf
-\EndWhile
-\end{algorithmic}
-\end{algorithm}
-
-\begin{minipage}{\hsize}
-
-\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}
-
-\begin{lstlisting}
-for i:=maxint to 0 do begin \{ do nothing \} end; Write(`Case
-insensitive'); Write(`Pascal keywords.');
-
-\end{lstlisting}
-
-\end{minipage}
-
-\hypertarget{sec8}{%
-\section{Cross referencing}\label{sec8}}
-
-Figures and tables are labeled with a prefix (fig or tab, respectively)
-plus the chunk label. Other environments such as equation and align can
-be labelled via the \texttt{\textbackslash{}label\{\#label\}} command
-inside or just below the \texttt{\textbackslash{}caption\{\}} command.
-You can then use the label for cross-reference. As an example, consider
-the chunk label declared for Figure~\ref{fig:fig1} which is fig1. To
-cross-reference it, use the command
-\texttt{Figure\ \textbackslash{}ref\{fig:fig1\}}, for which it comes up
-as ``Figure~\ref{fig:fig1}''.
-
-To reference line numbers in an algorithm, consider the label declared
-for the line number 2 of Algorithm~\ref{algo1} is
-\texttt{\textbackslash{}label\{algln2\}}. To cross-reference it, use the
-command \texttt{\textbackslash{}ref\{algln2\}} for which it comes up as
-line~\ref{algln2} of Algorithm~\ref{algo1}.
-
-\hypertarget{subsec7}{%
-\subsection{Details on reference citations}\label{subsec7}}
-
-For citations of references, use~\citet{bib1} or \citep{bib2}.
-
-\hypertarget{sec10}{%
-\section{Examples for theorem like environments}\label{sec10}}
-
-The documentclass for springer \texttt{sn-jnl.cls} contains 3 styling
-that you can use to set new default for theorems and proofs type
-
-\begin{description}
-\item[\texttt{thmstyleone}]
-Numbered, theorem head in bold font and theorem text in italic style
-\item[\texttt{thmstyletwo}]
-Numbered, theorem head in roman font and theorem text in italic style
-\item[\texttt{thmstylethree}]
-Numbered, theorem head in bold font and theorem text in roman style
-\end{description}
-
-For mathematics journals, theorem styles can be included as shown in the
-following examples.
-
-\begin{theorem}
-Example theorem text. Example theorem text. Example theorem text.
-Example theorem text. Example theorem text. Example theorem text.
-Example theorem text. Example theorem text. Example theorem text.
-Example theorem text. Example theorem text.
-
-\end{theorem}
-
-To add labels and subheadings, use LaTeX notation
-
-\begin{theorem}[Theorem subhead]\label{thm1}
-Example theorem text. Example theorem text. Example theorem text.
-Example theorem text. Example theorem text. Example theorem text.
-Example theorem text. Example theorem text. Example theorem text.
-Example theorem text. Example theorem text.
-
-\end{theorem}
-
-Other environments are proposition, example, remark, definition, proof
-and quote
-
-Sample body text. Sample body text. Sample body text. Sample body text.
-Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{proposition}
-Example proposition text. Example proposition text. Example proposition
-text. Example proposition text. Example proposition text. Example
-proposition text. Example proposition text. Example proposition text.
-Example proposition text. Example proposition text.
-
-\end{proposition}
-
-Sample body text. Sample body text. Sample body text. Sample body text.
-Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{example}
-Phasellus adipiscing semper elit. Proin fermentum massa ac quam. Sed
-diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas
-lacinia. Nam ipsum ligula, eleifend at, accumsan nec, suscipit a, ipsum.
-Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem.
-
-\end{example}
-
-Sample body text. Sample body text. Sample body text. Sample body text.
-Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{remark}
-Phasellus adipiscing semper elit. Proin fermentum massa ac quam. Sed
-diam turpis, molestie vitae, placerat a, molestie nec, leo. Maecenas
-lacinia. Nam ipsum ligula, eleifend at, accumsan nec, suscipit a, ipsum.
-Morbi blandit ligula feugiat magna. Nunc eleifend consequat lorem.
-
-\end{remark}
-
-Sample body text. Sample body text. Sample body text. Sample body text.
-Sample body text. Sample body text. Sample body text. Sample body text.
-
-\begin{definition}[Definition sub head]
-Example definition text. Example definition text. Example definition
-text. Example definition text. Example definition text. Example
-definition text. Example definition text. Example definition text.
-
-\end{definition}
-
-Additionally a predefined ``proof'' environment is available. This
-prints a ``Proof'' head in italic font style and the ``body text'' in
-roman font style with an open square at the end of each proof
-environment.
-
-\begin{proof}
-Example for proof text. Example for proof text. Example for proof text.
-Example for proof text. Example for proof text. Example for proof text.
-Example for proof text. Example for proof text. Example for proof text.
-Example for proof text.
-
-\end{proof}
-
-Sample body text. Sample body text. Sample body text. Sample body text.
-Sample body text. Sample body text. Sample body text. Sample body text.
+\begin{itemize}
+\tightlist
+\item
+  the aim, design and setting of the study
+\item
+  the characteristics of participants or description of materials
+\item
+  a clear description of all processes and methodologies employed.
+  Generic names should generally be used. When proprietary * brands are
+  used in research, include the brand names in parentheses
+\item
+  the type of statistical analysis used, including a power calculation
+  if appropriate
+\item
+  studies involving human participants, data or tissue or animals must
+  include statement on ethics approval and consent
+\end{itemize}
 
 \hypertarget{sec11}{%
-\section{Methods}\label{sec11}}
+\section{Results and Discussion}\label{sec11}}
 
@@
 
-\hypertarget{sec12}{%
-\section{Discussion}\label{sec12}}
-
-Discussions should be brief and focused. In some disciplines use of
-Discussion or `Conclusion' is interchangeable. It is not mandatory to
-use both. Some journals prefer a section `Results and Discussion'
-followed by a section `Conclusion'. Please refer to Journal-level
-guidance for any specific requirements.
-
 \hypertarget{sec13}{%
 \section{Conclusion}\label{sec13}}
+
+This should state clearly the main conclusions and provide an
+explanation of the importance and relevance of the study to the field.
 
@@
 
+This is a test \citet{felix2023} to check if they show up.
+
 \backmatter
-
-\bmhead{Supplementary information}
-
-If your article has accompanying supplementary file/s please state so
-here.
-
-Authors reporting data from electrophoretic gels and blots should supply
-the full unprocessed scans for key as part of their Supplementary
-information. This may be requested by the editorial team/s if it is
-missing.
-
-Please refer to Journal-level guidance for any specific requirements.
 
@@
 
-\begin{flushleft}
-Editorial Policies for:
-
-\noindent Springer journals and proceedings:
-\url{https://www.springer.com/gp/editorial-policies}
-
-\noindent Nature Portfolio journals:
-\url{https://www.nature.com/nature-research/editorial-policies}
-
-\noindent \textit{Scientific Reports}:
-\url{https://www.nature.com/srep/journal-policies/editorial-policies}
-
-\noindent BMC journals:
-\url{https://www.biomedcentral.com/getpublished/editorial-policies}
-
-\end{flushleft}
-
-\begin{appendices}
-
-\hypertarget{secA1}{%
-\section{Section title of first appendix}\label{secA1}}
-
-An appendix contains supplementary information that is not an essential
-part of the text itself but which may be helpful in providing a more
-comprehensive understanding of the research problem or it is information
-that is too cumbersome to be included in the body of the paper.
-
-For submissions to Nature Portfolio Journals please use the heading
-``Extended Data''.
-
-\end{appendices}
-
 \bibliography{bibliography.bib}

### 2023-09-21 paper/PaperTRA/PaperTRA.Rmd
structure paper tra

@@
   The results indicate that 20% of the current trips can be made with the bicycle + PT combination.
-Shifting to cycling for the first-and-last mile can reduce annual CO2eq emissions from 6,000 tons/day, and the 10-year socio-environmental benefits account from €230 million.
-For the PT leg, the transfer from car results in the avoidance of at lest 8,500 tons of CO2eq emissions per year.
+Shifting to cycling for the first-and-last mile can reduce annual CO~2~eq emissions from 6,000 tons/day, and the 10-year socio-environmental benefits account from €230 million.
+For the PT leg, the transfer from car results in the avoidance of at lest 8,500 tons of CO~2~eq emissions per year.
   The information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.
@@
     citation_package: default # or natbib
+# output: word_document
 ---
@@
 
+As características das viagens do IMob, constituem o cenário base deste projeto.
+Este inquérito à mobilidade foi realizado em 2017.
+Apesar de ter sido realizado em período pré-pandemia, este conjunto de dados é a melhor e mais recente informação que temos em termos de mobilidade urbana nas áreas metropolitanas.
+
+Segundo o IMob (2017), das cerca de 5.3 milhões de viagens diárias na área metropolitana de Lisboa, apenas
+25 479 das viagens são realizadas em bicicleta (0,5%), enquanto 3.1 milhões são feitas em automóvel (58.4%),
+1.3 milhões a pé (23.9%), 825 mil em transportes públicos (15.5%) e 96 mil em outros modos (1.8%).
+
+O número de viagens intramunicipais - ou que são realizadas com origem e destino no mesmo município (3.5
+milhões viagens) - é superior ao número de viagens intermunicipais - com origem e destino em municípios
+diferentes (1.8 milhões viagens). O automóvel e os transportes públicos são os modos maioritariamente
+utilizados em viagens intermunicipais. O automóvel é o modo mais utilizado em qualquer tipologia de viagem.
+
+Este é o cenário base, ou de referência, utilizado apenas para comparação com os cenários seguintes - de onde advém o potencial ciclável.
+
 ## Modeling Origin-Destination trips
+
+imob data
+
+Aplicou-se um método de desagregação das origens e destinos das viagens entre o centróide de uma freguesia para o centróide de outra freguesia, para que uma freguesia não esteja representada apenas por um único local de origem e destino das suas viagens (centróides).
+
+Ao agregar as viagens todas em centróides, tornamos o exercício menos realista pois elimina um conjunto
+importante de viagens de curta distância, o que é uma característica de viagens em modos ativos [@ef1, @ref2].
+Para tal, recorreu-se ao método OD Jittering, que utiliza uma desagregação de um único ponto (por exemplo o centróide de uma área) em vários pontos aleatórios da rede viária existente, com
+base no OpenStreetMap, e divide o volume de viagens dessa freguesia pelos pares origem-destino gerados
+aleatoriamente.
+
+Para este projeto, utilizou-se um nível de desagregação máxima de 100 viagens por par O-D.
+Isto significa que, por exemplo, para um par O-D de 2.000 viagens entre duas freguesias ligadas entre os seus centróides, o método jittering dispersa aleatoriamente as 2.000 viagens em 20 pares de 100 viagens entre 20 origens e 20 destinos nas duas freguesias.
+
+A figura 3 ilustra a diferença entre a representação de viagens utilizando o método tradicional de ligação entre um único local entre cada freguesia, e a representação com a aleatorização e desagregação de viagens entre freguesias, para a área metropolitana de Lisboa.
+Figura 3 - Representação de pares OD na área metropolitana de Lisboa entre freguesias, sem jittering (à esquerda), e com jittering (à direita).
+
+O resultado da utilização deste método, em vez do tradicional uso de centróides de grandes áreas, é que torna mais realistas a representação de viagens realizadas, não correspondendo, no entanto, aos pares O-D reais das viagens realizadas (por motivos de proteção de dados definidos pelo INE).
+Para uma melhor compreensão do método utilizado, consultar Jittering: A Computationally Efficient Method for
+Generating Realistic Route Networks from Origin-Destination Data (Lovelace, Félix, and Carlino 2022)
+
+## Modeling routes
+
+A identificação dos percursos entre as origens e destinos (para as linhas de desejo) são um aspeto fundamental da modelação da rede ciclável nesta metodologia, uma vez que o IMob não identifica os percursos realizados por cada viagem, sendo necessário estimá-los.
+A respetiva modelação depende do cenário escolhido e tem em conta toda a restante informação disponível (como os declives, ou localização de interfaces de transporte público), para além da informação viária base.
+
+A partir das linhas de desejo do cenário base, traçam-se os percursos modelados para o modo ciclável com base na rede viária (equivalente à afetação das viagens à rede, 4.º e último passo do modelo clássico dos 4 passos em sistemas de transportes).
+A modelação dos percursos cicláveis depende das variáveis consideradas, assim como das restrições pré-definidas.
+Estas restrições podem privilegiar aspetos como velocidade e volumes baixos, rotas mais diretas, percursos menos declivosos, entre outros, adequados à utilização da bicicleta, sendo que o algoritmo segue uma avaliação que resulta da ponderação das variáveis consideradas.
+Mesmo o percurso menos declivoso, mas com maior volume de tráfego, poderá ser o indicado se a função assim o definir.
+
+O algoritmo de escolha de percurso escolhido foi o do r5r que é um algoritmo que permite muita flexibilidade nas configurações de tipo de percursos estimados, nomeadamente por permitir estimar com probabilidade e incerteza os percursos que usam transportes públicos (considerando os seus horários).
+O r5r permite a identificação de percursos mais diretos ou mais seguros, recorrendo para isso ao nível de stress de tráfego (Level of Traffic Stress, ou LTS), que varia de 1 a 4, sendo o 1 mais tranquilo - correspondendo, por exemplo, a pistas cicláveis fora de estrada, e o 4 menos tranquilo - correspondendo, por exemplo, a percursos que partilham o tráfego motorizado.
+Os percursos foram estimados para o cenário base, para ambos os tipos de rede, direto e seguro, usando para isso o LTS 4 e LTS 3 respetivamente.
+
+O nível de tranquilidade^[see https://www.cyclestreets.net/help/journey/howitworks/#quietness] é outro indicador estimado pelo CycleStreets para cada troço, com base no número de vias, velocidade máxima permitida, arborização, hierarquia e tráfego (quando a informação correspondente existe nas etiquetas do OpenStreetMap).
+É apresentado numa escala de 0 a 100, em que o zero corresponde ao nível menos seguro para se circular de bicicleta, e o 100 ao nível mais seguro e tranquilo - normalmente correspondendo a troços que incluem já infraestrutura ciclável segregada do tráfego motorizado.
+
+Foi utilizado um modelo digital do terreno de 25m de resolução espacial e erro médio de altimetria vertical de 2.12m8 em Portugal, da Agência Espacial Europeia, na sua missão COPERNICUS, para incluir impedâncias 9 de declives no modelo.
+O modelo r5r utiliza a rede viária do OpenStreetMap (extraída a 22 de outubro 2022), e os dados de GTFS criados e validados para o efeito.
+Foi estimado o número de viagens em bicicleta potencial para as duas metas ENMAC (5% e 10%) a partir dos valores de viagens em bicicleta e em automóvel (como condutor e como passageiro) do cenário base de 2017.
+
+Os percursos ou rotas foram depois sobrepostos e agregados por segmentos, juntando a informação de várias rotas que se sobrepõem em certos segmentos (overline).
+Por exemplo, foram somados os volumes de viagens em bicicleta estimados e a transferência do automóvel, e foi feita a média da velocidade automóvel e do nível de tranquilidade, ponderadas pela distância dos segmentos.
 
@@
 
-```{r map1, fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area"}
-plot(pressure)
+Foram recolhidos os dados de transportes públicos da AML, em formato GTFS, tendo sido necessário o trabalho de limpeza, união, e validação prévio.
+Esta informação é indispensável para saber que terminais de transporte público (TP) se ligam a que origens ou destinos, e em que horários, para não se considerar de forma generalista que uma interface se liga a todo o sistema e que pode servir todas as viagens (origens e destinos), o que não seria realista.
+
+```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area"}
+knitr::include_graphics("img/map1.png")
 ```
@@
 
+Para a modelação dos percursos definiu-se como janela temporal viagens até 120 minutos (2h). Para além disso considerou-se como duração máxima 25 minutos em bicicleta.
+
+Map result (\ \ref{fig:map2}).
+
+```{r map2, out.width="80%", fig.cap="Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool)."}
+knitr::include_graphics("img/map2.png")
+```
+
 ## Assessing socio-environmental benefits
 
-Socio-environmental impacts were assessed using the HEAT for Cycling and the HEAT as a Service tools, from the WHO.
+Socio-environmental impacts were assessed using the HEAT for Cycling and the HEAT as a Service^[see https://github.com/HEAT-WHO/HEAT_heatr_api] tools, from the WHO.
 Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA’s COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.
 
-Table\ \ref{tab:table} gives a summary of all heading levels.
+Como referido na introdução, a ferramenta biclaR pretende estimar os impactes sociais e ambientais associados ao potencial ciclável dos vários cenários analisados.
+O horizonte temporal considerado para esta análise foi o curto prazo (i.e., 1 ano) e o longo prazo (i.e., 10 anos).
+Os impactes foram avaliados para cada cenário e para diferentes escalas territoriais:
 
-```{r table}
-knitr::kable(head(pressure), 
-             format = "latex",
-             caption = "Table captions should be placed above the tables.")
-```
+* Desagregado à escala municipal para cada segmento de rede:
+  + Ambientais (emissões de CO~2~eq evitadas);
+  + Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
+* Agregado à escala metropolitana:
+  + Ambientais (emissões de CO~2~eq evitadas);
+  + Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
+  + Ambientais (emissões de CO~2~eq, CO, PM10, NOx, e VOC11, evitadas por substituição dos modos motorizados, considerando as emissões geradas na transferência para transportes públicos).
+
+Para todos os cenários, recorreu-se à ferramenta HEAT for cycling, da Organização Mundial de Saúde, para estimativa dos impactes sociais e ambientais da transferência de viagens em automóvel para viagens em bicicleta, nas componentes de: a) Social - Saúde, Atividade Física, Exposição a poluição atmosférica, Exposição ao risco de sinistralidade rodoviária; b) Ambiental - Emissão de gases CO~2~eq.
+
+A estimativa dos impactes ambientais resulta em toneladas de CO~2~eq, e a estimativa dos impactes sociais resulta em mortalidade prematura evitada.
+Ambas as unidades são por fim monetizadas em €, segundo os valores da literatura utilizada em estudos semelhantes.
+Para além dos impactes sociais e ambientais resultantes da transferência do automóvel para a bicicleta (na primeira e na última parte da viagem, de e para a interface de TP), estimou-se também o impacte ambiental adicional, resultante da transferência do automóvel para os vários transportes públicos (na segunda etapa da viagem, entre as interfaces).
+Como tal, foi necessário caracterizar o universo dos modos motorizados a serem considerados para os cálculos dos respetivos fatores de emissões de gases poluentes e atmosféricos.
+
+Os fatores de consumo e de emissão dos automóveis e transportes públicos deve ser tido em conta relativamente ao número de passageiros transportados (por passageiro.km) e não ao veículo (que seria veículo.km).
+No caso do automóvel, para contabilizar a emissão de gases evitados pela transferência de viagens em automóvel para os transportes públicos, teve-se em conta o valor de referência de 1.61 passageiros/automóvel reportados pelo Inquérito à Mobilidade (INE 2017).
+
+As emissões evitadas por cada viagem em transportes públicos que substitui o automóvel correspondem às emissões equivalentes de um automóvel com características correspondentes à média da frota em circulação em Lisboa, para 2 tipos de combustível: gasolina e gasóleo.
+Utilizou-se a metodologia e valores de referência do software COPERT (Ntziachristos and Samaras 2020), v5.0, da Agência Europeia do Ambiente, para um nível de detalhe 3 (Tier 3) na estimativa de consumo de emissões para o automóvel, nestes dois tipos de combustível.
+Optou-se por usar um veículo de dimensão familiar, norma EURO, e tipo de combustível gasolina ou diesel.
+Considerou-se que as viagens foram todas realizadas em condições urbanas (com as respetivas implicações no regime médio de condução) e a uma velocidade média de 15km/h, nos períodos de hora de ponta.
+Uma vez que a distância média percorrida por viagem influencia o nível de sobreconsumo e emissões decorrentes da operação do motor a frio – ou seja, distâncias maiores diluem a importância deste sobreconsumo face ao fator de consumo com o motor a operar a quente, foram estimados os consumos para diferentes gamas de viagem, em intervalos de 500 metros.
+As emissões são estimadas para os seguintes poluentes atmosféricos: CO (monóxido de carbono), NO X (Óxidos de Azoto), COV (Compostos Orgânicos Voláteis), PM (material particulado).
+Também é estimada as emissões dos principais gases com efeito de estufa: CO2 (dióxido de carbono); CH4 (metano) e N2O (Óxido nitroso), assim como o CO~2~eq equivalente.
+
+Relativamente aos transportes públicos, recorreu-se aos fatores de emissão reportados nos relatórios de sustentabilidade dos respetivos operadores (Carris 2020; Metropolitano de Lisboa 2020; CP 2020, Grupo Transtejo 2014).
+
+a conversão das emissões evitadas em perda de bem-estar evitado, através da respetiva valorização monetária.
+a partir dos valores de referência para os vários gases contabilizados, atualizados para 2022 15, com base nas fontes: Bickel et al.
+(2006), Nash and others (2003), Sartori et al.
+(2014).
 
@@
 
+A Tabela 4 apresenta o número de viagens diárias do cenário base e potenciais, passíveis de serem realizadas
+até 5 km em bicicleta em complemento com TP, por tipo de percurso. Para as redes segura e direta, as viagens
+do cenário base até 5 km em bicicleta + TP correspondem a 20% do total de viagens reportadas no inquérito.
+
+Este cenário valoriza a utilização da bicicleta como complemento ao transporte público, com potencial de
+aumentar as viagens em TP realizadas na área metropolitana de Lisboa em até 12% (acrescentando às 825 mil
+do IMob 2017).
+Estes resultados sugerem que o potencial de transferência de viagens em automóvel para bicicleta + TP poderá
+ser quase ou tão importante quanto a transferência para viagens utilizando apenas a bicicleta.
+
+Para este cenário foram estimados os impactes ambientais de transferência do automóvel para os transportes
+públicos, para além dos impactes ambientais e sociais da transferência para a bicicleta
+
+Foram então calculadas as emissões dos poluentes para cada viagem substituída, tendo em conta o modo de transporte, distâncias e velocidade.
+A tabela 11 apresenta a estimativa anual de gases emitidos e evitados por transferência do automóvel para os outros modos de transporte público, como exemplo para o cenário com a meta de 4%, usando a rede ciclável “direta”.
+
+[Tabela 11]
+
+Para esse cenário com essas características, é estimada uma poupança de 8 702 toneladas de CO~2~eq por via da transferência de viagens motorizadas com combustíveis fósseis e eletricidade, numa perspetiva de ciclo de vida da produção de eletricidade.
+
+A valorização monetária das emissões (em toneladas) é apresentada na tabela seguinte, para o cenário 3 (apenas a segunda parte da viagem, em transportes públicos), com as metas de 4% e 10%, e usando as redes cicláveis “direta” e “segura”, para 365 dias (1 ano).
+
+[Tabela 13]
+
+<!-- Meta ENMAC	Cenário	Tipo de Percurso	Total Viagens /dia	Viagens Bicicleta Base	Viagens Bicicleta Potencial	CO~2~eq evitado (ton/ano)	Benefícios estimados em 10 anos (milhares €) -->
+<!-- 4%	intermodal até 5km	seguro	1 077 028	4 624	40 770	5 920	230 270 -->
+<!-- 4%	intermodal até 5km	direto	1 001 761	4 547	37 889	6 011	223 720 -->
+<!-- 10%	intermodal até 5km	seguro	1 077 028	4 624	104 647	15 192	591 790 -->
+<!-- 10%	intermodal até 5km	direto	1 001 761	4 547	97 218	15 414	574 200 -->
+
+```{r summary1, out.width="100%", fig.cap="Summary of the cycling potencial of intermodality scenario."}
+knitr::include_graphics("img/table1.png")
+```
+
+```{r summary2, out.width="100%", fig.cap="Potencial de transferência estimado para cada modo de transporte público, bem como as estimativas de consumo de CO~2~eq evitado anualmente por transferência do automóvel."}
+knitr::include_graphics("img/table2.png")
+```
+
+<!-- Tipo de Percurso	Meta ENMAC	Total	Barco	Comboio	Autocarro	Metro superfície	CO~2~eq	CO	PM10	NOx	VOC	Valorização monetária (€) -->
+<!-- seguro	4%	192 214	5 167	169 984	4 163	12 900	8 593	17	1.9	27	0.8	1 424 620 -->
+<!-- seguro	10%	224 152	5 594	197 857	5 042	15 659	20 627	42	4.6	65	2.0	3 430 552 -->
+<!-- direto	4%	189 846	5 795	171 534	4 879	7 639	8 702	18	2.0	28	0.8	1 452 820 -->
+<!-- direto	10%	219 511	6 263	198 373	5 806	9 069	20 793	42	4.7	66	1.9	3 487 282 -->
+
+Em suma, o impacte socioeconómico das emissões de gases poluentes e de gases de efeito estufa pode ser valorizado numa poupança potencial de entre 1.4 a 3.5 milhões €, a acrescentar aos impactes estimados na secção anterior.
+
 The results indicate that 20% of the current trips can be made with the bicycle + PT combination, with an additional 12% of PT trips being potentially replaced.
-Shifting to cycling for the first-and-last mile can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the 10-year socio-environmental benefits account for €230 to €590 million, depending on the cycling targets.
-For the PT leg, the transfer from car results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per year, or €1.4 to €3.5 million over 10 years, with trains offering the greatest potential for substitution (88%).
-
-Map result (\ \ref{fig:map2}).
-
-```{r map2, fig.cap="Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool)."}
-plot(pressure)
-```
+Shifting to cycling for the first-and-last mile can reduce annual CO~2~eq emissions by 6,000 to 15,000 tons/day, and the 10-year socio-environmental benefits account for €230 to €590 million, depending on the cycling targets.
+For the PT leg, the transfer from car results in the avoidance of 8,500 to 20,800 tons of CO~2~eq emissions per year, or €1.4 to €3.5 million over 10 years, with trains offering the greatest potential for substitution (88%).
 
@@
 
-# References {.unnumbered}
+# References (test) {.unnumbered}

### 2023-09-21 paper/PaperTRA/PaperTRA.tex
structure paper tra

@@
 current trips can be made with the bicycle + PT combination. Shifting to
-cycling for the first-and-last mile can reduce annual CO2eq emissions
-from 6,000 tons/day, and the 10-year socio-environmental benefits
-account from €230 million. For the PT leg, the transfer from car results
-in the avoidance of at lest 8,500 tons of CO2eq emissions per year. The
-information on socio-economic benefits can support policymakers in
-prioritizing interventions to reduce the reliance on individual
-motorized transportation and effectively communicate their decisions.
+cycling for the first-and-last mile can reduce annual
+CO\textsubscript{2}eq emissions from 6,000 tons/day, and the 10-year
+socio-environmental benefits account from €230 million. For the PT leg,
+the transfer from car results in the avoidance of at lest 8,500 tons of
+CO\textsubscript{2}eq emissions per year. The information on
+socio-economic benefits can support policymakers in prioritizing
+interventions to reduce the reliance on individual motorized
+transportation and effectively communicate their decisions.
 
@@
 \section{Introduction}\label{introduction}}
+
+\textbf{full paper}: 4-6 pages in length (typically up to 3,000 words).
 
@@
 
+As características das viagens do IMob, constituem o cenário base deste
+projeto. Este inquérito à mobilidade foi realizado em 2017. Apesar de
+ter sido realizado em período pré-pandemia, este conjunto de dados é a
+melhor e mais recente informação que temos em termos de mobilidade
+urbana nas áreas metropolitanas.
+
+Segundo o IMob (2017), das cerca de 5.3 milhões de viagens diárias na
+área metropolitana de Lisboa, apenas 25 479 das viagens são realizadas
+em bicicleta (0,5\%), enquanto 3.1 milhões são feitas em automóvel
+(58.4\%), 1.3 milhões a pé (23.9\%), 825 mil em transportes públicos
+(15.5\%) e 96 mil em outros modos (1.8\%).
+
+O número de viagens intramunicipais - ou que são realizadas com origem e
+destino no mesmo município (3.5 milhões viagens) - é superior ao número
+de viagens intermunicipais - com origem e destino em municípios
+diferentes (1.8 milhões viagens). O automóvel e os transportes públicos
+são os modos maioritariamente utilizados em viagens intermunicipais. O
+automóvel é o modo mais utilizado em qualquer tipologia de viagem.
+
+Este é o cenário base, ou de referência, utilizado apenas para
+comparação com os cenários seguintes - de onde advém o potencial
+ciclável.
+
 \hypertarget{modeling-origin-destination-trips}{%
@@
 trips}\label{modeling-origin-destination-trips}}
+
+imob data
+
+Aplicou-se um método de desagregação das origens e destinos das viagens
+entre o centróide de uma freguesia para o centróide de outra freguesia,
+para que uma freguesia não esteja representada apenas por um único local
+de origem e destino das suas viagens (centróides).
+
+Ao agregar as viagens todas em centróides, tornamos o exercício menos
+realista pois elimina um conjunto importante de viagens de curta
+distância, o que é uma característica de viagens em modos ativos
+(\textbf{ref2?}). Para tal, recorreu-se ao método OD Jittering, que
+utiliza uma desagregação de um único ponto (por exemplo o centróide de
+uma área) em vários pontos aleatórios da rede viária existente, com base
+no OpenStreetMap, e divide o volume de viagens dessa freguesia pelos
+pares origem-destino gerados aleatoriamente.
+
+Para este projeto, utilizou-se um nível de desagregação máxima de 100
+viagens por par O-D. Isto significa que, por exemplo, para um par O-D de
+2.000 viagens entre duas freguesias ligadas entre os seus centróides, o
+método jittering dispersa aleatoriamente as 2.000 viagens em 20 pares de
+100 viagens entre 20 origens e 20 destinos nas duas freguesias.
+
+A figura 3 ilustra a diferença entre a representação de viagens
+utilizando o método tradicional de ligação entre um único local entre
+cada freguesia, e a representação com a aleatorização e desagregação de
+viagens entre freguesias, para a área metropolitana de Lisboa. Figura 3
+- Representação de pares OD na área metropolitana de Lisboa entre
+freguesias, sem jittering (à esquerda), e com jittering (à direita).
+
+O resultado da utilização deste método, em vez do tradicional uso de
+centróides de grandes áreas, é que torna mais realistas a representação
+de viagens realizadas, não correspondendo, no entanto, aos pares O-D
+reais das viagens realizadas (por motivos de proteção de dados definidos
+pelo INE). Para uma melhor compreensão do método utilizado, consultar
+Jittering: A Computationally Efficient Method for Generating Realistic
+Route Networks from Origin-Destination Data (Lovelace, Félix, and
+Carlino 2022)
+
+\hypertarget{modeling-routes}{%
+\subsection{Modeling routes}\label{modeling-routes}}
+
+A identificação dos percursos entre as origens e destinos (para as
+linhas de desejo) são um aspeto fundamental da modelação da rede
+ciclável nesta metodologia, uma vez que o IMob não identifica os
+percursos realizados por cada viagem, sendo necessário estimá-los. A
+respetiva modelação depende do cenário escolhido e tem em conta toda a
+restante informação disponível (como os declives, ou localização de
+interfaces de transporte público), para além da informação viária base.
+
+A partir das linhas de desejo do cenário base, traçam-se os percursos
+modelados para o modo ciclável com base na rede viária (equivalente à
+afetação das viagens à rede, 4.º e último passo do modelo clássico dos 4
+passos em sistemas de transportes). A modelação dos percursos cicláveis
+depende das variáveis consideradas, assim como das restrições
+pré-definidas. Estas restrições podem privilegiar aspetos como
+velocidade e volumes baixos, rotas mais diretas, percursos menos
+declivosos, entre outros, adequados à utilização da bicicleta, sendo que
+o algoritmo segue uma avaliação que resulta da ponderação das variáveis
+consideradas. Mesmo o percurso menos declivoso, mas com maior volume de
+tráfego, poderá ser o indicado se a função assim o definir.
+
+O algoritmo de escolha de percurso escolhido foi o do r5r que é um
+algoritmo que permite muita flexibilidade nas configurações de tipo de
+percursos estimados, nomeadamente por permitir estimar com probabilidade
+e incerteza os percursos que usam transportes públicos (considerando os
+seus horários). O r5r permite a identificação de percursos mais diretos
+ou mais seguros, recorrendo para isso ao nível de stress de tráfego
+(Level of Traffic Stress, ou LTS), que varia de 1 a 4, sendo o 1 mais
+tranquilo - correspondendo, por exemplo, a pistas cicláveis fora de
+estrada, e o 4 menos tranquilo - correspondendo, por exemplo, a
+percursos que partilham o tráfego motorizado. Os percursos foram
+estimados para o cenário base, para ambos os tipos de rede, direto e
+seguro, usando para isso o LTS 4 e LTS 3 respetivamente.
+
+O nível de tranquilidade\footnote{see
+  \url{https://www.cyclestreets.net/help/journey/howitworks/\#quietness}}
+é outro indicador estimado pelo CycleStreets para cada troço, com base
+no número de vias, velocidade máxima permitida, arborização, hierarquia
+e tráfego (quando a informação correspondente existe nas etiquetas do
+OpenStreetMap). É apresentado numa escala de 0 a 100, em que o zero
+corresponde ao nível menos seguro para se circular de bicicleta, e o 100
+ao nível mais seguro e tranquilo - normalmente correspondendo a troços
+que incluem já infraestrutura ciclável segregada do tráfego motorizado.
+
+Foi utilizado um modelo digital do terreno de 25m de resolução espacial
+e erro médio de altimetria vertical de 2.12m8 em Portugal, da Agência
+Espacial Europeia, na sua missão COPERNICUS, para incluir impedâncias 9
+de declives no modelo. O modelo r5r utiliza a rede viária do
+OpenStreetMap (extraída a 22 de outubro 2022), e os dados de GTFS
+criados e validados para o efeito. Foi estimado o número de viagens em
+bicicleta potencial para as duas metas ENMAC (5\% e 10\%) a partir dos
+valores de viagens em bicicleta e em automóvel (como condutor e como
+passageiro) do cenário base de 2017.
+
+Os percursos ou rotas foram depois sobrepostos e agregados por
+segmentos, juntando a informação de várias rotas que se sobrepõem em
+certos segmentos (overline). Por exemplo, foram somados os volumes de
+viagens em bicicleta estimados e a transferência do automóvel, e foi
+feita a média da velocidade automóvel e do nível de tranquilidade,
+ponderadas pela distância dos segmentos.
 
@@
 
+Foram recolhidos os dados de transportes públicos da AML, em formato
+GTFS, tendo sido necessário o trabalho de limpeza, união, e validação
+prévio. Esta informação é indispensável para saber que terminais de
+transporte público (TP) se ligam a que origens ou destinos, e em que
+horários, para não se considerar de forma generalista que uma interface
+se liga a todo o sistema e que pode servir todas as viagens (origens e
+destinos), o que não seria realista.
+
 \begin{figure}
-\includegraphics{PaperTRA_files/figure-latex/map1-1} \caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
+\includegraphics[width=0.6\linewidth,]{img/map1} \caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
 \end{figure}
@@
 
+Para a modelação dos percursos definiu-se como janela temporal viagens
+até 120 minutos (2h). Para além disso considerou-se como duração máxima
+25 minutos em bicicleta.
+
+Map result (~\ref{fig:map2}).
+
+\begin{figure}
+\includegraphics[width=0.8\linewidth,]{img/map2} \caption{Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
+\end{figure}
+
 \hypertarget{assessing-socio-environmental-benefits}{%
@@
 Socio-environmental impacts were assessed using the HEAT for Cycling and
-the HEAT as a Service tools, from the WHO. Additionally, we estimate the
-impacts of shifting car trips to PT for the second leg of the journey
-with EMEP/EEA's COPERT methodology and monetize them with the EU Guide
-to cost-benefit analysis.
+the HEAT as a Service\footnote{see
+  \url{https://github.com/HEAT-WHO/HEAT_heatr_api}} tools, from the WHO.
+Additionally, we estimate the impacts of shifting car trips to PT for
+the second leg of the journey with EMEP/EEA's COPERT methodology and
+monetize them with the EU Guide to cost-benefit analysis.
 
-Table~\ref{tab:table} gives a summary of all heading levels.
+Como referido na introdução, a ferramenta biclaR pretende estimar os
+impactes sociais e ambientais associados ao potencial ciclável dos
+vários cenários analisados. O horizonte temporal considerado para esta
+análise foi o curto prazo (i.e., 1 ano) e o longo prazo (i.e., 10 anos).
+Os impactes foram avaliados para cada cenário e para diferentes escalas
+territoriais:
 
-\begin{table}
+\begin{itemize}
+\tightlist
+\item
+  Desagregado à escala municipal para cada segmento de rede:
 
-\caption{\label{tab:table}Table captions should be placed above the tables.}
-\centering
-\begin{tabular}[t]{r|r}
-\hline
-temperature & pressure\\
-\hline
-0 & 0.0002\\
-\hline
-20 & 0.0012\\
-\hline
-40 & 0.0060\\
-\hline
-60 & 0.0300\\
-\hline
-80 & 0.0900\\
-\hline
-100 & 0.2700\\
-\hline
-\end{tabular}
-\end{table}
+  \begin{itemize}
+  \tightlist
+  \item
+    Ambientais (emissões de CO\textsubscript{2}eq evitadas);
+  \item
+    Balanço monetarizado dos impactes socioeconómicos totais
+    (considerando o balanço dos impactes ambientais e sociais);
+  \end{itemize}
+\item
+  Agregado à escala metropolitana:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Ambientais (emissões de CO\textsubscript{2}eq evitadas);
+  \item
+    Balanço monetarizado dos impactes socioeconómicos totais
+    (considerando o balanço dos impactes ambientais e sociais);
+  \item
+    Ambientais (emissões de CO\textsubscript{2}eq, CO, PM10, NOx, e
+    VOC11, evitadas por substituição dos modos motorizados, considerando
+    as emissões geradas na transferência para transportes públicos).
+  \end{itemize}
+\end{itemize}
+
+Para todos os cenários, recorreu-se à ferramenta HEAT for cycling, da
+Organização Mundial de Saúde, para estimativa dos impactes sociais e
+ambientais da transferência de viagens em automóvel para viagens em
+bicicleta, nas componentes de: a) Social - Saúde, Atividade Física,
+Exposição a poluição atmosférica, Exposição ao risco de sinistralidade
+rodoviária; b) Ambiental - Emissão de gases CO\textsubscript{2}eq.
+
+A estimativa dos impactes ambientais resulta em toneladas de
+CO\textsubscript{2}eq, e a estimativa dos impactes sociais resulta em
+mortalidade prematura evitada. Ambas as unidades são por fim monetizadas
+em €, segundo os valores da literatura utilizada em estudos semelhantes.
+Para além dos impactes sociais e ambientais resultantes da transferência
+do automóvel para a bicicleta (na primeira e na última parte da viagem,
+de e para a interface de TP), estimou-se também o impacte ambiental
+adicional, resultante da transferência do automóvel para os vários
+transportes públicos (na segunda etapa da viagem, entre as interfaces).
+Como tal, foi necessário caracterizar o universo dos modos motorizados a
+serem considerados para os cálculos dos respetivos fatores de emissões
+de gases poluentes e atmosféricos.
+
+Os fatores de consumo e de emissão dos automóveis e transportes públicos
+deve ser tido em conta relativamente ao número de passageiros
+transportados (por passageiro.km) e não ao veículo (que seria
+veículo.km). No caso do automóvel, para contabilizar a emissão de gases
+evitados pela transferência de viagens em automóvel para os transportes
+públicos, teve-se em conta o valor de referência de 1.61
+passageiros/automóvel reportados pelo Inquérito à Mobilidade (INE 2017).
+
+As emissões evitadas por cada viagem em transportes públicos que
+substitui o automóvel correspondem às emissões equivalentes de um
+automóvel com características correspondentes à média da frota em
+circulação em Lisboa, para 2 tipos de combustível: gasolina e gasóleo.
+Utilizou-se a metodologia e valores de referência do software COPERT
+(Ntziachristos and Samaras 2020), v5.0, da Agência Europeia do Ambiente,
+para um nível de detalhe 3 (Tier 3) na estimativa de consumo de emissões
+para o automóvel, nestes dois tipos de combustível. Optou-se por usar um
+veículo de dimensão familiar, norma EURO, e tipo de combustível gasolina
+ou diesel. Considerou-se que as viagens foram todas realizadas em
+condições urbanas (com as respetivas implicações no regime médio de
+condução) e a uma velocidade média de 15km/h, nos períodos de hora de
+ponta. Uma vez que a distância média percorrida por viagem influencia o
+nível de sobreconsumo e emissões decorrentes da operação do motor a frio
+-- ou seja, distâncias maiores diluem a importância deste sobreconsumo
+face ao fator de consumo com o motor a operar a quente, foram estimados
+os consumos para diferentes gamas de viagem, em intervalos de 500
+metros. As emissões são estimadas para os seguintes poluentes
+atmosféricos: CO (monóxido de carbono), NO X (Óxidos de Azoto), COV
+(Compostos Orgânicos Voláteis), PM (material particulado). Também é
+estimada as emissões dos principais gases com efeito de estufa: CO2
+(dióxido de carbono); CH4 (metano) e N2O (Óxido nitroso), assim como o
+CO\textsubscript{2}eq equivalente.
+
+Relativamente aos transportes públicos, recorreu-se aos fatores de
+emissão reportados nos relatórios de sustentabilidade dos respetivos
+operadores (Carris 2020; Metropolitano de Lisboa 2020; CP 2020, Grupo
+Transtejo 2014).
+
+a conversão das emissões evitadas em perda de bem-estar evitado, através
+da respetiva valorização monetária. a partir dos valores de referência
+para os vários gases contabilizados, atualizados para 2022 15, com base
+nas fontes: Bickel et al.~(2006), Nash and others (2003), Sartori et
+al.~(2014).
 
@@
 
+A Tabela 4 apresenta o número de viagens diárias do cenário base e
+potenciais, passíveis de serem realizadas até 5 km em bicicleta em
+complemento com TP, por tipo de percurso. Para as redes segura e direta,
+as viagens do cenário base até 5 km em bicicleta + TP correspondem a
+20\% do total de viagens reportadas no inquérito.
+
+Este cenário valoriza a utilização da bicicleta como complemento ao
+transporte público, com potencial de aumentar as viagens em TP
+realizadas na área metropolitana de Lisboa em até 12\% (acrescentando às
+825 mil do IMob 2017). Estes resultados sugerem que o potencial de
+transferência de viagens em automóvel para bicicleta + TP poderá ser
+quase ou tão importante quanto a transferência para viagens utilizando
+apenas a bicicleta.
+
+Para este cenário foram estimados os impactes ambientais de
+transferência do automóvel para os transportes públicos, para além dos
+impactes ambientais e sociais da transferência para a bicicleta
+
+Foram então calculadas as emissões dos poluentes para cada viagem
+substituída, tendo em conta o modo de transporte, distâncias e
+velocidade. A tabela 11 apresenta a estimativa anual de gases emitidos e
+evitados por transferência do automóvel para os outros modos de
+transporte público, como exemplo para o cenário com a meta de 4\%,
+usando a rede ciclável ``direta''.
+
+{[}Tabela 11{]}
+
+Para esse cenário com essas características, é estimada uma poupança de
+8 702 toneladas de CO\textsubscript{2}eq por via da transferência de
+viagens motorizadas com combustíveis fósseis e eletricidade, numa
+perspetiva de ciclo de vida da produção de eletricidade.
+
+A valorização monetária das emissões (em toneladas) é apresentada na
+tabela seguinte, para o cenário 3 (apenas a segunda parte da viagem, em
+transportes públicos), com as metas de 4\% e 10\%, e usando as redes
+cicláveis ``direta'' e ``segura'', para 365 dias (1 ano).
+
+{[}Tabela 13{]}
+
+\begin{figure}
+\includegraphics[width=1\linewidth,]{img/table1} \caption{Summary of the cycling potencial of intermodality scenario.}\label{fig:summary1}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=1\linewidth,]{img/table2} \caption{Potencial de transferência estimado para cada modo de transporte público, bem como as estimativas de consumo de CO~2~eq evitado anualmente por transferência do automóvel.}\label{fig:summary2}
+\end{figure}
+
+Em suma, o impacte socioeconómico das emissões de gases poluentes e de
+gases de efeito estufa pode ser valorizado numa poupança potencial de
+entre 1.4 a 3.5 milhões €, a acrescentar aos impactes estimados na
+secção anterior.
+
 The results indicate that 20\% of the current trips can be made with the
@@
 potentially replaced. Shifting to cycling for the first-and-last mile
-can reduce annual CO2eq emissions by 6,000 to 15,000 tons/day, and the
-10-year socio-environmental benefits account for €230 to €590 million,
-depending on the cycling targets. For the PT leg, the transfer from car
-results in the avoidance of 8,500 to 20,800 tons of CO2eq emissions per
-year, or €1.4 to €3.5 million over 10 years, with trains offering the
-greatest potential for substitution (88\%).
-
-Map result (~\ref{fig:map2}).
-
-\begin{figure}
-\includegraphics{PaperTRA_files/figure-latex/map2-1} \caption{Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
-\end{figure}
+can reduce annual CO\textsubscript{2}eq emissions by 6,000 to 15,000
+tons/day, and the 10-year socio-environmental benefits account for €230
+to €590 million, depending on the cycling targets. For the PT leg, the
+transfer from car results in the avoidance of 8,500 to 20,800 tons of
+CO\textsubscript{2}eq emissions per year, or €1.4 to €3.5 million over
+10 years, with trains offering the greatest potential for substitution
+(88\%).
 

### 2023-09-26 code/biclarwww/index.Rmd
add video separator

@@
 
+## Videos
+
+Vídeos auxiliares sobre a metodologia adotada no biclaR, e sobre como se usa esta ferramenta.
+
+### Parte I: Metodologia
+
+<div align="center"><iframe width="1000" height="563" src="https://www.youtube.com/embed/SsvoYPZwrFs" title="YouTube video player" frameborder="0" data-external="1" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe></div>
+
+### Parte II: Como usar o biclaR?
+
+(*comming soon!*)
+
+<!-- ------------------------------------------------------------------------ -->
+
 ## FAQ

### 2023-09-26 paper/PaperTRA/PaperTRA.Rmd
paper TRA - translt
stopped at  section 2.4 (will resume soon)

@Robinlovelace can you take a look at the paragraph in the introduction about PCT?

@@
   In metropolitan areas, car trips can be replaced by a combination of public transit and cycling for the first-and-last mile.
-This paper focuses on estimating the potential for cycling + PT as a substitute for car trips in the Lisbon metropolitan area and assessing its socio-environmental impacts using open data and open source tools.
+This paper estimates the potential for cycling + public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
   A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_).
 A scenario of intermodality introduced, and its socio-environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
-Additionally, the impacts of shifting car trips to PT were estimated and monetized.
+The impacts of shifting car trips to PT were also estimated and monetized.
   The results indicate that 20% of the current trips can be made with the bicycle + PT combination.
 Shifting to cycling for the first-and-last mile can reduce annual CO~2~eq emissions from 6,000 tons/day, and the 10-year socio-environmental benefits account from €230 million.
-For the PT leg, the transfer from car results in the avoidance of at lest 8,500 tons of CO~2~eq emissions per year.
+For the PT leg, the transfer from car avoids of at least 8,500 tons of CO~2~eq emissions per year.
   The information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.
@@
 # Use csl when using default citation_package for Pandoc citeproc
-# csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
+csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
 # citations with arabic numbers is prefered when using natbib
@@
 header-includes:
-  - \usepackage[hidelinks]{hyperref}
+  # - \usepackage[hidelinks]{hyperref} #this is to remove the annoying boxes. but how to make them blue?
+  - \usepackage{hyperref}
+  - \hypersetup{colorlinks = TRUE,  urlcolor = blue, linkcolor = blue, citecolor = blue}
+   #Colours links instead of ugly boxes
+  #Colour for external hyperlinks
+  #Colour of internal links
+  #Colour of citations
 
@@
   rticles::lncs_article:
-    citation_package: default # or natbib
+    citation_package: default # or natbib DEFINE LATER
 # output: word_document
+
+urlcolor: blue
+linkcolor: blue
+
 ---
@@
 
-**full paper**: 4-6 pages in length (typically up to 3,000 words).
+<!-- **full paper**: 4-6 pages in length (typically up to 3,000 words). -->
 
-In metropolitan areas, car trips can be replaced by a combination of public transit (PT) and cycling for the first-and-last mile.
+Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can significantly replace private car trips.
 This approach requires interventions and programs to make bicycling more appealing, and the resulting public investments can have significant social and environmental benefits.
-This paper focuses on estimating the potential for cycling + PT as a substitute for car trips in the Lisbon metropolitan area (LMA) and assessing its socio-environmental impacts using open data and open source tools.
 
-According to the latest mobility survey conducted in 2018, the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle. Car modal share is 58.4%, while PT accounts for 15.5%.
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively), the Department of Transport introduced biclaR, a decision support tool that facilitates the design and development of a metropolitan cycling network.
+According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.
+Car modal share was 58.4%, while PT accounted for 15.5%.
+The number of intra-municipal trips - with origin and destination in the same municipality - amounts to 3.5 million trips, exceeding the number of inter-municipal trips (1.8 million trips) - involving travel between different municipalities.
+Cars and public transport are the most used modes for intercity trips, with cars being the predominant choice for all journeys.
 
-This research aims to present and discuss the methods used to estimate…
+<!-- Plot of  AML trips by mode and inter/intra %? -->
 
-Propensity to Cycle Tool
+To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced _biclaR_^[see [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/)], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
 
-adding up an intermodality scenario to estimate cycling potential to public transit interfaces, and thus to support planning and prioritize investments in the cycling network.
+<!-- Robin: include here a paragraph about pct, and how biclar adds up -->
+**(RL)** The methods of Propensity to Cycle Tool  
+adding up an intermodality scenario to estimate cycling potential to public transit interfaces.
+
+This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
+After presenting the methods used, it assesses its socio-environmental impacts using open data and open-source tools.
 
@@
 
-## Case Study
-
-As características das viagens do IMob, constituem o cenário base deste projeto.
-Este inquérito à mobilidade foi realizado em 2017.
-Apesar de ter sido realizado em período pré-pandemia, este conjunto de dados é a melhor e mais recente informação que temos em termos de mobilidade urbana nas áreas metropolitanas.
-
-Segundo o IMob (2017), das cerca de 5.3 milhões de viagens diárias na área metropolitana de Lisboa, apenas
-25 479 das viagens são realizadas em bicicleta (0,5%), enquanto 3.1 milhões são feitas em automóvel (58.4%),
-1.3 milhões a pé (23.9%), 825 mil em transportes públicos (15.5%) e 96 mil em outros modos (1.8%).
-
-O número de viagens intramunicipais - ou que são realizadas com origem e destino no mesmo município (3.5
-milhões viagens) - é superior ao número de viagens intermunicipais - com origem e destino em municípios
-diferentes (1.8 milhões viagens). O automóvel e os transportes públicos são os modos maioritariamente
-utilizados em viagens intermunicipais. O automóvel é o modo mais utilizado em qualquer tipologia de viagem.
-
-Este é o cenário base, ou de referência, utilizado apenas para comparação com os cenários seguintes - de onde advém o potencial ciclável.
-
 ## Modeling Origin-Destination trips
 
-imob data
+The mobility survey data [@IMOB] is the basis for this project and defines the baseline scenario.
+Despite being conducted in the pre-pandemic period (2017), this dataset represents the most comprehensive and up-to-date information on urban mobility in Portuguese metropolitan areas (Lisbon and Porto).
 
-Aplicou-se um método de desagregação das origens e destinos das viagens entre o centróide de uma freguesia para o centróide de outra freguesia, para que uma freguesia não esteja representada apenas por um único local de origem e destino das suas viagens (centróides).
+We used a method for disaggregating the origins and destinations of trips between the centroids of two districts (same as "parish") to ensure that a district is not solely characterized by a single point of origin and destination for its trips.
+Aggregating all trips into centroids renders the exercise less realistic, as it excludes a significant portion of short-distance trips, a prevalent characteristic of active mode travel [@Lovelace2022Jittering].
+The OD Jittering method breaks down a single point (i.e., the centroid of an area) into multiple random points on the existing and neighboring road network, using OpenStreetMap as a reference.
+This method then distributes the volume of trips within the district among the randomly generated origin-destination pairs.
 
-Ao agregar as viagens todas em centróides, tornamos o exercício menos realista pois elimina um conjunto
-importante de viagens de curta distância, o que é uma característica de viagens em modos ativos [@ef1, @ref2].
-Para tal, recorreu-se ao método OD Jittering, que utiliza uma desagregação de um único ponto (por exemplo o centróide de uma área) em vários pontos aleatórios da rede viária existente, com
-base no OpenStreetMap, e divide o volume de viagens dessa freguesia pelos pares origem-destino gerados
-aleatoriamente.
+Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we employed a maximum disaggregation level of 100 trips per O-D pair for this project.
+Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.
 
-Para este projeto, utilizou-se um nível de desagregação máxima de 100 viagens por par O-D.
-Isto significa que, por exemplo, para um par O-D de 2.000 viagens entre duas freguesias ligadas entre os seus centróides, o método jittering dispersa aleatoriamente as 2.000 viagens em 20 pares de 100 viagens entre 20 origens e 20 destinos nas duas freguesias.
+```{r jitter, fig.align='center', fig.cap="Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
 
-A figura 3 ilustra a diferença entre a representação de viagens utilizando o método tradicional de ligação entre um único local entre cada freguesia, e a representação com a aleatorização e desagregação de viagens entre freguesias, para a área metropolitana de Lisboa.
-Figura 3 - Representação de pares OD na área metropolitana de Lisboa entre freguesias, sem jittering (à esquerda), e com jittering (à direita).
+# od_all = readRDS("paper/paperTRA/load/od_all.Rds")
+# od_jittered_filter = readRDS("paper/paperTRA/load/od_jittered_filter.Rds")
+# 
+# par(mfrow = c(1, 2))
+# plot(od_all$geometry, lwd = 0.1)
+# plot(od_jittered_filter$geometry, lwd = 0.1)
 
-O resultado da utilização deste método, em vez do tradicional uso de centróides de grandes áreas, é que torna mais realistas a representação de viagens realizadas, não correspondendo, no entanto, aos pares O-D reais das viagens realizadas (por motivos de proteção de dados definidos pelo INE).
-Para uma melhor compreensão do método utilizado, consultar Jittering: A Computationally Efficient Method for
-Generating Realistic Route Networks from Origin-Destination Data (Lovelace, Félix, and Carlino 2022)
+knitr::include_graphics("img/jitter.png", error = FALSE)
+
+```
+
+Although this method provides a more realistic representation of the trips undertaken compared to the traditional approach, it does not fully align with the actual O-D pairs of trips, which remain unknown due to data privacy regulations.
 
@@
 
-A identificação dos percursos entre as origens e destinos (para as linhas de desejo) são um aspeto fundamental da modelação da rede ciclável nesta metodologia, uma vez que o IMob não identifica os percursos realizados por cada viagem, sendo necessário estimá-los.
-A respetiva modelação depende do cenário escolhido e tem em conta toda a restante informação disponível (como os declives, ou localização de interfaces de transporte público), para além da informação viária base.
+The mobility survey collects the origin and destination of trips but does not include the respective routes.
+Modeling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
+Other constraints regarding the behavior of potential cyclists determine the routing results.
+For example, such restrictions can favor low speed, low traffic streets, more direct routes, and less steep paths, among others, suitable for cycling.
 
-A partir das linhas de desejo do cenário base, traçam-se os percursos modelados para o modo ciclável com base na rede viária (equivalente à afetação das viagens à rede, 4.º e último passo do modelo clássico dos 4 passos em sistemas de transportes).
-A modelação dos percursos cicláveis depende das variáveis consideradas, assim como das restrições pré-definidas.
-Estas restrições podem privilegiar aspetos como velocidade e volumes baixos, rotas mais diretas, percursos menos declivosos, entre outros, adequados à utilização da bicicleta, sendo que o algoritmo segue uma avaliação que resulta da ponderação das variáveis consideradas.
-Mesmo o percurso menos declivoso, mas com maior volume de tráfego, poderá ser o indicado se a função assim o definir.
+<!-- When comparing ground-truth dataset from 67 count locations distributed throughout the city of Lisbon with differemt routing providers and options, @Lovelace2022exploring found that jittering and disaggregating OD data, combined with routing using low level of traffic stress (quieter) preferences resulted in the most accurate route networks.
+-->
 
-O algoritmo de escolha de percurso escolhido foi o do r5r que é um algoritmo que permite muita flexibilidade nas configurações de tipo de percursos estimados, nomeadamente por permitir estimar com probabilidade e incerteza os percursos que usam transportes públicos (considerando os seus horários).
-O r5r permite a identificação de percursos mais diretos ou mais seguros, recorrendo para isso ao nível de stress de tráfego (Level of Traffic Stress, ou LTS), que varia de 1 a 4, sendo o 1 mais tranquilo - correspondendo, por exemplo, a pistas cicláveis fora de estrada, e o 4 menos tranquilo - correspondendo, por exemplo, a percursos que partilham o tráfego motorizado.
-Os percursos foram estimados para o cenário base, para ambos os tipos de rede, direto e seguro, usando para isso o LTS 4 e LTS 3 respetivamente.
+The selected route choice algorithm was the [`r5r` R package](https://ipeagit.github.io/r5r/) [@r5r], which allows for great flexibility in configuring estimated route types, and which proven to provide most accurate route networks for the city of Lisbon [@Lovelace2022exploring].
+`r5r` can calculate multi-modal routes using PT combined with other modes.
+It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress^[see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
+The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.
 
-O nível de tranquilidade^[see https://www.cyclestreets.net/help/journey/howitworks/#quietness] é outro indicador estimado pelo CycleStreets para cada troço, com base no número de vias, velocidade máxima permitida, arborização, hierarquia e tráfego (quando a informação correspondente existe nas etiquetas do OpenStreetMap).
-É apresentado numa escala de 0 a 100, em que o zero corresponde ao nível menos seguro para se circular de bicicleta, e o 100 ao nível mais seguro e tranquilo - normalmente correspondendo a troços que incluem já infraestrutura ciclável segregada do tráfego motorizado.
+<!-- The quietness level^[see https://www.cyclestreets.net/help/journey/howitworks/#quietness] was another indicator estimated by [CycleStreets](https://www.cyclestreets.org/) for each route section, based on road network variables xisting on OpenStreetMap labels, such as the number of lanes, maximum allowed speed, trees, or road hierarchy.
+Values vary on a scale of 0 to 100, where zero corresponds to the least safe level for cycling and 100 corresponds to the safest and quietest level, usually corresponding to sections that already include segregated cycling infrastructure away from motorized traffic.
+I THINK THIS IS NOT RELEVANT FOR THIS SHORT PAPER, AS IT IS MOST USEFUL FOR VIZUALIZATION PURPOSES ON BICLAR TOOL-->
 
-Foi utilizado um modelo digital do terreno de 25m de resolução espacial e erro médio de altimetria vertical de 2.12m8 em Portugal, da Agência Espacial Europeia, na sua missão COPERNICUS, para incluir impedâncias 9 de declives no modelo.
-O modelo r5r utiliza a rede viária do OpenStreetMap (extraída a 22 de outubro 2022), e os dados de GTFS criados e validados para o efeito.
-Foi estimado o número de viagens em bicicleta potencial para as duas metas ENMAC (5% e 10%) a partir dos valores de viagens em bicicleta e em automóvel (como condutor e como passageiro) do cenário base de 2017.
+The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data agregated and validated.
+This information is crucial for an accurate PT trip and route estimation.
+A digital elevation model from the European Space Agency's COPERNICUS mission, with a 25m spatial resolution, was used to include street gradient information, as a weight in cycling routing.
+The cycling potential trips for the two ENMAC targets (4% and 10%) was estimated from the values for cycling and car trips (both as a driver and as a passenger) from the 2017 base scenario.
 
-Os percursos ou rotas foram depois sobrepostos e agregados por segmentos, juntando a informação de várias rotas que se sobrepõem em certos segmentos (overline).
-Por exemplo, foram somados os volumes de viagens em bicicleta estimados e a transferência do automóvel, e foi feita a média da velocidade automóvel e do nível de tranquilidade, ponderadas pela distância dos segmentos.
+The routes were then overlaid and aggregated by segments, using [`stplanr overline()` R function](https://docs.ropensci.org/stplanr/reference/overline.html).
+<!-- For example, the estimated volumes of bicycle trips and car transfers were aggregated, and the average car speed and quietness level, weighted by the distance of the segments, were calculated.
+-->
 
@@
 
-The intermodality scenario considers trips that can combine PT and cycling for the first-and-last legs.
-Conservatively, we considered the sum of first-and-last legs up to 5 km.
-Furthermore, we restricted PT use to unimodal trips without transfers (although they can be included in future modeling).
-[@bib2, @felix2023]
+The intermodality scenario considers trips combining PT and cycling for the first and last legs.
+In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km or up to 25 minutes on bike.
+Furthermore, we have imposed restrictions on PT usage to include only trips with no PT transfers, and up to 2 hours (120 min).
+Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
 
-Finally, we only included PT modes that can practically accommodate bicycles, such as trains, ferries, trams, and intermunicipal bus lines with bike racks (\ \ref{fig:map1}).
-
-Foram recolhidos os dados de transportes públicos da AML, em formato GTFS, tendo sido necessário o trabalho de limpeza, união, e validação prévio.
-Esta informação é indispensável para saber que terminais de transporte público (TP) se ligam a que origens ou destinos, e em que horários, para não se considerar de forma generalista que uma interface se liga a todo o sistema e que pode servir todas as viagens (origens e destinos), o que não seria realista.
-
-```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area"}
-knitr::include_graphics("img/map1.png")
+```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area", fig.align='center'}
+knitr::include_graphics("img/map1.png", error = FALSE)
 ```
 
-To obtain reliable results, we used the OpenStreetMap road network and GTFS data.
-The r5r R package estimated the trip duration and distance for both the original modes and the bicycle + PT combination, while the od jittering R package estimated the OD locations based on a centroid-based OD matrix.
+Figure \ref{fig:map2} illustrates the resulting bicycle routes to access the main PT interfaces in the LMA.
 
-Para a modelação dos percursos definiu-se como janela temporal viagens até 120 minutos (2h). Para além disso considerou-se como duração máxima 25 minutos em bicicleta.
-
-Map result (\ \ref{fig:map2}).
-
-```{r map2, out.width="80%", fig.cap="Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool)."}
-knitr::include_graphics("img/map2.png")
+```{r map2, out.width="80%", fig.cap="Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).", fig.align='center'}
+knitr::include_graphics("img/map2.png", error = FALSE)
 ```
 
-## Assessing socio-environmental benefits
+## Assessing socio-environmental benefits - CONTINUE FROM HERE
 
-Socio-environmental impacts were assessed using the HEAT for Cycling and the HEAT as a Service^[see https://github.com/HEAT-WHO/HEAT_heatr_api] tools, from the WHO.
-Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA’s COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.
+Socio-environmental impacts were estimated for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
+using the HEAT for Cycling and the HEAT as a Service^[see https://github.com/HEAT-WHO/HEAT_heatr_api] tools, from the WHO.
 
-Como referido na introdução, a ferramenta biclaR pretende estimar os impactes sociais e ambientais associados ao potencial ciclável dos vários cenários analisados.
-O horizonte temporal considerado para esta análise foi o curto prazo (i.e., 1 ano) e o longo prazo (i.e., 10 anos).
  Os impactes foram avaliados para cada cenário e para diferentes escalas territoriais:
@@
 Para todos os cenários, recorreu-se à ferramenta HEAT for cycling, da Organização Mundial de Saúde, para estimativa dos impactes sociais e ambientais da transferência de viagens em automóvel para viagens em bicicleta, nas componentes de: a) Social - Saúde, Atividade Física, Exposição a poluição atmosférica, Exposição ao risco de sinistralidade rodoviária; b) Ambiental - Emissão de gases CO~2~eq.
+
+Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA’s COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.
 
@@
 ```{r summary1, out.width="100%", fig.cap="Summary of the cycling potencial of intermodality scenario."}
-knitr::include_graphics("img/table1.png")
+knitr::include_graphics("img/table1.png", error = FALSE)
 ```
@@
 ```{r summary2, out.width="100%", fig.cap="Potencial de transferência estimado para cada modo de transporte público, bem como as estimativas de consumo de CO~2~eq evitado anualmente por transferência do automóvel."}
-knitr::include_graphics("img/table2.png")
+knitr::include_graphics("img/table2.png", error = FALSE)
 ```
@@
 
-# References (test) {.unnumbered}
+# References {.unnumbered}

### 2023-09-26 paper/PaperTRA/PaperTRA.tex
paper TRA - translt
stopped at  section 2.4 (will resume soon)

@Robinlovelace can you take a look at the paragraph in the introduction about PCT?

@@
 
-\usepackage[hidelinks]{hyperref}
+\usepackage{hyperref}
+\hypersetup{colorlinks = TRUE,  urlcolor = blue, linkcolor = blue, citecolor = blue}
 
@@
 public transit and cycling for the first-and-last mile. This paper
-focuses on estimating the potential for cycling + PT as a substitute for
-car trips in the Lisbon metropolitan area and assessing its
-socio-environmental impacts using open data and open source tools. A
+estimates the potential for cycling + public transit (PT) as a
+substitute for car trips in the Lisbon metropolitan area and assesses
+its socio-environmental impacts using open data and open source tools. A
 decision support tool that facilitates the design and development of a
@@
 assessed using the \emph{HEAT for Cycling} and the \emph{HEAT as a
-Service} tools. Additionally, the impacts of shifting car trips to PT
-were estimated and monetized. The results indicate that 20\% of the
-current trips can be made with the bicycle + PT combination. Shifting to
-cycling for the first-and-last mile can reduce annual
-CO\textsubscript{2}eq emissions from 6,000 tons/day, and the 10-year
-socio-environmental benefits account from €230 million. For the PT leg,
-the transfer from car results in the avoidance of at lest 8,500 tons of
-CO\textsubscript{2}eq emissions per year. The information on
-socio-economic benefits can support policymakers in prioritizing
-interventions to reduce the reliance on individual motorized
-transportation and effectively communicate their decisions.
+Service} tools. The impacts of shifting car trips to PT were also
+estimated and monetized. The results indicate that 20\% of the current
+trips can be made with the bicycle + PT combination. Shifting to cycling
+for the first-and-last mile can reduce annual CO\textsubscript{2}eq
+emissions from 6,000 tons/day, and the 10-year socio-environmental
+benefits account from €230 million. For the PT leg, the transfer from
+car avoids of at least 8,500 tons of CO\textsubscript{2}eq emissions per
+year. The information on socio-economic benefits can support
+policymakers in prioritizing interventions to reduce the reliance on
+individual motorized transportation and effectively communicate their
+decisions.
 
@@
 
-\textbf{full paper}: 4-6 pages in length (typically up to 3,000 words).
+Combining public transportation (PT) and cycling for the first and last
+mile in metropolitan areas can significantly replace private car trips.
+This approach requires interventions and programs to make bicycling more
+appealing, and the resulting public investments can have significant
+social and environmental benefits.
 
-In metropolitan areas, car trips can be replaced by a combination of
-public transit (PT) and cycling for the first-and-last mile. This
-approach requires interventions and programs to make bicycling more
-appealing, and the resulting public investments can have significant
-social and environmental benefits. This paper focuses on estimating the
-potential for cycling + PT as a substitute for car trips in the Lisbon
-metropolitan area (LMA) and assessing its socio-environmental impacts
-using open data and open source tools.
+According to the latest mobility survey conducted in 2018 {[}1{]}, the
+LMA registered a total of 5.3 million daily trips, with only 0.5\% by
+bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%. The
+number of intra-municipal trips - with origin and destination in the
+same municipality - amounts to 3.5 million trips, exceeding the number
+of inter-municipal trips (1.8 million trips) - involving travel between
+different municipalities. Cars and public transport are the most used
+modes for intercity trips, with cars being the predominant choice for
+all journeys.
 
-According to the latest mobility survey conducted in 2018, the LMA
-registered a total of 5.3 million daily trips, with only 0.5\% by
-bicycle. Car modal share is 58.4\%, while PT accounts for 15.5\%. To
-achieve the cycling targets set by the Portuguese national cycling
-strategy for 2025 and 2030 (4\% and 10\%, respectively), the Department
-of Transport introduced biclaR, a decision support tool that facilitates
-the design and development of a metropolitan cycling network.
+To achieve the cycling targets set by the Portuguese national cycling
+strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}2{]}, the
+Lisbon's Metropolitan Department of Transport introduced
+\emph{biclaR}\footnote{see
+  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}}, a
+decision support tool that facilitates the design and development of a
+metropolitan cycling network {[}3{]}.
 
-This research aims to present and discuss the methods used to
-estimate\ldots{}
+\textbf{(RL)} The methods of Propensity to Cycle Tool\\
+adding up an intermodality scenario to estimate cycling potential to
+public transit interfaces.
 
-Propensity to Cycle Tool
-
-adding up an intermodality scenario to estimate cycling potential to
-public transit interfaces, and thus to support planning and prioritize
-investments in the cycling network.
+This paper estimates the potential for combining cycling and PT to
+substitute car trips in the LMA. After presenting the methods used, it
+assesses its socio-environmental impacts using open data and open-source
+tools.
 
@@
 \section{Methods}\label{methods}}
-
-\hypertarget{case-study}{%
-\subsection{Case Study}\label{case-study}}
-
-As características das viagens do IMob, constituem o cenário base deste
-projeto. Este inquérito à mobilidade foi realizado em 2017. Apesar de
-ter sido realizado em período pré-pandemia, este conjunto de dados é a
-melhor e mais recente informação que temos em termos de mobilidade
-urbana nas áreas metropolitanas.
-
-Segundo o IMob (2017), das cerca de 5.3 milhões de viagens diárias na
-área metropolitana de Lisboa, apenas 25 479 das viagens são realizadas
-em bicicleta (0,5\%), enquanto 3.1 milhões são feitas em automóvel
-(58.4\%), 1.3 milhões a pé (23.9\%), 825 mil em transportes públicos
-(15.5\%) e 96 mil em outros modos (1.8\%).
-
-O número de viagens intramunicipais - ou que são realizadas com origem e
-destino no mesmo município (3.5 milhões viagens) - é superior ao número
-de viagens intermunicipais - com origem e destino em municípios
-diferentes (1.8 milhões viagens). O automóvel e os transportes públicos
-são os modos maioritariamente utilizados em viagens intermunicipais. O
-automóvel é o modo mais utilizado em qualquer tipologia de viagem.
-
-Este é o cenário base, ou de referência, utilizado apenas para
-comparação com os cenários seguintes - de onde advém o potencial
-ciclável.
 
@@
 
-imob data
+The mobility survey data {[}1{]} is the basis for this project and
+defines the baseline scenario. Despite being conducted in the
+pre-pandemic period (2017), this dataset represents the most
+comprehensive and up-to-date information on urban mobility in Portuguese
+metropolitan areas (Lisbon and Porto).
 
-Aplicou-se um método de desagregação das origens e destinos das viagens
-entre o centróide de uma freguesia para o centróide de outra freguesia,
-para que uma freguesia não esteja representada apenas por um único local
-de origem e destino das suas viagens (centróides).
+We used a method for disaggregating the origins and destinations of
+trips between the centroids of two districts (same as ``parish'') to
+ensure that a district is not solely characterized by a single point of
+origin and destination for its trips. Aggregating all trips into
+centroids renders the exercise less realistic, as it excludes a
+significant portion of short-distance trips, a prevalent characteristic
+of active mode travel {[}4{]}. The OD Jittering method breaks down a
+single point (i.e., the centroid of an area) into multiple random points
+on the existing and neighboring road network, using OpenStreetMap as a
+reference. This method then distributes the volume of trips within the
+district among the randomly generated origin-destination pairs.
 
-Ao agregar as viagens todas em centróides, tornamos o exercício menos
-realista pois elimina um conjunto importante de viagens de curta
-distância, o que é uma característica de viagens em modos ativos
-(\textbf{ref2?}). Para tal, recorreu-se ao método OD Jittering, que
-utiliza uma desagregação de um único ponto (por exemplo o centróide de
-uma área) em vários pontos aleatórios da rede viária existente, com base
-no OpenStreetMap, e divide o volume de viagens dessa freguesia pelos
-pares origem-destino gerados aleatoriamente.
+Using the
+\href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
+package}, we employed a maximum disaggregation level of 100 trips per
+O-D pair for this project. Figure \ref{fig:jitter} illustrates the
+contrast between trip representation through the traditional method,
+which connects a single desire line between each district, and the
+presentation achieved through the randomization and disaggregation of
+trips between districts, specifically for the Lisbon metropolitan area.
 
-Para este projeto, utilizou-se um nível de desagregação máxima de 100
-viagens por par O-D. Isto significa que, por exemplo, para um par O-D de
-2.000 viagens entre duas freguesias ligadas entre os seus centróides, o
-método jittering dispersa aleatoriamente as 2.000 viagens em 20 pares de
-100 viagens entre 20 origens e 20 destinos nas duas freguesias.
+\begin{figure}
 
-A figura 3 ilustra a diferença entre a representação de viagens
-utilizando o método tradicional de ligação entre um único local entre
-cada freguesia, e a representação com a aleatorização e desagregação de
-viagens entre freguesias, para a área metropolitana de Lisboa. Figura 3
-- Representação de pares OD na área metropolitana de Lisboa entre
-freguesias, sem jittering (à esquerda), e com jittering (à direita).
+{\centering \includegraphics[width=1\linewidth,]{img/jitter}
 
-O resultado da utilização deste método, em vez do tradicional uso de
-centróides de grandes áreas, é que torna mais realistas a representação
-de viagens realizadas, não correspondendo, no entanto, aos pares O-D
-reais das viagens realizadas (por motivos de proteção de dados definidos
-pelo INE). Para uma melhor compreensão do método utilizado, consultar
-Jittering: A Computationally Efficient Method for Generating Realistic
-Route Networks from Origin-Destination Data (Lovelace, Félix, and
-Carlino 2022)
+}
+
+\caption{Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
+\end{figure}
+
+Although this method provides a more realistic representation of the
+trips undertaken compared to the traditional approach, it does not fully
+align with the actual O-D pairs of trips, which remain unknown due to
+data privacy regulations.
 
@@
 
-A identificação dos percursos entre as origens e destinos (para as
-linhas de desejo) são um aspeto fundamental da modelação da rede
-ciclável nesta metodologia, uma vez que o IMob não identifica os
-percursos realizados por cada viagem, sendo necessário estimá-los. A
-respetiva modelação depende do cenário escolhido e tem em conta toda a
-restante informação disponível (como os declives, ou localização de
-interfaces de transporte público), para além da informação viária base.
+The mobility survey collects the origin and destination of trips but
+does not include the respective routes. Modeling the realistic cycling +
+PT routes between OD pairs depends on assumptions regarding the
+characteristics of the cycling and road networks and the location of
+public transport interfaces. Other constraints regarding the behavior of
+potential cyclists determine the routing results. For example, such
+restrictions can favor low speed, low traffic streets, more direct
+routes, and less steep paths, among others, suitable for cycling.
 
-A partir das linhas de desejo do cenário base, traçam-se os percursos
-modelados para o modo ciclável com base na rede viária (equivalente à
-afetação das viagens à rede, 4.º e último passo do modelo clássico dos 4
-passos em sistemas de transportes). A modelação dos percursos cicláveis
-depende das variáveis consideradas, assim como das restrições
-pré-definidas. Estas restrições podem privilegiar aspetos como
-velocidade e volumes baixos, rotas mais diretas, percursos menos
-declivosos, entre outros, adequados à utilização da bicicleta, sendo que
-o algoritmo segue uma avaliação que resulta da ponderação das variáveis
-consideradas. Mesmo o percurso menos declivoso, mas com maior volume de
-tráfego, poderá ser o indicado se a função assim o definir.
+The selected route choice algorithm was the
+\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}5{]},
+which allows for great flexibility in configuring estimated route types,
+and which proven to provide most accurate route networks for the city of
+Lisbon {[}6{]}. \texttt{r5r} can calculate multi-modal routes using PT
+combined with other modes. It enables the identification of the most
+direct or safest cycling routes, using the Level of Traffic
+Stress\footnote{see
+  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}}
+(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
+(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
+routes shared with motorized traffic). The routes were estimated for the
+base scenario for both types of networks: \emph{direct} and \emph{safe},
+using LTS 4 and LTS 3, respectively.
 
-O algoritmo de escolha de percurso escolhido foi o do r5r que é um
-algoritmo que permite muita flexibilidade nas configurações de tipo de
-percursos estimados, nomeadamente por permitir estimar com probabilidade
-e incerteza os percursos que usam transportes públicos (considerando os
-seus horários). O r5r permite a identificação de percursos mais diretos
-ou mais seguros, recorrendo para isso ao nível de stress de tráfego
-(Level of Traffic Stress, ou LTS), que varia de 1 a 4, sendo o 1 mais
-tranquilo - correspondendo, por exemplo, a pistas cicláveis fora de
-estrada, e o 4 menos tranquilo - correspondendo, por exemplo, a
-percursos que partilham o tráfego motorizado. Os percursos foram
-estimados para o cenário base, para ambos os tipos de rede, direto e
-seguro, usando para isso o LTS 4 e LTS 3 respetivamente.
+The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
+metropolitan data agregated and validated. This information is crucial
+for an accurate PT trip and route estimation. A digital elevation model
+from the European Space Agency's COPERNICUS mission, with a 25m spatial
+resolution, was used to include street gradient information, as a weight
+in cycling routing. The cycling potential trips for the two ENMAC
+targets (4\% and 10\%) was estimated from the values for cycling and car
+trips (both as a driver and as a passenger) from the 2017 base scenario.
 
-O nível de tranquilidade\footnote{see
-  \url{https://www.cyclestreets.net/help/journey/howitworks/\#quietness}}
-é outro indicador estimado pelo CycleStreets para cada troço, com base
-no número de vias, velocidade máxima permitida, arborização, hierarquia
-e tráfego (quando a informação correspondente existe nas etiquetas do
-OpenStreetMap). É apresentado numa escala de 0 a 100, em que o zero
-corresponde ao nível menos seguro para se circular de bicicleta, e o 100
-ao nível mais seguro e tranquilo - normalmente correspondendo a troços
-que incluem já infraestrutura ciclável segregada do tráfego motorizado.
-
-Foi utilizado um modelo digital do terreno de 25m de resolução espacial
-e erro médio de altimetria vertical de 2.12m8 em Portugal, da Agência
-Espacial Europeia, na sua missão COPERNICUS, para incluir impedâncias 9
-de declives no modelo. O modelo r5r utiliza a rede viária do
-OpenStreetMap (extraída a 22 de outubro 2022), e os dados de GTFS
-criados e validados para o efeito. Foi estimado o número de viagens em
-bicicleta potencial para as duas metas ENMAC (5\% e 10\%) a partir dos
-valores de viagens em bicicleta e em automóvel (como condutor e como
-passageiro) do cenário base de 2017.
-
-Os percursos ou rotas foram depois sobrepostos e agregados por
-segmentos, juntando a informação de várias rotas que se sobrepõem em
-certos segmentos (overline). Por exemplo, foram somados os volumes de
-viagens em bicicleta estimados e a transferência do automóvel, e foi
-feita a média da velocidade automóvel e do nível de tranquilidade,
-ponderadas pela distância dos segmentos.
+The routes were then overlaid and aggregated by segments, using
+\href{https://docs.ropensci.org/stplanr/reference/overline.html}{\texttt{stplanr\ overline()}
+R function}.
 
@@
 
-The intermodality scenario considers trips that can combine PT and
-cycling for the first-and-last legs. Conservatively, we considered the
-sum of first-and-last legs up to 5 km. Furthermore, we restricted PT use
-to unimodal trips without transfers (although they can be included in
-future modeling). Félix, Lovelace, and Moura (2022)
-
-Finally, we only included PT modes that can practically accommodate
-bicycles, such as trains, ferries, trams, and intermunicipal bus lines
-with bike racks (~\ref{fig:map1}).
-
-Foram recolhidos os dados de transportes públicos da AML, em formato
-GTFS, tendo sido necessário o trabalho de limpeza, união, e validação
-prévio. Esta informação é indispensável para saber que terminais de
-transporte público (TP) se ligam a que origens ou destinos, e em que
-horários, para não se considerar de forma generalista que uma interface
-se liga a todo o sistema e que pode servir todas as viagens (origens e
-destinos), o que não seria realista.
+The intermodality scenario considers trips combining PT and cycling for
+the first and last legs. In a conservative approach, we have restricted
+our analysis to the first and last legs with a combined length of up to
+5 km or up to 25 minutes on bike. Furthermore, we have imposed
+restrictions on PT usage to include only trips with no PT transfers, and
+up to 2 hours (120 min). Additionally, we have only included PT modes
+that can easily accommodate bicycles, such as trains, ferries, trams,
+and inter-municipal bus lines equipped with bike racks (Figure
+\ref{fig:map1}).
 
 \begin{figure}
-\includegraphics[width=0.6\linewidth,]{img/map1} \caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
+
+{\centering \includegraphics[width=0.6\linewidth,]{img/map1}
+
+}
+
+\caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
 \end{figure}
 
-To obtain reliable results, we used the OpenStreetMap road network and
-GTFS data. The r5r R package estimated the trip duration and distance
-for both the original modes and the bicycle + PT combination, while the
-od jittering R package estimated the OD locations based on a
-centroid-based OD matrix.
-
-Para a modelação dos percursos definiu-se como janela temporal viagens
-até 120 minutos (2h). Para além disso considerou-se como duração máxima
-25 minutos em bicicleta.
-
-Map result (~\ref{fig:map2}).
+Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
+the main PT interfaces in the LMA.
 
 \begin{figure}
-\includegraphics[width=0.8\linewidth,]{img/map2} \caption{Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
+
+{\centering \includegraphics[width=0.8\linewidth,]{img/map2}
+
+}
+
+\caption{Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
 \end{figure}
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{INE: Mobilidade e funcionalidade do território nas
+{Áreas Metropolitanas do Porto e de Lisboa}: 2017,
+\url{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt},
+(2018).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{Presidência do Conselho de Ministros: Resolução do
+conselho de ministros n.º 131/2019,
+\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}, (2019).}
+
 \leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
-Félix, Rosa, Robin Lovelace, and Filipe Moura. 2022. {``{biclaR -
-Ferramenta de apoio ao planeamento da rede ciclável na área
-metropolitana de Lisboa}.''} {CERIS - Instituto Superior Técnico and
-Transportes Metropolitanos de Lisboa}.
-\url{https://biclar.tmlmobilidade.pt}.
+\CSLLeftMargin{3. }%
+\CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
+de apoio ao planeamento da rede ciclável na área metropolitana de
+Lisboa}, \url{https://biclar.tmlmobilidade.pt}, (2022).}
 
-\leavevmode\vadjust pre{\hypertarget{ref-bib2}{}}%
-Slifka, M. K., and J. L. Whitton. 2000. {``Clinical Implications of
-Dysregulated Cytokine Production.''} \emph{J. {M}ol. {M}ed.} 78: 74--80.
-\url{https://doi.org/10.1007/s001090000086}.
+\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
+computationally efficient method for generating realistic route networks
+from origin-destination data. Findings. (2022).
+https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
+C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
+transport networks with R5 in r. Findings. (2021).
+https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
+jittering and routing options for converting origin-destination data
+into route networks: Towards accurate estimates of movement at the
+street level. The International Archives of the Photogrammetry, Remote
+Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
+(2022).
+https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}
 

### 2023-09-26 paper/PaperTRA/PaperTRA.Rmd
Not just metro areas
Update abstract

@@
 abstract: | 
-  In metropolitan areas, car trips can be replaced by a combination of public transit and cycling for the first-and-last mile.
-This paper estimates the potential for cycling + public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
+  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile.
+This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
   A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_).
-A scenario of intermodality introduced, and its socio-environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
+The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
 The impacts of shifting car trips to PT were also estimated and monetized.
-  The results indicate that 20% of the current trips can be made with the bicycle + PT combination.
-Shifting to cycling for the first-and-last mile can reduce annual CO~2~eq emissions from 6,000 tons/day, and the 10-year socio-environmental benefits account from €230 million.
+  The results indicate that 20% of cart trips could switch to the bicycle + PT combination.
+Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of €230 million.
 For the PT leg, the transfer from car avoids of at least 8,500 tons of CO~2~eq emissions per year.
-  The information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.
+  This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
 keywords: 

### 2023-09-26 paper/PaperTRA/PaperTRA.Rmd
Fix cart typo
Minor text tweaks

@@
 The impacts of shifting car trips to PT were also estimated and monetized.
-  The results indicate that 20% of cart trips could switch to the bicycle + PT combination.
+  The results indicate that 20% of car trips could switch to the bicycle + PT combination.
 Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of €230 million.
@@
 Car modal share was 58.4%, while PT accounted for 15.5%.
-The number of intra-municipal trips - with origin and destination in the same municipality - amounts to 3.5 million trips, exceeding the number of inter-municipal trips (1.8 million trips) - involving travel between different municipalities.
+The number of intra-municipal trips --- with origin and destination in the same municipality --- amounts to 3.5 million trips.
+This exceeds the number of inter-municipal trips (1.8 million trips), involving travel between different municipalities.
 Cars and public transport are the most used modes for intercity trips, with cars being the predominant choice for all journeys.
@@
 
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced _biclaR_^[see [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/)], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
+To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced _biclaR_^[See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/)], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
 

### 2023-09-26 paper/PaperTRA/PaperTRA.Rmd
Add PCT content

@@
 
-<!-- Robin: include here a paragraph about pct, and how biclar adds up -->
-**(RL)** The methods of Propensity to Cycle Tool  
-adding up an intermodality scenario to estimate cycling potential to public transit interfaces.
+*biclar* builds on the Propensity to Cycle Tool (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
+The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@Lovelace2017].
 

### 2023-09-26 paper/PaperTRA/PaperTRA.tex
Add PCT content

@@
 \begin{abstract}
-In metropolitan areas, car trips can be replaced by a combination of
+A high proportion of car trips can be replaced by a combination of
 public transit and cycling for the first-and-last mile. This paper
-estimates the potential for cycling + public transit (PT) as a
-substitute for car trips in the Lisbon metropolitan area and assesses
+estimates the potential for cycling combined with public transit (PT) as
+a substitute for car trips in the Lisbon metropolitan area and assesses
 its socio-environmental impacts using open data and open source tools. A
 decision support tool that facilitates the design and development of a
-metropolitan cycling network was developed (\emph{biclaR}). A scenario
-of intermodality introduced, and its socio-environmental impacts were
-assessed using the \emph{HEAT for Cycling} and the \emph{HEAT as a
-Service} tools. The impacts of shifting car trips to PT were also
-estimated and monetized. The results indicate that 20\% of the current
-trips can be made with the bicycle + PT combination. Shifting to cycling
-for the first-and-last mile can reduce annual CO\textsubscript{2}eq
-emissions from 6,000 tons/day, and the 10-year socio-environmental
-benefits account from €230 million. For the PT leg, the transfer from
-car avoids of at least 8,500 tons of CO\textsubscript{2}eq emissions per
-year. The information on socio-economic benefits can support
-policymakers in prioritizing interventions to reduce the reliance on
-individual motorized transportation and effectively communicate their
-decisions.
+metropolitan cycling network was developed (\emph{biclaR}). The social
+and environmental impacts were assessed using the \emph{HEAT for
+Cycling} and the \emph{HEAT as a Service} tools. The impacts of shifting
+car trips to PT were also estimated and monetized. The results indicate
+that 20\% of car trips could switch to the bicycle + PT combination.
+Shifting to cycling for the first-and-last mile stages can reduce annual
+CO\textsubscript{2}eq emissions from 6,000 tons/day, with benefits over
+10 years of €230 million. For the PT leg, the transfer from car avoids
+of at least 8,500 tons of CO\textsubscript{2}eq emissions per year. This
+evidence can support policymakers to prioritize interventions that
+reduce the reliance on private motor vehicles.
 
@@
 bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%. The
-number of intra-municipal trips - with origin and destination in the
-same municipality - amounts to 3.5 million trips, exceeding the number
-of inter-municipal trips (1.8 million trips) - involving travel between
-different municipalities. Cars and public transport are the most used
-modes for intercity trips, with cars being the predominant choice for
-all journeys.
+number of intra-municipal trips --- with origin and destination in the
+same municipality --- amounts to 3.5 million trips. This exceeds the
+number of inter-municipal trips (1.8 million trips), involving travel
+between different municipalities. Cars and public transport are the most
+used modes for intercity trips, with cars being the predominant choice
+for all journeys.
 
@@
 Lisbon's Metropolitan Department of Transport introduced
-\emph{biclaR}\footnote{see
+\emph{biclaR}\footnote{See
   \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}}, a
@@
 
-\textbf{(RL)} The methods of Propensity to Cycle Tool\\
-adding up an intermodality scenario to estimate cycling potential to
-public transit interfaces.
+\emph{biclar} builds on the Propensity to Cycle Tool (PCT), a web
+application and research project funded by the UK's Department for
+Transport in 2015 which launched nationally in 2017 as part of the
+government's Cycling and Walking Investment Strategy. The PCT initially
+used only origin-destination data for commuting trips as the basis of
+estimates of cycling potential at zone, route and route network levels
+{[}\textbf{Lovelace2017?}{]}.
 
@@
 
-\hypertarget{assessing-socio-environmental-benefits}{%
-\subsection{Assessing socio-environmental
-benefits}\label{assessing-socio-environmental-benefits}}
+\hypertarget{assessing-socio-environmental-benefits---continue-from-here}{%
+\subsection{Assessing socio-environmental benefits - CONTINUE FROM
+HERE}\label{assessing-socio-environmental-benefits---continue-from-here}}
 
-Socio-environmental impacts were assessed using the HEAT for Cycling and
-the HEAT as a Service\footnote{see
+Socio-environmental impacts were estimated for a short term time horizon
+(i.e., one year) and the long term (i.e., ten years). using the HEAT for
+Cycling and the HEAT as a Service\footnote{see
   \url{https://github.com/HEAT-WHO/HEAT_heatr_api}} tools, from the WHO.
-Additionally, we estimate the impacts of shifting car trips to PT for
-the second leg of the journey with EMEP/EEA's COPERT methodology and
-monetize them with the EU Guide to cost-benefit analysis.
 
-Como referido na introdução, a ferramenta biclaR pretende estimar os
-impactes sociais e ambientais associados ao potencial ciclável dos
-vários cenários analisados. O horizonte temporal considerado para esta
-análise foi o curto prazo (i.e., 1 ano) e o longo prazo (i.e., 10 anos).
 Os impactes foram avaliados para cada cenário e para diferentes escalas
@@
 rodoviária; b) Ambiental - Emissão de gases CO\textsubscript{2}eq.
+
+Additionally, we estimate the impacts of shifting car trips to PT for
+the second leg of the journey with EMEP/EEA's COPERT methodology and
+monetize them with the EU Guide to cost-benefit analysis.
 

### 2023-09-27 paper/PaperTRA/PaperTRA.Rmd

@@
 linkcolor: blue
-
 ---
@@
 
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced _biclaR_^[See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/)], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
+To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced *biclaR*[^1], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
+
+[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/)
 
 *biclar* builds on the Propensity to Cycle Tool (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
-The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@Lovelace2017].
+The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
+The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^2]
+However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.
+
+[^2]:  See <https://www.npt.scot> and <https://cruse.bike> for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.
 
@@
 `r5r` can calculate multi-modal routes using PT combined with other modes.
-It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress^[see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
+It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^3] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
 The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.
+
+[^3]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)
 
@@
 Socio-environmental impacts were estimated for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
-using the HEAT for Cycling and the HEAT as a Service^[see https://github.com/HEAT-WHO/HEAT_heatr_api] tools, from the WHO.
+using the HEAT for Cycling and the HEAT as a Service[^4] tools, from the WHO.
+
+[^4]: see https://github.com/HEAT-WHO/HEAT_heatr_api
 
@@
 
-* Desagregado à escala municipal para cada segmento de rede:
-  + Ambientais (emissões de CO~2~eq evitadas);
-  + Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
-* Agregado à escala metropolitana:
-  + Ambientais (emissões de CO~2~eq evitadas);
-  + Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
-  + Ambientais (emissões de CO~2~eq, CO, PM10, NOx, e VOC11, evitadas por substituição dos modos motorizados, considerando as emissões geradas na transferência para transportes públicos).
+-   Desagregado à escala municipal para cada segmento de rede:
+    -   Ambientais (emissões de CO~2~eq evitadas);
+    -   Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
+-   Agregado à escala metropolitana:
+    -   Ambientais (emissões de CO~2~eq evitadas);
+    -   Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
+    -   Ambientais (emissões de CO~2~eq, CO, PM10, NOx, e VOC11, evitadas por substituição dos modos motorizados, considerando as emissões geradas na transferência para transportes públicos).
 
@@
 
-Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA’s COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.
+Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA's COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.
 
@@
 Considerou-se que as viagens foram todas realizadas em condições urbanas (com as respetivas implicações no regime médio de condução) e a uma velocidade média de 15km/h, nos períodos de hora de ponta.
-Uma vez que a distância média percorrida por viagem influencia o nível de sobreconsumo e emissões decorrentes da operação do motor a frio – ou seja, distâncias maiores diluem a importância deste sobreconsumo face ao fator de consumo com o motor a operar a quente, foram estimados os consumos para diferentes gamas de viagem, em intervalos de 500 metros.
+Uma vez que a distância média percorrida por viagem influencia o nível de sobreconsumo e emissões decorrentes da operação do motor a frio -- ou seja, distâncias maiores diluem a importância deste sobreconsumo face ao fator de consumo com o motor a operar a quente, foram estimados os consumos para diferentes gamas de viagem, em intervalos de 500 metros.
 As emissões são estimadas para os seguintes poluentes atmosféricos: CO (monóxido de carbono), NO X (Óxidos de Azoto), COV (Compostos Orgânicos Voláteis), PM (material particulado).
@@
 a conversão das emissões evitadas em perda de bem-estar evitado, através da respetiva valorização monetária.
-a partir dos valores de referência para os vários gases contabilizados, atualizados para 2022 15, com base nas fontes: Bickel et al.
-(2006), Nash and others (2003), Sartori et al.
-(2014).
+a partir dos valores de referência para os vários gases contabilizados, atualizados para 2022 15, com base nas fontes: Bickel et al. (2006), Nash and others (2003), Sartori et al. (2014).
 
@@
 
-A Tabela 4 apresenta o número de viagens diárias do cenário base e potenciais, passíveis de serem realizadas
-até 5 km em bicicleta em complemento com TP, por tipo de percurso. Para as redes segura e direta, as viagens
-do cenário base até 5 km em bicicleta + TP correspondem a 20% do total de viagens reportadas no inquérito.
+A Tabela 4 apresenta o número de viagens diárias do cenário base e potenciais, passíveis de serem realizadas até 5 km em bicicleta em complemento com TP, por tipo de percurso.
+Para as redes segura e direta, as viagens do cenário base até 5 km em bicicleta + TP correspondem a 20% do total de viagens reportadas no inquérito.
 
-Este cenário valoriza a utilização da bicicleta como complemento ao transporte público, com potencial de
-aumentar as viagens em TP realizadas na área metropolitana de Lisboa em até 12% (acrescentando às 825 mil
-do IMob 2017).
-Estes resultados sugerem que o potencial de transferência de viagens em automóvel para bicicleta + TP poderá
-ser quase ou tão importante quanto a transferência para viagens utilizando apenas a bicicleta.
+Este cenário valoriza a utilização da bicicleta como complemento ao transporte público, com potencial de aumentar as viagens em TP realizadas na área metropolitana de Lisboa em até 12% (acrescentando às 825 mil do IMob 2017).
+Estes resultados sugerem que o potencial de transferência de viagens em automóvel para bicicleta + TP poderá ser quase ou tão importante quanto a transferência para viagens utilizando apenas a bicicleta.
 
-Para este cenário foram estimados os impactes ambientais de transferência do automóvel para os transportes
-públicos, para além dos impactes ambientais e sociais da transferência para a bicicleta
+Para este cenário foram estimados os impactes ambientais de transferência do automóvel para os transportes públicos, para além dos impactes ambientais e sociais da transferência para a bicicleta
 
 Foram então calculadas as emissões dos poluentes para cada viagem substituída, tendo em conta o modo de transporte, distâncias e velocidade.
-A tabela 11 apresenta a estimativa anual de gases emitidos e evitados por transferência do automóvel para os outros modos de transporte público, como exemplo para o cenário com a meta de 4%, usando a rede ciclável “direta”.
+A tabela 11 apresenta a estimativa anual de gases emitidos e evitados por transferência do automóvel para os outros modos de transporte público, como exemplo para o cenário com a meta de 4%, usando a rede ciclável "direta".
 
-[Tabela 11]
+\[Tabela 11\]
 
@@
 
-A valorização monetária das emissões (em toneladas) é apresentada na tabela seguinte, para o cenário 3 (apenas a segunda parte da viagem, em transportes públicos), com as metas de 4% e 10%, e usando as redes cicláveis “direta” e “segura”, para 365 dias (1 ano).
+A valorização monetária das emissões (em toneladas) é apresentada na tabela seguinte, para o cenário 3 (apenas a segunda parte da viagem, em transportes públicos), com as metas de 4% e 10%, e usando as redes cicláveis "direta" e "segura", para 365 dias (1 ano).
 
-[Tabela 13]
+\[Tabela 13\]
 
 <!-- Meta ENMAC Cenário Tipo de Percurso    Total Viagens /dia  Viagens Bicicleta Base  Viagens Bicicleta Potencial CO~2~eq evitado (ton/ano)   Benefícios estimados em 10 anos (milhares €) -->
+
 <!-- 4% intermodal até 5km  seguro  1 077 028   4 624   40 770  5 920   230 270 -->
+
 <!-- 4% intermodal até 5km  direto  1 001 761   4 547   37 889  6 011   223 720 -->
+
 <!-- 10%    intermodal até 5km  seguro  1 077 028   4 624   104 647 15 192  591 790 -->
+
 <!-- 10%    intermodal até 5km  direto  1 001 761   4 547   97 218  15 414  574 200 -->
@@
 <!-- Tipo de Percurso   Meta ENMAC  Total   Barco   Comboio Autocarro   Metro superfície    CO~2~eq CO  PM10    NOx VOC Valorização monetária (€) -->
+
 <!-- seguro 4%  192 214 5 167   169 984 4 163   12 900  8 593   17  1.9 27  0.8 1 424 620 -->
+
 <!-- seguro 10% 224 152 5 594   197 857 5 042   15 659  20 627  42  4.6 65  2.0 3 430 552 -->
+
 <!-- direto 4%  189 846 5 795   171 534 4 879   7 639   8 702   18  2.0 28  0.8 1 452 820 -->
+
 <!-- direto 10% 219 511 6 263   198 373 5 806   9 069   20 793  42  4.7 66  1.9 3 487 282 -->
@@
 ### Acknowledgements {.unnumbered}
+
 Please place your acknowledgments at the end of the paper, preceded by an unnumbered run-in heading (i.e. 3rd-level heading).

### 2023-09-27 paper/PaperTRA/PaperTRA.Rmd
papertra knit and minor updates

@@
 classoption: "runningheads"
-title: "Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
-shorttitle: "Estimating the SE impacts of car substitution by bicycle and public transit"
+title: "biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
+shorttitle: "biclaR: SE impacts of car substitution by bicycle and public transit"
 
@@
 
-*biclar* builds on the Propensity to Cycle Tool (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
+*biclaR* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
 The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
-The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^2]
+The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^3]
 However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.
 
-[^2]:  See <https://www.npt.scot> and <https://cruse.bike> for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.
+[^2]: See [pct.bike](https://www.pct.bike/)
+
+[^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.
 
@@
 `r5r` can calculate multi-modal routes using PT combined with other modes.
-It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^3] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
+It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^4] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
 The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.
 
-[^3]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)
+[^4]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)
 
@@
 A digital elevation model from the European Space Agency's COPERNICUS mission, with a 25m spatial resolution, was used to include street gradient information, as a weight in cycling routing.
-The cycling potential trips for the two ENMAC targets (4% and 10%) was estimated from the values for cycling and car trips (both as a driver and as a passenger) from the 2017 base scenario.
+The cycling potential trips for the two national strategic targets (4% and 10%) was estimated from the values for cycling and car trips (both as a driver and as a passenger) from the 2017 base scenario.
 
@@
 Socio-environmental impacts were estimated for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
-using the HEAT for Cycling and the HEAT as a Service[^4] tools, from the WHO.
+using the HEAT for Cycling and the HEAT as a Service[^5] tools, from the WHO.
 
-[^4]: see https://github.com/HEAT-WHO/HEAT_heatr_api
+[^5]: see https://github.com/HEAT-WHO/HEAT_heatr_api
 
@@
 By making the research process publicly accessible in a code repository, this study enables the replication of similar estimates for socio-environmental impacts resulting from a modal shift from cars to bicycles + PT in other metropolitan areas.
+
+The train with most potential (as seen from fig 2)
+
+The information available at the open access website can be downloaded, and used with any GIS software to, for instance, understant which cycling connections have the highest socioenvironmental impacts, in tons of avoided CO2eq emissions, or in long term social costs.
+
 The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.
@@
 
-Please place your acknowledgments at the end of the paper, preceded by an unnumbered run-in heading (i.e. 3rd-level heading).
-
 Thomas Götshi - HAAS.

### 2023-09-27 paper/PaperTRA/PaperTRA.tex
papertra knit and minor updates

@@
 
-\title{Estimating the socio-environmental impacts of car substitution by
-bicycle and public transit using open tools}
+\title{biclaR: Estimating the socio-environmental impacts of car
+substitution by bicycle and public transit using open tools}
 %
-\titlerunning{Estimating the SE impacts of car substitution by bicycle
-and public transit}
+\titlerunning{biclaR: SE impacts of car substitution by bicycle and
+public transit}
 % If the paper title is too long for the running head, you can set
@@
 
-\emph{biclar} builds on the Propensity to Cycle Tool (PCT), a web
-application and research project funded by the UK's Department for
-Transport in 2015 which launched nationally in 2017 as part of the
-government's Cycling and Walking Investment Strategy. The PCT initially
-used only origin-destination data for commuting trips as the basis of
-estimates of cycling potential at zone, route and route network levels
-{[}\textbf{Lovelace2017?}{]}.
+\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
+  \href{https://www.pct.bike/}{pct.bike}} (PCT), a web application and
+research project funded by the UK's Department for Transport in 2015
+which launched nationally in 2017 as part of the government's Cycling
+and Walking Investment Strategy. The PCT initially used only
+origin-destination data for commuting trips as the basis of estimates of
+cycling potential at zone, route and route network levels {[}4{]}. The
+PCT has been extended to include cycling potential for travel to school
+in England {[}5{]} and other trip types in other countries.\footnote{See
+  \href{https://www.npt.scot}{npt.scot} and
+  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
+  Scotland and Ireland that include estimates of cycling for other
+  purposes.} However, to the best of our knowledge, this is the first
+time that the method has been integrated with public transport data
+using multi-modal routing to estimate the potential and benefits of
+multi-stage cycling and PT trips.
 
@@
 significant portion of short-distance trips, a prevalent characteristic
-of active mode travel {[}4{]}. The OD Jittering method breaks down a
+of active mode travel {[}6{]}. The OD Jittering method breaks down a
 single point (i.e., the centroid of an area) into multiple random points
@@
 The selected route choice algorithm was the
-\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}5{]},
+\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}7{]},
 which allows for great flexibility in configuring estimated route types,
 and which proven to provide most accurate route networks for the city of
-Lisbon {[}6{]}. \texttt{r5r} can calculate multi-modal routes using PT
+Lisbon {[}8{]}. \texttt{r5r} can calculate multi-modal routes using PT
 combined with other modes. It enables the identification of the most
@@
 resolution, was used to include street gradient information, as a weight
-in cycling routing. The cycling potential trips for the two ENMAC
-targets (4\% and 10\%) was estimated from the values for cycling and car
-trips (both as a driver and as a passenger) from the 2017 base scenario.
+in cycling routing. The cycling potential trips for the two national
+strategic targets (4\% and 10\%) was estimated from the values for
+cycling and car trips (both as a driver and as a passenger) from the
+2017 base scenario.
 
@@
 
-{[}Tabela 11{]}
+\[Tabela 11\]
 
@@
 
-{[}Tabela 13{]}
+\[Tabela 13\]
 
@@
 socio-environmental impacts resulting from a modal shift from cars to
-bicycles + PT in other metropolitan areas. The provided information on
-socio-economic benefits can support policymakers in prioritizing
-interventions to reduce the reliance on individual motorized
-transportation and effectively communicate their decisions.
+bicycles + PT in other metropolitan areas.
+
+The train with most potential (as seen from fig 2)
+
+The information available at the open access website can be downloaded,
+and used with any GIS software to, for instance, understant which
+cycling connections have the highest socioenvironmental impacts, in tons
+of avoided CO2eq emissions, or in long term social costs.
+
+The provided information on socio-economic benefits can support
+policymakers in prioritizing interventions to reduce the reliance on
+individual motorized transportation and effectively communicate their
+decisions.
 
@@
 \addcontentsline{toc}{subsubsection}{Acknowledgements}
-
-Please place your acknowledgments at the end of the paper, preceded by
-an unnumbered run-in heading (i.e.~3rd-level heading).
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
+Abbas, A., Woodcock, J.: The Propensity to Cycle Tool: An open source
+online system for sustainable transport planning. Journal of Transport
+and Land Use. 10, (2017).
+https://doi.org/\href{https://doi.org/10.5198/jtlu.2016.862}{10.5198/jtlu.2016.862}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
+Berkoff, N., Morgan, M., Abbas, A., Lovelace, R.: Scenarios of cycling
+to school in england, and associated health and carbon impacts:
+Application of the {`}propensity to cycle tool{'}. Journal of Transport
+\& Health. 12, 263--278 (2019).
+https://doi.org/\href{https://doi.org/10.1016/j.jth.2019.01.008}{10.1016/j.jth.2019.01.008}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{6. }%
 \CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
@@
 \leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
-\CSLLeftMargin{5. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring

### 2023-09-27 paper/PaperTRA/PaperTRA.Rmd
finish methods

@@
 
-[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/)
+[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/).
 
@@
 
-[^2]: See [pct.bike](https://www.pct.bike/)
+[^2]: See [pct.bike](https://www.pct.bike/).
 
@@
 
-[^4]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress)
+[^4]: see [docs.conveyal.com/learn-more/traffic-stress](https://docs.conveyal.com/learn-more/traffic-stress).
 
@@
 
-## Assessing socio-environmental benefits - CONTINUE FROM HERE
+## Assessing socio-environmental benefits
 
-Socio-environmental impacts were estimated for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
-using the HEAT for Cycling and the HEAT as a Service[^5] tools, from the WHO.
+<!-- The impacts were assessed for the total metro area. -->
 
-[^5]: see https://github.com/HEAT-WHO/HEAT_heatr_api
+For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.0 [@HEAT] and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5] from the World Health Organization.
+The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
+We considered two dimensions: *social* — including the physical activity of cyclists, air pollution exposure, and road casualties; *environmental* — including CO~2~eq emissions and other pollutants.
 
-Os impactes foram avaliados para cada cenário e para diferentes escalas territoriais:
+[^5]: `HEATaaS` is under development. For more information contact [heatwalkingcycling.org](https://heatwalkingcycling.org).
 
--   Desagregado à escala municipal para cada segmento de rede:
-    -   Ambientais (emissões de CO~2~eq evitadas);
-    -   Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
--   Agregado à escala metropolitana:
-    -   Ambientais (emissões de CO~2~eq evitadas);
-    -   Balanço monetarizado dos impactes socioeconómicos totais (considerando o balanço dos impactes ambientais e sociais);
-    -   Ambientais (emissões de CO~2~eq, CO, PM10, NOx, e VOC11, evitadas por substituição dos modos motorizados, considerando as emissões geradas na transferência para transportes públicos).
+<!-- For the full paper, consider refer the input HEAT values (tabela 6) -->
+<!-- Explain why the use of the HaaS tool instead of the website (3 scenarios, 18 municipalities, 2 routing profiles) -->
+<!-- Also consider put the FE formula and emmission factor values - see report -->
 
-Para todos os cenários, recorreu-se à ferramenta HEAT for cycling, da Organização Mundial de Saúde, para estimativa dos impactes sociais e ambientais da transferência de viagens em automóvel para viagens em bicicleta, nas componentes de: a) Social - Saúde, Atividade Física, Exposição a poluição atmosférica, Exposição ao risco de sinistralidade rodoviária; b) Ambiental - Emissão de gases CO~2~eq.
+For the second leg of the journey, we estimate the additional environmental impacts of shifting car trips to PT (between the PT interfaces).
 
-Additionally, we estimate the impacts of shifting car trips to PT for the second leg of the journey with EMEP/EEA's COPERT methodology and monetize them with the EU Guide to cost-benefit analysis.
+To estimate the car emissions, we used the EMEP/EEA's COPERT software v5 methods and reference values [@COPERT] for a Tier 3 detail level. 
+We used a family-size vehicle, EURO standard, and gasoline or diesel fuel.
+All trips were considered to be made under urban conditions and at an average speed of 15 km/h during rush hour periods.
+Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
+In addition, we assumed an occupancy rate of 1.61 passengers *per* car [@IMOB].
 
-A estimativa dos impactes ambientais resulta em toneladas de CO~2~eq, e a estimativa dos impactes sociais resulta em mortalidade prematura evitada.
-Ambas as unidades são por fim monetizadas em €, segundo os valores da literatura utilizada em estudos semelhantes.
-Para além dos impactes sociais e ambientais resultantes da transferência do automóvel para a bicicleta (na primeira e na última parte da viagem, de e para a interface de TP), estimou-se também o impacte ambiental adicional, resultante da transferência do automóvel para os vários transportes públicos (na segunda etapa da viagem, entre as interfaces).
-Como tal, foi necessário caracterizar o universo dos modos motorizados a serem considerados para os cálculos dos respetivos fatores de emissões de gases poluentes e atmosféricos.
+Regarding the PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014].
 
-Os fatores de consumo e de emissão dos automóveis e transportes públicos deve ser tido em conta relativamente ao número de passageiros transportados (por passageiro.km) e não ao veículo (que seria veículo.km).
-No caso do automóvel, para contabilizar a emissão de gases evitados pela transferência de viagens em automóvel para os transportes públicos, teve-se em conta o valor de referência de 1.61 passageiros/automóvel reportados pelo Inquérito à Mobilidade (INE 2017).
+Emissions were estimated for the following atmospheric pollutants: CO~2~eq, CO, PM10, NOx, and VOC.
+In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally.
 
-As emissões evitadas por cada viagem em transportes públicos que substitui o automóvel correspondem às emissões equivalentes de um automóvel com características correspondentes à média da frota em circulação em Lisboa, para 2 tipos de combustível: gasolina e gasóleo.
-Utilizou-se a metodologia e valores de referência do software COPERT (Ntziachristos and Samaras 2020), v5.0, da Agência Europeia do Ambiente, para um nível de detalhe 3 (Tier 3) na estimativa de consumo de emissões para o automóvel, nestes dois tipos de combustível.
-Optou-se por usar um veículo de dimensão familiar, norma EURO, e tipo de combustível gasolina ou diesel.
-Considerou-se que as viagens foram todas realizadas em condições urbanas (com as respetivas implicações no regime médio de condução) e a uma velocidade média de 15km/h, nos períodos de hora de ponta.
-Uma vez que a distância média percorrida por viagem influencia o nível de sobreconsumo e emissões decorrentes da operação do motor a frio -- ou seja, distâncias maiores diluem a importância deste sobreconsumo face ao fator de consumo com o motor a operar a quente, foram estimados os consumos para diferentes gamas de viagem, em intervalos de 500 metros.
-As emissões são estimadas para os seguintes poluentes atmosféricos: CO (monóxido de carbono), NO X (Óxidos de Azoto), COV (Compostos Orgânicos Voláteis), PM (material particulado).
-Também é estimada as emissões dos principais gases com efeito de estufa: CO2 (dióxido de carbono); CH4 (metano) e N2O (Óxido nitroso), assim como o CO~2~eq equivalente.
+The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases [@EuropeanCommission2014; @bickel2006; @UNITE].
+The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal [@ANSR2021].
+We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a with a discount rate of 5% and inflation of 3%.
 
-Relativamente aos transportes públicos, recorreu-se aos fatores de emissão reportados nos relatórios de sustentabilidade dos respetivos operadores (Carris 2020; Metropolitano de Lisboa 2020; CP 2020, Grupo Transtejo 2014).
+[^6]: See [Statistics Portugal tool for inflation rate estimates between years](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc).
 
-a conversão das emissões evitadas em perda de bem-estar evitado, através da respetiva valorização monetária.
-a partir dos valores de referência para os vários gases contabilizados, atualizados para 2022 15, com base nas fontes: Bickel et al. (2006), Nash and others (2003), Sartori et al. (2014).
+<!-- For the environmental impacts, using conversion values by the ton, suggested by Bickel et al.
+(2006), Nash et al.
+(2003), Sartori et al.
+(2014) : 8.44 €/ton for CO, 2,867.85 €/ton for NOX, 340,969.27 €/ton for PM10, 7,169.62 €/ton for VOC and 35.85 €/ton for CO2eq.
+-->
 
@@
 Foram então calculadas as emissões dos poluentes para cada viagem substituída, tendo em conta o modo de transporte, distâncias e velocidade.
-A tabela 11 apresenta a estimativa anual de gases emitidos e evitados por transferência do automóvel para os outros modos de transporte público, como exemplo para o cenário com a meta de 4%, usando a rede ciclável "direta".
-
-\[Tabela 11\]
 

### 2023-09-27 paper/PaperTRA/PaperTRA.tex
finish methods

@@
 \emph{biclaR}\footnote{See
-  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}}, a
+  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
 decision support tool that facilitates the design and development of a
@@
 \emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
-  \href{https://www.pct.bike/}{pct.bike}} (PCT), a web application and
+  \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
 research project funded by the UK's Department for Transport in 2015
@@
 Stress\footnote{see
-  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}}
+  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
 (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
@@
 
-\hypertarget{assessing-socio-environmental-benefits---continue-from-here}{%
-\subsection{Assessing socio-environmental benefits - CONTINUE FROM
-HERE}\label{assessing-socio-environmental-benefits---continue-from-here}}
+\hypertarget{assessing-socio-environmental-benefits}{%
+\subsection{Assessing socio-environmental
+benefits}\label{assessing-socio-environmental-benefits}}
 
-Socio-environmental impacts were estimated for a short term time horizon
-(i.e., one year) and the long term (i.e., ten years). using the HEAT for
-Cycling and the HEAT as a Service\footnote{see
-  \url{https://github.com/HEAT-WHO/HEAT_heatr_api}} tools, from the WHO.
+For the cycling legs of the journey (first and last legs),
+socio-environmental impacts were estimated, using the HEAT for Cycling
+tool v5.0 {[}9{]} and the
+\href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
+package}\footnote{\texttt{HEATaaS} is under development. For more
+  information contact
+  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.} from
+the World Health Organization. The HEAT tool provided estimates on the
+shifting from car to cycling for a short term time horizon (i.e., one
+year) and the long term (i.e., ten years). We considered two dimensions:
+\emph{social} --- including the physical activity of cyclists, air
+pollution exposure, and road casualties; \emph{environmental} ---
+including CO\textsubscript{2}eq emissions and other pollutants.
 
-Os impactes foram avaliados para cada cenário e para diferentes escalas
-territoriais:
+For the second leg of the journey, we estimate the additional
+environmental impacts of shifting car trips to PT (between the PT
+interfaces).
 
-\begin{itemize}
-\tightlist
-\item
-  Desagregado à escala municipal para cada segmento de rede:
+To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
+methods and reference values {[}10{]} for a Tier 3 detail level. We used
+a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
+trips were considered to be made under urban conditions and at an
+average speed of 15 km/h during rush hour periods. Since the average
+distance traveled per trip influences the overconsumption and emissions
+from cold-start engine operation, we estimated energy and emission
+factors for different ranges of trips at 500-meter intervals. In
+addition, we assumed an occupancy rate of 1.61 passengers \emph{per} car
+{[}1{]}.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Ambientais (emissões de CO\textsubscript{2}eq evitadas);
-  \item
-    Balanço monetarizado dos impactes socioeconómicos totais
-    (considerando o balanço dos impactes ambientais e sociais);
-  \end{itemize}
-\item
-  Agregado à escala metropolitana:
+Regarding the PT, we considered the emission factor values reported in
+the environmental and sustainability reports of the PT operators in the
+LMA {[}11--14{]}.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    Ambientais (emissões de CO\textsubscript{2}eq evitadas);
-  \item
-    Balanço monetarizado dos impactes socioeconómicos totais
-    (considerando o balanço dos impactes ambientais e sociais);
-  \item
-    Ambientais (emissões de CO\textsubscript{2}eq, CO, PM10, NOx, e
-    VOC11, evitadas por substituição dos modos motorizados, considerando
-    as emissões geradas na transferência para transportes públicos).
-  \end{itemize}
-\end{itemize}
+Emissions were estimated for the following atmospheric pollutants:
+CO\textsubscript{2}eq, CO, PM10, NOx, and VOC. In particular, for the
+urban train and tram -- with 100\% electric traction -- only
+CO\textsubscript{2}eq emissions were considered (resulting from the
+production of electricity, considering a ``well-to-tank'' approach),
+since the other pollutants are not emitted locally.
 
-Para todos os cenários, recorreu-se à ferramenta HEAT for cycling, da
-Organização Mundial de Saúde, para estimativa dos impactes sociais e
-ambientais da transferência de viagens em automóvel para viagens em
-bicicleta, nas componentes de: a) Social - Saúde, Atividade Física,
-Exposição a poluição atmosférica, Exposição ao risco de sinistralidade
-rodoviária; b) Ambiental - Emissão de gases CO\textsubscript{2}eq.
-
-Additionally, we estimate the impacts of shifting car trips to PT for
-the second leg of the journey with EMEP/EEA's COPERT methodology and
-monetize them with the EU Guide to cost-benefit analysis.
-
-A estimativa dos impactes ambientais resulta em toneladas de
-CO\textsubscript{2}eq, e a estimativa dos impactes sociais resulta em
-mortalidade prematura evitada. Ambas as unidades são por fim monetizadas
-em €, segundo os valores da literatura utilizada em estudos semelhantes.
-Para além dos impactes sociais e ambientais resultantes da transferência
-do automóvel para a bicicleta (na primeira e na última parte da viagem,
-de e para a interface de TP), estimou-se também o impacte ambiental
-adicional, resultante da transferência do automóvel para os vários
-transportes públicos (na segunda etapa da viagem, entre as interfaces).
-Como tal, foi necessário caracterizar o universo dos modos motorizados a
-serem considerados para os cálculos dos respetivos fatores de emissões
-de gases poluentes e atmosféricos.
-
-Os fatores de consumo e de emissão dos automóveis e transportes públicos
-deve ser tido em conta relativamente ao número de passageiros
-transportados (por passageiro.km) e não ao veículo (que seria
-veículo.km). No caso do automóvel, para contabilizar a emissão de gases
-evitados pela transferência de viagens em automóvel para os transportes
-públicos, teve-se em conta o valor de referência de 1.61
-passageiros/automóvel reportados pelo Inquérito à Mobilidade (INE 2017).
-
-As emissões evitadas por cada viagem em transportes públicos que
-substitui o automóvel correspondem às emissões equivalentes de um
-automóvel com características correspondentes à média da frota em
-circulação em Lisboa, para 2 tipos de combustível: gasolina e gasóleo.
-Utilizou-se a metodologia e valores de referência do software COPERT
-(Ntziachristos and Samaras 2020), v5.0, da Agência Europeia do Ambiente,
-para um nível de detalhe 3 (Tier 3) na estimativa de consumo de emissões
-para o automóvel, nestes dois tipos de combustível. Optou-se por usar um
-veículo de dimensão familiar, norma EURO, e tipo de combustível gasolina
-ou diesel. Considerou-se que as viagens foram todas realizadas em
-condições urbanas (com as respetivas implicações no regime médio de
-condução) e a uma velocidade média de 15km/h, nos períodos de hora de
-ponta. Uma vez que a distância média percorrida por viagem influencia o
-nível de sobreconsumo e emissões decorrentes da operação do motor a frio
--- ou seja, distâncias maiores diluem a importância deste sobreconsumo
-face ao fator de consumo com o motor a operar a quente, foram estimados
-os consumos para diferentes gamas de viagem, em intervalos de 500
-metros. As emissões são estimadas para os seguintes poluentes
-atmosféricos: CO (monóxido de carbono), NO X (Óxidos de Azoto), COV
-(Compostos Orgânicos Voláteis), PM (material particulado). Também é
-estimada as emissões dos principais gases com efeito de estufa: CO2
-(dióxido de carbono); CH4 (metano) e N2O (Óxido nitroso), assim como o
-CO\textsubscript{2}eq equivalente.
-
-Relativamente aos transportes públicos, recorreu-se aos fatores de
-emissão reportados nos relatórios de sustentabilidade dos respetivos
-operadores (Carris 2020; Metropolitano de Lisboa 2020; CP 2020, Grupo
-Transtejo 2014).
-
-a conversão das emissões evitadas em perda de bem-estar evitado, através
-da respetiva valorização monetária. a partir dos valores de referência
-para os vários gases contabilizados, atualizados para 2022 15, com base
-nas fontes: Bickel et al.~(2006), Nash and others (2003), Sartori et
-al.~(2014).
+The conversion of avoided emissions into avoided welfare loss and
+respective monetary valuation was based on the EU Guide to Cost-benefit
+Analysis {[}15{]} and the best up-to-date reference values for the
+various gases {[}15--17{]}. The social impacts are in avoided premature
+mortality. This result is finally monetized using the \emph{Statistical
+Value of Life} for Portugal {[}18{]}. We updated all the monetary
+reference values of the literature based on the annual inflation rate in
+Portugal for 2022\footnote{See
+  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
+  Portugal tool for inflation rate estimates between years}.}, and our
+10-years estimations assumed a with a discount rate of 5\% and inflation
+of 3\%.
 
@@
 substituída, tendo em conta o modo de transporte, distâncias e
-velocidade. A tabela 11 apresenta a estimativa anual de gases emitidos e
-evitados por transferência do automóvel para os outros modos de
-transporte público, como exemplo para o cenário com a meta de 4\%,
-usando a rede ciclável ``direta''.
-
-\[Tabela 11\]
+velocidade.
 
@@
 \CSLLeftMargin{1. }%
-\CSLRightInline{INE: Mobilidade e funcionalidade do território nas
-{Áreas Metropolitanas do Porto e de Lisboa}: 2017,
-\url{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt},
-(2018).}
+\CSLRightInline{INE:
+\href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}{Mobilidade
+e funcionalidade do território nas {Áreas Metropolitanas do Porto e de
+Lisboa}: 2017}. {Instituto National de Estatística}, Lisboa (2018).}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
+Fernandez, A., Brand, C., Rojas Rueda, D., Woodcock, J., Kelly, P.,
+Lieb, C., Oja, P., others:
+\href{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}{Health
+economic assessment tool ({HEAT}) for walking and for cycling: Methods
+and user guide on physical activity, air pollution, injuries and carbon
+impact assessments}. (2017).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
+emission inventory guidebook 2019,
+\url{https://www.emisia.com/utilities/copert/documentation/}, (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{Carris:
+\href{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}{Relatório
+de sustentabilidade 2019 - demostração não financeira}. {Carris -
+Companhia Carris de Ferro de Lisboa, E.M., S.A.} (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{Metropolitano de Lisboa:
+\href{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}{Relatório
+integrado 2019}. {Metropolitano de Lisboa, E.P.E.} (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
+\CSLLeftMargin{13. }%
+\CSLRightInline{CP:
+\href{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}{Relatório
+de sustentabilidade 2019}. {CP - Comboios de Portugal, E.P.E.} (2020).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{Transtejo:
+\href{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}{Relatório
+de sustentabilidade 2014}. {Grupo Transtejo, S.A.} (2014).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
+\CSLLeftMargin{15. }%
+\CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
+Sirtori, E., Vignetti, S., Del Bo, C.:
+\href{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}{Guide
+to cost-benefit analysis of investment projects. Economic apraisal tool
+for cohesion policy 2014-2020}. {European Commission - Directorate
+General for Regional and Urban Policy} (2014).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
+\CSLLeftMargin{16. }%
+\CSLRightInline{Bickel, P., Friedrich, R., Burgess, A., Fagiani, P.,
+Hunt, A., Jong, G.D., Laird, J.:
+\href{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}{{HEATCO
+- Developing Harmonised European Approaches for Transport Costing and
+Project Assessment. Deliverable 5, Proposal for Harmonised Guidelines.}}
+(2006).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{Nash, C., others:
+\href{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}{UNIfication
+of accounts and marginal costs for transport efficiency: Final report}.
+{Institute for Transport Studies, University of Leeds} (2003).}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
+\CSLLeftMargin{18. }%
+\CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
+\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
+Económico e Social da Sinistralidade Rodoviária em Portugal}}. {CEGE -
+Centro de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
+Rodoviária (ANSR)} (2021).}
+
 \end{CSLReferences}

### 2023-09-27 paper/PaperTRA/PaperTRA.Rmd
paper table 1

@@
   # - \usepackage[hidelinks]{hyperref} #this is to remove the annoying boxes. but how to make them blue?
+  - \usepackage{booktabs}
+  - \usepackage{siunitx}
   - \usepackage{hyperref}
@@
                       out.extra = "")  # This forces knitr to label all figures.
+library(dplyr)
 ```
@@
 
-A Tabela 4 apresenta o número de viagens diárias do cenário base e potenciais, passíveis de serem realizadas até 5 km em bicicleta em complemento com TP, por tipo de percurso.
-Para as redes segura e direta, as viagens do cenário base até 5 km em bicicleta + TP correspondem a 20% do total de viagens reportadas no inquérito.
+Table \ref{tab:summary1} presents the LMA total daily trips, the cycling combined with PT trips in the baseline scenario and corresponding potential trips to achieve the national strategy targets (4% and 10%), for different route profiles.
+For the cycling legs of the journey (first and last legs), the environmental avoided emmissions and monetized socio-environment benefits are also presented, resulting from replacing car trips with cycling.
 
-Este cenário valoriza a utilização da bicicleta como complemento ao transporte público, com potencial de aumentar as viagens em TP realizadas na área metropolitana de Lisboa em até 12% (acrescentando às 825 mil do IMob 2017).
-Estes resultados sugerem que o potencial de transferência de viagens em automóvel para bicicleta + TP poderá ser quase ou tão importante quanto a transferência para viagens utilizando apenas a bicicleta.
-
-Para este cenário foram estimados os impactes ambientais de transferência do automóvel para os transportes públicos, para além dos impactes ambientais e sociais da transferência para a bicicleta
-
-Foram então calculadas as emissões dos poluentes para cada viagem substituída, tendo em conta o modo de transporte, distâncias e velocidade.
-
-Para esse cenário com essas características, é estimada uma poupança de 8 702 toneladas de CO~2~eq por via da transferência de viagens motorizadas com combustíveis fósseis e eletricidade, numa perspetiva de ciclo de vida da produção de eletricidade.
-
-A valorização monetária das emissões (em toneladas) é apresentada na tabela seguinte, para o cenário 3 (apenas a segunda parte da viagem, em transportes públicos), com as metas de 4% e 10%, e usando as redes cicláveis "direta" e "segura", para 365 dias (1 ano).
-
-\[Tabela 13\]
+For both *direct* and *safe* route profiles, 20% of the daily car trips have the potential to be replaced by a combination of PT and cycling (up to 5 km on bike).
 
@@
 
-```{r summary1, out.width="100%", fig.cap="Summary of the cycling potencial of intermodality scenario."}
+```{r summary1old, eval=FALSE, out.width="100%", fig.cap="Summary of the cycling potencial of intermodality scenario."}
 knitr::include_graphics("img/table1.png", error = FALSE)
 ```
+```{r summary1}
+summary1 = readRDS("load/summary1.Rds")
+
+summary1 = summary1 %>%
+  mutate(LTS = recode(LTS, `3` = "safe", `4` = "direct")) |> 
+  mutate(Estrategia = paste0(Estrategia*100,"%"))
+
+knitr::kable(summary1,
+      digits = 0,
+      format.args = list(big.mark = " "),
+      col.names = c("Target", "Routing", "Total trips",
+                    "Baseline Cycling + PT", "Potencial Cycling + PT",
+                    "Avoided CO2eq (ton/yr)", "SE Benefits for 10 years (thousand €)"),
+      caption = "\\label{summary1old}Summary of the cycling potencial of intermodality scenario.",
+      format="latex",
+      booktabs = TRUE) |> 
+  kableExtra::column_spec(column = c(4:7), width = "6em")
+  # kableExtra::kable_styling(full_width = TRUE)
+
+```
+
+This scenario unveils the potential of bicycles as a complementary mode of PT, with the prospective to uptake the number of PT trips within the LMA area by as much as 12% (in addition to the 825,000 reported in the mobility survey).
+These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips.
+
+A valorização monetária das emissões (em toneladas) é apresentada na tabela seguinte, para o cenário 3 (apenas a segunda parte da viagem, em transportes públicos), com as metas de 4% e 10%, e usando as redes cicláveis "direta" e "segura", para 365 dias (1 ano).
+
+<!-- this co2 needs to be added to the second table and then discuss -->
+We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity. These estimations used an electricity production life cycle approach.
+
+impactes ambientais de transferência do automóvel para os transportes públicos (tabela 5), para além dos impactes ambientais e sociais da transferência para a bicicleta (tabela 4)
 
@@
 
+The estimated socio-economic impacts stemming from the avoided pollutants and greenhouse gases vary from €1.4 million to €3.5 million.
+Our findings suggest that a bicycle and PT combination could viably replace 20% of current trips, with an additional 12% of PT journeys prone to further substitution.
+The adoption of cycling for first-and-last-mile travel can curtail annual CO2 equivalent emissions by 6,000 to 15,000 tons per day.
+Over a decade, the resulting socio-environmental benefits could accrue to an estimated range of €230 million to €590 million, contingent upon the targeted cycling modalities.
+Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO2 equivalent emissions to 8,500 to 20,800 tons annually, translating to a monetary value of €1.4 million to €3.5 million over a decade.
+Trains exhibit the highest potential for substitution, accounting for 88% of this benefit.
+
 # Conclusion
 
-By making the research process publicly accessible in a code repository, this study enables the replication of similar estimates for socio-environmental impacts resulting from a modal shift from cars to bicycles + PT in other metropolitan areas.
+By making the research process publicly accessible in a code repository, this study enables the replication of similar estimates for socio-environmental impacts resulting from a modal shift from car to bicycle in combination with PT in other metropolitan areas.
 
@@
 
+Different routing profiles for different city cycling maturity levels [@felix2017].
+
 The information available at the open access website can be downloaded, and used with any GIS software to, for instance, understant which cycling connections have the highest socioenvironmental impacts, in tons of avoided CO2eq emissions, or in long term social costs.
 
-The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and effectively communicate their decisions.
+The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and better communicate their decisions.
 

### 2023-09-27 paper/PaperTRA/PaperTRA.tex
paper table 1

@@
 
+\usepackage{booktabs}
+\usepackage{siunitx}
 \usepackage{hyperref}
 \hypersetup{colorlinks = TRUE,  urlcolor = blue, linkcolor = blue, citecolor = blue}
+
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
@@
 
-A Tabela 4 apresenta o número de viagens diárias do cenário base e
-potenciais, passíveis de serem realizadas até 5 km em bicicleta em
-complemento com TP, por tipo de percurso. Para as redes segura e direta,
-as viagens do cenário base até 5 km em bicicleta + TP correspondem a
-20\% do total de viagens reportadas no inquérito.
+Table \ref{tab:summary1} presents the LMA total daily trips, the cycling
+combined with PT trips in the baseline scenario and corresponding
+potential trips to achieve the national strategy targets (4\% and 10\%),
+for different route profiles. For the cycling legs of the journey (first
+and last legs), the environmental avoided emmissions and monetized
+socio-environment benefits are also presented, resulting from replacing
+car trips with cycling.
 
-Este cenário valoriza a utilização da bicicleta como complemento ao
-transporte público, com potencial de aumentar as viagens em TP
-realizadas na área metropolitana de Lisboa em até 12\% (acrescentando às
-825 mil do IMob 2017). Estes resultados sugerem que o potencial de
-transferência de viagens em automóvel para bicicleta + TP poderá ser
-quase ou tão importante quanto a transferência para viagens utilizando
-apenas a bicicleta.
+For both \emph{direct} and \emph{safe} route profiles, 20\% of the daily
+car trips have the potential to be replaced by a combination of PT and
+cycling (up to 5 km on bike).
 
-Para este cenário foram estimados os impactes ambientais de
-transferência do automóvel para os transportes públicos, para além dos
-impactes ambientais e sociais da transferência para a bicicleta
+\begin{table}
 
-Foram então calculadas as emissões dos poluentes para cada viagem
-substituída, tendo em conta o modo de transporte, distâncias e
-velocidade.
+\caption{\label{tab:summary1}\label{summary1old}Summary of the cycling potencial of intermodality scenario.}
+\centering
+\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}}
+\toprule
+Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT & Avoided CO2eq (ton/yr) & SE Benefits for 10 years (thousand €)\\
+\midrule
+4\% & safe & 1 077 028 & 4 624 & 40 770 & 5 920 & 230 270\\
+4\% & direct & 1 001 761 & 4 547 & 37 889 & 6 011 & 223 720\\
+10\% & safe & 1 077 028 & 4 624 & 104 647 & 15 192 & 591 790\\
+10\% & direct & 1 001 761 & 4 547 & 97 218 & 15 414 & 574 200\\
+\bottomrule
+\end{tabular}
+\end{table}
 
-Para esse cenário com essas características, é estimada uma poupança de
-8 702 toneladas de CO\textsubscript{2}eq por via da transferência de
-viagens motorizadas com combustíveis fósseis e eletricidade, numa
-perspetiva de ciclo de vida da produção de eletricidade.
+This scenario unveils the potential of bicycles as a complementary mode
+of PT, with the prospective to uptake the number of PT trips within the
+LMA area by as much as 12\% (in addition to the 825,000 reported in the
+mobility survey). These findings indicate that transferring car trips to
+a combination of bicycle and PT could be nearly as substantial, if not
+equally significant, as the shift towards bicycle-only trips.
 
@@
 
-\[Tabela 13\]
+We estimate a reduction of 8,702 tons of CO2eq emissions by substituting
+motorized trips relying on fossil fuels and electricity. These
+estimations used an electricity production life cycle approach.
 
-\begin{figure}
-\includegraphics[width=1\linewidth,]{img/table1} \caption{Summary of the cycling potencial of intermodality scenario.}\label{fig:summary1}
-\end{figure}
+impactes ambientais de transferência do automóvel para os transportes
+públicos (tabela 5), para além dos impactes ambientais e sociais da
+transferência para a bicicleta (tabela 4)
 
@@
 
+The estimated socio-economic impacts stemming from the avoided
+pollutants and greenhouse gases vary from €1.4 million to €3.5 million.
+Our findings suggest that a bicycle and PT combination could viably
+replace 20\% of current trips, with an additional 12\% of PT journeys
+prone to further substitution. The adoption of cycling for
+first-and-last-mile travel can curtail annual CO2 equivalent emissions
+by 6,000 to 15,000 tons per day. Over a decade, the resulting
+socio-environmental benefits could accrue to an estimated range of €230
+million to €590 million, contingent upon the targeted cycling
+modalities. Regarding the PT segment, the shift from private car usage
+would lead to the mitigation of CO2 equivalent emissions to 8,500 to
+20,800 tons annually, translating to a monetary value of €1.4 million to
+€3.5 million over a decade. Trains exhibit the highest potential for
+substitution, accounting for 88\% of this benefit.
+
 \hypertarget{conclusion}{%
@@
 this study enables the replication of similar estimates for
-socio-environmental impacts resulting from a modal shift from cars to
-bicycles + PT in other metropolitan areas.
+socio-environmental impacts resulting from a modal shift from car to
+bicycle in combination with PT in other metropolitan areas.
 
 The train with most potential (as seen from fig 2)
+
+Different routing profiles for different city cycling maturity levels
+{[}19{]}.
 
@@
 policymakers in prioritizing interventions to reduce the reliance on
-individual motorized transportation and effectively communicate their
+individual motorized transportation and better communicate their
 decisions.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
+\CSLLeftMargin{19. }%
+\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
+cyclists: Review of market segmentation methods for planning practice.
+Transportation Research Record. 2662, 125--133 (2017).
+https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}
+
 \end{CSLReferences}

### 2023-09-27 paper/PaperTRA/PaperTRA.Rmd
bug values TP total
try to understand why there are only max new 104k trips in scenario 3, but there are 224k trips that can be replaced cyc + TP

@@
 The intermodality scenario considers trips combining PT and cycling for the first and last legs.
-In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km or up to 25 minutes on bike.
+In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for instance: 1 km from origin to interface A plus 4 km from interface B to destination) or up to 25 minutes on bike.
 Furthermore, we have imposed restrictions on PT usage to include only trips with no PT transfers, and up to 2 hours (120 min).
@@
 
-Table \ref{tab:summary1} presents the LMA total daily trips, the cycling combined with PT trips in the baseline scenario and corresponding potential trips to achieve the national strategy targets (4% and 10%), for different route profiles.
+Table \ref{tab:summary1} presents the LMA total daily trips, the cycling combined with PT trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles.
 For the cycling legs of the journey (first and last legs), the environmental avoided emmissions and monetized socio-environment benefits are also presented, resulting from replacing car trips with cycling.
@@
 For both *direct* and *safe* route profiles, 20% of the daily car trips have the potential to be replaced by a combination of PT and cycling (up to 5 km on bike). 
-
-<!-- Meta ENMAC Cenário Tipo de Percurso    Total Viagens /dia  Viagens Bicicleta Base  Viagens Bicicleta Potencial CO~2~eq evitado (ton/ano)   Benefícios estimados em 10 anos (milhares €) -->
-
-<!-- 4% intermodal até 5km  seguro  1 077 028   4 624   40 770  5 920   230 270 -->
-
-<!-- 4% intermodal até 5km  direto  1 001 761   4 547   37 889  6 011   223 720 -->
-
-<!-- 10%    intermodal até 5km  seguro  1 077 028   4 624   104 647 15 192  591 790 -->
-
-<!-- 10%    intermodal até 5km  direto  1 001 761   4 547   97 218  15 414  574 200 -->
+<!-- place in the end - here is not clear -->
 
@@
                     "Avoided CO2eq (ton/yr)", "SE Benefits for 10 years (thousand €)"),
-      caption = "\\label{summary1old}Summary of the cycling potencial of intermodality scenario.",
+      caption = "\\label{summary1}Summary of the cycling potencial of intermodality scenario.",
       format="latex",
@@
 
-This scenario unveils the potential of bicycles as a complementary mode of PT, with the prospective to uptake the number of PT trips within the LMA area by as much as 12% (in addition to the 825,000 reported in the mobility survey).
+This scenario unveils the potential of bicycles as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 12% (in addition to the 825,000 reported in the mobility survey).
 These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips.
@@
 
-```{r summary2, out.width="100%", fig.cap="Potencial de transferência estimado para cada modo de transporte público, bem como as estimativas de consumo de CO~2~eq evitado anualmente por transferência do automóvel."}
+Table \ref{tab:summary21} shows something that doesn't make sense.
+<!-- ???? -->
+
+```{r summary2old, eval=FALSE, out.width="100%", fig.cap="Potencial de transferência estimado para cada modo de transporte público, bem como as estimativas de consumo de CO~2~eq evitado anualmente por transferência do automóvel."}
 knitr::include_graphics("img/table2.png", error = FALSE)
+```
+
+```{r summary21}
+summary2 = readRDS("load/TPtable_aml.Rds")
+
+summary21 = summary2 %>%
+  mutate(LTS = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
+  select(Meta_ENMAC, LTS, Total, BUS, FERRY, RAIL, TRAM)
+
+knitr::kable(summary21,
+      digits = 0,
+      format.args = list(big.mark = " "),
+      col.names = c("Target", "Routing", "Total",
+                    "Bus", "Ferry", "Train", "Tram"),
+      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.",
+      format="latex",
+      booktabs = TRUE)
+  # kableExtra::column_spec(column = c(4:7), width = "6em")
+  # kableExtra::kable_styling(full_width = TRUE)
+
+```
+
+```{r summary22}
+
+summary22 = summary2 %>%
+  mutate(LTS = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
+  select(Meta_ENMAC, LTS, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, money_emissions_eur)
 ```

### 2023-09-27 paper/PaperTRA/intermodality_testbug.R

@@ -1,0 +1,23 @@
+## new values fot number of trips that can be replaced from car to Bike + PT (instead of total trops old+new)
+
+TESTEbug1 = CAR_TP_TRANSF %>%
+  group_by(LTS, mode_TP) %>% 
+  summarise(TP_4 = sum(Bike_new4), TP_10 = sum(Bike_new10)) %>% 
+  ungroup() |> 
+  pivot_longer(c(3,4), names_to = "ENMAC", values_to = "viagens") %>% 
+  pivot_wider(names_from = mode_TP, values_from = viagens) %>% 
+  mutate (Total = rowSums(.[3:6])) %>% 
+  as.data.frame()
+
+TESTEbug2 = CAR_TP_TRANSF %>%
+  group_by(LTS, mode_TP) %>% 
+  summarise(TP_4 = sum(TP_4), TP_10 = sum(TP_10)) %>% 
+  ungroup() |> 
+  pivot_longer(c(3,4), names_to = "ENMAC", values_to = "viagens") %>% 
+  pivot_wider(names_from = mode_TP, values_from = viagens) %>% 
+  mutate (Total = rowSums(.[3:6])) %>% 
+  as.data.frame()
+
+# none of them make sense
+
+# What am I missing here?

### 2023-09-28 paper/PaperTRA/PaperTRA.Rmd
table 22

@@
 [^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.
+
+<!-- For the full paper, explain that we had 3 scenarios, and describe them. But in this paper we focous on the intermodality one. -->
 
@@
   kableExtra::column_spec(column = c(4:7), width = "6em")
-  # kableExtra::kable_styling(full_width = TRUE)
 
@@
 ```{r summary21}
-summary2 = readRDS("load/TPtable_aml.Rds")
+summary2 = readRDS("load/summary2.Rds")
 
 summary21 = summary2 %>%
-  mutate(LTS = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
-  select(Meta_ENMAC, LTS, Total, BUS, FERRY, RAIL, TRAM)
+  mutate(LTS = recode(LTS, "seguro" = "safe", "direto" = "direct")) |> 
+  mutate(Estrategia = paste0(`Meta ENMAC`*100,"%")) |> 
+  select(Estrategia, LTS, Total, Autocarro, Barco, Comboio, `Metro superfície`)
 
@@
       format="latex",
-      booktabs = TRUE)
-  # kableExtra::column_spec(column = c(4:7), width = "6em")
-  # kableExtra::kable_styling(full_width = TRUE)
+      booktabs = TRUE) |> 
+  kableExtra::column_spec(column = c(3:7), width = "4em")
 
 ```
+
+Table \ref{tab:summary22} is fine.
 
@@
 summary22 = summary2 %>%
-  mutate(LTS = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
-  select(Meta_ENMAC, LTS, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, money_emissions_eur)
+  mutate(LTS = recode(LTS, "seguro" = "safe", "direto" = "direct")) |> 
+  mutate(Estrategia = paste0(`Meta ENMAC`*100,"%")) |> 
+  mutate(Value = `Valorização monetária (€)`/1000) |> 
+  select(Estrategia, LTS, CO2eq, CO, PM10, NOx, VOC, Value)
+
+knitr::kable(summary22,
+      digits = c(0,0,0,0,1,0,1,0),
+      format.args = list(big.mark = " "),
+      col.names = c("Target", "Routing", "CO2eq",
+                    "CO", "PM10", "NOx", "VOC", "Value (k€)"),
+      caption = "\\label{summary21}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
+      format="latex",
+      booktabs = TRUE) |> 
+  kableExtra::column_spec(column = c(3:7), width = "3.5em")
 ```
-
-<!-- Tipo de Percurso   Meta ENMAC  Total   Barco   Comboio Autocarro   Metro superfície    CO~2~eq CO  PM10    NOx VOC Valorização monetária (€) -->
-
-<!-- seguro 4%  192 214 5 167   169 984 4 163   12 900  8 593   17  1.9 27  0.8 1 424 620 -->
-
-<!-- seguro 10% 224 152 5 594   197 857 5 042   15 659  20 627  42  4.6 65  2.0 3 430 552 -->
-
-<!-- direto 4%  189 846 5 795   171 534 4 879   7 639   8 702   18  2.0 28  0.8 1 452 820 -->
-
-<!-- direto 10% 219 511 6 263   198 373 5 806   9 069   20 793  42  4.7 66  1.9 3 487 282 -->
 

### 2023-09-28 paper/PaperTRA/PaperTRA.tex
table 22

@@
 our analysis to the first and last legs with a combined length of up to
-5 km or up to 25 minutes on bike. Furthermore, we have imposed
-restrictions on PT usage to include only trips with no PT transfers, and
-up to 2 hours (120 min). Additionally, we have only included PT modes
-that can easily accommodate bicycles, such as trains, ferries, trams,
-and inter-municipal bus lines equipped with bike racks (Figure
-\ref{fig:map1}).
+5 km (for instance: 1 km from origin to interface A plus 4 km from
+interface B to destination) or up to 25 minutes on bike. Furthermore, we
+have imposed restrictions on PT usage to include only trips with no PT
+transfers, and up to 2 hours (120 min). Additionally, we have only
+included PT modes that can easily accommodate bicycles, such as trains,
+ferries, trams, and inter-municipal bus lines equipped with bike racks
+(Figure \ref{fig:map1}).
 
@@
 Table \ref{tab:summary1} presents the LMA total daily trips, the cycling
-combined with PT trips in the baseline scenario and corresponding
-potential trips to achieve the national strategy targets (4\% and 10\%),
-for different route profiles. For the cycling legs of the journey (first
-and last legs), the environmental avoided emmissions and monetized
+combined with PT trips in the baseline scenario and corresponding new
+trips to achieve the national strategy targets (4\% and 10\%), for
+different route profiles. For the cycling legs of the journey (first and
+last legs), the environmental avoided emmissions and monetized
 socio-environment benefits are also presented, resulting from replacing
@@
 
-\caption{\label{tab:summary1}\label{summary1old}Summary of the cycling potencial of intermodality scenario.}
+\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of intermodality scenario.}
 \centering
@@
 This scenario unveils the potential of bicycles as a complementary mode
-of PT, with the prospective to uptake the number of PT trips within the
+of PT, with the potential to uptake the number of PT trips within the
 LMA area by as much as 12\% (in addition to the 825,000 reported in the
@@
 
-\begin{figure}
-\includegraphics[width=1\linewidth,]{img/table2} \caption{Potencial de transferência estimado para cada modo de transporte público, bem como as estimativas de consumo de CO~2~eq evitado anualmente por transferência do automóvel.}\label{fig:summary2}
-\end{figure}
+Table \ref{tab:summary21} shows something that doesn't make sense.
+
+\begin{table}
+
+\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.}
+\centering
+\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{4em}}
+\toprule
+Target & Routing & Total & Bus & Ferry & Train & Tram\\
+\midrule
+4\% & safe & 192 214 & 4 163 & 5 167 & 169 984 & 12 900\\
+10\% & safe & 224 152 & 5 042 & 5 594 & 197 857 & 15 659\\
+4\% & direct & 189 846 & 4 879 & 5 795 & 171 534 & 7 639\\
+10\% & direct & 219 511 & 5 806 & 6 263 & 198 373 & 9 069\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Table \ref{tab:summary22} is fine.
+
+\begin{table}
+
+\caption{\label{tab:summary22}\label{summary21}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.}
+\centering
+\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}r}
+\toprule
+Target & Routing & CO2eq & CO & PM10 & NOx & VOC & Value (k€)\\
+\midrule
+4\% & safe & 8 593 & 17 & 1.9 & 27 & 0.8 & 1 425\\
+10\% & safe & 20 627 & 42 & 4.6 & 65 & 2.0 & 3 431\\
+4\% & direct & 8 702 & 18 & 2.0 & 28 & 0.8 & 1 453\\
+10\% & direct & 20 793 & 42 & 4.7 & 66 & 1.9 & 3 487\\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2023-09-28 paper/PaperTRA/PaperTRA.Rmd
finish writing paper
@Robinlovelace  have a look if feel like.
there is one last sentence in the discussion that i am not quite sure how to write it (tagged).
maybe you need to add something in the acknowledgments?

i still need TO-DO some things, as described on the heading.

@@
 
+<!-- TODO: -->
+<!-- improve sentence in the end of discussion -->
+<!-- update € values in 10yr period with inflation and discount rate (i have that at home) -->
+<!-- understand where the values from table 2 come from -->
+<!-- make sure of the reference cls -->
+
 ```{r setup, include=FALSE}
@@
 
-<!-- **full paper**: 4-6 pages in length (typically up to 3,000 words). -->
+<!-- **full paper**: 4-6 pages in length (typically up to 3,000 words). we have 2849 words before references, or 2278 words without tables -->
 
@@
 The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal [@ANSR2021].
-We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a with a discount rate of 5% and inflation of 3%.
+We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a discount rate of 5% and inflation of 3%.
 
@@
 Table \ref{tab:summary1} presents the LMA total daily trips, the cycling combined with PT trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles.
-For the cycling legs of the journey (first and last legs), the environmental avoided emmissions and monetized socio-environment benefits are also presented, resulting from replacing car trips with cycling.
-
-For both *direct* and *safe* route profiles, 20% of the daily car trips have the potential to be replaced by a combination of PT and cycling (up to 5 km on bike). 
-<!-- place in the end - here is not clear -->
+For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment benefits are also presented, resulting from replacing car trips with cycling.
 
@@
 
-This scenario unveils the potential of bicycles as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 12% (in addition to the 825,000 reported in the mobility survey).
-These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips.
+For both *direct* and *safe* route profiles, 20% of the daily car trips have the potential to be replaced by a combination of PT and cycling (up to 5 km on bike). 
+<!-- place in the end - here is not clear -->
 
-A valorização monetária das emissões (em toneladas) é apresentada na tabela seguinte, para o cenário 3 (apenas a segunda parte da viagem, em transportes públicos), com as metas de 4% e 10%, e usando as redes cicláveis "direta" e "segura", para 365 dias (1 ano).
-
-<!-- this co2 needs to be added to the second table and then discuss -->
-We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity. These estimations used an electricity production life cycle approach.
-
-impactes ambientais de transferência do automóvel para os transportes públicos (tabela 5), para além dos impactes ambientais e sociais da transferência para a bicicleta (tabela 4)
-
-Table \ref{tab:summary21} shows something that doesn't make sense.
+*Table \ref{tab:summary21} shows something that doesn't make sense.*
 <!-- ???? -->
@@
   mutate(Estrategia = paste0(`Meta ENMAC`*100,"%")) |> 
-  select(Estrategia, LTS, Total, Autocarro, Barco, Comboio, `Metro superfície`)
+  select(Estrategia, LTS, Total, Autocarro, Barco, Comboio, `Metro superfície`) |> 
+  arrange(desc(Estrategia))
 
@@
 
-Table \ref{tab:summary22} is fine.
+This scenario unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 12% (in addition to the 825 thousand PT trips reported in the mobility survey).
+These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips.
+
+When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the PT train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.
+
+Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips.
+<!-- We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity.
+These estimations used an electricity production life cycle approach.
+-->
+<!-- this co2 needs to be added to the second table and then discuss -->
+Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly, with trains offering the greatest potential for substitution (88%).
 
@@
   mutate(Value = `Valorização monetária (€)`/1000) |> 
-  select(Estrategia, LTS, CO2eq, CO, PM10, NOx, VOC, Value)
+  select(Estrategia, LTS, CO2eq, CO, PM10, NOx, VOC, Value) |> 
+  arrange(desc(Estrategia))
 
@@
                     "CO", "PM10", "NOx", "VOC", "Value (k€)"),
-      caption = "\\label{summary21}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
+      caption = "\\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
       format="latex",
@@
 
-Em suma, o impacte socioeconómico das emissões de gases poluentes e de gases de efeito estufa pode ser valorizado numa poupança potencial de entre 1.4 a 3.5 milhões €, a acrescentar aos impactes estimados na secção anterior.
+The sum of CO~2~eq avoided emissions from the potential car trips shifted to bike (first-and-last legs) in combination with PT (second leg) in the LMA is presented on Table \ref{tab:summaryall}, for both national cycling strategy targets and routing profiles, and the socio-environmental benefits monetized in €, for a 1-year and 10-year time periods.
 
-The results indicate that 20% of the current trips can be made with the bicycle + PT combination, with an additional 12% of PT trips being potentially replaced.
-Shifting to cycling for the first-and-last mile can reduce annual CO~2~eq emissions by 6,000 to 15,000 tons/day, and the 10-year socio-environmental benefits account for €230 to €590 million, depending on the cycling targets.
-For the PT leg, the transfer from car results in the avoidance of 8,500 to 20,800 tons of CO~2~eq emissions per year, or €1.4 to €3.5 million over 10 years, with trains offering the greatest potential for substitution (88%).
+```{r summaryall, message=FALSE, warning=FALSE}
+summaryall = summary1 |> left_join(summary22) |> 
+  select(c(1, 2, 6, 7, 8, 13)) |> 
+  mutate(CO2eq = co2 + CO2eq,
+         value1 = monet/1000 + Value, # this is NOT CORRECT, need to inflate - see original table
+         value10 = monet + Value*10) |> # this is NOT CORRECT, need to deflate
+  select(Estrategia, LTS, CO2eq, value1, value10)
 
-The estimated socio-economic impacts stemming from the avoided pollutants and greenhouse gases vary from €1.4 million to €3.5 million.
-Our findings suggest that a bicycle and PT combination could viably replace 20% of current trips, with an additional 12% of PT journeys prone to further substitution.
-The adoption of cycling for first-and-last-mile travel can curtail annual CO2 equivalent emissions by 6,000 to 15,000 tons per day.
-Over a decade, the resulting socio-environmental benefits could accrue to an estimated range of €230 million to €590 million, contingent upon the targeted cycling modalities.
-Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO2 equivalent emissions to 8,500 to 20,800 tons annually, translating to a monetary value of €1.4 million to €3.5 million over a decade.
-Trains exhibit the highest potential for substitution, accounting for 88% of this benefit.
+knitr::kable(summaryall,
+      digits = 0,
+      format.args = list(big.mark = " "),
+      col.names = c("Target", "Routing", "Avoided CO2eq (tons)",
+                    "SE Benefits 1yr (k€)", "SE Benefits 10yrs (k€)"),
+      caption = "\\label{summaryall}Summary of the avoided CO2eq emmissions (ton/year) and the estimated social and environmental benefits (monetized in thousand €) by replacing car trips with cycling in combination with PT.",
+      format="latex",
+      booktabs = TRUE) |> 
+  kableExtra::column_spec(column = c(3:5), width = "8em")
+
+```
+
+<!-- correct the € numbers after inflation -->
+Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year, 
+and the 10-year socio-environmental benefits account for €238 to €626 million, depending on the cycling targets.
+
+*Say something about the avoided C02eq emissions in the first and last legs (bike) is about 3/4 of the PT leg avoided emissions, which is (expected?) and should not be overlooked when promoting the use of PT.
+Making the accessibility to PT interfaces more atractive to cycling and providing bicycle-friendly equipments (parking) can reduce more 75% CO2eq emissions than the shifted car to PT trip* (FM/RL help)
+
+Our findings suggest that a bicycle and PT combination could viably replace 20% of current LMA trips, with an additional 12% of PT journeys prone to further substitution. 
+<!-- repeated? -->
 
@@
 
-By making the research process publicly accessible in a code repository, this study enables the replication of similar estimates for socio-environmental impacts resulting from a modal shift from car to bicycle in combination with PT in other metropolitan areas.
+The information on socio-economic benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation, and better communicate their decisions by providing the expected avoided GHG and air pollutant emissions and the monetized socio-economic benefits for short and long terms.
+Different routing profiles enable decision-makers to plan for different bicycle user typologies and/or for different city cycling maturity levels [@felix2017].
 
-The train with most potential (as seen from fig 2)
+The information available at *biclaR* tool -- an open access website -- can be downloaded and used with any GIS software to, for instance, understand which potential cycling connections have the highest socio-environmental impacts, in tons of avoided CO~2~eq emissions, or in long term social benefits.
 
-Different routing profiles for different city cycling maturity levels [@felix2017].
-
-The information available at the open access website can be downloaded, and used with any GIS software to, for instance, understant which cycling connections have the highest socioenvironmental impacts, in tons of avoided CO2eq emissions, or in long term social costs.
-
-The provided information on socio-economic benefits can support policymakers in prioritizing interventions to reduce the reliance on individual motorized transportation and better communicate their decisions.
+By making the research process publicly accessible in a code repository, this research enables the replication of similar estimates for socio-environmental impacts, resulting from a modal shift from car to bicycle in combination with PT, in other metropolitan areas.
 
 ### Acknowledgements {.unnumbered}
-
-Thomas Götshi - HAAS.
+This research was funded by the Lisbon's Metropolitan Department of Transport (TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.), under the *biclaR* Project.
+This work is part of the research activity carried out at Civil Engineering Research and Innovation for Sustainability (CERIS) and has been funded by Fundação para a Ciência e a Tecnologia (FCT), Portugal in the framework of project UIDB/04625/2020.
+The authors thank Thomas Götschi (HEAT for Cycling) for providing access to HaaS tool, which is under development.
 

### 2023-09-28 paper/PaperTRA/PaperTRA.tex
finish writing paper
@Robinlovelace  have a look if feel like.
there is one last sentence in the discussion that i am not quite sure how to write it (tagged).
maybe you need to add something in the acknowledgments?

i still need TO-DO some things, as described on the heading.

@@
   Portugal tool for inflation rate estimates between years}.}, and our
-10-years estimations assumed a with a discount rate of 5\% and inflation
-of 3\%.
+10-years estimations assumed a discount rate of 5\% and inflation of
+3\%.
 
@@
 different route profiles. For the cycling legs of the journey (first and
-last legs), the environmental avoided emmissions and monetized
+last legs), the environmental avoided emissions and monetized
 socio-environment benefits are also presented, resulting from replacing
 car trips with cycling.
-
-For both \emph{direct} and \emph{safe} route profiles, 20\% of the daily
-car trips have the potential to be replaced by a combination of PT and
-cycling (up to 5 km on bike).
 
@@
 
-This scenario unveils the potential of bicycles as a complementary mode
-of PT, with the potential to uptake the number of PT trips within the
-LMA area by as much as 12\% (in addition to the 825,000 reported in the
-mobility survey). These findings indicate that transferring car trips to
-a combination of bicycle and PT could be nearly as substantial, if not
-equally significant, as the shift towards bicycle-only trips.
+For both \emph{direct} and \emph{safe} route profiles, 20\% of the daily
+car trips have the potential to be replaced by a combination of PT and
+cycling (up to 5 km on bike).
 
-A valorização monetária das emissões (em toneladas) é apresentada na
-tabela seguinte, para o cenário 3 (apenas a segunda parte da viagem, em
-transportes públicos), com as metas de 4\% e 10\%, e usando as redes
-cicláveis ``direta'' e ``segura'', para 365 dias (1 ano).
-
-We estimate a reduction of 8,702 tons of CO2eq emissions by substituting
-motorized trips relying on fossil fuels and electricity. These
-estimations used an electricity production life cycle approach.
-
-impactes ambientais de transferência do automóvel para os transportes
-públicos (tabela 5), para além dos impactes ambientais e sociais da
-transferência para a bicicleta (tabela 4)
-
-Table \ref{tab:summary21} shows something that doesn't make sense.
+\emph{Table \ref{tab:summary21} shows something that doesn't make
+sense.}
 
@@
 4\% & safe & 192 214 & 4 163 & 5 167 & 169 984 & 12 900\\
+4\% & direct & 189 846 & 4 879 & 5 795 & 171 534 & 7 639\\
 10\% & safe & 224 152 & 5 042 & 5 594 & 197 857 & 15 659\\
-4\% & direct & 189 846 & 4 879 & 5 795 & 171 534 & 7 639\\
 10\% & direct & 219 511 & 5 806 & 6 263 & 198 373 & 9 069\\
@@
 
-Table \ref{tab:summary22} is fine.
+This scenario unveils the potential of cycling as a complementary mode
+of PT, with the potential to uptake the number of PT trips within the
+LMA area by as much as 12\% (in addition to the 825 thousand PT trips
+reported in the mobility survey). These findings indicate that
+transferring car trips to a combination of bicycle and PT could be
+nearly as substantial, if not equally significant, as the shift towards
+bicycle-only trips.
+
+When comparing the existing PT interfaces (Figure \ref{fig:map1}) with
+the bike routes with highest potential to serve as first and last legs
+(Figure \ref{fig:map2}) it becomes clear that the PT train interfaces
+are the ones that have the highest potential to attract car-to-PT
+substituting trips, if their accessibility by bicycle is improved to be
+safe.
+
+Table \ref{tab:summary22} presents the results of the avoided emissions
+and its monetization for the second leg of the journey, by replacing car
+trips with potential TP trips. Regarding the PT segment, the shift from
+private car usage would lead to the mitigation of CO\textsubscript{2}
+equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4
+million to €3.5 million yearly, with trains offering the greatest
+potential for substitution (88\%).
 
@@
 
-\caption{\label{tab:summary22}\label{summary21}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.}
+\caption{\label{tab:summary22}\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.}
 \centering
@@
 4\% & safe & 8 593 & 17 & 1.9 & 27 & 0.8 & 1 425\\
+4\% & direct & 8 702 & 18 & 2.0 & 28 & 0.8 & 1 453\\
 10\% & safe & 20 627 & 42 & 4.6 & 65 & 2.0 & 3 431\\
-4\% & direct & 8 702 & 18 & 2.0 & 28 & 0.8 & 1 453\\
 10\% & direct & 20 793 & 42 & 4.7 & 66 & 1.9 & 3 487\\
@@
 
-Em suma, o impacte socioeconómico das emissões de gases poluentes e de
-gases de efeito estufa pode ser valorizado numa poupança potencial de
-entre 1.4 a 3.5 milhões €, a acrescentar aos impactes estimados na
-secção anterior.
+The sum of CO\textsubscript{2}eq avoided emissions from the potential
+car trips shifted to bike (first-and-last legs) in combination with PT
+(second leg) in the LMA is presented on Table \ref{tab:summaryall}, for
+both national cycling strategy targets and routing profiles, and the
+socio-environmental benefits monetized in €, for a 1-year and 10-year
+time periods.
 
-The results indicate that 20\% of the current trips can be made with the
-bicycle + PT combination, with an additional 12\% of PT trips being
-potentially replaced. Shifting to cycling for the first-and-last mile
-can reduce annual CO\textsubscript{2}eq emissions by 6,000 to 15,000
-tons/day, and the 10-year socio-environmental benefits account for €230
-to €590 million, depending on the cycling targets. For the PT leg, the
-transfer from car results in the avoidance of 8,500 to 20,800 tons of
-CO\textsubscript{2}eq emissions per year, or €1.4 to €3.5 million over
-10 years, with trains offering the greatest potential for substitution
-(88\%).
+\begin{table}
 
-The estimated socio-economic impacts stemming from the avoided
-pollutants and greenhouse gases vary from €1.4 million to €3.5 million.
+\caption{\label{tab:summaryall}\label{summaryall}Summary of the avoided CO2eq emmissions (ton/year) and the estimated social and environmental benefits (monetized in thousand €) by replacing car trips with cycling in combination with PT.}
+\centering
+\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{8em}>{\raggedleft\arraybackslash}p{8em}>{\raggedleft\arraybackslash}p{8em}}
+\toprule
+Target & Routing & Avoided CO2eq (tons) & SE Benefits 1yr (k€) & SE Benefits 10yrs (k€)\\
+\midrule
+4\% & safe & 14 513 & 1 655 & 244 516\\
+4\% & direct & 14 713 & 1 677 & 238 248\\
+10\% & safe & 35 819 & 4 022 & 626 096\\
+10\% & direct & 36 207 & 4 061 & 609 073\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Shifting from car to cycling + PT can reduce annual
+CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year, and
+the 10-year socio-environmental benefits account for €238 to €626
+million, depending on the cycling targets.
+
+\emph{Say something about the avoided C02eq emissions in the first and
+last legs (bike) is about 3/4 of the PT leg avoided emissions, which is
+(expected?) and should not be overlooked when promoting the use of PT.
+Making the accessibility to PT interfaces more atractive to cycling and
+providing bicycle-friendly equipments (parking) can reduce more 75\%
+CO2eq emissions than the shifted car to PT trip} (FM/RL help)
+
 Our findings suggest that a bicycle and PT combination could viably
-replace 20\% of current trips, with an additional 12\% of PT journeys
-prone to further substitution. The adoption of cycling for
-first-and-last-mile travel can curtail annual CO2 equivalent emissions
-by 6,000 to 15,000 tons per day. Over a decade, the resulting
-socio-environmental benefits could accrue to an estimated range of €230
-million to €590 million, contingent upon the targeted cycling
-modalities. Regarding the PT segment, the shift from private car usage
-would lead to the mitigation of CO2 equivalent emissions to 8,500 to
-20,800 tons annually, translating to a monetary value of €1.4 million to
-€3.5 million over a decade. Trains exhibit the highest potential for
-substitution, accounting for 88\% of this benefit.
+replace 20\% of current LMA trips, with an additional 12\% of PT
+journeys prone to further substitution.
 
@@
 
-By making the research process publicly accessible in a code repository,
-this study enables the replication of similar estimates for
-socio-environmental impacts resulting from a modal shift from car to
-bicycle in combination with PT in other metropolitan areas.
-
-The train with most potential (as seen from fig 2)
-
-Different routing profiles for different city cycling maturity levels
+The information on socio-economic benefits can support policy-makers in
+prioritizing interventions to reduce the reliance on individual
+motorized transportation, and better communicate their decisions by
+providing the expected avoided GHG and air pollutant emissions and the
+monetized socio-economic benefits for short and long terms. Different
+routing profiles enable decision-makers to plan for different bicycle
+user typologies and/or for different city cycling maturity levels
 {[}19{]}.
 
-The information available at the open access website can be downloaded,
-and used with any GIS software to, for instance, understant which
-cycling connections have the highest socioenvironmental impacts, in tons
-of avoided CO2eq emissions, or in long term social costs.
+The information available at \emph{biclaR} tool -- an open access
+website -- can be downloaded and used with any GIS software to, for
+instance, understand which potential cycling connections have the
+highest socio-environmental impacts, in tons of avoided
+CO\textsubscript{2}eq emissions, or in long term social benefits.
 
-The provided information on socio-economic benefits can support
-policymakers in prioritizing interventions to reduce the reliance on
-individual motorized transportation and better communicate their
-decisions.
+By making the research process publicly accessible in a code repository,
+this research enables the replication of similar estimates for
+socio-environmental impacts, resulting from a modal shift from car to
+bicycle in combination with PT, in other metropolitan areas.
 
@@
 
-Thomas Götshi - HAAS.
+This research was funded by the Lisbon's Metropolitan Department of
+Transport (TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.),
+under the \emph{biclaR} Project. This work is part of the research
+activity carried out at Civil Engineering Research and Innovation for
+Sustainability (CERIS) and has been funded by Fundação para a Ciência e
+a Tecnologia (FCT), Portugal in the framework of project
+UIDB/04625/2020. The authors thank Thomas Götschi (HEAT for Cycling) for
+providing access to HaaS tool, which is under development.
 

### 2023-09-28 paper/PaperTRA/PaperTRA.Rmd
update 1yr SE

@@
   mutate(CO2eq = co2 + CO2eq,
-         value1 = monet/1000 + Value, # this is NOT CORRECT, need to inflate - see original table
+         value1 = monet/10 + Value, # this is NOT CORRECT, need to inflate - see original table
          value10 = monet + Value*10) |> # this is NOT CORRECT, need to deflate

### 2023-10-02 paper/PaperTRA/PaperTRA.Rmd
deflation rate

@@
 ```{r summaryall, message=FALSE, warning=FALSE}
+summary0 = readRDS("load/HEAT_aml_results.Rds") |> slice_tail(n = 4)
+rate = ((1+0.03)^10)/((1+0.05)^10) # inflation rate 3% discount rate 5%
+
 summaryall = summary1 |> left_join(summary22) |> 
-  select(c(1, 2, 6, 7, 8, 13)) |> 
+  mutate(Code = c(3304, 3404, 3310, 3410)) |> 
+  left_join(summary0 |> select(Code, Economic), by = "Code") |> 
+  select(c(1, 2, 6, 7, 8, 13, 15)) |> 
   mutate(CO2eq = co2 + CO2eq,
-         value1 = monet/10 + Value, # this is NOT CORRECT, need to inflate - see original table
-         value10 = monet + Value*10) |> # this is NOT CORRECT, need to deflate
+         value1 = Economic/1000 + Value, # this is CORRECT
+         value10 = monet + Value*10*rate) |> # this is CORRECT
   select(Estrategia, LTS, CO2eq, value1, value10)
@@
 
-<!-- correct the € numbers after inflation -->
+<!-- separate the Socio impacts from the Environemtal ones? It shouldn't be ok to sum them -->
 Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year, 
-and the 10-year socio-environmental benefits account for €238 to €626 million, depending on the cycling targets.
+and the 10-year socio-environmental benefits account for €235 million to €620 million, depending on the cycling targets.
 

### 2023-10-02 paper/PaperTRA/PaperTRA.tex
update 1yr SE
deflation rate

@@
 \midrule
-4\% & safe & 14 513 & 1 655 & 244 516\\
-4\% & direct & 14 713 & 1 677 & 238 248\\
-10\% & safe & 35 819 & 4 022 & 626 096\\
-10\% & direct & 36 207 & 4 061 & 609 073\\
+4\% & safe & 14 513 & 27 299 & 242 024\\
+4\% & direct & 14 713 & 26 572 & 235 706\\
+10\% & safe & 35 819 & 69 938 & 620 094\\
+10\% & direct & 36 207 & 67 956 & 602 972\\
 \bottomrule
@@
 CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year, and
-the 10-year socio-environmental benefits account for €238 to €626
-million, depending on the cycling targets.
+the 10-year socio-environmental benefits account for €235 million to
+€620 million, depending on the cycling targets.
 

### 2023-10-02 paper/PaperTRA/PaperTRA.Rmd
discussion + litrev

@@
 <!-- TODO: -->
-<!-- improve sentence in the end of discussion -->
-<!-- update € values in 10yr period with inflation and discount rate (i have that at home) -->
 <!-- understand where the values from table 2 come from -->
+<!-- check the % of social impacts vs. environmental -->
 <!-- make sure of the reference cls -->
+<!-- see where we can cut to 1 less paragraph -->
 
@@
 
-Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can significantly replace private car trips.
+Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can significantly replace private car trips [@MARTENS2007326; @RIETVELD200071].
 This approach requires interventions and programs to make bicycling more appealing, and the resulting public investments can have significant social and environmental benefits.
+<!-- Many studies refer to the potential of cycling and bike-sharing to replace car and PT trips, but not to the potential of serving as first and last miles. -->
 
@@
 
-<!-- For the full paper, explain that we had 3 scenarios, and describe them. But in this paper we focous on the intermodality one. -->
+<!-- For the full paper, explain that we had 3 scenarios, and describe them. But in this paper we focus on the intermodality one. -->
 
@@
 The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.
+Different routing profiles enable decision-makers to plan for different bicycle user typologies and/or for different city cycling maturity levels [@felix2017]. <!-- move this to conclusions? -->
 
@@
 Table \ref{tab:summary1} presents the LMA total daily trips, the cycling combined with PT trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles.
-For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment benefits are also presented, resulting from replacing car trips with cycling.
+For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are also presented, resulting from replacing car trips with cycling.
 
@@
 <!-- place in the end - here is not clear -->
-
 *Table \ref{tab:summary21} shows something that doesn't make sense.*
@@
 
-*Say something about the avoided C02eq emissions in the first and last legs (bike) is about 3/4 of the PT leg avoided emissions, which is (expected?) and should not be overlooked when promoting the use of PT.
-Making the accessibility to PT interfaces more atractive to cycling and providing bicycle-friendly equipments (parking) can reduce more 75% CO2eq emissions than the shifted car to PT trip* (FM/RL help)
+The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 75% of the emissions avoided during the PT segment.
+This finding, while expected -- due the zero cycling emissions, should not be overlooked when promoting the PT use.
+Improving the safe accessibility to PT interfaces to cyclists and providing bicycle-friendly amenities such as parking facilities can potentially lead to a higher reduction in CO~2~eq emissions, compared to a scenario where individuals shift from car travel to car + PT combination.
 
+The social impacts from replacing car trips to bicycle in first-and-last legs, represent 95% of the total socio-environmental benefits estimated for all journey legs.
+<!-- sCheck the % of SOCIAL - original heat tables -->
 Our findings suggest that a bicycle and PT combination could viably replace 20% of current LMA trips, with an additional 12% of PT journeys prone to further substitution. 
@@
 
-The information on socio-economic benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation, and better communicate their decisions by providing the expected avoided GHG and air pollutant emissions and the monetized socio-economic benefits for short and long terms.
-Different routing profiles enable decision-makers to plan for different bicycle user typologies and/or for different city cycling maturity levels [@felix2017].
+The information on socio-economic benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation, and to better communicate their decisions by providing the expected avoided GHG and air pollutant emissions and the monetized socio-economic benefits for short and long terms.
 
 The information available at *biclaR* tool -- an open access website -- can be downloaded and used with any GIS software to, for instance, understand which potential cycling connections have the highest socio-environmental impacts, in tons of avoided CO~2~eq emissions, or in long term social benefits.
+<!-- move to introduction?
+-->
 
@@
 
-### Acknowledgements {.unnumbered}
+### Acknowledgements. {.unnumbered}
 This research was funded by the Lisbon's Metropolitan Department of Transport (TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.), under the *biclaR* Project.

### 2023-10-02 paper/PaperTRA/PaperTRA.tex
discussion + litrev

@@
 Combining public transportation (PT) and cycling for the first and last
-mile in metropolitan areas can significantly replace private car trips.
-This approach requires interventions and programs to make bicycling more
-appealing, and the resulting public investments can have significant
-social and environmental benefits.
+mile in metropolitan areas can significantly replace private car trips
+{[}1, 2{]}. This approach requires interventions and programs to make
+bicycling more appealing, and the resulting public investments can have
+significant social and environmental benefits.
 
-According to the latest mobility survey conducted in 2018 {[}1{]}, the
+According to the latest mobility survey conducted in 2018 {[}3{]}, the
 LMA registered a total of 5.3 million daily trips, with only 0.5\% by
@@
 To achieve the cycling targets set by the Portuguese national cycling
-strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}2{]}, the
+strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}4{]}, the
 Lisbon's Metropolitan Department of Transport introduced
@@
 decision support tool that facilitates the design and development of a
-metropolitan cycling network {[}3{]}.
+metropolitan cycling network {[}5{]}.
 
@@
 origin-destination data for commuting trips as the basis of estimates of
-cycling potential at zone, route and route network levels {[}4{]}. The
+cycling potential at zone, route and route network levels {[}6{]}. The
 PCT has been extended to include cycling potential for travel to school
-in England {[}5{]} and other trip types in other countries.\footnote{See
+in England {[}7{]} and other trip types in other countries.\footnote{See
   \href{https://www.npt.scot}{npt.scot} and
@@
 
-The mobility survey data {[}1{]} is the basis for this project and
+The mobility survey data {[}3{]} is the basis for this project and
 defines the baseline scenario. Despite being conducted in the
@@
 significant portion of short-distance trips, a prevalent characteristic
-of active mode travel {[}6{]}. The OD Jittering method breaks down a
+of active mode travel {[}8{]}. The OD Jittering method breaks down a
 single point (i.e., the centroid of an area) into multiple random points
@@
 The selected route choice algorithm was the
-\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}7{]},
+\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}9{]},
 which allows for great flexibility in configuring estimated route types,
 and which proven to provide most accurate route networks for the city of
-Lisbon {[}8{]}. \texttt{r5r} can calculate multi-modal routes using PT
+Lisbon {[}10{]}. \texttt{r5r} can calculate multi-modal routes using PT
 combined with other modes. It enables the identification of the most
@@
 base scenario for both types of networks: \emph{direct} and \emph{safe},
-using LTS 4 and LTS 3, respectively.
+using LTS 4 and LTS 3, respectively. Different routing profiles enable
+decision-makers to plan for different bicycle user typologies and/or for
+different city cycling maturity levels {[}11{]}.
 
@@
 socio-environmental impacts were estimated, using the HEAT for Cycling
-tool v5.0 {[}9{]} and the
+tool v5.0 {[}12{]} and the
 \href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
@@
 To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
-methods and reference values {[}10{]} for a Tier 3 detail level. We used
+methods and reference values {[}13{]} for a Tier 3 detail level. We used
 a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
@@
 addition, we assumed an occupancy rate of 1.61 passengers \emph{per} car
-{[}1{]}.
+{[}3{]}.
 
@@
 the environmental and sustainability reports of the PT operators in the
-LMA {[}11--14{]}.
+LMA {[}14--17{]}.
 
@@
 respective monetary valuation was based on the EU Guide to Cost-benefit
-Analysis {[}15{]} and the best up-to-date reference values for the
-various gases {[}15--17{]}. The social impacts are in avoided premature
+Analysis {[}18{]} and the best up-to-date reference values for the
+various gases {[}18--20{]}. The social impacts are in avoided premature
 mortality. This result is finally monetized using the \emph{Statistical
-Value of Life} for Portugal {[}18{]}. We updated all the monetary
+Value of Life} for Portugal {[}21{]}. We updated all the monetary
 reference values of the literature based on the annual inflation rate in
@@
 last legs), the environmental avoided emissions and monetized
-socio-environment benefits are also presented, resulting from replacing
-car trips with cycling.
+socio-environment (SE) benefits are also presented, resulting from
+replacing car trips with cycling.
 
@@
 car trips have the potential to be replaced by a combination of PT and
-cycling (up to 5 km on bike).
-
-\emph{Table \ref{tab:summary21} shows something that doesn't make
-sense.}
+cycling (up to 5 km on bike). \emph{Table \ref{tab:summary21} shows
+something that doesn't make sense.}
 
@@
 
-\emph{Say something about the avoided C02eq emissions in the first and
-last legs (bike) is about 3/4 of the PT leg avoided emissions, which is
-(expected?) and should not be overlooked when promoting the use of PT.
-Making the accessibility to PT interfaces more atractive to cycling and
-providing bicycle-friendly equipments (parking) can reduce more 75\%
-CO2eq emissions than the shifted car to PT trip} (FM/RL help)
+The emissions of CO\textsubscript{2}eq that are avoided during both the
+initial and final journey segments account for about 75\% of the
+emissions avoided during the PT segment. This finding, while expected --
+due the zero cycling emissions, should not be overlooked when promoting
+the PT use. Improving the safe accessibility to PT interfaces to
+cyclists and providing bicycle-friendly amenities such as parking
+facilities can potentially lead to a higher reduction in
+CO\textsubscript{2}eq emissions, compared to a scenario where
+individuals shift from car travel to car + PT combination.
 
-Our findings suggest that a bicycle and PT combination could viably
-replace 20\% of current LMA trips, with an additional 12\% of PT
-journeys prone to further substitution.
+The social impacts from replacing car trips to bicycle in first-and-last
+legs, represent 95\% of the total socio-environmental benefits estimated
+for all journey legs. Our findings suggest that a bicycle and PT
+combination could viably replace 20\% of current LMA trips, with an
+additional 12\% of PT journeys prone to further substitution.
 
@@
 prioritizing interventions to reduce the reliance on individual
-motorized transportation, and better communicate their decisions by
+motorized transportation, and to better communicate their decisions by
 providing the expected avoided GHG and air pollutant emissions and the
-monetized socio-economic benefits for short and long terms. Different
-routing profiles enable decision-makers to plan for different bicycle
-user typologies and/or for different city cycling maturity levels
-{[}19{]}.
+monetized socio-economic benefits for short and long terms.
 
@@
 
-\hypertarget{acknowledgements}{%
-\subsubsection*{Acknowledgements}\label{acknowledgements}}
-\addcontentsline{toc}{subsubsection}{Acknowledgements}
+\hypertarget{acknowledgements.}{%
+\subsubsection*{Acknowledgements.}\label{acknowledgements.}}
+\addcontentsline{toc}{subsubsection}{Acknowledgements.}
 
@@
 \begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-MARTENS2007326}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{Martens, K.: Promoting bike-and-ride: The dutch
+experience. Transportation Research Part A: Policy and Practice. 41,
+326--338 (2007).
+https://doi.org/\href{https://doi.org/10.1016/j.tra.2006.09.010}{10.1016/j.tra.2006.09.010}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-RIETVELD200071}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{Rietveld, P.: The accessibility of railway stations: The
+role of the bicycle in the netherlands. Transportation Research Part D:
+Transport and Environment. 5, 71--75 (2000).
+https://doi.org/\href{https://doi.org/10.1016/S1361-9209(99)00019-X}{10.1016/S1361-9209(99)00019-X}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
-\CSLLeftMargin{1. }%
+\CSLLeftMargin{3. }%
 \CSLRightInline{INE:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
-\CSLLeftMargin{2. }%
+\CSLLeftMargin{4. }%
 \CSLRightInline{Presidência do Conselho de Ministros: Resolução do
@@
 \leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
-\CSLLeftMargin{3. }%
+\CSLLeftMargin{5. }%
 \CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{6. }%
 \CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
-\CSLLeftMargin{5. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
@@
 \leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
-\CSLLeftMargin{7. }%
+\CSLLeftMargin{9. }%
 \CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
+cyclists: Review of market segmentation methods for planning practice.
+Transportation Research Record. 2662, 125--133 (2017).
+https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
@@
 \leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{14. }%
 \CSLRightInline{Carris:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{15. }%
 \CSLRightInline{Metropolitano de Lisboa:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{16. }%
 \CSLRightInline{CP:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{17. }%
 \CSLRightInline{Transtejo:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
-\CSLLeftMargin{16. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{Bickel, P., Friedrich, R., Burgess, A., Fagiani, P.,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{Nash, C., others:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
@@
 Rodoviária (ANSR)} (2021).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
-\CSLLeftMargin{19. }%
-\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
-cyclists: Review of market segmentation methods for planning practice.
-Transportation Research Record. 2662, 125--133 (2017).
-https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}
 

### 2023-10-02 paper/PaperTRA/PaperTRA.Rmd
social vs environmental

@@
 csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
+# cls : svproc.csl
 # citations with arabic numbers is prefered when using natbib
@@
 <!-- understand where the values from table 2 come from -->
-<!-- check the % of social impacts vs. environmental -->
-<!-- make sure of the reference cls -->
+<!-- make sure of the reference cls - try the other cls -->
 <!-- see where we can cut to 1 less paragraph -->
+<!-- should we ignore the social benefits component? and only estimate and add the environmental ones? or estimate the car to tp transfer Social benefits? maybe to the next paper? -->
 
@@
 
-The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data agregated and validated.
+The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
 This information is crucial for an accurate PT trip and route estimation.
-A digital elevation model from the European Space Agency's COPERNICUS mission, with a 25m spatial resolution, was used to include street gradient information, as a weight in cycling routing.
-The cycling potential trips for the two national strategic targets (4% and 10%) was estimated from the values for cycling and car trips (both as a driver and as a passenger) from the 2017 base scenario.
+A digital elevation model, from the European Space Agency's COPERNICUS mission, was used to include street gradient information, as a weight in cycling routing.
+<!--  with a 25 m spatial resolution, -->
+The cycling potential trips for the two national strategic targets (4% and 10%) were estimated from the 2017 cycling and car trips (both as a driver and as a passenger), the baseline scenario.
 
@@
 
+The social impacts represent 98% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs.
+<!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper -->
 The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 75% of the emissions avoided during the PT segment.
@@
 
-The social impacts from replacing car trips to bicycle in first-and-last legs, represent 95% of the total socio-environmental benefits estimated for all journey legs.
-<!-- sCheck the % of SOCIAL - original heat tables -->
 Our findings suggest that a bicycle and PT combination could viably replace 20% of current LMA trips, with an additional 12% of PT journeys prone to further substitution. 

### 2023-10-02 paper/PaperTRA/PaperTRA.Rmd
found the bug
add table 2 corrected

@@
 knitr::opts_chunk$set(echo = FALSE,
+                      # root.dir = "paper/PaperTRA/",
                       out.extra = "")  # This forces knitr to label all figures.
+
 library(dplyr)
@@
 
-```{r summary1old, eval=FALSE, out.width="100%", fig.cap="Summary of the cycling potencial of intermodality scenario."}
-knitr::include_graphics("img/table1.png", error = FALSE)
-```
 ```{r summary1}
@@
 
-For both *direct* and *safe* route profiles, 20% of the daily car trips have the potential to be replaced by a combination of PT and cycling (up to 5 km on bike). 
+For both *direct* and *safe* route profiles, 20% of the daily car trips have the potential to be replaced by a combination of PT and cycling (up to 5 km on bike). <!-- NOT CORRECT -->
 <!-- place in the end - here is not clear -->
-*Table \ref{tab:summary21} shows something that doesn't make sense.*
+Table \ref{tab:summary21} shows the PT modes with most potential of  in combination with cycling, by mode.
 <!-- ???? -->
-
-```{r summary2old, eval=FALSE, out.width="100%", fig.cap="Potencial de transferência estimado para cada modo de transporte público, bem como as estimativas de consumo de CO~2~eq evitado anualmente por transferência do automóvel."}
-knitr::include_graphics("img/table2.png", error = FALSE)
-```
 
@@
 
-summary21 = summary2 %>%
-  mutate(LTS = recode(LTS, "seguro" = "safe", "direto" = "direct")) |> 
-  mutate(Estrategia = paste0(`Meta ENMAC`*100,"%")) |> 
-  select(Estrategia, LTS, Total, Autocarro, Barco, Comboio, `Metro superfície`) |> 
-  arrange(desc(Estrategia))
+summary21 = summary21 %>%
+  # mutate(LTS = recode(LTS, "seguro" = "safe", "direto" = "direct")) |> 
+  # mutate(Estrategia = paste0(`Meta ENMAC`*100,"%")) |> 
+  # select(Estrategia, LTS, Total, Autocarro, Barco, Comboio, `Metro superfície`) |> 
+  arrange(desc(Target))
 
-knitr::kable(summary21,
+knitr::kable(summary21 |> select(-Baseline),
       digits = 0,
       format.args = list(big.mark = " "),
-      col.names = c("Target", "Routing", "Total",
-                    "Bus", "Ferry", "Train", "Tram"),
+      # col.names = c("Target", "Routing", "Total",
+      #               "Bus", "Ferry", "Train", "Tram"),
       caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.",
@@
       booktabs = TRUE) |> 
-  kableExtra::column_spec(column = c(3:7), width = "4em")
+  kableExtra::column_spec(column = c(1:7), width = "4.5em")
 
@@
 
-This scenario unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 12% (in addition to the 825 thousand PT trips reported in the mobility survey).
-These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips.
+This scenario unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).
+<!-- These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips.
+WE CANNOT SHOW THIS IN THIS PAPER WITH ONLY SCENARIO 3 DATA -->
 
@@
 Our findings suggest that a bicycle and PT combination could viably replace 20% of current LMA trips, with an additional 12% of PT journeys prone to further substitution. 
-<!-- repeated? -->
+<!-- repeated? not correct -->
 

### 2023-10-02 paper/PaperTRA/intermodality_testbug.R

@@ -21,3 +21,120 @@ TESTEbug2 = CAR_TP_TRANSF %>%
 # none of them make sense
 
 # What am I missing here?
+
+
+# left_join the intermoddality routes with the pt segment
+
+INTERMODALcar_check = routes_r5r_100jit_lts3__intermod_NoSub_elev |>
+  filter(segment == 2) |> 
+  left_join(routes_r5r_100jit_car_TP_lts3_ONLYTPsegment, by = "id")
+
+table(INTERMODALcar_check$mode.x)
+# BUS FERRY  RAIL  TRAM 
+# 474   208 10871   605 
+
+INTERMODALcar_check = routes_r5r_100jit_lts3__intermod_NoSub_elev |>
+  filter(segment == 2) |> 
+  left_join(routes_r5r_100jit_car_TP_lts3_ONLYTPsegment, by = "id") |> 
+  filter(!is.na(mode.y))
+
+# ok, já entendi. nós não calculámos a rota em automóvel só da segunda perna. isto porque a rota do início ao fim pode ser completamente diferente da rota 1 + 2+ 3.
+
+# Mas calculámo sim
+# ver routes_allmodesNSub3_TPonly_preoverline
+são 5731 rotas da leg em TP
+
+TESTEbutTP3 = routes_allmodesNSub3_TPonly_preoverline |> 
+  group_by(mode) |> 
+  summarise(Bike = sum(Bike),
+            Car = sum(Car + CarP),
+            TP = sum(Other),
+            Total = sum(Total),
+            new_cyc4 = sum(new_cyc4),
+            new_cyc10 = sum(new_cyc10))
+table(routes_allmodesNSub3_TPonly_preoverline$mode)
+
+
+# > 3589+573
+# [1] 4162
+# > 4882+285
+# [1] 5167
+# > sum(TESTEbutTP3$Bike)
+# [1] 2311.753
+# > sum(TESTEbutTP3$new_cyc4)
+# [1] 20385.23
+# > sum(TESTEbutTP3$new_cyc4)*2
+# [1] 40770.45
+# > sum(TESTEbutTP3$Total)
+# [1] 538514
+# > sum(TESTEbutTP3$Total)*2
+# [1] 1077028
+
+# O erro está aqui. Duplicámos o número de viagens em bicicleta (2 legs!) quando calculámos o potential cycling para a tabela 1 do paper
+
[…]

### 2023-10-02 code/test-code/heat-prep.R
found the duplicate & re-run HEAT for intermodality AML
not returning any resp output https://github.com/HEAT-WHO/HEAT_heatr_api/issues/11#issuecomment-1743834047

@@ -251,11 +251,11 @@ HEATbind_aml_PPP = HEATbind
 
 
 ## For Intermodality ---------------------------------------------------------------------------
-
+# RUN THIS AGAIN WHEN API GETS LIVE
 
 Code_Hass = CENARIOS$Code[9:12] #fazer por enquanto só AML e senários 1 e 2
 
-HEATbind_intermodal = HEAT
+HEATbind_intermodal_correct = HEAT
 
 for (i in Code_Hass){
   
@@ -271,14 +271,15 @@ for (i in Code_Hass){
     mutate(ENMAC = ENMAC/100)
   
   HEAT_bike = readRDS(HEAT$routes_filepath) %>% 
-    ungroup() %>% 
     st_drop_geometry() %>%
-    group_by(id) %>% 
-    summarise(distance = sum(distance), 
+    ungroup() %>% 
+    group_by(id) %>%  # it shouldn't return two segments again.
+    summarise(distance = sum(distance),  # we need to put both segments distance together!
               Bike = Bike,
               Car = Car + CarP,
               Total = Total,
               Bikeper = Bike / Total) %>% 
+    unique() |> # NOT THE IDEAL BUT WORKS FOR THE PURPUSE (this was missing in the first version 2023)
     ungroup() %>% 
     mutate(
       cyc = ifelse(Bikeper >= HEAT$ENMAC, Bike, HEAT$ENMAC * Total),
@@ -438,7 +439,7 @@ for (i in Code_Hass){
     mutate(value_newcyc = round(Economic10/Bike_new),
            value_newcyc_eur = round(Economic10/Bike_new*usd_to_eur))
   
-  HEATbind_intermodal = rbind(HEATbind_intermodal, HEAT)
+  HEATbind_intermodal_correct = rbind(HEATbind_intermodal_correct, HEAT)
   
   print(paste0(i, " done"))
   
@@ -446,6 +447,6 @@ for (i in Code_Hass){
   
 }
 
-# HEATbind_intermodal = HEATbind_intermodal[-1,]
-# HEATbind_intermodal_PPP = HEATbind_intermodal
-saveRDS(HEATbind_intermodal_PPP, "HEAT/HEAT_AML_intermodal_ppp.Rds")
+# HEATbind_intermodal_correct = HEATbind_intermodal_correct[-1,]
+# HEATbind_intermodal_correct_PPP = HEATbind_intermodal_correct
[…]

### 2023-10-02 paper/PaperTRA/intermodality_testbug.R

@@ -78,7 +78,11 @@ pois terá sido feito um sum em todos.
 Este erro será apenas referente ao intermodal, para a parte da bicicleta.
 Isto deve-se a ter deixado as rotas separadas - o que faz todo o sentido - e depois ter pegado nesse ficheiro para o heat.
 
-
+O código estava de facto bem escrito (heat-prep.R, linha 273-286)! mas ele está a dar o seguinte aviso, que indica que continuam a aparecer 2 linhas por segmento
+# Warning message:
+#   Returning more (or less) than 1 row per `summarise()` group was deprecated in dplyr 1.1.0.
+# ℹ Please use `reframe()` instead.
+# ℹ When switching from `summarise()` to `reframe()`, remember that `reframe()` always returns an ungrouped data frame and adjust accordingly.
 
 
 ### Por enquanto refazer a tabela 2. Depois tentar calcular de novo as emissões

### 2023-10-02 paper/PaperTRA/PaperTRA.Rmd
Update PaperTRA.Rmd

@@
 
-For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.0 [@HEAT] and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5] from the World Health Organization.
+For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.0 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
 The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years).

### 2023-10-03 paper/PaperTRA/PaperTRA.Rmd
correct summary 2 and 3

@@
 
-summary21 = summary21 %>%
+summary21 = summary2 %>%
   # mutate(LTS = recode(LTS, "seguro" = "safe", "direto" = "direct")) |> 
@@
 ```{r summary22}
-
-summary22 = summary2 %>%
-  mutate(LTS = recode(LTS, "seguro" = "safe", "direto" = "direct")) |> 
-  mutate(Estrategia = paste0(`Meta ENMAC`*100,"%")) |> 
-  mutate(Value = `Valorização monetária (€)`/1000) |> 
-  select(Estrategia, LTS, CO2eq, CO, PM10, NOx, VOC, Value) |> 
-  arrange(desc(Estrategia))
+summary22 = readRDS("load/TPtable_aml.Rds")
+summary22 = summary22 %>%
+  mutate(LTS = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
+  mutate(Value = money_emissions_eur/1000) |> 
+  select(Meta_ENMAC, LTS, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, Value) |> 
+  arrange(Meta_ENMAC)
 
@@
 
-summaryall = summary1 |> left_join(summary22) |> 
+summaryall = summary1 |> 
+  mutate(Meta_ENMAC = Estrategia) |>
+  left_join(summary22) |> 
   mutate(Code = c(3304, 3404, 3310, 3410)) |> 
   left_join(summary0 |> select(Code, Economic), by = "Code") |> 
-  select(c(1, 2, 6, 7, 8, 13, 15)) |> 
-  mutate(CO2eq = co2 + CO2eq,
+  select(Estrategia, LTS, co2, CO2eq_TP, monet, Value, Economic) |> 
+  mutate(CO2eq = co2 + CO2eq_TP,
          value1 = Economic/1000 + Value, # this is CORRECT

### 2023-10-03 paper/PaperTRA/PaperTRA.tex
correct summary 2 and 3

@@
 The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
-metropolitan data agregated and validated. This information is crucial
-for an accurate PT trip and route estimation. A digital elevation model
-from the European Space Agency's COPERNICUS mission, with a 25m spatial
-resolution, was used to include street gradient information, as a weight
-in cycling routing. The cycling potential trips for the two national
-strategic targets (4\% and 10\%) was estimated from the values for
-cycling and car trips (both as a driver and as a passenger) from the
-2017 base scenario.
+metropolitan data aggregated and validated. This information is crucial
+for an accurate PT trip and route estimation. A digital elevation model,
+from the European Space Agency's COPERNICUS mission, was used to include
+street gradient information, as a weight in cycling routing. The cycling
+potential trips for the two national strategic targets (4\% and 10\%)
+were estimated from the 2017 cycling and car trips (both as a driver and
+as a passenger), the baseline scenario.
 
@@
 socio-environmental impacts were estimated, using the HEAT for Cycling
-tool v5.0 {[}12{]} and the
+tool v5.0 {[}12{]} from the World Health Organization, and the
 \href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
@@
   information contact
-  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.} from
-the World Health Organization. The HEAT tool provided estimates on the
-shifting from car to cycling for a short term time horizon (i.e., one
-year) and the long term (i.e., ten years). We considered two dimensions:
-\emph{social} --- including the physical activity of cyclists, air
-pollution exposure, and road casualties; \emph{environmental} ---
-including CO\textsubscript{2}eq emissions and other pollutants.
+  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
+HEAT tool provided estimates on the shifting from car to cycling for a
+short term time horizon (i.e., one year) and the long term (i.e., ten
+years). We considered two dimensions: \emph{social} --- including the
+physical activity of cyclists, air pollution exposure, and road
+casualties; \emph{environmental} --- including CO\textsubscript{2}eq
+emissions and other pollutants.
 
@@
 car trips have the potential to be replaced by a combination of PT and
-cycling (up to 5 km on bike). \emph{Table \ref{tab:summary21} shows
-something that doesn't make sense.}
+cycling (up to 5 km on bike). Table \ref{tab:summary21} shows the PT
+modes with most potential of in combination with cycling, by mode.
 
@@
 \centering
-\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{4em}}
+\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
 \toprule
-Target & Routing & Total & Bus & Ferry & Train & Tram\\
+Target & Routing & Potential & Bus & Ferry & Train & Tram\\
 \midrule
-4\% & safe & 192 214 & 4 163 & 5 167 & 169 984 & 12 900\\
-4\% & direct & 189 846 & 4 879 & 5 795 & 171 534 & 7 639\\
-10\% & safe & 224 152 & 5 042 & 5 594 & 197 857 & 15 659\\
-10\% & direct & 219 511 & 5 806 & 6 263 & 198 373 & 9 069\\
+4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
+4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
+10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
+10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
 \bottomrule
@@
 of PT, with the potential to uptake the number of PT trips within the
-LMA area by as much as 12\% (in addition to the 825 thousand PT trips
-reported in the mobility survey). These findings indicate that
-transferring car trips to a combination of bicycle and PT could be
-nearly as substantial, if not equally significant, as the shift towards
-bicycle-only trips.
+LMA area by as much as 6.3\% (in addition to the 825 thousand PT trips
+reported in the mobility survey).
 
@@
 
+The social impacts represent 98\% of the socio-environmental benefits
+(in value) from replacing car trips to bicycle in first-and-last legs.
 The emissions of CO\textsubscript{2}eq that are avoided during both the
@@
 
-The social impacts from replacing car trips to bicycle in first-and-last
-legs, represent 95\% of the total socio-environmental benefits estimated
-for all journey legs. Our findings suggest that a bicycle and PT
-combination could viably replace 20\% of current LMA trips, with an
-additional 12\% of PT journeys prone to further substitution.
+Our findings suggest that a bicycle and PT combination could viably
+replace 20\% of current LMA trips, with an additional 12\% of PT
+journeys prone to further substitution.
 

### 2023-10-03 paper/PaperTRA/PaperTRA.Rmd
wrap paper to TRA
I think the paper is ok for the full paper TRA.
There is a lot of room for improvement, as marked in comments, when submitting to https://www.sciencedirect.com/journal/computers-environment-and-urban-systems/about/call-for-papers#environmentally-and-socially-sustainable-mobility in late november.
@Robinlovelace do you want to take a look at the output, before the submission next week?

Still need to HEAT to be in good heath to make sure that the SE estimates remain the same after the changes.

@@
 The impacts of shifting car trips to PT were also estimated and monetized.
-  The results indicate that 20% of car trips could switch to the bicycle + PT combination.
-Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of €230 million.
-For the PT leg, the transfer from car avoids of at least 8,500 tons of CO~2~eq emissions per year.
+  The results indicate that 10% of trips could be made by bicycle + PT combination.
+Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of at least €230 million.
+For the PT leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
   This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
@@
 <!-- TODO: -->
-<!-- understand where the values from table 2 come from -->
-<!-- make sure of the reference cls - try the other cls -->
 <!-- see where we can cut to 1 less paragraph -->
@@
 
-Table \ref{tab:summary1} presents the LMA total daily trips, the cycling combined with PT trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles.
+Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles.
 For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are also presented, resulting from replacing car trips with cycling.
@@
   mutate(LTS = recode(LTS, `3` = "safe", `4` = "direct")) |> 
-  mutate(Estrategia = paste0(Estrategia*100,"%"))
+  mutate(Estrategia = paste0(Estrategia*100,"%")) |> 
+  mutate(TotalTrips = TotalTrips/2,
+         Bikebase = Bikebase/2,
+         Bikepotential = Bikepotential/2) #this is a quick correction while we donr run biclar entirely again.
+The error was found and it was adding up the 2 bike legs, instead of making the mean.
+The SE estimates remain the same because they are from the bikeshare and carshare percentages, that did not change.
 
@@
 
-For both *direct* and *safe* route profiles, 20% of the daily car trips have the potential to be replaced by a combination of PT and cycling (up to 5 km on bike). <!-- NOT CORRECT -->
+For both *direct* and *safe* route profiles, 10% of the daily trips have the potential to be made by a combination of PT and cycling (up to 5 km on bike).
+This unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).
 <!-- place in the end - here is not clear -->
-Table \ref{tab:summary21} shows the PT modes with most potential of  in combination with cycling, by mode.
-<!-- ???? -->
+
+Table \ref{tab:summary21} shows the potential trips by PT mode to replace the second leg of the journey, in combination with cycling.
+Trains offer the greatest potential for substitution (88%).
+When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the Train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.
 
@@
 
-This scenario unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).
 <!-- These findings indicate that transferring car trips to a combination of bicycle and PT could be nearly as substantial, if not equally significant, as the shift towards bicycle-only trips.
 WE CANNOT SHOW THIS IN THIS PAPER WITH ONLY SCENARIO 3 DATA -->
-
-When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the PT train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.
 
@@
 -->
-<!-- this co2 needs to be added to the second table and then discuss -->
-Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly, with trains offering the greatest potential for substitution (88%).
+<!-- this table could be split into Social benefits and Environmental benefits -->
+Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly.
 
@@
 The social impacts represent 98% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs.
-<!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper -->
+For the PT segment, we did not estimate the social impacts from substituting car trips, although its health benefits would not be as high as shifting to cycling.<!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper --> <!--Overall there was very little evidence available on mode switch to public transport.
+-->
+
 The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 75% of the emissions avoided during the PT segment.
@@
 
-The information available at *biclaR* tool -- an open access website -- can be downloaded and used with any GIS software to, for instance, understand which potential cycling connections have the highest socio-environmental impacts, in tons of avoided CO~2~eq emissions, or in long term social benefits.
+The information available at *biclaR* tool -- an open access website -- can be downloaded and used with any GIS software.
+This allows users to, for example, gain insights into which potential cycling connections have the highest socio-environmental impacts, quantified in tons of avoided CO~2~eq emissions, or in long term social benefits.
 <!-- move to introduction?

### 2023-10-03 paper/PaperTRA/PaperTRA.tex
wrap paper to TRA
I think the paper is ok for the full paper TRA.
There is a lot of room for improvement, as marked in comments, when submitting to https://www.sciencedirect.com/journal/computers-environment-and-urban-systems/about/call-for-papers#environmentally-and-socially-sustainable-mobility in late november.
@Robinlovelace do you want to take a look at the output, before the submission next week?

Still need to HEAT to be in good heath to make sure that the SE estimates remain the same after the changes.

@@
 car trips to PT were also estimated and monetized. The results indicate
-that 20\% of car trips could switch to the bicycle + PT combination.
-Shifting to cycling for the first-and-last mile stages can reduce annual
+that 10\% of trips could be made by bicycle + PT combination. Shifting
+to cycling for the first-and-last mile stages can reduce annual
 CO\textsubscript{2}eq emissions from 6,000 tons/day, with benefits over
-10 years of €230 million. For the PT leg, the transfer from car avoids
-of at least 8,500 tons of CO\textsubscript{2}eq emissions per year. This
-evidence can support policymakers to prioritize interventions that
+10 years of at least €230 million. For the PT leg, the transfer from car
+avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per year.
+This evidence can support policymakers to prioritize interventions that
 reduce the reliance on private motor vehicles.
@@
 
-Table \ref{tab:summary1} presents the LMA total daily trips, the cycling
-combined with PT trips in the baseline scenario and corresponding new
+Table \ref{tab:summary1} presents the LMA total daily trips that can be
+made with cycling + TP combination (with the aforementioned
+restrictions), the trips in the baseline scenario and corresponding new
 trips to achieve the national strategy targets (4\% and 10\%), for
@@
 \midrule
-4\% & safe & 1 077 028 & 4 624 & 40 770 & 5 920 & 230 270\\
-4\% & direct & 1 001 761 & 4 547 & 37 889 & 6 011 & 223 720\\
-10\% & safe & 1 077 028 & 4 624 & 104 647 & 15 192 & 591 790\\
-10\% & direct & 1 001 761 & 4 547 & 97 218 & 15 414 & 574 200\\
+4\% & safe & 538 514 & 2 312 & 20 385 & 5 920 & 230 270\\
+4\% & direct & 500 880 & 2 274 & 18 944 & 6 011 & 223 720\\
+10\% & safe & 538 514 & 2 312 & 52 324 & 15 192 & 591 790\\
+10\% & direct & 500 880 & 2 274 & 48 609 & 15 414 & 574 200\\
 \bottomrule
@@
 
-For both \emph{direct} and \emph{safe} route profiles, 20\% of the daily
-car trips have the potential to be replaced by a combination of PT and
-cycling (up to 5 km on bike). Table \ref{tab:summary21} shows the PT
-modes with most potential of in combination with cycling, by mode.
+For both \emph{direct} and \emph{safe} route profiles, 10\% of the daily
+trips have the potential to be made by a combination of PT and cycling
+(up to 5 km on bike). This unveils the potential of cycling as a
+complementary mode of PT, with the potential to uptake the number of PT
+trips within the LMA area by as much as 6.3\% (in addition to the 825
+thousand PT trips reported in the mobility survey).
+
+Table \ref{tab:summary21} shows the potential trips by PT mode to
+replace the second leg of the journey, in combination with cycling.
+Trains offer the greatest potential for substitution (88\%). When
+comparing the existing PT interfaces (Figure \ref{fig:map1}) with the
+bike routes with highest potential to serve as first and last legs
+(Figure \ref{fig:map2}) it becomes clear that the Train interfaces are
+the ones that have the highest potential to attract car-to-PT
+substituting trips, if their accessibility by bicycle is improved to be
+safe.
 
@@
 
-This scenario unveils the potential of cycling as a complementary mode
-of PT, with the potential to uptake the number of PT trips within the
-LMA area by as much as 6.3\% (in addition to the 825 thousand PT trips
-reported in the mobility survey).
-
-When comparing the existing PT interfaces (Figure \ref{fig:map1}) with
-the bike routes with highest potential to serve as first and last legs
-(Figure \ref{fig:map2}) it becomes clear that the PT train interfaces
-are the ones that have the highest potential to attract car-to-PT
-substituting trips, if their accessibility by bicycle is improved to be
-safe.
-
 Table \ref{tab:summary22} presents the results of the avoided emissions
@@
 equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4
-million to €3.5 million yearly, with trains offering the greatest
-potential for substitution (88\%).
+million to €3.5 million yearly.
 
@@
 (in value) from replacing car trips to bicycle in first-and-last legs.
+For the PT segment, we did not estimate the social impacts from
+substituting car trips, although its health benefits would not be as
+high as shifting to cycling.
+
 The emissions of CO\textsubscript{2}eq that are avoided during both the
@@
 The information available at \emph{biclaR} tool -- an open access
-website -- can be downloaded and used with any GIS software to, for
-instance, understand which potential cycling connections have the
-highest socio-environmental impacts, in tons of avoided
-CO\textsubscript{2}eq emissions, or in long term social benefits.
+website -- can be downloaded and used with any GIS software. This allows
+users to, for example, gain insights into which potential cycling
+connections have the highest socio-environmental impacts, quantified in
+tons of avoided CO\textsubscript{2}eq emissions, or in long term social
+benefits.
 
@@
 \href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
-Económico e Social da Sinistralidade Rodoviária em Portugal}}. {CEGE -
-Centro de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
+Económico e Social da Sinistralidade Rodoviária em Portugal}}. {Centro
+de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
 Rodoviária (ANSR)} (2021).}

### 2023-10-10 paper/PaperTRA/PaperTRA.Rmd
update to-do
minor updates
there are some things that we will improve in the other paper

@@
 <!-- TODO: -->
-<!-- see where we can cut to 1 less paragraph -->
+<!-- check if HEAT estimates remain the same (the tool is not working atm) -->
 <!-- should we ignore the social benefits component? and only estimate and add the environmental ones? or estimate the car to tp transfer Social benefits? maybe to the next paper? -->
@@
 
-Emissions were estimated for the following atmospheric pollutants: CO~2~eq, CO, PM10, NOx, and VOC.
+Emissions were estimated for the following atmospheric pollutants: CO, PM10, NOx, and VOC; and for the main green house gases: CO~2, CH~4, and N~2~O, converted in CO~2~eq.
 In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally.
@@
                     "Avoided CO2eq (ton/yr)", "SE Benefits for 10 years (thousand €)"),
-      caption = "\\label{summary1}Summary of the cycling potencial of intermodality scenario.",
+      caption = "\\label{summary1}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.",
       format="latex",
@@
 
-<!-- separate the Socio impacts from the Environemtal ones? It shouldn't be ok to sum them -->
+<!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? It shouldn't be ok to sum them -->
 Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year, 

### 2023-10-10 HEAT/code/haas_reprex.R
try HaaS again - no luck

@@ -48,6 +48,8 @@ resp <-
 
 # convert from "jsonl-serialized" to R format
 from_serialized_json <- jsonlite::unserializeJSON(httr::content(resp, "text"))
+
+
 str(from_serialized_json$results)
 
 # Check final result value
@@ -76,6 +78,8 @@ resp <-
 
 # convert to R
 from_rds <- unserialize(httr::content(resp, "raw"))
+# check the response
+str(resp)
 
 # Check final result value
 res = as.numeric(gsub(" ","",from_rds$results$moneytotaldisc)) #remove spaces and convert to to numeric
@@ -84,6 +88,30 @@ res[26]-res[27]-res[28]+res[29]
 #> 
 #> # ESTE É o unico que dá resultados semelhantes com o do UI, pois usa o genericdata (onde tem o occupancy car rate, velocidades assumidas, e % de shifting de que modos
 # 
+
+
+ # TRY TO USE OTHER VALUES
+webapp_input[["raw_activemode_bike_ref_alltrips"]] # it use to be the sum of all TRIPS per day (not per person/day)
+
+totaltripspax = 2.6
+webapp_input[["raw_activemode_bike_ref_alltrips"]] = totaltripspax
+webapp_input[["raw_activemode_bike_cf_alltrips"]] = totaltripspax
+webapp_input[["raw_motorizedmode_car_ref_alltrips"]] = totaltripspax
+webapp_input[["raw_motorizedmode_car_cf_alltrips"]] = totaltripspax
+
+
+# DOESNT WORK
+resp_reprex <- 
+  httr::POST("https://api.heatwalkingcycling.org/apiv1/heat_5_0/calc_results?output_format=rds",
+             httr::set_cookies("user" = Sys.getenv("HEAT_COOKIE")),
+             body = jsonlite::toJSON(
+               list("webapp_input" = webapp_input,
+                    "generic_data_relevant" = genericdata)),
+             encode = "multipart")
+
+# convert to R
+from_rds_reprex <- unserialize(httr::content(resp_reprex, "raw"))
+
 
 
 

### 2023-10-10 code/test-code/heat-prep.R

@@ -253,7 +253,7 @@ HEATbind_aml_PPP = HEATbind
 ## For Intermodality ---------------------------------------------------------------------------
 # RUN THIS AGAIN WHEN API GETS LIVE
 
-Code_Hass = CENARIOS$Code[9:12] #fazer por enquanto só AML e senários 1 e 2
+Code_Hass = CENARIOS$Code[9:12] #fazer por enquanto só AML e senários 3
 
 HEATbind_intermodal_correct = HEAT
 
@@ -275,11 +275,11 @@ for (i in Code_Hass){
     ungroup() %>% 
     group_by(id) %>%  # it shouldn't return two segments again.
     summarise(distance = sum(distance),  # we need to put both segments distance together!
-              Bike = Bike,
-              Car = Car + CarP,
-              Total = Total,
-              Bikeper = Bike / Total) %>% 
-    unique() |> # NOT THE IDEAL BUT WORKS FOR THE PURPUSE (this was missing in the first version 2023)
+              Bike = mean(Bike),
+              Car = mean(Car) + mean(CarP),
+              Total = mean(Total),
+              Bikeper = mean(Bike) / mean(Total)) %>% 
+    # unique() |> # NOT THE IDEAL BUT WORKS FOR THE PURPUSE (this was missing in the first version 2023)
     ungroup() %>% 
     mutate(
       cyc = ifelse(Bikeper >= HEAT$ENMAC, Bike, HEAT$ENMAC * Total),
@@ -292,10 +292,10 @@ for (i in Code_Hass){
     filter(id %in% unique(HEAT_bike$id)) %>% 
     group_by(id) %>% 
     summarise(distance = sum(distance), 
-              Bike = Bike,
-              Car = Car + CarP,
-              Total = Total,
-              Bikeper = Bike / Total) %>% 
+              Bike = mean(Bike),
+              Car = mean(Car) + mean(CarP),
+              Total = mean(Total),
+              Bikeper = mean(Bike) / mean(Total)) %>% 
     mutate(
       cyc = ifelse(Bikeper >= HEAT$ENMAC, Bike, HEAT$ENMAC * Total),
       new_cyc = ifelse(Bikeper >= HEAT$ENMAC, 0, cyc - Bike)) %>% 

### 2023-10-11 paper/PaperTRA/PaperTRA.Rmd
Add minor tweak

@@
 
-Our findings suggest that a bicycle and PT combination could viably replace 20% of current LMA trips, with an additional 12% of PT journeys prone to further substitution. 
+Our findings suggest that cycling and PT *in combination* could viably replace 20% of current LMA trips, with an additional 12% of PT journeys prone to further substitution. 
 <!-- repeated? not correct -->

### 2023-10-11 paper/PaperTRA/PaperTRA.Rmd
minor changes and knit

--- 2023-10-10 paper/PaperTRA/PaperTRA.Rmd
@@
 Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
-In addition, we assumed an occupancy rate of 1.61 passengers *per* car [@IMOB].
+In addition, we assumed an occupancy rate of 1.6 passengers *per* car [@IMOB].
 
-Regarding the PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014].
+Regarding PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014].
 
-Emissions were estimated for the following atmospheric pollutants: CO, PM10, NOx, and VOC; and for the main green house gases: CO~2, CH~4, and N~2~O, converted in CO~2~eq.
+Emissions were estimated for the following atmospheric pollutants: CO, PM10, NOx, and VOC; and for the main green house gases: CO~2~, CH~4~, and N~2~O, converted in CO~2~eq.
 In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally.
@@
 
-Our findings suggest that a bicycle and PT combination could viably replace 20% of current LMA trips, with an additional 12% of PT journeys prone to further substitution. 
+Our findings suggest that cycling and PT *in combination* could viably replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution. 
 <!-- repeated? not correct -->

### 2023-10-11 paper/PaperTRA/PaperTRA.tex
minor changes and knit

@@
 factors for different ranges of trips at 500-meter intervals. In
-addition, we assumed an occupancy rate of 1.61 passengers \emph{per} car
+addition, we assumed an occupancy rate of 1.6 passengers \emph{per} car
 {[}3{]}.
 
-Regarding the PT, we considered the emission factor values reported in
-the environmental and sustainability reports of the PT operators in the
-LMA {[}14--17{]}.
+Regarding PT, we considered the emission factor values reported in the
+environmental and sustainability reports of the PT operators in the LMA
+{[}14--17{]}.
 
-Emissions were estimated for the following atmospheric pollutants:
-CO\textsubscript{2}eq, CO, PM10, NOx, and VOC. In particular, for the
-urban train and tram -- with 100\% electric traction -- only
-CO\textsubscript{2}eq emissions were considered (resulting from the
-production of electricity, considering a ``well-to-tank'' approach),
-since the other pollutants are not emitted locally.
+Emissions were estimated for the following atmospheric pollutants: CO,
+PM10, NOx, and VOC; and for the main green house gases:
+CO\textsubscript{2}, CH\textsubscript{4}, and N\textsubscript{2}O,
+converted in CO\textsubscript{2}eq. In particular, for the urban train
+and tram -- with 100\% electric traction -- only CO\textsubscript{2}eq
+emissions were considered (resulting from the production of electricity,
+considering a ``well-to-tank'' approach), since the other pollutants are
+not emitted locally.
 
@@
 
-\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of intermodality scenario.}
+\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.}
 \centering
@@
 
-Our findings suggest that a bicycle and PT combination could viably
-replace 20\% of current LMA trips, with an additional 12\% of PT
+Our findings suggest that cycling and PT \emph{in combination} could
+viably replace 10\% of current LMA trips, with an additional 6\% of PT
 journeys prone to further substitution.

### 2023-11-22 paper/PaperCEUS/PaperCEUS.Rmd
paper CEUS setup

--- 2023-10-11 paper/PaperTRA/PaperTRA.Rmd
@@
 ---
-classoption: "runningheads"
 title: "biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
-shorttitle: "biclaR: SE impacts of car substitution by bicycle and public transit"
-
-authors: 
-  - name: "Rosa Félix"
-    inst: "1"
-    orcid: "0000-0002-5642-6006"
-  - name: "Filipe Moura"
-    inst: "1"
-    orcid: "0000-0001-7749-8490"
-  - name: "Robin Lovelace"
-    inst: "2"
-    orcid: "0000-0001-5679-6536"
-# First names are abbreviated in the running head.
-# If there are more than two authors, 'et al.' is used.
-authorrunning: "R. Félix et al." 
-institutes: 
-  - name: "CERIS - Instituto Superior Técnico, University of Lisbon. Av Rovisco Pais 1049-001 Lisboa, Portugal"
-    email: "rosamfelix@tecnico.ulisboa.pt"
-  - name: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
-
+author:
+  - name: Rosa Félix\corref{cor1}
+    affiliation: CERIS
+    correspondingauthor: true
+    # footnote: 1
+    email: rosamfelix@tecnico.ulisboa.pt
+  - name: Filipe Moura
+    affiliation: CERIS
+    email: fmoura@tecnico.ulisboa.pt
+  - name: Robin Lovelace
+    affiliation: ITS
+    email: r.lovelace@uleeds.uk
+address:
+  - code: CERIS
+    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
+  - code: ITS
+    address: "Institute for Transport Studies, University of Leeds. 34-40 University Rd, Leeds LS2 9JT, UK"
 abstract: |
@@
   This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
+  \
 keywords: 
@@
   - Open data and methods
+journal: "Computers, Environment and Urban Studies"
+date: "`r Sys.Date()`"
+classoption: review, doubleblind, 3p, authoryear #change later to preprint instead of review
 bibliography: bibliography.bib
-# Use csl when using default citation_package for Pandoc citeproc
-csl: https://www.zotero.org/styles/springer-lecture-notes-in-computer-science
-# cls : svproc.csl
-# citations with arabic numbers is prefered when using natbib
-natbiboptions: numbers
+biblio-style: elsarticle-harv
+linenumbers: true
+numbersections: true
+# Use a CSL with `citation_package = "default"`
+csl: https://www.zotero.org/styles/computers-environment-and-urban-systems
+# csl: computers-environment-and-urban-systems.csl
+# csl: trd_test.csl
+preamble: |
+  \biboptions{authoryear}
+  \usepackage{url}
+  \usepackage{float}
+  \usepackage{booktabs}
+  \usepackage{longtable}
+  \usepackage{makecell}
+  \usepackage{multirow}
+output: 
+  rticles::elsevier_article:
+    keep_tex: yes
+    citation_package: "default" #ou nenhum
+    latex_engine: xelatex
+    extra_dependencies: "subfig"
+#   bookdown::word_document2:
+#     fig_caption: true
+# always_allow_html: true
+urlcolor: blue
+citecolor: blue
+linkcolor: blue
+# natbiboptions: longnamesfirst,angle,semicolon
+---
 
-header-includes:
-  # - \usepackage[hidelinks]{hyperref} #this is to remove the annoying boxes. but how to make them blue?
-  - \usepackage{booktabs}
-  - \usepackage{siunitx}
-  - \usepackage{hyperref}
-  - \hypersetup{colorlinks = TRUE,  urlcolor = blue, linkcolor = blue, citecolor = blue}
-   #Colours links instead of ugly boxes
-  #Colour for external hyperlinks
-  #Colour of internal links
-  #Colour of citations
+Please make sure that your manuscript follows the guidelines in the 
+Guide for Authors of the relevant journal. It is not necessary to 
+typeset your manuscript in exactly the same way as an article, 
+unless you are submitting to a camera-ready copy (CRC) journal.
 
-output:
-  rticles::lncs_article:
-    citation_package: default # or natbib DEFINE LATER
-# output: word_document
-
-urlcolor: blue
-linkcolor: blue
----
+For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>
 

### 2023-11-22 paper/PaperCEUS/PaperCEUS.tex
paper CEUS setup

--- 2023-10-11 paper/PaperTRA/PaperTRA.tex
@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.21 of 2022/01/12
-%
-\documentclass[runningheads]{llncs}
-%
+\documentclass[review, doubleblind, 3p,
+authoryear]{elsarticle} %review=doublespace preprint=single 5p=2 column
+%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
+
+\usepackage[hyphens]{url}
+
+  \journal{Computers, Environment and Urban Studies} % Sets Journal name
+
+\usepackage{graphicx}
+%%%%%%%%%%%%%%%% end my additions to header
+
 \usepackage[T1]{fontenc}
-% T1 fonts will be used to generate the final print and online PDFs,
-% so please use T1 fonts in your manuscript whenever possible.
-% Other font encondings may result in incorrect characters.
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+% TODO: Currently lineno needs to be loaded after amsmath because of conflict
+% https://github.com/latex-lineno/lineno/issues/5
+\usepackage{lineno} % add
+  \linenumbers % turns line numbering on
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \usepackage{fontspec}
+  \ifxetex
+    \usepackage{xltxtra,xunicode}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use microtype if available
+\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={},
+            pdftitle={biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools},
+            colorlinks=true,
+            urlcolor=blue,
+            linkcolor=blue,
+            pdfborder={0 0 0}}
+
+\setcounter{secnumdepth}{5}
+% Pandoc toggle for numbering sections (defaults to be off)
 
@@
 
-\usepackage{booktabs}
-\usepackage{siunitx}
-\usepackage{hyperref}
-\hypersetup{colorlinks = TRUE,  urlcolor = blue, linkcolor = blue, citecolor = blue}
-
+\usepackage{subfig}
 \usepackage{booktabs}
@@
 
-\usepackage{graphicx}
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following two lines
-% to display URLs in blue roman font according to Springer's eBook style:
-\usepackage{hyperref}
-\usepackage{color}
-\renewcommand\UrlFont{\color{blue}\rmfamily}
+\biboptions{authoryear}
+\usepackage{url}
+\usepackage{float}
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{makecell}
+\usepackage{multirow}
 
@@
 
+\begin{frontmatter}
+
   \title{biclaR: Estimating the socio-environmental impacts of car
 substitution by bicycle and public transit using open tools}
+    \author[CERIS]{Rosa Félix\corref{cor1}%
+  \corref{cor1}%
+  }
+   \ead{rosamfelix@tecnico.ulisboa.pt} 
+    \author[CERIS]{Filipe Moura%
   %
-\titlerunning{biclaR: SE impacts of car substitution by bicycle and
-public transit}
-% If the paper title is too long for the running head, you can set
-% an abbreviated paper title here
+  }
+   \ead{fmoura@tecnico.ulisboa.pt} 
+    \author[ITS]{Robin Lovelace%
   %
-\author{Rosa Félix\inst{1}\orcidID{0000-0002-5642-6006} \and Filipe
-Moura\inst{1}\orcidID{0000-0001-7749-8490} \and Robin
-Lovelace\inst{2}\orcidID{0000-0001-5679-6536}}
+  }
+   \ead{r.lovelace@uleeds.uk} 
+      \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
+Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
+    \affiliation[ITS]{Institute for Transport Studies, University of
+Leeds. 34-40 University Rd, Leeds LS2 9JT, UK}
+    \cortext[cor1]{Corresponding author}
 
-\authorrunning{R. Félix et al.}
-% First names are abbreviated in the running head.
-% If there are more than two authors, 'et al.' is used.
-%
-
-\institute{CERIS - Instituto Superior Técnico, University of Lisbon. Av
-Rovisco Pais 1049-001 Lisboa, Portugal\\
-\email{\href{mailto:rosamfelix@tecnico.ulisboa.pt}{\nolinkurl{rosamfelix@tecnico.ulisboa.pt}}}\\ \and Institute
-for Transport Studies, University of Leeds. 34-40 University Rd, Leeds
-LS2 9JT, UK}
-
-\maketitle              % typeset the header of the contribution
-%
   \begin{abstract}
@@
   public transit and cycling for the first-and-last mile. This paper
-estimates the potential for cycling combined with public transit (PT) as
-a substitute for car trips in the Lisbon metropolitan area and assesses
-its socio-environmental impacts using open data and open source tools. A
-decision support tool that facilitates the design and development of a
-metropolitan cycling network was developed (\emph{biclaR}). The social
-and environmental impacts were assessed using the \emph{HEAT for
-Cycling} and the \emph{HEAT as a Service} tools. The impacts of shifting
-car trips to PT were also estimated and monetized. The results indicate
-that 10\% of trips could be made by bicycle + PT combination. Shifting
-to cycling for the first-and-last mile stages can reduce annual
-CO\textsubscript{2}eq emissions from 6,000 tons/day, with benefits over
-10 years of at least €230 million. For the PT leg, the transfer from car
-avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per year.
-This evidence can support policymakers to prioritize interventions that
-reduce the reliance on private motor vehicles.
+  estimates the potential for cycling combined with public transit (PT)
+  as a substitute for car trips in the Lisbon metropolitan area and
+  assesses its socio-environmental impacts using open data and open
+  source tools. A decision support tool that facilitates the design and
+  development of a metropolitan cycling network was developed
+  (\emph{biclaR}). The social and environmental impacts were assessed
+  using the \emph{HEAT for Cycling} and the \emph{HEAT as a Service}
+  tools. The impacts of shifting car trips to PT were also estimated and
+  monetized. The results indicate that 10\% of trips could be made by
+  bicycle + PT combination. Shifting to cycling for the first-and-last
+  mile stages can reduce annual CO\textsubscript{2}eq emissions from
+  6,000 tons/day, with benefits over 10 years of at least €230 million.
+  For the PT leg, the transfer from car avoids of up to 20,500 tons of
+  CO\textsubscript{2}eq emissions per year. This evidence can support
+  policymakers to prioritize interventions that reduce the reliance on
+  private motor vehicles.\\
+  \end{abstract}
+    \begin{keyword}
+    Active transport \sep Intermodality \sep First and last
+mile \sep Health economic assessment \sep Environmental impacts \sep 
+    Open data and methods
+  \end{keyword}
 
-\keywords{Active transport \and Intermodality \and First and last
-mile \and Health economic assessment \and Environmental
-impacts \and Open data and methods}
+ \end{frontmatter}
 
-\end{abstract}
+Please make sure that your manuscript follows the guidelines in the
+Guide for Authors of the relevant journal. It is not necessary to
+typeset your manuscript in exactly the same way as an article, unless
+you are submitting to a camera-ready copy (CRC) journal.
+
+For detailed instructions regarding the elsevier article class, see
+\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}
 
@@
 mile in metropolitan areas can significantly replace private car trips
-{[}1, 2{]}. This approach requires interventions and programs to make
-bicycling more appealing, and the resulting public investments can have
-significant social and environmental benefits.
+(Martens, 2007; Rietveld, 2000). This approach requires interventions
+and programs to make bicycling more appealing, and the resulting public
+investments can have significant social and environmental benefits.
 
-According to the latest mobility survey conducted in 2018 {[}3{]}, the
-LMA registered a total of 5.3 million daily trips, with only 0.5\% by
-bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%. The
-number of intra-municipal trips --- with origin and destination in the
-same municipality --- amounts to 3.5 million trips. This exceeds the
+According to the latest mobility survey conducted in 2018 (INE, 2018),
+the LMA registered a total of 5.3 million daily trips, with only 0.5\%
+by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%.
+The number of intra-municipal trips --- with origin and destination in
+the same municipality --- amounts to 3.5 million trips. This exceeds the
 number of inter-municipal trips (1.8 million trips), involving travel
@@
 To achieve the cycling targets set by the Portuguese national cycling
-strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}4{]}, the
-Lisbon's Metropolitan Department of Transport introduced
-\emph{biclaR}\footnote{See
+strategy for 2025 and 2030 (4\% and 10\%, respectively) (Presidência do
+Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
+Transport introduced \emph{biclaR}\footnote{See
   \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
 decision support tool that facilitates the design and development of a
-metropolitan cycling network {[}5{]}.
+metropolitan cycling network (Félix et al., 2022).
 
@@
 origin-destination data for commuting trips as the basis of estimates of
-cycling potential at zone, route and route network levels {[}6{]}. The
-PCT has been extended to include cycling potential for travel to school
-in England {[}7{]} and other trip types in other countries.\footnote{See
-  \href{https://www.npt.scot}{npt.scot} and
-  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
+cycling potential at zone, route and route network levels (Lovelace et
+al., 2017). The PCT has been extended to include cycling potential for
+travel to school in England (Goodman et al., 2019) and other trip types
+in other countries.\footnote{See \href{https://www.npt.scot}{npt.scot}
+  and \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
   Scotland and Ireland that include estimates of cycling for other
@@
 
-The mobility survey data {[}3{]} is the basis for this project and
+The mobility survey data (INE, 2018) is the basis for this project and
 defines the baseline scenario. Despite being conducted in the
@@
 significant portion of short-distance trips, a prevalent characteristic
-of active mode travel {[}8{]}. The OD Jittering method breaks down a
-single point (i.e., the centroid of an area) into multiple random points
-on the existing and neighboring road network, using OpenStreetMap as a
-reference. This method then distributes the volume of trips within the
-district among the randomly generated origin-destination pairs.
+of active mode travel (Lovelace et al., 2022b). The OD Jittering method
+breaks down a single point (i.e., the centroid of an area) into multiple
+random points on the existing and neighboring road network, using
+OpenStreetMap as a reference. This method then distributes the volume of
+trips within the district among the randomly generated
+origin-destination pairs.
 
@@
 The selected route choice algorithm was the
-\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}9{]},
-which allows for great flexibility in configuring estimated route types,
-and which proven to provide most accurate route networks for the city of
-Lisbon {[}10{]}. \texttt{r5r} can calculate multi-modal routes using PT
-combined with other modes. It enables the identification of the most
-direct or safest cycling routes, using the Level of Traffic
-Stress\footnote{see
+\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} (Pereira
+et al., 2021), which allows for great flexibility in configuring
+estimated route types, and which proven to provide most accurate route
+networks for the city of Lisbon (Lovelace et al., 2022a). \texttt{r5r}
+can calculate multi-modal routes using PT combined with other modes. It
+enables the identification of the most direct or safest cycling routes,
+using the Level of Traffic Stress\footnote{see
   \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
@@
 decision-makers to plan for different bicycle user typologies and/or for
-different city cycling maturity levels {[}11{]}.
+different city cycling maturity levels (Félix et al., 2017).
 
@@
 socio-environmental impacts were estimated, using the HEAT for Cycling
-tool v5.0 {[}12{]} from the World Health Organization, and the
+tool v5.0 (Kahlmeier et al., 2017) from the World Health Organization,
+and the
 \href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
@@
 To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
-methods and reference values {[}13{]} for a Tier 3 detail level. We used
-a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
-trips were considered to be made under urban conditions and at an
-average speed of 15 km/h during rush hour periods. Since the average
-distance traveled per trip influences the overconsumption and emissions
-from cold-start engine operation, we estimated energy and emission
-factors for different ranges of trips at 500-meter intervals. In
-addition, we assumed an occupancy rate of 1.6 passengers \emph{per} car
-{[}3{]}.
+methods and reference values (Ntziachristos \& Samaras, 2020) for a Tier
+3 detail level. We used a family-size vehicle, EURO standard, and
+gasoline or diesel fuel. All trips were considered to be made under
+urban conditions and at an average speed of 15 km/h during rush hour
+periods. Since the average distance traveled per trip influences the
+overconsumption and emissions from cold-start engine operation, we
+estimated energy and emission factors for different ranges of trips at
+500-meter intervals. In addition, we assumed an occupancy rate of 1.6
+passengers \emph{per} car (INE, 2018).
 
@@
 environmental and sustainability reports of the PT operators in the LMA
-{[}14--17{]}.
+(Carris, 2020; CP, 2020; Metropolitano de Lisboa, 2020; Transtejo,
+2014).
 
@@
 respective monetary valuation was based on the EU Guide to Cost-benefit
-Analysis {[}18{]} and the best up-to-date reference values for the
-various gases {[}18--20{]}. The social impacts are in avoided premature
-mortality. This result is finally monetized using the \emph{Statistical
-Value of Life} for Portugal {[}21{]}. We updated all the monetary
+Analysis (Sartori et al., 2014) and the best up-to-date reference values
+for the various gases (Bickel et al., 2006; Nash et al., 2003; Sartori
+et al., 2014). The social impacts are in avoided premature mortality.
+This result is finally monetized using the \emph{Statistical Value of
+Life} for Portugal (Silva et al., 2021). We updated all the monetary
 reference values of the literature based on the annual inflation rate in
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
+Bickel, P., Friedrich, R., Burgess, A., Fagiani, P., Hunt, A., Jong, G.
+D., \& Laird, J. (2006). \emph{{HEATCO - Developing Harmonised European
+Approaches for Transport Costing and Project Assessment. Deliverable 5,
+Proposal for Harmonised Guidelines.}}
+\url{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
+Carris. (2020). \emph{Relatório de sustentabilidade 2019 - demostração
+não financeira}. {Carris - Companhia Carris de Ferro de Lisboa, E.M.,
+S.A.} \url{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}
+
+\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
+CP. (2020). \emph{Relatório de sustentabilidade 2019}. {CP - Comboios de
+Portugal, E.P.E.}
+\url{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}
+
+\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
+Félix, R., Lovelace, R., \& Moura, F. (2022). \emph{{biclaR - Ferramenta
+de apoio ao planeamento da rede ciclável na área metropolitana de
+Lisboa}}. {CERIS - Instituto Superior Técnico and Transportes
+Metropolitanos de Lisboa}. \url{https://biclar.tmlmobilidade.pt}
+
+\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
+Félix, R., Moura, F., \& Clifton, K. J. (2017). Typologies of urban
+cyclists: Review of market segmentation methods for planning practice.
+\emph{Transportation Research Record}, \emph{2662}(1), 125--133.
+\url{https://doi.org/10.3141/2662-14}
+
+\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
+Goodman, A., Rojas, I. F., Woodcock, J., Aldred, R., Berkoff, N.,
+Morgan, M., Abbas, A., \& Lovelace, R. (2019). Scenarios of cycling to
+school in england, and associated health and carbon impacts: Application
+of the {`}propensity to cycle tool{'}. \emph{Journal of Transport and
+Health}, \emph{12}, 263--278.
+\url{https://doi.org/10.1016/j.jth.2019.01.008}
+
+\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
+INE. (2018). \emph{Mobilidade e funcionalidade do território nas {Áreas
+Metropolitanas do Porto e de Lisboa}: 2017}. {Instituto National de
+Estatística}.
+\url{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}
+
+\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
+Kahlmeier, S., Götschi, T., Cavill, N., Castro Fernandez, A., Brand, C.,
+Rojas Rueda, D., Woodcock, J., Kelly, P., Lieb, C., Oja, P., et al.
+(2017). \emph{Health economic assessment tool ({HEAT}) for walking and
+for cycling: Methods and user guide on physical activity, air pollution,
+injuries and carbon impact assessments}.
+\url{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
+Lovelace, R., Félix, R., \& Carlino, D. (2022a). Exploring jittering and
+routing options for converting origin-destination data into route
+networks: Towards accurate estimates of movement at the street level.
+\emph{The International Archives of the Photogrammetry, Remote Sensing
+and Spatial Information Sciences}, \emph{XLVIII-4/W1-2022}, 279--286.
+\url{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
+Lovelace, R., Félix, R., \& Carlino, D. (2022b). Jittering: A
+computationally efficient method for generating realistic route networks
+from origin-destination data. \emph{Findings}.
+\url{https://doi.org/10.32866/001c.33873}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
+Lovelace, R., Goodman, A., Aldred, R., Berkoff, N., Abbas, A., \&
+Woodcock, J. (2017). The Propensity to Cycle Tool: An open source online
+system for sustainable transport planning. \emph{Journal of Transport
+and Land Use}, \emph{10}(1). \url{https://doi.org/10.5198/jtlu.2016.862}
+
 \leavevmode\vadjust pre{\hypertarget{ref-MARTENS2007326}{}}%
-\CSLLeftMargin{1. }%
-\CSLRightInline{Martens, K.: Promoting bike-and-ride: The dutch
-experience. Transportation Research Part A: Policy and Practice. 41,
-326--338 (2007).
-https://doi.org/\href{https://doi.org/10.1016/j.tra.2006.09.010}{10.1016/j.tra.2006.09.010}.}
+Martens, K. (2007). Promoting bike-and-ride: The dutch experience.
+\emph{Transportation Research Part A: Policy and Practice},
+\emph{41}(4), 326--338. \url{https://doi.org/10.1016/j.tra.2006.09.010}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
+Metropolitano de Lisboa. (2020). \emph{Relatório integrado 2019}.
+{Metropolitano de Lisboa, E.P.E.}
+\url{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}
+
+\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
+Nash, C. et al. (2003). \emph{UNIfication of accounts and marginal costs
+for transport efficiency: Final report}. {Institute for Transport
+Studies, University of Leeds}.
+\url{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}
+
+\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
+Ntziachristos, L., \& Samaras, Z. (2020). \emph{{EMEP/EEA} air pollutant
+emission inventory guidebook 2019}. Luxembourg: European Environment
+Agency. \url{https://www.emisia.com/utilities/copert/documentation/}
+
+\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
+Pereira, R. H. M., Saraiva, M., Herszenhut, D., Braga, C. K. V., \&
+Conway, M. W. (2021). r5r: Rapid realistic routing on multimodal
+transport networks with R5 in r. \emph{Findings}.
+\url{https://doi.org/10.32866/001c.21262}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
+Presidência do Conselho de Ministros. (2019). Resolução do conselho de
+ministros n.º 131/2019. In \emph{Diário da República, 1ª série} (Vol.
+147, pp. 46--81).
+\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}
 
 \leavevmode\vadjust pre{\hypertarget{ref-RIETVELD200071}{}}%
-\CSLLeftMargin{2. }%
-\CSLRightInline{Rietveld, P.: The accessibility of railway stations: The
-role of the bicycle in the netherlands. Transportation Research Part D:
-Transport and Environment. 5, 71--75 (2000).
-https://doi.org/\href{https://doi.org/10.1016/S1361-9209(99)00019-X}{10.1016/S1361-9209(99)00019-X}.}
+Rietveld, P. (2000). The accessibility of railway stations: The role of
+the bicycle in the netherlands. \emph{Transportation Research Part D:
+Transport and Environment}, \emph{5}(1), 71--75.
+\url{https://doi.org/10.1016/S1361-9209(99)00019-X}
 
-\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
-\CSLLeftMargin{3. }%
-\CSLRightInline{INE:
-\href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}{Mobilidade
-e funcionalidade do território nas {Áreas Metropolitanas do Porto e de
-Lisboa}: 2017}. {Instituto National de Estatística}, Lisboa (2018).}
+\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
+Sartori, D., Catalano, G., Genco, M., Pancotti, C., Sirtori, E.,
+Vignetti, S., \& Del Bo, C. (2014). \emph{Guide to cost-benefit analysis
+of investment projects. Economic apraisal tool for cohesion policy
+2014-2020}. {European Commission - Directorate General for Regional and
+Urban Policy}.
+\url{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}
 
-\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
-\CSLLeftMargin{4. }%
-\CSLRightInline{Presidência do Conselho de Ministros: Resolução do
-conselho de ministros n.º 131/2019,
-\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}, (2019).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
-\CSLLeftMargin{5. }%
-\CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
-de apoio ao planeamento da rede ciclável na área metropolitana de
-Lisboa}, \url{https://biclar.tmlmobilidade.pt}, (2022).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
-\CSLLeftMargin{6. }%
-\CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
-Abbas, A., Woodcock, J.: The Propensity to Cycle Tool: An open source
-online system for sustainable transport planning. Journal of Transport
-and Land Use. 10, (2017).
-https://doi.org/\href{https://doi.org/10.5198/jtlu.2016.862}{10.5198/jtlu.2016.862}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
-\CSLLeftMargin{7. }%
-\CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
-Berkoff, N., Morgan, M., Abbas, A., Lovelace, R.: Scenarios of cycling
-to school in england, and associated health and carbon impacts:
-Application of the {`}propensity to cycle tool{'}. Journal of Transport
-\& Health. 12, 263--278 (2019).
-https://doi.org/\href{https://doi.org/10.1016/j.jth.2019.01.008}{10.1016/j.jth.2019.01.008}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
-\CSLLeftMargin{8. }%
-\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
-computationally efficient method for generating realistic route networks
-from origin-destination data. Findings. (2022).
-https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
-\CSLLeftMargin{9. }%
-\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
-C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
-transport networks with R5 in r. Findings. (2021).
-https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
-\CSLLeftMargin{10. }%
-\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
-jittering and routing options for converting origin-destination data
-into route networks: Towards accurate estimates of movement at the
-street level. The International Archives of the Photogrammetry, Remote
-Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
-(2022).
-https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
-\CSLLeftMargin{11. }%
-\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
-cyclists: Review of market segmentation methods for planning practice.
-Transportation Research Record. 2662, 125--133 (2017).
-https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
-\CSLLeftMargin{12. }%
-\CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
-Fernandez, A., Brand, C., Rojas Rueda, D., Woodcock, J., Kelly, P.,
-Lieb, C., Oja, P., others:
-\href{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}{Health
-economic assessment tool ({HEAT}) for walking and for cycling: Methods
-and user guide on physical activity, air pollution, injuries and carbon
-impact assessments}. (2017).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
-\CSLLeftMargin{13. }%
-\CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
-emission inventory guidebook 2019,
-\url{https://www.emisia.com/utilities/copert/documentation/}, (2020).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
-\CSLLeftMargin{14. }%
-\CSLRightInline{Carris:
-\href{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}{Relatório
-de sustentabilidade 2019 - demostração não financeira}. {Carris -
-Companhia Carris de Ferro de Lisboa, E.M., S.A.} (2020).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
-\CSLLeftMargin{15. }%
-\CSLRightInline{Metropolitano de Lisboa:
-\href{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}{Relatório
-integrado 2019}. {Metropolitano de Lisboa, E.P.E.} (2020).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
-\CSLLeftMargin{16. }%
-\CSLRightInline{CP:
-\href{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}{Relatório
-de sustentabilidade 2019}. {CP - Comboios de Portugal, E.P.E.} (2020).}
+\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
+Silva, C., Bravo, J. V., \& Gonçalves, J. (2021). \emph{{Impacto
+Económico e Social da Sinistralidade Rodoviária em Portugal}}. {Centro
+de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
+Rodoviária (ANSR)}.
+\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O
+Impacto Economico e Social da Sinistralidade - PT.pdf}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
-\CSLLeftMargin{17. }%
-\CSLRightInline{Transtejo:
-\href{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}{Relatório
-de sustentabilidade 2014}. {Grupo Transtejo, S.A.} (2014).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
-\CSLLeftMargin{18. }%
-\CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
-Sirtori, E., Vignetti, S., Del Bo, C.:
-\href{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}{Guide
-to cost-benefit analysis of investment projects. Economic apraisal tool
-for cohesion policy 2014-2020}. {European Commission - Directorate
-General for Regional and Urban Policy} (2014).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
-\CSLLeftMargin{19. }%
-\CSLRightInline{Bickel, P., Friedrich, R., Burgess, A., Fagiani, P.,
-Hunt, A., Jong, G.D., Laird, J.:
-\href{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}{{HEATCO
-- Developing Harmonised European Approaches for Transport Costing and
-Project Assessment. Deliverable 5, Proposal for Harmonised Guidelines.}}
-(2006).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
-\CSLLeftMargin{20. }%
-\CSLRightInline{Nash, C., others:
-\href{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}{UNIfication
-of accounts and marginal costs for transport efficiency: Final report}.
-{Institute for Transport Studies, University of Leeds} (2003).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
-\CSLLeftMargin{21. }%
-\CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
-\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
-Económico e Social da Sinistralidade Rodoviária em Portugal}}. {Centro
-de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
-Rodoviária (ANSR)} (2021).}
+Transtejo. (2014). \emph{Relatório de sustentabilidade 2014}. {Grupo
+Transtejo, S.A.}
+\url{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}
 
@@
 
-%
-% ---- Bibliography ----
-
 \end{document}

### 2023-11-22 paper/PaperCEUS/PaperCEUS.Rmd
remember things

@@
 
-Please make sure that your manuscript follows the guidelines in the 
-Guide for Authors of the relevant journal. It is not necessary to 
-typeset your manuscript in exactly the same way as an article, 
-unless you are submitting to a camera-ready copy (CRC) journal.
-
-For detailed instructions regarding the elsevier article class, see   <https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions>
-
 <!-- TODO: -->
 <!-- check if HEAT estimates remain the same (the tool is not working atm) -->
-<!-- should we ignore the social benefits component? and only estimate and add the environmental ones? or estimate the car to tp transfer Social benefits? maybe to the next paper? -->
+<!-- should we ignore the social benefits component? and only estimate and add the environmental ones? or estimate the car to PT transfer Social benefits? maybe to the next paper? -->
+<!-- should we keep both direct and safe routing is is it just noise? -->
 
@@
 
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced *biclaR*[^1], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
+To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].
 
@@
 
+There are also social benefits form shifting car trips to PT, nevertheless those were not estimated in this research.
+One of the main socioenvironmetal benefits, valued after monetization, comes from the increase in physical activity [@Felix2023ES].
+The literature shows that the Metabolic Equivalent Tasks (MET) for "riding in a bus or a train" is 1.3 plus the "walking for transportation" is 3.5, while driving a car is 2.5.
+see https://golf.procon.org/met-values-for-800-activities/
+The difference is not that much between them.
+Future works should also encompass the estimation of the social impacts for the PT leg of the journey, shifting from car.
+
 # Conclusion

### 2023-11-22 paper/PaperCEUS/PaperCEUS.tex
remember things

@@
 
-Please make sure that your manuscript follows the guidelines in the
-Guide for Authors of the relevant journal. It is not necessary to
-typeset your manuscript in exactly the same way as an article, unless
-you are submitting to a camera-ready copy (CRC) journal.
-
-For detailed instructions regarding the elsevier article class, see
-\url{https://www.elsevier.com/authors/policies-and-guidelines/latex-instructions}
-
 \hypertarget{introduction}{%
@@
   \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
-decision support tool that facilitates the design and development of a
-metropolitan cycling network (Félix et al., 2022).
+decision support tool that facilitates the planning, design, and
+development of a metropolitan cycling network (Félix et al., 2022).
 
@@
 
+There are also social benefits form shifting car trips to PT,
+nevertheless those were not estimated in this research. One of the main
+socioenvironmetal benefits, valued after monetization, comes from the
+increase in physical activity (Félix et al., 2023). The literature shows
+that the Metabolic Equivalent Tasks (MET) for ``riding in a bus or a
+train'' is 1.3 plus the ``walking for transportation'' is 3.5, while
+driving a car is 2.5. see
+https://golf.procon.org/met-values-for-800-activities/ The difference is
+not that much between them. Future works should also encompass the
+estimation of the social impacts for the PT leg of the journey, shifting
+from car.
+
 \hypertarget{conclusion}{%
@@
 \url{https://doi.org/10.3141/2662-14}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Felix2023ES}{}}%
+Félix, R., Orozco-Fontalvo, M., \& Moura, F. (2023). Socio-economic
+assessment of shared e-scooters: Do the benefits overcome the
+externalities? \emph{Transportation Research Part D: Transport and
+Environment}, \emph{118}, 103714.
+https://doi.org/\url{https://doi.org/10.1016/j.trd.2023.103714}
 

### 2023-11-22 paper/PaperCEUS/PaperCEUS.Rmd
minor paper changes

@@
 Table \ref{tab:summary21} shows the potential trips by PT mode to replace the second leg of the journey, in combination with cycling.
-Trains offer the greatest potential for substitution (88%).
+Train offers the greatest potential for substitution (88%).
 When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the Train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.
@@
 <!-- this table could be split into Social benefits and Environmental benefits -->
-Regarding the PT segment, the shift from private car usage would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly.
+Regarding the PT segment, the shift from private car would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly.
 
@@
 <!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? It shouldn't be ok to sum them -->
-Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year, 
-and the 10-year socio-environmental benefits account for €235 million to €620 million, depending on the cycling targets.
+Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year. 
+The 10-year socio-environmental benefits account for €235 million to €620 million, depending on the cycling targets.
 
@@
 
-The information on socio-economic benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation, and to better communicate their decisions by providing the expected avoided GHG and air pollutant emissions and the monetized socio-economic benefits for short and long terms.
+The information on socio-economic benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation, and to better communicate their decisions by providing the expected avoided GHG and air pollutant emissions, and the monetized socio-economic benefits for short and long terms.
 
 The information available at *biclaR* tool -- an open access website -- can be downloaded and used with any GIS software.
-This allows users to, for example, gain insights into which potential cycling connections have the highest socio-environmental impacts, quantified in tons of avoided CO~2~eq emissions, or in long term social benefits.
+This allows practitioners to, for example, gain insights into which potential cycling connections have the highest socio-environmental impacts, quantified in tons of avoided CO~2~eq emissions, or in long term social benefits.
 <!-- move to introduction?
@@
 
-By making the research process publicly accessible in a code repository, this research enables the replication of similar estimates for socio-environmental impacts, resulting from a modal shift from car to bicycle in combination with PT, in other metropolitan areas.
+By making the research process publicly accessible in a code repository, it enables the replication of similar estimates for socio-environmental impacts, resulting from a modal shift from car to bicycle in combination with PT, in other metropolitan areas.
 
 ### Acknowledgements. {.unnumbered}
-This research was funded by the Lisbon's Metropolitan Department of Transport (TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.), under the *biclaR* Project.
-This work is part of the research activity carried out at Civil Engineering Research and Innovation for Sustainability (CERIS) and has been funded by Fundação para a Ciência e a Tecnologia (FCT), Portugal in the framework of project UIDB/04625/2020.
-The authors thank Thomas Götschi (HEAT for Cycling) for providing access to HaaS tool, which is under development.
+[*blind*]
+<!-- This research was funded by the Lisbon's Metropolitan Department of Transport (TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.), under the *biclaR* Project. -->
+<!-- This work is part of the research activity carried out at Civil Engineering Research and Innovation for Sustainability (CERIS) and has been funded by Fundação para a Ciência e a Tecnologia (FCT), Portugal in the framework of project UIDB/04625/2020.
+-->
+<!-- The authors thank Thomas Götschi (HEAT for Cycling) for providing access to HaaS tool, which is under development. -->
 

### 2023-11-22 paper/PaperCEUS/PaperCEUS.tex
minor paper changes

@@
 replace the second leg of the journey, in combination with cycling.
-Trains offer the greatest potential for substitution (88\%). When
+Train offers the greatest potential for substitution (88\%). When
 comparing the existing PT interfaces (Figure \ref{fig:map1}) with the
@@
 trips with potential TP trips. Regarding the PT segment, the shift from
-private car usage would lead to the mitigation of CO\textsubscript{2}
+private car would lead to the mitigation of CO\textsubscript{2}
 equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4
@@
 Shifting from car to cycling + PT can reduce annual
-CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year, and
-the 10-year socio-environmental benefits account for €235 million to
-€620 million, depending on the cycling targets.
+CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year. The
+10-year socio-environmental benefits account for €235 million to €620
+million, depending on the cycling targets.
 
@@
 motorized transportation, and to better communicate their decisions by
-providing the expected avoided GHG and air pollutant emissions and the
+providing the expected avoided GHG and air pollutant emissions, and the
 monetized socio-economic benefits for short and long terms.
@@
 website -- can be downloaded and used with any GIS software. This allows
-users to, for example, gain insights into which potential cycling
-connections have the highest socio-environmental impacts, quantified in
-tons of avoided CO\textsubscript{2}eq emissions, or in long term social
-benefits.
+practitioners to, for example, gain insights into which potential
+cycling connections have the highest socio-environmental impacts,
+quantified in tons of avoided CO\textsubscript{2}eq emissions, or in
+long term social benefits.
 
 By making the research process publicly accessible in a code repository,
-this research enables the replication of similar estimates for
-socio-environmental impacts, resulting from a modal shift from car to
-bicycle in combination with PT, in other metropolitan areas.
+it enables the replication of similar estimates for socio-environmental
+impacts, resulting from a modal shift from car to bicycle in combination
+with PT, in other metropolitan areas.
 
@@
 
-This research was funded by the Lisbon's Metropolitan Department of
-Transport (TML - Transportes Metropolitanos de Lisboa, E.M.T., S.A.),
-under the \emph{biclaR} Project. This work is part of the research
-activity carried out at Civil Engineering Research and Innovation for
-Sustainability (CERIS) and has been funded by Fundação para a Ciência e
-a Tecnologia (FCT), Portugal in the framework of project
-UIDB/04625/2020. The authors thank Thomas Götschi (HEAT for Cycling) for
-providing access to HaaS tool, which is under development.
+{[}\emph{blind}{]}
 
@@
 Environment}, \emph{118}, 103714.
-https://doi.org/\url{https://doi.org/10.1016/j.trd.2023.103714}
+\url{https://doi.org/10.1016/j.trd.2023.103714}
 

### 2023-11-23 HEAT/code/haas_reprex.R
HaaS update

@@ -1,8 +1,13 @@
 
 
-# read RDS exported from UI
-webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_webapp_input.rds"))
-genericdata = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_generic_data_relevant_with_calc_params.rds"))
+# read RDS exported from UI _OLD
+# webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_webapp_input.rds"))
+# genericdata = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/1304_generic_data_relevant_with_calc_params.rds"))
+
+
+# read RDS exported from UI - NEW VERSION
+webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/2023_52_1304_webapp_input.rds"))
+genericdata = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/2023_52_1304_generic_data_relevant_with_calc_params.rds"))
 
 
 ## 1. Input and returned data as json
@@ -117,6 +122,32 @@ from_rds_reprex <- unserialize(httr::content(resp_reprex, "raw"))
 
 
 
+# REPREX ------------------------------------------------------------------
+
+# read RDS exported from UI - NEW VERSION
+webapp_input = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/2023_52_1304_webapp_input.rds"))
+genericdata = readRDS(url("https://web.tecnico.ulisboa.pt/~rosamfelix/heat/2023_52_1304_generic_data_relevant_with_calc_params.rds"))
+
+# to get data prepared by "calc_results"
+# input data as "json", two inputs - "generic_data_relevant" from previous step 
+# returned data as "rds"
+resp <- 
+  httr::POST("https://api.heatwalkingcycling.org/apiv1/heat_5_0/calc_results?output_format=rds",
+             httr::set_cookies("user" = Sys.getenv("HEAT_COOKIE")),
+             body = jsonlite::toJSON(
+               list("webapp_input" = webapp_input,
+                    "generic_data_relevant" = genericdata)),
+             encode = "multipart")
+
+# convert to R
+from_rds <- unserialize(httr::content(resp, "raw"))
+
+
+# check the response
[…]

### 2023-11-26 code/test-code/heat-prep_50.R
old version of heat haas 5.0

@@ -1,0 +1,452 @@
+# prepare tables for HEAT SE estimations
+
+
+# locations id
+library(readr)
+HEAT_all_locations_data <- read_csv("D:/GIS/TML/HEAT_heatr_api/all_locations_data.csv") #data from https://github.com/HEAT-WHO/HEAT_data/blob/main/data/heatr/locations_data/all_locations_data.csv
+HEAT_all_locations_data = HEAT_all_locations_data %>% filter(country_location_name == "Portugal")
+HEAT_all_locations_data = HEAT_all_locations_data %>% filter(sub1_location_name %in% c("Distrito de Lisboa", "Distrito de Setúbal"))
+HEAT_all_locations_data = HEAT_all_locations_data %>% filter(is.na(city_location_id))
+saveRDS(HEAT_all_locations_data, "HEAT/HEAT_locations_data_AML.Rds")
+
+
+
+# PM25 por municipio, incluindo AML
+PM25 = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "PM25")
+PM25$PM25 = round(PM25$PM25, 2)
+PM25$location_id = as.character(PM25$location_id)
+
+# VSL Portugal 2017
+usd_to_eur = 0.976430 #21nov2022
+VSL_intusdPPP = 3055358/usd_to_eur #fonte: ANSR 2021. original = 3355000
+# VSL_eurotousd = 2323000 #? 
+# VSL_usdMER = 2194000
+#https://www.oecd-ilibrary.org/environment/mortality-risk-valuation-in-environment-health-and-transport-policies_9789264130807-en
+
+# CENARIOS combinations
+CENARIOS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "CENARIOS")
+CENARIOS$dist_max[CENARIOS$Estrategia == "Convencional"] = 5000 
+CENARIOS$dist_max[CENARIOS$Estrategia == "Ebike"] = 10000 
+CENARIOS$dist_max[CENARIOS$Estrategia == "Intermodal"] = 5000 
+
+# File paths
+PATHS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "paths")
+
+
+# proportion in traffic vs. away from major roads. carspeed > 30 ?? 50 is too high, as a road in 40 is still in traffic.
+# População movel AML = 80.4%
+# Deslocacoes / pessoa / dia = 2.6 AML
+
+
+
+# script to get values -------------------------------------------------------------------------
+
+# Code == X
+
[…]

### 2023-11-27 code/test-code/heat_prep_52_originalvalues.R
found the compatibility between HaaS 5.0 and 5.2
see
https://github.com/HEAT-WHO/heatr_api/issues/17

@@ -1,0 +1,1048 @@
+library(tidyverse)
+library(sf)
+
+## For Intermodality ---------------------------------------------------------------------------
+ # ORIGINAL VALUES - but using 5.2 version
+
+Code_Hass = CENARIOS$Code[9:12] #fazer por enquanto só AML e senários 1 e 2
+
+HEATbind_intermodal_original = HEAT
+
+for (i in Code_Hass){
+  
+  # i = 3304
+  
+  HEAT = CENARIOS %>% 
+    mutate(Year_ref = `0_Ano`,
+           Year_cf = `1_Ano`) %>% 
+    select(NMunicipio, Municipio, Code, ENMAC, dist_max, Year_ref, Year_cf) %>%
+    filter(Code == i) %>% 
+    left_join(PATHS) %>% 
+    left_join(PM25) %>% 
+    mutate(ENMAC = ENMAC/100)
+  
+  HEAT_bike = readRDS(HEAT$routes_filepath) %>% 
+    ungroup() %>% 
+    st_drop_geometry() %>%
+    group_by(id) %>% 
+    summarise(distance = sum(distance), 
+              Bike = Bike,
+              Car = Car + CarP,
+              Total = Total,
+              Bikeper = Bike / Total) %>% 
+    ungroup() %>% 
+    mutate(
+      cyc = ifelse(Bikeper >= HEAT$ENMAC, Bike, HEAT$ENMAC * Total),
+      new_cyc = ifelse(Bikeper >= HEAT$ENMAC, 0, cyc - Bike)
+    )
+  
+  
+  HEAT_car = readRDS(HEAT$car_routes_filepath) %>%
+    ungroup() %>% 
+    filter(id %in% unique(HEAT_bike$id)) %>% 
+    group_by(id) %>% 
+    summarise(distance = sum(distance), 
+              Bike = Bike,
+              Car = Car + CarP,
+              Total = Total,
+              Bikeper = Bike / Total) %>% 
+    mutate(
+      cyc = ifelse(Bikeper >= HEAT$ENMAC, Bike, HEAT$ENMAC * Total),
+      new_cyc = ifelse(Bikeper >= HEAT$ENMAC, 0, cyc - Bike)) %>% 
+    mutate(Car_new = Car - new_cyc)
+  
+  Trips_rnet = readRDS(HEAT$rnet_filepath) %>% filter(Bike>0)
+  InTraffic = Trips_rnet %>% filter(carspeed > 30)
+  
[…]

### 2023-11-27 code/test-code/heat-prep_52.R
estimates intermodal v5.2
re-run estimates using only half trips
#43

@@ -1,0 +1,454 @@
+# prepare tables for HEAT SE estimations
+
+
+# locations id
+library(readr)
+HEAT_all_locations_data <- read_csv("D:/GIS/TML/HEAT_heatr_api/all_locations_data.csv") #data from https://github.com/HEAT-WHO/HEAT_data/blob/main/data/heatr/locations_data/all_locations_data.csv
+HEAT_all_locations_data = HEAT_all_locations_data %>% filter(country_location_name == "Portugal")
+HEAT_all_locations_data = HEAT_all_locations_data %>% filter(sub1_location_name %in% c("Distrito de Lisboa", "Distrito de Setúbal"))
+HEAT_all_locations_data = HEAT_all_locations_data %>% filter(is.na(city_location_id))
+saveRDS(HEAT_all_locations_data, "HEAT/HEAT_locations_data_AML.Rds")
+
+
+
+# PM25 por municipio, incluindo AML
+PM25 = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "PM25")
+PM25$PM25 = round(PM25$PM25, 2)
+PM25$location_id = as.character(PM25$location_id)
+
+# VSL Portugal 2017
+usd_to_eur = 0.976430 #21nov2022
+VSL_intusdPPP = 3055358/usd_to_eur #fonte: ANSR 2021. original = 3355000
+# VSL_eurotousd = 2323000 #? 
+# VSL_usdMER = 2194000
+#https://www.oecd-ilibrary.org/environment/mortality-risk-valuation-in-environment-health-and-transport-policies_9789264130807-en
+
+# CENARIOS combinations
+CENARIOS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "CENARIOS")
+CENARIOS$dist_max[CENARIOS$Estrategia == "Convencional"] = 5000 
+CENARIOS$dist_max[CENARIOS$Estrategia == "Ebike"] = 10000 
+CENARIOS$dist_max[CENARIOS$Estrategia == "Intermodal"] = 5000 
+
+# File paths
+PATHS = readxl::read_excel("export2/Scenarios_Routing.xlsx", sheet = "paths")
+
+
+# proportion in traffic vs. away from major roads. carspeed > 30 ?? 50 is too high, as a road in 40 is still in traffic.
+# População movel AML = 80.4%
+# Deslocacoes / pessoa / dia = 2.6 AML
+
+
+
+# script to get values -------------------------------------------------------------------------
+
+# Code == X
+
[…]

### 2023-11-27 code/test-code/heat-prep_52.R
piggyback heat intermodality

@@ -452,3 +452,5 @@ for (i in Code_Hass){
 HEATbind_intermodal_correct = HEATbind_intermodal_correct[-1,]
 # HEATbind_intermodal_correct_PPP = HEATbind_intermodal_correct
 saveRDS(HEATbind_intermodal_correct, "HEAT/HEAT_AML_intermodal_ppp_NEW.Rds")
+
+piggyback::pb_upload("HEAT/HEAT_AML_intermodal_ppp_NEW.Rds")

### 2023-11-27 paper/PaperCEUS/PaperCEUS.Rmd
replace tables paper

@@
 
-For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.0 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
+For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.2 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
 The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
@@
 Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles.
-For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are also presented, resulting from replacing car trips with cycling.
+For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in \ref{tab:summary1b}, resulting from replacing car trips with cycling.
+
+```{r summary0}
+summary0 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/HEAT_AML_intermodal_ppp_NEW.Rds"))
+# see code/test-code/heat-prep.R#intermodality
+
+summary0 = summary0 |> 
+    mutate(Target = paste0(ENMAC*100,"%"),
+           Routing = c("safe", "safe", "direct", "direct"),
+           TotalTrips = Total,
+           Bikebase = Bike,
+           Bikepotential = Bike_new,
+           Social = Economic - Environmenal_only,
+           Social10 = Economic10 - Environmenal10_only) 
+```
 
 ```{r summary1}
-summary1 = readRDS("load/summary1.Rds")
-
-summary1 = summary1 %>%
-  mutate(LTS = recode(LTS, `3` = "safe", `4` = "direct")) |> 
-  mutate(Estrategia = paste0(Estrategia*100,"%")) |> 
-  mutate(TotalTrips = TotalTrips/2,
-         Bikebase = Bikebase/2,
-         Bikepotential = Bikepotential/2) #this is a quick correction while we donr run biclar entirely again.
-The error was found and it was adding up the 2 bike legs, instead of making the mean.
-The SE estimates remain the same because they are from the bikeshare and carshare percentages, that did not change.
+summary1 = summary0 |> 
+  select(Target, Routing, TotalTrips, Bikebase, Bikepotential) |> 
+    arrange(desc(Target))
 
@@
       col.names = c("Target", "Routing", "Total trips",
-                    "Baseline Cycling + PT", "Potencial Cycling + PT",
-                    "Avoided CO2eq (ton/yr)", "SE Benefits for 10 years (thousand €)"),
-      caption = "\\label{summary1}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.",
+                    "Baseline Cycling + PT", "Potencial Cycling + PT"),
+      caption = "\\label{summary1}Summary of the cycling potencial of the intermodality scenario.",
       format="latex",
       booktabs = TRUE) |> 
-  kableExtra::column_spec(column = c(4:7), width = "6em")
+  kableExtra::column_spec(column = c(4:5), width = "7em")
+
+```
+
+```{r summary1b}
+summary1b = summary0 |> 
+  select(Target, Routing, Mortality, Social, CO2eq, Environmenal_only) |> 
+  mutate(Social = Social/1000,
+         Environmenal_only = Environmenal_only/1000) |> 
+    arrange(desc(Target))
+
+knitr::kable(summary1b,
+      digits = c(0,0,1,0,0,0),
+      format.args = list(big.mark = " "),
+      col.names = c("Target", "Routing",
+                    "Avoided Mortality (deaths/yr)", "Social benefits (k€/yr)", 
+                    "Avoided CO2eq (ton/yr)", "Environmental benefits (k€/yr)"),
+      caption = "\\label{summary1b}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.",
+      format="latex",
+      booktabs = TRUE) |> 
+  kableExtra::column_spec(column = c(3:6), width = "6em")
 
@@
 summary22 = summary22 %>%
-  mutate(LTS = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
-  mutate(Value = money_emissions_eur/1000) |> 
-  select(Meta_ENMAC, LTS, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, Value) |> 
-  arrange(Meta_ENMAC)
+  mutate(Routing = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
+  mutate(Value = money_emissions_eur/1000,
+         Target = Meta_ENMAC) |> 
+  select(Target, Routing, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, Value) |> 
+  arrange(Target)
 
@@
 ```{r summaryall, message=FALSE, warning=FALSE}
-summary0 = readRDS("load/HEAT_aml_results.Rds") |> slice_tail(n = 4)
+
 rate = ((1+0.03)^10)/((1+0.05)^10) # inflation rate 3% discount rate 5%
 
-summaryall = summary1 |> 
-  mutate(Meta_ENMAC = Estrategia) |>
+summaryall = summary0 |> 
   left_join(summary22) |> 
-  mutate(Code = c(3304, 3404, 3310, 3410)) |> 
-  left_join(summary0 |> select(Code, Economic), by = "Code") |> 
-  select(Estrategia, LTS, co2, CO2eq_TP, monet, Value, Economic) |> 
-  mutate(CO2eq = co2 + CO2eq_TP,
-         value1 = Economic/1000 + Value, # this is CORRECT
-         value10 = monet + Value*10*rate) |> # this is CORRECT
-  select(Estrategia, LTS, CO2eq, value1, value10)
+  select(Target, Routing, CO2eq, CO2eq_TP, Environmenal_only, Environmenal10_only, Social, Social10, Value) |> 
+    mutate(Social = Social/1000,
+           Social10 = Social10/1000,
+         Environmenal_only = Environmenal_only/1000,
+         Environmenal10_only = Environmenal10_only/1000) |> 
+  mutate(CO2eqT = CO2eq + CO2eq_TP,
+         value1 = Environmenal_only + Value + Social, # this is CORRECT
+         value10 = Environmenal10_only + Value*10*rate + Social10) |> # this is CORRECT
+  select(Target, Routing, CO2eqT, value1, value10)
 

### 2023-11-27 paper/PaperCEUS/PaperCEUS.tex
replace tables paper

@@
 socio-environmental impacts were estimated, using the HEAT for Cycling
-tool v5.0 (Kahlmeier et al., 2017) from the World Health Organization,
+tool v5.2 (Kahlmeier et al., 2017) from the World Health Organization,
 and the
@@
 last legs), the environmental avoided emissions and monetized
-socio-environment (SE) benefits are also presented, resulting from
-replacing car trips with cycling.
+socio-environment (SE) benefits are presented in \ref{tab:summary1b},
+resulting from replacing car trips with cycling.
 
@@
 
-\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.}
+\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario.}
 \centering
-\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}}
+\begin{tabular}[t]{llr>{\raggedleft\arraybackslash}p{7em}>{\raggedleft\arraybackslash}p{7em}}
 \toprule
-Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT & Avoided CO2eq (ton/yr) & SE Benefits for 10 years (thousand €)\\
+Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT\\
 \midrule
-4\% & safe & 538 514 & 2 312 & 20 385 & 5 920 & 230 270\\
-4\% & direct & 500 880 & 2 274 & 18 944 & 6 011 & 223 720\\
-10\% & safe & 538 514 & 2 312 & 52 324 & 15 192 & 591 790\\
-10\% & direct & 500 880 & 2 274 & 48 609 & 15 414 & 574 200\\
+4\% & safe & 538 514 & 2 312 & 20 385\\
+4\% & direct & 500 880 & 2 274 & 18 944\\
+10\% & safe & 538 514 & 2 312 & 52 323\\
+10\% & direct & 500 880 & 2 274 & 48 609\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:summary1b}\label{summary1b}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.}
+\centering
+\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}>{\raggedleft\arraybackslash}p{6em}}
+\toprule
+Target & Routing & Avoided Mortality (deaths/yr) & Social benefits (k€/yr) & Avoided CO2eq (ton/yr) & Environmental benefits (k€/yr)\\
+\midrule
+4\% & safe & 4.1 & 12 717 & 2 958 & 238\\
+4\% & direct & 4.0 & 12 441 & 3 004 & 241\\
+10\% & safe & 10.0 & 32 820 & 7 590 & 610\\
+10\% & direct & 10.0 & 31 800 & 7 694 & 618\\
 \bottomrule
@@
 \midrule
-4\% & safe & 14 513 & 27 299 & 242 024\\
-4\% & direct & 14 713 & 26 572 & 235 706\\
-10\% & safe & 35 819 & 69 938 & 620 094\\
-10\% & direct & 36 207 & 67 956 & 602 972\\
+4\% & safe & 11 551 & 14 380 & 127 534\\
+10\% & safe & 28 217 & 36 861 & 325 814\\
+4\% & direct & 11 706 & 14 135 & 125 016\\
+10\% & direct & 28 487 & 35 905 & 318 062\\
 \bottomrule

### 2023-11-27 paper/PaperCEUS/PaperCEUS.Rmd
paper values update + MET

@@
   The results indicate that 10% of trips could be made by bicycle + PT combination.
-Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of at least €230 million.
-For the PT leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
+Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the PT leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
+The estimated socio-environmental benefits are of €125 million to €325 million over 10 years.
   This evidence can support policymakers to prioritize interventions that reduce the reliance on private motor vehicles.
@@
 <!-- TODO: -->
-<!-- check if HEAT estimates remain the same (the tool is not working atm) -->
 <!-- should we ignore the social benefits component? and only estimate and add the environmental ones? or estimate the car to PT transfer Social benefits? maybe to the next paper? -->
@@
 # Introduction
-
-<!-- **full paper**: 4-6 pages in length (typically up to 3,000 words). we have 2849 words before references, or 2278 words without tables -->
 
@@
 
-<!-- The quietness level^[see https://www.cyclestreets.net/help/journey/howitworks/#quietness] was another indicator estimated by [CycleStreets](https://www.cyclestreets.org/) for each route section, based on road network variables xisting on OpenStreetMap labels, such as the number of lanes, maximum allowed speed, trees, or road hierarchy.
-Values vary on a scale of 0 to 100, where zero corresponds to the least safe level for cycling and 100 corresponds to the safest and quietest level, usually corresponding to sections that already include segregated cycling infrastructure away from motorized traffic.
-I THINK THIS IS NOT RELEVANT FOR THIS SHORT PAPER, AS IT IS MOST USEFUL FOR VIZUALIZATION PURPOSES ON BICLAR TOOL-->
-
 The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
@@
 The routes were then overlaid and aggregated by segments, using [`stplanr overline()` R function](https://docs.ropensci.org/stplanr/reference/overline.html).
-<!-- For example, the estimated volumes of bicycle trips and car transfers were aggregated, and the average car speed and quietness level, weighted by the distance of the segments, were calculated.
--->
 
@@
 The intermodality scenario considers trips combining PT and cycling for the first and last legs.
-In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for instance: 1 km from origin to interface A plus 4 km from interface B to destination) or up to 25 minutes on bike.
+In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for instance: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike.
 Furthermore, we have imposed restrictions on PT usage to include only trips with no PT transfers, and up to 2 hours (120 min).
@@
 
-```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area", fig.align='center'}
+```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area.", fig.align='center'}
 knitr::include_graphics("img/map1.png", error = FALSE)
@@
 
-For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.2 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
+For the *cycling legs of the journey* (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.2 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
+The use of this package made it possible to run multiple scenarios with few changes in input values, making the interaction with HEAT more reliable when reproducing runs.    
 The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
-We considered two dimensions: *social* — including the physical activity of cyclists, air pollution exposure, and road casualties; *environmental* — including CO~2~eq emissions and other pollutants.
+It estimates the differences between two considered scenarios.
+In this case: one baseline scenario, with data from the mobility survey, and one cycling potential scenario in which targets of 4% and 10% of cycling levels were achieved, transferred from car trips.
+We considered two dimensions: *social* — including the physical activity, air pollution exposure, and road casualties; and *environmental* — including CO~2~eq emissions and other pollutants.
 
@@
 <!-- For the full paper, consider refer the input HEAT values (tabela 6) -->
-<!-- Explain why the use of the HaaS tool instead of the website (3 scenarios, 18 municipalities, 2 routing profiles) -->
+
+<!-- include a map or image with the legs and all trip? by mode  -->
+
 <!-- Also consider put the FE formula and emmission factor values - see report -->
 
-For the second leg of the journey, we estimate the additional environmental impacts of shifting car trips to PT (between the PT interfaces).
+For the *second leg of the journey*, we estimate the environmental impacts of shifting car trips to PT (between the PT interfaces).
 
@@
          value10 = Environmenal10_only + Value*10*rate + Social10) |> # this is CORRECT
-  select(Target, Routing, CO2eqT, value1, value10)
+  select(Target, Routing, CO2eqT, value1, value10) |> 
+  arrange(desc(Target))
 
@@
 
-<!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? It shouldn't be ok to sum them -->
+<!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? -->
 Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year. 
-The 10-year socio-environmental benefits account for €235 million to €620 million, depending on the cycling targets.
+The 10-year socio-environmental benefits account for €125 million to €325 million, depending on the cycling targets.
 
-The social impacts represent 98% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs.
-For the PT segment, we did not estimate the social impacts from substituting car trips, although its health benefits would not be as high as shifting to cycling.<!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper --> <!--Overall there was very little evidence available on mode switch to public transport.
--->
+The environmental impacts represent less than 2% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs. 
+For the PT segment, we did not estimate the social impacts from substituting car trips. 
+One of the main socio-environmetal benefits, valued after monetization, comes from the increase in physical activity [@Felix2023ES].
+Although are also social benefits form shifting car trips to PT, its health benefits would not be as high as shifting to cycling. 
+The literature shows that the Metabolic Equivalent Tasks (MET) for "riding in a bus or a train" is 1.3 plus the "walking for transportation" as 3.5, while "driving a car" is 2.5 [@MET2011].
+The difference between these activities is not very obvious when compared to cycling, whose equivalent MET is between 6.8 to 7.5.
+Nevertheless, future works should also encompass the estimation of the social impacts for the PT leg of the journey, shifting from car.
+ <!--Overall there was very little evidence available on mode switch to public transport. -->
 
-The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 75% of the emissions avoided during the PT segment.
+The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 74% of the emissions avoided during the PT segment.
 This finding, while expected -- due the zero cycling emissions, should not be overlooked when promoting the PT use.
@@
 Our findings suggest that cycling and PT *in combination* could viably replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution. 
-<!-- repeated? not correct -->
-
-There are also social benefits form shifting car trips to PT, nevertheless those were not estimated in this research.
-One of the main socioenvironmetal benefits, valued after monetization, comes from the increase in physical activity [@Felix2023ES].
-The literature shows that the Metabolic Equivalent Tasks (MET) for "riding in a bus or a train" is 1.3 plus the "walking for transportation" is 3.5, while driving a car is 2.5.
-see https://golf.procon.org/met-values-for-800-activities/
-The difference is not that much between them.
-Future works should also encompass the estimation of the social impacts for the PT leg of the journey, shifting from car.
+<!-- repeated? -->
 

### 2023-11-27 paper/PaperCEUS/PaperCEUS.tex
paper values update + MET

@@
   mile stages can reduce annual CO\textsubscript{2}eq emissions from
-  6,000 tons/day, with benefits over 10 years of at least €230 million.
-  For the PT leg, the transfer from car avoids of up to 20,500 tons of
-  CO\textsubscript{2}eq emissions per year. This evidence can support
-  policymakers to prioritize interventions that reduce the reliance on
-  private motor vehicles.\\
+  3,000 to 7,500 tons/day, while for the PT leg, the transfer from car
+  avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per
+  year. The estimated socio-environmental benefits are of €125 million
+  to €325 million over 10 years. This evidence can support policymakers
+  to prioritize interventions that reduce the reliance on private motor
+  vehicles.\\
   \end{abstract}
@@
 our analysis to the first and last legs with a combined length of up to
-5 km (for instance: 1 km from origin to interface A plus 4 km from
-interface B to destination) or up to 25 minutes on bike. Furthermore, we
-have imposed restrictions on PT usage to include only trips with no PT
-transfers, and up to 2 hours (120 min). Additionally, we have only
-included PT modes that can easily accommodate bicycles, such as trains,
-ferries, trams, and inter-municipal bus lines equipped with bike racks
-(Figure \ref{fig:map1}).
+5 km (for instance: 1 km from origin to interface \emph{A} plus 4 km
+from interface \emph{B} to destination) or up to 25 minutes on bike.
+Furthermore, we have imposed restrictions on PT usage to include only
+trips with no PT transfers, and up to 2 hours (120 min). Additionally,
+we have only included PT modes that can easily accommodate bicycles,
+such as trains, ferries, trams, and inter-municipal bus lines equipped
+with bike racks (Figure \ref{fig:map1}).
 
@@
 
-\caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
+\caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area.}\label{fig:map1}
 \end{figure}
@@
 
-For the cycling legs of the journey (first and last legs),
+For the \emph{cycling legs of the journey} (first and last legs),
 socio-environmental impacts were estimated, using the HEAT for Cycling
@@
   \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
-HEAT tool provided estimates on the shifting from car to cycling for a
-short term time horizon (i.e., one year) and the long term (i.e., ten
-years). We considered two dimensions: \emph{social} --- including the
-physical activity of cyclists, air pollution exposure, and road
-casualties; \emph{environmental} --- including CO\textsubscript{2}eq
-emissions and other pollutants.
+use of this package made it possible to run multiple scenarios with few
+changes in input values, making the interaction with HEAT more reliable
+when reproducing runs.\\
+The HEAT tool provided estimates on the shifting from car to cycling for
+a short term time horizon (i.e., one year) and the long term (i.e., ten
+years). It estimates the differences between two considered scenarios.
+In this case: one baseline scenario, with data from the mobility survey,
+and one cycling potential scenario in which targets of 4\% and 10\% of
+cycling levels were achieved, transferred from car trips. We considered
+two dimensions: \emph{social} --- including the physical activity, air
+pollution exposure, and road casualties; and \emph{environmental} ---
+including CO\textsubscript{2}eq emissions and other pollutants.
 
-For the second leg of the journey, we estimate the additional
-environmental impacts of shifting car trips to PT (between the PT
-interfaces).
+For the \emph{second leg of the journey}, we estimate the environmental
+impacts of shifting car trips to PT (between the PT interfaces).
 
@@
 and its monetization for the second leg of the journey, by replacing car
-trips with potential TP trips. Regarding the PT segment, the shift from
-private car would lead to the mitigation of CO\textsubscript{2}
-equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4
-million to €3.5 million yearly.
+trips with potential TP trips.\\
+Regarding the PT segment, the shift from private car would lead to the
+mitigation of CO\textsubscript{2} equivalent emissions to 8,500 to
+20,800 tons annually, valued in €1.4 million to €3.5 million yearly.
 
@@
 4\% & safe & 11 551 & 14 380 & 127 534\\
+4\% & direct & 11 706 & 14 135 & 125 016\\
 10\% & safe & 28 217 & 36 861 & 325 814\\
-4\% & direct & 11 706 & 14 135 & 125 016\\
 10\% & direct & 28 487 & 35 905 & 318 062\\
@@
 CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year. The
-10-year socio-environmental benefits account for €235 million to €620
+10-year socio-environmental benefits account for €125 million to €325
 million, depending on the cycling targets.
 
-The social impacts represent 98\% of the socio-environmental benefits
-(in value) from replacing car trips to bicycle in first-and-last legs.
-For the PT segment, we did not estimate the social impacts from
-substituting car trips, although its health benefits would not be as
-high as shifting to cycling.
+The environmental impacts represent less than 2\% of the
+socio-environmental benefits (in value) from replacing car trips to
+bicycle in first-and-last legs. For the PT segment, we did not estimate
+the social impacts from substituting car trips. One of the main
+socio-environmetal benefits, valued after monetization, comes from the
+increase in physical activity (Félix et al., 2023). Although are also
+social benefits form shifting car trips to PT, its health benefits would
+not be as high as shifting to cycling. The literature shows that the
+Metabolic Equivalent Tasks (MET) for ``riding in a bus or a train'' is
+1.3 plus the ``walking for transportation'' as 3.5, while ``driving a
+car'' is 2.5 (Ainsworth et al., 2011). The difference between these
+activities is not very obvious when compared to cycling, whose
+equivalent MET is between 6.8 to 7.5. Nevertheless, future works should
+also encompass the estimation of the social impacts for the PT leg of
+the journey, shifting from car.
 
 The emissions of CO\textsubscript{2}eq that are avoided during both the
-initial and final journey segments account for about 75\% of the
+initial and final journey segments account for about 74\% of the
 emissions avoided during the PT segment. This finding, while expected --
@@
 journeys prone to further substitution.
-
-There are also social benefits form shifting car trips to PT,
-nevertheless those were not estimated in this research. One of the main
-socioenvironmetal benefits, valued after monetization, comes from the
-increase in physical activity (Félix et al., 2023). The literature shows
-that the Metabolic Equivalent Tasks (MET) for ``riding in a bus or a
-train'' is 1.3 plus the ``walking for transportation'' is 3.5, while
-driving a car is 2.5. see
-https://golf.procon.org/met-values-for-800-activities/ The difference is
-not that much between them. Future works should also encompass the
-estimation of the social impacts for the PT leg of the journey, shifting
-from car.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-MET2011}{}}%
+Ainsworth, B. E., Haskell, W. L., Herrmann, S. D., Meckes, N., Bassett
+Jr, D. R., Tudor-Locke, C., Greer, J. L., Vezina, J., Whitt-Glover, M.
+C., \& Leon, A. S. (2011). {2011 Compendium of Physical Activities: A
+second update of codes and MET values}. \emph{Medicine \& Science in
+Sports \& Exercise}, \emph{43}(8), 1575--1581.
+\url{https://doi.org/10.1249/MSS.0b013e31821ece12}
+
 \leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
@@
 school in england, and associated health and carbon impacts: Application
-of the {`}propensity to cycle tool{'}. \emph{Journal of Transport and
+of the {`}propensity to cycle tool{'}. \emph{Journal of Transport \&
 Health}, \emph{12}, 263--278.

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
paper revs from FM
@filmoura

@@
 ---
-title: "biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
+title: "XXX: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
 author:
@@
 This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
-  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_).
+  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_XXX_).
 The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
@@
 
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].
+To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced *XXX*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network (ref X).
+<!-- [@felix2023]. -->
 
-[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/).
+[^1]: See https://xxx.xxx. *For double blind peer-review process, the authors decided to truncate the name and references to this tool*.
+<!-- [^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/). -->
 
-*biclaR* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
+*XXX* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
 The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
@@
 We used a method for disaggregating the origins and destinations of trips between the centroids of two districts (same as "parish") to ensure that a district is not solely characterized by a single point of origin and destination for its trips.
-Aggregating all trips into centroids renders the exercise less realistic, as it excludes a significant portion of short-distance trips, a prevalent characteristic of active mode travel [@Lovelace2022Jittering].
+Aggregating all trips into centroids renders the exercise less realistic, as it excludes a significant portion of short-distance trips, a prevalent characteristic of active travel [@Lovelace2022Jittering].
 The OD Jittering method breaks down a single point (i.e., the centroid of an area) into multiple random points on the existing and neighboring road network, using OpenStreetMap as a reference.
@@
 
-```{r jitter, fig.align='center', fig.cap="Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
+```{r jitter, fig.align='center', fig.cap="Representation of desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
 
@@
 Other constraints regarding the behavior of potential cyclists determine the routing results.
-For example, such restrictions can favor low speed, low traffic streets, more direct routes, and less steep paths, among others, suitable for cycling.
+For example, such restrictions can favor low speed, low traffic streets, more direct routes, and less steep paths, among others, which are suitable for cycling.
 
@@
 The intermodality scenario considers trips combining PT and cycling for the first and last legs.
-In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for instance: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike.
-Furthermore, we have imposed restrictions on PT usage to include only trips with no PT transfers, and up to 2 hours (120 min).
+In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike travel-time.
+Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
 Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
+These restrictions can be eased in the future when testing more developed policy interventions to enhance intermodality between cycling and PT, considering both the vehicle and infrastructure perspectives.
 
@@
 
-```{r map2, out.width="80%", fig.cap="Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).", fig.align='center'}
+```{r map2, out.width="80%", fig.cap="Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).", fig.align='center'}
 knitr::include_graphics("img/map2.png", error = FALSE)
@@
 
-For the *cycling legs of the journey* (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.2 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
+For the *cycling legs of the journey* (first and last legs), socio-environmental impacts were estimated, using the Health Economic Assessment Tool (HEAT) for Cycling v5.2 [@HEAT], from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
 The use of this package made it possible to run multiple scenarios with few changes in input values, making the interaction with HEAT more reliable when reproducing runs.    
@@
 Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles.
-For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in \ref{tab:summary1b}, resulting from replacing car trips with cycling.
+For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in Table \ref{tab:summary1b}, resulting from replacing car trips with cycling.
 
@@
 Train offers the greatest potential for substitution (88%).
-When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the Train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.
+When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the Train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to become safer.
 
@@
 
-The sum of CO~2~eq avoided emissions from the potential car trips shifted to bike (first-and-last legs) in combination with PT (second leg) in the LMA is presented on Table \ref{tab:summaryall}, for both national cycling strategy targets and routing profiles, and the socio-environmental benefits monetized in €, for a 1-year and 10-year time periods.
+The sum of CO~2~eq avoided emissions from the potential car trips shifted to bike (first-and-last legs) in combination with PT (second leg) in the LMA is presented in Table \ref{tab:summaryall}, for both national cycling strategy targets and routing profiles, and the socio-environmental benefits monetized in €, for a 1-year and 10-year time periods.
 
@@
 One of the main socio-environmetal benefits, valued after monetization, comes from the increase in physical activity [@Felix2023ES].
-Although are also social benefits form shifting car trips to PT, its health benefits would not be as high as shifting to cycling. 
+Although there are also social benefits form shifting car trips to PT, its health benefits would not be as high as shifting to cycling. 
 The literature shows that the Metabolic Equivalent Tasks (MET) for "riding in a bus or a train" is 1.3 plus the "walking for transportation" as 3.5, while "driving a car" is 2.5 [@MET2011].
-The difference between these activities is not very obvious when compared to cycling, whose equivalent MET is between 6.8 to 7.5.
+The difference between these activities - shifting from car to PT - is not very obvious when compared to shifting from car to cycling, whose MET is about 6.8.
 Nevertheless, future works should also encompass the estimation of the social impacts for the PT leg of the journey, shifting from car.
@@
 
-Our findings suggest that cycling and PT *in combination* could viably replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution. 
+Our findings suggest that cycling *in combination* with PT could potentially replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution. 
 <!-- repeated? -->
@@
 
-By making the research process publicly accessible in a code repository, it enables the replication of similar estimates for socio-environmental impacts, resulting from a modal shift from car to bicycle in combination with PT, in other metropolitan areas.
+By making the research process publicly accessible in a code repository (see [Research Data](#research-data)), it enables the replication of similar estimates for socio-environmental impacts, resulting from a modal shift from car to bicycle in combination with PT, in other metropolitan areas.
 
-### Acknowledgements. {.unnumbered}
+## Acknowledgements {.unnumbered}
 [*blind*]
@@
 
+## Research data {.unnumbered}
+The data and the code to produce XXX and this paper are available at: https://github.com/[*blind*].
+
 # References {.unnumbered}

### 2023-11-28 paper/PaperCEUS/PaperCEUS.tex
paper revs from FM
@filmoura

@@
             pdfauthor={},
-            pdftitle={biclaR: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools},
+            pdftitle={XXX: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools},
             colorlinks=true,
@@
 
-  \title{biclaR: Estimating the socio-environmental impacts of car
+  \title{XXX: Estimating the socio-environmental impacts of car
 substitution by bicycle and public transit using open tools}
@@
   development of a metropolitan cycling network was developed
-  (\emph{biclaR}). The social and environmental impacts were assessed
-  using the \emph{HEAT for Cycling} and the \emph{HEAT as a Service}
-  tools. The impacts of shifting car trips to PT were also estimated and
+  (\emph{XXX}). The social and environmental impacts were assessed using
+  the \emph{HEAT for Cycling} and the \emph{HEAT as a Service} tools.
+  The impacts of shifting car trips to PT were also estimated and
   monetized. The results indicate that 10\% of trips could be made by
@@
 Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
-Transport introduced \emph{biclaR}\footnote{See
-  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
-decision support tool that facilitates the planning, design, and
-development of a metropolitan cycling network (Félix et al., 2022).
+Transport introduced \emph{XXX}\footnote{See https://xxx.xxx. \emph{For
+  double blind peer-review process, the authors decided to truncate the
+  name and references to this tool}.}, a decision support tool that
+facilitates the planning, design, and development of a metropolitan
+cycling network (ref X).
 
-\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
+\emph{XXX} builds on the Propensity to Cycle Tool\footnote{See
   \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
@@
 significant portion of short-distance trips, a prevalent characteristic
-of active mode travel (Lovelace et al., 2022b). The OD Jittering method
+of active travel (Lovelace et al., 2022b). The OD Jittering method
 breaks down a single point (i.e., the centroid of an area) into multiple
@@
 
-\caption{Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
+\caption{Representation of desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
 \end{figure}
@@
 restrictions can favor low speed, low traffic streets, more direct
-routes, and less steep paths, among others, suitable for cycling.
+routes, and less steep paths, among others, which are suitable for
+cycling.
 
@@
 our analysis to the first and last legs with a combined length of up to
-5 km (for instance: 1 km from origin to interface \emph{A} plus 4 km
-from interface \emph{B} to destination) or up to 25 minutes on bike.
-Furthermore, we have imposed restrictions on PT usage to include only
-trips with no PT transfers, and up to 2 hours (120 min). Additionally,
-we have only included PT modes that can easily accommodate bicycles,
-such as trains, ferries, trams, and inter-municipal bus lines equipped
-with bike racks (Figure \ref{fig:map1}).
+5 km (for example: 1 km from origin to interface \emph{A} plus 4 km from
+interface \emph{B} to destination) or up to 25 minutes on bike
+travel-time. Furthermore, we have imposed restrictions on PT usage,
+limiting it to trips without PT transfers, and within a duration of up
+to 2 hours (120 minutes). Additionally, we have only included PT modes
+that can easily accommodate bicycles, such as trains, ferries, trams,
+and inter-municipal bus lines equipped with bike racks (Figure
+\ref{fig:map1}). These restrictions can be eased in the future when
+testing more developed policy interventions to enhance intermodality
+between cycling and PT, considering both the vehicle and infrastructure
+perspectives.
 
@@
 
-\caption{Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
+\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
 \end{figure}
@@
 For the \emph{cycling legs of the journey} (first and last legs),
-socio-environmental impacts were estimated, using the HEAT for Cycling
-tool v5.2 (Kahlmeier et al., 2017) from the World Health Organization,
-and the
+socio-environmental impacts were estimated, using the Health Economic
+Assessment Tool (HEAT) for Cycling v5.2 (Kahlmeier et al., 2017), from
+the World Health Organization, and the
 \href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
@@
 last legs), the environmental avoided emissions and monetized
-socio-environment (SE) benefits are presented in \ref{tab:summary1b},
-resulting from replacing car trips with cycling.
+socio-environment (SE) benefits are presented in Table
+\ref{tab:summary1b}, resulting from replacing car trips with cycling.
 
@@
 the ones that have the highest potential to attract car-to-PT
-substituting trips, if their accessibility by bicycle is improved to be
-safe.
+substituting trips, if their accessibility by bicycle is improved to
+become safer.
 
@@
 car trips shifted to bike (first-and-last legs) in combination with PT
-(second leg) in the LMA is presented on Table \ref{tab:summaryall}, for
+(second leg) in the LMA is presented in Table \ref{tab:summaryall}, for
 both national cycling strategy targets and routing profiles, and the
@@
 socio-environmetal benefits, valued after monetization, comes from the
-increase in physical activity (Félix et al., 2023). Although are also
-social benefits form shifting car trips to PT, its health benefits would
-not be as high as shifting to cycling. The literature shows that the
-Metabolic Equivalent Tasks (MET) for ``riding in a bus or a train'' is
-1.3 plus the ``walking for transportation'' as 3.5, while ``driving a
+increase in physical activity (Félix et al., 2023). Although there are
+also social benefits form shifting car trips to PT, its health benefits
+would not be as high as shifting to cycling. The literature shows that
+the Metabolic Equivalent Tasks (MET) for ``riding in a bus or a train''
+is 1.3 plus the ``walking for transportation'' as 3.5, while ``driving a
 car'' is 2.5 (Ainsworth et al., 2011). The difference between these
-activities is not very obvious when compared to cycling, whose
-equivalent MET is between 6.8 to 7.5. Nevertheless, future works should
-also encompass the estimation of the social impacts for the PT leg of
-the journey, shifting from car.
+activities - shifting from car to PT - is not very obvious when compared
+to shifting from car to cycling, whose MET is about 6.8. Nevertheless,
+future works should also encompass the estimation of the social impacts
+for the PT leg of the journey, shifting from car.
 
@@
 
-Our findings suggest that cycling and PT \emph{in combination} could
-viably replace 10\% of current LMA trips, with an additional 6\% of PT
-journeys prone to further substitution.
+Our findings suggest that cycling \emph{in combination} with PT could
+potentially replace 10\% of current LMA trips, with an additional 6\% of
+PT journeys prone to further substitution.
 
@@
 
-By making the research process publicly accessible in a code repository,
-it enables the replication of similar estimates for socio-environmental
-impacts, resulting from a modal shift from car to bicycle in combination
-with PT, in other metropolitan areas.
+By making the research process publicly accessible in a code repository
+(see \protect\hyperlink{research-data}{Research Data}), it enables the
+replication of similar estimates for socio-environmental impacts,
+resulting from a modal shift from car to bicycle in combination with PT,
+in other metropolitan areas.
 
-\hypertarget{acknowledgements.}{%
-\subsubsection*{Acknowledgements.}\label{acknowledgements.}}
-\addcontentsline{toc}{subsubsection}{Acknowledgements.}
+\hypertarget{acknowledgements}{%
+\subsection*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{subsection}{Acknowledgements}
 
 {[}\emph{blind}{]}
+
+\hypertarget{research-data}{%
+\subsection*{Research data}\label{research-data}}
+\addcontentsline{toc}{subsection}{Research data}
+
+The data and the code to produce XXX and this paper are available at:
+https://github.com/{[}\emph{blind}{]}.
 
@@
 \url{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
-Félix, R., Lovelace, R., \& Moura, F. (2022). \emph{{biclaR - Ferramenta
-de apoio ao planeamento da rede ciclável na área metropolitana de
-Lisboa}}. {CERIS - Instituto Superior Técnico and Transportes
-Metropolitanos de Lisboa}. \url{https://biclar.tmlmobilidade.pt}
 

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
title and plot aml trips

@@
 ---
-title: "XXX: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools"
+title: "Modelling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from the XXX project"
+# subtitle: "Reproducible methods, results and actionable evidence from the XXX project"
 author:
@@
 library(dplyr)
+library(ggplot2)
 ```
@@
 According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.
-Car modal share was 58.4%, while PT accounted for 15.5%.
+Car modal share was 58.4%, while PT accounted for 15.5% (see Figure \ref{fig:mododist}).
 The number of intra-municipal trips --- with origin and destination in the same municipality --- amounts to 3.5 million trips.
@@
 <!-- Plot of  AML trips by mode and inter/intra %? -->
+```{r dadosaml, message=FALSE, warning=FALSE}
+TRIPSmode_municipio = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/TRIPSmode_municipal.Rds"))
+
+TRIPSmode_aml = TRIPSmode_municipio %>%
+  group_by(modo) %>%
+  summarise(viagens = round(sum(viagens))) %>% 
+  mutate(percentagem = round(100*viagens/sum(viagens),1))
+
+TRIPSmode_aml_intr = TRIPSmode_municipio
+TRIPSmode_aml_intr$Inter[TRIPSmode_aml_intr$Origem == TRIPSmode_aml_intr$Destino] = "Intra-municipal"
+TRIPSmode_aml_intr$Inter[is.na(TRIPSmode_aml_intr$Inter)] = "Inter-municipal"
+TRIPSmode_aml_intr = TRIPSmode_aml_intr %>% 
+  group_by(modo, Inter) %>%
+  summarise(viagens = round(sum(viagens)))
+
+viagensinter = sum(TRIPSmode_aml_intr$viagens[TRIPSmode_aml_intr$Inter == "Inter-municipal"])
+viagensintra = sum(TRIPSmode_aml_intr$viagens[TRIPSmode_aml_intr$Inter == "Intra-municipal"])
+
+TRIPSmode_aml_intr$percentagem[TRIPSmode_aml_intr$Inter == "Inter-municipal"] =  round(100*TRIPSmode_aml_intr$viagens[TRIPSmode_aml_intr$Inter == "Inter-municipal"]/viagensinter, 1)
+
+TRIPSmode_aml_intr$percentagem[TRIPSmode_aml_intr$Inter == "Intra-municipal"] =  round(100*TRIPSmode_aml_intr$viagens[TRIPSmode_aml_intr$Inter == "Intra-municipal"]/viagensintra, 1)
+
+TRIPSmode_aml_intr = bind_rows(TRIPSmode_aml %>% mutate(Inter = "All"),  TRIPSmode_aml_intr)
+
+```
+```{r mododist, fig.ncol=2, out.width="50%", fig.cap="\\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.", fig.subcap=c("In percentage","In total")}
+#gráfico com bistribuição de viagens por modo na AML segundo IMob 2018
+
+TRIPSmode_aml_intr$modo = factor(TRIPSmode_aml_intr$modo, 
+                                 levels = c("Car", "Transit", "Motorcycle", 
+                                            "Other", "Walk", "Bike"))
+
+col_modes = c("#fe5f55", "#ffd166", "#8E794A", "grey", "#457b9d", "#90be6d")
+
+ggplot(TRIPSmode_aml_intr, aes(fill = modo, y = percentagem/100, x = Inter)) +
+  geom_bar(position = position_fill(reverse = FALSE), stat = "identity") +
+  scale_fill_manual(values = col_modes, name = "Mode") +
+  scale_y_continuous(labels = scales::percent) +
+  geom_text(aes(label = ifelse(percentagem > 6, paste0(percentagem,"%"), "")),
+    position = position_stack(vjust = 0.5, reverse = FALSE))+
+  theme_minimal() +
+  theme(axis.title.y = element_blank(),
+        axis.title.x = element_blank(),
+        legend.title = element_blank(),
+        legend.position = "none") +
+  coord_flip()
+
+ggplot(TRIPSmode_aml_intr, aes(fill = modo, y = viagens, x = Inter)) +
+  geom_bar(stat = "identity") +
+  scale_fill_manual(values = col_modes, name = "Mode") +
+  scale_y_continuous("Trips / day", labels = scales::comma_format(big.mark = ' ')) +
+  geom_text(aes(label = ifelse(viagens > 500000, paste0(round(viagens/1000000,1)," M"), "")),
+  position = position_stack(vjust = 0.5, reverse = FALSE))+
+  theme_minimal() +
+  theme(
+    axis.title.y = element_blank(),
+        legend.position = "right") +
+  coord_flip()
+
+```
 
@@
 
-This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
+This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA, while achieving the national cycling targets.
 After presenting the methods used, it assesses its socio-environmental impacts using open data and open-source tools.

### 2023-11-28 paper/PaperCEUS/PaperCEUS.tex
title and plot aml trips

@@
             pdfauthor={},
-            pdftitle={XXX: Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools},
+            pdftitle={Modelling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from the XXX project},
             colorlinks=true,
@@
 
-  \title{XXX: Estimating the socio-environmental impacts of car
-substitution by bicycle and public transit using open tools}
+  \title{Modelling the impacts of replacing car trips with combined
+public transport and cycling: Reproducible methods, results and
+actionable evidence from the XXX project}
     \author[CERIS]{Rosa Félix\corref{cor1}%
@@
 the LMA registered a total of 5.3 million daily trips, with only 0.5\%
-by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%.
-The number of intra-municipal trips --- with origin and destination in
-the same municipality --- amounts to 3.5 million trips. This exceeds the
-number of inter-municipal trips (1.8 million trips), involving travel
-between different municipalities. Cars and public transport are the most
-used modes for intercity trips, with cars being the predominant choice
-for all journeys.
+by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%
+(see Figure \ref{fig:mododist}). The number of intra-municipal trips ---
+with origin and destination in the same municipality --- amounts to 3.5
+million trips. This exceeds the number of inter-municipal trips (1.8
+million trips), involving travel between different municipalities. Cars
+and public transport are the most used modes for intercity trips, with
+cars being the predominant choice for all journeys.
+
+\begin{figure}
+\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
+\end{figure}
 
@@
 This paper estimates the potential for combining cycling and PT to
-substitute car trips in the LMA. After presenting the methods used, it
-assesses its socio-environmental impacts using open data and open-source
-tools.
+substitute car trips in the LMA, while achieving the national cycling
+targets. After presenting the methods used, it assesses its
+socio-environmental impacts using open data and open-source tools.
 

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
EF equations + conclusion

@@
 Cars and public transport are the most used modes for intercity trips, with cars being the predominant choice for all journeys.  
+`r round(100*2948314/5569731)`% trips are up to 5 km distance, and `r round(100*3958667/5569731)`% up to 10 km.
+Nevertheless, 29% of trips are longer than 10 km, which requires the use of motorized modes, or active modes in combination with public transport.
 
@@
 
-## Modeling Origin-Destination trips
+## Modelling Origin-Destination trips
 
@@
 
-## Modeling routes
+## Modelling routes
 
 The mobility survey collects the origin and destination of trips but does not include the respective routes.
-Modeling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
+Modelling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
 Other constraints regarding the behavior of potential cyclists determine the routing results.
@@
 
-## Modeling intermodality
+## Modelling intermodality
 
@@
 Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
-In addition, we assumed an occupancy rate of 1.6 passengers *per* car [@IMOB].
+An equation was then used to calculate emission factors for the two types of fuel, for each type of pollutant, whose explanatory variables are driving speed ($speed$, in km/h) and average trip distance ($l_{trip}$, in km/trip).
+Thus, the emission factors ($EF_{fuel,l_{trip},speed}$, in g/km) can be calculated using equation \ref{eq:fatoremissaoauto}.
+
+\begin{equation}\label{eq:fatoremissaoauto}
+{EF}_{fuel,l_{trip},speed} = a + b\cdot {speed} + c\cdot l_{trip}
+\end{equation}
+
+Emission factors are estimated for the following air pollutants: CO, NO~X~, VOC, and PM.
+Emission factors of the main greenhouse gases (GHG) are also estimated: CO~2~, CH~4~ and N~2~O, converted in CO~2~eq by the following relationship[^9]: $EF_{CO_2eq} = EF_{CO_2} + 28\cdot EF_{CH_4} + 265\cdot EF_{N_2O}$.
+The CH~4~ and N~2~O emission factors do not vary with travel speed.
+The PM10 emission factor does not vary with trip distance.
+
+The used values consider that 64% of the car fleet was diesel in 2022[^6]. In addition, we assumed an occupancy rate of 1.6 passengers *per* car [@IMOB].
+Finally, the final emissions for each trip ($E_{pollutant}$, in g/trip) are derived from the equation \ref{eq:emissaoauto}.
+
+[^6]: See [Statistics Portugal: Stock road vehicles statistic](https://smi.ine.pt/Indicador/Detalhes/10837?LANG=EN).
+[^9]: The weights correspond to the Global Warming Potentials (GWP) defined for a 100-year period by the IPCC in its [5th Assessment Report](https://www.ipcc.ch/report/ar5/).
+
+\begin{equation}\label{eq:emissaoauto}
+{E}_{pollutant} = {EF}_{fuel,l_{trip},speed}\cdot l_{trip}
+\end{equation}
 
 Regarding PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014].
-
-Emissions were estimated for the following atmospheric pollutants: CO, PM10, NOx, and VOC; and for the main green house gases: CO~2~, CH~4~, and N~2~O, converted in CO~2~eq.
 In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally.
 
-The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases [@EuropeanCommission2014; @bickel2006; @UNITE].
-The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal [@ANSR2021].
-We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a discount rate of 5% and inflation of 3%.
+The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases [@EuropeanCommission2014; @bickel2006; @UNITE]: 8.44 €/ton for CO, 2,867.85 €/ton for NO~X~, 340,969.27 €/ton for PM10, 7,169.62 €/ton for VOC and 35.85 €/ton for CO~2~eq.
+The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal: €3,055,358/fatality [@ANSR2021].
+We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^7], and our 10-years estimations assumed a discount rate of 5% and inflation of 3%.
 
-[^6]: See [Statistics Portugal tool for inflation rate estimates between years](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc).
+[^7]: See [Statistics Portugal tool for inflation rate estimates between years](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc).
 
-<!-- For the environmental impacts, using conversion values by the ton, suggested by Bickel et al.
-(2006), Nash et al.
-(2003), Sartori et al.
-(2014) : 8.44 €/ton for CO, 2,867.85 €/ton for NOX, 340,969.27 €/ton for PM10, 7,169.62 €/ton for VOC and 35.85 €/ton for CO2eq.
--->
+<!-- make tables of emission factors available in attachment? they are already in the github, if you look for them -->
 
@@
 Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips.  
-<!-- We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity.
-These estimations used an electricity production life cycle approach.
--->
-<!-- this table could be split into Social benefits and Environmental benefits -->
 Regarding the PT segment, the shift from private car would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly.
@@
 
+It has become progressively more common to establish strategic plans, at national, regional or municipal level, to mitigate climate change.
+Among these, the Sustainable Urban Mobility Plans (SUMP[^10]), promoted by the European Commission, are becoming popular in Europe, although authorities are designing documents of this sort all over the world.
+The definition of targets associated with a timeframe for reducing dependence on the individual motorized vehicle, or targets for the use of active modes such as walking and cycling, are too often not accompanied by estimates of their social, environmental and economic impacts.
+It is important for authorities and practitioners to know how to estimate those impacts, which tools are available to support them in the process, and what results to expect.
+
+This paper attempts to estimate the socio-environmental impacts associated with the modal shift from car to intermodal trips by bicycle in combination with public transport, in the Lisbon metropolitan area, in order to meet the national targets set for bicycle use of 4% and 10% by 2025 and 2030, respectively.
+
+[^10]: See [eltis.org/mobility-plans/sump-concept](https://www.eltis.org/mobility-plans/sump-concept)
+
 The information on socio-economic benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation, and to better communicate their decisions by providing the expected avoided GHG and air pollutant emissions, and the monetized socio-economic benefits for short and long terms.
 
-The information available at *biclaR* tool -- an open access website -- can be downloaded and used with any GIS software.
+<!-- Será que não se deve incluir uma explicação das sessões de formação com as câmaras e que estas ferramentas digitais são extremamente potenciadoras deste tipo de iniciativas e promovem o empowerment dos técnicos e das populações?
+-->
+
+The information available at *XXX* tool -- an open access website -- can be downloaded and used with any GIS software.
 This allows practitioners to, for example, gain insights into which potential cycling connections have the highest socio-environmental impacts, quantified in tons of avoided CO~2~eq emissions, or in long term social benefits.
@@
 
+We used open-source tools and transparent methods.
 By making the research process publicly accessible in a code repository (see [Research Data](#research-data)), it enables the replication of similar estimates for socio-environmental impacts, resulting from a modal shift from car to bicycle in combination with PT, in other metropolitan areas.

### 2023-11-28 paper/PaperCEUS/PaperCEUS.tex
EF equations + conclusion

@@
 and public transport are the most used modes for intercity trips, with
-cars being the predominant choice for all journeys.
+cars being the predominant choice for all journeys.\\
+53\% trips are up to 5 km distance, and 71\% up to 10 km. Nevertheless,
+29\% of trips are longer than 10 km, which requires the use of motorized
+modes, or active modes in combination with public transport.
 
@@
 
-\hypertarget{modeling-origin-destination-trips}{%
-\subsection{Modeling Origin-Destination
-trips}\label{modeling-origin-destination-trips}}
+\hypertarget{modelling-origin-destination-trips}{%
+\subsection{Modelling Origin-Destination
+trips}\label{modelling-origin-destination-trips}}
 
@@
 
-\hypertarget{modeling-routes}{%
-\subsection{Modeling routes}\label{modeling-routes}}
+\hypertarget{modelling-routes}{%
+\subsection{Modelling routes}\label{modelling-routes}}
 
 The mobility survey collects the origin and destination of trips but
-does not include the respective routes. Modeling the realistic cycling +
-PT routes between OD pairs depends on assumptions regarding the
+does not include the respective routes. Modelling the realistic cycling
++ PT routes between OD pairs depends on assumptions regarding the
 characteristics of the cycling and road networks and the location of
@@
 
-\hypertarget{modeling-intermodality}{%
-\subsection{Modeling intermodality}\label{modeling-intermodality}}
+\hypertarget{modelling-intermodality}{%
+\subsection{Modelling intermodality}\label{modelling-intermodality}}
 
@@
 estimated energy and emission factors for different ranges of trips at
-500-meter intervals. In addition, we assumed an occupancy rate of 1.6
-passengers \emph{per} car (INE, 2018).
+500-meter intervals.\\
+An equation was then used to calculate emission factors for the two
+types of fuel, for each type of pollutant, whose explanatory variables
+are driving speed (\(speed\), in km/h) and average trip distance
+(\(l_{trip}\), in km/trip). Thus, the emission factors
+(\(EF_{fuel,l_{trip},speed}\), in g/km) can be calculated using equation
+\ref{eq:fatoremissaoauto}.
+
+\begin{equation}\label{eq:fatoremissaoauto}
+{EF}_{fuel,l_{trip},speed} = a + b\cdot {speed} + c\cdot l_{trip}
+\end{equation}
+
+Emission factors are estimated for the following air pollutants: CO,
+NO\textsubscript{X}, VOC, and PM. Emission factors of the main
+greenhouse gases (GHG) are also estimated: CO\textsubscript{2},
+CH\textsubscript{4} and N\textsubscript{2}O, converted in
+CO\textsubscript{2}eq by the following relationship\footnote{The weights
+  correspond to the Global Warming Potentials (GWP) defined for a
+  100-year period by the IPCC in its
+  \href{https://www.ipcc.ch/report/ar5/}{5th Assessment Report}.}:
+\(EF_{CO_2eq} = EF_{CO_2} + 28\cdot EF_{CH_4} + 265\cdot EF_{N_2O}\).
+The CH\textsubscript{4} and N\textsubscript{2}O emission factors do not
+vary with travel speed. The PM10 emission factor does not vary with trip
+distance.
+
+The used values consider that 64\% of the car fleet was diesel in
+2022\footnote{See
+  \href{https://smi.ine.pt/Indicador/Detalhes/10837?LANG=EN}{Statistics
+  Portugal: Stock road vehicles statistic}.}. In addition, we assumed an
+occupancy rate of 1.6 passengers \emph{per} car (INE, 2018). Finally,
+the final emissions for each trip (\(E_{pollutant}\), in g/trip) are
+derived from the equation \ref{eq:emissaoauto}.
+
+\begin{equation}\label{eq:emissaoauto}
+{E}_{pollutant} = {EF}_{fuel,l_{trip},speed}\cdot l_{trip}
+\end{equation}
 
@@
 (Carris, 2020; CP, 2020; Metropolitano de Lisboa, 2020; Transtejo,
-2014).
-
-Emissions were estimated for the following atmospheric pollutants: CO,
-PM10, NOx, and VOC; and for the main green house gases:
-CO\textsubscript{2}, CH\textsubscript{4}, and N\textsubscript{2}O,
-converted in CO\textsubscript{2}eq. In particular, for the urban train
-and tram -- with 100\% electric traction -- only CO\textsubscript{2}eq
-emissions were considered (resulting from the production of electricity,
-considering a ``well-to-tank'' approach), since the other pollutants are
-not emitted locally.
+2014). In particular, for the urban train and tram -- with 100\%
+electric traction -- only CO\textsubscript{2}eq emissions were
+considered (resulting from the production of electricity, considering a
+``well-to-tank'' approach), since the other pollutants are not emitted
+locally.
 
@@
 for the various gases (Bickel et al., 2006; Nash et al., 2003; Sartori
-et al., 2014). The social impacts are in avoided premature mortality.
-This result is finally monetized using the \emph{Statistical Value of
-Life} for Portugal (Silva et al., 2021). We updated all the monetary
-reference values of the literature based on the annual inflation rate in
-Portugal for 2022\footnote{See
+et al., 2014): 8.44 €/ton for CO, 2,867.85 €/ton for
+NO\textsubscript{X}, 340,969.27 €/ton for PM10, 7,169.62 €/ton for VOC
+and 35.85 €/ton for CO\textsubscript{2}eq. The social impacts are in
+avoided premature mortality. This result is finally monetized using the
+\emph{Statistical Value of Life} for Portugal: €3,055,358/fatality
+(Silva et al., 2021). We updated all the monetary reference values of
+the literature based on the annual inflation rate in Portugal for
+2022\footnote{See
   \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
@@
 
+It has become progressively more common to establish strategic plans, at
+national, regional or municipal level, to mitigate climate change. Among
+these, the Sustainable Urban Mobility Plans (SUMP\footnote{See
+  \href{https://www.eltis.org/mobility-plans/sump-concept}{eltis.org/mobility-plans/sump-concept}}),
+promoted by the European Commission, are becoming popular in Europe,
+although authorities are designing documents of this sort all over the
+world. The definition of targets associated with a timeframe for
+reducing dependence on the individual motorized vehicle, or targets for
+the use of active modes such as walking and cycling, are too often not
+accompanied by estimates of their social, environmental and economic
+impacts. It is important for authorities and practitioners to know how
+to estimate those impacts, which tools are available to support them in
+the process, and what results to expect.
+
+This paper attempts to estimate the socio-environmental impacts
+associated with the modal shift from car to intermodal trips by bicycle
+in combination with public transport, in the Lisbon metropolitan area,
+in order to meet the national targets set for bicycle use of 4\% and
+10\% by 2025 and 2030, respectively.
+
 The information on socio-economic benefits can support policy-makers in
@@
 
-The information available at \emph{biclaR} tool -- an open access
-website -- can be downloaded and used with any GIS software. This allows
+The information available at \emph{XXX} tool -- an open access website
+-- can be downloaded and used with any GIS software. This allows
 practitioners to, for example, gain insights into which potential
@@
 
-By making the research process publicly accessible in a code repository
-(see \protect\hyperlink{research-data}{Research Data}), it enables the
+We used open-source tools and transparent methods. By making the
+research process publicly accessible in a code repository (see
+\protect\hyperlink{research-data}{Research Data}), it enables the
 replication of similar estimates for socio-environmental impacts,
@@
 Rodoviária (ANSR)}.
-\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O
-Impacto Economico e Social da Sinistralidade - PT.pdf}
+\url{http://www.ansr.pt/Documents/Impacto_Economico_Social_Sinistralidade_Rodoviaria.pdf}
 

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Modeling 1L

@@
 ---
-title: "Modelling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from the XXX project"
+title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from the XXX project"
 # subtitle: "Reproducible methods, results and actionable evidence from the XXX project"
@@
 
-[^1]: See https://xxx.xxx. *For double blind peer-review process, the authors decided to truncate the name and references to this tool*.
+[^1]: See xxx.xxx. *For double blind peer-review process, the authors decided to truncate the name and references to this tool*.
 <!-- [^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/). -->
@@
 
-## Modelling Origin-Destination trips
+## Modeling Origin-Destination trips
 
@@
 
-## Modelling routes
+## Modeling routes
 
 The mobility survey collects the origin and destination of trips but does not include the respective routes.
-Modelling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
+Modeling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
 Other constraints regarding the behavior of potential cyclists determine the routing results.
@@
 
-## Modelling intermodality
+## Modeling intermodality
 

### 2023-11-28 paper/PaperCEUS/PTcar_legs.R
start PT car legs fig

@@ -1,0 +1,31 @@
+# see heat-prep_car.R to see where these data comes from
+
+carlegs_2_or = OD_2legs_lts3 |> st_as_sf(sf_column_name = "origins")
+carlegs_2_de = OD_2legs_lts3 |> st_as_sf(sf_column_name = "destinations")
+
+# mapview::mapview(carlegs_2)
+# id = 5123
+
+carlegs_2_od = st_coordinates(OD_2legs_lts3 |> st_as_sf(sf_column_name = "origins")) |> as.data.frame() |> 
+  cbind(st_coordinates(OD_2legs_lts3 |> st_as_sf(sf_column_name = "destinations")) |> as.data.frame()) |> 
+  cbind(OD_2legs_lts3$id)
+
+carlegs_2_od = stplanr::od_coords2line(carlegs_2_od)
+
+mapview::mapview(carlegs_2_od)
+id_route = 45563
+
+
+###### LEG 2 in PT
+mapview::mapview(carlegs_2_od[669,])
+
+
+
+carlegs_13_od = OD_13legs_lts4
+
+###### LEG 2 in Bike
+mapview::mapview(carlegs_13_od[carlegs_13_od$id_route == 45563,])
+
+
+# Try to run the routing model WITH grometry to get car routing
+

### 2023-11-28 paper/PaperCEUS/PaperCEUS.tex
knit

@@
             pdfauthor={},
-            pdftitle={Modelling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from the XXX project},
+            pdftitle={Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from the XXX project},
             colorlinks=true,
@@
 
-  \title{Modelling the impacts of replacing car trips with combined
+  \title{Modeling the impacts of replacing car trips with combined
 public transport and cycling: Reproducible methods, results and
@@
 Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
-Transport introduced \emph{XXX}\footnote{See https://xxx.xxx. \emph{For
-  double blind peer-review process, the authors decided to truncate the
-  name and references to this tool}.}, a decision support tool that
+Transport introduced \emph{XXX}\footnote{See xxx.xxx. \emph{For double
+  blind peer-review process, the authors decided to truncate the name
+  and references to this tool}.}, a decision support tool that
 facilitates the planning, design, and development of a metropolitan
@@
 
-\hypertarget{modelling-origin-destination-trips}{%
-\subsection{Modelling Origin-Destination
-trips}\label{modelling-origin-destination-trips}}
+\hypertarget{modeling-origin-destination-trips}{%
+\subsection{Modeling Origin-Destination
+trips}\label{modeling-origin-destination-trips}}
 
@@
 
-\hypertarget{modelling-routes}{%
-\subsection{Modelling routes}\label{modelling-routes}}
+\hypertarget{modeling-routes}{%
+\subsection{Modeling routes}\label{modeling-routes}}
 
 The mobility survey collects the origin and destination of trips but
-does not include the respective routes. Modelling the realistic cycling
-+ PT routes between OD pairs depends on assumptions regarding the
+does not include the respective routes. Modeling the realistic cycling +
+PT routes between OD pairs depends on assumptions regarding the
 characteristics of the cycling and road networks and the location of
@@
 
-\hypertarget{modelling-intermodality}{%
-\subsection{Modelling intermodality}\label{modelling-intermodality}}
+\hypertarget{modeling-intermodality}{%
+\subsection{Modeling intermodality}\label{modeling-intermodality}}
 
@@
 10-years estimations assumed a discount rate of 5\% and inflation of
-3\%.
+3\%. See \protect\hyperlink{research-data}{Research Data} for all the
+input values we used.
 

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
start PT car legs fig
Remove xxx

@@
 ---
-title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from the XXX project"
-# subtitle: "Reproducible methods, results and actionable evidence from the XXX project"
+title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from the biclaR project"
 author:
@@
 This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
-  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_XXX_).
+  A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_).
 The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
@@
 
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced *XXX*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network (ref X).
+To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network (ref X).
 <!-- [@felix2023]. -->
@@
 
-*XXX* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
+*biclaR* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
 The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
@@
 We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^7], and our 10-years estimations assumed a discount rate of 5% and inflation of 3%.
+See [Research Data](#research-data) for all the input values we used.
 
@@
 
-The information available at *XXX* tool -- an open access website -- can be downloaded and used with any GIS software.
+The results can can be downloaded from the *biclaR* and used with any GIS software.
 This allows practitioners to, for example, gain insights into which potential cycling connections have the highest socio-environmental impacts, quantified in tons of avoided CO~2~eq emissions, or in long term social benefits.
@@
 ## Research data {.unnumbered}
-The data and the code to produce XXX and this paper are available at: https://github.com/[*blind*].
+The data and the code to reproduce the results will be made available upon publication.
 

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Add results from Rietveld

@@
   A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile.
-This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
+This paper estimates the potential for cycling combined with public transit as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
   A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_).
 The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
-The impacts of shifting car trips to PT were also estimated and monetized.
-  The results indicate that 10% of trips could be made by bicycle + PT combination.
-Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the PT leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
+The impacts of shifting car trips to public transport were also estimated and monetized.
+  The results show that 10% of trips could be made by cycling in combination with public transport.
+Shifting to cycling for the shorter first and last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the public transport leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
 The estimated socio-environmental benefits are of €125 million to €325 million over 10 years.
@@
 
-Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can significantly replace private car trips [@MARTENS2007326; @RIETVELD200071].
+Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326].
+In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system to reduce car dependency and associated externalities [@RIETVELD200071].
 This approach requires interventions and programs to make bicycling more appealing, and the resulting public investments can have significant social and environmental benefits.

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Add new references

@@
 Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326].
-In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system to reduce car dependency and associated externalities [@RIETVELD200071].
-This approach requires interventions and programs to make bicycling more appealing, and the resulting public investments can have significant social and environmental benefits.
+In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system [@RIETVELD200071].
+This approach to reducing car dependency and associated externalities requires interventions and programs to make bicycling more appealing [@lapaix_role_2021].
+The resulting public investments can have significant social and environmental benefits [@internationaltransportforum_integrating_2017].
+Despite the benefits of cycling-PT intermodality, the potential of this combination is often overlooked in transport planning [@lapaix_role_2021].
 <!-- Many studies refer to the potential of cycling and bike-sharing to replace car and PT trips, but not to the potential of serving as first and last miles. -->
 
+The potential of cycling as a complementary mode of PT is substantial worldwide, especially in cities with established public transport networks or substantial ambitions to develop them.
+In the Lisbon metropolitan area (LMA) the largest metropolitan area in Portugal, the modal share of cycling is low, but the potential for cycling as a complementary mode of PT is high.
 According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Update description of jittering

@@
 
-The mobility survey data [@IMOB] is the basis for this project and defines the baseline scenario.
-Despite being conducted in the pre-pandemic period (2017), this dataset represents the most comprehensive and up-to-date information on urban mobility in Portuguese metropolitan areas (Lisbon and Porto).
+The mobility survey data [@IMOB] is the basis of the baseline scenario and trip rates presented in this paper.
+Conducted in the pre-pandemic period (2017), this OD dataset represents the most comprehensive and up-to-date information on urban mobility in Portuguese metropolitan areas (Lisbon and Porto).
 
-We used a method for disaggregating the origins and destinations of trips between the centroids of two districts (same as "parish") to ensure that a district is not solely characterized by a single point of origin and destination for its trips.
-Aggregating all trips into centroids renders the exercise less realistic, as it excludes a significant portion of short-distance trips, a prevalent characteristic of active travel [@Lovelace2022Jittering].
-The OD Jittering method breaks down a single point (i.e., the centroid of an area) into multiple random points on the existing and neighboring road network, using OpenStreetMap as a reference.
-This method then distributes the volume of trips within the district among the randomly generated origin-destination pairs.
+We used 'jittering' to disaggregate the OD data, resulting in a wide spatial distribution trip origins and destinations [@Lovelace2022Jittering].
+The method works by sampling 'sub-points' (nodes on the transport network represented in OpenStreetMap in this case) and using these instead of a single point (typically the centroid) to represent trip origins and destinations for each zone.
+This method then distributes the trips to desire lines connecting the subpoints based on a 'disaggregation threshold' which determines the maximum number of trips that can be represented by a single desire line.
 
-Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we employed a maximum disaggregation level of 100 trips per O-D pair for this project.
+Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we disaggregated the OD data into desire lines reprenting a maximum of 100 trips each.
 Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Fix type: OD

@@
 
-Although this method provides a more realistic representation of the trips undertaken compared to the traditional approach, it does not fully align with the actual O-D pairs of trips, which remain unknown due to data privacy regulations.
+Although this method provides a more realistic representation of the trips undertaken compared to the traditional approach, it does not fully align with the actual OD pairs of trips, which remain unknown due to data privacy regulations.
 

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Tweak modeling prose

@@
 
-Although this method provides a more realistic representation of the trips undertaken compared to the traditional approach, it does not fully align with the actual OD pairs of trips, which remain unknown due to data privacy regulations.
+The jittering pre-processing stage generates a more realistic representation of the trips undertaken than the traditional centroid-based approach but does precisely capture the exactly spatial distribution of trips.
+Even where such datasets, they cannot be shared for research due to data privacy regulations.
 
@@
 
-The intermodality scenario considers trips combining PT and cycling for the first and last legs.
+The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.
 In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike travel-time.

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Add reference in support of multi-modality statement

@@
 The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.
-In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike travel-time.
+We restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike travel-time.
+This conservative approach was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
 Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Clarify results

@@
 
-Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips.  
-Regarding the PT segment, the shift from private car would lead to the mitigation of CO~2~ equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly.
+Table \ref{tab:summary22} presents emissions reductions and associated economic benefits associated with the second (PT) leg of trips.
+The shift from private car associated with thes PT segments would reduce CO~2~ equivalent emissions by 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly, for the 4% and 10% targets, respectively.
 

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Cycling-PT

@@
 The mobility survey collects the origin and destination of trips but does not include the respective routes.
-Modeling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
+Modeling the realistic cycling-PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
 Other constraints regarding the behavior of potential cyclists determine the routing results.
@@
 <!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? -->
-Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year. 
+Shifting from car to cycling in combination with PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year. 
 The 10-year socio-environmental benefits account for €125 million to €325 million, depending on the cycling targets.

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Beef-up results

@@
 
-Our findings suggest that cycling *in combination* with PT could potentially replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution. 
-<!-- repeated? -->
+Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.
 

### 2023-11-28 paper/PaperCEUS/PaperCEUS.Rmd
Refine conclusions

@@
 
+This paper estimated the potential for combining cycling and PT to substitute car trips in the LMA, while achieving the national cycling targets and supporting decarbonization goals.
 It has become progressively more common to establish strategic plans, at national, regional or municipal level, to mitigate climate change.
@@
 
-This paper attempts to estimate the socio-environmental impacts associated with the modal shift from car to intermodal trips by bicycle in combination with public transport, in the Lisbon metropolitan area, in order to meet the national targets set for bicycle use of 4% and 10% by 2025 and 2030, respectively.
+This paper quantifies the benefits of replacing car trips with cycling, in combination with public transport.
+The case study of the Lisbon metropolitan area demonstrates that cycling-PT integration can help meet the national targets set for bicycle use of 4% and 10% by 2025 and 2030, respectively.
 
@@
 
-The information on socio-economic benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation, and to better communicate their decisions by providing the expected avoided GHG and air pollutant emissions, and the monetized socio-economic benefits for short and long terms.
+The quantification of benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation.
+The presentation of the results in an open access web application will help  communicate decisions.
+Furthermore, the provision of datasets resulting from this project provides a foundation for further research and development of new tools and methods.
 
 <!-- Será que não se deve incluir uma explicação das sessões de formação com as câmaras e que estas ferramentas digitais são extremamente potenciadoras deste tipo de iniciativas e promovem o empowerment dos técnicos e das populações?
--->
-
-The results can can be downloaded from the *biclaR* and used with any GIS software.
-This allows practitioners to, for example, gain insights into which potential cycling connections have the highest socio-environmental impacts, quantified in tons of avoided CO~2~eq emissions, or in long term social benefits.
-<!-- move to introduction?
 -->

### 2023-11-29 paper/PaperCEUS/PaperCEUS.Rmd

@@
 Furthermore, the provision of datasets resulting from this project provides a foundation for further research and development of new tools and methods.
+<!-- Punchy final sentence: -->
+The methods are reproducible and based on open source software, which can be applied to other cities and metropolitan areas, supporting the decarbonization of transport systems internationally.
 
@@
 
-We used open-source tools and transparent methods.
+<!-- We used open-source tools and transparent methods.
 By making the research process publicly accessible in a code repository (see [Research Data](#research-data)), it enables the replication of similar estimates for socio-environmental impacts, resulting from a modal shift from car to bicycle in combination with PT, in other metropolitan areas.
+-->
 

### 2023-11-29 paper/PaperCEUS/PaperCEUS.Rmd
xxx and minor improvs
cover letter and title page

@@
 ---
-title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from the biclaR project"
+title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from biclaR"
 author:
@@
 
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network (ref X).
-<!-- [@felix2023]. -->
+To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].
 
-[^1]: See xxx.xxx. *For double blind peer-review process, the authors decided to truncate the name and references to this tool*.
-<!-- [^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/). -->
+<!-- [^1]: See xxx.xxx. *For double blind peer-review process, the authors decided to truncate the name and references to this tool*. -->
+[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/).
 
@@
 The jittering pre-processing stage generates a more realistic representation of the trips undertaken than the traditional centroid-based approach but does precisely capture the exactly spatial distribution of trips.
-Even where such datasets, they cannot be shared for research due to data privacy regulations.
+Even where such datasets exist, they cannot be shared for research due to data privacy regulations.
 
@@
 -->
-<!-- The authors thank Thomas Götschi (HEAT for Cycling) for providing access to HaaS tool, which is under development. -->
+<!-- The authors thank Thomas Götschi (HEAT for Cycling) for providing access to _HaaS_ tool, which is under development. -->
 

### 2023-11-29 paper/PaperCEUS/PaperCEUS.tex
xxx and minor improvs
cover letter and title page

@@
             pdfauthor={},
-            pdftitle={Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from the XXX project},
+            pdftitle={Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from biclaR},
             colorlinks=true,
@@
 public transport and cycling: Reproducible methods, results and
-actionable evidence from the XXX project}
+actionable evidence from biclaR}
     \author[CERIS]{Rosa Félix\corref{cor1}%
@@
   public transit and cycling for the first-and-last mile. This paper
-  estimates the potential for cycling combined with public transit (PT)
-  as a substitute for car trips in the Lisbon metropolitan area and
-  assesses its socio-environmental impacts using open data and open
-  source tools. A decision support tool that facilitates the design and
-  development of a metropolitan cycling network was developed
-  (\emph{XXX}). The social and environmental impacts were assessed using
-  the \emph{HEAT for Cycling} and the \emph{HEAT as a Service} tools.
-  The impacts of shifting car trips to PT were also estimated and
-  monetized. The results indicate that 10\% of trips could be made by
-  bicycle + PT combination. Shifting to cycling for the first-and-last
-  mile stages can reduce annual CO\textsubscript{2}eq emissions from
-  3,000 to 7,500 tons/day, while for the PT leg, the transfer from car
-  avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per
-  year. The estimated socio-environmental benefits are of €125 million
-  to €325 million over 10 years. This evidence can support policymakers
-  to prioritize interventions that reduce the reliance on private motor
-  vehicles.\\
+  estimates the potential for cycling combined with public transit as a
+  substitute for car trips in the Lisbon metropolitan area and assesses
+  its socio-environmental impacts using open data and open source tools.
+  A decision support tool that facilitates the design and development of
+  a metropolitan cycling network was developed (\emph{biclaR}). The
+  social and environmental impacts were assessed using the \emph{HEAT
+  for Cycling} and the \emph{HEAT as a Service} tools. The impacts of
+  shifting car trips to public transport were also estimated and
+  monetized. The results show that 10\% of trips could be made by
+  cycling in combination with public transport. Shifting to cycling for
+  the shorter first and last mile stages can reduce annual
+  CO\textsubscript{2}eq emissions from 3,000 to 7,500 tons/day, while
+  for the public transport leg, the transfer from car avoids of up to
+  20,500 tons of CO\textsubscript{2}eq emissions per year. The estimated
+  socio-environmental benefits are of €125 million to €325 million over
+  10 years. This evidence can support policymakers to prioritize
+  interventions that reduce the reliance on private motor vehicles.\\
   \end{abstract}
@@
 Combining public transportation (PT) and cycling for the first and last
-mile in metropolitan areas can significantly replace private car trips
-(Martens, 2007; Rietveld, 2000). This approach requires interventions
-and programs to make bicycling more appealing, and the resulting public
-investments can have significant social and environmental benefits.
+mile in metropolitan areas can replace a high proportion of private car
+trips (Martens, 2007). In The Netherlands, which has the highest mode
+share of cycling in the world, cycling accounts for more than a third of
+all trips to and from rail stations at the `home' end of the journey,
+greatly increasing the ability of the transport system (Rietveld, 2000).
+This approach to reducing car dependency and associated externalities
+requires interventions and programs to make bicycling more appealing (La
+Paix et al., 2021). The resulting public investments can have
+significant social and environmental benefits (International Transport
+Forum, 2017). Despite the benefits of cycling-PT intermodality, the
+potential of this combination is often overlooked in transport planning
+(La Paix et al., 2021).
 
-According to the latest mobility survey conducted in 2018 (INE, 2018),
-the LMA registered a total of 5.3 million daily trips, with only 0.5\%
-by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%
-(see Figure \ref{fig:mododist}). The number of intra-municipal trips ---
-with origin and destination in the same municipality --- amounts to 3.5
-million trips. This exceeds the number of inter-municipal trips (1.8
-million trips), involving travel between different municipalities. Cars
-and public transport are the most used modes for intercity trips, with
-cars being the predominant choice for all journeys.\\
+The potential of cycling as a complementary mode of PT is substantial
+worldwide, especially in cities with established public transport
+networks or substantial ambitions to develop them. In the Lisbon
+metropolitan area (LMA) the largest metropolitan area in Portugal, the
+modal share of cycling is low, but the potential for cycling as a
+complementary mode of PT is high. According to the latest mobility
+survey conducted in 2018 (INE, 2018), the LMA registered a total of 5.3
+million daily trips, with only 0.5\% by bicycle. Car modal share was
+58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
+The number of intra-municipal trips --- with origin and destination in
+the same municipality --- amounts to 3.5 million trips. This exceeds the
+number of inter-municipal trips (1.8 million trips), involving travel
+between different municipalities. Cars and public transport are the most
+used modes for intercity trips, with cars being the predominant choice
+for all journeys.\\
 53\% trips are up to 5 km distance, and 71\% up to 10 km. Nevertheless,
@@
 Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
-Transport introduced \emph{XXX}\footnote{See xxx.xxx. \emph{For double
-  blind peer-review process, the authors decided to truncate the name
-  and references to this tool}.}, a decision support tool that
-facilitates the planning, design, and development of a metropolitan
-cycling network (ref X).
+Transport commissioned \emph{biclaR}\footnote{See
+  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
+decision support tool that facilitates the planning, design, and
+development of a metropolitan cycling network (Félix et al., 2022).
 
-\emph{XXX} builds on the Propensity to Cycle Tool\footnote{See
+\emph{biclaR} builds on the Propensity to Cycle Tool\footnote{See
   \href{https://www.pct.bike/}{pct.bike}.} (PCT), a web application and
@@
 
-The mobility survey data (INE, 2018) is the basis for this project and
-defines the baseline scenario. Despite being conducted in the
-pre-pandemic period (2017), this dataset represents the most
+The mobility survey data (INE, 2018) is the basis of the baseline
+scenario and trip rates presented in this paper. Conducted in the
+pre-pandemic period (2017), this OD dataset represents the most
 comprehensive and up-to-date information on urban mobility in Portuguese
@@
 
-We used a method for disaggregating the origins and destinations of
-trips between the centroids of two districts (same as ``parish'') to
-ensure that a district is not solely characterized by a single point of
-origin and destination for its trips. Aggregating all trips into
-centroids renders the exercise less realistic, as it excludes a
-significant portion of short-distance trips, a prevalent characteristic
-of active travel (Lovelace et al., 2022b). The OD Jittering method
-breaks down a single point (i.e., the centroid of an area) into multiple
-random points on the existing and neighboring road network, using
-OpenStreetMap as a reference. This method then distributes the volume of
-trips within the district among the randomly generated
-origin-destination pairs.
+We used `jittering' to disaggregate the OD data, resulting in a wide
+spatial distribution trip origins and destinations (Lovelace et al.,
+2022b). The method works by sampling `sub-points' (nodes on the
+transport network represented in OpenStreetMap in this case) and using
+these instead of a single point (typically the centroid) to represent
+trip origins and destinations for each zone. This method then
+distributes the trips to desire lines connecting the subpoints based on
+a `disaggregation threshold' which determines the maximum number of
+trips that can be represented by a single desire line.
 
@@
 \href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
-package}, we employed a maximum disaggregation level of 100 trips per
-O-D pair for this project. Figure \ref{fig:jitter} illustrates the
+package}, we disaggregated the OD data into desire lines reprenting a
+maximum of 100 trips each. Figure \ref{fig:jitter} illustrates the
 contrast between trip representation through the traditional method,
@@
 
-Although this method provides a more realistic representation of the
-trips undertaken compared to the traditional approach, it does not fully
-align with the actual O-D pairs of trips, which remain unknown due to
-data privacy regulations.
+The jittering pre-processing stage generates a more realistic
+representation of the trips undertaken than the traditional
+centroid-based approach but does precisely capture the exactly spatial
+distribution of trips. Even where such datasets exist, they cannot be
+shared for research due to data privacy regulations.
 
@@
 The mobility survey collects the origin and destination of trips but
-does not include the respective routes. Modeling the realistic cycling +
-PT routes between OD pairs depends on assumptions regarding the
+does not include the respective routes. Modeling the realistic
+cycling-PT routes between OD pairs depends on assumptions regarding the
 characteristics of the cycling and road networks and the location of
@@
 
-The intermodality scenario considers trips combining PT and cycling for
-the first and last legs. In a conservative approach, we have restricted
-our analysis to the first and last legs with a combined length of up to
-5 km (for example: 1 km from origin to interface \emph{A} plus 4 km from
-interface \emph{B} to destination) or up to 25 minutes on bike
-travel-time. Furthermore, we have imposed restrictions on PT usage,
-limiting it to trips without PT transfers, and within a duration of up
-to 2 hours (120 minutes). Additionally, we have only included PT modes
-that can easily accommodate bicycles, such as trains, ferries, trams,
-and inter-municipal bus lines equipped with bike racks (Figure
-\ref{fig:map1}). These restrictions can be eased in the future when
-testing more developed policy interventions to enhance intermodality
-between cycling and PT, considering both the vehicle and infrastructure
-perspectives.
+The intermodality scenario considers trips made by PT in which cycling
+is used for the first and last legs. We restricted our analysis to the
+first and last legs with a combined length of up to 5 km (for example: 1
+km from origin to interface \emph{A} plus 4 km from interface \emph{B}
+to destination) or up to 25 minutes on bike travel-time. This
+conservative approach was adopted to capture the fact that cycling
+stages as part of a multi-modal trip are likely to be shorter than
+cycling-only trips (van Mil et al., 2021). Furthermore, we have imposed
+restrictions on PT usage, limiting it to trips without PT transfers, and
+within a duration of up to 2 hours (120 minutes). Additionally, we have
+only included PT modes that can easily accommodate bicycles, such as
+trains, ferries, trams, and inter-municipal bus lines equipped with bike
+racks (Figure \ref{fig:map1}). These restrictions can be eased in the
+future when testing more developed policy interventions to enhance
+intermodality between cycling and PT, considering both the vehicle and
+infrastructure perspectives.
 
@@
 
-Table \ref{tab:summary22} presents the results of the avoided emissions
-and its monetization for the second leg of the journey, by replacing car
-trips with potential TP trips.\\
-Regarding the PT segment, the shift from private car would lead to the
-mitigation of CO\textsubscript{2} equivalent emissions to 8,500 to
-20,800 tons annually, valued in €1.4 million to €3.5 million yearly.
+Table \ref{tab:summary22} presents emissions reductions and associated
+economic benefits associated with the second (PT) leg of trips. The
+shift from private car associated with thes PT segments would reduce
+CO\textsubscript{2} equivalent emissions by 8,500 to 20,800 tons
+annually, valued in €1.4 million to €3.5 million yearly, for the 4\% and
+10\% targets, respectively.
 
@@
 
-Shifting from car to cycling + PT can reduce annual
+Shifting from car to cycling in combination with PT can reduce annual
 CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year. The
@@
 
-Our findings suggest that cycling \emph{in combination} with PT could
-potentially replace 10\% of current LMA trips, with an additional 6\% of
-PT journeys prone to further substitution.
+Our findings show that cycling \emph{in combination} with PT could
+replace 10\% of current LMA trips, with an additional 6\% of PT journeys
+prone to further substitution, based on conservative assumptions.
 
@@
 
-It has become progressively more common to establish strategic plans, at
-national, regional or municipal level, to mitigate climate change. Among
-these, the Sustainable Urban Mobility Plans (SUMP\footnote{See
+This paper estimated the potential for combining cycling and PT to
+substitute car trips in the LMA, while achieving the national cycling
+targets and supporting decarbonization goals. It has become
+progressively more common to establish strategic plans, at national,
+regional or municipal level, to mitigate climate change. Among these,
+the Sustainable Urban Mobility Plans (SUMP\footnote{See
   \href{https://www.eltis.org/mobility-plans/sump-concept}{eltis.org/mobility-plans/sump-concept}}),
@@
 
-This paper attempts to estimate the socio-environmental impacts
-associated with the modal shift from car to intermodal trips by bicycle
-in combination with public transport, in the Lisbon metropolitan area,
-in order to meet the national targets set for bicycle use of 4\% and
-10\% by 2025 and 2030, respectively.
+This paper quantifies the benefits of replacing car trips with cycling,
+in combination with public transport. The case study of the Lisbon
+metropolitan area demonstrates that cycling-PT integration can help meet
+the national targets set for bicycle use of 4\% and 10\% by 2025 and
+2030, respectively.
 
-The information on socio-economic benefits can support policy-makers in
-prioritizing interventions to reduce the reliance on individual
-motorized transportation, and to better communicate their decisions by
-providing the expected avoided GHG and air pollutant emissions, and the
-monetized socio-economic benefits for short and long terms.
-
-The information available at \emph{XXX} tool -- an open access website
--- can be downloaded and used with any GIS software. This allows
-practitioners to, for example, gain insights into which potential
-cycling connections have the highest socio-environmental impacts,
-quantified in tons of avoided CO\textsubscript{2}eq emissions, or in
-long term social benefits.
-
-We used open-source tools and transparent methods. By making the
-research process publicly accessible in a code repository (see
-\protect\hyperlink{research-data}{Research Data}), it enables the
-replication of similar estimates for socio-environmental impacts,
-resulting from a modal shift from car to bicycle in combination with PT,
-in other metropolitan areas.
+The quantification of benefits can support policy-makers in prioritizing
+interventions to reduce the reliance on individual motorized
+transportation. The presentation of the results in an open access web
+application will help communicate decisions. Furthermore, the provision
+of datasets resulting from this project provides a foundation for
+further research and development of new tools and methods. The methods
+are reproducible and based on open source software, which can be applied
+to other cities and metropolitan areas, supporting the decarbonization
+of transport systems internationally.
 
@@
 
-The data and the code to produce XXX and this paper are available at:
-https://github.com/{[}\emph{blind}{]}.
+The data and the code to reproduce the results will be made available
+upon publication.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
+Félix, R., Lovelace, R., \& Moura, F. (2022). \emph{{biclaR - Ferramenta
+de apoio ao planeamento da rede ciclável na área metropolitana de
+Lisboa}}. {CERIS - Instituto Superior Técnico and Transportes
+Metropolitanos de Lisboa}. \url{https://biclar.tmlmobilidade.pt}
+
 \leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-internationaltransportforum_integrating_2017}{}}%
+International Transport Forum. (2017). \emph{Integrating {Urban Public
+Transport Systems} and {Cycling}}. {Organisation for Economic
+Co-operation and Development}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
@@
 \url{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lapaix_role_2021}{}}%
+La Paix, L., Cherchi, E., \& Geurs, K. (2021). Role of perception of
+bicycle infrastructure on the choice of the bicycle as a train feeder
+mode. \emph{International Journal of Sustainable Transportation},
+\emph{15}(6), 486--499.
+\url{https://doi.org/10.1080/15568318.2020.1765223}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-vanmil_insights_2021}{}}%
+van Mil, J. F. P., Leferink, T. S., Annema, J. A., \& van Oort, N.
+(2021). Insights into factors affecting the combined bicycle-transit
+mode. \emph{Public Transport}, \emph{13}(3), 649--673.
+\url{https://doi.org/10.1007/s12469-020-00240-2}
+
 \end{CSLReferences}

### 2023-11-30 paper/PaperCEUS/PaperCEUS.Rmd
last revision

@@
 The impacts of shifting car trips to public transport were also estimated and monetized.
-  The results show that 10% of trips could be made by cycling in combination with public transport.
+  The results show that 10% of all trips could be made by cycling in combination with public transport.
 Shifting to cycling for the shorter first and last mile stages can reduce annual CO~2~eq emissions from 3,000 to 7,500 tons/day, while for the public transport leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
@@
 
-We used 'jittering' to disaggregate the OD data, resulting in a wide spatial distribution trip origins and destinations [@Lovelace2022Jittering].
+We used 'jittering' to disaggregate the OD data, resulting in a wide spatial distribution of trip origins and destinations [@Lovelace2022Jittering].
 The method works by sampling 'sub-points' (nodes on the transport network represented in OpenStreetMap in this case) and using these instead of a single point (typically the centroid) to represent trip origins and destinations for each zone.
@@
 
-The jittering pre-processing stage generates a more realistic representation of the trips undertaken than the traditional centroid-based approach but does precisely capture the exactly spatial distribution of trips.
+The jittering pre-processing stage generates a more realistic representation of the trips undertaken than the traditional centroid-based approach but does not precisely capture the exact spatial distribution of trips.
 Even where such datasets exist, they cannot be shared for research due to data privacy regulations.
@@
 <!-- The impacts were assessed for the total metro area. -->
+
+<!-- Include a map with the legs of journeys, and which modes are replacing which, and what impacts are bein estimated at each one. See PTcar_legs.R -->
 
@@
 
-For both *direct* and *safe* route profiles, 10% of the daily trips have the potential to be made by a combination of PT and cycling (up to 5 km on bike).
+For both *direct* and *safe* route profiles, 10% of the daily trips have the potential to be made by a combination of PT and cycling, even given the travel restrictions considered (up to 5 km on bike, up to 2 hours, no possible transfers between PT).
 This unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).
-<!-- place in the end - here is not clear -->
 
@@
 
-The quantification of benefits can support policy-makers in prioritizing interventions to reduce the reliance on individual motorized transportation.
-The presentation of the results in an open access web application will help  communicate decisions.
+The quantification of benefits can support policy-makers in prioritizing interventions to reduce the reliance on private motorized modes of transportation.
+The presentation of the results in an open access web application will help to inform and explain decisions.
 Furthermore, the provision of datasets resulting from this project provides a foundation for further research and development of new tools and methods.

### 2023-11-30 paper/PaperCEUS/PaperCEUS.tex
last revision

@@
   shifting car trips to public transport were also estimated and
-  monetized. The results show that 10\% of trips could be made by
+  monetized. The results show that 10\% of all trips could be made by
   cycling in combination with public transport. Shifting to cycling for
@@
 We used `jittering' to disaggregate the OD data, resulting in a wide
-spatial distribution trip origins and destinations (Lovelace et al.,
+spatial distribution of trip origins and destinations (Lovelace et al.,
 2022b). The method works by sampling `sub-points' (nodes on the
@@
 representation of the trips undertaken than the traditional
-centroid-based approach but does precisely capture the exactly spatial
+centroid-based approach but does not precisely capture the exact spatial
 distribution of trips. Even where such datasets exist, they cannot be
@@
 For both \emph{direct} and \emph{safe} route profiles, 10\% of the daily
-trips have the potential to be made by a combination of PT and cycling
-(up to 5 km on bike). This unveils the potential of cycling as a
-complementary mode of PT, with the potential to uptake the number of PT
-trips within the LMA area by as much as 6.3\% (in addition to the 825
-thousand PT trips reported in the mobility survey).
+trips have the potential to be made by a combination of PT and cycling,
+even given the travel restrictions considered (up to 5 km on bike, up to
+2 hours, no possible transfers between PT). This unveils the potential
+of cycling as a complementary mode of PT, with the potential to uptake
+the number of PT trips within the LMA area by as much as 6.3\% (in
+addition to the 825 thousand PT trips reported in the mobility survey).
 
@@
 The quantification of benefits can support policy-makers in prioritizing
-interventions to reduce the reliance on individual motorized
+interventions to reduce the reliance on private motorized modes of
 transportation. The presentation of the results in an open access web
-application will help communicate decisions. Furthermore, the provision
-of datasets resulting from this project provides a foundation for
-further research and development of new tools and methods. The methods
-are reproducible and based on open source software, which can be applied
-to other cities and metropolitan areas, supporting the decarbonization
-of transport systems internationally.
+application will help to inform and explain decisions. Furthermore, the
+provision of datasets resulting from this project provides a foundation
+for further research and development of new tools and methods. The
+methods are reproducible and based on open source software, which can be
+applied to other cities and metropolitan areas, supporting the
+decarbonization of transport systems internationally.
 

### 2023-11-30 paper/PaperCEUS/PTcar_legs.R

@@ -21,11 +21,19 @@ mapview::mapview(carlegs_2_od[669,])
 
 
 
-carlegs_13_od = OD_13legs_lts4
+carlegs_13_od = OD_13legs_lts3
 
 ###### LEG 2 in Bike
 mapview::mapview(carlegs_13_od[carlegs_13_od$id_route == 45563,])
 
 
+
+
+
 # Try to run the routing model WITH grometry to get car routing
+
+mapview::mapview(carlegs_13_od[carlegs_13_od$id_route == 45563,], color = "red") + mapview::mapview(carlegs_2_od[669,])
+
+
+
 

### 2023-11-30 code/test-code/r5r_reprex.R
Create r5r_reprex.R

@@ -1,0 +1,10 @@
+r5r::download_r5(force_update = TRUE)
+r5r::r5r_sitrep()
+tools::R_user_dir("r5r", which = "cache")
+
+# sudo Rscript -e "r5r::download_r5(force_update = TRUE)"
+
+options(java.parameters = '-Xmx32G') 
+library(r5r)
+data_path <- system.file("extdata/poa", package = "r5r")
+r5r_core <- r5r::setup_r5(data_path = data_path)

### 2023-11-30 paper/PaperCEUS/PaperCEUS.tex
final paper to submit

@@
 Emission factors are estimated for the following air pollutants: CO,
-NO\textsubscript{X}, VOC, and PM. Emission factors of the main
-greenhouse gases (GHG) are also estimated: CO\textsubscript{2},
+NO\textsubscript{X}, VOC, and PM\textsubscript{10}. Emission factors of
+the main greenhouse gases (GHG) are also estimated: CO\textsubscript{2},
 CH\textsubscript{4} and N\textsubscript{2}O, converted in
@@
 The CH\textsubscript{4} and N\textsubscript{2}O emission factors do not
-vary with travel speed. The PM10 emission factor does not vary with trip
-distance.
+vary with travel speed. The PM\textsubscript{10} emission factor does
+not vary with trip distance.
 
@@
 et al., 2014): 8.44 €/ton for CO, 2,867.85 €/ton for
-NO\textsubscript{X}, 340,969.27 €/ton for PM10, 7,169.62 €/ton for VOC
-and 35.85 €/ton for CO\textsubscript{2}eq. The social impacts are in
-avoided premature mortality. This result is finally monetized using the
-\emph{Statistical Value of Life} for Portugal: €3,055,358/fatality
-(Silva et al., 2021). We updated all the monetary reference values of
-the literature based on the annual inflation rate in Portugal for
-2022\footnote{See
+NO\textsubscript{X}, 340,969.27 €/ton for PM\textsubscript{10}, 7,169.62
+€/ton for VOC and 35.85 €/ton for CO\textsubscript{2}eq. The social
+impacts are in avoided premature mortality. This result is finally
+monetized using the \emph{Statistical Value of Life} for Portugal:
+€3,055,358/fatality (Silva et al., 2021). We updated all the monetary
+reference values of the literature based on the annual inflation rate in
+Portugal for 2022\footnote{See
   \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
@@
 Shifting from car to cycling in combination with PT can reduce annual
-CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year. The
-10-year socio-environmental benefits account for €125 million to €325
-million, depending on the cycling targets.
+CO\textsubscript{2}eq emissions by 11,500 to 28,500 tons per year. These
+figures represent a 2.7\% reduction in Lisbon's transport emissions
+(Lisboa E-Nova, 2023), a small but important component of wider
+transport decarbonization measures. The 10-year socio-environmental
+benefits account for €125 million to €325 million, depending on the
+cycling targets.
 
@@
 \hypertarget{acknowledgements}{%
-\subsection*{Acknowledgements}\label{acknowledgements}}
-\addcontentsline{toc}{subsection}{Acknowledgements}
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
 
@@
 \hypertarget{research-data}{%
-\subsection*{Research data}\label{research-data}}
-\addcontentsline{toc}{subsection}{Research data}
+\section*{Research data}\label{research-data}}
+\addcontentsline{toc}{section}{Research data}
 
@@
 \url{https://doi.org/10.1080/15568318.2020.1765223}
+
+\leavevmode\vadjust pre{\hypertarget{ref-LisboaENova}{}}%
+Lisboa E-Nova. (2023). \emph{{Observatórios Lisboa - Infos Emissões
+Gases com Efeito Estufa}}. {Agência de Energia e Ambiente de Lisboa}.
+\url{https://observatorios-lisboa.pt/info_emissoes.html}
 

### 2023-12-21 paper/PaperTRA/PaperTRA_rev1.Rmd
setup PaperTRA_rev1 redux 6p
as requested by reviewers
reduce do 6+1

--- 2023-10-11 paper/PaperTRA/PaperTRA.Rmd
@@
 abstract: | 
-  A high proportion of car trips can be replaced by a combination of public transit and cycling for the first-and-last mile.
  This paper estimates the potential for cycling combined with public transit (PT) as a substitute for car trips in the Lisbon metropolitan area and assesses its socio-environmental impacts using open data and open source tools.
@@
   The results indicate that 10% of trips could be made by bicycle + PT combination.
-Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 6,000 tons/day, with benefits over 10 years of at least €230 million.
+Shifting to cycling for the first-and-last mile stages can reduce annual CO~2~eq emissions from 3,000 tons/day, with benefits over 10 years of at least €125 million.
 For the PT leg, the transfer from car avoids of up to 20,500 tons of CO~2~eq emissions per year.
@@
   - Environmental impacts
-  - Open data and methods
+  - Open methods
 bibliography: bibliography.bib
@@
 <!-- **full paper**: 4-6 pages in length (typically up to 3,000 words). we have 2849 words before references, or 2278 words without tables -->
+<!-- REDUCE TO MAX 6pages -->
 
@@
 
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport introduced *biclaR*[^1], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
+To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively), the Lisbon's Metropolitan Department of Transport introduced *biclaR*[^1], a decision support tool that facilitates the design and development of a metropolitan cycling network [@felix2023].
+<!-- [@ENMAC] -->
 
@@
 The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
-The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^3]
+<!-- The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^3] -->
 However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.
@@
 
-[^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.
+<!-- [^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes. -->
 
@@
 We used a method for disaggregating the origins and destinations of trips between the centroids of two districts (same as "parish") to ensure that a district is not solely characterized by a single point of origin and destination for its trips.
-Aggregating all trips into centroids renders the exercise less realistic, as it excludes a significant portion of short-distance trips, a prevalent characteristic of active mode travel [@Lovelace2022Jittering].
-The OD Jittering method breaks down a single point (i.e., the centroid of an area) into multiple random points on the existing and neighboring road network, using OpenStreetMap as a reference.
-This method then distributes the volume of trips within the district among the randomly generated origin-destination pairs.
+<!-- Aggregating all trips into centroids renders the exercise less realistic, as it excludes a significant portion of short-distance trips, a prevalent characteristic of active mode travel. -->
+The OD Jittering method breaks down a single point (i.e., the centroid of an area) into multiple random points on the existing and neighboring road network, using OpenStreetMap as a reference [@Lovelace2022Jittering].
+<!-- This method then distributes the volume of trips within the district among the randomly generated origin-destination pairs. -->
+Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we employed a maximum disaggregation level of 100 trips per O-D pair for this project.
+<!-- Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.
+-->
+Although this method provides a more realistic representation of the trips undertaken compared to the traditional approach, it does not fully align with the actual O-D pairs of trips, which remain unknown due to data privacy regulations.
 
-Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we employed a maximum disaggregation level of 100 trips per O-D pair for this project.
-Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.
-
-```{r jitter, fig.align='center', fig.cap="Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
+```{r jitter, fig.align='center', fig.cap="Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%", eval=FALSE}
 
@@
 
-Although this method provides a more realistic representation of the trips undertaken compared to the traditional approach, it does not fully align with the actual O-D pairs of trips, which remain unknown due to data privacy regulations.
-
 ## Modeling routes
@@
 Modeling the realistic cycling + PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
-Other constraints regarding the behavior of potential cyclists determine the routing results.
-For example, such restrictions can favor low speed, low traffic streets, more direct routes, and less steep paths, among others, suitable for cycling.
+<!-- Other constraints regarding the behavior of potential cyclists determine the routing results. -->
+<!-- For example, such restrictions can favor low speed, low traffic streets, more direct routes, and less steep paths, among others, suitable for cycling. -->
 
@@
 
-The selected route choice algorithm was the [`r5r` R package](https://ipeagit.github.io/r5r/) [@r5r], which allows for great flexibility in configuring estimated route types, and which proven to provide most accurate route networks for the city of Lisbon [@Lovelace2022exploring].
+The selected route choice algorithm was the [`r5r` R package](https://ipeagit.github.io/r5r/) [@r5r], which allows for great flexibility in configuring estimated route types.
+<!-- and which proven to provide most accurate route networks for the city of Lisbon [@Lovelace2022exploring]. -->
 `r5r` can calculate multi-modal routes using PT combined with other modes.
-It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^4] (LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic).
+It enables the identification of the most direct or safest cycling routes, using the Level of Traffic Stress[^4] (LTS) scale.
+<!-- ranging from 1 to 4, where 1 corresponds to the quietest (e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g., routes shared with motorized traffic). -->
 The routes were estimated for the base scenario for both types of networks: *direct* and *safe*, using LTS 4 and LTS 3, respectively.
-Different routing profiles enable decision-makers to plan for different bicycle user typologies and/or for different city cycling maturity levels [@felix2017]. <!-- move this to conclusions? -->
+<!-- Different routing profiles enable decision-makers to plan for different bicycle user typologies and/or for different city cycling maturity levels [@felix2017]. <!-- move this to conclusions? -->
 
@@
 This information is crucial for an accurate PT trip and route estimation.
-A digital elevation model, from the European Space Agency's COPERNICUS mission, was used to include street gradient information, as a weight in cycling routing.
+<!-- A digital elevation model, from the European Space Agency's COPERNICUS mission, was used to include street gradient information, as a weight in cycling routing.
 <!--  with a 25 m spatial resolution, -->
-The cycling potential trips for the two national strategic targets (4% and 10%) were estimated from the 2017 cycling and car trips (both as a driver and as a passenger), the baseline scenario.
+<!-- The cycling potential trips for the two national strategic targets (4% and 10%) were estimated from the 2017 cycling and car trips (both as a driver and as a passenger), the baseline scenario.
+-->
 
-The routes were then overlaid and aggregated by segments, using [`stplanr overline()` R function](https://docs.ropensci.org/stplanr/reference/overline.html).
+<!-- The routes were then overlaid and aggregated by segments, using [`stplanr overline()` R function](https://docs.ropensci.org/stplanr/reference/overline.html). -->
 <!-- For example, the estimated volumes of bicycle trips and car transfers were aggregated, and the average car speed and quietness level, weighted by the distance of the segments, were calculated.
@@
 The intermodality scenario considers trips combining PT and cycling for the first and last legs.
-In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for instance: 1 km from origin to interface A plus 4 km from interface B to destination) or up to 25 minutes on bike.
+In a conservative approach, we have restricted our analysis to the first and last legs with a combined length of up to 5 km (for instance: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike.
 Furthermore, we have imposed restrictions on PT usage to include only trips with no PT transfers, and up to 2 hours (120 min).
 Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
+Figure \ref{fig:map2} illustrates the resulting bicycle routes to access the main PT interfaces in the LMA.
 
-```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area", fig.align='center'}
+```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area", fig.align='center', fig.pos="h"}
 knitr::include_graphics("img/map1.png", error = FALSE)
@@
 
-Figure \ref{fig:map2} illustrates the resulting bicycle routes to access the main PT interfaces in the LMA.
-
-```{r map2, out.width="80%", fig.cap="Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).", fig.align='center'}
+```{r map2, out.width="80%", fig.cap="Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).", fig.align='center', fig.pos="h"}
 knitr::include_graphics("img/map2.png", error = FALSE)
@@
 
-For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated, using the HEAT for Cycling tool v5.0 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
-The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years).
+For the cycling legs of the journey (first and last legs), socio-environmental impacts were estimated on the shifting from car to cycling, using the HEAT for Cycling tool v5.0 [@HEAT] from the World Health Organization, and the [`HEATaaS` R package](https://github.com/HEAT-WHO/HEAT_heatr_api)[^5].
+<!-- The HEAT tool provided estimates on the shifting from car to cycling for a short term time horizon (i.e., one year) and the long term (i.e., ten years). -->
 We considered two dimensions: *social* — including the physical activity of cyclists, air pollution exposure, and road casualties; *environmental* — including CO~2~eq emissions and other pollutants.
@@
 To estimate the car emissions, we used the EMEP/EEA's COPERT software v5 methods and reference values [@COPERT] for a Tier 3 detail level. 
-We used a family-size vehicle, EURO standard, and gasoline or diesel fuel.
-All trips were considered to be made under urban conditions and at an average speed of 15 km/h during rush hour periods.
-Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
-In addition, we assumed an occupancy rate of 1.6 passengers *per* car [@IMOB].
+<!-- We used a family-size vehicle, EURO standard, and gasoline or diesel fuel. -->
+<!-- All trips were considered to be made under urban conditions and at an average speed of 15 km/h during rush hour periods. -->
+<!-- Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
+-->
+<!-- In addition, we assumed an occupancy rate of 1.6 passengers *per* car [@IMOB]. -->
+Regarding PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA.
+<!-- [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014]. -->
+<!-- Emissions were estimated for the following atmospheric pollutants: CO, PM10, NOx, and VOC; and for the main green house gases: CO~2~, CH~4~, and N~2~O, converted in CO~2~eq. -->
+<!-- In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally.
+-->
+The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases.
+<!-- [@EuropeanCommission2014; @bickel2006; @UNITE]. -->
+<!-- The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal [@ANSR2021]. -->
+<!-- We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a discount rate of 5% and inflation of 3%.
+-->
 
-Regarding PT, we considered the emission factor values reported in the environmental and sustainability reports of the PT operators in the LMA [@Carris2019s; @Metro2019s; @CP2019s; @Transtejo2014].
-
-Emissions were estimated for the following atmospheric pollutants: CO, PM10, NOx, and VOC; and for the main green house gases: CO~2~, CH~4~, and N~2~O, converted in CO~2~eq.
-In particular, for the urban train and tram -- with 100% electric traction -- only CO~2~eq emissions were considered (resulting from the production of electricity, considering a "well-to-tank" approach), since the other pollutants are not emitted locally.
-
-The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases [@EuropeanCommission2014; @bickel2006; @UNITE].
-The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal [@ANSR2021].
-We updated all the monetary reference values of the literature based on the annual inflation rate in Portugal for 2022[^6], and our 10-years estimations assumed a discount rate of 5% and inflation of 3%.
-
-[^6]: See [Statistics Portugal tool for inflation rate estimates between years](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc).
+<!-- [^6]: See [Statistics Portugal tool for inflation rate estimates between years](https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc). -->
 
@@
 
+```{r summary0}
+summary0 = readRDS(url("https://github.com/U-Shift/biclar/releases/download/0.0.1/HEAT_AML_intermodal_ppp_NEW.Rds"))
+# see code/test-code/heat-prep.R#intermodality
+
+summary0 = summary0 |> 
+    mutate(Target = paste0(ENMAC*100,"%"),
+           Routing = c("safe", "safe", "direct", "direct"),
+           TotalTrips = Total,
+           Bikebase = Bike,
+           Bikepotential = Bike_new,
+           Social = Economic - Environmenal_only,
+           Social10 = Economic10 - Environmenal10_only) 
+```
+
 ```{r summary1}
-summary1 = readRDS("load/summary1.Rds")
+summary1 = summary0 |> 
+  select(Target, Routing, TotalTrips, Bikebase, Bikepotential, CO2eq, Economic10) |> 
+    arrange(desc(Target))
 
-summary1 = summary1 %>%
-  mutate(LTS = recode(LTS, `3` = "safe", `4` = "direct")) |> 
-  mutate(Estrategia = paste0(Estrategia*100,"%")) |> 
-  mutate(TotalTrips = TotalTrips/2,
-         Bikebase = Bikebase/2,
-         Bikepotential = Bikepotential/2) #this is a quick correction while we donr run biclar entirely again.
-The error was found and it was adding up the 2 bike legs, instead of making the mean.
-The SE estimates remain the same because they are from the bikeshare and carshare percentages, that did not change.
-
-knitr::kable(summary1,
+knitr::kable(summary1 |> mutate(Economic10 = Economic10/1000),
       digits = 0,
@@
 
-Table \ref{tab:summary21} shows the potential trips by PT mode to replace the second leg of the journey, in combination with cycling.
-Trains offer the greatest potential for substitution (88%).
+<!-- Table \ref{tab:summary21} shows the potential trips by PT mode to replace the second leg of the journey, in combination with cycling. -->
+Regarding the PT mode to replace the second leg of the journey, in combination with cycling, trains offer the greatest potential for substitution (88%).
 When comparing the existing PT interfaces (Figure \ref{fig:map1}) with the bike routes with highest potential to serve as first and last legs (Figure \ref{fig:map2}) it becomes clear that the Train interfaces are the ones that have the highest potential to attract car-to-PT substituting trips, if their accessibility by bicycle is improved to be safe.
 
-```{r summary21}
+```{r summary21, eval =FALSE}
 summary2 = readRDS("load/summary2.Rds")
@@
 
-Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips.
+<!-- Table \ref{tab:summary22} presents the results of the avoided emissions and its monetization for the second leg of the journey, by replacing car trips with potential TP trips. -->
 <!-- We estimate a reduction of 8,702 tons of CO2eq emissions by substituting motorized trips relying on fossil fuels and electricity.
@@
 summary22 = summary22 %>%
-  mutate(LTS = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
-  mutate(Value = money_emissions_eur/1000) |> 
-  select(Meta_ENMAC, LTS, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, Value) |> 
-  arrange(Meta_ENMAC)
+  mutate(Routing = recode(Percursos, "seguro" = "safe", "direto" = "direct")) |> 
+  mutate(Value = money_emissions_eur/1000,
+         Target = Meta_ENMAC) |> 
+  select(Target, Routing, CO2eq_TP, CO_PT, PM10_PT, NOx_PT, VOC_PT, Value) |> 
+  arrange(Target)
 
-knitr::kable(summary22,
-      digits = c(0,0,0,0,1,0,1,0),
-      format.args = list(big.mark = " "),
-      col.names = c("Target", "Routing", "CO2eq",
-                    "CO", "PM10", "NOx", "VOC", "Value (k€)"),
-      caption = "\\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
-      format="latex",
-      booktabs = TRUE) |> 
-  kableExtra::column_spec(column = c(3:7), width = "3.5em")
+# knitr::kable(summary22,
+#       digits = c(0,0,0,0,1,0,1,0),
+#       format.args = list(big.mark = " "),
+#       col.names = c("Target", "Routing", "CO2eq",
+#                     "CO", "PM10", "NOx", "VOC", "Value (k€)"),
+#       caption = "\\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
+#       format="latex",
+#       booktabs = TRUE) |> 
+#   kableExtra::column_spec(column = c(3:7), width = "3.5em")
 ```
@@
 ```{r summaryall, message=FALSE, warning=FALSE}
-summary0 = readRDS("load/HEAT_aml_results.Rds") |> slice_tail(n = 4)
 rate = ((1+0.03)^10)/((1+0.05)^10) # inflation rate 3% discount rate 5%
 
-summaryall = summary1 |> 
-  mutate(Meta_ENMAC = Estrategia) |>
+summaryall = summary0 |> 
   left_join(summary22) |> 
-  mutate(Code = c(3304, 3404, 3310, 3410)) |> 
-  left_join(summary0 |> select(Code, Economic), by = "Code") |> 
-  select(Estrategia, LTS, co2, CO2eq_TP, monet, Value, Economic) |> 
-  mutate(CO2eq = co2 + CO2eq_TP,
-         value1 = Economic/1000 + Value, # this is CORRECT
-         value10 = monet + Value*10*rate) |> # this is CORRECT
-  select(Estrategia, LTS, CO2eq, value1, value10)
+  select(Target, Routing, CO2eq, CO2eq_TP, Environmenal_only, Environmenal10_only, Social, Social10, Value) |> 
+    mutate(Social = Social/1000,
+           Social10 = Social10/1000,
+         Environmenal_only = Environmenal_only/1000,
+         Environmenal10_only = Environmenal10_only/1000) |> 
+  mutate(CO2eqT = CO2eq + CO2eq_TP,
+         value1 = Environmenal_only + Value + Social, # this is CORRECT
+         value10 = Environmenal10_only + Value*10*rate + Social10) |> # this is CORRECT
+  select(Target, Routing, CO2eqT, value1, value10) |> 
+  arrange(desc(Target))
 
@@
 <!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? It shouldn't be ok to sum them -->
-Shifting from car to cycling + PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year, 
-and the 10-year socio-environmental benefits account for €235 million to €620 million, depending on the cycling targets.
+Shifting from car to cycling + PT can reduce annual CO~2~eq emissions emissions by 11,500 to 28,500 tons per year, 
+and the 10-year socio-environmental benefits account for €125 million to €325 million, depending on the cycling targets
 
-The social impacts represent 98% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs.
+The environmental impacts represent less than 2% of the socio-environmental benefits (in value) from replacing car trips to bicycle in first-and-last legs.
 For the PT segment, we did not estimate the social impacts from substituting car trips, although its health benefits would not be as high as shifting to cycling.<!-- We should estimate the social impacts from replacing Car to PT in the second leg, for the other paper --> <!--Overall there was very little evidence available on mode switch to public transport.
@@
 
-The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 75% of the emissions avoided during the PT segment.
+The emissions of CO~2~eq that are avoided during both the initial and final journey segments account for about 74% of the emissions avoided during the PT segment.
 This finding, while expected -- due the zero cycling emissions, should not be overlooked when promoting the PT use.
@@
 
-Our findings suggest that cycling and PT *in combination* could viably replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution. 
-<!-- repeated? not correct -->
+Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.
 

### 2023-12-21 paper/PaperTRA/PaperTRA_rev1.tex
reduce do 6+1

--- 2023-10-11 paper/PaperTRA/PaperTRA.tex
@@
 \begin{abstract}
-A high proportion of car trips can be replaced by a combination of
-public transit and cycling for the first-and-last mile. This paper
-estimates the potential for cycling combined with public transit (PT) as
-a substitute for car trips in the Lisbon metropolitan area and assesses
-its socio-environmental impacts using open data and open source tools. A
-decision support tool that facilitates the design and development of a
-metropolitan cycling network was developed (\emph{biclaR}). The social
-and environmental impacts were assessed using the \emph{HEAT for
-Cycling} and the \emph{HEAT as a Service} tools. The impacts of shifting
-car trips to PT were also estimated and monetized. The results indicate
-that 10\% of trips could be made by bicycle + PT combination. Shifting
-to cycling for the first-and-last mile stages can reduce annual
-CO\textsubscript{2}eq emissions from 6,000 tons/day, with benefits over
-10 years of at least €230 million. For the PT leg, the transfer from car
-avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per year.
-This evidence can support policymakers to prioritize interventions that
-reduce the reliance on private motor vehicles.
+This paper estimates the potential for cycling combined with public
+transit (PT) as a substitute for car trips in the Lisbon metropolitan
+area and assesses its socio-environmental impacts using open data and
+open source tools. A decision support tool that facilitates the design
+and development of a metropolitan cycling network was developed
+(\emph{biclaR}). The social and environmental impacts were assessed
+using the \emph{HEAT for Cycling} and the \emph{HEAT as a Service}
+tools. The impacts of shifting car trips to PT were also estimated and
+monetized. The results indicate that 10\% of trips could be made by
+bicycle + PT combination. Shifting to cycling for the first-and-last
+mile stages can reduce annual CO\textsubscript{2}eq emissions from 3,000
+tons/day, with benefits over 10 years of at least €125 million. For the
+PT leg, the transfer from car avoids of up to 20,500 tons of
+CO\textsubscript{2}eq emissions per year. This evidence can support
+policymakers to prioritize interventions that reduce the reliance on
+private motor vehicles.
 
@@
 mile \and Health economic assessment \and Environmental
-impacts \and Open data and methods}
+impacts \and Open methods}
 
@@
 To achieve the cycling targets set by the Portuguese national cycling
-strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}4{]}, the
-Lisbon's Metropolitan Department of Transport introduced
+strategy for 2025 and 2030 (4\% and 10\%, respectively), the Lisbon's
+Metropolitan Department of Transport introduced
 \emph{biclaR}\footnote{See
@@
 decision support tool that facilitates the design and development of a
-metropolitan cycling network {[}5{]}.
+metropolitan cycling network {[}4{]}.
 
@@
 origin-destination data for commuting trips as the basis of estimates of
-cycling potential at zone, route and route network levels {[}6{]}. The
-PCT has been extended to include cycling potential for travel to school
-in England {[}7{]} and other trip types in other countries.\footnote{See
-  \href{https://www.npt.scot}{npt.scot} and
-  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
-  Scotland and Ireland that include estimates of cycling for other
-  purposes.} However, to the best of our knowledge, this is the first
-time that the method has been integrated with public transport data
-using multi-modal routing to estimate the potential and benefits of
-multi-stage cycling and PT trips.
+cycling potential at zone, route and route network levels {[}5{]}.
+However, to the best of our knowledge, this is the first time that the
+method has been integrated with public transport data using multi-modal
+routing to estimate the potential and benefits of multi-stage cycling
+and PT trips.
 
@@
 ensure that a district is not solely characterized by a single point of
-origin and destination for its trips. Aggregating all trips into
-centroids renders the exercise less realistic, as it excludes a
-significant portion of short-distance trips, a prevalent characteristic
-of active mode travel {[}8{]}. The OD Jittering method breaks down a
-single point (i.e., the centroid of an area) into multiple random points
-on the existing and neighboring road network, using OpenStreetMap as a
-reference. This method then distributes the volume of trips within the
-district among the randomly generated origin-destination pairs.
-
-Using the
+origin and destination for its trips. The OD Jittering method breaks
+down a single point (i.e., the centroid of an area) into multiple random
+points on the existing and neighboring road network, using OpenStreetMap
+as a reference {[}6{]}. Using the
 \href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
 package}, we employed a maximum disaggregation level of 100 trips per
-O-D pair for this project. Figure \ref{fig:jitter} illustrates the
-contrast between trip representation through the traditional method,
-which connects a single desire line between each district, and the
-presentation achieved through the randomization and disaggregation of
-trips between districts, specifically for the Lisbon metropolitan area.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth,]{img/jitter}
-
-}
-
-\caption{Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
-\end{figure}
-
-Although this method provides a more realistic representation of the
-trips undertaken compared to the traditional approach, it does not fully
-align with the actual O-D pairs of trips, which remain unknown due to
-data privacy regulations.
+O-D pair for this project. Although this method provides a more
+realistic representation of the trips undertaken compared to the
+traditional approach, it does not fully align with the actual O-D pairs
+of trips, which remain unknown due to data privacy regulations.
 
@@
 characteristics of the cycling and road networks and the location of
-public transport interfaces. Other constraints regarding the behavior of
-potential cyclists determine the routing results. For example, such
-restrictions can favor low speed, low traffic streets, more direct
-routes, and less steep paths, among others, suitable for cycling.
+public transport interfaces.
 
 The selected route choice algorithm was the
-\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}9{]},
-which allows for great flexibility in configuring estimated route types,
-and which proven to provide most accurate route networks for the city of
-Lisbon {[}10{]}. \texttt{r5r} can calculate multi-modal routes using PT
-combined with other modes. It enables the identification of the most
-direct or safest cycling routes, using the Level of Traffic
-Stress\footnote{see
+\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}7{]},
+which allows for great flexibility in configuring estimated route types.
+\texttt{r5r} can calculate multi-modal routes using PT combined with
+other modes. It enables the identification of the most direct or safest
+cycling routes, using the Level of Traffic Stress\footnote{see
   \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}.}
-(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
-(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
-routes shared with motorized traffic). The routes were estimated for the
-base scenario for both types of networks: \emph{direct} and \emph{safe},
-using LTS 4 and LTS 3, respectively. Different routing profiles enable
-decision-makers to plan for different bicycle user typologies and/or for
-different city cycling maturity levels {[}11{]}.
+(LTS) scale. The routes were estimated for the base scenario for both
+types of networks: \emph{direct} and \emph{safe}, using LTS 4 and LTS 3,
+respectively.
 
@@
 metropolitan data aggregated and validated. This information is crucial
-for an accurate PT trip and route estimation. A digital elevation model,
-from the European Space Agency's COPERNICUS mission, was used to include
-street gradient information, as a weight in cycling routing. The cycling
-potential trips for the two national strategic targets (4\% and 10\%)
-were estimated from the 2017 cycling and car trips (both as a driver and
-as a passenger), the baseline scenario.
-
-The routes were then overlaid and aggregated by segments, using
-\href{https://docs.ropensci.org/stplanr/reference/overline.html}{\texttt{stplanr\ overline()}
-R function}.
+for an accurate PT trip and route estimation.
 
@@
 ferries, trams, and inter-municipal bus lines equipped with bike racks
-(Figure \ref{fig:map1}).
+(Figure \ref{fig:map1}). Figure \ref{fig:map2} illustrates the resulting
+bicycle routes to access the main PT interfaces in the LMA.
 
-\begin{figure}
+\begin{figure}[h]
 
@@
 
-Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
-the main PT interfaces in the LMA.
-
-\begin{figure}
+\begin{figure}[h]
 
@@
 For the cycling legs of the journey (first and last legs),
-socio-environmental impacts were estimated, using the HEAT for Cycling
-tool v5.0 {[}12{]} from the World Health Organization, and the
+socio-environmental impacts were estimated on the shifting from car to
+cycling, using the HEAT for Cycling tool v5.0 {[}8{]} from the World
+Health Organization, and the
 \href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
@@
   information contact
-  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. The
-HEAT tool provided estimates on the shifting from car to cycling for a
-short term time horizon (i.e., one year) and the long term (i.e., ten
-years). We considered two dimensions: \emph{social} --- including the
-physical activity of cyclists, air pollution exposure, and road
-casualties; \emph{environmental} --- including CO\textsubscript{2}eq
-emissions and other pollutants.
+  \href{https://heatwalkingcycling.org}{heatwalkingcycling.org}.}. We
+considered two dimensions: \emph{social} --- including the physical
+activity of cyclists, air pollution exposure, and road casualties;
+\emph{environmental} --- including CO\textsubscript{2}eq emissions and
+other pollutants.
 
@@
 To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
-methods and reference values {[}13{]} for a Tier 3 detail level. We used
-a family-size vehicle, EURO standard, and gasoline or diesel fuel. All
-trips were considered to be made under urban conditions and at an
-average speed of 15 km/h during rush hour periods. Since the average
-distance traveled per trip influences the overconsumption and emissions
-from cold-start engine operation, we estimated energy and emission
-factors for different ranges of trips at 500-meter intervals. In
-addition, we assumed an occupancy rate of 1.6 passengers \emph{per} car
-{[}3{]}.
-
+methods and reference values {[}9{]} for a Tier 3 detail level.
 Regarding PT, we considered the emission factor values reported in the
-environmental and sustainability reports of the PT operators in the LMA
-{[}14--17{]}.
-
-Emissions were estimated for the following atmospheric pollutants: CO,
-PM10, NOx, and VOC; and for the main green house gases:
-CO\textsubscript{2}, CH\textsubscript{4}, and N\textsubscript{2}O,
-converted in CO\textsubscript{2}eq. In particular, for the urban train
-and tram -- with 100\% electric traction -- only CO\textsubscript{2}eq
-emissions were considered (resulting from the production of electricity,
-considering a ``well-to-tank'' approach), since the other pollutants are
-not emitted locally.
-
+environmental and sustainability reports of the PT operators in the LMA.
 The conversion of avoided emissions into avoided welfare loss and
 respective monetary valuation was based on the EU Guide to Cost-benefit
-Analysis {[}18{]} and the best up-to-date reference values for the
-various gases {[}18--20{]}. The social impacts are in avoided premature
-mortality. This result is finally monetized using the \emph{Statistical
-Value of Life} for Portugal {[}21{]}. We updated all the monetary
-reference values of the literature based on the annual inflation rate in
-Portugal for 2022\footnote{See
-  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
-  Portugal tool for inflation rate estimates between years}.}, and our
-10-years estimations assumed a discount rate of 5\% and inflation of
-3\%.
+Analysis {[}10{]} and the best up-to-date reference values for the
+various gases.
 
@@
 \midrule
-4\% & safe & 538 514 & 2 312 & 20 385 & 5 920 & 230 270\\
-4\% & direct & 500 880 & 2 274 & 18 944 & 6 011 & 223 720\\
-10\% & safe & 538 514 & 2 312 & 52 324 & 15 192 & 591 790\\
-10\% & direct & 500 880 & 2 274 & 48 609 & 15 414 & 574 200\\
+4\% & safe & 538 514 & 2 312 & 20 385 & 2 958 & 115 780\\
+4\% & direct & 500 880 & 2 274 & 18 944 & 3 004 & 113 030\\
+10\% & safe & 538 514 & 2 312 & 52 323 & 7 590 & 297 510\\
+10\% & direct & 500 880 & 2 274 & 48 609 & 7 694 & 289 290\\
 \bottomrule
@@
 
-Table \ref{tab:summary21} shows the potential trips by PT mode to
-replace the second leg of the journey, in combination with cycling.
-Trains offer the greatest potential for substitution (88\%). When
-comparing the existing PT interfaces (Figure \ref{fig:map1}) with the
-bike routes with highest potential to serve as first and last legs
-(Figure \ref{fig:map2}) it becomes clear that the Train interfaces are
-the ones that have the highest potential to attract car-to-PT
-substituting trips, if their accessibility by bicycle is improved to be
-safe.
+Regarding the PT mode to replace the second leg of the journey, in
+combination with cycling, trains offer the greatest potential for
+substitution (88\%). When comparing the existing PT interfaces (Figure
+\ref{fig:map1}) with the bike routes with highest potential to serve as
+first and last legs (Figure \ref{fig:map2}) it becomes clear that the
+Train interfaces are the ones that have the highest potential to attract
+car-to-PT substituting trips, if their accessibility by bicycle is
+improved to be safe.
 
-\begin{table}
-
-\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.}
-\centering
-\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4.5em}>{\raggedright\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}>{\raggedleft\arraybackslash}p{4.5em}}
-\toprule
-Target & Routing & Potential & Bus & Ferry & Train & Tram\\
-\midrule
-4\% & safe & 20 385 & 573 & 285 & 17 716 & 1 811\\
-4\% & direct & 18 944 & 593 & 313 & 17 093 & 946\\
-10\% & safe & 52 323 & 1 452 & 712 & 45 588 & 4 571\\
-10\% & direct & 48 609 & 1 520 & 781 & 43 932 & 2 375\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-Table \ref{tab:summary22} presents the results of the avoided emissions
-and its monetization for the second leg of the journey, by replacing car
-trips with potential TP trips. Regarding the PT segment, the shift from
-private car usage would lead to the mitigation of CO\textsubscript{2}
-equivalent emissions to 8,500 to 20,800 tons annually, valued in €1.4
-million to €3.5 million yearly.
-
-\begin{table}
-
-\caption{\label{tab:summary22}\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.}
-\centering
-\begin{tabular}[t]{ll>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{3.5em}r}
-\toprule
-Target & Routing & CO2eq & CO & PM10 & NOx & VOC & Value (k€)\\
-\midrule
-4\% & safe & 8 593 & 17 & 1.9 & 27 & 0.8 & 1 425\\
-4\% & direct & 8 702 & 18 & 2.0 & 28 & 0.8 & 1 453\\
-10\% & safe & 20 627 & 42 & 4.6 & 65 & 2.0 & 3 431\\
-10\% & direct & 20 793 & 42 & 4.7 & 66 & 1.9 & 3 487\\
-\bottomrule
-\end{tabular}
-\end{table}
+Regarding the PT segment, the shift from private car usage would lead to
+the mitigation of CO\textsubscript{2} equivalent emissions to 8,500 to
+20,800 tons annually, valued in €1.4 million to €3.5 million yearly.
 
@@
 \midrule
-4\% & safe & 14 513 & 27 299 & 242 024\\
-4\% & direct & 14 713 & 26 572 & 235 706\\
-10\% & safe & 35 819 & 69 938 & 620 094\\
-10\% & direct & 36 207 & 67 956 & 602 972\\
+4\% & safe & 11 551 & 14 380 & 127 534\\
+4\% & direct & 11 706 & 14 135 & 125 016\\
+10\% & safe & 28 217 & 36 861 & 325 814\\
+10\% & direct & 28 487 & 35 905 & 318 062\\
 \bottomrule
@@
 Shifting from car to cycling + PT can reduce annual
-CO\textsubscript{2}eq emissions by 14,000 to 36,000 tons per year, and
-the 10-year socio-environmental benefits account for €235 million to
-€620 million, depending on the cycling targets.
+CO\textsubscript{2}eq emissions emissions by 11,500 to 28,500 tons per
+year, and the 10-year socio-environmental benefits account for €125
+million to €325 million, depending on the cycling targets
 
-The social impacts represent 98\% of the socio-environmental benefits
-(in value) from replacing car trips to bicycle in first-and-last legs.
-For the PT segment, we did not estimate the social impacts from
-substituting car trips, although its health benefits would not be as
-high as shifting to cycling.
+The environmental impacts represent less than 2\% of the
+socio-environmental benefits (in value) from replacing car trips to
+bicycle in first-and-last legs. For the PT segment, we did not estimate
+the social impacts from substituting car trips, although its health
+benefits would not be as high as shifting to cycling.
 
 The emissions of CO\textsubscript{2}eq that are avoided during both the
-initial and final journey segments account for about 75\% of the
+initial and final journey segments account for about 74\% of the
 emissions avoided during the PT segment. This finding, while expected --
@@
 
-Our findings suggest that cycling and PT \emph{in combination} could
-viably replace 10\% of current LMA trips, with an additional 6\% of PT
-journeys prone to further substitution.
+Our findings show that cycling \emph{in combination} with PT could
+replace 10\% of current LMA trips, with an additional 6\% of PT journeys
+prone to further substitution, based on conservative assumptions.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
 \CSLLeftMargin{4. }%
-\CSLRightInline{Presidência do Conselho de Ministros: Resolução do
-conselho de ministros n.º 131/2019,
-\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}, (2019).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
-\CSLLeftMargin{5. }%
 \CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{5. }%
 \CSLRightInline{Lovelace, R., Goodman, A., Aldred, R., Berkoff, N.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
-\CSLLeftMargin{7. }%
-\CSLRightInline{Goodman, A., Rojas, I.F., Woodcock, J., Aldred, R.,
-Berkoff, N., Morgan, M., Abbas, A., Lovelace, R.: Scenarios of cycling
-to school in england, and associated health and carbon impacts:
-Application of the {`}propensity to cycle tool{'}. Journal of Transport
-\& Health. 12, 263--278 (2019).
-https://doi.org/\href{https://doi.org/10.1016/j.jth.2019.01.008}{10.1016/j.jth.2019.01.008}.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{6. }%
 \CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
@@
 \leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
-\CSLLeftMargin{10. }%
-\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
-jittering and routing options for converting origin-destination data
-into route networks: Towards accurate estimates of movement at the
-street level. The International Archives of the Photogrammetry, Remote
-Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
-(2022).
-https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
-\CSLLeftMargin{11. }%
-\CSLRightInline{Félix, R., Moura, F., Clifton, K.J.: Typologies of urban
-cyclists: Review of market segmentation methods for planning practice.
-Transportation Research Record. 2662, 125--133 (2017).
-https://doi.org/\href{https://doi.org/10.3141/2662-14}{10.3141/2662-14}.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{Kahlmeier, S., Götschi, T., Cavill, N., Castro
@@
 \leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{9. }%
 \CSLRightInline{Ntziachristos, L., Samaras, Z.: {EMEP/EEA} air pollutant
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
-\CSLLeftMargin{14. }%
-\CSLRightInline{Carris:
-\href{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}{Relatório
-de sustentabilidade 2019 - demostração não financeira}. {Carris -
-Companhia Carris de Ferro de Lisboa, E.M., S.A.} (2020).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
-\CSLLeftMargin{15. }%
-\CSLRightInline{Metropolitano de Lisboa:
-\href{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}{Relatório
-integrado 2019}. {Metropolitano de Lisboa, E.P.E.} (2020).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
-\CSLLeftMargin{16. }%
-\CSLRightInline{CP:
-\href{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}{Relatório
-de sustentabilidade 2019}. {CP - Comboios de Portugal, E.P.E.} (2020).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
-\CSLLeftMargin{17. }%
-\CSLRightInline{Transtejo:
-\href{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}{Relatório
-de sustentabilidade 2014}. {Grupo Transtejo, S.A.} (2014).}
-
 \leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{Sartori, D., Catalano, G., Genco, M., Pancotti, C.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
-\CSLLeftMargin{19. }%
-\CSLRightInline{Bickel, P., Friedrich, R., Burgess, A., Fagiani, P.,
-Hunt, A., Jong, G.D., Laird, J.:
-\href{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}{{HEATCO
-- Developing Harmonised European Approaches for Transport Costing and
-Project Assessment. Deliverable 5, Proposal for Harmonised Guidelines.}}
-(2006).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
-\CSLLeftMargin{20. }%
-\CSLRightInline{Nash, C., others:
-\href{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}{UNIfication
-of accounts and marginal costs for transport efficiency: Final report}.
-{Institute for Transport Studies, University of Leeds} (2003).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
-\CSLLeftMargin{21. }%
-\CSLRightInline{Silva, C., Bravo, J.V., Gonçalves, J.:
-\href{http://www.ansr.pt/Estatisticas/RelatoriosTematicos/Documents/O\%20Impacto\%20Economico\%20e\%20Social\%20da\%20Sinistralidade\%20-\%20PT.pdf}{{Impacto
-Económico e Social da Sinistralidade Rodoviária em Portugal}}. {Centro
-de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
-Rodoviária (ANSR)} (2021).}
-
 \end{CSLReferences}

### 2023-12-21 paper/PaperTRA/PaperTRA_rev1.tex
shorter aknowlegments

@@
 our analysis to the first and last legs with a combined length of up to
-5 km (for instance: 1 km from origin to interface A plus 4 km from
-interface B to destination) or up to 25 minutes on bike. Furthermore, we
-have imposed restrictions on PT usage to include only trips with no PT
-transfers, and up to 2 hours (120 min). Additionally, we have only
-included PT modes that can easily accommodate bicycles, such as trains,
-ferries, trams, and inter-municipal bus lines equipped with bike racks
-(Figure \ref{fig:map1}). Figure \ref{fig:map2} illustrates the resulting
-bicycle routes to access the main PT interfaces in the LMA.
+5 km (for instance: 1 km from origin to interface \emph{A} plus 4 km
+from interface \emph{B} to destination) or up to 25 minutes on bike.
+Furthermore, we have imposed restrictions on PT usage to include only
+trips with no PT transfers, and up to 2 hours (120 min). Additionally,
+we have only included PT modes that can easily accommodate bicycles,
+such as trains, ferries, trams, and inter-municipal bus lines equipped
+with bike racks (Figure \ref{fig:map1}). Figure \ref{fig:map2}
+illustrates the resulting bicycle routes to access the main PT
+interfaces in the LMA.
 
@@
 a Tecnologia (FCT), Portugal in the framework of project
-UIDB/04625/2020. The authors thank Thomas Götschi (HEAT for Cycling) for
-providing access to HaaS tool, which is under development.
+UIDB/04625/2020.
 

### 2024-04-17 paper/PaperCEUS/PaperCEUS_rev1.Rmd
Create PaperCEUS_rev1.Rmd

--- 2023-11-30 paper/PaperCEUS/PaperCEUS.Rmd
@@
 
-Emission factors are estimated for the following air pollutants: CO, NO~X~, VOC, and PM.
+Emission factors are estimated for the following air pollutants: CO, NO~X~, VOC, and PM~10~.
 Emission factors of the main greenhouse gases (GHG) are also estimated: CO~2~, CH~4~ and N~2~O, converted in CO~2~eq by the following relationship[^9]: $EF_{CO_2eq} = EF_{CO_2} + 28\cdot EF_{CH_4} + 265\cdot EF_{N_2O}$.
 The CH~4~ and N~2~O emission factors do not vary with travel speed.
-The PM10 emission factor does not vary with trip distance.
+The PM~10~ emission factor does not vary with trip distance.
 
@@
 
-The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases [@EuropeanCommission2014; @bickel2006; @UNITE]: 8.44 €/ton for CO, 2,867.85 €/ton for NO~X~, 340,969.27 €/ton for PM10, 7,169.62 €/ton for VOC and 35.85 €/ton for CO~2~eq.
+The conversion of avoided emissions into avoided welfare loss and respective monetary valuation was based on the EU Guide to Cost-benefit Analysis [@EuropeanCommission2014] and the best up-to-date reference values for the various gases [@EuropeanCommission2014; @bickel2006; @UNITE]: 8.44 €/ton for CO, 2,867.85 €/ton for NO~X~, 340,969.27 €/ton for PM~10~, 7,169.62 €/ton for VOC and 35.85 €/ton for CO~2~eq.
 The social impacts are in avoided premature mortality. This result is finally monetized using the *Statistical Value of Life* for Portugal: €3,055,358/fatality [@ANSR2021].
@@
 <!-- FOR OTHER PAPER: separate the Socio impacts from the Environmental ones? -->
-Shifting from car to cycling in combination with PT can reduce annual CO~2~eq emissions by 14,000 to 36,000 tons per year. 
+Shifting from car to cycling in combination with PT can reduce annual CO~2~eq emissions by 11,500 to 28,500 tons per year. 
+These figures represent a 2.7% reduction in Lisbon's transport emissions [@LisboaENova], a small but important component of wider transport decarbonization measures.
 The 10-year socio-environmental benefits account for €125 million to €325 million, depending on the cycling targets.
@@
 
-## Acknowledgements {.unnumbered}
+# Acknowledgements {.unnumbered}
 [*blind*]
@@
 
-## Research data {.unnumbered}
+# Research data {.unnumbered}
 The data and the code to reproduce the results will be made available upon publication.

### 2024-04-18 paper/PaperCEUS/PaperCEUS_rev1.Rmd
Update with tracked changes code
jitter fig 2

@@
 
+```{r eval=FALSE, echo=FALSE}
+system("latexdiff paper/PaperCEUS.tex paper/PaperCEUS_rev1.tex > paper/PaperCEUS_diff.tex")
+
+```
+
 <!-- TODO: -->
@@
 Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326].
+
+test
+
 In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system [@RIETVELD200071].
@@
 
-```{r jitter, fig.align='center', fig.cap="Representation of desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
+```{r jitter, fig.align='center', fig.cap="Representation of a flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
 
@@
 # par(mfrow = c(1, 2))
-# plot(od_all$geometry, lwd = 0.1)
+# plot(od_all$geometry, lwd = 0.1,)
 # plot(od_jittered_filter$geometry, lwd = 0.1)
+# knitr::include_graphics("img/jitter.png", error = FALSE)
+# 
+# 
+# odcombined = rbind(
+#   od_all |> 
+#     transmute(Total, Disaggregation = " Original"),
+#   od_jittered_filter |> 
+#     transmute(Total, Disaggregation = "Jittered")
+# )
+# odcombined |> 
+#   sample_n(10000, weight = odcombined$Total) |>
+#   mutate(Total = Total / 1000) |> 
+#   ggplot() +
+#   geom_sf(aes(alpha = Total)) +
+#   facet_wrap(~Disaggregation) +
+#   theme_void() +
+#   geom_sf(data = MUNICIPIOSgeo, fill = NA, color = "gray60")
 
-knitr::include_graphics("img/jitter.png", error = FALSE)
-
+knitr::include_graphics("img/jitter_rev.png", error = FALSE)
 ```

### 2024-04-18 paper/PaperCEUS/PaperCEUS_rev1.Rmd
Update diff

@@
 ```{r eval=FALSE, echo=FALSE}
-system("latexdiff paper/PaperCEUS.tex paper/PaperCEUS_rev1.tex > paper/PaperCEUS_diff.tex")
-
+setwd("paper/PaperCEUS/")
+system("latexdiff PaperCEUS.tex PaperCEUS_rev1.tex > tracked_changes_rev1.tex")
+system("pdflatex tracked_changes_rev1.tex")
 ```
@@
 Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326].
-
-test
-
 In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system [@RIETVELD200071].

### 2024-04-19 paper/PaperCEUS/PaperCEUS_rev1.Rmd
Shorten fig 2 caption

@@
 
-```{r jitter, fig.align='center', fig.cap="Representation of a flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
+```{r jitter, fig.align='center', fig.cap="Flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
 

### 2024-05-03 paper/PaperCEUS/PaperCEUS_rev1.Rmd
solve bib

--- 2024-04-18 paper/PaperCEUS/PaperCEUS_rev1.Rmd
@@
 # csl: computers-environment-and-urban-systems.csl
-# csl: trd_test.csl
 preamble: |
@@
     keep_tex: yes
-    citation_package: "default" #ou nenhum
+    # citation_package: "default" #or none
     latex_engine: xelatex
@@
 
-```{r eval=FALSE, echo=FALSE}
-system("latexdiff paper/PaperCEUS.tex paper/PaperCEUS_rev1.tex > paper/PaperCEUS_diff.tex")
-
+```{r eval=TRUE, echo=FALSE}
+# setwd("paper/PaperCEUS/") # knit directory: Document
+# system("latexdiff PaperCEUS.tex PaperCEUS_rev1.tex > tracked_changes_rev1.tex")
+# system("pdflatex tracked_changes_rev1.tex")
 ```
@@
 knitr::opts_chunk$set(echo = FALSE,
-                      # root.dir = "paper/PaperTRA/",
+                      # root.dir = "paper/PaperCEUS/",
                       out.extra = "")  # This forces knitr to label all figures.
@@
 
-```{r jitter, fig.align='center', fig.cap="Representation of a flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).", out.width="100%"}
+```{r jitter, fig.align='center', fig.cap="Flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts of the 18 municipalities, without jittering (left) and with jittering (right).", fig.height=4, out.width="100%"}
 
@@
 # par(mfrow = c(1, 2))
-# plot(od_all$geometry, lwd = 0.1,)
-# plot(od_jittered_filter$geometry, lwd = 0.1)
-# knitr::include_graphics("img/jitter.png", error = FALSE)
-# 
 # 
@@
 #   theme_void() +
+#   theme(legend.position="none",
+#         strip.text.x = element_blank()) +
 #   geom_sf(data = MUNICIPIOSgeo, fill = NA, color = "gray60")
+# 
+# ggsave(filename = "img/jitter_rev2.png", plot = last_plot(), width = 1000, height = 400, units = "px", dpi = 300, type = "cairo")
 
-knitr::include_graphics("img/jitter_rev.png", error = FALSE)
+knitr::include_graphics("img/jitter_new.png", error = FALSE)
 ```
@@
 Figure \ref{fig:map2} illustrates the resulting bicycle routes to access the main PT interfaces in the LMA.
+It is noticeable the high potential of the train interfaces to attract car-to-PT substituting trips.
 
-```{r map2, out.width="80%", fig.cap="Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).", fig.align='center'}
+```{r map2, out.width="80%", fig.cap="Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).
+Larger the line width, higher the segment's car to bicycle shift potential.
+Darker the line color, lower the quietness level.", fig.align='center'}
 knitr::include_graphics("img/map2.png", error = FALSE)

### 2024-05-03 paper/PaperCEUS/PaperCEUS_rev1.tex
solve bib

--- 2023-11-30 paper/PaperCEUS/PaperCEUS.tex
@@
 \IfFileExists{microtype.sty}{\usepackage{microtype}}{}
+\usepackage[]{natbib}
+\bibliographystyle{elsarticle-harv}
 
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-% Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 mile in metropolitan areas can replace a high proportion of private car
-trips (Martens, 2007). In The Netherlands, which has the highest mode
-share of cycling in the world, cycling accounts for more than a third of
-all trips to and from rail stations at the `home' end of the journey,
-greatly increasing the ability of the transport system (Rietveld, 2000).
-This approach to reducing car dependency and associated externalities
-requires interventions and programs to make bicycling more appealing (La
-Paix et al., 2021). The resulting public investments can have
-significant social and environmental benefits (International Transport
-Forum, 2017). Despite the benefits of cycling-PT intermodality, the
-potential of this combination is often overlooked in transport planning
-(La Paix et al., 2021).
+trips \citep{MARTENS2007326}. In The Netherlands, which has the highest
+mode share of cycling in the world, cycling accounts for more than a
+third of all trips to and from rail stations at the `home' end of the
+journey, greatly increasing the ability of the transport system
+\citep{RIETVELD200071}. This approach to reducing car dependency and
+associated externalities requires interventions and programs to make
+bicycling more appealing \citep{lapaix_role_2021}. The resulting public
+investments can have significant social and environmental benefits
+\citep{internationaltransportforum_integrating_2017}. Despite the
+benefits of cycling-PT intermodality, the potential of this combination
+is often overlooked in transport planning \citep{lapaix_role_2021}.
 
@@
 complementary mode of PT is high. According to the latest mobility
-survey conducted in 2018 (INE, 2018), the LMA registered a total of 5.3
+survey conducted in 2018 \citep{IMOB}, the LMA registered a total of 5.3
 million daily trips, with only 0.5\% by bicycle. Car modal share was
@@
 \begin{figure}
-\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
+\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
 \end{figure}
@@
 To achieve the cycling targets set by the Portuguese national cycling
-strategy for 2025 and 2030 (4\% and 10\%, respectively) (Presidência do
-Conselho de Ministros, 2019), the Lisbon's Metropolitan Department of
-Transport commissioned \emph{biclaR}\footnote{See
+strategy for 2025 and 2030 (4\% and 10\%, respectively) \citep{ENMAC},
+the Lisbon's Metropolitan Department of Transport commissioned
+\emph{biclaR}\footnote{See
   \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}.}, a
 decision support tool that facilitates the planning, design, and
-development of a metropolitan cycling network (Félix et al., 2022).
+development of a metropolitan cycling network \citep{felix2023}.
 
@@
 origin-destination data for commuting trips as the basis of estimates of
-cycling potential at zone, route and route network levels (Lovelace et
-al., 2017). The PCT has been extended to include cycling potential for
-travel to school in England (Goodman et al., 2019) and other trip types
-in other countries.\footnote{See \href{https://www.npt.scot}{npt.scot}
-  and \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
+cycling potential at zone, route and route network levels
+\citep{lovelace2017}. The PCT has been extended to include cycling
+potential for travel to school in England \citep{goodman2019} and other
+trip types in other countries.\footnote{See
+  \href{https://www.npt.scot}{npt.scot} and
+  \href{https://cruse.bike}{cruse.bike} for examples of the PCT in
   Scotland and Ireland that include estimates of cycling for other
@@
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\section{Methods}\label{methods}
 
-\hypertarget{modeling-origin-destination-trips}{%
 \subsection{Modeling Origin-Destination
-trips}\label{modeling-origin-destination-trips}}
+trips}\label{modeling-origin-destination-trips}
 
-The mobility survey data (INE, 2018) is the basis of the baseline
+The mobility survey data \citep{IMOB} is the basis of the baseline
 scenario and trip rates presented in this paper. Conducted in the
@@
 We used `jittering' to disaggregate the OD data, resulting in a wide
-spatial distribution of trip origins and destinations (Lovelace et al.,
-2022b). The method works by sampling `sub-points' (nodes on the
-transport network represented in OpenStreetMap in this case) and using
-these instead of a single point (typically the centroid) to represent
-trip origins and destinations for each zone. This method then
-distributes the trips to desire lines connecting the subpoints based on
-a `disaggregation threshold' which determines the maximum number of
-trips that can be represented by a single desire line.
+spatial distribution of trip origins and destinations
+\citep{Lovelace2022Jittering}. The method works by sampling `sub-points'
+(nodes on the transport network represented in OpenStreetMap in this
+case) and using these instead of a single point (typically the centroid)
+to represent trip origins and destinations for each zone. This method
+then distributes the trips to desire lines connecting the subpoints
+based on a `disaggregation threshold' which determines the maximum
+number of trips that can be represented by a single desire line.
 
@@
 
-{\centering \includegraphics[width=1\linewidth,]{img/jitter}
+{\centering \includegraphics[width=1\linewidth,]{img/jitter_new}
 
@@
 
-\caption{Representation of desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
+\caption{Flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts of the 18 municipalities, without jittering (left) and with jittering (right).}\label{fig:jitter}
 \end{figure}
@@
 
-\hypertarget{modeling-routes}{%
-\subsection{Modeling routes}\label{modeling-routes}}
+\subsection{Modeling routes}\label{modeling-routes}
 
@@
 The selected route choice algorithm was the
-\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} (Pereira
-et al., 2021), which allows for great flexibility in configuring
-estimated route types, and which proven to provide most accurate route
-networks for the city of Lisbon (Lovelace et al., 2022a). \texttt{r5r}
-can calculate multi-modal routes using PT combined with other modes. It
+\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package}
+\citep{r5r}, which allows for great flexibility in configuring estimated
+route types, and which proven to provide most accurate route networks
+for the city of Lisbon \citep{Lovelace2022exploring}. \texttt{r5r} can
+calculate multi-modal routes using PT combined with other modes. It
 enables the identification of the most direct or safest cycling routes,
@@
 decision-makers to plan for different bicycle user typologies and/or for
-different city cycling maturity levels (Félix et al., 2017).
+different city cycling maturity levels \citep{felix2017}.
 
@@
 
-\hypertarget{modeling-intermodality}{%
-\subsection{Modeling intermodality}\label{modeling-intermodality}}
+\subsection{Modeling intermodality}\label{modeling-intermodality}
 
@@
 stages as part of a multi-modal trip are likely to be shorter than
-cycling-only trips (van Mil et al., 2021). Furthermore, we have imposed
-restrictions on PT usage, limiting it to trips without PT transfers, and
-within a duration of up to 2 hours (120 minutes). Additionally, we have
-only included PT modes that can easily accommodate bicycles, such as
-trains, ferries, trams, and inter-municipal bus lines equipped with bike
-racks (Figure \ref{fig:map1}). These restrictions can be eased in the
-future when testing more developed policy interventions to enhance
-intermodality between cycling and PT, considering both the vehicle and
-infrastructure perspectives.
+cycling-only trips \citep{vanmil_insights_2021}. Furthermore, we have
+imposed restrictions on PT usage, limiting it to trips without PT
+transfers, and within a duration of up to 2 hours (120 minutes).
+Additionally, we have only included PT modes that can easily accommodate
+bicycles, such as trains, ferries, trams, and inter-municipal bus lines
+equipped with bike racks (Figure \ref{fig:map1}). These restrictions can
+be eased in the future when testing more developed policy interventions
+to enhance intermodality between cycling and PT, considering both the
+vehicle and infrastructure perspectives.
 
@@
 Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
-the main PT interfaces in the LMA.
+the main PT interfaces in the LMA. It is noticeable the high potential
+of the train interfaces to attract car-to-PT substituting trips.
 
@@
 
-\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
+\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).
+Larger the line width, higher the segment's car to bicycle shift potential.
+Darker the line color, lower the quietness level.}\label{fig:map2}
 \end{figure}
 
-\hypertarget{assessing-socio-environmental-benefits}{%
 \subsection{Assessing socio-environmental
-benefits}\label{assessing-socio-environmental-benefits}}
+benefits}\label{assessing-socio-environmental-benefits}
 
@@
 socio-environmental impacts were estimated, using the Health Economic
-Assessment Tool (HEAT) for Cycling v5.2 (Kahlmeier et al., 2017), from
-the World Health Organization, and the
+Assessment Tool (HEAT) for Cycling v5.2 \citep{HEAT}, from the World
+Health Organization, and the
 \href{https://github.com/HEAT-WHO/HEAT_heatr_api}{\texttt{HEATaaS} R
@@
 To estimate the car emissions, we used the EMEP/EEA's COPERT software v5
-methods and reference values (Ntziachristos \& Samaras, 2020) for a Tier
-3 detail level. We used a family-size vehicle, EURO standard, and
-gasoline or diesel fuel. All trips were considered to be made under
-urban conditions and at an average speed of 15 km/h during rush hour
-periods. Since the average distance traveled per trip influences the
-overconsumption and emissions from cold-start engine operation, we
-estimated energy and emission factors for different ranges of trips at
-500-meter intervals.\\
+methods and reference values \citep{COPERT} for a Tier 3 detail level.
+We used a family-size vehicle, EURO standard, and gasoline or diesel
+fuel. All trips were considered to be made under urban conditions and at
+an average speed of 15 km/h during rush hour periods. Since the average
+distance traveled per trip influences the overconsumption and emissions
+from cold-start engine operation, we estimated energy and emission
+factors for different ranges of trips at 500-meter intervals.\\
 An equation was then used to calculate emission factors for the two
@@
   Portugal: Stock road vehicles statistic}.}. In addition, we assumed an
-occupancy rate of 1.6 passengers \emph{per} car (INE, 2018). Finally,
+occupancy rate of 1.6 passengers \emph{per} car \citep{IMOB}. Finally,
 the final emissions for each trip (\(E_{pollutant}\), in g/trip) are
@@
 environmental and sustainability reports of the PT operators in the LMA
-(Carris, 2020; CP, 2020; Metropolitano de Lisboa, 2020; Transtejo,
-2014). In particular, for the urban train and tram -- with 100\%
-electric traction -- only CO\textsubscript{2}eq emissions were
-considered (resulting from the production of electricity, considering a
-``well-to-tank'' approach), since the other pollutants are not emitted
-locally.
+\citep{Carris2019s, Metro2019s, CP2019s, Transtejo2014}. In particular,
+for the urban train and tram -- with 100\% electric traction -- only
+CO\textsubscript{2}eq emissions were considered (resulting from the
+production of electricity, considering a ``well-to-tank'' approach),
+since the other pollutants are not emitted locally.
 
@@
 respective monetary valuation was based on the EU Guide to Cost-benefit
-Analysis (Sartori et al., 2014) and the best up-to-date reference values
-for the various gases (Bickel et al., 2006; Nash et al., 2003; Sartori
-et al., 2014): 8.44 €/ton for CO, 2,867.85 €/ton for
-NO\textsubscript{X}, 340,969.27 €/ton for PM\textsubscript{10}, 7,169.62
-€/ton for VOC and 35.85 €/ton for CO\textsubscript{2}eq. The social
-impacts are in avoided premature mortality. This result is finally
-monetized using the \emph{Statistical Value of Life} for Portugal:
-€3,055,358/fatality (Silva et al., 2021). We updated all the monetary
-reference values of the literature based on the annual inflation rate in
-Portugal for 2022\footnote{See
-  \href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ipc}{Statistics
+Analysis \citep{EuropeanCommission2014} and the best up-to-date
+reference values for the various gases
+\citep{EuropeanCommission2014, bickel2006, UNITE}: 8.44 €/ton for CO,
+2,867.85 €/ton for NO\textsubscript{X}, 340,969.27 €/ton for
+PM\textsubscript{10}, 7,169.62 €/ton for VOC and 35.85 €/ton for
+CO\textsubscript{2}eq. The social impacts are in avoided premature
+mortality. This result is finally monetized using the \emph{Statistical
+Value of Life} for Portugal: €3,055,358/fatality \citep{ANSR2021}. We
+updated all the monetary reference values of the literature based on the
+annual inflation rate in Portugal for 2022\footnote{See
+  \href{https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ipc}{Statistics
   Portugal tool for inflation rate estimates between years}.}, and our
 10-years estimations assumed a discount rate of 5\% and inflation of
-3\%. See \protect\hyperlink{research-data}{Research Data} for all the
-input values we used.
+3\%. See \hyperref[research-data]{Research Data} for all the input
+values we used.
 
-\hypertarget{results-and-discussion}{%
-\section{Results and Discussion}\label{results-and-discussion}}
+\section{Results and Discussion}\label{results-and-discussion}
 
@@
 figures represent a 2.7\% reduction in Lisbon's transport emissions
-(Lisboa E-Nova, 2023), a small but important component of wider
-transport decarbonization measures. The 10-year socio-environmental
-benefits account for €125 million to €325 million, depending on the
-cycling targets.
+\citep{LisboaENova}, a small but important component of wider transport
+decarbonization measures. The 10-year socio-environmental benefits
+account for €125 million to €325 million, depending on the cycling
+targets.
 
@@
 socio-environmetal benefits, valued after monetization, comes from the
-increase in physical activity (Félix et al., 2023). Although there are
+increase in physical activity \citep{Felix2023ES}. Although there are
 also social benefits form shifting car trips to PT, its health benefits
@@
 is 1.3 plus the ``walking for transportation'' as 3.5, while ``driving a
-car'' is 2.5 (Ainsworth et al., 2011). The difference between these
-activities - shifting from car to PT - is not very obvious when compared
-to shifting from car to cycling, whose MET is about 6.8. Nevertheless,
-future works should also encompass the estimation of the social impacts
-for the PT leg of the journey, shifting from car.
+car'' is 2.5 \citep{MET2011}. The difference between these activities -
+shifting from car to PT - is not very obvious when compared to shifting
+from car to cycling, whose MET is about 6.8. Nevertheless, future works
+should also encompass the estimation of the social impacts for the PT
+leg of the journey, shifting from car.
 
@@
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+\section{Conclusion}\label{conclusion}
 
@@
 
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgements}\label{acknowledgements}}
+\section*{Acknowledgements}\label{acknowledgements}
 \addcontentsline{toc}{section}{Acknowledgements}
@@
 
-\hypertarget{research-data}{%
-\section*{Research data}\label{research-data}}
+\section*{Research data}\label{research-data}
 \addcontentsline{toc}{section}{Research data}
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-MET2011}{}}%
-Ainsworth, B. E., Haskell, W. L., Herrmann, S. D., Meckes, N., Bassett
-Jr, D. R., Tudor-Locke, C., Greer, J. L., Vezina, J., Whitt-Glover, M.
-C., \& Leon, A. S. (2011). {2011 Compendium of Physical Activities: A
-second update of codes and MET values}. \emph{Medicine \& Science in
-Sports \& Exercise}, \emph{43}(8), 1575--1581.
-\url{https://doi.org/10.1249/MSS.0b013e31821ece12}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bickel2006}{}}%
-Bickel, P., Friedrich, R., Burgess, A., Fagiani, P., Hunt, A., Jong, G.
-D., \& Laird, J. (2006). \emph{{HEATCO - Developing Harmonised European
-Approaches for Transport Costing and Project Assessment. Deliverable 5,
-Proposal for Harmonised Guidelines.}}
-\url{https://trimis.ec.europa.eu/sites/default/files/project/documents/20130122_113653_88902_HEATCO_D5_summary.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Carris2019s}{}}%
-Carris. (2020). \emph{Relatório de sustentabilidade 2019 - demostração
-não financeira}. {Carris - Companhia Carris de Ferro de Lisboa, E.M.,
-S.A.} \url{https://www.carris.pt/media/dkdp2wbg/dnf_carris2019_rv5.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CP2019s}{}}%
-CP. (2020). \emph{Relatório de sustentabilidade 2019}. {CP - Comboios de
-Portugal, E.P.E.}
-\url{https://www.cp.pt/StaticFiles/Institucional/2_gestao_sustentavel/1_RelatoriosSustentabilidade/relatorio-de-sustentabilidade-2019.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
-Félix, R., Lovelace, R., \& Moura, F. (2022). \emph{{biclaR - Ferramenta
-de apoio ao planeamento da rede ciclável na área metropolitana de
-Lisboa}}. {CERIS - Instituto Superior Técnico and Transportes
-Metropolitanos de Lisboa}. \url{https://biclar.tmlmobilidade.pt}
-
-\leavevmode\vadjust pre{\hypertarget{ref-felix2017}{}}%
-Félix, R., Moura, F., \& Clifton, K. J. (2017). Typologies of urban
-cyclists: Review of market segmentation methods for planning practice.
-\emph{Transportation Research Record}, \emph{2662}(1), 125--133.
-\url{https://doi.org/10.3141/2662-14}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Felix2023ES}{}}%
-Félix, R., Orozco-Fontalvo, M., \& Moura, F. (2023). Socio-economic
-assessment of shared e-scooters: Do the benefits overcome the
-externalities? \emph{Transportation Research Part D: Transport and
-Environment}, \emph{118}, 103714.
-\url{https://doi.org/10.1016/j.trd.2023.103714}
-
-\leavevmode\vadjust pre{\hypertarget{ref-goodman2019}{}}%
-Goodman, A., Rojas, I. F., Woodcock, J., Aldred, R., Berkoff, N.,
-Morgan, M., Abbas, A., \& Lovelace, R. (2019). Scenarios of cycling to
-school in england, and associated health and carbon impacts: Application
-of the {`}propensity to cycle tool{'}. \emph{Journal of Transport \&
-Health}, \emph{12}, 263--278.
-\url{https://doi.org/10.1016/j.jth.2019.01.008}
-
-\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
-INE. (2018). \emph{Mobilidade e funcionalidade do território nas {Áreas
-Metropolitanas do Porto e de Lisboa}: 2017}. {Instituto National de
-Estatística}.
-\url{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}
-
-\leavevmode\vadjust pre{\hypertarget{ref-internationaltransportforum_integrating_2017}{}}%
-International Transport Forum. (2017). \emph{Integrating {Urban Public
-Transport Systems} and {Cycling}}. {Organisation for Economic
-Co-operation and Development}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
-Kahlmeier, S., Götschi, T., Cavill, N., Castro Fernandez, A., Brand, C.,
-Rojas Rueda, D., Woodcock, J., Kelly, P., Lieb, C., Oja, P., et al.
-(2017). \emph{Health economic assessment tool ({HEAT}) for walking and
-for cycling: Methods and user guide on physical activity, air pollution,
-injuries and carbon impact assessments}.
-\url{https://www.euro.who.int/__data/assets/pdf_file/0010/352963/Heat.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lapaix_role_2021}{}}%
-La Paix, L., Cherchi, E., \& Geurs, K. (2021). Role of perception of
-bicycle infrastructure on the choice of the bicycle as a train feeder
-mode. \emph{International Journal of Sustainable Transportation},
-\emph{15}(6), 486--499.
-\url{https://doi.org/10.1080/15568318.2020.1765223}
-
-\leavevmode\vadjust pre{\hypertarget{ref-LisboaENova}{}}%
-Lisboa E-Nova. (2023). \emph{{Observatórios Lisboa - Infos Emissões
-Gases com Efeito Estufa}}. {Agência de Energia e Ambiente de Lisboa}.
-\url{https://observatorios-lisboa.pt/info_emissoes.html}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
-Lovelace, R., Félix, R., \& Carlino, D. (2022a). Exploring jittering and
-routing options for converting origin-destination data into route
-networks: Towards accurate estimates of movement at the street level.
-\emph{The International Archives of the Photogrammetry, Remote Sensing
-and Spatial Information Sciences}, \emph{XLVIII-4/W1-2022}, 279--286.
-\url{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
-Lovelace, R., Félix, R., \& Carlino, D. (2022b). Jittering: A
-computationally efficient method for generating realistic route networks
-from origin-destination data. \emph{Findings}.
-\url{https://doi.org/10.32866/001c.33873}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
-Lovelace, R., Goodman, A., Aldred, R., Berkoff, N., Abbas, A., \&
-Woodcock, J. (2017). The Propensity to Cycle Tool: An open source online
-system for sustainable transport planning. \emph{Journal of Transport
-and Land Use}, \emph{10}(1). \url{https://doi.org/10.5198/jtlu.2016.862}
-
-\leavevmode\vadjust pre{\hypertarget{ref-MARTENS2007326}{}}%
-Martens, K. (2007). Promoting bike-and-ride: The dutch experience.
-\emph{Transportation Research Part A: Policy and Practice},
-\emph{41}(4), 326--338. \url{https://doi.org/10.1016/j.tra.2006.09.010}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Metro2019s}{}}%
-Metropolitano de Lisboa. (2020). \emph{Relatório integrado 2019}.
-{Metropolitano de Lisboa, E.P.E.}
-\url{https://www.metrolisboa.pt/wp-content/uploads/2021/01/relatorio_integrado_2019.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-UNITE}{}}%
-Nash, C. et al. (2003). \emph{UNIfication of accounts and marginal costs
-for transport efficiency: Final report}. {Institute for Transport
-Studies, University of Leeds}.
-\url{https://www.its.leeds.ac.uk/projects/unite/downloads/Unite\%20Final\%20Report.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
-Ntziachristos, L., \& Samaras, Z. (2020). \emph{{EMEP/EEA} air pollutant
-emission inventory guidebook 2019}. Luxembourg: European Environment
-Agency. \url{https://www.emisia.com/utilities/copert/documentation/}
-
-\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
-Pereira, R. H. M., Saraiva, M., Herszenhut, D., Braga, C. K. V., \&
-Conway, M. W. (2021). r5r: Rapid realistic routing on multimodal
-transport networks with R5 in r. \emph{Findings}.
-\url{https://doi.org/10.32866/001c.21262}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
-Presidência do Conselho de Ministros. (2019). Resolução do conselho de
-ministros n.º 131/2019. In \emph{Diário da República, 1ª série} (Vol.
-147, pp. 46--81).
-\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-RIETVELD200071}{}}%
-Rietveld, P. (2000). The accessibility of railway stations: The role of
-the bicycle in the netherlands. \emph{Transportation Research Part D:
-Transport and Environment}, \emph{5}(1), 71--75.
-\url{https://doi.org/10.1016/S1361-9209(99)00019-X}
-
-\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
-Sartori, D., Catalano, G., Genco, M., Pancotti, C., Sirtori, E.,
-Vignetti, S., \& Del Bo, C. (2014). \emph{Guide to cost-benefit analysis
-of investment projects. Economic apraisal tool for cohesion policy
-2014-2020}. {European Commission - Directorate General for Regional and
-Urban Policy}.
-\url{https://ec.europa.eu/regional_policy/sources/docgener/studies/pdf/cba_guide.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ANSR2021}{}}%
-Silva, C., Bravo, J. V., \& Gonçalves, J. (2021). \emph{{Impacto
-Económico e Social da Sinistralidade Rodoviária em Portugal}}. {Centro
-de Estudos de Gestão do ISEG, Autoridade Nacional de Segurança
-Rodoviária (ANSR)}.
-\url{http://www.ansr.pt/Documents/Impacto_Economico_Social_Sinistralidade_Rodoviaria.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Transtejo2014}{}}%
-Transtejo. (2014). \emph{Relatório de sustentabilidade 2014}. {Grupo
-Transtejo, S.A.}
-\url{https://ttsl.pt/wp-content/uploads/2018/01/rs_2014_min.pdf}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vanmil_insights_2021}{}}%
-van Mil, J. F. P., Leferink, T. S., Annema, J. A., \& van Oort, N.
-(2021). Insights into factors affecting the combined bicycle-transit
-mode. \emph{Public Transport}, \emph{13}(3), 649--673.
-\url{https://doi.org/10.1007/s12469-020-00240-2}
-
-\end{CSLReferences}
+\renewcommand\refname{References}
+\bibliography{bibliography.bib}
 

### 2024-05-03 paper/PaperCEUS/PaperCEUS_prerev.Rmd
prepare review latexdiff

--- 2023-11-30 paper/PaperCEUS/PaperCEUS.Rmd
@@
     keep_tex: yes
-    citation_package: "default" #ou nenhum
+    # citation_package: "default" #ou nenhum
     latex_engine: xelatex
@@
 urlcolor: blue
-citecolor: blue
+citecolor: black
 linkcolor: blue
@@
 ---
+### THIS IS A VERSION BEFORE CHANGES, SO WE CAN COMPARE WITH latexdiff ###
+
+\hypersetup{citecolor=black}
 

### 2024-05-03 paper/PaperCEUS/PaperCEUS_rev1.Rmd
responses 1.2, 1.3, 1.4, 2.2, 2.8
not happy with the blurred jittering lines

@@
     keep_tex: yes
-    # citation_package: "default" #or none
+    # citation_package: "default" #ou nenhum
     latex_engine: xelatex
@@
 ---
+\hypersetup{citecolor=black}
 
-```{r eval=TRUE, echo=FALSE}
+```{r eval=FALSE, echo=FALSE}
 # setwd("paper/PaperCEUS/") # knit directory: Document
-# system("latexdiff PaperCEUS.tex PaperCEUS_rev1.tex > tracked_changes_rev1.tex")
-# system("pdflatex tracked_changes_rev1.tex")
+system("latexdiff paper/PaperCEUS/PaperCEUS_prerev.tex paper/PaperCEUS/PaperCEUS_rev1.tex > paper/PaperCEUS/PaperCEUS_diff1.tex")
+#and change in tex the blue for another color (green? 0,0.5,0) and remove \uwave
+# \definecolor{green}{rgb}{0,0.5,0}
+# {\protect\color{green}{#1}}}
+#and compile PDF!
+system("pdflatex paper/PaperCEUS/PaperCEUS_diff1.tex")
 ```
@@
 knitr::opts_chunk$set(echo = FALSE,
-                      # root.dir = "paper/PaperCEUS/",
+                      # root.dir = "paper/PaperCEUS/", # knit directory: Document, instead of the default
                       out.extra = "")  # This forces knitr to label all figures.
@@
 
-To achieve the cycling targets set by the Portuguese national cycling strategy for 2025 and 2030 (4% and 10%, respectively) [@ENMAC], the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].
+The Portuguese Government has established a national cycling strategy, which outlines targets for the percentage of trips to be made by bicycle in urban areas.
+Specifically, these targets aim for 4% of all trips in 2025 and 10% by 2030 [@ENMAC].
+Moreover, the strategy emphasizes that this increase should be achieved by substituting car trips for bicycle trips, as a way to reduce the environmental and social impacts of car use.
+
+To support the implementation of this strategy, the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].
 
@@
 # 
-# par(mfrow = c(1, 2))
-# 
 # odcombined = rbind(
@@
 # )
+# 
+# cairo_pdf(filename = "img/jitter_cairo.pdf", onefile= T, width=10, height=4) 
+# 
 # odcombined |>
@@
 #   ggplot() +
-#   geom_sf(aes(alpha = Total)) +
+#   geom_sf(aes(alpha = Total)) + # blured
 #   facet_wrap(~Disaggregation) +
@@
 # 
+# dev.off()
+# 
+# png::writePNG(pdftools::pdf_render_page("img/jitter_cairo.pdf", dpi = 300), "img/jitter_cairo.png")
+
 # ggsave(filename = "img/jitter_rev2.png", plot = last_plot(), width = 1000, height = 400, units = "px", dpi = 300, type = "cairo")
 
-knitr::include_graphics("img/jitter_new.png", error = FALSE)
+knitr::include_graphics("img/jitter_cairo.png", error = FALSE)
 ```
@@
 <!-- Include a map with the legs of journeys, and which modes are replacing which, and what impacts are bein estimated at each one. See PTcar_legs.R -->
+Following the clearly established national strategy, we estimated the socio-environmental potential impacts of shifting car trips by bicycle in combination with PT, in the LMA, for the two targets (4% and 10%).
 
@@
 
-Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new trips to achieve the national strategy targets (4% and 10%), for different route profiles.
+Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new daily trips to achieve the national strategy targets (4% and 10%), for different route profiles.
 For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in Table \ref{tab:summary1b}, resulting from replacing car trips with cycling.
@@
                     "Baseline Cycling + PT", "Potencial Cycling + PT"),
-      caption = "\\label{summary1}Summary of the cycling potencial of the intermodality scenario.",
+      caption = "\\label{summary1}Summary of the cycling potencial of the intermodality scenario. Values in 'trips/day'.",
       format="latex",
@@
       #               "Bus", "Ferry", "Train", "Tram"),
-      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.",
+      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.",
       format="latex",
@@
 Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.
+Although this paper does not address the designing specific solutions, our findings indicate a potential surge in demand (+6.3%) of public transport interfaces to align with the national strategies.
+This added demand is conceptually driven from individuals utilising bicycles, either personal or shared, which would be parked at the PT stations.
+Consequently, operators and public space managers would face the challenge of accommodating this increased demand for bicycle parking and/or docks from a bike sharing system.
+Additionally, they may need to explore options to enhance bicycle carriage capacity on public transport services, ensuring alignment with national targets.
+
+This analysis can be applied to other metropolitan areas, providing a basis for decision-making and policy development to support a modal shift from car to PT.
+This process involves a minimum of data requirements, such as the origin-destination data, road network, and the PT GTFS data, which are available in most metropolitan areas.
+Furthermore, it is essential to incorporate existing cycling level targets specific to those areas or consider the possibility of utilising the same targets as outlined in the Portuguese national strategy, in order to compare results with different areas.
+Nevertheless, the extent of the potential shift from car to PT is highly dependent on the distances typically travelled in such contexts, and the availability of PT services - which can vary significantly between regions, and for which we are not assuming any change in this study.
+This paper focus primarily on the potential of making cycling and PT a viable alternative to car trips, and the socio-environmental benefits of such a shift.
 

### 2024-05-03 paper/PaperCEUS/PaperCEUS_rev1.tex
responses 1.2, 1.3, 1.4, 2.2, 2.8
not happy with the blurred jittering lines

@@
 
+\hypersetup{citecolor=black}
+
 \section{Introduction}\label{introduction}
@@
 
-To achieve the cycling targets set by the Portuguese national cycling
-strategy for 2025 and 2030 (4\% and 10\%, respectively) \citep{ENMAC},
-the Lisbon's Metropolitan Department of Transport commissioned
+The Portuguese Government has established a national cycling strategy,
+which outlines targets for the percentage of trips to be made by bicycle
+in urban areas. Specifically, these targets aim for 4\% of all trips in
+2025 and 10\% by 2030 \citep{ENMAC}. Moreover, the strategy emphasizes
+that this increase should be achieved by substituting car trips for
+bicycle trips, as a way to reduce the environmental and social impacts
+of car use.
+
+To support the implementation of this strategy, the Lisbon's
+Metropolitan Department of Transport commissioned
 \emph{biclaR}\footnote{See
@@
 
-{\centering \includegraphics[width=1\linewidth,]{img/jitter_new}
+{\centering \includegraphics[width=1\linewidth,]{img/jitter_cairo}
 
@@
 
+Following the clearly established national strategy, we estimated the
+socio-environmental potential impacts of shifting car trips by bicycle
+in combination with PT, in the LMA, for the two targets (4\% and 10\%).
+
 For the \emph{cycling legs of the journey} (first and last legs),
@@
 restrictions), the trips in the baseline scenario and corresponding new
-trips to achieve the national strategy targets (4\% and 10\%), for
+daily trips to achieve the national strategy targets (4\% and 10\%), for
 different route profiles. For the cycling legs of the journey (first and
@@
 
-\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario.}
+\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario. Values in 'trips/day'.}
 \centering
@@
 
-\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.}
+\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.}
 \centering
@@
 prone to further substitution, based on conservative assumptions.
+Although this paper does not address the designing specific solutions,
+our findings indicate a potential surge in demand (+6.3\%) of public
+transport interfaces to align with the national strategies. This added
+demand is conceptually driven from individuals utilising bicycles,
+either personal or shared, which would be parked at the PT stations.
+Consequently, operators and public space managers would face the
+challenge of accommodating this increased demand for bicycle parking
+and/or docks from a bike sharing system. Additionally, they may need to
+explore options to enhance bicycle carriage capacity on public transport
+services, ensuring alignment with national targets.
+
+This analysis can be applied to other metropolitan areas, providing a
+basis for decision-making and policy development to support a modal
+shift from car to PT. This process involves a minimum of data
+requirements, such as the origin-destination data, road network, and the
+PT GTFS data, which are available in most metropolitan areas.
+Furthermore, it is essential to incorporate existing cycling level
+targets specific to those areas or consider the possibility of utilising
+the same targets as outlined in the Portuguese national strategy, in
+order to compare results with different areas. Nevertheless, the extent
+of the potential shift from car to PT is highly dependent on the
+distances typically travelled in such contexts, and the availability of
+PT services - which can vary significantly between regions, and for
+which we are not assuming any change in this study. This paper focus
+primarily on the potential of making cycling and PT a viable alternative
+to car trips, and the socio-environmental benefits of such a shift.
 

### 2024-05-03 paper/PaperCEUS/PaperCEUS_rev1.Rmd
continue from 2.5 - complexity

@@
 
+## Model replicability and complexity
+
 This analysis can be applied to other metropolitan areas, providing a basis for decision-making and policy development to support a modal shift from car to PT.
@@
 Nevertheless, the extent of the potential shift from car to PT is highly dependent on the distances typically travelled in such contexts, and the availability of PT services - which can vary significantly between regions, and for which we are not assuming any change in this study.
+
+<!-- CONTINUE HERE -->
+A further refinement could be to differentiate between ‘simpler’ no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods: we could set a different maximum cycling stage distance for the former and a shorter maximum cycling stage distance for the latter.
+However, this would have an impact on the explainability of the results to policy makers and the overall complexity.
+We now discuss this and other potential refinements to the methods using the same input datasets and open, reproducible methods in the discussion with reference to the **likelihood of diminishing returns with additional model complexity**.
+
 This paper focus primarily on the potential of making cycling and PT a viable alternative to car trips, and the socio-environmental benefits of such a shift.

### 2024-05-06 paper/PaperCEUS/PaperCEUS_rev1.Rmd
map municipalities plot
something not right regarding the percentages

@@
   - code: CERIS
-    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
+    address: "CERIS, Instituto Superior Técnico - Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal"
   - code: ITS
@@
 
+<!-- INSERT ZONAL MAP HERE -->
+
+```{r mapzones, fig.cap="Potential Bike + PT daily trips for the 10% target, by municipality"}
+CAR_PT_TRANSF = readRDS("load/CAR_TP_TRANSF.Rds")
+
+# summarize by OD pair
+CAR_PT_TRANSF_ODs = CAR_PT_TRANSF |>
+  filter(LTS == 4) |> 
+  group_by(DICOFREor11, DICOFREde11) |>
+  summarise(Total = sum(Total),
+            Car = sum(Car),
+            Bike = sum(Bike),
+            Walk = sum(Walk),
+            TP = sum(TP),
+            Bike_new4 = sum(Bike_new4),
+            Bike_new10 = sum(Bike_new10),
+            Bike_4 = sum(Bike_4),
+            Bike_10 = sum(Bike_10),
+            TP_4 = sum(TP_4),
+            TP_10 = sum(TP_10)) |>
+  mutate(Bikeper = 100*Bike/Total,
+         Bike_new4per = 100*Bike_new4/Total,
+         Bike_new10per = 100*Bike_new10/Total,
+         TPper = 100*TP/Total,
+         TP_4per = 100*TP_4/Total,
+         TP_10per = 100*TP_10/Total) |> 
+  ungroup()
+
+# summarize by origin
+CAR_PT_TRANSF_mun_or = CAR_PT_TRANSF_ODs |> 
+  group_by(DICOFREor11) |>
+  summarise(Total = sum(Total),
+            Car = sum(Car),
+            Bike = sum(Bike),
+            Walk = sum(Walk),
+            TP = sum(TP),
+            Bike_new4 = sum(Bike_new4),
+            Bike_new10 = sum(Bike_new10),
+            Bike_4 = sum(Bike_4),
+            Bike_10 = sum(Bike_10),
+            TP_4 = sum(TP_4),
+            TP_10 = sum(TP_10)) |>
+  mutate(Bikeper = 100*Bike/Total,
+         Bike_new4per = 100*Bike_new4/Total,
+         Bike_new10per = 100*Bike_new10/Total,
+         TPper = 100*TP/Total,
+         TP_4per = 100*TP_4/Total,
+         TP_10per = 100*TP_10) |> 
+  ungroup()
+
+# join with municipalities geo
+CAR_PT_TRANSF_mun_or_geo = CAR_PT_TRANSF_mun_or |> 
+  left_join(FREGUESIASgeo_zones |> select(Dicofre, Concelho) |> sf::st_drop_geometry(),
+            by = c("DICOFREor11" = "Dicofre"))
+CAR_PT_TRANSF_mun_or_geo$Concelho[CAR_PT_TRANSF_mun_or_geo$Concelho == "Setúbal"] = "Setubal"
+CAR_PT_TRANSF_mun_or_geo = CAR_PT_TRANSF_mun_or_geo |> 
+  group_by(Concelho) |>
+  summarise(Total = sum(Total),
+            Car = sum(Car),
+            Bike = sum(Bike),
+            Walk = sum(Walk),
+            TP = sum(TP),
+            Bike_new4 = sum(Bike_new4),
+            Bike_new10 = sum(Bike_new10),
+            Bike_4 = sum(Bike_4),
+            Bike_10 = sum(Bike_10),
+            TP_4 = sum(TP_4),
+            TP_10 = sum(TP_10)) |>
+  mutate(Bikeper = 100*Bike/Total,
+         Bike_new4per = 100*Bike_new4/Total,
+         Bike_new10per = 100*Bike_new10/Total,
+         TPper = 100*TP/Total,
+         TP_4per = 100*TP_4/Total,
+         TP_10per = 100*TP_10/Total) |>
+  left_join(MUNICIPIOSgeo, by = "Concelho") |> 
+  ungroup()
+
+# bind municipalities with zero potential
+CAR_PT_TRANSF_mun_or_geo_18 = CAR_PT_TRANSF_mun_or_geo |> 
+  dplyr::bind_rows(as.data.frame(
+   MUNICIPIOSgeo |> filter(!Concelho %in% CAR_PT_TRANSF_mun_or_geo$Concelho))) |> 
+  sf::st_as_sf()
+
+# re-calculating the percentage of trips
+CAR_PT_TRANSF_mun_or_geo_18 = CAR_PT_TRANSF_mun_or_geo_18 |> 
+  left_join(TRIPSall_mun, by = c("Concelho" = "Origem")) |>
+
+  #CONTINUE FROM HERE ---  LOOK AT DIFRERENCES TOTAL AND VIAGENS
+
+mutate(Car_4 = Car - Bike_new4,
+         Car_10 = Car - Bike_new10,
+         TP_4 = TP + Bike_new4,
+         TP_10 = TP + Bike_new10,)
+
+# plot 1
+ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
+  geom_sf(aes(fill = Bike_new10)) +
+  viridis::scale_fill_viridis(option = "G", direction = -1, name = "Bike + PT trips", na.value = "white") +
+  theme_void()
+  # theme(legend.position = "bottom") +
+  # labs(title = "Potential Bike + PT for 10% target, by municipality")
+
+# plot 2
+ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
+  geom_sf(aes(fill = TP_10per)) +
+  viridis::scale_fill_viridis(option = "G", direction = -1, name = "Bike + PT trips %", na.value = "white") +
+  theme_void()
+  # theme(legend.position = "bottom") +
+  # labs(title = "Potential Bike + PT for 10% target, by municipality")
+
+```
+
 Table \ref{tab:summary22} presents emissions reductions and associated economic benefits associated with the second (PT) leg of trips.

### 2024-05-06 paper/PaperCEUS/PaperCEUS_rev1.Rmd
review percentages bikeper conditional

@@
 CAR_PT_TRANSF_mun_or_geo_18 = CAR_PT_TRANSF_mun_or_geo_18 |> 
-  left_join(TRIPSall_mun, by = c("Concelho" = "Origem")) |>
+  left_join(TRIPSall_mun, by = c("Concelho" = "Origem"))
 
   #CONTINUE FROM HERE ---  LOOK AT DIFRERENCES TOTAL AND VIAGENS
-
-mutate(Car_4 = Car - Bike_new4,
-         Car_10 = Car - Bike_new10,
+ENMAC4 = 0.04 # 4%
+ENMAC10 = 0.10 # 10#
+CAR_PT_TRANSF_mun_or_geo_18 = CAR_PT_TRANSF_mun_or_geo_18 |> 
+  mutate(Bikeper = Bike / viagens) |>  #should it be "viagens" or "Total  -- all trips, not only the found ones
+  mutate(Bike_new4 = ifelse(Bikeper >= ENMAC4*100, Bike, ENMAC4 * viagens),
+         Bike_new10 = ifelse(Bikeper >= ENMAC10*100, Bike, ENMAC10 * viagens)) |> 
+  mutate(Bike4 = ifelse(Bikeper >= ENMAC4*100, Bike, Bike + Bike_new4),
+         Bike10 = ifelse(Bikeper >= ENMAC10*100, Bike, Bike + Bike_new10),
+         Car_4 = ifelse(Bikeper >= ENMAC4*100, Car, Car - Bike_new4),
+         Car_10 = ifelse(Bikeper >= ENMAC10*100, Car, Car - Bike_new10),
          TP_4 = TP + Bike_new4,
-         TP_10 = TP + Bike_new10,)
+         TP_10 = TP + Bike_new10) |> 
+  mutate(Bike_new4per = 100*Bike_new4 / viagens,
+         Bike_new10per = 100*Bike_new10 / viagens,
+         TPper = 100*TP / viagens,
+         TP_4per = 100*TP_4 / viagens,
+         TP_10per = 100*TP_10 / viagens)
 
@@
 
-# plot 2
+# plot 2 - the total % of potential bike + TP trips
 ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
@@
   theme_void()
-  # theme(legend.position = "bottom") +
-  # labs(title = "Potential Bike + PT for 10% target, by municipality")
+
+# plot 3 - the added % of potential bike + TP trips - doesn't make sense to show
+ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
+  geom_sf(aes(fill = Bike_new10per)) + #???
+  viridis::scale_fill_viridis(option = "G", direction = -1, name = "new Bike + PT trips %", na.value = "white") +
+  theme_void()
 

### 2024-05-06 paper/PaperCEUS/PaperCEUS_rev1.Rmd
percentages in trips - still to conclude

@@
 
-  #CONTINUE FROM HERE ---  LOOK AT DIFRERENCES TOTAL AND VIAGENS
 ENMAC4 = 0.04 # 4%
@@
 CAR_PT_TRANSF_mun_or_geo_18 = CAR_PT_TRANSF_mun_or_geo_18 |> 
-  mutate(Bikeper = Bike / viagens) |>  #should it be "viagens" or "Total  -- all trips, not only the found ones
-  mutate(Bike_new4 = ifelse(Bikeper >= ENMAC4*100, Bike, ENMAC4 * viagens),
-         Bike_new10 = ifelse(Bikeper >= ENMAC10*100, Bike, ENMAC10 * viagens)) |> 
-  mutate(Bike4 = ifelse(Bikeper >= ENMAC4*100, Bike, Bike + Bike_new4),
-         Bike10 = ifelse(Bikeper >= ENMAC10*100, Bike, Bike + Bike_new10),
-         Car_4 = ifelse(Bikeper >= ENMAC4*100, Car, Car - Bike_new4),
-         Car_10 = ifelse(Bikeper >= ENMAC10*100, Car, Car - Bike_new10),
+  mutate(Bikeper = Bike / viagens,
+         Carper = Car / viagens) |>  #should it be "viagens" or "Total  -- all trips, not only the found ones
+  mutate(Bike_new4 = ifelse(Bikeper >= ENMAC4 | Carper <= ENMAC4, Bike, ENMAC4 * viagens),
+         Bike_new10 = ifelse(Bikeper >= ENMAC10 | Carper <= ENMAC4, Bike, ENMAC10 * viagens)) |> 
+  mutate(Bike4 = ifelse(Bikeper >= ENMAC4 | Carper <= ENMAC4, Bike, Bike + Bike_new4),
+         Bike10 = ifelse(Bikeper >= ENMAC10 | Carper <= ENMAC4, Bike, Bike + Bike_new10),
+         Car_4 = ifelse(Bikeper >= ENMAC4 | Carper <= ENMAC4, Car, Car - Bike_new4),
+         Car_10 = ifelse(Bikeper >= ENMAC10 | Carper <= ENMAC4, Car, Car - Bike_new10),
          TP_4 = TP + Bike_new4,
@@
          TP_10per = 100*TP_10 / viagens)
+
+  #CONTINUE FROM HERE ---  LOOK AT DIFRERENCES TOTAL AND VIAGENS
 

### 2024-05-07 paper/PaperCEUS/PaperCEUS_rev1.Rmd
respond to 2.2, 2.3, 2.4, 2.5

@@
 
-Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326].
+Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326; @van2021insights].
 In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system [@RIETVELD200071].
@@
 Nevertheless, 29% of trips are longer than 10 km, which requires the use of motorized modes, or active modes in combination with public transport.
+
+Regarding cycling trips in the LMA, 55% are up to 5 km, and 88% are up to 10 km.
+These values are not in line with the typically found in cycling distance decay curves, which show a high percentage of trips are up to 5 km (~75%), and a smaller proportion up to 10 km [@krizek2007detailed; @larsen2010beyond; @DDfunction2023].
+This may be due to the low cycling modal share in the LMA - even more limited in the sample survey, which may be biased towards shorter trips.
 
@@
 
+**TO-DO**
+
+-   mais mapas:
+-   Densidade populational - nº de viagens
+-   Pares OD - desire lines width
+
 The Portuguese Government has established a national cycling strategy, which outlines targets for the percentage of trips to be made by bicycle in urban areas.
@@
 
-This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA, while achieving the national cycling targets.
-After presenting the methods used, it assesses its socio-environmental impacts using open data and open-source tools.
+This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
+After presenting the methods used, we assess its socio-environmental impacts using open data and open-source tools.
+
+It is essential to note that our approach is not predictive but scenario-based, providing a range of potential outcomes regarding the target (% of cycling tips) and the routing profile (safe or direct routes), rather than a single forecast.
+These scenarios are rooted in the explicit national cycling targets, assuming that the potential for cycling is proportional to the reduction of car trips.
+Thus said, we do not predict the generation of new trips, but the substitution of car trips by cycling in combination with PT, considering the potential of the first and last mile of these journeys to be made by bicycle.
+Additionally, we do not model changes in the baseline PT trips, under the implicit assumption that the initial and final stages of these trips remain unaltered.
+A strength of this _backcasting_ approach is that we can simulate the aggregate travel patterns that would meet particular policy objectives, and explore what the implications could be for the public transport system and provision of safe cycling infrastructure to public transport nodes at regional, local authority and, importantly from a local transport planning practitioner perspective, corridor level.
 
 # Methods
+
+## Case Study
 
@@
 The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.
-We restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination) or up to 25 minutes on bike travel-time.
-This conservative approach was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
+We restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination).
 Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
 Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
+
+This conservative approach was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
+Additionally, @Leferink2017 found that the bicycle is most attractive option as a first and last mile mode when the distance ranges from 3 to 5 km, as observed across various comparative studies.
+This restrictions are based on the assumption that 5 km is a plausible maximum distance that people are willing to travel by bike as part of a longer journey.
+This applies especially for cycling beginners shifting from cars, who are less likely to be confident and experienced cyclists than the existing cyclists.
+
 These restrictions can be eased in the future when testing more developed policy interventions to enhance intermodality between cycling and PT, considering both the vehicle and infrastructure perspectives.
@@
 
-```{r mapzones, fig.cap="Potential Bike + PT daily trips for the 10% target, by municipality"}
+```{r mapzones, eval=FALSE, fig.cap="Potential Bike + PT daily trips for the 10% target, by municipality"}
 CAR_PT_TRANSF = readRDS("load/CAR_TP_TRANSF.Rds")
@@
 This added demand is conceptually driven from individuals utilising bicycles, either personal or shared, which would be parked at the PT stations.
-Consequently, operators and public space managers would face the challenge of accommodating this increased demand for bicycle parking and/or docks from a bike sharing system.
+Consequently, operators and public space managers would face the challenge of accommodating this increased demand for bicycle parking and/or docks from a bike sharing system, since this availability plays a crucial role in the decision-making process of individuals to use combined bike-train [@jonkeren2021bicycle].
 Additionally, they may need to explore options to enhance bicycle carriage capacity on public transport services, ensuring alignment with national targets.
 
-## Model replicability and complexity
+## Model replicability and limitations
+
+This paper focus primarily on the potential of making cycling and PT a viable alternative to car trips, and the socio-environmental benefits of such a shift.
 
@@
 Furthermore, it is essential to incorporate existing cycling level targets specific to those areas or consider the possibility of utilising the same targets as outlined in the Portuguese national strategy, in order to compare results with different areas.
+
 Nevertheless, the extent of the potential shift from car to PT is highly dependent on the distances typically travelled in such contexts, and the availability of PT services - which can vary significantly between regions, and for which we are not assuming any change in this study.
+For instance, we considered only the inter-municipal bus services that accommodate bicycles on board.
+Future research could explore the potential of equipping other regional bus services with this feature and analyse their attractiveness, in conjunction with cycling, to address current car trip demands.
 
-<!-- CONTINUE HERE -->
-A further refinement could be to differentiate between ‘simpler’ no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods: we could set a different maximum cycling stage distance for the former and a shorter maximum cycling stage distance for the latter.
-However, this would have an impact on the explainability of the results to policy makers and the overall complexity.
-We now discuss this and other potential refinements to the methods using the same input datasets and open, reproducible methods in the discussion with reference to the **likelihood of diminishing returns with additional model complexity**.
+A strength of the approach is that there is no requirement for sensitive individual trip level data to reproduce the approach in other cities.
+However, we acknowledge that this approach has limitations in terms of being able to identify exactly which trips are being replaced -- in terms of the age of people making the trips, the types of car they drive, and other individual or vehicle level variables.
+We assumed uniformity in the type of vehicle for all replaced car journeys, albeit with variations in the fuels utilized.
+Additionally, we assumed consistency in the health impacts across all individuals for the replaced car trips
+This approach is a simplification of the real-world scenario, but it is a necessary one to make the analysis feasible and reproducible.
 
-This paper focus primarily on the potential of making cycling and PT a viable alternative to car trips, and the socio-environmental benefits of such a shift.
+A further refinement could be to differentiate between ‘simpler’ no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods, by relaxing the assumption of no-transfers between PT modes in `r5r`.
+This would led to a larger number of total daily trips that can be made by cycling + PT, in particular in locations where the PT network is less complex and requires more transfers.
+We could also set a different maximum lenght of cycling stage, extending the bicycle catchment area.
+However, despite the added value of a sensibility analyses, they would also increase the overall complexity, and would have an impact on the explainability of the results to be used by policy makers.
 

### 2024-05-07 paper/PaperCEUS/PaperCEUS_rev1.tex
respond to 2.2, 2.3, 2.4, 2.5

@@
       \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
-Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
+Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
     \affiliation[ITS]{Institute for Transport Studies, University of
@@
 mile in metropolitan areas can replace a high proportion of private car
-trips \citep{MARTENS2007326}. In The Netherlands, which has the highest
-mode share of cycling in the world, cycling accounts for more than a
-third of all trips to and from rail stations at the `home' end of the
-journey, greatly increasing the ability of the transport system
-\citep{RIETVELD200071}. This approach to reducing car dependency and
-associated externalities requires interventions and programs to make
+trips \citep{MARTENS2007326, van2021insights}. In The Netherlands, which
+has the highest mode share of cycling in the world, cycling accounts for
+more than a third of all trips to and from rail stations at the `home'
+end of the journey, greatly increasing the ability of the transport
+system \citep{RIETVELD200071}. This approach to reducing car dependency
+and associated externalities requires interventions and programs to make
 bicycling more appealing \citep{lapaix_role_2021}. The resulting public
@@
 used modes for intercity trips, with cars being the predominant choice
-for all journeys.\\
-53\% trips are up to 5 km distance, and 71\% up to 10 km. Nevertheless,
-29\% of trips are longer than 10 km, which requires the use of motorized
-modes, or active modes in combination with public transport.
+for all journeys. 53\% trips are up to 5 km distance, and 71\% up to 10
+km. Nevertheless, 29\% of trips are longer than 10 km, which requires
+the use of motorized modes, or active modes in combination with public
+transport.
+
+Regarding cycling trips in the LMA, 55\% are up to 5 km, and 88\% are up
+to 10 km. These values are not in line with the typically found in
+cycling distance decay curves, which show a high percentage of trips are
+up to 5 km (\textasciitilde75\%), and a smaller proportion up to 10 km
+\citep{krizek2007detailed, larsen2010beyond, DDfunction2023}. This may
+be due to the low cycling modal share in the LMA - even more limited in
+the sample survey, which may be biased towards shorter trips.
 
@@
 \end{figure}
+
+\textbf{TO-DO}
+
+\begin{itemize}
+\tightlist
+\item
+  mais mapas:
+\item
+  Densidade populational - nº de viagens
+\item
+  Pares OD - desire lines width
+\end{itemize}
 
@@
 This paper estimates the potential for combining cycling and PT to
-substitute car trips in the LMA, while achieving the national cycling
-targets. After presenting the methods used, it assesses its
-socio-environmental impacts using open data and open-source tools.
+substitute car trips in the LMA. After presenting the methods used, we
+assess its socio-environmental impacts using open data and open-source
+tools.
+
+It is essential to note that our approach is not predictive but
+scenario-based, providing a range of potential outcomes regarding the
+target (\% of cycling tips) and the routing profile (safe or direct
+routes), rather than a single forecast. These scenarios are rooted in
+the explicit national cycling targets, assuming that the potential for
+cycling is proportional to the reduction of car trips. Thus said, we do
+not predict the generation of new trips, but the substitution of car
+trips by cycling in combination with PT, considering the potential of
+the first and last mile of these journeys to be made by bicycle.
+Additionally, we do not model changes in the baseline PT trips, under
+the implicit assumption that the initial and final stages of these trips
+remain unaltered. A strength of this \emph{backcasting} approach is that
+we can simulate the aggregate travel patterns that would meet particular
+policy objectives, and explore what the implications could be for the
+public transport system and provision of safe cycling infrastructure to
+public transport nodes at regional, local authority and, importantly
+from a local transport planning practitioner perspective, corridor
+level.
 
 \section{Methods}\label{methods}
+
+\subsection{Case Study}\label{case-study}
 
@@
 km from origin to interface \emph{A} plus 4 km from interface \emph{B}
-to destination) or up to 25 minutes on bike travel-time. This
-conservative approach was adopted to capture the fact that cycling
+to destination). Furthermore, we have imposed restrictions on PT usage,
+limiting it to trips without PT transfers, and within a duration of up
+to 2 hours (120 minutes). Additionally, we have only included PT modes
+that can easily accommodate bicycles, such as trains, ferries, trams,
+and inter-municipal bus lines equipped with bike racks (Figure
+\ref{fig:map1}).
+
+This conservative approach was adopted to capture the fact that cycling
 stages as part of a multi-modal trip are likely to be shorter than
-cycling-only trips \citep{vanmil_insights_2021}. Furthermore, we have
-imposed restrictions on PT usage, limiting it to trips without PT
-transfers, and within a duration of up to 2 hours (120 minutes).
-Additionally, we have only included PT modes that can easily accommodate
-bicycles, such as trains, ferries, trams, and inter-municipal bus lines
-equipped with bike racks (Figure \ref{fig:map1}). These restrictions can
-be eased in the future when testing more developed policy interventions
-to enhance intermodality between cycling and PT, considering both the
-vehicle and infrastructure perspectives.
+cycling-only trips \citep{vanmil_insights_2021}. Additionally,
+\citet{Leferink2017} found that the bicycle is most attractive option as
+a first and last mile mode when the distance ranges from 3 to 5 km, as
+observed across various comparative studies. This restrictions are based
+on the assumption that 5 km is a plausible maximum distance that people
+are willing to travel by bike as part of a longer journey. This applies
+especially for cycling beginners shifting from cars, who are less likely
+to be confident and experienced cyclists than the existing cyclists.
+
+These restrictions can be eased in the future when testing more
+developed policy interventions to enhance intermodality between cycling
+and PT, considering both the vehicle and infrastructure perspectives.
 
@@
 challenge of accommodating this increased demand for bicycle parking
-and/or docks from a bike sharing system. Additionally, they may need to
-explore options to enhance bicycle carriage capacity on public transport
-services, ensuring alignment with national targets.
+and/or docks from a bike sharing system, since this availability plays a
+crucial role in the decision-making process of individuals to use
+combined bike-train \citep{jonkeren2021bicycle}. Additionally, they may
+need to explore options to enhance bicycle carriage capacity on public
+transport services, ensuring alignment with national targets.
+
+\subsection{Model replicability and
+limitations}\label{model-replicability-and-limitations}
+
+This paper focus primarily on the potential of making cycling and PT a
+viable alternative to car trips, and the socio-environmental benefits of
+such a shift.
 
@@
 the same targets as outlined in the Portuguese national strategy, in
-order to compare results with different areas. Nevertheless, the extent
-of the potential shift from car to PT is highly dependent on the
-distances typically travelled in such contexts, and the availability of
-PT services - which can vary significantly between regions, and for
-which we are not assuming any change in this study. This paper focus
-primarily on the potential of making cycling and PT a viable alternative
-to car trips, and the socio-environmental benefits of such a shift.
+order to compare results with different areas.
+
+Nevertheless, the extent of the potential shift from car to PT is highly
+dependent on the distances typically travelled in such contexts, and the
+availability of PT services - which can vary significantly between
+regions, and for which we are not assuming any change in this study. For
+instance, we considered only the inter-municipal bus services that
+accommodate bicycles on board. Future research could explore the
+potential of equipping other regional bus services with this feature and
+analyse their attractiveness, in conjunction with cycling, to address
+current car trip demands.
+
+A strength of the approach is that there is no requirement for sensitive
+individual trip level data to reproduce the approach in other cities.
+However, we acknowledge that this approach has limitations in terms of
+being able to identify exactly which trips are being replaced -- in
+terms of the age of people making the trips, the types of car they
+drive, and other individual or vehicle level variables. We assumed
+uniformity in the type of vehicle for all replaced car journeys, albeit
+with variations in the fuels utilized. Additionally, we assumed
+consistency in the health impacts across all individuals for the
+replaced car trips This approach is a simplification of the real-world
+scenario, but it is a necessary one to make the analysis feasible and
+reproducible.
+
+A further refinement could be to differentiate between `simpler'
+no-transfer PT trips (which require less person effort) and more arduous
+multi-transfer PT trips, using the same methods, by relaxing the
+assumption of no-transfers between PT modes in \texttt{r5r}. This would
+led to a larger number of total daily trips that can be made by cycling
++ PT, in particular in locations where the PT network is less complex
+and requires more transfers. We could also set a different maximum
+lenght of cycling stage, extending the bicycle catchment area. However,
+despite the added value of a sensibility analyses, they would also
+increase the overall complexity, and would have an impact on the
+explainability of the results to be used by policy makers.
 

### 2024-05-07 paper/PaperCEUS/PaperCEUS_rev1.Rmd
imptove map 3

@@
 ```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area.", fig.align='center'}
-knitr::include_graphics("img/map1.png", error = FALSE)
+# knitr::include_graphics("img/map1.png", error = FALSE) #old version
+
+MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
+stopsAML = sf::st_read("load/Stops_AML.geojson", quiet = TRUE)
+stopsAML = stopsAML |> filter(mode != "Metropolitano") |>
+  mutate(mode = factor(mode, levels = c("Comboio", "Ferry", "Metro Superficie", "Autocarro"),
+                       labels = c("Train", "Ferry", "Tram", "Bus")))
+routesAML = sf::st_read("load/Routes_AML.geojson", quiet = TRUE)
+routesAML = routesAML |> 
+  mutate(mode = factor(mode, levels = c("Comboio", "Ferry", "Metro Superficie", "Autocarro"),
+                       labels = c("Train", "Ferry", "Tram", "Bus")))
+
+ggplot() +
+  geom_sf(data = stopsAML, aes(color = mode), size = 2) +
+  geom_sf(data = routesAML, aes(color = mode), size = 1) +
+  scale_color_manual(values = c("purple", "darkgreen", "orange", "dodgerblue")) +
+  geom_sf(data = MUNICIPIOSgeo, fill = NA, color = "gray60") +
+  theme_void() +
+  labs(color = "Mode")
 ```
@@
 CAR_PT_TRANSF_mun_or_geo_18 = CAR_PT_TRANSF_mun_or_geo_18 |> 
-  left_join(TRIPSall_mun, by = c("Concelho" = "Origem"))
+  left_join(TRIPSall_mun, by = c("Concelho" = "Origem")) |> 
+  select(Concelho, Total, Car, Bike, Walk, TP, viagens, geometry)
 
@@
          Carper = Car / viagens) |>  #should it be "viagens" or "Total  -- all trips, not only the found ones
-  mutate(Bike_new4 = ifelse(Bikeper >= ENMAC4 | Carper <= ENMAC4, Bike, ENMAC4 * viagens),
-         Bike_new10 = ifelse(Bikeper >= ENMAC10 | Carper <= ENMAC4, Bike, ENMAC10 * viagens)) |> 
-  mutate(Bike4 = ifelse(Bikeper >= ENMAC4 | Carper <= ENMAC4, Bike, Bike + Bike_new4),
-         Bike10 = ifelse(Bikeper >= ENMAC10 | Carper <= ENMAC4, Bike, Bike + Bike_new10),
-         Car_4 = ifelse(Bikeper >= ENMAC4 | Carper <= ENMAC4, Car, Car - Bike_new4),
-         Car_10 = ifelse(Bikeper >= ENMAC10 | Carper <= ENMAC4, Car, Car - Bike_new10),
+  mutate(Bike_new4 = ifelse(Bikeper <= ENMAC4 & Carper >= ENMAC4, ENMAC4 * viagens, ifelse(Carper <= ENMAC4, Car, Bike)),
+         Bike_new10 = ifelse(Bikeper <= ENMAC10 & Carper >= ENMAC10, ENMAC10 * viagens, ifelse(Carper <= ENMAC10, Car, Bike))) |> 
+  mutate(Bike4 = Bike +Bike_new4,
+         Bike10 = Bike + Bike_new10,
+         Car_4 = Car - Bike_new4,
+         Car_10 = Car - Bike_new10,
          TP_4 = TP + Bike_new4,
@@
          Bike_new10per = 100*Bike_new10 / viagens,
+         Bike_4per = 100*Bike4 / viagens,
+         Bike_10per = 100*Bike10 / viagens,
          TPper = 100*TP / viagens,
@@
 
-  #CONTINUE FROM HERE ---  LOOK AT DIFRERENCES TOTAL AND VIAGENS
+# other way  
+CAR_PT_TRANSF_mun_or_geo_18_extrap = CAR_PT_TRANSF_mun_or_geo_18 |> 
+  mutate(Bikeper = Bike / Total,
+         Carper = Car / Total) |>  #extrapolating to the total trips
+  mutate(Bike_new4 = ifelse(Bikeper <= ENMAC4 & Carper >= ENMAC4, ENMAC4 * Total, ifelse(Carper <= ENMAC4, Car, Bike)),
+         Bike_new10 = ifelse(Bikeper <= ENMAC10 & Carper >= ENMAC10, ENMAC10 * Total, ifelse(Carper <= ENMAC10, Car, Bike))) |> 
+  mutate(Bike4 = Bike +Bike_new4,
+         Bike10 = Bike + Bike_new10,
+         Car_4 = Car - Bike_new4,
+         Car_10 = Car - Bike_new10,
+         TP_4 = TP + Bike_new4,
+         TP_10 = TP + Bike_new10) |> 
+  mutate(Bike_new4per = 100*Bike_new4 / Total,
+         Bike_new10per = 100*Bike_new10 / Total,
+         Bike_4per = 100*Bike4 / Total,
+         Bike_10per = 100*Bike10 / Total,
+         TPper = 100*TP / Total,
+         TP_4per = 100*TP_4 / Total,
+         TP_10per = 100*TP_10 / Total)
 
@@
 ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
-  geom_sf(aes(fill = Bike_new10per)) + #???
+  geom_sf(aes(fill = TP_10per- TPper)) + #???
   viridis::scale_fill_viridis(option = "G", direction = -1, name = "new Bike + PT trips %", na.value = "white") +

### 2024-05-07 paper/PaperCEUS/PaperCEUS_rev1.Rmd
Mention jittering subpoints in the limitations, close #48

--- 2024-05-07 paper/PaperCEUS/PaperCEUS_rev1.Rmd
@@
 
-## Model replicability and limitations
+## Generalizability and limitations
 
-This paper focus primarily on the potential of making cycling and PT a viable alternative to car trips, and the socio-environmental benefits of such a shift.
+This paper focus primarily on the potential for replacing trips currently made by car with combined cycling and public transport trips, and the associated social and environmental benefits.
+While the results are specific to the Lisbon Metropolitan Area, the approach is designed to be generalizable to other metropolitan areas.
 
-This analysis can be applied to other metropolitan areas, providing a basis for decision-making and policy development to support a modal shift from car to PT.
-This process involves a minimum of data requirements, such as the origin-destination data, road network, and the PT GTFS data, which are available in most metropolitan areas.
-Furthermore, it is essential to incorporate existing cycling level targets specific to those areas or consider the possibility of utilising the same targets as outlined in the Portuguese national strategy, in order to compare results with different areas.
+An important aspect of the approach from a generalizability perspective is that it has specific data requirements, including origin-destination, road network, and the GTFS data.
+Such datasets are available to transport planners in many areas.
+In areas where such datasets do not exist, the value of research such as this could provide a motivation to collect and open-up transport data.
+For this study we have used Portugal's national cycling targets as the basis of the scenarios; the approach can be adapted to other targets, or to other policy goals, such as reducing congestion or improving air quality.
 
@@
 A strength of the approach is that there is no requirement for sensitive individual trip level data to reproduce the approach in other cities.
-However, we acknowledge that this approach has limitations in terms of being able to identify exactly which trips are being replaced -- in terms of the age of people making the trips, the types of car they drive, and other individual or vehicle level variables.
+However, we acknowledge that this approach has limitations.
+The approach cannot show exactly which trips are being replaced, in terms of the age of people making the trips, the types of car they drive, and other individual or vehicle level variables.
 We assumed uniformity in the type of vehicle for all replaced car journeys, albeit with variations in the fuels utilized.
-Additionally, we assumed consistency in the health impacts across all individuals for the replaced car trips
-This approach is a simplification of the real-world scenario, but it is a necessary one to make the analysis feasible and reproducible.
+Additionally, we assumed consistency in the health impacts across all individuals for the replaced car trips.
+This approach is a simplification of reality to enable the analysis of a large region on available computing resources.
+Future work could seek to increase the resolution of the results, and move towards a more individual-level analysis, for example by using agent and activity-based models using tools such as MATSim.
+This direction of travel would require substantial resources outside the scope of this study; another way to improve the spatial resolution of results would be to use higher resolution 'subpoints' as the basis of the desire line 'jittering' process.
+We would also like to explore the potential for probabilistic routing to improve the spatial resolution of the results.
 
-A further refinement could be to differentiate between ‘simpler’ no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods, by relaxing the assumption of no-transfers between PT modes in `r5r`.
+A further refinement could be to differentiate between 'simpler' no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods, by relaxing the assumption of no-transfers between PT modes in the routing engine.
 This would led to a larger number of total daily trips that can be made by cycling + PT, in particular in locations where the PT network is less complex and requires more transfers.

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
matsim ref

@@
 # plot 1
-ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
+ggplot(data = CAR_PT_TRANSF_mun_or_geo_18_extrap) +
   geom_sf(aes(fill = Bike_new10)) +
@@
 
+The benefits are spatially unevenly distributed across the LMA, as shown in Figure \ref{fig:mapzones}.
+
+<!-- The potential of the intermodality scenario is not only in the number of trips that can be made, but also in the potential to reduce emissions and improve the quality of life in the LMA. -->
 Table \ref{tab:summary22} presents emissions reductions and associated economic benefits associated with the second (PT) leg of trips.
@@
 
-## Generalizability and limitations
+# Generalizability and limitations
 
@@
 This approach is a simplification of reality to enable the analysis of a large region on available computing resources.
-Future work could seek to increase the resolution of the results, and move towards a more individual-level analysis, for example by using agent and activity-based models using tools such as MATSim.
+Future work could seek to increase the resolution of the results, and move towards a more individual-level analysis, for example by using agent and activity-based models using tools such as MATSim [@MATSim].
 This direction of travel would require substantial resources outside the scope of this study; another way to improve the spatial resolution of results would be to use higher resolution 'subpoints' as the basis of the desire line 'jittering' process.

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
imptove map 3
matsim ref

@@
 
-{\centering \includegraphics[width=0.6\linewidth,]{img/map1}
+{\centering \includegraphics[width=0.6\linewidth,]{PaperCEUS_rev1_files/figure-latex/map1-1}
 
@@
 
+The benefits are spatially unevenly distributed across the LMA, as shown
+in Figure \ref{fig:mapzones}.
+
 Table \ref{tab:summary22} presents emissions reductions and associated
@@
 
-\subsection{Model replicability and
-limitations}\label{model-replicability-and-limitations}
+\section{Generalizability and
+limitations}\label{generalizability-and-limitations}
 
-This paper focus primarily on the potential of making cycling and PT a
-viable alternative to car trips, and the socio-environmental benefits of
-such a shift.
+This paper focus primarily on the potential for replacing trips
+currently made by car with combined cycling and public transport trips,
+and the associated social and environmental benefits. While the results
+are specific to the Lisbon Metropolitan Area, the approach is designed
+to be generalizable to other metropolitan areas.
 
-This analysis can be applied to other metropolitan areas, providing a
-basis for decision-making and policy development to support a modal
-shift from car to PT. This process involves a minimum of data
-requirements, such as the origin-destination data, road network, and the
-PT GTFS data, which are available in most metropolitan areas.
-Furthermore, it is essential to incorporate existing cycling level
-targets specific to those areas or consider the possibility of utilising
-the same targets as outlined in the Portuguese national strategy, in
-order to compare results with different areas.
+An important aspect of the approach from a generalizability perspective
+is that it has specific data requirements, including origin-destination,
+road network, and the GTFS data. Such datasets are available to
+transport planners in many areas. In areas where such datasets do not
+exist, the value of research such as this could provide a motivation to
+collect and open-up transport data. For this study we have used
+Portugal's national cycling targets as the basis of the scenarios; the
+approach can be adapted to other targets, or to other policy goals, such
+as reducing congestion or improving air quality.
 
@@
 individual trip level data to reproduce the approach in other cities.
-However, we acknowledge that this approach has limitations in terms of
-being able to identify exactly which trips are being replaced -- in
-terms of the age of people making the trips, the types of car they
-drive, and other individual or vehicle level variables. We assumed
-uniformity in the type of vehicle for all replaced car journeys, albeit
-with variations in the fuels utilized. Additionally, we assumed
-consistency in the health impacts across all individuals for the
-replaced car trips This approach is a simplification of the real-world
-scenario, but it is a necessary one to make the analysis feasible and
-reproducible.
+However, we acknowledge that this approach has limitations. The approach
+cannot show exactly which trips are being replaced, in terms of the age
+of people making the trips, the types of car they drive, and other
+individual or vehicle level variables. We assumed uniformity in the type
+of vehicle for all replaced car journeys, albeit with variations in the
+fuels utilized. Additionally, we assumed consistency in the health
+impacts across all individuals for the replaced car trips. This approach
+is a simplification of reality to enable the analysis of a large region
+on available computing resources. Future work could seek to increase the
+resolution of the results, and move towards a more individual-level
+analysis, for example by using agent and activity-based models using
+tools such as MATSim \citep{MATSim}. This direction of travel would
+require substantial resources outside the scope of this study; another
+way to improve the spatial resolution of results would be to use higher
+resolution `subpoints' as the basis of the desire line `jittering'
+process. We would also like to explore the potential for probabilistic
+routing to improve the spatial resolution of the results.
 
@@
 multi-transfer PT trips, using the same methods, by relaxing the
-assumption of no-transfers between PT modes in \texttt{r5r}. This would
-led to a larger number of total daily trips that can be made by cycling
-+ PT, in particular in locations where the PT network is less complex
-and requires more transfers. We could also set a different maximum
-lenght of cycling stage, extending the bicycle catchment area. However,
-despite the added value of a sensibility analyses, they would also
-increase the overall complexity, and would have an impact on the
+assumption of no-transfers between PT modes in the routing engine. This
+would led to a larger number of total daily trips that can be made by
+cycling + PT, in particular in locations where the PT network is less
+complex and requires more transfers. We could also set a different
+maximum lenght of cycling stage, extending the bicycle catchment area.
+However, despite the added value of a sensibility analyses, they would
+also increase the overall complexity, and would have an impact on the
 explainability of the results to be used by policy makers.

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
case study and methods structrure

@@
 ---
-title: "Modeling the impacts of replacing car trips with combined public transport and cycling: \ Reproducible methods, results and actionable evidence from biclaR"
+title: "Reproducible methods for modeling combined public transport and cycling trips and associated benefits: \ evidence from the biclaR tool"
 author:
@@
   A decision support tool that facilitates the design and development of a metropolitan cycling network was developed (_biclaR_).
-The social and environmental impacts were assessed using the _HEAT for Cycling_ and the _HEAT as a Service_ tools.
+The social and environmental impacts were assessed using Health World Organization tools.
 The impacts of shifting car trips to public transport were also estimated and monetized.
@@
 In the Lisbon metropolitan area (LMA) the largest metropolitan area in Portugal, the modal share of cycling is low, but the potential for cycling as a complementary mode of PT is high.
+
+The Portuguese Government has established a national cycling strategy, which outlines targets for the percentage of trips to be made by bicycle in urban areas.
+Specifically, these targets aim for 4% of all trips in 2025 and 10% by 2030 [@ENMAC].
+Moreover, the strategy emphasizes that this increase should be achieved by substituting car trips for bicycle trips, as a way to reduce the environmental and social impacts of car use.
+
+To support the implementation of this strategy, the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].
+
+<!-- [^1]: See xxx.xxx. *For double blind peer-review process, the authors decided to truncate the name and references to this tool*. -->
+[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/).
+
+*biclaR* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
+The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
+The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^3]
+However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.
+
+[^2]: See [pct.bike](https://www.pct.bike/).
+
+[^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.
+
+<!-- For the full paper, explain that we had 3 scenarios, and describe them. But in this paper we focus on the intermodality one. -->
+
+This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
+After presenting the methods used, we assess its socio-environmental impacts using open data and open-source tools.
+
+# Methods
+
+## Case Study
+
 According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.
@@
 
-The Portuguese Government has established a national cycling strategy, which outlines targets for the percentage of trips to be made by bicycle in urban areas.
-Specifically, these targets aim for 4% of all trips in 2025 and 10% by 2030 [@ENMAC].
-Moreover, the strategy emphasizes that this increase should be achieved by substituting car trips for bicycle trips, as a way to reduce the environmental and social impacts of car use.
-
-To support the implementation of this strategy, the Lisbon's Metropolitan Department of Transport commissioned *biclaR*[^1], a decision support tool that facilitates the planning, design, and development of a metropolitan cycling network [@felix2023].
-
-<!-- [^1]: See xxx.xxx. *For double blind peer-review process, the authors decided to truncate the name and references to this tool*. -->
-[^1]: See [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt/).
-
-*biclaR* builds on the Propensity to Cycle Tool[^2] (PCT), a web application and research project funded by the UK's Department for Transport in 2015 which launched nationally in 2017 as part of the government's Cycling and Walking Investment Strategy.
-The PCT initially used only origin-destination data for commuting trips as the basis of estimates of cycling potential at zone, route and route network levels [@lovelace2017].
-The PCT has been extended to include cycling potential for travel to school in England [@goodman2019] and other trip types in other countries.[^3]
-However, to the best of our knowledge, this is the first time that the method has been integrated with public transport data using multi-modal routing to estimate the potential and benefits of multi-stage cycling and PT trips.
-
-[^2]: See [pct.bike](https://www.pct.bike/).
-
-[^3]:  See [npt.scot](https://www.npt.scot) and [cruse.bike](https://cruse.bike) for examples of the PCT in Scotland and Ireland that include estimates of cycling for other purposes.
-
-<!-- For the full paper, explain that we had 3 scenarios, and describe them. But in this paper we focus on the intermodality one. -->
-
-This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
-After presenting the methods used, we assess its socio-environmental impacts using open data and open-source tools.
-
-It is essential to note that our approach is not predictive but scenario-based, providing a range of potential outcomes regarding the target (% of cycling tips) and the routing profile (safe or direct routes), rather than a single forecast.
-These scenarios are rooted in the explicit national cycling targets, assuming that the potential for cycling is proportional to the reduction of car trips.
-Thus said, we do not predict the generation of new trips, but the substitution of car trips by cycling in combination with PT, considering the potential of the first and last mile of these journeys to be made by bicycle.
-Additionally, we do not model changes in the baseline PT trips, under the implicit assumption that the initial and final stages of these trips remain unaltered.
-A strength of this _backcasting_ approach is that we can simulate the aggregate travel patterns that would meet particular policy objectives, and explore what the implications could be for the public transport system and provision of safe cycling infrastructure to public transport nodes at regional, local authority and, importantly from a local transport planning practitioner perspective, corridor level.
-
-# Methods
-
-## Case Study
-
 ## Modeling Origin-Destination trips
@@
 # od_jittered_filter = readRDS("paper/paperTRA/load/od_jittered_filter.Rds")
+# MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
 #
@@
 ```
+
+## Modelling car shift to bicycle-PT - here or before??
+
+It is essential to note that our approach is not predictive but scenario-based, providing a range of potential outcomes regarding the target (% of cycling tips) and the routing profile (safe or direct routes), rather than a single forecast.
+These scenarios are rooted in the explicit national cycling targets, assuming that the potential for cycling is proportional to the reduction of car trips.
+Thus said, we do not predict the generation of new trips, but the substitution of car trips by cycling in combination with PT, considering the potential of the first and last mile of these journeys to be made by bicycle.
+Additionally, we do not model changes in the baseline PT trips, under the implicit assumption that the initial and final stages of these trips remain unaltered.
+
+A strength of this _backcasting_ approach is that we can simulate the aggregate travel patterns that would meet particular policy objectives, and explore what the implications could be for the public transport system and provision of safe cycling infrastructure to public transport nodes at regional, local authority and, importantly from a local transport planning practitioner perspective, corridor level.
 
@@
 
-This paper focus primarily on the potential for replacing trips currently made by car with combined cycling and public transport trips, and the associated social and environmental benefits.
+This paper focus primarily on the potential for replacing trips currently made by car with combined cycling and public transport trips, and the associated social and environmental benefits, showcasing the potential impact of facilitating the first-and-last mile to PT interfaces.
 While the results are specific to the Lisbon Metropolitan Area, the approach is designed to be generalizable to other metropolitan areas.
@@
 For instance, we considered only the inter-municipal bus services that accommodate bicycles on board.
-Future research could explore the potential of equipping other regional bus services with this feature and analyse their attractiveness, in conjunction with cycling, to address current car trip demands.
+Future research could explore the potential of equipping other regional bus services with this feature and analyse their attractiveness, combined with cycling, to address current car trip demands.
 
@@
 The approach cannot show exactly which trips are being replaced, in terms of the age of people making the trips, the types of car they drive, and other individual or vehicle level variables.
-We assumed uniformity in the type of vehicle for all replaced car journeys, albeit with variations in the fuels utilized.
+We assumed uniformity in the type of vehicle for all replaced car journeys, albeit with variations in the fuels utilised.
 Additionally, we assumed consistency in the health impacts across all individuals for the replaced car trips.
@@
 Future work could seek to increase the resolution of the results, and move towards a more individual-level analysis, for example by using agent and activity-based models using tools such as MATSim [@MATSim].
-This direction of travel would require substantial resources outside the scope of this study; another way to improve the spatial resolution of results would be to use higher resolution 'subpoints' as the basis of the desire line 'jittering' process.
+This option would require substantial resources outside the scope of this study; another way to improve the spatial resolution of results would be to use higher resolution 'subpoints' as the basis of the desire line 'jittering' process.
 We would also like to explore the potential for probabilistic routing to improve the spatial resolution of the results.
@@
 A further refinement could be to differentiate between 'simpler' no-transfer PT trips (which require less person effort) and more arduous multi-transfer PT trips, using the same methods, by relaxing the assumption of no-transfers between PT modes in the routing engine.
-This would led to a larger number of total daily trips that can be made by cycling + PT, in particular in locations where the PT network is less complex and requires more transfers.
-We could also set a different maximum lenght of cycling stage, extending the bicycle catchment area.
-However, despite the added value of a sensibility analyses, they would also increase the overall complexity, and would have an impact on the explainability of the results to be used by policy makers.
+This would lead to a larger number of total daily trips that can be made by cycling + PT, in particular in locations where the PT network is less complex and requires more transfers.
+We could also set a different maximum length of cycling stage, extending the bicycle catchment area.
+However, despite the added value of a sensitivity analyses, they would also increase the overall complexity, and would have an impact on the explainability of the results to be used by policy makers.
 

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
case study and methods structrure

@@
             pdfauthor={},
-            pdftitle={Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from biclaR},
+            pdftitle={Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool},
             colorlinks=true,
@@
 
-  \title{Modeling the impacts of replacing car trips with combined
-public transport and cycling: Reproducible methods, results and
-actionable evidence from biclaR}
+  \title{Reproducible methods for modeling combined public transport and
+cycling trips and associated benefits: evidence from the biclaR tool}
     \author[CERIS]{Rosa Félix\corref{cor1}%
@@
   a metropolitan cycling network was developed (\emph{biclaR}). The
-  social and environmental impacts were assessed using the \emph{HEAT
-  for Cycling} and the \emph{HEAT as a Service} tools. The impacts of
-  shifting car trips to public transport were also estimated and
-  monetized. The results show that 10\% of all trips could be made by
-  cycling in combination with public transport. Shifting to cycling for
-  the shorter first and last mile stages can reduce annual
-  CO\textsubscript{2}eq emissions from 3,000 to 7,500 tons/day, while
-  for the public transport leg, the transfer from car avoids of up to
-  20,500 tons of CO\textsubscript{2}eq emissions per year. The estimated
-  socio-environmental benefits are of €125 million to €325 million over
-  10 years. This evidence can support policymakers to prioritize
-  interventions that reduce the reliance on private motor vehicles.\\
+  social and environmental impacts were assessed using Health World
+  Organization tools. The impacts of shifting car trips to public
+  transport were also estimated and monetized. The results show that
+  10\% of all trips could be made by cycling in combination with public
+  transport. Shifting to cycling for the shorter first and last mile
+  stages can reduce annual CO\textsubscript{2}eq emissions from 3,000 to
+  7,500 tons/day, while for the public transport leg, the transfer from
+  car avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per
+  year. The estimated socio-environmental benefits are of €125 million
+  to €325 million over 10 years. This evidence can support policymakers
+  to prioritize interventions that reduce the reliance on private motor
+  vehicles.\\
   \end{abstract}
@@
 modal share of cycling is low, but the potential for cycling as a
-complementary mode of PT is high. According to the latest mobility
-survey conducted in 2018 \citep{IMOB}, the LMA registered a total of 5.3
-million daily trips, with only 0.5\% by bicycle. Car modal share was
-58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
-The number of intra-municipal trips --- with origin and destination in
-the same municipality --- amounts to 3.5 million trips. This exceeds the
-number of inter-municipal trips (1.8 million trips), involving travel
-between different municipalities. Cars and public transport are the most
-used modes for intercity trips, with cars being the predominant choice
-for all journeys. 53\% trips are up to 5 km distance, and 71\% up to 10
-km. Nevertheless, 29\% of trips are longer than 10 km, which requires
-the use of motorized modes, or active modes in combination with public
-transport.
-
-Regarding cycling trips in the LMA, 55\% are up to 5 km, and 88\% are up
-to 10 km. These values are not in line with the typically found in
-cycling distance decay curves, which show a high percentage of trips are
-up to 5 km (\textasciitilde75\%), and a smaller proportion up to 10 km
-\citep{krizek2007detailed, larsen2010beyond, DDfunction2023}. This may
-be due to the low cycling modal share in the LMA - even more limited in
-the sample survey, which may be biased towards shorter trips.
-
-\begin{figure}
-\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
-\end{figure}
-
-\textbf{TO-DO}
-
-\begin{itemize}
-\tightlist
-\item
-  mais mapas:
-\item
-  Densidade populational - nº de viagens
-\item
-  Pares OD - desire lines width
-\end{itemize}
+complementary mode of PT is high.
 
@@
 
-It is essential to note that our approach is not predictive but
-scenario-based, providing a range of potential outcomes regarding the
-target (\% of cycling tips) and the routing profile (safe or direct
-routes), rather than a single forecast. These scenarios are rooted in
-the explicit national cycling targets, assuming that the potential for
-cycling is proportional to the reduction of car trips. Thus said, we do
-not predict the generation of new trips, but the substitution of car
-trips by cycling in combination with PT, considering the potential of
-the first and last mile of these journeys to be made by bicycle.
-Additionally, we do not model changes in the baseline PT trips, under
-the implicit assumption that the initial and final stages of these trips
-remain unaltered. A strength of this \emph{backcasting} approach is that
-we can simulate the aggregate travel patterns that would meet particular
-policy objectives, and explore what the implications could be for the
-public transport system and provision of safe cycling infrastructure to
-public transport nodes at regional, local authority and, importantly
-from a local transport planning practitioner perspective, corridor
-level.
-
 \section{Methods}\label{methods}
@@
 \subsection{Case Study}\label{case-study}
+
+According to the latest mobility survey conducted in 2018 \citep{IMOB},
+the LMA registered a total of 5.3 million daily trips, with only 0.5\%
+by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%
+(see Figure \ref{fig:mododist}). The number of intra-municipal trips ---
+with origin and destination in the same municipality --- amounts to 3.5
+million trips. This exceeds the number of inter-municipal trips (1.8
+million trips), involving travel between different municipalities. Cars
+and public transport are the most used modes for intercity trips, with
+cars being the predominant choice for all journeys. 53\% trips are up to
+5 km distance, and 71\% up to 10 km. Nevertheless, 29\% of trips are
+longer than 10 km, which requires the use of motorized modes, or active
+modes in combination with public transport.
+
+Regarding cycling trips in the LMA, 55\% are up to 5 km, and 88\% are up
+to 10 km. These values are not in line with the typically found in
+cycling distance decay curves, which show a high percentage of trips are
+up to 5 km (\textasciitilde75\%), and a smaller proportion up to 10 km
+\citep{krizek2007detailed, larsen2010beyond, DDfunction2023}. This may
+be due to the low cycling modal share in the LMA - even more limited in
+the sample survey, which may be biased towards shorter trips.
+
+\begin{figure}
+\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
+\end{figure}
+
+\textbf{TO-DO}
+
+\begin{itemize}
+\tightlist
+\item
+  mais mapas:
+\item
+  Densidade populational - nº de viagens
+\item
+  Pares OD - desire lines width
+\end{itemize}
 
@@
 \end{figure}
+
+\subsection{Modelling car shift to bicycle-PT - here or
+before??}\label{modelling-car-shift-to-bicycle-pt---here-or-before}
+
+It is essential to note that our approach is not predictive but
+scenario-based, providing a range of potential outcomes regarding the
+target (\% of cycling tips) and the routing profile (safe or direct
+routes), rather than a single forecast. These scenarios are rooted in
+the explicit national cycling targets, assuming that the potential for
+cycling is proportional to the reduction of car trips. Thus said, we do
+not predict the generation of new trips, but the substitution of car
+trips by cycling in combination with PT, considering the potential of
+the first and last mile of these journeys to be made by bicycle.
+Additionally, we do not model changes in the baseline PT trips, under
+the implicit assumption that the initial and final stages of these trips
+remain unaltered.
+
+A strength of this \emph{backcasting} approach is that we can simulate
+the aggregate travel patterns that would meet particular policy
+objectives, and explore what the implications could be for the public
+transport system and provision of safe cycling infrastructure to public
+transport nodes at regional, local authority and, importantly from a
+local transport planning practitioner perspective, corridor level.
 
@@
 currently made by car with combined cycling and public transport trips,
-and the associated social and environmental benefits. While the results
-are specific to the Lisbon Metropolitan Area, the approach is designed
-to be generalizable to other metropolitan areas.
+and the associated social and environmental benefits, showcasing the
+potential impact of facilitating the first-and-last mile to PT
+interfaces. While the results are specific to the Lisbon Metropolitan
+Area, the approach is designed to be generalizable to other metropolitan
+areas.
 
@@
 potential of equipping other regional bus services with this feature and
-analyse their attractiveness, in conjunction with cycling, to address
-current car trip demands.
+analyse their attractiveness, combined with cycling, to address current
+car trip demands.
 
@@
 of vehicle for all replaced car journeys, albeit with variations in the
-fuels utilized. Additionally, we assumed consistency in the health
+fuels utilised. Additionally, we assumed consistency in the health
 impacts across all individuals for the replaced car trips. This approach
@@
 analysis, for example by using agent and activity-based models using
-tools such as MATSim \citep{MATSim}. This direction of travel would
-require substantial resources outside the scope of this study; another
-way to improve the spatial resolution of results would be to use higher
+tools such as MATSim \citep{MATSim}. This option would require
+substantial resources outside the scope of this study; another way to
+improve the spatial resolution of results would be to use higher
 resolution `subpoints' as the basis of the desire line `jittering'
@@
 assumption of no-transfers between PT modes in the routing engine. This
-would led to a larger number of total daily trips that can be made by
+would lead to a larger number of total daily trips that can be made by
 cycling + PT, in particular in locations where the PT network is less
 complex and requires more transfers. We could also set a different
-maximum lenght of cycling stage, extending the bicycle catchment area.
-However, despite the added value of a sensibility analyses, they would
+maximum length of cycling stage, extending the bicycle catchment area.
+However, despite the added value of a sensitivity analyses, they would
 also increase the overall complexity, and would have an impact on the

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
doi and urls bib

@@
 
-Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326; @van2021insights].
+Combining public transportation (PT) and cycling for the first and last mile in metropolitan areas can replace a high proportion of private car trips [@MARTENS2007326; @vanmil_insights_2021].
 In The Netherlands, which has the highest mode share of cycling in the world, cycling accounts for more than a third of all trips to and from rail stations at the 'home' end of the journey, greatly increasing the ability of the transport system [@RIETVELD200071].

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
maps pop and trips

@@
 library(ggplot2)
+library(sf)
 ```
@@
 
+The Lisbon metropolitan area (LMA), in Portugal, is the case study, comprising 18 municipalities, with a total area of 3,015 km² and a population of 2.8 million residents
+The LMA is the most populous and economically active region in Portugal, with the city of Lisbon as the central node of the LMA, with a high concentration of jobs and services, and a high demand for mobility (see Figure \ref{fig:lma}).
+As shown, the number of trips is proportional to the population of each municipality, with the highest number of trips originating in Lisbon, followed by the municipalities of Sintra, Cascais, Loures, and Almada.
+
+The Public Transport network is well developed, although very much Lisbon-centric, with a high level of service, including inter-municipal trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a metro system that serves Lisbon and surrounding municipalities.
+
+```{r lma, fig.cap="\\label{lma}Population and daily trips (origins) at each Municipality of the LMA according to census and the travel survey.", fig.subcap=c("Population in the LMA","Trips/day in the LMA"), fig.ncol=2, message=FALSE, warning=FALSE, out.width="50%"}
+MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
+MUNICIPIOSgeo_mapa = readRDS("load/MUNICIPIOSgeo_mapa.Rds")
+MUNICIPIOS_dtcc = readRDS("load/MUNICIPIOS_dtcc.Rds")
+POPMunicipios = readRDS("load/POPMunicipios.Rds")
+
+# labels
+cities = MUNICIPIOSgeo |> 
+  filter(Concelho %in% c("Lisboa", "Sintra", "Cascais", "Loures", "Almada"))
+cities$point = sf::st_centroid(cities$geometry)
+cities = sf::st_set_geometry(cities, cities$point)
+
+# population
+pop_mun = POPMunicipios |> 
+  left_join(MUNICIPIOS_dtcc, by = c("DTCC" = "DTMN"))
+pop_mun$Concelho[pop_mun$Concelho == "Setúbal"] = "Setubal"
+pop_mun = MUNICIPIOSgeo |>
+  left_join(pop_mun, by = "Concelho")
+pop_mun$area = sf::st_area(pop_mun) |> units::drop_units()
+pop_mun$popdens = pop_mun$Populacao / pop_mun$area * 1000000
+
+# residents
+  ggplot() +
+  geom_sf(data = pop_mun,
+          aes(fill = Populacao),
+          color = "gray60") +
+  scale_fill_gradient(low = "white", high = "darkred",
+                      labels = scales::unit_format(unit = "k", scale = 1e-3)
+                      ) +
+  theme_void() +
+  # theme(legend.position = "bottom")+
+  labs(fill = "Residents") +
+  geom_sf_text(data = cities,
+               aes(label = Concelho),
+               size = 2.3)
+
+# #denstity
+# ggplot() +
+#   geom_sf(data = pop_mun,
+#           aes(fill = popdens),
+#           color = "gray60") +
+#   scale_fill_gradient(low = "white", high = "darkred") +
+#   theme_void() +
+#   labs(fill = "inhabitants/km²")
+
+# trips
+ggplot() +
+  geom_sf(data = MUNICIPIOSgeo_mapa,
+          aes(fill = viagens),
+          color = "gray60") +
+  viridis::scale_fill_viridis(option = "F", direction = -1,
+                              labels = scales::unit_format(unit = "k", scale = 1e-3)) +
+  theme_void() +
+  labs(fill = "Trips / day")
+
+```
+
+The LMA is a low cycling maturity area, characterised by a sparse and fragmented cycling network totalling 378 km, primarily concentrated within Lisbon, and a low cycling modal share.
 According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.
-Car modal share was 58.4%, while PT accounted for 15.5% (see Figure \ref{fig:mododist}).
+
+The survey also showed that 58.4% of trips were made by car, 15.5% by public transport, and 23.9% by walking (see Figure \ref{fig:mododist}).
 The number of intra-municipal trips --- with origin and destination in the same municipality --- amounts to 3.5 million trips.
@@
 Nevertheless, 29% of trips are longer than 10 km, which requires the use of motorized modes, or active modes in combination with public transport.
-
-Regarding cycling trips in the LMA, 55% are up to 5 km, and 88% are up to 10 km.
-These values are not in line with the typically found in cycling distance decay curves, which show a high percentage of trips are up to 5 km (~75%), and a smaller proportion up to 10 km [@krizek2007detailed; @larsen2010beyond; @DDfunction2023].
-This may be due to the low cycling modal share in the LMA - even more limited in the sample survey, which may be biased towards shorter trips.
 
@@
 
-**TO-DO**
-
--   mais mapas:
--   Densidade populational - nº de viagens
--   Pares OD - desire lines width
+Regarding cycling trips in the LMA, 55% are up to 5 km, and 88% are up to 10 km.
+These values are not in line with the typically found in cycling distance decay curves, which show a high percentage of trips are up to 5 km (~75%), and a smaller proportion up to 10 km [@krizek2007detailed; @larsen2010beyond; @DDfunction2023].
+This may be due to the low cycling modal share in the LMA - even more limited in the sample survey, which may be biased towards shorter trips.
 
@@
 Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.
+As shown, the city of Lisbon (in the centre) is the main attractor of trips, with a high number of trips to and from the other municipalities.
 
@@
 
-MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
 stopsAML = sf::st_read("load/Stops_AML.geojson", quiet = TRUE)

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
maps pop and trips

@@
 mile in metropolitan areas can replace a high proportion of private car
-trips \citep{MARTENS2007326, van2021insights}. In The Netherlands, which
-has the highest mode share of cycling in the world, cycling accounts for
-more than a third of all trips to and from rail stations at the `home'
-end of the journey, greatly increasing the ability of the transport
-system \citep{RIETVELD200071}. This approach to reducing car dependency
-and associated externalities requires interventions and programs to make
-bicycling more appealing \citep{lapaix_role_2021}. The resulting public
-investments can have significant social and environmental benefits
+trips \citep{MARTENS2007326, vanmil_insights_2021}. In The Netherlands,
+which has the highest mode share of cycling in the world, cycling
+accounts for more than a third of all trips to and from rail stations at
+the `home' end of the journey, greatly increasing the ability of the
+transport system \citep{RIETVELD200071}. This approach to reducing car
+dependency and associated externalities requires interventions and
+programs to make bicycling more appealing \citep{lapaix_role_2021}. The
+resulting public investments can have significant social and
+environmental benefits
 \citep{internationaltransportforum_integrating_2017}. Despite the
@@
 
-According to the latest mobility survey conducted in 2018 \citep{IMOB},
-the LMA registered a total of 5.3 million daily trips, with only 0.5\%
-by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%
-(see Figure \ref{fig:mododist}). The number of intra-municipal trips ---
-with origin and destination in the same municipality --- amounts to 3.5
-million trips. This exceeds the number of inter-municipal trips (1.8
-million trips), involving travel between different municipalities. Cars
-and public transport are the most used modes for intercity trips, with
-cars being the predominant choice for all journeys. 53\% trips are up to
-5 km distance, and 71\% up to 10 km. Nevertheless, 29\% of trips are
-longer than 10 km, which requires the use of motorized modes, or active
-modes in combination with public transport.
+The Lisbon metropolitan area (LMA), in Portugal, is the case study,
+comprising 18 municipalities, with a total area of 3,015 km² and a
+population of 2.8 million residents The LMA is the most populous and
+economically active region in Portugal, with the city of Lisbon as the
+central node of the LMA, with a high concentration of jobs and services,
+and a high demand for mobility (see Figure \ref{fig:lma}). As shown, the
+number of trips is proportional to the population of each municipality,
+with the highest number of trips originating in Lisbon, followed by the
+municipalities of Sintra, Cascais, Loures, and Almada.
+
+The Public Transport network is well developed, although very much
+Lisbon-centric, with a high level of service, including inter-municipal
+trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a
+metro system that serves Lisbon and surrounding municipalities.
+
+\begin{figure}
+\subfloat[Population in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA according to census and the travel survey.}\label{fig:lma}
+\end{figure}
+
+The LMA is a low cycling maturity area, characterised by a sparse and
+fragmented cycling network totalling 378 km, primarily concentrated
+within Lisbon, and a low cycling modal share. According to the latest
+mobility survey conducted in 2018 \citep{IMOB}, the LMA registered a
+total of 5.3 million daily trips, with only 0.5\% by bicycle.
+
+The survey also showed that 58.4\% of trips were made by car, 15.5\% by
+public transport, and 23.9\% by walking (see Figure \ref{fig:mododist}).
+The number of intra-municipal trips --- with origin and destination in
+the same municipality --- amounts to 3.5 million trips. This exceeds the
+number of inter-municipal trips (1.8 million trips), involving travel
+between different municipalities. Cars and public transport are the most
+used modes for intercity trips, with cars being the predominant choice
+for all journeys. 53\% trips are up to 5 km distance, and 71\% up to 10
+km. Nevertheless, 29\% of trips are longer than 10 km, which requires
+the use of motorized modes, or active modes in combination with public
+transport.
+
+\begin{figure}
+\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
+\end{figure}
 
@@
 the sample survey, which may be biased towards shorter trips.
-
-\begin{figure}
-\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
-\end{figure}
-
-\textbf{TO-DO}
-
-\begin{itemize}
-\tightlist
-\item
-  mais mapas:
-\item
-  Densidade populational - nº de viagens
-\item
-  Pares OD - desire lines width
-\end{itemize}
 
@@
 trips between districts, specifically for the Lisbon metropolitan area.
+As shown, the city of Lisbon (in the centre) is the main attractor of
+trips, with a high number of trips to and from the other municipalities.
 

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
beef spatial results and re-structure conclusions

@@
 
+It has become progressively more common to establish strategic plans, at national, regional or municipal level, to mitigate climate change.
+Among these, the Sustainable Urban Mobility Plans (SUMP[^10]), promoted by the European Commission, are becoming popular in Europe, although authorities are designing documents of this sort all over the world.
+The definition of targets associated with a timeframe for reducing dependence on the individual motorized vehicle, or targets for the use of active modes such as walking and cycling, are too often not accompanied by estimates of their social, environmental and economic impacts.
+It is important for authorities and practitioners to know how to estimate those impacts, which tools are available to support them in the process, and what results to expect.
+
+[^10]: See [eltis.org/mobility-plans/sump-concept](https://www.eltis.org/mobility-plans/sump-concept)
+
 This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
@@
 The LMA is the most populous and economically active region in Portugal, with the city of Lisbon as the central node of the LMA, with a high concentration of jobs and services, and a high demand for mobility (see Figure \ref{fig:lma}).
+
 As shown, the number of trips is proportional to the population of each municipality, with the highest number of trips originating in Lisbon, followed by the municipalities of Sintra, Cascais, Loures, and Almada.
-
 The Public Transport network is well developed, although very much Lisbon-centric, with a high level of service, including inter-municipal trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a metro system that serves Lisbon and surrounding municipalities.
 
-```{r lma, fig.cap="\\label{lma}Population and daily trips (origins) at each Municipality of the LMA according to census and the travel survey.", fig.subcap=c("Population in the LMA","Trips/day in the LMA"), fig.ncol=2, message=FALSE, warning=FALSE, out.width="50%"}
+```{r lma, fig.cap="\\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to census and the travel survey.", fig.subcap=c("Population distribution in the LMA","Trips/day in the LMA"), fig.ncol=2, message=FALSE, warning=FALSE, out.width="50%"}
 MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
@@
 cities = MUNICIPIOSgeo |> 
-  filter(Concelho %in% c("Lisboa", "Sintra", "Cascais", "Loures", "Almada"))
+  filter(Concelho %in% c("Lisboa", "Sintra", "Cascais", "Loures", "Almada", "Vila Franca de Xira", "Seixal", "Setubal", "Oeiras", "Mafra", "Amadora"))
 cities$point = sf::st_centroid(cities$geometry)
@@
 Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
+This map shows that some of the municipalities are not served by PT modes that can easily accommodate bicycles, which may limit the potential for intermodality in these areas.
 
@@
 
-Figure \ref{fig:map2} illustrates the resulting bicycle routes to access the main PT interfaces in the LMA.
+Figure \ref{fig:map2} illustrates the routes with greatest cycling potential to access nodes on the public transport network.
+The interactive version of the map is publicly available in the *biclaR* tool, allowing detailed exploration of local areas.
+At *biclaR*, users and local authority practitioners have the option to download the results as a spatial dataset for further analysis concerning the prioritization of cycling interventions.
 It is noticeable the high potential of the train interfaces to attract car-to-PT substituting trips.
@@
 
-## Modelling car shift to bicycle-PT - here or before??
+## Modelling car shift to bicycle-PT
 
@@
 
-Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new daily trips to achieve the national strategy targets (4% and 10%), for different route profiles.
+Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new daily trips to achieve the national strategy targets (4% and 10%), for different route profiles, at an aggregate level of the LMA.
 For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in Table \ref{tab:summary1b}, resulting from replacing car trips with cycling.
@@
 
-<!-- INSERT ZONAL MAP HERE -->
+Another relevant aspect is that the higher potential (Figure \ref{fig:mapzones}.a) is in the municipalities with the highest number of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and Almada, which are also the ones with the highest number of PT interfaces (Figure \ref{fig:map1}).
+On another hand, the municipalities with little PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a lower potential for car to Bike-PT substitution (Figure \ref{fig:mapzones}.a), even if they have a high number of trips, such as Setubal.
+This findings are aligned with their local travel patterns, where the majority of trips are intra-municipal (67% for both municipalities), and not to the city of Lisbon, despite the high car-dependency (79% in Mafra, 67% in Setubal).
 
-```{r mapzones, eval=FALSE, fig.cap="Potential Bike + PT daily trips for the 10% target, by municipality"}
+The potential and the resulting benefits are spatially unevenly distributed across the LMA, as shown in Figure \ref{fig:mapzones}.b, concentrating 40% in Lisbon, where the highest number of trips are made.
+
+```{r mathmapzones, warning=FALSE, message=FALSE}
 CAR_PT_TRANSF = readRDS("load/CAR_TP_TRANSF.Rds")
+FREGUESIASgeo_zones = readRDS("load/FREGUESIASgeo_zones.Rds")
+TRIPSall_mun = readRDS("load/TRIPSall_mun.Rds")
 
@@
          TP_4per = 100*TP_4 / viagens,
-         TP_10per = 100*TP_10 / viagens)
+         TP_10per = 100*TP_10 / viagens) |> 
+  mutate(new_perc = 100*Bike_new10 / 290686.9) #sum(CAR_PT_TRANSF_mun_or_geo_18$Bike_new10, na.rm = T)
 
@@
          TP_4per = 100*TP_4 / Total,
-         TP_10per = 100*TP_10 / Total)
+         TP_10per = 100*TP_10 / Total) |> 
+  mutate(new_perc = 100*Bike_new10 / 50081.15) #sum(CAR_PT_TRANSF_mun_or_geo_18_extrap$Bike_new10, na.rm = T))
 
+cities2 = MUNICIPIOSgeo |> 
+  filter(Concelho %in% c("Lisboa", "Sintra", "Cascais", "Loures", "Almada", "Oeiras", "Amadora", "Seixal", "Vila Franca de Xira"))
+cities2$point = sf::st_centroid(cities2$geometry)
+cities2 = sf::st_set_geometry(cities2, cities2$point)
+cities2 = cities2 |> left_join(CAR_PT_TRANSF_mun_or_geo_18_extrap |>
+                                 sf::st_drop_geometry() |> 
+                                 select(Concelho, new_perc),
+                               by = "Concelho") |> 
+  mutate(new_perc_ext = round(new_perc)) |> 
+  select(-new_perc) |> 
+  left_join(CAR_PT_TRANSF_mun_or_geo_18 |>
+                                 sf::st_drop_geometry() |> 
+                                 select(Concelho, new_perc),
+                               by = "Concelho") |> 
+  mutate(new_perc = round(new_perc))
+
+```
+
+```{r mapzones, fig.cap="\\label{mapzones}Potential Bicycle + PT daily trips for the 10% target and direct routing, by municipality.
+The percentages are shown for municipalities with values above 4%.", warning=FALSE, fig.subcap=c("Potential new Bicycle + PT trips","Proportion of total potential modal shift in the LMA"), fig.ncol=2, message=FALSE, out.width="50%"}
 # plot 1
 ggplot(data = CAR_PT_TRANSF_mun_or_geo_18_extrap) +
-  geom_sf(aes(fill = Bike_new10)) +
-  viridis::scale_fill_viridis(option = "G", direction = -1, name = "Bike + PT trips", na.value = "white") +
+  geom_sf(aes(fill = Bike10)) +
+  viridis::scale_fill_viridis(option = "G",
+                              direction = -1,
+                              name = "Potential new \nBike + PT\ntrips/day",
+                              na.value = "white",
+                              labels = scales::unit_format(unit = "k", scale = 1e-3)) +
+
   theme_void()
+
   # theme(legend.position = "bottom") +
@@
 
-# plot 2 - the total % of potential bike + TP trips
+# # plot 2 - the total % of potential bike + TP trips
+# ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
+#   geom_sf(aes(fill = TP_10per)) +
+#   viridis::scale_fill_viridis(option = "G", direction = -1, name = "Bike + PT trips %", na.value = "white") +
+#   theme_void()
+# 
+# # plot 3 - the added % of potential bike + TP trips - doesn't make sense to show
+# ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
+#   geom_sf(aes(fill = TP_10per- TPper)) + #???
+#   viridis::scale_fill_viridis(option = "G", direction = -1, name = "new Bike + PT trips %", na.value = "white") +
+#   theme_void()
+
+# plot 3 - the % of potential bike + TP trips in the LMA context
 ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
-  geom_sf(aes(fill = TP_10per)) +
-  viridis::scale_fill_viridis(option = "G", direction = -1, name = "Bike + PT trips %", na.value = "white") +
-  theme_void()
-
-# plot 3 - the added % of potential bike + TP trips - doesn't make sense to show
-ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
-  geom_sf(aes(fill = TP_10per- TPper)) + #???
-  viridis::scale_fill_viridis(option = "G", direction = -1, name = "new Bike + PT trips %", na.value = "white") +
-  theme_void()
-
+  geom_sf(aes(fill = new_perc)) + #???
+  scale_fill_gradient(low = "white", high = "darkslateblue", name = "% of total \npotential\nmodal shift", na.value = "white") +
+  theme_void() +
+  geom_sf_text(data = cities2,
+    aes(label = paste0(new_perc, "%")),
+    size = 2.7,
+    color = ifelse(cities2$Concelho == "Lisboa", "white", "black"),
+    check_overlap = T)
 ```
-
-The benefits are spatially unevenly distributed across the LMA, as shown in Figure \ref{fig:mapzones}.
 
@@
 Table \ref{tab:summary22} presents emissions reductions and associated economic benefits associated with the second (PT) leg of trips.
-The shift from private car associated with thes PT segments would reduce CO~2~ equivalent emissions by 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly, for the 4% and 10% targets, respectively.
+The shift from private car associated with the PT segments would reduce CO~2~ equivalent emissions by 8,500 to 20,800 tons annually, valued in €1.4 million to €3.5 million yearly, for the 4% and 10% targets, respectively.
 
@@
 
-Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions.
+Our findings show that cycling *in combination* with PT could replace 10% of current LMA trips, with an additional 6% of PT journeys prone to further substitution, based on conservative assumptions, highlighting the municipalities of Lisbon, Sintra, Almada, Amadora, and Oeiras as the ones with the highest potential for car to bike-PT substitution.
 Although this paper does not address the designing specific solutions, our findings indicate a potential surge in demand (+6.3%) of public transport interfaces to align with the national strategies.
@@
 
-This paper estimated the potential for combining cycling and PT to substitute car trips in the LMA, while achieving the national cycling targets and supporting decarbonization goals.
-It has become progressively more common to establish strategic plans, at national, regional or municipal level, to mitigate climate change.
-Among these, the Sustainable Urban Mobility Plans (SUMP[^10]), promoted by the European Commission, are becoming popular in Europe, although authorities are designing documents of this sort all over the world.
-The definition of targets associated with a timeframe for reducing dependence on the individual motorized vehicle, or targets for the use of active modes such as walking and cycling, are too often not accompanied by estimates of their social, environmental and economic impacts.
-It is important for authorities and practitioners to know how to estimate those impacts, which tools are available to support them in the process, and what results to expect.
-
-This paper quantifies the benefits of replacing car trips with cycling, in combination with public transport.
+This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA, while achieving the national cycling targets and supporting decarbonization goals.
 The case study of the Lisbon metropolitan area demonstrates that cycling-PT integration can help meet the national targets set for bicycle use of 4% and 10% by 2025 and 2030, respectively.
 
-[^10]: See [eltis.org/mobility-plans/sump-concept](https://www.eltis.org/mobility-plans/sump-concept)
+This research also quantifies the socio-environmental benefits of achieving such targets, exploring an intermodality scenario, where car trips are potentially substituted by bicycle in combination with PT.
 
-The quantification of benefits can support policy-makers in prioritizing interventions to reduce the reliance on private motorized modes of transportation.
+The findings indicate that cycling combined with public transport has the potential to account for 10% of total trips.
+Transitioning to cycling for the first and last segments could lead to a decrease in annual CO~2~eq emissions by a substantial margin, ranging from 3,000 to 7,500 tons per day.
+Moreover, for the second trip segment, the shift from car to public transport contributes to avoiding up to 20,500 tons of CO~2~eq emissions annually.
+These changes are estimated to yield socio-environmental benefits totaling €125 million to €325 million over a decade.
+The quantification of such benefits can support policy-makers in prioritizing interventions to reduce the reliance on private motorized modes of transportation.
+
+Additionally, the results suggest that the potential for cycling-PT intermodality is spatially unevenly distributed across the LMA, with the highest potential in the municipalities with the highest number of trips and PT interfaces of PT modes connecting to Lisbon and where it is possible to carry bicycles on board, such as Lisbon, Sintra, Amadora, and Almada.
+
+We opted to detail all the steps of the methodology, from data requirements to the final results, to ensure transparency and generalization of the approach.
+We acknowledge the limitations of the used methods and propose avenues for future research.
+
 The presentation of the results in an open access web application will help to inform and explain decisions.

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
beef spatial results and re-structure conclusions

@@
 
+It has become progressively more common to establish strategic plans, at
+national, regional or municipal level, to mitigate climate change. Among
+these, the Sustainable Urban Mobility Plans (SUMP\footnote{See
+  \href{https://www.eltis.org/mobility-plans/sump-concept}{eltis.org/mobility-plans/sump-concept}}),
+promoted by the European Commission, are becoming popular in Europe,
+although authorities are designing documents of this sort all over the
+world. The definition of targets associated with a timeframe for
+reducing dependence on the individual motorized vehicle, or targets for
+the use of active modes such as walking and cycling, are too often not
+accompanied by estimates of their social, environmental and economic
+impacts. It is important for authorities and practitioners to know how
+to estimate those impacts, which tools are available to support them in
+the process, and what results to expect.
+
 This paper estimates the potential for combining cycling and PT to
@@
 central node of the LMA, with a high concentration of jobs and services,
-and a high demand for mobility (see Figure \ref{fig:lma}). As shown, the
-number of trips is proportional to the population of each municipality,
-with the highest number of trips originating in Lisbon, followed by the
-municipalities of Sintra, Cascais, Loures, and Almada.
+and a high demand for mobility (see Figure \ref{fig:lma}).
 
+As shown, the number of trips is proportional to the population of each
+municipality, with the highest number of trips originating in Lisbon,
+followed by the municipalities of Sintra, Cascais, Loures, and Almada.
 The Public Transport network is well developed, although very much
@@
 \begin{figure}
-\subfloat[Population in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA according to census and the travel survey.}\label{fig:lma}
+\subfloat[Population distribution in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to census and the travel survey.}\label{fig:lma}
 \end{figure}
@@
 and inter-municipal bus lines equipped with bike racks (Figure
-\ref{fig:map1}).
+\ref{fig:map1}). This map shows that some of the municipalities are not
+served by PT modes that can easily accommodate bicycles, which may limit
+the potential for intermodality in these areas.
 
@@
 
-Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
-the main PT interfaces in the LMA. It is noticeable the high potential
-of the train interfaces to attract car-to-PT substituting trips.
+Figure \ref{fig:map2} illustrates the routes with greatest cycling
+potential to access nodes on the public transport network. The
+interactive version of the map is publicly available in the
+\emph{biclaR} tool, allowing detailed exploration of local areas. At
+\emph{biclaR}, users and local authority practitioners have the option
+to download the results as a spatial dataset for further analysis
+concerning the prioritization of cycling interventions. It is noticeable
+the high potential of the train interfaces to attract car-to-PT
+substituting trips.
 
@@
 
-\subsection{Modelling car shift to bicycle-PT - here or
-before??}\label{modelling-car-shift-to-bicycle-pt---here-or-before}
+\subsection{Modelling car shift to
+bicycle-PT}\label{modelling-car-shift-to-bicycle-pt}
 
@@
 daily trips to achieve the national strategy targets (4\% and 10\%), for
-different route profiles. For the cycling legs of the journey (first and
-last legs), the environmental avoided emissions and monetized
-socio-environment (SE) benefits are presented in Table
-\ref{tab:summary1b}, resulting from replacing car trips with cycling.
+different route profiles, at an aggregate level of the LMA. For the
+cycling legs of the journey (first and last legs), the environmental
+avoided emissions and monetized socio-environment (SE) benefits are
+presented in Table \ref{tab:summary1b}, resulting from replacing car
+trips with cycling.
 
@@
 
-The benefits are spatially unevenly distributed across the LMA, as shown
-in Figure \ref{fig:mapzones}.
+Another relevant aspect is that the higher potential (Figure
+\ref{fig:mapzones}.a) is in the municipalities with the highest number
+of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and
+Almada, which are also the ones with the highest number of PT interfaces
+(Figure \ref{fig:map1}). On another hand, the municipalities with little
+PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a
+lower potential for car to Bike-PT substitution (Figure
+\ref{fig:mapzones}.a), even if they have a high number of trips, such as
+Setubal. This findings are aligned with their local travel patterns,
+where the majority of trips are intra-municipal (67\% for both
+municipalities), and not to the city of Lisbon, despite the high
+car-dependency (79\% in Mafra, 67\% in Setubal).
+
+The potential and the resulting benefits are spatially unevenly
+distributed across the LMA, as shown in Figure \ref{fig:mapzones}.b,
+concentrating 40\% in Lisbon, where the highest number of trips are
+made.
+
+\begin{figure}
+\subfloat[Potential new Bicycle + PT trips\label{fig:mapzones-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-1} }\subfloat[Proportion of total potential modal shift in the LMA\label{fig:mapzones-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-2} }\caption{\label{mapzones}Potential Bicycle + PT daily trips for the 10\% target and direct routing, by municipality.
+The percentages are shown for municipalities with values above 4\%.}\label{fig:mapzones}
+\end{figure}
 
@@
 economic benefits associated with the second (PT) leg of trips. The
-shift from private car associated with thes PT segments would reduce
+shift from private car associated with the PT segments would reduce
 CO\textsubscript{2} equivalent emissions by 8,500 to 20,800 tons
@@
 replace 10\% of current LMA trips, with an additional 6\% of PT journeys
-prone to further substitution, based on conservative assumptions.
-Although this paper does not address the designing specific solutions,
-our findings indicate a potential surge in demand (+6.3\%) of public
-transport interfaces to align with the national strategies. This added
-demand is conceptually driven from individuals utilising bicycles,
-either personal or shared, which would be parked at the PT stations.
-Consequently, operators and public space managers would face the
-challenge of accommodating this increased demand for bicycle parking
-and/or docks from a bike sharing system, since this availability plays a
-crucial role in the decision-making process of individuals to use
-combined bike-train \citep{jonkeren2021bicycle}. Additionally, they may
-need to explore options to enhance bicycle carriage capacity on public
-transport services, ensuring alignment with national targets.
+prone to further substitution, based on conservative assumptions,
+highlighting the municipalities of Lisbon, Sintra, Almada, Amadora, and
+Oeiras as the ones with the highest potential for car to bike-PT
+substitution. Although this paper does not address the designing
+specific solutions, our findings indicate a potential surge in demand
+(+6.3\%) of public transport interfaces to align with the national
+strategies. This added demand is conceptually driven from individuals
+utilising bicycles, either personal or shared, which would be parked at
+the PT stations. Consequently, operators and public space managers would
+face the challenge of accommodating this increased demand for bicycle
+parking and/or docks from a bike sharing system, since this availability
+plays a crucial role in the decision-making process of individuals to
+use combined bike-train \citep{jonkeren2021bicycle}. Additionally, they
+may need to explore options to enhance bicycle carriage capacity on
+public transport services, ensuring alignment with national targets.
 
@@
 
-This paper estimated the potential for combining cycling and PT to
+This paper estimates the potential for combining cycling and PT to
 substitute car trips in the LMA, while achieving the national cycling
-targets and supporting decarbonization goals. It has become
-progressively more common to establish strategic plans, at national,
-regional or municipal level, to mitigate climate change. Among these,
-the Sustainable Urban Mobility Plans (SUMP\footnote{See
-  \href{https://www.eltis.org/mobility-plans/sump-concept}{eltis.org/mobility-plans/sump-concept}}),
-promoted by the European Commission, are becoming popular in Europe,
-although authorities are designing documents of this sort all over the
-world. The definition of targets associated with a timeframe for
-reducing dependence on the individual motorized vehicle, or targets for
-the use of active modes such as walking and cycling, are too often not
-accompanied by estimates of their social, environmental and economic
-impacts. It is important for authorities and practitioners to know how
-to estimate those impacts, which tools are available to support them in
-the process, and what results to expect.
+targets and supporting decarbonization goals. The case study of the
+Lisbon metropolitan area demonstrates that cycling-PT integration can
+help meet the national targets set for bicycle use of 4\% and 10\% by
+2025 and 2030, respectively.
 
-This paper quantifies the benefits of replacing car trips with cycling,
-in combination with public transport. The case study of the Lisbon
-metropolitan area demonstrates that cycling-PT integration can help meet
-the national targets set for bicycle use of 4\% and 10\% by 2025 and
-2030, respectively.
+This research also quantifies the socio-environmental benefits of
+achieving such targets, exploring an intermodality scenario, where car
+trips are potentially substituted by bicycle in combination with PT.
 
-The quantification of benefits can support policy-makers in prioritizing
-interventions to reduce the reliance on private motorized modes of
-transportation. The presentation of the results in an open access web
-application will help to inform and explain decisions. Furthermore, the
-provision of datasets resulting from this project provides a foundation
-for further research and development of new tools and methods. The
-methods are reproducible and based on open source software, which can be
-applied to other cities and metropolitan areas, supporting the
-decarbonization of transport systems internationally.
+The findings indicate that cycling combined with public transport has
+the potential to account for 10\% of total trips. Transitioning to
+cycling for the first and last segments could lead to a decrease in
+annual CO\textsubscript{2}eq emissions by a substantial margin, ranging
+from 3,000 to 7,500 tons per day. Moreover, for the second trip segment,
+the shift from car to public transport contributes to avoiding up to
+20,500 tons of CO\textsubscript{2}eq emissions annually. These changes
+are estimated to yield socio-environmental benefits totaling €125
+million to €325 million over a decade. The quantification of such
+benefits can support policy-makers in prioritizing interventions to
+reduce the reliance on private motorized modes of transportation.
+
+Additionally, the results suggest that the potential for cycling-PT
+intermodality is spatially unevenly distributed across the LMA, with the
+highest potential in the municipalities with the highest number of trips
+and PT interfaces of PT modes connecting to Lisbon and where it is
+possible to carry bicycles on board, such as Lisbon, Sintra, Amadora,
+and Almada.
+
+We opted to detail all the steps of the methodology, from data
+requirements to the final results, to ensure transparency and
+generalization of the approach. We acknowledge the limitations of the
+used methods and propose avenues for future research.
+
+The presentation of the results in an open access web application will
+help to inform and explain decisions. Furthermore, the provision of
+datasets resulting from this project provides a foundation for further
+research and development of new tools and methods. The methods are
+reproducible and based on open source software, which can be applied to
+other cities and metropolitan areas, supporting the decarbonization of
+transport systems internationally.
 

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
review FM + diff

@@
 This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
-After presenting the methods used, we assess its socio-environmental impacts using open data and open-source tools.
+After presenting the [Methods] used in Chapter 2, we assess its socio-environmental impacts using open data and open-source tools, presenting the [Results and Discussion] in Chapter 3, followed by Chapter 4 where we discuss the [Generalizability and Limitations] of the methods.
+We [Conclude](#Conclusions) the paper in Chapter 5 with the main findings and policy implications.
 
@@
 
-The Lisbon metropolitan area (LMA), in Portugal, is the case study, comprising 18 municipalities, with a total area of 3,015 km² and a population of 2.8 million residents
+The Lisbon metropolitan area (LMA), in Portugal, is the case study, comprising 18 municipalities, with a total area of 3,015 km² and a population of 2.8 million residents, according to the 2021 census [@INEcensus].
 The LMA is the most populous and economically active region in Portugal, with the city of Lisbon as the central node of the LMA, with a high concentration of jobs and services, and a high demand for mobility (see Figure \ref{fig:lma}).
@@
 
-```{r lma, fig.cap="\\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to census and the travel survey.", fig.subcap=c("Population distribution in the LMA","Trips/day in the LMA"), fig.ncol=2, message=FALSE, warning=FALSE, out.width="50%"}
+```{r lma, fig.cap="\\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to the 2021 census and the 2017 travel survey.", fig.subcap=c("Population distribution in the LMA","Trips/day in the LMA"), fig.ncol=2, message=FALSE, warning=FALSE, out.width="50%"}
 MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
@@
 
-```{r mapzones, fig.cap="\\label{mapzones}Potential Bicycle + PT daily trips for the 10% target and direct routing, by municipality.
+```{r mapzones, fig.cap="\\label{mapzones}Potential Bicycle + PT daily trips for the 10% target and direct routing, by municipality (origins).
 The percentages are shown for municipalities with values above 4%.", warning=FALSE, fig.subcap=c("Potential new Bicycle + PT trips","Proportion of total potential modal shift in the LMA"), fig.ncol=2, message=FALSE, out.width="50%"}
@@
 # 
-# # plot 3 - the added % of potential bike + TP trips - doesn't make sense to show
+# # plot 3 - the added % of potential PT
 # ggplot(data = CAR_PT_TRANSF_mun_or_geo_18) +
@@
 
-# Generalizability and limitations
+# Generalizability and Limitations
 
@@
 
-# Conclusion
+# Conclusion {#Conclusions}
 
@@
 The findings indicate that cycling combined with public transport has the potential to account for 10% of total trips.
-Transitioning to cycling for the first and last segments could lead to a decrease in annual CO~2~eq emissions by a substantial margin, ranging from 3,000 to 7,500 tons per day.
+Transitioning to cycling for the first and last segments could lead to a decrease in annual CO~2~eq emissions by a substantial margin, ranging from 3,000 to 7,500 tons per day, depending on the cycling target and routing profile.
 Moreover, for the second trip segment, the shift from car to public transport contributes to avoiding up to 20,500 tons of CO~2~eq emissions annually.
@@
 
-Additionally, the results suggest that the potential for cycling-PT intermodality is spatially unevenly distributed across the LMA, with the highest potential in the municipalities with the highest number of trips and PT interfaces of PT modes connecting to Lisbon and where it is possible to carry bicycles on board, such as Lisbon, Sintra, Amadora, and Almada.
+Additionally, the results suggest that the potential for cycling-PT intermodality is spatially unevenly distributed across the LMA.
+The highest potential occurs in the municipalities with the highest number of trips and PT interfaces with connecting modes to Lisbon that allow for carrying bicycles on board, such as Lisbon, Sintra, Amadora, and Almada.
 
-We opted to detail all the steps of the methodology, from data requirements to the final results, to ensure transparency and generalization of the approach.
+We opted to detail all the steps of the methodology, from data requirements to the final results, to ensure transparency and generalization of the approach, for instance making sensitivity analysis to the bike trip distances and the PT number of transfers.
 We acknowledge the limitations of the used methods and propose avenues for future research.

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
review FM + diff

@@
 This paper estimates the potential for combining cycling and PT to
-substitute car trips in the LMA. After presenting the methods used, we
-assess its socio-environmental impacts using open data and open-source
-tools.
+substitute car trips in the LMA. After presenting the
+\hyperref[methods]{Methods} used in Chapter 2, we assess its
+socio-environmental impacts using open data and open-source tools,
+presenting the \hyperref[results-and-discussion]{Results and Discussion}
+in Chapter 3, followed by Chapter 4 where we discuss the
+\hyperref[generalizability-and-limitations]{Generalizability and
+Limitations} of the methods. We \hyperref[Conclusions]{Conclude} the
+paper in Chapter 5 with the main findings and policy implications.
 
@@
 comprising 18 municipalities, with a total area of 3,015 km² and a
-population of 2.8 million residents The LMA is the most populous and
-economically active region in Portugal, with the city of Lisbon as the
-central node of the LMA, with a high concentration of jobs and services,
-and a high demand for mobility (see Figure \ref{fig:lma}).
+population of 2.8 million residents, according to the 2021 census
+\citep{INEcensus}. The LMA is the most populous and economically active
+region in Portugal, with the city of Lisbon as the central node of the
+LMA, with a high concentration of jobs and services, and a high demand
+for mobility (see Figure \ref{fig:lma}).
 
@@
 \begin{figure}
-\subfloat[Population distribution in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to census and the travel survey.}\label{fig:lma}
+\subfloat[Population distribution in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to the 2021 census and the 2017 travel survey.}\label{fig:lma}
 \end{figure}
@@
 \begin{figure}
-\subfloat[Potential new Bicycle + PT trips\label{fig:mapzones-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-1} }\subfloat[Proportion of total potential modal shift in the LMA\label{fig:mapzones-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-2} }\caption{\label{mapzones}Potential Bicycle + PT daily trips for the 10\% target and direct routing, by municipality.
+\subfloat[Potential new Bicycle + PT trips\label{fig:mapzones-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-1} }\subfloat[Proportion of total potential modal shift in the LMA\label{fig:mapzones-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-2} }\caption{\label{mapzones}Potential Bicycle + PT daily trips for the 10\% target and direct routing, by municipality (origins).
 The percentages are shown for municipalities with values above 4\%.}\label{fig:mapzones}
@@
 \section{Generalizability and
-limitations}\label{generalizability-and-limitations}
+Limitations}\label{generalizability-and-limitations}
 
@@
 
-\section{Conclusion}\label{conclusion}
+\section{Conclusion}\label{Conclusions}
 
@@
 annual CO\textsubscript{2}eq emissions by a substantial margin, ranging
-from 3,000 to 7,500 tons per day. Moreover, for the second trip segment,
-the shift from car to public transport contributes to avoiding up to
-20,500 tons of CO\textsubscript{2}eq emissions annually. These changes
-are estimated to yield socio-environmental benefits totaling €125
-million to €325 million over a decade. The quantification of such
-benefits can support policy-makers in prioritizing interventions to
-reduce the reliance on private motorized modes of transportation.
+from 3,000 to 7,500 tons per day, depending on the cycling target and
+routing profile. Moreover, for the second trip segment, the shift from
+car to public transport contributes to avoiding up to 20,500 tons of
+CO\textsubscript{2}eq emissions annually. These changes are estimated to
+yield socio-environmental benefits totaling €125 million to €325 million
+over a decade. The quantification of such benefits can support
+policy-makers in prioritizing interventions to reduce the reliance on
+private motorized modes of transportation.
 
 Additionally, the results suggest that the potential for cycling-PT
-intermodality is spatially unevenly distributed across the LMA, with the
-highest potential in the municipalities with the highest number of trips
-and PT interfaces of PT modes connecting to Lisbon and where it is
-possible to carry bicycles on board, such as Lisbon, Sintra, Amadora,
-and Almada.
+intermodality is spatially unevenly distributed across the LMA. The
+highest potential occurs in the municipalities with the highest number
+of trips and PT interfaces with connecting modes to Lisbon that allow
+for carrying bicycles on board, such as Lisbon, Sintra, Amadora, and
+Almada.
 
@@
 requirements to the final results, to ensure transparency and
-generalization of the approach. We acknowledge the limitations of the
-used methods and propose avenues for future research.
+generalization of the approach, for instance making sensitivity analysis
+to the bike trip distances and the PT number of transfers. We
+acknowledge the limitations of the used methods and propose avenues for
+future research.
 

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1_word.Rmd
final rev1 CEUS paper
submitted! closes #47
closes #48
closes #49
closes #50

--- 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
@@
 biblio-style: elsarticle-harv
-linenumbers: true
+# linenumbers: true
 numbersections: true
@@
 This paper estimates the potential for combining cycling and PT to substitute car trips in the LMA.
-After presenting the [Methods] used in Chapter 2, we assess its socio-environmental impacts using open data and open-source tools, presenting the [Results and Discussion] in Chapter 3, followed by Chapter 4 where we discuss the [Generalizability and Limitations] of the methods.
-We [Conclude](#Conclusions) the paper in Chapter 5 with the main findings and policy implications.
+After presenting the [Methods] used in Section 2, we assess its socio-environmental impacts using open data and open-source tools, presenting the [Results and Discussion] in Section 3, followed by Section 4 where we discuss the [Generalizability and Limitations] of the methods.
+We [Conclude](#Conclusions) the paper in Section 5 with the main findings and policy implications.
 

### 2024-05-09 paper/PaperCEUS/PaperCEUS_prerev.tex

--- 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
@@
             pdfauthor={},
-            pdftitle={Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool},
+            pdftitle={Modeling the impacts of replacing car trips with combined public transport and cycling: Reproducible methods, results and actionable evidence from biclaR},
             colorlinks=true,
@@
 
-  \title{Reproducible methods for modeling combined public transport and
-cycling trips and associated benefits: evidence from the biclaR tool}
+  \title{Modeling the impacts of replacing car trips with combined
+public transport and cycling: Reproducible methods, results and
+actionable evidence from biclaR}
     \author[CERIS]{Rosa Félix\corref{cor1}%
@@
       \affiliation[CERIS]{CERIS, Instituto Superior Técnico -
-Universidade de Lisboa. Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
+Universidade de Lisboa, Av Rovisco Pais 1, 1049-001 Lisboa, Portugal}
     \affiliation[ITS]{Institute for Transport Studies, University of
@@
   a metropolitan cycling network was developed (\emph{biclaR}). The
-  social and environmental impacts were assessed using Health World
-  Organization tools. The impacts of shifting car trips to public
-  transport were also estimated and monetized. The results show that
-  10\% of all trips could be made by cycling in combination with public
-  transport. Shifting to cycling for the shorter first and last mile
-  stages can reduce annual CO\textsubscript{2}eq emissions from 3,000 to
-  7,500 tons/day, while for the public transport leg, the transfer from
-  car avoids of up to 20,500 tons of CO\textsubscript{2}eq emissions per
-  year. The estimated socio-environmental benefits are of €125 million
-  to €325 million over 10 years. This evidence can support policymakers
-  to prioritize interventions that reduce the reliance on private motor
-  vehicles.\\
+  social and environmental impacts were assessed using the \emph{HEAT
+  for Cycling} and the \emph{HEAT as a Service} tools. The impacts of
+  shifting car trips to public transport were also estimated and
+  monetized. The results show that 10\% of all trips could be made by
+  cycling in combination with public transport. Shifting to cycling for
+  the shorter first and last mile stages can reduce annual
+  CO\textsubscript{2}eq emissions from 3,000 to 7,500 tons/day, while
+  for the public transport leg, the transfer from car avoids of up to
+  20,500 tons of CO\textsubscript{2}eq emissions per year. The estimated
+  socio-environmental benefits are of €125 million to €325 million over
+  10 years. This evidence can support policymakers to prioritize
+  interventions that reduce the reliance on private motor vehicles.\\
   \end{abstract}
@@
 mile in metropolitan areas can replace a high proportion of private car
-trips \citep{MARTENS2007326, vanmil_insights_2021}. In The Netherlands,
-which has the highest mode share of cycling in the world, cycling
-accounts for more than a third of all trips to and from rail stations at
-the `home' end of the journey, greatly increasing the ability of the
-transport system \citep{RIETVELD200071}. This approach to reducing car
-dependency and associated externalities requires interventions and
-programs to make bicycling more appealing \citep{lapaix_role_2021}. The
-resulting public investments can have significant social and
-environmental benefits
+trips \citep{MARTENS2007326}. In The Netherlands, which has the highest
+mode share of cycling in the world, cycling accounts for more than a
+third of all trips to and from rail stations at the `home' end of the
+journey, greatly increasing the ability of the transport system
+\citep{RIETVELD200071}. This approach to reducing car dependency and
+associated externalities requires interventions and programs to make
+bicycling more appealing \citep{lapaix_role_2021}. The resulting public
+investments can have significant social and environmental benefits
 \citep{internationaltransportforum_integrating_2017}. Despite the
@@
 modal share of cycling is low, but the potential for cycling as a
-complementary mode of PT is high.
+complementary mode of PT is high. According to the latest mobility
+survey conducted in 2018 \citep{IMOB}, the LMA registered a total of 5.3
+million daily trips, with only 0.5\% by bicycle. Car modal share was
+58.4\%, while PT accounted for 15.5\% (see Figure \ref{fig:mododist}).
+The number of intra-municipal trips --- with origin and destination in
+the same municipality --- amounts to 3.5 million trips. This exceeds the
+number of inter-municipal trips (1.8 million trips), involving travel
+between different municipalities. Cars and public transport are the most
+used modes for intercity trips, with cars being the predominant choice
+for all journeys.\\
+53\% trips are up to 5 km distance, and 71\% up to 10 km. Nevertheless,
+29\% of trips are longer than 10 km, which requires the use of motorized
+modes, or active modes in combination with public transport.
 
-The Portuguese Government has established a national cycling strategy,
-which outlines targets for the percentage of trips to be made by bicycle
-in urban areas. Specifically, these targets aim for 4\% of all trips in
-2025 and 10\% by 2030 \citep{ENMAC}. Moreover, the strategy emphasizes
-that this increase should be achieved by substituting car trips for
-bicycle trips, as a way to reduce the environmental and social impacts
-of car use.
+\begin{figure}
+\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_prerev_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_prerev_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
+\end{figure}
 
-To support the implementation of this strategy, the Lisbon's
-Metropolitan Department of Transport commissioned
+To achieve the cycling targets set by the Portuguese national cycling
+strategy for 2025 and 2030 (4\% and 10\%, respectively) \citep{ENMAC},
+the Lisbon's Metropolitan Department of Transport commissioned
 \emph{biclaR}\footnote{See
@@
 This paper estimates the potential for combining cycling and PT to
-substitute car trips in the LMA. After presenting the methods used, we
-assess its socio-environmental impacts using open data and open-source
-tools.
+substitute car trips in the LMA, while achieving the national cycling
+targets. After presenting the methods used, it assesses its
+socio-environmental impacts using open data and open-source tools.
 
 \section{Methods}\label{methods}
-
-\subsection{Case Study}\label{case-study}
-
-The Lisbon metropolitan area (LMA), in Portugal, is the case study,
-comprising 18 municipalities, with a total area of 3,015 km² and a
-population of 2.8 million residents The LMA is the most populous and
-economically active region in Portugal, with the city of Lisbon as the
-central node of the LMA, with a high concentration of jobs and services,
-and a high demand for mobility (see Figure \ref{fig:lma}). As shown, the
-number of trips is proportional to the population of each municipality,
-with the highest number of trips originating in Lisbon, followed by the
-municipalities of Sintra, Cascais, Loures, and Almada.
-
-The Public Transport network is well developed, although very much
-Lisbon-centric, with a high level of service, including inter-municipal
-trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a
-metro system that serves Lisbon and surrounding municipalities.
-
-\begin{figure}
-\subfloat[Population in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA according to census and the travel survey.}\label{fig:lma}
-\end{figure}
-
-The LMA is a low cycling maturity area, characterised by a sparse and
-fragmented cycling network totalling 378 km, primarily concentrated
-within Lisbon, and a low cycling modal share. According to the latest
-mobility survey conducted in 2018 \citep{IMOB}, the LMA registered a
-total of 5.3 million daily trips, with only 0.5\% by bicycle.
-
-The survey also showed that 58.4\% of trips were made by car, 15.5\% by
-public transport, and 23.9\% by walking (see Figure \ref{fig:mododist}).
-The number of intra-municipal trips --- with origin and destination in
-the same municipality --- amounts to 3.5 million trips. This exceeds the
-number of inter-municipal trips (1.8 million trips), involving travel
-between different municipalities. Cars and public transport are the most
-used modes for intercity trips, with cars being the predominant choice
-for all journeys. 53\% trips are up to 5 km distance, and 71\% up to 10
-km. Nevertheless, 29\% of trips are longer than 10 km, which requires
-the use of motorized modes, or active modes in combination with public
-transport.
-
-\begin{figure}
-\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
-\end{figure}
-
-Regarding cycling trips in the LMA, 55\% are up to 5 km, and 88\% are up
-to 10 km. These values are not in line with the typically found in
-cycling distance decay curves, which show a high percentage of trips are
-up to 5 km (\textasciitilde75\%), and a smaller proportion up to 10 km
-\citep{krizek2007detailed, larsen2010beyond, DDfunction2023}. This may
-be due to the low cycling modal share in the LMA - even more limited in
-the sample survey, which may be biased towards shorter trips.
 
@@
 trips between districts, specifically for the Lisbon metropolitan area.
-As shown, the city of Lisbon (in the centre) is the main attractor of
-trips, with a high number of trips to and from the other municipalities.
 
@@
 
-{\centering \includegraphics[width=1\linewidth,]{img/jitter_cairo}
+{\centering \includegraphics[width=1\linewidth,]{img/jitter}
 
@@
 
-\caption{Flow-weighted random sample of 10,000 desire lines in the Lisbon metropolitan area between districts of the 18 municipalities, without jittering (left) and with jittering (right).}\label{fig:jitter}
+\caption{Representation of desire lines in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
 \end{figure}
@@
 km from origin to interface \emph{A} plus 4 km from interface \emph{B}
-to destination). Furthermore, we have imposed restrictions on PT usage,
-limiting it to trips without PT transfers, and within a duration of up
-to 2 hours (120 minutes). Additionally, we have only included PT modes
-that can easily accommodate bicycles, such as trains, ferries, trams,
-and inter-municipal bus lines equipped with bike racks (Figure
-\ref{fig:map1}).
-
-This conservative approach was adopted to capture the fact that cycling
+to destination) or up to 25 minutes on bike travel-time. This
+conservative approach was adopted to capture the fact that cycling
 stages as part of a multi-modal trip are likely to be shorter than
-cycling-only trips \citep{vanmil_insights_2021}. Additionally,
-\citet{Leferink2017} found that the bicycle is most attractive option as
-a first and last mile mode when the distance ranges from 3 to 5 km, as
-observed across various comparative studies. This restrictions are based
-on the assumption that 5 km is a plausible maximum distance that people
-are willing to travel by bike as part of a longer journey. This applies
-especially for cycling beginners shifting from cars, who are less likely
-to be confident and experienced cyclists than the existing cyclists.
-
-These restrictions can be eased in the future when testing more
-developed policy interventions to enhance intermodality between cycling
-and PT, considering both the vehicle and infrastructure perspectives.
+cycling-only trips \citep{vanmil_insights_2021}. Furthermore, we have
+imposed restrictions on PT usage, limiting it to trips without PT
+transfers, and within a duration of up to 2 hours (120 minutes).
+Additionally, we have only included PT modes that can easily accommodate
+bicycles, such as trains, ferries, trams, and inter-municipal bus lines
+equipped with bike racks (Figure \ref{fig:map1}). These restrictions can
+be eased in the future when testing more developed policy interventions
+to enhance intermodality between cycling and PT, considering both the
+vehicle and infrastructure perspectives.
 
@@
 
-{\centering \includegraphics[width=0.6\linewidth,]{PaperCEUS_rev1_files/figure-latex/map1-1}
+{\centering \includegraphics[width=0.6\linewidth,]{img/map1}
 
@@
 Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
-the main PT interfaces in the LMA. It is noticeable the high potential
-of the train interfaces to attract car-to-PT substituting trips.
+the main PT interfaces in the LMA.
 
@@
 
-\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).
-Larger the line width, higher the segment's car to bicycle shift potential.
-Darker the line color, lower the quietness level.}\label{fig:map2}
+\caption{Bike routes with the highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
 \end{figure}
-
-\subsection{Modelling car shift to bicycle-PT - here or
-before??}\label{modelling-car-shift-to-bicycle-pt---here-or-before}
-
-It is essential to note that our approach is not predictive but
-scenario-based, providing a range of potential outcomes regarding the
-target (\% of cycling tips) and the routing profile (safe or direct
-routes), rather than a single forecast. These scenarios are rooted in
-the explicit national cycling targets, assuming that the potential for
-cycling is proportional to the reduction of car trips. Thus said, we do
-not predict the generation of new trips, but the substitution of car
-trips by cycling in combination with PT, considering the potential of
-the first and last mile of these journeys to be made by bicycle.
-Additionally, we do not model changes in the baseline PT trips, under
-the implicit assumption that the initial and final stages of these trips
-remain unaltered.
-
-A strength of this \emph{backcasting} approach is that we can simulate
-the aggregate travel patterns that would meet particular policy
-objectives, and explore what the implications could be for the public
-transport system and provision of safe cycling infrastructure to public
-transport nodes at regional, local authority and, importantly from a
-local transport planning practitioner perspective, corridor level.
 
@@
 benefits}\label{assessing-socio-environmental-benefits}
-
-Following the clearly established national strategy, we estimated the
-socio-environmental potential impacts of shifting car trips by bicycle
-in combination with PT, in the LMA, for the two targets (4\% and 10\%).
 
@@
 restrictions), the trips in the baseline scenario and corresponding new
-daily trips to achieve the national strategy targets (4\% and 10\%), for
+trips to achieve the national strategy targets (4\% and 10\%), for
 different route profiles. For the cycling legs of the journey (first and
@@
 
-\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario. Values in 'trips/day'.}
+\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario.}
 \centering
@@
 
-\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.}
+\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode.}
 \centering
@@
 \end{table}
-
-The benefits are spatially unevenly distributed across the LMA, as shown
-in Figure \ref{fig:mapzones}.
 
@@
 prone to further substitution, based on conservative assumptions.
-Although this paper does not address the designing specific solutions,
-our findings indicate a potential surge in demand (+6.3\%) of public
-transport interfaces to align with the national strategies. This added
-demand is conceptually driven from individuals utilising bicycles,
-either personal or shared, which would be parked at the PT stations.
-Consequently, operators and public space managers would face the
-challenge of accommodating this increased demand for bicycle parking
-and/or docks from a bike sharing system, since this availability plays a
-crucial role in the decision-making process of individuals to use
-combined bike-train \citep{jonkeren2021bicycle}. Additionally, they may
-need to explore options to enhance bicycle carriage capacity on public
-transport services, ensuring alignment with national targets.
-
-\section{Generalizability and
-limitations}\label{generalizability-and-limitations}
-
-This paper focus primarily on the potential for replacing trips
-currently made by car with combined cycling and public transport trips,
-and the associated social and environmental benefits, showcasing the
-potential impact of facilitating the first-and-last mile to PT
-interfaces. While the results are specific to the Lisbon Metropolitan
-Area, the approach is designed to be generalizable to other metropolitan
-areas.
-
-An important aspect of the approach from a generalizability perspective
-is that it has specific data requirements, including origin-destination,
-road network, and the GTFS data. Such datasets are available to
-transport planners in many areas. In areas where such datasets do not
-exist, the value of research such as this could provide a motivation to
-collect and open-up transport data. For this study we have used
-Portugal's national cycling targets as the basis of the scenarios; the
-approach can be adapted to other targets, or to other policy goals, such
-as reducing congestion or improving air quality.
-
-Nevertheless, the extent of the potential shift from car to PT is highly
-dependent on the distances typically travelled in such contexts, and the
-availability of PT services - which can vary significantly between
-regions, and for which we are not assuming any change in this study. For
-instance, we considered only the inter-municipal bus services that
-accommodate bicycles on board. Future research could explore the
-potential of equipping other regional bus services with this feature and
-analyse their attractiveness, combined with cycling, to address current
-car trip demands.
-
-A strength of the approach is that there is no requirement for sensitive
-individual trip level data to reproduce the approach in other cities.
-However, we acknowledge that this approach has limitations. The approach
-cannot show exactly which trips are being replaced, in terms of the age
-of people making the trips, the types of car they drive, and other
-individual or vehicle level variables. We assumed uniformity in the type
-of vehicle for all replaced car journeys, albeit with variations in the
-fuels utilised. Additionally, we assumed consistency in the health
-impacts across all individuals for the replaced car trips. This approach
-is a simplification of reality to enable the analysis of a large region
-on available computing resources. Future work could seek to increase the
-resolution of the results, and move towards a more individual-level
-analysis, for example by using agent and activity-based models using
-tools such as MATSim \citep{MATSim}. This option would require
-substantial resources outside the scope of this study; another way to
-improve the spatial resolution of results would be to use higher
-resolution `subpoints' as the basis of the desire line `jittering'
-process. We would also like to explore the potential for probabilistic
-routing to improve the spatial resolution of the results.
-
-A further refinement could be to differentiate between `simpler'
-no-transfer PT trips (which require less person effort) and more arduous
-multi-transfer PT trips, using the same methods, by relaxing the
-assumption of no-transfers between PT modes in the routing engine. This
-would lead to a larger number of total daily trips that can be made by
-cycling + PT, in particular in locations where the PT network is less
-complex and requires more transfers. We could also set a different
-maximum length of cycling stage, extending the bicycle catchment area.
-However, despite the added value of a sensitivity analyses, they would
-also increase the overall complexity, and would have an impact on the
-explainability of the results to be used by policy makers.
 

### 2024-05-09 paper/PaperCEUS/PaperCEUS_rev1.tex

@@
 substitute car trips in the LMA. After presenting the
-\hyperref[methods]{Methods} used in Chapter 2, we assess its
+\hyperref[methods]{Methods} used in Section 2, we assess its
 socio-environmental impacts using open data and open-source tools,
 presenting the \hyperref[results-and-discussion]{Results and Discussion}
-in Chapter 3, followed by Chapter 4 where we discuss the
+in Section 3, followed by Section 4 where we discuss the
 \hyperref[generalizability-and-limitations]{Generalizability and
 Limitations} of the methods. We \hyperref[Conclusions]{Conclude} the
-paper in Chapter 5 with the main findings and policy implications.
+paper in Section 5 with the main findings and policy implications.
 

### 2024-08-31 paper/PaperCEUS/PaperCEUS_rev2_word.Rmd
setup paper ceus rev2

--- 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1_word.Rmd
@@
 # setwd("paper/PaperCEUS/") # knit directory: Document
-system("latexdiff paper/PaperCEUS/PaperCEUS_prerev.tex paper/PaperCEUS/PaperCEUS_rev1.tex > paper/PaperCEUS/PaperCEUS_diff1.tex")
+system("latexdiff paper/PaperCEUS/PaperCEUS_rev1.tex paper/PaperCEUS/PaperCEUS_rev2.tex > paper/PaperCEUS/PaperCEUS_diff2.tex")
 #and change in tex the blue for another color (green? 0,0.5,0) and remove \uwave
@@
 #and compile PDF!
-system("pdflatex paper/PaperCEUS/PaperCEUS_diff1.tex")
+system("pdflatex paper/PaperCEUS/PaperCEUS_diff2.tex")
 ```

### 2024-09-02 paper/PaperCEUS/PaperCEUS_rev2_word.Rmd
test load files
knit again

@@
 
-stopsAML = sf::st_read("load/Stops_AML.geojson", quiet = TRUE)
+stopsAML = sf::st_read("load/StopsAML.geojson", quiet = TRUE)
 stopsAML = stopsAML |> filter(mode != "Metropolitano") |>
@@
                        labels = c("Train", "Ferry", "Tram", "Bus")))
-routesAML = sf::st_read("load/Routes_AML.geojson", quiet = TRUE)
+routesAML = sf::st_read("load/RoutesAML.geojson", quiet = TRUE)
 routesAML = routesAML |> 

### 2024-09-02 paper/PaperCEUS/PaperCEUS_rev2_word.tex
knit again

--- 2024-05-09 paper/PaperCEUS/PaperCEUS_rev1.tex
@@
 \usepackage{lineno} % add
-  \linenumbers % turns line numbering on
 \usepackage{ifxetex,ifluatex}
@@
 \begin{figure}
-\subfloat[Population distribution in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to the 2021 census and the 2017 travel survey.}\label{fig:lma}
+\subfloat[Population distribution in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev2_word_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev2_word_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA, according to the 2021 census and the 2017 travel survey.}\label{fig:lma}
 \end{figure}
@@
 \begin{figure}
-\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
+\subfloat[In percentage\label{fig:mododist-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev2_word_files/figure-latex/mododist-1} }\subfloat[In total\label{fig:mododist-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev2_word_files/figure-latex/mododist-2} }\caption{\label{mododist}Trips in the LMA by inter/intra municipal and mode, according to the travel survey.}\label{fig:mododist}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.6\linewidth,]{PaperCEUS_rev1_files/figure-latex/map1-1}
+{\centering \includegraphics[width=0.6\linewidth,]{PaperCEUS_rev2_word_files/figure-latex/map1-1}
 
@@
 \begin{figure}
-\subfloat[Potential new Bicycle + PT trips\label{fig:mapzones-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-1} }\subfloat[Proportion of total potential modal shift in the LMA\label{fig:mapzones-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/mapzones-2} }\caption{\label{mapzones}Potential Bicycle + PT daily trips for the 10\% target and direct routing, by municipality (origins).
+\subfloat[Potential new Bicycle + PT trips\label{fig:mapzones-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev2_word_files/figure-latex/mapzones-1} }\subfloat[Proportion of total potential modal shift in the LMA\label{fig:mapzones-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev2_word_files/figure-latex/mapzones-2} }\caption{\label{mapzones}Potential Bicycle + PT daily trips for the 10\% target and direct routing, by municipality (origins).
 The percentages are shown for municipalities with values above 4\%.}\label{fig:mapzones}

### 2024-09-02 paper/PaperCEUS/PaperCEUS_rev2_word.Rmd
reply to 2.3 - why more safe than direct

--- 2024-08-31 paper/PaperCEUS/PaperCEUS_rev2_word.Rmd
@@
 
-The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
+The `r5r` model uses the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
 This information is crucial for an accurate PT trip and route estimation.
@@
 The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.
-We restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination).
-Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
-Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
+
+Trips were routed for all the OD jittered pairs, using `r5r`.
+The resulting estimated routes are the ones with the least-cost paths (shortest travel time) for each OD pair.
+For some routes, bicycle alone can be more attractive than the combination of bicycle and PT - for instance for shorter trips.
+
+Figure \ref{fig:trip_schematic} illustrates a schematic trip for different routing profiles (_safe_ and _direct_), and only using Bicycle or Bicycle in combination with Public Transportation, considering the same origin and destination.
+As we can see, for the same routing profile, the _direct_ route by bicycle can be more convenient (attractive) than the bicycle in combination with PT.
+The opposite can also be true, in particular for longer journeys, where the combination of bicycle and PT can be more attractive than the bicycle mode.
+When comparing the _direct_ with the _safe_ route profile, the _safe_ route is usually longer, as it avoids the most stressful roads (LTS 4), which are usually the most direct ones, considering only roads with LTS 3 or lower.
+In this case, bike (without combination) can be more attractive than the bike with PT, as the PT leg can be longer or take more time than the safe bike route.
+
+```{r trip_schematic, fig.align='center', fig.cap="Schematic trip for different routing profiles (safe and direct), and only using Bicycle or Bicycle with Public Transportation, considering the same origin and destination.
+The black lines represent the trip made by Bike, and the dashed orange lines represent the trip made by PT.", fig.height=4, out.width="70%"}
+# png::writePNG(pdftools::pdf_render_page("img/schematic_trips.pdf", dpi = 300), "img/schematic_trips_cairo.png")
+knitr::include_graphics("img/schematic_trips_cairo.png", error = FALSE)
+```
+
+We have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
 This map shows that some of the municipalities are not served by PT modes that can easily accommodate bicycles, which may limit the potential for intermodality in these areas.
 
-This conservative approach was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
+Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
+This resulted in that only 53.8% of the total trips between the jittered origins and destinations are possible to be made by bicycle or bicycle + PT (for the considered PT modes) for the _safe_ routing profile, and 69.2% for the _direct_ routing profile.
+
+After the estimation of routes, we considered only the Origin-Destination pairs for which the estimated routes were the ones that were more attractive (least-cost path / fastest) for the bicycle and PT combination, than for bicycle only.
+This process resulted in a subset of the OD pairs used in the intermodality scenario: 19.5% of the total OD pairs with _direct_ routing profile, and 21% of the total OD pairs with _safe_ routing profile.
+This difference is due to the fact that the direct routes can be more attractive for cycling only than for the combination of bicycle and PT.
+In particular, because main PT connections are along busy roads, and thus, for short trips (e.g.
+intramunicipal trips) using bicycle only at the main roads can be more attractive (shorter travel time) than the combination of bicycle and PT.
+For _safe_ routes, the alternative of using bicycle for the PT leg using safe routes can take longer - since the main roads are restricted in this routing profile -, and thus less attractive.
+
+Additionally, we restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination).
+This restriction resulted in 538.514 trips that can be made by bicycle in combination with PT (`r round(538514/sum(TRIPSmode_aml$viagens)*100,1)`% of all daily trips) considering the safe routing profile, and 500.880 trips considering the direct routing profile (`r round(500880/sum(TRIPSmode_aml$viagens)*100,1)`% of all daily trips).
+
+This conservative approach, regarding the assumed multi-modal restrictions, was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
 Additionally, @Leferink2017 found that the bicycle is most attractive option as a first and last mile mode when the distance ranges from 3 to 5 km, as observed across various comparative studies.
@@
 For the *second leg of the journey*, we estimate the environmental impacts of shifting car trips to PT (between the PT interfaces).
+Car routing was estimated for all the journeys with the potential to be replaced by bike in combination with PT.
 
@@
 Another relevant aspect is that the higher potential (Figure \ref{fig:mapzones}.a) is in the municipalities with the highest number of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and Almada, which are also the ones with the highest number of PT interfaces (Figure \ref{fig:map1}).
-On another hand, the municipalities with little PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a lower potential for car to Bike-PT substitution (Figure \ref{fig:mapzones}.a), even if they have a high number of trips, such as Setubal.
-This findings are aligned with their local travel patterns, where the majority of trips are intra-municipal (67% for both municipalities), and not to the city of Lisbon, despite the high car-dependency (79% in Mafra, 67% in Setubal).
+On another hand, the municipalities with little PT interfaces connecting to Lisbon, such as Mafra and Setúbal, have a lower potential for car to Bike-PT substitution (Figure \ref{fig:mapzones}.a), even if they have a high number of trips, such as Setúbal.
+This findings are aligned with their local travel patterns, where the majority of trips are intra-municipal (67% for both municipalities), and not to the city of Lisbon, despite the high car-dependency (79% in Mafra, 67% in Setúbal).
+
+A particular difference is noted regarding the attraction of trams to be used in combination with cycling, for _safe_ and _direct_ routes.
+In this case, Tram is available for intermodality only at one municipality (Almada) (see \ref{fig:map1}), and most of its stations are located on major roads.
+Their accessibility involves travelling along these major roads (with LTS 4), which limits the potential of valid routing options when considering _safe_ routes. 
+In this case, Train offers the highest potential for _safe_ routes, and Tram for _direct_ routes.
 

### 2024-09-02 paper/PaperCEUS/PaperCEUS_rev2_word.tex
reply to 2.3 - why more safe than direct

@@
 The intermodality scenario considers trips made by PT in which cycling
-is used for the first and last legs. We restricted our analysis to the
-first and last legs with a combined length of up to 5 km (for example: 1
-km from origin to interface \emph{A} plus 4 km from interface \emph{B}
-to destination). Furthermore, we have imposed restrictions on PT usage,
-limiting it to trips without PT transfers, and within a duration of up
-to 2 hours (120 minutes). Additionally, we have only included PT modes
-that can easily accommodate bicycles, such as trains, ferries, trams,
-and inter-municipal bus lines equipped with bike racks (Figure
-\ref{fig:map1}). This map shows that some of the municipalities are not
-served by PT modes that can easily accommodate bicycles, which may limit
-the potential for intermodality in these areas.
+is used for the first and last legs. The trips were routed for all the
+OD jittered pairs, using \texttt{r5r}. The estimated routes are the ones
+with the least-cost paths (shortest travel time). For some routes,
+bicycle alone can be more attractive than the combination of bicycle and
+PT - for instance for shorter trips.
 
-This conservative approach was adopted to capture the fact that cycling
-stages as part of a multi-modal trip are likely to be shorter than
-cycling-only trips \citep{vanmil_insights_2021}. Additionally,
-\citet{Leferink2017} found that the bicycle is most attractive option as
-a first and last mile mode when the distance ranges from 3 to 5 km, as
-observed across various comparative studies. This restrictions are based
-on the assumption that 5 km is a plausible maximum distance that people
-are willing to travel by bike as part of a longer journey. This applies
+Figure \ref{fig:trip_schematic} illustrates a schematic trip for
+different routing profiles (safe and direct), and only using Bicycle or
+Bicycle in combination with Public Transportation, considering the same
+origin and destination. As we can see, for the same routing profile, the
+direct route by bicycle can be more convenient (attractive) than the
+bicycle in combination with PT. The opposite can also be true, in
+particular for longer journeys, where the combination of bicycle and PT
+can be more attractive than the bicycle mode. When comparing the direct
+with the safe route profile, the safe route is usually longer, as it
+avoids the most stressful roads (LTS 4), which are usually the most
+direct ones, considering only roads with LTS 3 or lower. In this case,
+the bike route (without combination) can be more attractive than the
+bike with PT, as the PT leg can be longer or take more time than the
+safe bike route.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth,]{img/schematic_trips_cairo}
+
+}
+
+\caption{Schematic trip for different routing profiles (safe and direct), and only using Bicycle or Bicycle with Public Transportation, considering the same origin and destination.
+The black lines represent the trip made by Bike, and the dashed orange lines represent the trip made by PT.}\label{fig:trip_schematic}
+\end{figure}
+
+We have only included PT modes that can easily accommodate bicycles,
+such as trains, ferries, trams, and inter-municipal bus lines equipped
+with bike racks (Figure \ref{fig:map1}). This map shows that some of the
+municipalities are not served by PT modes that can easily accommodate
+bicycles, which may limit the potential for intermodality in these
+areas.
+
+Furthermore, we have imposed restrictions on PT usage, limiting it to
+trips without PT transfers, and within a duration of up to 2 hours (120
+minutes). This resulted in that only 53.8\% of the total trips between
+the jittered origins and destinations are possible to be made by bicycle
+or bicycle + PT (for the modelled PT modes), considering the safe
+routing profile, and 69.2\% considering the direct routing profile.
+
+After the routes were estimated, we considered only the
+Origin-Destination pairs for which the estimated routes were the ones
+that were more attractive (least-cost path / fastest) for the bicycle
+and PT combination, than for bicycle only. This process resulted in a
+subset of the OD pairs that were considered for the intermodality
+scenario: 19.5\% of the total OD pairs with direct routing profile, and
+21\% of the total OD pairs with safe routing profile. This difference is
+due to the fact that the direct routes are can be more attractive for
+the bicycle mode than the combination of bicycle and PT, in particular
+because main PT connections are along busy roads, and thus, for short
+trips (for instance intramunicipal trips) the bicycle mode can be more
+attractive than the combination of bicycle and PT. For safe routes, the
+alternative of using bicycle for the PT leg using safe routes can take
+longer - since the main roads are restricted in this routing profile -,
+and thus less attractive.
+
+Additionally, we restricted our analysis to the first and last legs with
+a combined length of up to 5 km (for example: 1 km from origin to
+interface \emph{A} plus 4 km from interface \emph{B} to destination).
+This restriction resulted in 538.514 trips that can be made by bicycle
+in combination with PT (10.1\% of all daily trips) considering the safe
+routing profile, and 696.000 trips considering the direct routing
+profile (9.4\% of all daily trips).
+
+This conservative approach, regarding the assumed multi-modal
+restrictions, was adopted to capture the fact that cycling stages as
+part of a multi-modal trip are likely to be shorter than cycling-only
+trips \citep{vanmil_insights_2021}. Additionally, \citet{Leferink2017}
+found that the bicycle is most attractive option as a first and last
+mile mode when the distance ranges from 3 to 5 km, as observed across
+various comparative studies. This restrictions are based on the
+assumption that 5 km is a plausible maximum distance that people are
+willing to travel by bike as part of a longer journey. This applies
 especially for cycling beginners shifting from cars, who are less likely
@@
 For the \emph{second leg of the journey}, we estimate the environmental
-impacts of shifting car trips to PT (between the PT interfaces).
+impacts of shifting car trips to PT (between the PT interfaces). Car
+routing was estimated for all the journeys with the potential to be
+replaced by bike in combination with PT.
 
@@
 (Figure \ref{fig:map1}). On another hand, the municipalities with little
-PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a
+PT interfaces connecting to Lisbon, such as Mafra and Setúbal, have a
 lower potential for car to Bike-PT substitution (Figure
 \ref{fig:mapzones}.a), even if they have a high number of trips, such as
-Setubal. This findings are aligned with their local travel patterns,
+Setúbal. This findings are aligned with their local travel patterns,
 where the majority of trips are intra-municipal (67\% for both
 municipalities), and not to the city of Lisbon, despite the high
-car-dependency (79\% in Mafra, 67\% in Setubal).
+car-dependency (79\% in Mafra, 67\% in Setúbal).
+
+A particular difference is noted regarding the attraction of trams to be
+used in combination with cycling, for \emph{safe} and \emph{direct}
+routes. In this case, Tram is available for intermodality only at one
+municipality (Almada) (see \ref{fig:map1}), and most of its stations are
+located on major roads. Their accessibility involves travelling along
+these major roads (with LTS 4), which limits the potential of valid
+routing options when considering \emph{safe} routes. In this case, Train
+offers the highest potential for \emph{safe} routes, and Tram for
+\emph{direct} routes.
 

### 2024-10-05 paper/PaperCEUS/PaperCEUS_rev2_word.Rmd
update rev 2

--- 2024-08-31 paper/PaperCEUS/PaperCEUS_rev2_word.Rmd
@@
 
-The `r5r` model used the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
+Trips were routed for all the OD jittered pairs, resulting in the least-cost paths (shortest travel time) for each pair.
+
+Figure \ref{fig:trip_schematic} illustrates a schematic comparison between *safe* and *direct* routing profiles, showing both bicycle-only and Bike + PT options.
+In the *direct* profile, bicycle-only trips can be more attractive due to shorter travel times, especially for shorter distances, although, for longer trips, combining cycling with PT can be faster.
+In the *safe* routing profile, where longer, less stressful routes (LTS 3 or lower), are used to avoid high-stress roads (LTS 4), cycling in combination with PT can be more efficient.
+
+```{r trip_schematic, fig.align='center', fig.cap="Schematic trip for safe and direct routing profiles, with bicycle-only and bicycle with PT options.
+The black lines represent the Biclcye legs, and the dashed orange lines represent the PT leg.", fig.height=4, out.width="70%"}
+# png::writePNG(pdftools::pdf_render_page("img/schematic_trips.pdf", dpi = 300), "img/schematic_trips_cairo.png")
+knitr::include_graphics("img/schematic_trips_cairo.png", error = FALSE)
+```
+
+The `r5r` model uses the OpenStreetMap road network and the GTFS metropolitan data aggregated and validated.
 This information is crucial for an accurate PT trip and route estimation.
@@
 The intermodality scenario considers trips made by PT in which cycling is used for the first and last legs.
-We restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination).
-Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
-Additionally, we have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
+
+We have only included PT modes that can easily accommodate bicycles, such as trains, ferries, trams, and inter-municipal bus lines equipped with bike racks (Figure \ref{fig:map1}).
 This map shows that some of the municipalities are not served by PT modes that can easily accommodate bicycles, which may limit the potential for intermodality in these areas.
 
-This conservative approach was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
+Furthermore, we have imposed restrictions on PT usage, limiting it to trips without PT transfers, and within a duration of up to 2 hours (120 minutes).
+Additionally, we restricted our analysis to the first and last legs with a combined length of up to 5 km (for example: 1 km from origin to interface _A_ plus 4 km from interface _B_ to destination).
+
+<!-- This resulted in that only 53.8% of the total trips between the jittered origins and destinations are possible to be made by bicycle-only or bicycle + PT (for the considered PT modes) for the _safe_ routing profile, and 69.2% for the _direct_ routing profile.
+-->
+
+<!-- After the estimation of routes, we considered only the Origin-Destination pairs for which the estimated routes were the ones that were more attractive (least-cost path / fastest) for the bicycle and PT combination, than for bicycle only.
+-->
+
+This conservative approach, regarding the assumed multi-modal restrictions, was adopted to capture the fact that cycling stages as part of a multi-modal trip are likely to be shorter than cycling-only trips [@vanmil_insights_2021].
 Additionally, @Leferink2017 found that the bicycle is most attractive option as a first and last mile mode when the distance ranges from 3 to 5 km, as observed across various comparative studies.
@@
 Table \ref{tab:summary1} presents the LMA total daily trips that can be made with cycling + TP combination (with the aforementioned restrictions), the trips in the baseline scenario and corresponding new daily trips to achieve the national strategy targets (4% and 10%), for different route profiles, at an aggregate level of the LMA.
+It should be noted that for the OD pairs with already 4% or 10% bike trips in 2018, those are not accounted as new potential bike trips, resulting in potential Cycling + PT numbers slightly below the 4% and 10% of the total trips.
+
 For the cycling legs of the journey (first and last legs), the environmental avoided emissions and monetized socio-environment (SE) benefits are presented in Table \ref{tab:summary1b}, resulting from replacing car trips with cycling.
@@
 
-For both *direct* and *safe* route profiles, 10% of the daily trips have the potential to be made by a combination of PT and cycling, even given the travel restrictions considered (up to 5 km on bike, up to 2 hours, no possible transfers between PT).
+Even given the travel restrictions considered (up to 5 km on bike, up to 2 hours, no possible transfers between PT), this resulted in 538.514 trips with potential to be made by bicycle in combination with PT (`r round(538514/sum(TRIPSmode_aml$viagens)*100,1)`% of all daily trips) for the *safe* routing profile, and 500.880 trips for the *direct* routing profile (`r round(500880/sum(TRIPSmode_aml$viagens)*100,1)`% of all daily trips).
 This unveils the potential of cycling as a complementary mode of PT, with the potential to uptake the number of PT trips within the LMA area by as much as 6.3% (in addition to the 825 thousand PT trips reported in the mobility survey).
@@
       #               "Bus", "Ferry", "Train", "Tram"),
-      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.",
+      caption = "\\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disaggregated by PT mode. Values in 'trips/day'.",
       format="latex",
@@
 
+The higher number of Bike + PT trips in the _safe_ routing profile can be explained by the reduced competitiveness of cycling alone in this profile, as it restricts cyclists to lower-traffic sections, with less directed options, and a subset of the total road network.
+In contrast, the _direct_ routing profile, which allows for direct cycling routes along the entire network, often results in faster single-mode bicycle trips - more competitive than using the Bike in combination with PT.
+As a result, more trips are made using Bike + PT under the _safe_ routing profile, while _direct_ routes favour cycling-only journeys.
+This justifies the aggregate difference in the number of potential trips between the two profiles.
+
+For trams, which primarily operate along busy main roads that are generally hostile for cycling (classified as LTS 4), the Bike + Tram combination is often less competitive compared to direct bike-only routes, especially for shorter trips.
+This is particularly evident in the case of Almada, the only municipality in this study with a tram system that can be integrated into Bike + PT trips (see Figure \ref{fig:map1}).
+For _direct_ routing, bike-only options are usually faster for shorter local trips within the municipality.
+However, under _safe_ routing assumptions (LTS 3), where cyclists are limited to less direct, low-stress routes, the Bike + Tram option becomes more attractive, as the bike-only option would be longer, albeit safer, avoiding the main roads.
+This explains the higher competitiveness of Tram + Bike in the safe profile in this municipality.
+
 Another relevant aspect is that the higher potential (Figure \ref{fig:mapzones}.a) is in the municipalities with the highest number of trips (Figure \ref{fig:lma}.b), such as Lisbon, Sintra, Amadora, and Almada, which are also the ones with the highest number of PT interfaces (Figure \ref{fig:map1}).
-On another hand, the municipalities with little PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a lower potential for car to Bike-PT substitution (Figure \ref{fig:mapzones}.a), even if they have a high number of trips, such as Setubal.
-This findings are aligned with their local travel patterns, where the majority of trips are intra-municipal (67% for both municipalities), and not to the city of Lisbon, despite the high car-dependency (79% in Mafra, 67% in Setubal).
+On another hand, the municipalities with little PT interfaces connecting to Lisbon, such as Mafra and Setúbal, have a lower potential for car to Bike-PT substitution (Figure \ref{fig:mapzones}.a), even if they have a high number of trips, such as Setúbal.
+This findings are aligned with their local travel patterns, where the majority of trips are intra-municipal (67% for both municipalities), and not to the city of Lisbon, despite the high car-dependency (79% in Mafra, 67% in Setúbal).
+
+<!-- A particular difference is noted regarding the attraction of trams to be used in combination with cycling, for _safe_ and _direct_ routes. -->
+<!-- In this case, Tram is available for intermodality only at one municipality (Almada) (see \ref{fig:map1}), and most of its stations are located on major roads. -->
+<!-- Their accessibility involves travelling along these major roads (with LTS 4), which limits the potential of valid routing options when considering _safe_ routes.  -->
+<!-- In this case, Train offers the highest potential for _safe_ routes, and Tram for _direct_ routes. -->
 

### 2024-10-05 paper/PaperCEUS/PaperCEUS_rev2_word.tex
update rev 2

--- 2024-09-02 paper/PaperCEUS/PaperCEUS_rev2_word.tex
@@
 
-The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
+Trips were routed for all the OD jittered pairs, resulting in the
+least-cost paths (shortest travel time) for each pair.
+
+Figure \ref{fig:trip_schematic} illustrates a schematic comparison
+between \emph{safe} and \emph{direct} routing profiles, showing both
+bicycle-only and Bike + PT options. In the \emph{direct} profile,
+bicycle-only trips can be more attractive due to shorter travel times,
+especially for shorter distances, although, for longer trips, combining
+cycling with PT can be faster. In the \emph{safe} routing profile, where
+longer, less stressful routes (LTS 3 or lower), are used to avoid
+high-stress roads (LTS 4), cycling in combination with PT can be more
+efficient.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth,]{img/schematic_trips_cairo}
+
+}
+
+\caption{Schematic trip for safe and direct routing profiles, with bicycle-only and bicycle with PT options.
+The black lines represent the Biclcye legs, and the dashed orange lines represent the PT leg.}\label{fig:trip_schematic}
+\end{figure}
+
+The \texttt{r5r} model uses the OpenStreetMap road network and the GTFS
 metropolitan data aggregated and validated. This information is crucial
@@
 The intermodality scenario considers trips made by PT in which cycling
-is used for the first and last legs. We restricted our analysis to the
-first and last legs with a combined length of up to 5 km (for example: 1
-km from origin to interface \emph{A} plus 4 km from interface \emph{B}
-to destination). Furthermore, we have imposed restrictions on PT usage,
-limiting it to trips without PT transfers, and within a duration of up
-to 2 hours (120 minutes). Additionally, we have only included PT modes
-that can easily accommodate bicycles, such as trains, ferries, trams,
-and inter-municipal bus lines equipped with bike racks (Figure
-\ref{fig:map1}). This map shows that some of the municipalities are not
-served by PT modes that can easily accommodate bicycles, which may limit
-the potential for intermodality in these areas.
+is used for the first and last legs.
 
-This conservative approach was adopted to capture the fact that cycling
-stages as part of a multi-modal trip are likely to be shorter than
-cycling-only trips \citep{vanmil_insights_2021}. Additionally,
-\citet{Leferink2017} found that the bicycle is most attractive option as
-a first and last mile mode when the distance ranges from 3 to 5 km, as
-observed across various comparative studies. This restrictions are based
-on the assumption that 5 km is a plausible maximum distance that people
-are willing to travel by bike as part of a longer journey. This applies
+We have only included PT modes that can easily accommodate bicycles,
+such as trains, ferries, trams, and inter-municipal bus lines equipped
+with bike racks (Figure \ref{fig:map1}). This map shows that some of the
+municipalities are not served by PT modes that can easily accommodate
+bicycles, which may limit the potential for intermodality in these
+areas.
+
+Furthermore, we have imposed restrictions on PT usage, limiting it to
+trips without PT transfers, and within a duration of up to 2 hours (120
+minutes). Additionally, we restricted our analysis to the first and last
+legs with a combined length of up to 5 km (for example: 1 km from origin
+to interface \emph{A} plus 4 km from interface \emph{B} to destination).
+
+This conservative approach, regarding the assumed multi-modal
+restrictions, was adopted to capture the fact that cycling stages as
+part of a multi-modal trip are likely to be shorter than cycling-only
+trips \citep{vanmil_insights_2021}. Additionally, \citet{Leferink2017}
+found that the bicycle is most attractive option as a first and last
+mile mode when the distance ranges from 3 to 5 km, as observed across
+various comparative studies. This restrictions are based on the
+assumption that 5 km is a plausible maximum distance that people are
+willing to travel by bike as part of a longer journey. This applies
 especially for cycling beginners shifting from cars, who are less likely
@@
 daily trips to achieve the national strategy targets (4\% and 10\%), for
-different route profiles, at an aggregate level of the LMA. For the
-cycling legs of the journey (first and last legs), the environmental
-avoided emissions and monetized socio-environment (SE) benefits are
-presented in Table \ref{tab:summary1b}, resulting from replacing car
-trips with cycling.
+different route profiles, at an aggregate level of the LMA. It should be
+noted that for the OD pairs with already 4\% or 10\% bike trips in 2018,
+those are not accounted as new potential bike trips, resulting in
+potential Cycling + PT numbers slightly below the 4\% and 10\% of the
+total trips.
+
+For the cycling legs of the journey (first and last legs), the
+environmental avoided emissions and monetized socio-environment (SE)
+benefits are presented in Table \ref{tab:summary1b}, resulting from
+replacing car trips with cycling.
 
@@
 
-For both \emph{direct} and \emph{safe} route profiles, 10\% of the daily
-trips have the potential to be made by a combination of PT and cycling,
-even given the travel restrictions considered (up to 5 km on bike, up to
-2 hours, no possible transfers between PT). This unveils the potential
-of cycling as a complementary mode of PT, with the potential to uptake
-the number of PT trips within the LMA area by as much as 6.3\% (in
-addition to the 825 thousand PT trips reported in the mobility survey).
+Even given the travel restrictions considered (up to 5 km on bike, up to
+2 hours, no possible transfers between PT), this resulted in 538.514
+trips with potential to be made by bicycle in combination with PT
+(10.1\% of all daily trips) for the \emph{safe} routing profile, and
+500.880 trips for the \emph{direct} routing profile (9.4\% of all daily
+trips). This unveils the potential of cycling as a complementary mode of
+PT, with the potential to uptake the number of PT trips within the LMA
+area by as much as 6.3\% (in addition to the 825 thousand PT trips
+reported in the mobility survey).
 
@@
 
-\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disagregated by PT mode. Values in 'trips/day'.}
+\caption{\label{tab:summary21}\label{summary21}Summary of the potential of replacing car trips with cycling in combination with PT, disaggregated by PT mode. Values in 'trips/day'.}
 \centering
@@
 
+The higher number of Bike + PT trips in the \emph{safe} routing profile
+can be explained by the reduced competitiveness of cycling alone in this
+profile, as it restricts cyclists to lower-traffic sections, with less
+directed options, and a subset of the total road network. In contrast,
+the \emph{direct} routing profile, which allows for direct cycling
+routes along the entire network, often results in faster single-mode
+bicycle trips - more competitive than using the Bike in combination with
+PT. As a result, more trips are made using Bike + PT under the
+\emph{safe} routing profile, while \emph{direct} routes favour
+cycling-only journeys. This justifies the aggregate difference in the
+number of potential trips between the two profiles.
+
+For trams, which primarily operate along busy main roads that are
+generally hostile for cycling (classified as LTS 4), the Bike + Tram
+combination is often less competitive compared to direct bike-only
+routes, especially for shorter trips. This is particularly evident in
+the case of Almada, the only municipality in this study with a tram
+system that can be integrated into Bike + PT trips (see Figure
+\ref{fig:map1}). For \emph{direct} routing, bike-only options are
+usually faster for shorter local trips within the municipality. However,
+under \emph{safe} routing assumptions (LTS 3), where cyclists are
+limited to less direct, low-stress routes, the Bike + Tram option
+becomes more attractive, as the bike-only option would be longer, albeit
+safer, avoiding the main roads. This explains the higher competitiveness
+of Tram + Bike in the safe profile in this municipality.
+
 Another relevant aspect is that the higher potential (Figure
@@
 (Figure \ref{fig:map1}). On another hand, the municipalities with little
-PT interfaces connecting to Lisbon, such as Mafra and Setubal, have a
+PT interfaces connecting to Lisbon, such as Mafra and Setúbal, have a
 lower potential for car to Bike-PT substitution (Figure
 \ref{fig:mapzones}.a), even if they have a high number of trips, such as
-Setubal. This findings are aligned with their local travel patterns,
+Setúbal. This findings are aligned with their local travel patterns,
 where the majority of trips are intra-municipal (67\% for both
 municipalities), and not to the city of Lisbon, despite the high
-car-dependency (79\% in Mafra, 67\% in Setubal).
+car-dependency (79\% in Mafra, 67\% in Setúbal).
 

### 2024-10-06 paper/PaperCEUS/PaperCEUS_rev2_word.tex
tex minor updates

@@
 \href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
-package}, we disaggregated the OD data into desire lines reprenting a
+package}, we disaggregated the OD data into desire lines representing a
 maximum of 100 trips each. Figure \ref{fig:jitter} illustrates the
@@
 characteristics of the cycling and road networks and the location of
-public transport interfaces. Other constraints regarding the behavior of
-potential cyclists determine the routing results. For example, such
-restrictions can favor low speed, low traffic streets, more direct
+public transport interfaces. Other constraints regarding the behaviour
+of potential cyclists determine the routing results. For example, such
+restrictions can favour low speed, low traffic streets, more direct
 routes, and less steep paths, among others, which are suitable for
@@
 an average speed of 15 km/h during rush hour periods. Since the average
-distance traveled per trip influences the overconsumption and emissions
+distance travelled per trip influences the overconsumption and emissions
 from cold-start engine operation, we estimated energy and emission
@@
 
-\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potencial of the intermodality scenario. Values in 'trips/day'.}
+\caption{\label{tab:summary1}\label{summary1}Summary of the cycling potential of the intermodality scenario. Values in 'trips/day'.}
 \centering
@@
 \toprule
-Target & Routing & Total trips & Baseline Cycling + PT & Potencial Cycling + PT\\
+Target & Routing & Total trips & Baseline Cycling + PT & Potential Cycling + PT\\
 \midrule
@@
 
-\caption{\label{tab:summary1b}\label{summary1b}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.}
+\caption{\label{tab:summary1b}\label{summary1b}Summary of the cycling potential of intermodality scenario and its socio-environmental benefits for the cycling legs.}
 \centering
@@
 
-\caption{\label{tab:summary22}\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.}
+\caption{\label{tab:summary22}\label{summary22}Summary of the avoided emissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.}
 \centering
@@
 
-\caption{\label{tab:summaryall}\label{summaryall}Summary of the avoided CO2eq emmissions (ton/year) and the estimated social and environmental benefits (monetized in thousand €) by replacing car trips with cycling in combination with PT.}
+\caption{\label{tab:summaryall}\label{summaryall}Summary of the avoided CO2eq emissions (ton/year) and the estimated social and environmental benefits (monetized in thousand €) by replacing car trips with cycling in combination with PT.}
 \centering
@@
 the social impacts from substituting car trips. One of the main
-socio-environmetal benefits, valued after monetization, comes from the
+socio-environmental benefits, valued after monetization, comes from the
 increase in physical activity \citep{Felix2023ES}. Although there are
@@
 CO\textsubscript{2}eq emissions annually. These changes are estimated to
-yield socio-environmental benefits totaling €125 million to €325 million
-over a decade. The quantification of such benefits can support
+yield socio-environmental benefits totalling €125 million to €325
+million over a decade. The quantification of such benefits can support
 policy-makers in prioritizing interventions to reduce the reliance on

### 2024-10-06 paper/PaperCEUS/PaperCEUS_rev2_word.Rmd
tiny typos

@@
 # setwd("paper/PaperCEUS/") # knit directory: Document
-system("latexdiff paper/PaperCEUS/PaperCEUS_rev1.tex paper/PaperCEUS/PaperCEUS_rev2.tex > paper/PaperCEUS/PaperCEUS_diff2.tex")
+system("latexdiff paper/PaperCEUS/PaperCEUS_rev1.tex paper/PaperCEUS/PaperCEUS_rev2_word.tex > paper/PaperCEUS/PaperCEUS_diff2.tex")
 #and change in tex the blue for another color (green? 0,0.5,0) and remove \uwave
@@
 
-Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we disaggregated the OD data into desire lines reprenting a maximum of 100 trips each.
+Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we disaggregated the OD data into desire lines representing a maximum of 100 trips each.
 Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.
@@
 Modeling the realistic cycling-PT routes between OD pairs depends on assumptions regarding the characteristics of the cycling and road networks and the location of public transport interfaces.
-Other constraints regarding the behavior of potential cyclists determine the routing results.
-For example, such restrictions can favor low speed, low traffic streets, more direct routes, and less steep paths, among others, which are suitable for cycling.
+Other constraints regarding the behaviour of potential cyclists determine the routing results.
+For example, such restrictions can favour low speed, low traffic streets, more direct routes, and less steep paths, among others, which are suitable for cycling.
 
@@
 All trips were considered to be made under urban conditions and at an average speed of 15 km/h during rush hour periods.
-Since the average distance traveled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
+Since the average distance travelled per trip influences the overconsumption and emissions from cold-start engine operation, we estimated energy and emission factors for different ranges of trips at 500-meter intervals.
 An equation was then used to calculate emission factors for the two types of fuel, for each type of pollutant, whose explanatory variables are driving speed ($speed$, in km/h) and average trip distance ($l_{trip}$, in km/trip).
@@
       col.names = c("Target", "Routing", "Total trips",
-                    "Baseline Cycling + PT", "Potencial Cycling + PT"),
-      caption = "\\label{summary1}Summary of the cycling potencial of the intermodality scenario. Values in 'trips/day'.",
+                    "Baseline Cycling + PT", "Potential Cycling + PT"),
+      caption = "\\label{summary1}Summary of the cycling potential of the intermodality scenario. Values in 'trips/day'.",
       format="latex",
@@
                     "Avoided CO2eq (ton/yr)", "Environmental benefits (k€/yr)"),
-      caption = "\\label{summary1b}Summary of the cycling potencial of intermodality scenario and its socio-environmental benefits for the cycling legs.",
+      caption = "\\label{summary1b}Summary of the cycling potential of intermodality scenario and its socio-environmental benefits for the cycling legs.",
       format="latex",
@@
                     "CO", "PM10", "NOx", "VOC", "Value (k€)"),
-      caption = "\\label{summary22}Summary of the avoided emmissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
+      caption = "\\label{summary22}Summary of the avoided emissions (ton/year) and corresponding monetization (thousand €) by replacing car trips with PT, in the second leg.",
       format="latex",
@@
                     "SE Benefits 1yr (k€)", "SE Benefits 10yrs (k€)"),
-      caption = "\\label{summaryall}Summary of the avoided CO2eq emmissions (ton/year) and the estimated social and environmental benefits (monetized in thousand €) by replacing car trips with cycling in combination with PT.",
+      caption = "\\label{summaryall}Summary of the avoided CO2eq emissions (ton/year) and the estimated social and environmental benefits (monetized in thousand €) by replacing car trips with cycling in combination with PT.",
       format="latex",
@@
 For the PT segment, we did not estimate the social impacts from substituting car trips. 
-One of the main socio-environmetal benefits, valued after monetization, comes from the increase in physical activity [@Felix2023ES].
+One of the main socio-environmental benefits, valued after monetization, comes from the increase in physical activity [@Felix2023ES].
 Although there are also social benefits form shifting car trips to PT, its health benefits would not be as high as shifting to cycling. 
@@
 Moreover, for the second trip segment, the shift from car to public transport contributes to avoiding up to 20,500 tons of CO~2~eq emissions annually.
-These changes are estimated to yield socio-environmental benefits totaling €125 million to €325 million over a decade.
+These changes are estimated to yield socio-environmental benefits totalling €125 million to €325 million over a decade.
 The quantification of such benefits can support policy-makers in prioritizing interventions to reduce the reliance on private motorized modes of transportation.

### 2024-11-21 paper/PaperTRA/PaperTRA_rev2.Rmd
add proper basemap caption on TRA Springer paper

--- 2023-12-21 paper/PaperTRA/PaperTRA_rev1.Rmd
@@
 
-```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area", fig.align='center', fig.pos="h"}
+```{r map1, out.width="60%", fig.cap="Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area.
+Author created map, using the mapview R package, with OpenStreetMap CARTO basemap.", fig.align='center', fig.pos="h"}
 knitr::include_graphics("img/map1.png", error = FALSE)
@@
 This work is part of the research activity carried out at Civil Engineering Research and Innovation for Sustainability (CERIS) and has been funded by Fundação para a Ciência e a Tecnologia (FCT), Portugal in the framework of project UIDB/04625/2020.
-The authors thank Thomas Götschi (HEAT for Cycling) for providing access to HaaS tool, which is under development.
+<!-- The authors thank Thomas Götschi (HEAT for Cycling) for providing access to HaaS tool, which is under development. -->
 

### 2024-11-21 paper/PaperTRA/PaperTRA_rev2.tex
add proper basemap caption on TRA Springer paper

--- 2023-12-21 paper/PaperTRA/PaperTRA_rev1.tex
@@
 % Pandoc citation processing
+%From Pandoc 3.1.8
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
@@
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\section{Methods}\label{methods}
 
-\hypertarget{modeling-origin-destination-trips}{%
 \subsection{Modeling Origin-Destination
-trips}\label{modeling-origin-destination-trips}}
+trips}\label{modeling-origin-destination-trips}
 
@@
 
-\caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
+\caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area. Author created map, using the mapview R package, with OpenStreetMap CARTO basemap.}\label{fig:map1}
 \end{figure}
@@
 
-\hypertarget{assessing-socio-environmental-benefits}{%
 \subsection{Assessing socio-environmental
-benefits}\label{assessing-socio-environmental-benefits}}
+benefits}\label{assessing-socio-environmental-benefits}
 
@@
 
-\hypertarget{results-and-discussion}{%
-\section{Results and Discussion}\label{results-and-discussion}}
+\section{Results and Discussion}\label{results-and-discussion}
 
@@
 
-\hypertarget{acknowledgements.}{%
-\subsubsection*{Acknowledgements.}\label{acknowledgements.}}
+\subsubsection*{Acknowledgements.}\label{acknowledgements.}
 \addcontentsline{toc}{subsubsection}{Acknowledgements.}
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-MARTENS2007326}{}}%
+\bibitem[\citeproctext]{ref-MARTENS2007326}
 \CSLLeftMargin{1. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-RIETVELD200071}{}}%
+\bibitem[\citeproctext]{ref-RIETVELD200071}
 \CSLLeftMargin{2. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
+\bibitem[\citeproctext]{ref-IMOB}
 \CSLLeftMargin{3. }%
 \CSLRightInline{INE:
-\href{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt}{Mobilidade
+\href{https://www.ine.pt/xportal/xmain?xpid=INE&xpgid=ine_publicacoes&PUBLICACOESpub_boui=349495406&PUBLICACOESmodo=2&xlang=pt}{Mobilidade
 e funcionalidade do território nas {Áreas Metropolitanas do Porto e de
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
+\bibitem[\citeproctext]{ref-felix2023}
 \CSLLeftMargin{4. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lovelace2017}{}}%
+\bibitem[\citeproctext]{ref-lovelace2017}
 \CSLLeftMargin{5. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
+\bibitem[\citeproctext]{ref-Lovelace2022Jittering}
 \CSLLeftMargin{6. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
+\bibitem[\citeproctext]{ref-r5r}
 \CSLLeftMargin{7. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-HEAT}{}}%
+\bibitem[\citeproctext]{ref-HEAT}
 \CSLLeftMargin{8. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-COPERT}{}}%
+\bibitem[\citeproctext]{ref-COPERT}
 \CSLLeftMargin{9. }%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-EuropeanCommission2014}{}}%
+\bibitem[\citeproctext]{ref-EuropeanCommission2014}
 \CSLLeftMargin{10. }%

### 2024-12-05 code/biclarwww/index.Rmd
insert video 2 and 3

@@
 
-### Parte II: Como usar o biclaR?
+### Parte II: Como usar o biclaR
 
-(*comming soon!*)
+<div align="center"><iframe width="1000" height="563" src="https://www.youtube.com/embed/PPZMcD0MrZI" title="YouTube video player" frameborder="0" data-external="1" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe></div>
+
+### Parte III: Como ler os ficheiros e fazer análises em SIG
+
+<div align="center"><iframe width="1000" height="563" src="https://www.youtube.com/embed/d1wu1Rpl5y8" title="YouTube video player" frameborder="0" data-external="1" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe></div>
 

### 2025-01-04 README.Rmd
Add links in readme and biclar.bib, close #53

@@
 output: github_document
-bibliography: refs.bib
-
+bibliography:
+  - refs.bib
+  - biclar.bib
 ---
@@
 
-The goal of `biclar` is to store code and (in the releases) data for estimating cycling potential and influencing policy.
+**biclar** is a tool for the design and assessment of different scenarios of the cycling network models in the Lisbon metropolitan area (LMA).
+The web application can be found at [u-shift.github.io/biclar](https://u-shift.github.io/biclarwww/) and on the [TML website](https://www.tmlmobilidade.pt/projetos-e-atividades/planos-e-estudos/rede-ciclavel-metropolitana-estudo-modelacao-e-ferramenta-de-apoio-ao-planeamento-e-decisao/).
 
-**biclar** is a tool for the design and assessment of different scenarios of the cycling network models in the Lisbon metropolitan area (LMA).
+**biclar** is an open source project, see the source code on [github.com/U-Shift/biclar](https://github.com/U-Shift/biclar).
+
+For a description of the methods and research underlying the project, see the [paper](https://www.sciencedirect.com/science/article/pii/S0198971524001595) *Reproducible methods for modeling combined public transport and cycling trips and associated benefits: Evidence from the biclaR tool" published in *Computers, Environment and Urban Systems* [@felix_reproducible_2025].
 

### 2025-01-06 README.Rmd
update scenarrios' maps and knit #53

--- 2022-07-17 README.Rmd
@@
 ```
+
 <!-- README.md is generated from README.Rmd. Please edit that file -->
@@
 
-# biclar
+# biclaR
 
@@
 
-The goal of `biclar` is to store code and (in the releases) data for estimating cycling potential and influencing policy.
+**biclaR** is a tool for the design and assessment of different scenarios of the cycling network models in the Lisbon metropolitan area (LMA).
+The web application can be found at [u-shift.github.io/biclarwww](https://u-shift.github.io/biclarwww/).
 
-**biclar** is a tool for the design and assessment of different scenarios of the cycling network models in the Lisbon metropolitan area (LMA).
+**biclaR** is an open source project, see the source code on [github.com/U-Shift/biclar](https://github.com/U-Shift/biclar).
+
+For a description of the methods and research underlying the project, see the [paper](https://www.sciencedirect.com/science/article/pii/S0198971524001595) "Reproducible methods for modeling combined public transport and cycling trips and associated benefits: Evidence from the biclaR tool" published in *Computers, Environment and Urban Systems* [@felix_reproducible_2025], and the methodological report [PT] on the [TML website](https://www.tmlmobilidade.pt/projetos-e-atividades/planos-e-estudos/rede-ciclavel-metropolitana-estudo-modelacao-e-ferramenta-de-apoio-ao-planeamento-e-decisao/).
 
@@
 
--   Trips dataset with Origin and Destination, at *Freguesia* level, disaggregated by transport mode, from @IMOB  
+-   Trips dataset with Origin and Destination, at *Freguesia* level, disaggregated by transport mode, from @IMOB\
 -   [CAOP 2020](https://www.dgterritorio.gov.pt/cartografia/cartografia-tematica/caop?language=en) - Official limits of Portuguese areas.
--   Road network from [OpenStreetMap](https://www.openstreetmap.org/#map=11/38.7053/-9.1585)  
--   Main public transport interfaces at Lisbon Metro Region, provided by [Transportes Metropolitanos de Lisboa](https://www.tmlmobilidade.pt/)
+-   Road network from [OpenStreetMap](https://www.openstreetmap.org/#map=11/38.7053/-9.1585)\
+-   Main public transport interfaces at Lisbon Metropolitan Area, provided by [Transportes Metropolitanos de Lisboa](https://www.tmlmobilidade.pt/)
 
@@
 
-The baseline scenario makes use of the 2018 mobility survey data in LMA.  
-We considered all trips between _Freguesias_.
+The baseline scenario makes use of the 2018 mobility survey data in LMA.\
+We considered all trips between *Freguesias*.
 
@@
 
-For the disagregation of OD pairs at _Freguesias_ level, we use [OD Jittering](https://github.com/atumworld/odrust) [@Jittering2022] method, which better suits walking and cycling trips modelling (shorter distances), instead of relying on centroids that concentrate all the trips between areas.
+For the disagregation of OD pairs at *Freguesias* level, we use [OD Jittering](https://github.com/atumworld/odrust) [@Jittering2022] method, which better suits walking and cycling trips modelling (shorter distances), instead of relying on centroids that concentrate all the trips between areas.
 
@@
 
-#### ENMAC Scenario
+#### ENMAC Scenario up to 5km
 
-See [here](region_enmac4_quiet.html) for full map.
+See [here](https://u-shift.github.io/biclarwww/aml/mapa_cenario1.html) for full map.
 
-```{r}
-# knitr::include_url("m_rnet_enmac_region_fastest.html")
-```
+<iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario1.html" width="800" height="400px" data-external="1">
 
-<iframe src="region_enmac4_quiet.html" width="800" height="400px" data-external="1">
 </iframe>
 
-See [here](articles/propotype.html) for results for each Municipality.
+See [here](https://u-shift.github.io/biclarwww/) for results for each Municipality.
 
-#### E-bike Scenario
+#### ENMAC Scenario up to 10km (E-bike)
 
-<!-- todo -->
+See [here](https://u-shift.github.io/biclarwww/aml/mapa_cenario2.html) for full map.
+
+<iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario2.html" width="800" height="400px" data-external="1">
+
+</iframe>
 
@@
 
-<!-- todo -->
+See [here](https://u-shift.github.io/biclarwww/aml/mapa_cenario3.html) for full map.
+
+<iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario3.html" width="800" height="400px" data-external="1">
+
+</iframe>
 
@@
 
-# References {-}
+# References {.unnumbered}

### 2025-01-06 README.Rmd
replace interactive by images in readme #53

@@
 
-<iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario1.html" width="800" height="400px" data-external="1">
+<!--iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario1.html" width="800" height="400px" data-external="1"-->
 
-</iframe>
+<!--/iframe-->
+
+![](images/clipboard-917840690.png)
 
@@
 
-<iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario2.html" width="800" height="400px" data-external="1">
+<!--iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario2.html" width="800" height="400px" data-external="1"-->
 
-</iframe>
+<!--/iframe-->
+
+![](images/clipboard-2091156617.png)
 
@@
 
-<iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario3.html" width="800" height="400px" data-external="1">
+<!--iframe src="https://u-shift.github.io/biclarwww/lisboa/mapa_cenario3.html" width="800" height="400px" data-external="1"-->
 
-</iframe>
+<!--/iframe-->
+
+![](images/clipboard-1852439731.png)
 

### 2025-02-13 code/biclarwww/index.Rmd
auxiliar files for render www

@@
       after_body: img/footer.html
+    # self_contained: false
 ---

### 2025-02-13 code/biclarwww/downloads.R

@@ -1,0 +1,455 @@
+library(dplyr)
+library(sf)
+
+for (i in municipios){
+  
+  # i = "barreiro"
+  mun = stringr::str_to_title(i)
+  # i = "vila franca de xira"
+  # mun = "Vila Franca de Xira" #problema com o De
+  
+  # cenario 1
+  mapa11r = sf::st_read("export_rnet/rnet_aml_1404_15.gpkg", quiet = TRUE)
+  mapa11r = mapa11r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+           quietness = as.integer(quietness)) 
+  mapa12r = sf::st_read("export_rnet/rnet_aml_1410_15.gpkg", quiet = TRUE)
+  mapa12r = mapa12r[MUNICIPIOSgeo %>% filter(Concelho == mun) %>% sf::st_buffer(300), , op = sf::st_within] %>% 
+    mutate(tCO2eq = tCO2eq10/10,
+           quietness = as.integer(quietness)) 
+  
+  mapa1r = mapa11r %>%
+    rename(ViagensTotal_base = Total,
+           Ciclistas_base = Bike,
+           Ciclistas_4 = cyc,
+           NovosCiclistas_4 = new_cyc,
+           Beneficios10_keuro_4 = Beneficios10_keuro,
+           tCO2eqanual_4 = tCO2eq) %>% 
+    select(-c(tCO2eq10, Bikeper)) %>% 
+    bind_cols(mapa12r %>% sf::st_drop_geometry()) %>% 
+    rename(Ciclistas_10 = cyc,
+           NovosCiclistas_10 = new_cyc,
+           Beneficios10_keuro_10 = Beneficios10_keuro,
+           tCO2eqanual_10 = tCO2eq) %>% 
+    select(-c(tCO2eq10, Bikeper, Total, Bike)) %>% 
+    select(-c(12:14)) %>% 
+    rename(tranquilidade = quietness...5,
+           carspeed = carspeed...6,
+           declive = slope...7) %>%
+    select(c(1:4, 8:13, 5:7)) %>% 
+    mutate(declive = as.numeric(gsub("%", "", declive)))
+  
+  mapa11l = sf::st_read("export_rnet/rnet_aml_1304_15.gpkg", quiet = TRUE)
[…]

### 2025-02-13 code/biclarwww/tabelas.R

@@ -1,0 +1,34 @@
+# result tables by mun (and aml)
+
+tabelas_mun = municipios_index %>% st_drop_geometry() %>%
+  select(-url) %>% 
+  dplyr::left_join(HEAT_municipios_PPP_redux) %>% 
+  dplyr::mutate(Economic10 = Economic10/1000,
+                CO2eq = as.numeric(gsub(" ","", CO2eq10))/10) %>%
+  dplyr::select(Municipio, Estrategia, LTS, Meta_ENMAC, Pop21, Total, Bike, Bike_new, CO2eq, Economic10) 
+saveRDS(tabelas_mun, "HEAT/tabelas_mun.Rds")
+
+tabelas_aml = HEAT_aml_redux %>% 
+  dplyr::mutate(Economic10 = Economic10/1000,
+                CO2eq = as.numeric(gsub(" ","", CO2eq10))/10) %>%
+  dplyr::select(Estrategia, LTS, Meta_ENMAC, Pop21, Total, Bike, Bike_new, CO2eq, Economic10) 
+saveRDS(tabelas_aml, "HEAT/tabelas_aml.Rds")
+
+
+
+
+
+
+  filter(Municipio == mun)
+
+
+  # dplyr::mutate_if(is.numeric, round, 1) %>%
+  DT::datatable(escape = -1, #para colocar os links, a contar com os rownames
+                filter = 'none',
+                rownames = FALSE,
+                colnames = c("Município", "População 2021", "Viagens /dia", "Viagens Bicicleta", "Novas Viagens Bicicleta 4%",
+                             "CO2eq evitado em 10 anos (ton)", "Benefícios estimados em 10 anos (milhares €)"),
+                options = list(dom = "t",
+                               pageLength = 18,
+                               responsive = TRUE,
+                               extensions = c('Responsive'))) %>% 
